---
title: "GreenWaste at a glance"
subtitle: "Preview of the case introduction for the simple dataset"
author: "3ie"
format:
  revealjs:
    theme: [theme_dge.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 192
---

```{r setup}
gw <- read.csv("evaluation_data_GreenWaste_simple.csv")
source("greenwaste_case.R")
ACCENT <- "#215a9e"; LIGHT <- "#c9d6e6"; MUTED <- "#545860"; ALERT <- "#B8272C"
f <- function(x) format(round(abs(x)), big.mark = ",")
FENCE <- strrep("`", 3)
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")
```

## GreenWaste at a glance

```{=html}
<style>
.case-grid { display: grid; grid-template-columns: 37% 63%; gap: 22px; margin-top: 0; }
.case-card .b { color: #215a9e; font-weight: 700; } .case-card .l { color: #7f97b5; font-weight: 700; }
.case-card { border: 1px solid #D5DEE8; border-radius: 6px; padding: 12px 16px 10px; background: #fff; }
.case-card h3 { font-size: 0.95em; margin: 0 0 2px; color: #063360; }
.case-card .how { font-size: 0.72em; line-height: 1.35; color: #545860; margin: 0 0 6px; }
.case-card img { width: 100%; display: block; }
.case-card .uses { font-size: 0.66em; line-height: 1.35; margin-top: 6px; color: #111418; }
.case-card .uses b { color: #215a9e; }
.case-foot { display: grid; grid-template-columns: 1fr 1fr 1fr; gap: 22px; margin-top: 16px; font-size: 0.68em; line-height: 1.35; }
.case-foot div { border-top: 3px solid #215a9e; padding-top: 6px; color: #111418; }
.case-foot div.rule { border-top-color: #B8272C; }
.case-foot .k { display: block; font-size: 0.85em; color: #545860; text-transform: uppercase; letter-spacing: 0.06em; }
</style>
```

```{r pilot-plot, include=FALSE}
png("case_pilot.png", width = 520, height = 330, res = 96, bg = "transparent")
par(mar = c(0.5, 0.5, 0.5, 0.5))
set.seed(4)
drawn <- sample(rep(c(TRUE, FALSE), 200))
xy <- expand.grid(x = 1:25, y = 16:1)
plot(xy$x, xy$y, pch = 15, cex = 2.1, col = ifelse(drawn, ACCENT, LIGHT),
     axes = FALSE, xlab = "", ylab = "", asp = 1)
dev.off()
```

```{r pool-plot, include=FALSE}
png("case_pool.png", width = 900, height = 330, res = 96, bg = "transparent")
par(mar = c(3.4, 0.5, 2.4, 0.5), family = "sans", col.axis = MUTED, fg = MUTED)
set.seed(5)
show <- pool[sample(nrow(pool), 1400), ]
y <- runif(nrow(show))
plot(show$score, y, pch = 16, cex = 0.75, col = ifelse(show$took_part == 1, ACCENT, LIGHT),
     axes = FALSE, xlab = "", ylab = "", xlim = c(20, 100), ylim = c(-0.02, 1.02))
abline(v = 58, col = ALERT, lwd = 3, lty = 2)
mtext("cut-off: 58", side = 3, at = 58, line = 0.6, col = ALERT, font = 2, cex = 1.15)
axis(1, at = seq(20, 100, 10), cex.axis = 1.1)
mtext("Efficiency score (higher = more efficient)", side = 1, line = 2.3, col = MUTED, cex = 1.05)
mtext("took part", side = 3, at = 39, line = 0.6, col = ACCENT, font = 2, cex = 1.1)
mtext("did not", side = 3, at = 79, line = 0.6, col = MUTED, font = 2, cex = 1.1)
dev.off()
```

```{r card-html, results='asis'}
img64 <- function(p) paste0("data:image/png;base64,", base64enc::base64encode(p))
html_out(sprintf('
<div class="case-grid">
  <div class="case-card">
    <h3>The pilot &middot; %s businesses</h3>
    <p class="how">A <b>lottery</b> decided who took part: <span class="b">%s did</span>, <span class="l">%s did not</span>.</p>
    <img src="%s" alt="400 squares, half blue">
    <p class="uses">Used for: <b>the randomised comparison</b></p>
  </div>
  <div class="case-card">
    <h3>The pool &middot; %s businesses</h3>
    <p class="how">A <b>rule</b> decided: every business scoring 58 or below took part (%s), none above (%s).</p>
    <img src="%s" alt="businesses along the efficiency score, blue below 58">
    <p class="uses">Used for: <b>before and after &middot; with and without &middot; difference-in-differences &middot; regression discontinuity &middot; matching</b></p>
  </div>
</div>
<div class="case-foot">
  <div><span class="k">What we measure</span>Waste-management costs per business, in AED, before and after the programme</div>
  <div><span class="k">What we know about each business</span>Efficiency score, manager age, staff, premises area, filtration</div>
  <div class="rule"><span class="k">The decision rule</span>Scale up only if costs fall by at least <b>%s AED</b></div>
</div>',
  format(nrow(pilot), big.mark = ","), sum(pilot$took_part), sum(pilot$took_part == 0), img64("case_pilot.png"),
  format(nrow(pool), big.mark = ","), format(sum(took), big.mark = ","), format(sum(!took), big.mark = ","),
  img64("case_pool.png"), f(RULE)))
```

::: {.notes}
About 3 minutes, the first time Module 2 opens GreenWaste. Two ways the
programme reached businesses: a lottery in the pilot, a rule in the wider pool.
Everything this week comes from these two groups. Ask: "Which group would you
trust more to tell you what the programme did, and why?" Keep the answer for
the randomised comparison.
:::

## One row, one business

```{r one-row, results='asis'}
cols <- c("business", "setting", "score", "took_part", "cost_before", "cost_after",
          "manager_age", "staff", "area", "filtration")
ex <- rbind(head(pilot[pilot$took_part == 1, cols], 1), head(pilot[pilot$took_part == 0, cols], 1),
            head(pool[pool$took_part == 1, cols], 1), head(pool[pool$took_part == 0, cols], 1))
cat(knitr::kable(ex, format = "html", row.names = FALSE, align = "llrrrrrrrr"), sep = "\n")
```

```{=html}
<div class="facts" style="margin-top:18px;">
  <div class="k">setting</div>     <div class="v">pilot (the lottery) or pool (the rule)</div>
  <div class="k">took_part</div>   <div class="v">1 if the business took part in GreenWaste</div>
  <div class="k">cost_before, cost_after</div> <div class="v">waste-management costs, AED, before and after</div>
  <div class="k">the last four</div> <div class="v">what we know about each business, used for matching</div>
</div>
```

::: {.notes}
Ten columns, one row per business. No other file is needed all week.
:::
