---
title: "The GreenWaste case"
subtitle: "Preview of the case introduction (simple dataset)"
author: "3ie"
format:
  revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    center: false
    width: 1280
    height: 720
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r setup}
gw <- read.csv("evaluation_data_GreenWaste_simple.csv")
source("greenwaste_case.R")
n  <- function(x) format(x, big.mark = ",")
av <- function(x) format(round(mean(x)), big.mark = ",")
FENCE <- strrep("`", 3)
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")
```


## The GreenWaste programme

```{=html}
<style>
.gw-cards { display: grid; grid-template-columns: repeat(3, 1fr); gap: 22px; margin-top: 10px; }
.gw-card { border-top: 5px solid #215a9e; background: #f4f7fb; padding: 16px 18px; border-radius: 0 0 6px 6px; }
.gw-card.rule { border-top-color: #B8272C; }
.gw-card .k { font-size: 0.55em; letter-spacing: 0.08em; text-transform: uppercase; color: #545860; }
.gw-card .v { font-size: 0.72em; line-height: 1.4; color: #111418; margin-top: 6px; }
.gw-card .big { font-size: 1.5em; font-weight: 700; color: #063360; line-height: 1.1; margin-top: 4px; }
.gw-card.rule .big { color: #B8272C; }

.lanes { display: grid; grid-template-columns: 220px 0.9fr 1.5fr 0.8fr; gap: 16px 16px; margin-top: 6px; align-items: stretch; }
.lanes .head { font-size: 0.58em; letter-spacing: 0.08em; text-transform: uppercase; color: #545860; padding-bottom: 2px; border-bottom: 2px solid #D5DEE8; }
.lanes .who { font-size: 0.78em; line-height: 1.3; color: #063360; font-weight: 700; padding: 10px 0; }
.lanes .who span { display: block; font-weight: 400; color: #545860; font-size: 0.85em; }
.lanes .step { font-size: 0.74em; line-height: 1.35; background: #f4f7fb; border-radius: 6px; padding: 14px 14px; position: relative; }
.lanes .step.key { background: #215a9e; color: #fff; }
.lanes .step.key b { color: #fff; }
.lanes .step b { color: #063360; }
.lanes .arrow::after { content: "\2192"; position: absolute; right: -14px; top: 40%; color: #7da1c4; font-size: 1.1em; }

.groups { display: grid; grid-template-columns: 230px 1fr 1fr; gap: 10px; margin-top: 8px; }
.groups .h { font-size: 0.62em; font-weight: 700; color: #063360; text-align: center; padding: 6px; border-bottom: 3px solid #215a9e; }
.groups .h.not { border-bottom-color: #7da1c4; }
.groups .row { font-size: 0.66em; font-weight: 700; color: #063360; padding: 12px 6px; line-height: 1.3; }
.groups .row span { display: block; font-weight: 400; color: #545860; font-size: 0.85em; }
.groups .cell { background: #f4f7fb; border-radius: 6px; padding: 10px 14px; text-align: center; }
.groups .cell.took { background: #e3ebf5; }
.groups .cell .count { font-size: 0.58em; color: #545860; }
.groups .cell .costs { font-size: 0.95em; font-weight: 700; color: #111418; margin-top: 2px; }
.groups .cell .costs small { font-size: 0.55em; font-weight: 400; color: #545860; display: block; }
</style>
```

```{r cards, results='asis'}
html_out(sprintf('
<div class="gw-cards">
  <div class="gw-card">
    <div class="k">What it does</div>
    <div class="big">Cuts waste costs</div>
    <div class="v">A subsidy for new waste equipment, technical help to install it, and training for staff.</div>
  </div>
  <div class="gw-card">
    <div class="k">Who it is for</div>
    <div class="big">Score 58 or below</div>
    <div class="v">Every business has an <b>efficiency score</b> from 0 to 100 (higher is more efficient). GreenWaste targets the less efficient ones.</div>
  </div>
  <div class="gw-card rule">
    <div class="k">The decision</div>
    <div class="big">At least %s AED</div>
    <div class="v">Scale up nationally only if costs fall by at least %s AED per business per year.</div>
  </div>
</div>', n(RULE), n(RULE)))
```

::: {.notes}
Invented case, used all week. Read the three cards; the decision rule on the
right is the bar every estimate is judged against.
:::

## How it reached businesses

```{r lanes, results='asis'}
html_out(sprintf('
<div class="lanes">
  <div class="head"></div><div class="head">Before</div><div class="head">The programme</div><div class="head">12 months later</div>

  <div class="who">Pilot district<span>%s businesses, all scoring 58 or below</span></div>
  <div class="step arrow">Waste costs measured for every business</div>
  <div class="step key arrow">A <b>lottery</b> picked <b>%s</b> to take part; the other %s waited</div>
  <div class="step">Costs measured again</div>

  <div class="who">Rest of the city<span>%s businesses, every score</span></div>
  <div class="step arrow">Waste costs measured for every business</div>
  <div class="step key arrow">A <b>rule</b>: every business scoring 58 or below took part (<b>%s</b>); the %s above did not</div>
  <div class="step">Costs measured again</div>
</div>
<p style="font-size:0.74em; color:#545860; margin-top:26px;">Same programme, same measurements, two ways of deciding who took part.</p>',
  n(nrow(pilot)), n(sum(pilot$took_part)), n(sum(pilot$took_part == 0)),
  n(nrow(city)), n(sum(took)), n(sum(!took))))
```

::: {.notes}
The pilot came first, in one district, to test the programme: a lottery decided.
In the rest of the city the programme was simply given to every business
scoring 58 or below. Ask: "In which of the two would you trust a simple
comparison of those who took part with those who did not? Why?"
:::

## Four groups of businesses

```{r groups, results='asis'}
cell <- function(d, cls) sprintf('<div class="cell %s"><div class="count">%s businesses</div><div class="costs">%s &rarr; %s<small>average waste costs, AED: before &rarr; after</small></div></div>',
                                 cls, n(nrow(d)), av(d$cost_before), av(d$cost_after))
html_out(paste0('
<div class="groups">
  <div></div><div class="h">Took part</div><div class="h not">Did not take part</div>
  <div class="row">Pilot district<span>a lottery decided</span></div>',
  cell(pilot[pilot$took_part == 1, ], "took"), cell(pilot[pilot$took_part == 0, ], ""), '
  <div class="row">Rest of the city<span>the score rule decided</span></div>',
  cell(city[took, ], "took"), cell(city[!took, ], ""), '
</div>'))
```

::: {.ask}
Every method this week compares businesses that took part with one of the other
groups. **Which group would you compare them with, and why?**
:::

::: {.notes}
This grid comes back at the start of every GreenWaste session, with that
session's comparison highlighted. Do not compute anything yet; the point is
that the four groups started from different places.
:::

## One row, one business

```{r one-row, results='asis'}
cols <- c("business", "setting", "score", "took_part", "cost_before", "cost_after",
          "manager_age", "staff", "area", "filtration")
ex <- rbind(head(pilot[pilot$took_part == 1, cols], 1), head(pilot[pilot$took_part == 0, cols], 1),
            head(city[city$took_part == 1, cols], 1), head(city[city$took_part == 0, cols], 1))
cat(knitr::kable(ex, format = "html", row.names = FALSE, align = "llrrrrrrrr"), sep = "\n")
```

```{=html}
<div class="facts" style="margin-top:18px;">
  <div class="k">setting</div>     <div class="v">pilot (the lottery) or city (the rule)</div>
  <div class="k">took_part</div>   <div class="v">1 if the business took part in GreenWaste</div>
  <div class="k">cost_before, cost_after</div> <div class="v">waste-management costs, AED, before and 12 months after</div>
  <div class="k">the last four</div> <div class="v">what else we know about each business</div>
</div>
```

::: {.notes}
Ten columns, one row per business. This is the only data file used all week.
:::
