---
title: "Was It Worth It? Reading a Cost-Benefit Result"
subtitle: "Module 2 · Day 3, Session 1: The Assumption That Decides the Answer"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html. Static knitr chunks work as before.
#
# One example all session: GreenWaste. The Abu Dhabi case slot of the earlier
# version is dropped (one example per session); the group exercise now runs
# scenarios on the GreenWaste model instead.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: []
  render-df: kable
  resources:
    - evaluation_data_GreenWaste.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"
PL_W <- 1100; PL_WC <- 520

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54) {
  p %>%
    layout(width = width, height = height,
           font = list(family = CHART_FONT, size = 13, color = INK),
           paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
           margin = list(l = 70, r = 24, t = t, b = b),
           hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                             font = list(color = INK, size = 12))) %>%
    config(displayModeBar = FALSE)
}
ax <- function(title = "", ...) {
  modifyList(list(title = list(text = title, font = list(size = 12, color = MUTED)),
                  gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
                  tickfont = list(size = 12, color = MUTED)), list(...))
}

gfmt  <- function(x) format(round(x), big.mark = ",")
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_d3s1.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0)); plot(qr_code(url, ecl = "H")); path
}
MENTI_QR <- menti_qr()

# ---- The GreenWaste cost-benefit model ---------------------------------------
# The saving is the difference-in-differences estimate from Day 2 (816 AED),
# recomputed here from the data so the deck cannot drift from it. It missed the
# week's 1,000 AED rule; using the rule as the saving would inflate the ratio.
gw <- read.csv("./evaluation_data_GreenWaste.csv")
tn <- subset(gw, treatment_neighborhood == 1)
did_fit <- lm(waste_management_costs ~ enrolled * round, data = tn)
SAVING     <- -coef(did_fit)[["enrolled:round"]]                 # 816
SAVING_CI  <- sort(-confint(did_fit)["enrolled:round", ])        # 753 to 880
ENROLLED   <- sum(tn$enrolled == 1 & tn$round == 0)              # 2,964
COST       <- 1800    # AED per business, one-off delivery cost (programme fact)
LAG        <- 1       # no saving in year 1: rollout and adoption take time
RATE_BASE  <- 0.05    # the authority's default discount rate
YEARS_BASE <- 5       # years of saving assumed by default
RULEBAR    <- 1000

# Present value of 1 AED a year for `years` years, starting after the lag.
pv <- function(r, years, lag = LAG) {
  t <- seq_len(lag + years)
  sum(ifelse(t <= lag, 0, 1) / (1 + r)^t)
}
bcr <- function(r = RATE_BASE, years = YEARS_BASE, saving = SAVING, cost = COST) {
  saving * pv(r, years) / cost
}
bcr_base  <- bcr()
pv_base   <- SAVING * pv(RATE_BASE, YEARS_BASE)
total_net <- (pv_base - COST) * ENROLLED
flip_rate <- uniroot(function(r) bcr(r) - 1, c(0.05, 1))$root          # about 24%
extra_flip <- pv_base - COST                                           # about 1,565
rate_grid  <- c(0.03, 0.05, 0.08, 0.12, 0.20)
years_grid <- c(2, 3, 5, 8, 10)
grid_tbl <- expand_grid(years = years_grid, r = rate_grid) %>%
  mutate(bcr = map2_dbl(r, years, ~ bcr(.x, .y)))
```

# 0 · Where the week has got to {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## From "how big" to "worth it"

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens.
gw <- read.csv("evaluation_data_GreenWaste.csv")
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)
clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)
COST <- 1800; SAVING <- 816; LAG <- 1

# Year by year: the saving, the discount factor, and what it is worth today.
# `extra` adds costs the headline left out; `decay` shrinks the saving each year.
savings_table <- function(rate, years, saving = SAVING, decay = 0, lag = LAG) {
  year <- seq_len(lag + years)
  s <- ifelse(year <= lag, 0, saving * (1 - decay / 100)^pmax(year - lag - 1, 0))
  f <- 1 / (1 + rate / 100)^year
  data.frame(year, saving = s, discount_factor = f, worth_today = s * f)
}
ratio <- function(tb, cost = COST) sum(tb$worth_today) / cost

# One line of arithmetic, then the payback chart.
calculator <- function(rate, years, ...) {
  tb <- savings_table(rate, years, ...)
  cat(sprintf("%d%% discount, %d years: savings worth %s AED today. Ratio = %s / 1,800 = %.2f\n",
              rate, years, format(round(sum(tb$worth_today)), big.mark = ","),
              format(round(sum(tb$worth_today)), big.mark = ","), ratio(tb)))
  payback_plot(tb)
}

# Running total of the savings (today's money) against what was spent.
payback_plot <- function(tb, cost = COST) {
  clean_par()
  run <- cumsum(tb$worth_today)
  b <- barplot(run, names.arg = tb$year, col = ifelse(run >= cost, ACCENT, ACCENT2),
               border = NA, ylim = c(0, max(cost, run) * 1.18),
               xlab = "Year", ylab = "AED, today's money")
  abline(h = cost, col = ALERT, lwd = 3)
  text(b[1], cost, paste("cost", format(cost, big.mark = ",")), col = ALERT, pos = 3, adj = 0)
}
```

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">three questions for any number; what a fair comparison is</div>
  <div class="k">Day 2</div>  <div class="v">three ways to estimate an effect: DiD, RDD, matching</div>
  <div class="k">Today</div>  <div class="v">what the effect is <em>worth</em>, and which assumption decides it</div>
</div>
```

::: {.ask}
A programme can have a real effect and still not be worth funding. Today the bar
is a ratio instead of a threshold.
:::

::: {.notes}
About 2 minutes. Open the deck a couple of minutes before the session starts: R and its statistics package take about a minute to load in the browser, after which every live cell runs in a few seconds.
:::

## Where the 816 AED comes from

```{webr}
offered <- subset(gw, treatment_neighborhood == 1)
m <- tapply(offered$waste_management_costs, list(offered$enrolled, offered$round), mean)
round(data.frame(before = m[, "0"], after = m[, "1"], change = m[, "1"] - m[, "0"],
                 row.names = c("did not enrol", "enrolled")))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
From Day 2: which two numbers do you subtract to get the saving?
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
The two changes: **−665 − 151 = −816 AED** a year per business, with a 95%
interval of about `r gfmt(SAVING_CI[1])` to `r gfmt(SAVING_CI[2])`. It missed the
1,000 AED rule. Today we ask a different question of the same number.
:::
:::::

::::::

# 1 · The headline {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The decision, and the ratio

:::::: {.columns}

::::: {.column width="54%"}
GreenWaste cost the authority **`r gfmt(COST)` AED per business** to deliver,
once. It saved each business about **`r gfmt(SAVING)` AED a year**.

A **benefit-cost ratio** compares the two: what you get back, divided by what you
spend, with future money counted as worth less than money today.
:::::

::::: {.column width="46%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Cost</div>          <div class="v">%s AED per business, one-off</div>
  <div class="k">Saving</div>        <div class="v">%s AED per business per year</div>
  <div class="k">Decision rule</div> <div class="v">a ratio above 1 pays for itself</div>
  <div class="k">Businesses</div>    <div class="v">%s enrolled</div>
</div>', gfmt(COST), gfmt(SAVING), gfmt(ENROLLED)))
```
:::::

::::::

## The headline number

```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Benefit-cost ratio, GreenWaste</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">AED spent per business</span></div>
  <div><span class="n">%s</span><span class="k">AED returned per business</span></div>
  <div><span class="n">%s</span><span class="k">AED net, whole programme</span></div>
</div>', bcr_base, gfmt(COST), gfmt(pv_base), gfmt(total_net)))
```

::: {.panel}
For every AED spent, about `r sprintf("%.2f", bcr_base)` AED comes back. This is
the programme that missed the Day 2 rule, and both statements are correct.
:::

::: {.warn}
The rule asks about the saving **in one year**. The ratio adds up the savings
**over several years** and sets them against a one-off cost. Different
questions can give different answers.
:::

## `[MENTI PLACEHOLDER]` Would you approve it?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
On a ratio of **`r sprintf("%.2f", bcr_base)`**, what would you do?

Approve &nbsp;·&nbsp; Reject &nbsp;·&nbsp; I need to know something else first
:::

Go to **menti.com**, code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='300px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.notes}
If most pick "something else first", ask two people what, write it up, and
come back to it at the three questions.
:::

## What went into that number

```{r}
tibble(
  Step = c("Cost per business", "Annual saving", "Years of saving assumed",
           "Discount rate assumed", "Savings, in today's money", "Benefit-cost ratio"),
  Value = c(sprintf("%s AED", gfmt(COST)), sprintf("%s AED", gfmt(SAVING)),
            "5, starting in year 2", "5% a year", sprintf("%s AED", gfmt(pv_base)),
            sprintf("%s / %s = %.2f", gfmt(pv_base), gfmt(COST), bcr_base)),
  Source = c("programme records", "measured, Day 2", "assumed", "assumed",
             "calculated", "calculated")
) %>%
  mutate(across(everything(), ~ ifelse(Source == "assumed",
                                        sprintf('<strong style="color:%s">%s</strong>', ALERT, .x), .x))) %>%
  kable(align = "lrl", escape = FALSE)
```

::: {.ask}
Two rows are neither costs nor savings. They are **choices** made by the analyst.
The next twenty minutes are about those two rows.
:::

::: {.notes}
Paper (Oct14_session1_materials.docx): Hand out "What went into the ratio" (pack page 1, one per group). Groups keep it: they paste the table into AI later.
:::

# 2 · Build the ratio yourself {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## A saving next year is worth less than one today

```{webr}
round(savings_table(rate = 5, years = 5), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Before I press Run: 816 AED saved in **year 4**. At 5% a year, what is it worth
in today's money? Guess a number.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
About 670 AED: 816 × 0.82. The **discount rate** says how fast future money loses
value. Year 1 shows 0 because the saving starts a year after rollout.
:::
:::::

::::::

## The calculator

:::::: {.columns}

::::: {.column width="50%"}
```{ojs}
//| echo: false
viewof rate  = Inputs.range([0, 40], {value: 5, step: 1, label: "Discount rate (%)"})
viewof years = Inputs.range([1, 10], {value: 5, step: 1, label: "Years the saving lasts"})
```
:::::

::::: {.column width="50%"}
::: {.panel}
Bars: savings so far, in today's money. It pays for itself once a bar reaches
the red line.
:::
:::::

::::::

```{webr}
#| input:
#|   - rate
#|   - years
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 2.9
calculator(rate, years)
```

::: {.notes}
Start at the base case (5%, 5 years): ratio 1.87. Do not move the sliders yet;
the next two slides move one at a time.
:::

## Choice 1: the discount rate, on its own

```{ojs}
//| echo: false
viewof rate1 = Inputs.range([0, 40], {value: 5, step: 1, label: "Discount rate (%)"})
```

```{webr}
#| input:
#|   - rate1
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 2.4
calculator(rate1, years = 5)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Agencies typically use 3% to 8%. **At what rate does the ratio fall below 1?**
Guess, then slide.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Around **`r sprintf("%.0f%%", 100 * flip_rate)`**. Across 3% to 8% the ratio only
moves from `r sprintf("%.2f", bcr(0.03))` to `r sprintf("%.2f", bcr(0.08))`. The
rate alone does not flip this decision.
:::
:::::

::::::

## Choice 2: how long the saving lasts

```{ojs}
//| echo: false
viewof years2 = Inputs.range([1, 10], {value: 5, step: 1, label: "Years the saving lasts (5% rate)"})
```

```{webr}
#| input:
#|   - years2
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 2
calculator(rate = 5, years = years2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
The Day 2 evaluations measured the saving for **one year**. Slide down to 2.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
At two years the ratio is **`r sprintf("%.2f", bcr(years = 2))`**: below
break-even. Below about 2.5 years the programme does not pay for itself, and
nothing in the evaluation measured how long the saving lasts.
:::
:::::

::::::

## Both choices together

```{r fig.height=3.1, fig.width=11.5}
grid_tbl %>%
  mutate(r_lab = factor(sprintf("%.0f%%", 100 * r), levels = sprintf("%.0f%%", 100 * rate_grid)),
         y_lab = factor(sprintf("%d years", years), levels = sprintf("%d years", rev(years_grid)))) %>%
  plot_ly(x = ~r_lab, y = ~y_lab, z = ~bcr, type = "heatmap",
          colorscale = list(c(0, ALERT), c(0.14, "#E3A6A8"), c(0.15, "#DCE7F2"), c(1, ACCENT)),
          zmin = 0.55, zmax = 3.4, showscale = FALSE,
          text = ~sprintf("%.2f", bcr), texttemplate = "%{text}",
          textfont = list(size = 16, color = INK),
          hovertemplate = "%{y}, discount %{x}: ratio %{z:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Discount rate", zeroline = FALSE),
         yaxis = ax("Years of saving assumed", zeroline = FALSE)) %>%
  pl_m2(height = 300)
```

::: {.ask}
Same programme, same cost, same measured saving in every cell; only the two
choices differ. **Find the row that is below 1 whatever the rate.**
:::

::: {.panel .fragment}
Two years. If the saving lasts two years, no discount rate makes this worth
doing. Read down the columns, not across the rows: the horizon matters far more
than the rate.
:::

# 3 · Three questions for any ratio {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## How to judge a cost-benefit claim

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Are the benefits plausible?</div>
  <div class="k">2</div> <div class="v">Are all the costs included?</div>
  <div class="k">3</div> <div class="v">What would flip the result?</div>
</div>
```

::: {.panel}
None of these needs a spreadsheet. They follow the pattern of the week: what is
the number compared to, what is missing, and what would change the answer.
:::

## Question 1: are the benefits plausible?

```{webr}
low  <- savings_table(5, 5, saving = 753)     # the interval end closest to zero
high <- savings_table(5, 5, saving = 880)
round(data.frame(saving = c(753, 816, 880),
                 ratio = c(ratio(low), ratio(savings_table(5, 5)), ratio(high))), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
The saving here is **measured**. The three rows carry its confidence interval
from Day 2 into the ratio.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The ratio stays between `r sprintf("%.2f", bcr(saving = SAVING_CI[1]))` and
`r sprintf("%.2f", bcr(saving = SAVING_CI[2]))`. The measured uncertainty barely
moves it. The horizon, which nobody measured, moves it from
`r sprintf("%.2f", bcr(years = 2))` to `r sprintf("%.2f", bcr_base)`.
:::
:::::

::::::

## The measured part and the assumed part

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Measured</div> <div class="v">the annual saving, %s AED (interval %s to %s): moves the ratio by %.2f</div>
  <div class="k">Measured</div> <div class="v">the cost, %s AED: a programme fact</div>
  <div class="k">Assumed</div>  <div class="v">how many years the saving lasts: moves the ratio by %.2f</div>
  <div class="k">Assumed</div>  <div class="v">the discount rate: across 3%% to 8%%, moves it by %.2f</div>
</div>',
  gfmt(SAVING), gfmt(SAVING_CI[1]), gfmt(SAVING_CI[2]),
  bcr(saving = SAVING_CI[2]) - bcr(saving = SAVING_CI[1]), gfmt(COST),
  bcr_base - bcr(years = 2), bcr(0.03) - bcr(0.08)))
```

::: {.ask}
The assumed lines move the answer more than the measured ones. The numbers that
were carefully estimated are not the ones deciding it.
:::

## Question 2: are all the costs included?

```{ojs}
//| echo: false
viewof extra = Inputs.range([0, 2000], {value: 0, step: 100, label: "Costs left out, per business (AED)"})
```

```{webr}
#| input:
#|   - extra
#| autorun: true
#| edit: false
tb <- savings_table(5, 5)
data.frame(cost_counted = 1800 + extra, ratio = round(ratio(tb, cost = 1800 + extra), 2))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
The 1,800 AED is the delivery cost only. Not counted: the authority's set-up and
administration, the businesses' own time, anything the programme displaced.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
How much would the missing costs have to add, per business, before the ratio
falls below 1?
:::

::: {.warn .fragment}
About **`r gfmt(round(extra_flip, -2))` AED**. Every cost left out makes the
ratio look better, so ask what is in the cost line.
:::
:::::

::::::

## Question 3: what would flip it?

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
For this programme it is not the discount rate. **If the saving lasts two years
instead of five, the ratio falls from `r sprintf("%.2f", bcr_base)` to
`r sprintf("%.2f", bcr(years = 2))`.**
:::

::: {.ask}
The question to put to the evaluator: **"How long do you assume the benefit
lasts, and what evidence do you have for that?"**
:::
:::::

::::: {.column width="46%"}
```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%.2f &nbsp;&rarr;&nbsp; %.2f</div>
  <div class="label">The same evidence, five years or two</div>
</div>', bcr_base, bcr(years = 2)))
```
:::::

::::::

## Your turn: stress-test the ratio

```{ojs}
//| echo: false
viewof s_rate  = Inputs.range([0, 20], {value: 5, step: 1, label: "Discount rate (%)"})
viewof s_years = Inputs.range([1, 10], {value: 5, step: 1, label: "Years the saving lasts"})
viewof s_decay = Inputs.range([0, 50], {value: 0, step: 5, label: "Saving shrinks each year by (%)"})
viewof s_extra = Inputs.range([0, 2000], {value: 0, step: 100, label: "Costs left out (AED)"})
```

```{webr}
#| input:
#|   - s_rate
#|   - s_years
#|   - s_decay
#|   - s_extra
#| autorun: true
#| edit: false
tb <- savings_table(s_rate, s_years, decay = s_decay)
data.frame(ratio = round(ratio(tb, cost = 1800 + s_extra), 2),
           verdict = ifelse(ratio(tb, cost = 1800 + s_extra) >= 1, "pays for itself", "does not pay"))
```

::: {.ask}
Each group takes one card and tells me the settings. Is your scenario plausible
for a waste-cost programme, and what is your verdict?
:::

::: {.notes}
Cards for the partner to print (one per group): (1) businesses drift back to old
habits, saving shrinks 30% a year; (2) set-up and admin add 600 AED per
business; (3) the ministry of finance insists on 8%; (4) the saving lasts three
years; (5) all of 1 and 2 together. About 10 minutes including report-back.

Paper (Oct14_session1_materials.docx): Hand out the scenario cards (pack page 2, cut, one card per group) and a recording sheet per group (page 3). Ratios for each card are in the facilitator key.
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: find the scenario that breaks it

:::::: {.columns}

::::: {.column width="42%"}
::: {.panel}
Groups: paste the cost-benefit table and this prompt.

**"Generate one plausible scenario in which this programme's benefit-cost ratio
falls below 1. Then tell me whether that scenario is realistic."**
:::

::: {.ask}
Check each claim against the calculator. Which is right, which is wrong, and
which is true but no use to you? Four minutes.
:::
:::::

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

The ratio falls below 1 if the discount rate is raised. [At a discount rate of
around 40%, the ratio drops below break-even]{.flag}, which would mean the
programme does not pay for itself and should not be funded.

[Alternatively, if the programme's benefits are delayed by a few years, the
ratio also falls below 1]{.flag}. Both scenarios are [realistic and commonly
observed in infrastructure appraisal]{.flag}.
:::
:::::

::::::

::: {.notes}
Paper (Oct14_session1_materials.docx): Groups paste the table from their "What went into the ratio" handout; both prompts are printed on it.
:::

## Debrief

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The arithmetic holds.** At 40% the ratio is below 1; in fact it crosses at
about `r sprintf("%.0f%%", 100 * flip_rate)`.
:::

::: {.warn}
**1 · It picked the easiest lever.** No agency uses a rate that high. The model
answered "name a scenario", not "name the one that matters".
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**2 · It came near the right lever and gave no number.** "Benefits delayed" is
close to the horizon, which decides this programme, but it never says how many
years or what the ratio becomes.
:::

::: {.warn}
**3 · "Commonly observed in infrastructure appraisal."** Unsourced, and about a
different kind of programme.
:::
:::::

::::::

::: {.panel}
The model invented no number and misread no table. It chose what was easy to
compute. Ask for the assumption that matters most, and ask for the number.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Generate one plausible scenario in which this benefit-cost ratio falls below 1.

"Plausible" is not defined, so the model may pick whatever is easiest to compute.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Here is a cost-benefit model. The cost and annual saving are measured; the
> discount rate and the number of benefit years are assumptions. Tell me which
> **single assumption** moves the ratio most, give the ratio under a realistic
> alternative, and say whether a government agency would normally make that
> assumption. **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both. Does the second rank the assumptions before choosing one? What did
each explain well?
:::

# 5 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Two tests, one programme

:::::: {.columns}

::::: {.column width="50%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Day 2 rule</div>     <div class="v">cut costs by 1,000 AED in a year</div>
  <div class="k">What was found</div> <div class="v">%s AED, real and short of the bar</div>
  <div class="k">Verdict</div>        <div class="v">do not scale on that evidence</div>
</div>', gfmt(SAVING)))
```
:::::

::::: {.column width="50%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Today\'s ratio</div>   <div class="v">%.2f AED back per AED spent</div>
  <div class="k">What it assumes</div> <div class="v">the saving lasts five years</div>
  <div class="k">Verdict</div>         <div class="v">worth funding, if that holds</div>
</div>', bcr_base))
```
:::::

::::::

::: {.ask}
Same programme, estimate and cost. The rule asks whether the effect is big
enough to scale; the ratio asks whether the spending pays for itself over time.
A report that gives only one of them answers half the question.
:::

::: {.warn}
The rule rests on a measurement. The ratio rests mostly on assumptions.
:::

::: {.notes}
Re-run the opening approval poll now ("On a ratio of 1.87, what would you do?")
and compare with the first vote. Ask someone who moved what moved them: it is
usually the horizon or the missing costs.
:::

## What today gives you

```{=html}
<div class="facts">
  <div class="k">A ratio is a model</div> <div class="v">it contains choices as well as evidence</div>
  <div class="k">The rate</div>           <div class="v">3% to 8% barely moves this one</div>
  <div class="k">The horizon</div>        <div class="v">decides it, and the evaluation did not measure it</div>
  <div class="k">The costs</div>          <div class="v">anything left out flatters the ratio</div>
</div>
```

::: {.ask}
The sentence to take to a meeting: **"What assumption would have to change for
this ratio to fall below 1, and what is your evidence for it?"**
:::

::: {.notes}
Paper (Oct14_session1_materials.docx): Hand out the take-away cards (pack page 4). Give a minute to fill in "On my desk" on the card, then take two or three examples from the room.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1</div>        <div class="v">is the number answering your question, and is it big enough?</div>
  <div class="k">Day 2</div>        <div class="v">which method produced it, and what does that method assume?</div>
  <div class="k">Today</div>        <div class="v">what is it worth, and which assumption decides that?</div>
  <div class="k">Next session</div> <div class="v">reading the charts that carry these claims into a meeting</div>
</div>
```
