---
title: "Reading a Regression Discontinuity Result"
subtitle: "Module 2 · Day 2, Session 2: What Happens at the Cut-off"
author: "3ie"
always_allow_html: yes
# live-revealjs (quarto-live): the baked plotly/kable charts still render at knit time
# exactly as before; the teaching code cells (Checking the data, Running the
# regression, Seeing the jump, Check 2 density, Check 3 change-the-window) are
# {webr} cells that run live in the browser via WebAssembly. Needs internet at
# presentation time to load webR. Export to PDF for a portable copy (live cells
# then show their code but do not run).
format:
  live-revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    # Single-file deck. The live (webR) cells work fine with this over http,
    # via `quarto preview` or a published URL, but NOT from a file:// path.
    # Do not open the rendered .html directly from disk; serve it or publish it.
    embed-resources: true
engine: knitr
resources:
  - evaluation_data_GreenWaste.csv
webr:
  packages:
    - dplyr
    - ggplot2
    - estimatr
    - broom
  cell-options:
    autorun: false
    edit: true
    warning: false
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / eligible side
ACCENT2 <- "#A9561F"   # series 2: comparison / other side
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000     # must cut costs by at least this many AED

# ---- The GreenWaste eligibility rule ---------------------------------------
# The programme was offered only to businesses whose efficiency index was at or
# BELOW 58. Read that again: the LESS efficient businesses qualified. This is
# the opposite direction to a "higher score wins" rule, and it is the first
# thing to get right when reading the chart.
#
# MEASURED, not assumed: the maximum index among eligible businesses is 57.9598
# and the minimum among ineligible is 58.0371, so the rule is genuinely sharp
# and the cut-off sits in between. Verified agreement: 1.000000.
gw <- read.csv("./evaluation_data_GreenWaste.csv")
CUT <- 58

# ---- Is the rule sharp, and is take-up full? -------------------------------
# The OFFER is a clean function of the index. TAKE-UP is not: only about half of
# the eligible businesses enrolled. That makes this a FUZZY design, and it is
# the distinction that decides whether the headline number is the effect on
# those offered or the effect on those who took it.
gw1 <- gw %>% filter(round == 1)

rule_sharp <- mean(as.integer(gw$efficiency_index <= CUT) == gw$eligible,
                   na.rm = TRUE)

# ---- The local comparison, at a chosen bandwidth ---------------------------
# Everything below is done at +/-2 index points. That is a teaching choice: wide
# enough to have power, narrow enough that the two sides are genuinely alike.
# The deck shows what happens when the bandwidth changes, because that choice
# is the one a commissioner can most easily be misled by.
BAND <- 2

win <- gw1 %>%
  filter(abs(efficiency_index - CUT) <= BAND) %>%
  mutate(below   = as.integer(efficiency_index <= CUT),   # the eligible side
         x       = efficiency_index - CUT)

rdd_fit <- lm_robust(waste_management_costs ~ below * x, data = win)
itt     <- coef(rdd_fit)[["below"]]
itt_ci  <- confint(rdd_fit)["below", ]
itt_p   <- rdd_fit$p.value[["below"]]

# Take-up difference at the cut-off: this is what turns the ITT into the effect
# on those who actually took the programme.
takeup <- mean(win$enrolled[win$below == 1]) - mean(win$enrolled[win$below == 0])
scaled <- itt / takeup

# The decision question. The rule asks the programme to AVOID at least RULEBAR.
# Sign trap: the coefficient is negative, so conf.low is the LARGER effect and
# abs(conf.low) is the OPTIMISTIC end. Sorting the absolute values makes the
# first element the end closest to zero, which is what the decision turns on.
itt_avoided       <- sort(abs(itt_ci))
itt_avoided_least <- itt_avoided[1]
itt_avoided_most  <- itt_avoided[2]
itt_clears        <- itt_avoided_least >= RULEBAR

# ---- What the two sides look like, either side of the line -----------------
side_means <- win %>%
  group_by(below) %>%
  summarise(n = n(), cost = mean(waste_management_costs),
            enrolled_share = mean(enrolled), .groups = "drop")

just_below <- side_means$cost[side_means$below == 1]   # eligible
just_above <- side_means$cost[side_means$below == 0]   # not eligible
raw_gap    <- just_below - just_above

# ---- Sensitivity: the estimate at other bandwidths -------------------------
# Built from the data so the slide cannot drift from the numbers.
band_tbl <- map_dfr(c(1, 2, 3, 5, 8), function(w) {
  d <- gw1 %>% filter(abs(efficiency_index - CUT) <= w) %>%
    mutate(below = as.integer(efficiency_index <= CUT), x = efficiency_index - CUT)
  f <- lm_robust(waste_management_costs ~ below * x, data = d)
  tk <- mean(d$enrolled[d$below == 1]) - mean(d$enrolled[d$below == 0])
  tibble(band = w, n = nrow(d),
         itt = coef(f)[["below"]],
         lo  = confint(f)["below", 1],
         hi  = confint(f)["below", 2],
         takeup = tk, scaled = coef(f)[["below"]] / tk)
})

# ---- Placebo cut-offs: the same test somewhere there is no rule ------------
# If the method reports a jump where no programme exists, its jump at the real
# cut-off is not evidence of anything.
placebo_tbl <- map_dfr(c(50, 54, 62, 66), function(cut) {
  d <- gw1 %>% filter(abs(efficiency_index - cut) <= BAND) %>%
    mutate(below = as.integer(efficiency_index <= cut), x = efficiency_index - cut)
  f <- lm_robust(waste_management_costs ~ below * x, data = d)
  tibble(cut = cut, n = nrow(d), jump = coef(f)[["below"]],
         lo = confint(f)["below", 1], hi = confint(f)["below", 2])
})

# ---- Density at the cut-off: did anyone bend the rule? ---------------------
dens_tbl <- tibble(
  bucket = c("56 to 57", "57 to 58", "58 to 59", "59 to 60"),
  n = c(sum(gw1$efficiency_index >= 56 & gw1$efficiency_index < 57, na.rm = TRUE),
        sum(gw1$efficiency_index >= 57 & gw1$efficiency_index < 58, na.rm = TRUE),
        sum(gw1$efficiency_index >= 58 & gw1$efficiency_index < 59, na.rm = TRUE),
        sum(gw1$efficiency_index >= 59 & gw1$efficiency_index < 60, na.rm = TRUE))
)

# ---- Covariates at the cut-off, at several bandwidths ----------------------
# A covariate that jumps at the cut-off means the two sides are not alike and
# the estimate is not credible. One of these appears to jump - but only at some
# bandwidths, which is the lesson this slide exists to teach.
cov_bands <- c(1, 2, 3, 5, 8)
cov_vars  <- c("recycling_center_distance", "staff_size", "age_manager",
               "business_area")
cov_labs  <- c(recycling_center_distance = "Distance to recycling centre (m)",
               staff_size = "Staff size",
               age_manager = "Manager age (years)",
               business_area = "Business area (units)")

cov_tbl <- map_dfr(cov_vars, function(v) {
  row <- tibble(Characteristic = cov_labs[[v]])
  for (w in cov_bands) {
    d <- gw1 %>% filter(abs(efficiency_index - CUT) <= w) %>%
      mutate(below = as.integer(efficiency_index <= CUT), x = efficiency_index - CUT)
    f <- lm_robust(as.formula(paste(v, "~ below * x")), data = d)
    sig <- confint(f)["below", 1] > 0 | confint(f)["below", 2] < 0
    row[[paste0("b", w)]] <- sprintf("%+.2f%s", coef(f)[["below"]],
                                     if (sig) "*" else "")
  }
  row
})

# ---- Who is near the line, versus everyone else ----------------------------
near_line <- gw1 %>% filter(abs(efficiency_index - CUT) <= BAND)
far_line  <- gw1 %>% filter(abs(efficiency_index - CUT) > BAND)
reach_tbl <- tibble(
  Group = c("Just either side of the cut-off", "Everybody else"),
  n = c(nrow(near_line), nrow(far_line)),
  staff = c(mean(near_line$staff_size), mean(far_line$staff_size)),
  area  = c(mean(near_line$business_area), mean(far_line$business_area)),
  cost  = c(mean(near_line$waste_management_costs),
            mean(far_line$waste_management_costs))
)

# ---- The table an evaluator would report ----------------------------------
rdd_tbl <- rdd_fit %>% tidy() %>%
  mutate(` ` = c("(Intercept)", "Just below the cut-off",
                 "Distance from the cut-off",
                 "Below x Distance"),
         CI = sprintf("[%s, %s]", gfmt(conf.low), gfmt(conf.high))) %>%
  select(` `, estimate, std.error, p.value, CI)
```

:::: {style="display:none;"}
```{webr}
#| autorun: true
#| edit: false
#| echo: false
#| output: false
#| warning: false
#| message: false
# ---- Live (webR) setup ----------------------------------------------------
# Runs once in the browser when the deck opens, priming the shared webR session
# so the teaching cells have their data and helpers ready even if the trainer
# skips one. Separate from the knitr `setup` chunk above, which prepares the
# baked charts and tables.
library(dplyr)
library(ggplot2)
library(estimatr)
library(broom)

gw  <- read.csv("evaluation_data_GreenWaste.csv")
gw1 <- gw %>% filter(round == 1)          # RDD uses the after-programme data
CUT <- 58

win <- gw1 %>%
  filter(abs(efficiency_index - CUT) <= 2) %>%
  mutate(x     = efficiency_index - CUT,
         below = as.integer(efficiency_index <= CUT),
         side  = ifelse(below == 1, "Eligible", "Not eligible"))
rdd_fit <- lm_robust(waste_management_costs ~ below * x, data = win)
top  <- coef(rdd_fit)[["(Intercept)"]]   # cost just above the line
jump <- coef(rdd_fit)[["below"]]         # the step at the line

# Average cost in quarter-point bins, for "Seeing the jump"
bins <- win %>%
  mutate(bin = CUT + floor(x * 4) / 4 + 0.125) %>%
  group_by(side, bin) %>%
  summarise(cost = mean(waste_management_costs), .groups = "drop")

# Businesses from 54 to 62, for the density check
near <- gw1 %>%
  filter(efficiency_index >= 54, efficiency_index < 62) %>%
  mutate(side = ifelse(efficiency_index <= CUT, "Eligible", "Not eligible"))

ACCENT  <- "#0E6E80"; ACCENT2 <- "#A9561F"; ALERT <- "#8E2413"
INK <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
side_cols <- c("Eligible" = ACCENT, "Not eligible" = ACCENT2)

theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(text = element_text(colour = INK),
          axis.title = element_text(colour = MUTED, size = rel(0.85)),
          axis.text = element_text(colour = MUTED),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
          legend.position = "bottom", legend.title = element_blank(),
          plot.margin = margin(4, 10, 4, 4))
}
```
::::

```{=html}
<style>
/* This deck: smaller live-code editors and output, so code plus result fits
   on a 1280x720 slide. The extension default is 1.3rem. */
.reveal .card.exercise-editor { --exercise-font-size: 1.1rem; margin: 0.3rem 0; }
.reveal .card.exercise-editor .card-header { padding: 0.25rem 0.75rem; }
.reveal .cell-output-container pre code,
.reveal .exercise-cell-output pre code { font-size: 0.9rem; line-height: 1.3; }
.reveal .cell-output-container pre { margin: 0.2rem 0; }
.reveal div.sourceCode pre code { font-size: 1.05rem; line-height: 1.35; }
.reveal .cell-output-display img,
.reveal .cell-output-display canvas { max-height: 280px; width: auto !important; }
</style>
```

# 1 · A different way to find a comparison

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1 S1</div> <div class="v">compared to what? &middot; the three questions</div>
  <div class="k">Day 1 S2</div> <div class="v">reading a descriptive output</div>
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons, then randomisation</div>
  <div class="k">Yesterday</div> <div class="v">difference-in-differences &middot; and an assumption that could not be tested</div>
  <div class="k">Today</div>     <div class="v">regression discontinuity &middot; a comparison the programme created</div>
</div>
```

::: {.ask}
Yesterday's method needed a comparison group that was heading the same way, and
we could not check whether it was.

Today's method does not need you to believe anything about a comparison group.
It builds one out of a rule.
:::

## The rule that decided who got in

:::::: {.columns}

::::: {.column width="54%"}
GreenWaste was not offered to everyone. It went to businesses whose
**efficiency index** was **at or below 58**: the less efficient ones.

Two businesses either side of that line are almost identical. One got the
programme, one did not. That is the whole idea.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Running variable</div> <div class="v">Efficiency index</div>
  <div class="k">Cut-off</div>         <div class="v">58</div>
  <div class="k">Eligible</div>        <div class="v">at or below 58</div>
  <div class="k">Rule followed?</div>  <div class="v">we check, next slide</div>
</div>
```
:::::

::::::

::: {.warn}
**Read the direction carefully.** The rule offers the programme to the **less**
efficient businesses. A "higher is better" rule would put the eligible side on
the right. Getting this backwards inverts every number that follows.
:::

## What RDD needs from the data {.dense}

:::::: {.columns}

::::: {.column width="44%"}
::: {.panel}
1. **A score** that decides eligibility
2. **A known cut-off**, and which side is eligible
3. **The outcome**, after the programme
4. **Who was offered it**, and **who took it up**
5. **Enough businesses near the line**
:::

::: {.ask}
Same habit as yesterday: ask AI *what data do I need, and does mine have it?*
before you ask for a result.
:::
:::::

::::: {.column width="56%"}
```{webr}
gw1 %>%
  mutate(side = ifelse(efficiency_index <= 58,
                       "At or below 58", "Above 58")) %>%
  group_by(side) %>%
  summarise(businesses = n(),
            offered    = mean(eligible),
            took_part  = mean(enrolled))
```

::: {.panel .fragment}
**Offered**: 1 on one side, 0 on the other. The rule was followed exactly.

**Took part**: only about half. Hold on to that.
:::
:::::

::::::

::: {.notes}
Press Run. Read the two rows aloud. "offered" is a share: 1 means every
business on that side was offered the programme, 0 means none were. So the
offer follows the rule perfectly (a sharp rule). "took_part" is the share that
enrolled: about 53% below the line and nobody above.

This is the setup for "The number that matters: who actually took it up" in
Section 2. Do not resolve it here.
:::

## `[MENTI PLACEHOLDER]` Which side is which?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
Businesses with an efficiency index **at or below 58** were offered the
programme.

On the chart we are about to draw, is the eligible side on the **left** or the
**right**?
:::

Left &nbsp;·&nbsp; Right &nbsp;·&nbsp; Depends how the axis is drawn

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel .fragment}
**Left.** A low index is on the left of a normal axis, and low is eligible.

Everyone who answered "right" has just demonstrated exactly the error this
session is about: a reader who assumes the programme went to the *good*
businesses will misread the entire chart, and the estimate's sign with it.
:::

## The picture the method is built on

```{r fig.height=3.2, fig.width=11.5}
# Binned means either side of the cut-off, plus the fitted lines.
binned <- gw1 %>%
  filter(between(efficiency_index, CUT - 10, CUT + 10)) %>%
  mutate(side = if_else(efficiency_index <= CUT, "Eligible", "Not eligible")) %>%
  group_by(side, bucket = round(efficiency_index)) %>%
  summarise(cost = mean(waste_management_costs),
            n = n(), .groups = "drop")

plot_ly() %>%
  add_trace(data = binned, x = ~bucket, y = ~cost, color = ~side,
            type = "scatter", mode = "markers",
            colors = c("Eligible" = ACCENT, "Not eligible" = ACCENT2),
            marker = list(size = ~pmax(8, pmin(18, n / 12))),
            text = ~sprintf("%d businesses", n),
            hovertemplate = "%{x} index: %{y:,.0f} AED<br>%{text}<extra></extra>") %>%
  add_segments(x = CUT, xend = CUT, y = 0, yend = ~max(cost),
               line = list(color = ALERT, width = 3, dash = "dot"),
               showlegend = FALSE, hoverinfo = "skip") %>%
  layout(xaxis = ax("Efficiency index  (eligible is at or below 58)",
                    zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2600))) %>%
  pl_m2(height = 310)
```

::: {.ask}
Hover the dots. Is there a step at the dotted line, the cut-off at 58?

Markers are sized by how many businesses sit in each bucket, so you can see
where the evidence is thicker.
:::

# 2 · From the picture to a number

## Regression discontinuity in one sentence

::: {.ask style="font-size: 120%;"}
Compare businesses **just below** the cut-off, who were offered the programme,
with businesses **just above** it, who were not.
:::

Right at the line, whether a business scores 57.9 or 58.1 is close to a matter
of chance. So the two sides are alike in every way except one: which side was
offered the programme.

::: {.panel .fragment}
The **step in costs at the cut-off** is the estimated effect. Everything in this
section is about measuring the height of that step.
:::

::: {.notes}
Say the sentence, then say it slowly again. The intuition to land: far from the
line the groups are different, but as you zoom in on the line they become as
good as randomly assigned. A business at 57.9 and a business at 58.1 are
near-twins who happened to fall on opposite sides of an administrative line.

Point back at the previous chart: the "step" is the gap between the two clouds
of dots right at the dotted line.
:::

## A first, rough number

:::::: {.columns}

::::: {.column width="48%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, eligible side vs just above the line</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">At or below 58</span></div>
  <div><span class="n">%s</span><span class="k">Just above 58</span></div>
  <div><span class="n">%d</span><span class="k">Within 2 points</span></div>
</div>', gfmt(raw_gap), gfmt(just_below), gfmt(just_above), nrow(win)))
```
:::::

::::: {.column width="52%"}
::: {.panel}
Take every business within two index points of the line, and average the costs
on each side. Two averages, one subtraction: `r gfmt(abs(raw_gap))` AED.

This is the same move as yesterday's two subtractions.
:::

::: {.warn .fragment}
It is a **starting point**, not the answer. Costs drift as the index rises, so
a plain average across the window mixes that drift into the gap. The regression
removes it by reading the step **exactly at the line**.
:::
:::::

::::::

## Implementing RDD: three steps

:::::: {.columns}

::::: {.column width="54%"}
```{r}
#| echo: true
#| eval: false
# Step 1. Keep only businesses near the line
#         (within 2 index points of 58).
# Step 2. Measure each score from the line,
#         so 0 = the cut-off, and mark the
#         eligible side.
win <- gw1 %>%
  filter(abs(efficiency_index - 58) <= 2) %>%
  mutate(x     = efficiency_index - 58,
         below = as.integer(efficiency_index <= 58))

# Step 3. Fit a line on each side and read
#         the step between them at 0.
rdd_fit <- lm_robust(
  waste_management_costs ~ below * x,
  data = win)
```
:::::

::::: {.column width="46%"}
::: {.panel}
- `x`: the score measured **from the line** (0 = 58)
- `below`: 1 offered, 0 not offered
- `below * x`: a line on each side; the **step between them at 0** is the
  effect
:::

::: {.warn}
The **2-point window** is the analyst's choice. Section 3 tests it.
:::
:::::

::::::

::: {.notes}
Do not dwell on syntax. The one idea that matters: measuring the score from 58
means the coefficient we care about is read exactly at the line, not at a score
of zero.

Note for questions: `below` is the OFFER, not whether the business took part.
That is deliberate, the rule decides the offer. Take-up comes two slides later.
:::

## Running the regression

:::::: {.columns}

::::: {.column width="55%"}
```{webr}
rdd_fit <- lm_robust(waste_management_costs ~ below * x,
                     data = win)

res <- tidy(rdd_fit)[, c("term", "estimate",
                         "std.error", "p.value")]
res[, 2:3]  <- round(res[, 2:3], 1)
res$p.value <- round(res$p.value, 3)
res
```
:::::

::::: {.column width="45%"}
::: {.ask}
Press **Run**. Only one row is the programme's effect. Which one?
:::

::: {.panel .fragment}
**Row 2, `below`: the jump, `r gfmt(itt)` AED.**

The other rows say where the lines sit and how they slope.
:::

::: {.warn .fragment}
**Trap:** `x`, about `r sprintf("%+.0f", coef(rdd_fit)[["x"]])` AED per point,
is the slope **along** the line, not the step **at** it.
:::
:::::

::::::

::: {.notes}
Run the chunk live, then take a vote across the four rows before clicking.
Many will pick `x` because it is large and significant, or the interaction
because it looks technical. That is the RDD version of yesterday's baseline-gap
trap.

Reminder for the room (vocab from Day 1 Session 1): the **coefficient** is the
effect size, here in AED; the **standard error** is how much the estimate would
move from sample to sample; the **95% confidence interval** is the range of
plausible true values, not a margin of certainty; the **p-value** is how
surprising this result would be if the programme had truly done nothing.

**The estimate is local.** It is what the programme did for businesses within
two index points of 58. Check 4 returns to this.
:::

## Seeing the jump: the red gap is row 2 {.scrollable}

```{webr}
#| fig-width: 11
#| fig-height: 3.1
# Press Run. The picture and the table are the same number.
ggplot(bins, aes(bin, cost, colour = side)) +
  geom_point(size = 3) +                          # average cost per bin
  geom_smooth(data = win, aes(efficiency_index, waste_management_costs),
              method = "lm", formula = y ~ x, se = FALSE) +  # a line per side
  annotate("segment", x = 58, xend = 58, y = top, yend = top + jump,
           colour = ALERT, linewidth = 2) +       # the jump, in red
  annotate("label", x = 58.5, y = top + jump / 2, colour = ALERT,
           label = sprintf("Jump = %.0f AED", jump)) +
  scale_colour_manual(values = side_cols) +
  labs(x = "Efficiency index", y = "Waste costs (AED)") + theme_m2()
```

::: {.notes}
Two lines, one per side, fitted to the businesses within two points of 58.

The dots are averages in quarter-point bins, so the room can see the data
behind each line. The y-axis is zoomed to 1,200 to 2,300 AED so the step is
visible; say so if anyone asks. The lines slope upward: more efficient
businesses (further right) have higher costs here, which is the `x` slope from
the previous slide, and exactly why the raw average overstated the gap.
:::

## The number that matters: who actually took it up

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
As the data check showed: just below the line, take-up is
**`r sprintf("%.0f", 100 * mean(win$enrolled[win$below == 1]))`%**; just above,
**`r sprintf("%.0f", 100 * mean(win$enrolled[win$below == 0]))`%**.

So the `r gfmt(abs(itt))` AED jump is the effect of being **offered** the
programme, not of taking it.
:::

::: {.warn .fragment}
Divide by the take-up difference: **`r gfmt(abs(scaled))` AED** per business
that **took** it, roughly twice the size.
:::
:::::

::::: {.column width="46%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED per business offered</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f%%</span><span class="k">Take-up below the line</span></div>
  <div><span class="n">%s</span><span class="k">AED per business that took it</span></div>
</div>',
  gfmt(abs(itt)),
  100 * mean(win$enrolled[win$below == 1]),
  gfmt(abs(scaled))))
```
:::::

::::::

::: {.ask .fragment}
**A decision, not a technicality.** A hundred offered with 48 taking it up is a
different proposition from a hundred taking it up. Which number goes in the
briefing?
:::

## Does it clear the rule?

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the estimated cut in costs near the cut-off</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">No</span><span class="k">Clears the bar</span></div>
</div>', gfmt(abs(itt)), gfmt(RULEBAR), gfmt(itt_avoided_least)))
```

::: {.warn}
`r gfmt(abs(itt))` AED is clearly not zero: a real effect. But the rule asks for
`r gfmt(RULEBAR)`, the interval runs from `r gfmt(itt_avoided_least)` to
`r gfmt(itt_avoided_most)`, and even the effect on those who took it up
(`r gfmt(abs(scaled))`) falls short. **It does not clear the bar.**
:::

::: {.ask .fragment}
Fourth session, same verdict. Day 1 S2, Day 1 S3, yesterday's DiD, and now this:
every result has been **real**, and none has been **enough**.

That is not a coincidence in the data. It is what evidence usually looks like
when it is asked to clear a bar set in advance.
:::

::: {.notes}
Take-up comes before this slide on purpose, so the verdict covers both numbers:
the effect of the offer (`r gfmt(abs(itt))`) and the effect on those who took
part (`r gfmt(abs(scaled))`). Neither reaches `r gfmt(RULEBAR)`.
:::

# 3 · Four things that must be checked

## This method's one assumption

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Nothing else jumps at the cut-off.</div>
</div>')
```

::: {.panel}
A business one point below 58 and one point above should be alike in every way
**except** whether it was offered the programme.

If something else also changes at 58, if the businesses differ or the rules
were bent, the jump is not the programme.
:::

::: {.ask}
It is the same shape as yesterday's assumption. Yesterday it was *the groups
were moving in parallel*. Today it is *nothing else changes at the line*.

Both are statements about a world you cannot see. Both have to be probed.
:::

## Check 1: does the jump survive a different window?

```{r fig.height=3, fig.width=11.5}
bt <- band_tbl %>%
  mutate(lab = sprintf("+/- %d", band)) %>%
  mutate(lab = factor(lab, levels = sprintf("+/- %d", sort(unique(band)))))

plot_ly() %>%
  add_segments(data = bt, x = ~abs(lo), xend = ~abs(hi), y = ~lab, yend = ~lab,
               line = list(color = ACCENT, width = 4), showlegend = FALSE,
               hovertemplate = "avoids %{x:,.0f}<extra></extra>") %>%
  add_markers(data = bt, x = ~abs(itt), y = ~lab,
              marker = list(color = ACCENT, size = 11), showlegend = FALSE,
              hovertemplate = "avoids %{x:,.0f}<extra></extra>") %>%
  add_segments(x = RULEBAR, xend = RULEBAR, y = -0.5, yend = 4.5,
               line = list(color = ALERT, width = 3),
               name = sprintf("The rule: %s", gfmt(RULEBAR)),
               hoverinfo = "name") %>%
  layout(xaxis = ax("AED avoided per business", range = c(0, 1250)),
         yaxis = ax("Bandwidth: index points either side of 58",
                    zeroline = FALSE)) %>%
  pl_m2(height = 290)
```

::: {.ask}
Hover the lines. The estimate moves as the window widens: roughly
`r gfmt(abs(band_tbl$itt[band_tbl$band == 2]))` at two points,
`r gfmt(abs(band_tbl$itt[band_tbl$band == 5]))` at five.

**Which window is the right one?** There is no rule. That choice is the
analyst's, and it is yours to interrogate.
:::

::: {.notes}
This is Step 1 of "Implementing RDD" coming back. Every row is the same
regression with a different window. The point to land: the estimate is never
anywhere near the 1,000 AED rule, whichever window you pick, so the verdict is
robust even though the number moves.
:::

## Check 1, continued: is there a jump where there is no rule?

:::::: {.columns}

::::: {.column width="56%"}
```{r}
placebo_tbl %>%
  mutate(across(c(jump, lo, hi), ~ round(.x, 0))) %>%
  kable(align = "lrrrr", digits = 0,
        col.names = c("Cut-off tested", "n", "Jump", "CI low", "CI high")) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="44%"}
::: {.panel}
The real cut-off is 58. The table tests **50, 54, 62 and 66**: places where no
rule exists, so there should be no jump.

Every confidence interval contains zero. Good news.
:::

::: {.warn}
If the method found a jump where there is no programme, its jump at 58 would be
worth nothing.
:::
:::::

::::::

## Check 2: did anyone bend the rule? {.scrollable}

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 3.3
# Businesses per one-point slice, 54 to 62
ggplot(near, aes(efficiency_index, fill = side)) +
  geom_histogram(binwidth = 1, boundary = 58,
                 colour = "white") +
  geom_vline(xintercept = 58, linetype = "dashed") +
  scale_fill_manual(values = side_cols) +
  labs(x = "Efficiency index", y = "Businesses") +
  theme_m2()
```
:::::

::::: {.column width="44%"}
::: {.panel}
If businesses could nudge their score below 58 to qualify, there would be a
**pile-up** just below the line.
:::

::: {.panel .fragment}
There is none: `r dens_tbl$n[2]` and `r dens_tbl$n[3]` businesses in the slices
either side of 58.
:::

::: {.warn .fragment}
**Ask anyway.** A pile-up is invisible unless you look, and it would bias the
estimate.
:::
:::::

::::::

::: {.notes}
Press Run. Each bar is the number of businesses in a one-point slice of the
index. Ask the room what a pile-up would look like before clicking: a tall bar
just left of the dashed line and a dip just right of it. The bars go up and
down a little everywhere; the question is whether the bar at the line is out of
pattern, and it is not.
:::

## Check 3: are the two sides alike?

This is the check that **does not fully pass**, and it is the most useful slide
in the session.

```{r}
cov_tbl %>%
  kable(align = "l", col.names = c("Characteristic",
        "+/- 1", "+/- 2", "+/- 3", "+/- 5", "+/- 8")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  footnote(general = "* marks a difference whose 95% confidence interval excludes zero.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.ask}
Look along the **top row** as the window widens:
`r cov_tbl$b1[1]`, then `r cov_tbl$b2[1]`, `r cov_tbl$b3[1]`,
`r cov_tbl$b5[1]`, `r cov_tbl$b8[1]`.

What do you notice about when the star appears and when it goes?
:::

::: {.notes}
Each cell runs the same RDD regression, but with a business characteristic as
the outcome instead of costs. If the two sides are alike, there should be no
jump in any of them. The next slide runs the top row live.
:::

## Check 3, live: change the window

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
window <- 2    # try 1, 3 and 5

d <- gw1 %>%
  filter(abs(efficiency_index - CUT) <= window) %>%
  mutate(x     = efficiency_index - CUT,
         below = as.integer(efficiency_index <= CUT))

fit <- lm_robust(recycling_center_distance ~ below * x,
                 data = d)
tidy(fit)[2, c("term", "estimate",
               "conf.low", "conf.high")]
```
:::::

::::: {.column width="42%"}
::: {.panel}
The same regression as before, with **distance to a recycling centre** as the
outcome instead of costs.

If the whole interval sits below zero, the two sides differ.
:::

::: {.ask}
Press **Run** at 2. Then change the window to **1**, run again, then **5**.
Does the gap stay?
:::
:::::

::::::

::: {.notes}
Expected results: at 2 the interval is roughly -20 to -2 (a "difference");
at 1 it is roughly -22 to +6 (no difference); at 5 roughly -3 to +8 (no
difference). The editor is live: click into the cell, change the 2, press Run.
If the live cell will not start (no internet), the table on the previous slide
shows the same numbers.
:::

## What that row means

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**At +/-2 and +/-3 the distance to a recycling centre appears to jump.**

Taken at face value, that says the two sides are not alike, and the estimate
would not be credible.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
But it is absent at +/-1, and absent again at +/-5 and +/-8.

A real discontinuity does not appear at two windows and vanish at the
neighbouring ones. That pattern is the signature of a **specification
artefact**: a result that depends on an analyst's choice rather than on the
data.
:::
:::::

::::::

::: {.ask .fragment}
So the honest reading is: **there is no strong evidence of a jump in the
covariates, but the check is not clean either.**

Which is a different claim from "the check passed", and a different claim from
"the estimate is wrong". Hold all three apart.
:::

## Check 4: who does the result apply to?

:::::: {.columns}

::::: {.column width="52%"}
```{r}
reach_tbl %>%
  mutate(across(c(staff, area, cost), ~ round(.x, 2))) %>%
  kable(align = "lrrr", col.names = c("", "n", "Staff", "Area", "Cost (AED)")) %>%
  kable_styling(full_width = FALSE, font_size = 20)
```
:::::

::::: {.column width="48%"}
::: {.panel .fragment}
The businesses near the line are a particular group: smaller staff
(`r sprintf("%.1f", reach_tbl$staff[1])` against
`r sprintf("%.1f", reach_tbl$staff[2])`) and smaller premises.

The estimate describes **them**.
:::

::: {.warn .fragment}
It does **not** describe the large, efficient businesses far from the line:
the single most common misreading of an RDD result.
:::
:::::

::::::

::: {.ask .fragment}
This is an estimate for businesses *near a line*. If your plan is to extend the
programme to everyone, what have you actually learned?
:::

# 4 · Put AI to work

## AI Snapshot: who does this apply to?

::: {.panel}
Paste the table and the four checks into an AI tool with this prompt:

**"Explain, without jargon, who this result actually applies to."**
:::

::: {.ai-output}
**AI response:**

This study found that the GreenWaste programme reduced costs by around
**249 AED** for participating businesses. [Because the study uses a
cut-off at an efficiency index of 58, the findings apply to all businesses in
the programme]{.flag}.

[The cut-off design means the result generalises to the wider population of
inefficient businesses]{.flag}. The estimate is therefore a reasonable basis for
[scaling the programme to every business below average efficiency]{.flag}.
:::

::: {.ask}
Mark every claim the study supports. Four minutes.
:::

## Debrief: one error, repeated three ways {.dense}

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "Applies to all businesses in the programme."**

The estimate is local to businesses near the line. It says nothing about those
far from it, whether they were in the programme or not.
:::

::: {.warn}
**2 · "Generalises to the wider population of inefficient businesses."**

Backwards. RDD is **least** able to generalise, precisely because it uses only
the businesses closest to the line.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "A reasonable basis for scaling to every business."**

A scale-up decision needs exactly the thing the model has discarded: what
happens away from the cut-off.
:::
:::::

::::::

::: {.panel}
**The pattern to notice.** The model made the same mistake three times in
different words, and each restatement sounded more confident than the last.

The question it was asked, *who does this apply to?*, is the one it answered
least well. A narrow question about scope is where a fluent model is at its
most dangerous, because a broad answer reads better than a careful one.
:::

## The same question, a better prompt {.dense}

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain, without jargon, who this result actually applies to.

Nothing about the design, so the model reaches for the most natural-sounding
story: the programme, everyone in it, and what to do next.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> This is a regression discontinuity result with the cut-off at an efficiency
> index of 58, estimated within 2 points either side. The businesses near the
> line are smaller than average. Name precisely which businesses the estimate
> describes, and state what it cannot tell us. **Ask me questions before you
> answer.**
:::
:::::

::::::

::: {.ask}
Give the model the design and the local nature of the estimate, then run both.
Does error 1 survive? Does error 2?

The habit: **supply the constraint the model cannot infer, then ask a narrower
question.**
:::

# 5 · The Abu Dhabi case

## `[PLACEHOLDER]` The DCD social protection evaluation

::: {.warn}
**Awaiting the real findings.** The Abu Dhabi Department of Community
Development social protection evaluation has not been shared with this workspace
yet, so nothing here is presented as its result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>      <div class="v">[PLACEHOLDER]</div>
  <div class="k">Running variable</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Cut-off</div>        <div class="v">[PLACEHOLDER]</div>
  <div class="k">Outcome</div>        <div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed. When the DCD findings arrive they drop into these four
fields and the two slides that follow, without redesigning the session.
:::

## Four questions for the DCD result

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, and which side is eligible?</div>
  <div class="k">2</div> <div class="v">Is the rule sharp, and if not, what is take-up on the eligible side?</div>
  <div class="k">3</div> <div class="v">Is the density smooth at the cut-off, or did people bend the rule?</div>
  <div class="k">4</div> <div class="v">Who sits near the line, and is that who you want to help?</div>
</div>
```

::: {.ask}
The same four checks we just ran on GreenWaste, in the same order.

The programme changes; the interrogation does not. That transferable habit is
what you are here for.
:::

# 6 · Close

## Four questions to ask any evaluator presenting RDD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, which side is eligible, and is the rule sharp or fuzzy?</div>
  <div class="k">2</div> <div class="v">Does the jump survive at other bandwidths, and are there placebo cut-offs that show nothing?</div>
  <div class="k">3</div> <div class="v">Is the density smooth at the line, and are the covariates balanced either side?</div>
  <div class="k">4</div> <div class="v">Who is near the line, and does the estimate clear our rule for them?</div>
</div>
```

::: {.panel}
Question 4 has two halves, and both matter. **Who it applies to** is the RDD
question. **Does it clear the rule** is the question every session this week has
ended on.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1 S2</div> <div class="v">descriptive reading &middot; -1.13 against a 2.0 rule</div>
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons and an RCT</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences &middot; -816 against a 1,000 rule</div>
  <div class="k">Today</div>  <div class="v">regression discontinuity &middot; -249 against a 1,000 rule, and local</div>
</div>
```

::: {.ask}
Notice how each method buys its credibility, and what it pays:

**DiD** buys a comparison group by assuming parallel trends.
**RDD** buys credibility from the rule, and pays with **reach**: a tight
estimate about a narrow group.

Next session does the same trade a third time, with **matching**: a much wider
reach, bought by assuming that everything that matters was measured.
:::
