---
title: "GreenWaste: reading an evaluation result"
format:
  html:
    embed-resources: true
    toc: false
    html-math-method: plain
    theme: none
    include-in-header:
      text: |
        <style>
        body{font:16px/1.45 Arial,sans-serif;color:#111418;margin:0;background:#eef2f6}
        main{max-width:850px;margin:auto}h1{font-size:26px;color:#063360;margin:12px 0}h2{font-size:22px;color:#063360;margin:0 0 12px}h3{font-size:18px;margin:14px 0 6px}
        .sheet-page{background:white;padding:28px 40px;margin:20px 0;box-sizing:border-box;min-height:950px}p{margin:8px 0}.caption{font-size:13px;color:#545860}.reading-table{border-collapse:collapse;width:100%;font-size:14px}.reading-table td,.reading-table th{padding:8px;border-bottom:1px solid #D5DEE8;text-align:left}.reading-table thead{background:#063360;color:white}.writing{height:44px;border-bottom:1px solid #c8d6e5;margin:6px 0}.writing.tall{height:75px}.costs{font-size:19px;word-spacing:9px;border-top:2px solid #215a9e;padding:12px 0}.ai-source{border-left:3px solid #7da1c4;padding-left:16px}.prompt{border-top:2px solid #215a9e;padding-top:8px;font-size:15px}.source-note{font-size:14px}.writing.part{height:30px;margin:2px 0}.table-title{margin:10px 0 4px}
        header button{font:inherit;margin:12px 0;padding:8px 15px;border:1px solid #215a9e;background:white;color:#063360;cursor:pointer}
        @page{size:A4;margin:14mm}@media print{body{background:white;font-size:11pt;line-height:1.35}main{max-width:none}.sheet-page{padding:0;margin:0;min-height:0;break-after:page;box-shadow:none}.sheet-page:last-child{break-after:auto}header{display:none}h1{font-size:19pt}h2{font-size:17pt}h3{font-size:12pt}.reading-table{font-size:10pt}.writing{height:10mm}.writing.tall{height:17mm}.writing.part{height:8mm}.ai-source{font-size:11pt}.prompt{font-size:10pt}#title-block-header{display:none}}
        </style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('session1_review.R', encoding='UTF-8')
```

<header><button onclick="window.print()">Print worksheet</button></header>

<div class="sheet-page">

## Page 1: The before-and-after table

GreenWaste subsidises waste equipment and provides installation help and staff training. In the pilot district, a lottery chose which businesses took part. In the rest of the city, businesses with an efficiency score of 58 or below took part. In this session we read results about each business's annual waste management cost, in AED. GreenWaste is a fictional training case.

### H1-A: The before-and-after table for city participant businesses

The same 4,794 participant businesses were measured before GreenWaste and 12 months after. This table does not include any comparison businesses that did not receive GreenWaste. Our rule is a reduction in annual waste management cost of at least 1,000 AED per business per year. Meeting this rule does not by itself show that GreenWaste is good value for money.

```{r, results='asis'}
html_out(review_table())
```

### Q1 | Mean and what it hides

These ten businesses were selected from the GreenWaste data. The values below are their annual waste management cost before GreenWaste, in AED. They are not representative of all city businesses.

```{r, results='asis'}
html_out(paste0('<div class="costs">',paste(fmt(review_costs),collapse=' · '),'</div>'))
```

(a) Calculate the mean annual waste management cost of the ten businesses.

<div class="writing part"></div>

(b) Calculate the mean and the median again after removing the business with the largest cost (4,306 AED).

<div class="writing part"></div>

(c) Which changes more, the mean or the median, and why?

<div class="writing part"></div>

### Q2 | Treatment effect

The costs of the participant businesses fell after GreenWaste.

(a) Does this tell us what would have happened to these businesses without GreenWaste?

<div class="writing part"></div>

(b) Explain what information is missing. This missing information is called the counterfactual.

<div class="writing part"></div>

### Q3 | Coefficient and sign

Use H1-A.

(a) Mark the row that shows the change in cost.

(b) Write its value, its units, what it compares, and whether it is positive or negative.

<div class="writing part"></div>

(c) Use both rows of the table to work out the mean annual waste management cost 12 months after GreenWaste.

<div class="writing part"></div>

</div>

<div class="sheet-page">

## Page 2: Uncertainty and your own interpretation

Keep page 1 open, because you need H1-A (the before-and-after table). Answer each question when we reach that idea in the session.

### Q4 | Confidence interval

(a) Draw a box around the 95% interval in the change row of H1-A.

(b) What does this interval describe?

<div class="writing part"></div>

(c) Does it show how much each individual business saved? Explain.

<div class="writing part"></div>

Now compare three invented estimates of the annual reduction in cost. Each estimate is 1,100 AED. Their 95% intervals are A: 1,080 to 1,120 AED, B: 900 to 1,300 AED, and C: 400 to 1,800 AED.

(d) Which interval is entirely above our rule of a reduction of at least 1,000 AED?

<div class="writing part"></div>

(e) What is still uncertain for B and C?

<div class="writing part"></div>

### Q5 | P-value

(a) Find the p-value in the change row of H1-A.

(b) What does the p-value assume about the change?

<div class="writing part"></div>

(c) Does the p-value show that GreenWaste caused the fall, or the probability that the programme works? Explain.

<div class="writing part"></div>

Now compare two invented results. Result A is a reduction of 50 AED (interval 30 to 70 AED, p < 0.001). Result B is a reduction of 1,100 AED (interval -100 to 2,300 AED, p = 0.072).

(d) Which result is statistically significant?

<div class="writing part"></div>

(e) Which result could be large enough to matter for our rule?

<div class="writing part"></div>

(f) Does Result B prove that there is no reduction? Explain.

<div class="writing part"></div>

### Q6 | Pair check and reading repair

(a) Compare your answers to Q1 to Q5 with a partner. Correct one explanation if it needs it.

(b) Explain what the estimate and the interval in H1-A show, and compare them with our rule of a reduction of at least 1,000 AED per business per year.

<div class="writing part"></div>

(c) Explain why this table cannot show that GreenWaste caused the change.

<div class="writing part"></div>

### My own reading and evidence request

Close the AI responses and your notes on the terms, and keep H1-A open. Write three sentences. (1) What the table shows, with the units and the period. (2) What the table cannot show. (3) One piece of evidence you would ask for, and why.

<div class="writing tall"></div>

</div>

<div class="sheet-page">

## Page 3: AI exercise

### Part A: Mark the response

Naive prompt (a prompt with no context, as a busy person might type it): "Explain H1-A and advise whether to expand GreenWaste." This response was written for the exercise. It is not a real AI answer. Underline the claims that H1-A supports. Cross out the claims that H1-A does not support.

```{r, results='asis'}
html_out(paste0('<div class="ai-source">',review_paragraphs(review_ai),'</div>'))
```

### Part B: Run the improved prompt, then compare

Copy the prompt and the H1-A table into Microsoft Copilot, and keep your AI good-practice card beside you. If you are not using Copilot, read the prepared response on page 4 after you finish Part A.

```{r, results='asis'}
html_out(paste0('<div class="prompt"><strong>Improved prompt</strong>',review_paragraphs(review_prompt),'</div>'))
```

(a) Which question that the AI asked helped you most?

<div class="writing part"></div>

(b) What improved compared with the response in Part A?

<div class="writing part"></div>

(c) Which claim still needs checking against H1-A?

<div class="writing part"></div>

</div>

<div class="sheet-page">

## Page 4: Prepared response

Use this response if you are not using Copilot. Read it after you have marked the response on page 3. It is an example of a better answer. An AI tool may answer differently.

```{r, results='asis'}
html_out(paste0('<div class="ai-source">',review_paragraphs(review_fallback),'</div>'))
```

(a) Which sentences does H1-A (the before-and-after table) support?

<div class="writing part"></div>

(b) What additional evidence would answer the clarifying questions at the start of the response?

<div class="writing part"></div>

### Records used in Q1

These ten participant businesses were selected from the GreenWaste data for teaching. They are not representative of all participant businesses. Values are annual waste management cost before GreenWaste, in AED.

```{r, results='asis'}
html_out(paste0('<p class="source-note">',paste(paste(review_ten$business,fmt(review_costs),'AED'),collapse='; '),'.</p>'))
```

</div>
