---
title: "A practice lottery: 20 manager-age slips"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    header-includes: |
      <style>
      body{font:17px/1.4 Arial,sans-serif;color:#111418;max-width:850px;margin:25px auto;padding:0 25px}h1{font-size:28px;color:#063360}.slips{display:grid;grid-template-columns:repeat(4,1fr);gap:0;margin:18px 0}.slip{border:1px dashed #7f8995;padding:13px 10px;text-align:center}.number{font-size:24px;font-weight:bold;color:#215a9e;margin:0}.age{font-size:23px;margin:5px 0}.id{font-size:12px;color:#545860;margin:0}.caption{font-size:14px;color:#545860}button{font:inherit;padding:8px;color:#063360;background:white;border:1px solid #7da1c4}@page{size:A4;margin:14mm}@media print{body{margin:0;padding:0;font-size:11pt;max-width:none}h1{font-size:19pt}.number{font-size:17pt}.age{font-size:16pt}.slip{padding:10px}.caption{font-size:9pt}button{display:none}}</style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('session3_review.R', encoding='UTF-8')
```

There are 20 businesses and 10 places. Cut out the slips along the dashed lines (one set per group). Draw ten slips for the treatment group. The other ten slips are the control group. Calculate each group's mean manager age. Put all the slips back before you draw again.

<button onclick="window.print()">Print age slips</button>

```{r, results='asis'}
html_out(paste0('<div class="slips">',paste(vapply(1:20,function(i)paste0('<div class="slip"><p class="number">',i,'</p><p class="age">Age ',s3_ages[i],' years</p><p class="id">',pilot$business[i],'</p></div>'),''),collapse=''),'</div>'))
```

The slips use manager ages from 20 businesses in the GreenWaste pilot data. A draw compares ages before the programme, not effects on cost. It is not the lottery used in the pilot, which involved 400 businesses and 200 places. Record both draws on worksheet page 3.

::: {.caption}
GreenWaste is a fictional training case. Ages are in years.
:::
