---
title: "Compared to What? Know the Language First"
subtitle: "Module 2 · Day 1, Session 1: Five Words for Reading Any Result"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html. Static knitr chunks work as before.
#
# One example all session: the fictional traffic camera programme
# (make_traffic_camera_data.R), the same data Session 2 opens in full. This
# session works on a ten-road slice and on average SPEED, so it does not give
# away Session 2's collision results.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  render-df: kable
  resources:
    - evaluation_data_TrafficCameras.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(knitr)
library(qrcode)

gfmt  <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = ",")
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_s1.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- Numbers the static slides quote (the live cells recompute them) -------
tc  <- read.csv("./evaluation_data_TrafficCameras.csv")
cam <- subset(tc, treated == 1)
tot <- function(col, yr) sum(cam[[col]][cam$year == yr])

fatal_before <- tot("fatal_collisions", 2019)          # 233
fatal_after  <- tot("fatal_collisions", 2021)          # 159
fatal_pct    <- 100 * (fatal_after - fatal_before) / fatal_before
vkm_before   <- tot("vehicles_km_millions", 2019)
vkm_after    <- tot("vehicles_km_millions", 2021)
vkm_pct      <- 100 * (vkm_after - vkm_before) / vkm_before
rate_before  <- 1000 * fatal_before / vkm_before       # per 1,000 million km
rate_after   <- 1000 * fatal_after  / vkm_after
rate_pct     <- 100 * (rate_after - rate_before) / rate_before

# Ten camera roads, 2019, chosen once (seed 12 in the planning script) so the
# room can do the arithmetic by hand. The last one is the busiest road in the
# programme, which is what makes the mean and the median part company.
TEN <- c("SEC08-013", "SEC10-021", "SEC10-024", "SEC12-020", "SEC04-033",
         "SEC02-023", "SEC10-047", "SEC08-002", "SEC06-011", "SEC05-018")
ten <- subset(tc, year == 2019 & segment_id %in% TEN)
ten <- ten[order(ten$injury_collisions), ]

# Average speed, the 2x2 for Word 2.
spd <- function(t, yr) mean(tc$ave_speed_kmh[tc$treated == t & tc$year == yr])
sp_c0 <- spd(1, 2019); sp_c1 <- spd(1, 2021)     # camera roads
sp_o0 <- spd(0, 2019); sp_o1 <- spd(0, 2021)     # roads without cameras
r1 <- function(x) round(x, 2)   # two decimals keep the hand arithmetic exact
sp_effect <- (r1(sp_c1) - r1(sp_c0)) - (r1(sp_o1) - r1(sp_o0))
```

# 0 · Where Module 1 left us {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The word Module 1 was built on

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Every live cell below uses
# what is defined here, so each one on a slide stays two or three lines long.
tc <- read.csv("evaluation_data_TrafficCameras.csv")
library(estimatr)
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Warm up estimatr now, while the opening slides are talked through: its first
# call in the browser takes about 10 seconds, every later one is instant.
invisible(estimatr::lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1)),
                              clusters = c(1, 1, 2, 2)))

clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)
aed <- function(x) format(round(x), big.mark = ",")

TEN <- c("SEC08-013", "SEC10-021", "SEC10-024", "SEC12-020", "SEC04-033",
         "SEC02-023", "SEC10-047", "SEC08-002", "SEC06-011", "SEC05-018")
ten <- subset(tc, year == 2019 & segment_id %in% TEN)
cam <- subset(tc, treated == 1)

# Bars with the value printed on each (handles negative values).
bars <- function(v, col = c(ACCENT, ACCENT2, ALERT)) {
  clean_par(); par(mar = c(2.5, 1, 1.5, 1))
  b <- barplot(v, col = col, border = NA, axes = FALSE,
               ylim = range(c(0, v)) + c(-1.2, 1.2))
  abline(h = 0, col = MUTED)
  text(b, v, sprintf("%.1f km/h", v), pos = ifelse(v < 0, 1, 3), cex = 1.2, xpd = NA)
}

# One row per road: speed in 2019, speed in 2021, and the change.
wide <- reshape(tc[, c("segment_id", "treated", "year", "ave_speed_kmh",
                       "school_within_500m", "lanes")],
                idvar = c("segment_id", "treated", "school_within_500m", "lanes"),
                timevar = "year", direction = "wide")
names(wide) <- sub("ave_speed_kmh.", "speed_", names(wide))
wide$change <- wide$speed_2021 - wide$speed_2019

# Dots for a handful of values, with the mean (red) and the median (dashed).
dots <- function(x, labels = NULL) {
  clean_par()
  stripchart(x, method = "stack", pch = 19, cex = 2.2, col = ACCENT, offset = 0.7,
             xlim = c(0, 45), xlab = "Injury collisions, 2019", at = 0.2, ylim = c(0, 1.2))
  abline(v = mean(x), col = ALERT, lwd = 3)
  abline(v = median(x), col = MUTED, lwd = 3, lty = 2)
  text(mean(x), 1.1, paste("mean", round(mean(x), 1)), col = ALERT, pos = 4)
  text(median(x), 0.9, paste("median", median(x)), col = MUTED, pos = 2)
}

# A study of n roads (half with cameras), always the same draw for a given n.
study <- function(n, seed = 3) {
  set.seed(seed)
  rbind(wide[sample(which(wide$treated == 1), n / 2), ],
        wide[sample(which(wide$treated == 0), n / 2), ])
}
gap <- function(d) mean(d$change[d$treated == 1]) - mean(d$change[d$treated == 0])

# Shuffle the camera labels k times and count how often chance alone beats
# the gap we measured.
shuffle_plot <- function(d, k) {
  real <- gap(d)
  set.seed(1)
  sh <- replicate(max(k, 1), { e <- d; e$treated <- sample(d$treated); gap(e) })
  if (k == 0) sh <- numeric(0)
  clean_par()
  hist(sh, breaks = seq(-6, 6, by = 0.25), col = ACCENT2, border = "white",
       main = "", xlab = "Gap in speed change, shuffled labels (km/h)",
       ylab = "Shuffles", xlim = c(-5, 5), ylim = c(0, max(12, k / 8)))
  abline(v = c(real, -real), col = ALERT, lwd = 3, lty = c(1, 2))
  cat(sprintf("Measured gap: %.1f km/h.  Shuffles: %d.  As far from zero by chance: %d  ->  p = %s\n",
              real, k, sum(abs(sh) >= abs(real)),
              if (k == 0) "?" else format(round(mean(abs(sh) >= abs(real)), 3))))
}

# Estimate and 95% interval against zero, one line.
one_interval <- function(est, lo, hi, xlim = c(-5, 1)) {
  clean_par()
  plot(NA, xlim = xlim, ylim = c(0.5, 1.5), yaxt = "n", bty = "n",
       xlab = "Effect on average speed (km/h)", ylab = "")
  abline(v = 0, col = MUTED, lwd = 2, lty = 2)
  segments(lo, 1, hi, 1, col = ACCENT, lwd = 8)
  points(est, 1, pch = 19, cex = 2.5, col = ACCENT)
}

# "Suppose the cameras had cut speed by only `cut_by` km/h": shift the camera
# roads' changes so the effect is exactly that, keep all 1,200 roads, re-test.
significance_demo <- function(cut_by) {
  what_if <- wide
  what_if$change <- what_if$change + what_if$treated * (-gap(wide) - cut_by)
  f  <- lm(change ~ treated, data = what_if)
  ci <- confint(f)["treated", ]
  p  <- summary(f)$coefficients["treated", 4]
  cat(sprintf("All 1,200 roads. Effect %.1f km/h, 95%% interval %.2f to %.2f, p = %s
",
              coef(f)[["treated"]], ci[1], ci[2], format.pval(p, digits = 2, eps = 0.001)))
  one_interval(coef(f)[["treated"]], ci[1], ci[2])
}

# 100 studies of n roads each, with their intervals. Red: misses the value we
# get from all 1,200 roads.
many_studies <- function(n = 40, runs = 100, seed = 1) {
  truth <- gap(wide)
  set.seed(seed)
  res <- t(replicate(runs, {
    d <- rbind(wide[sample(which(wide$treated == 1), n / 2), ],
               wide[sample(which(wide$treated == 0), n / 2), ])
    f <- lm(change ~ treated, data = d)
    c(coef(f)[["treated"]], confint(f)["treated", ])
  }))
  res  <- res[order(res[, 1]), ]
  miss <- res[, 2] > truth | res[, 3] < truth
  clean_par()
  plot(NA, xlim = c(-7, 1), ylim = c(1, runs), yaxt = "n", bty = "n",
       xlab = "Effect on average speed (km/h), one line per study", ylab = "")
  segments(res[, 2], 1:runs, res[, 3], 1:runs, col = ifelse(miss, ALERT, ACCENT2), lwd = 2)
  points(res[, 1], 1:runs, pch = 19, cex = 0.5, col = ifelse(miss, ALERT, ACCENT))
  abline(v = truth, col = MUTED, lwd = 2)
  cat(sprintf("%d studies of %d roads. %d intervals contain the all-roads value (%.2f); %d miss it.\n",
              runs, n, sum(!miss), truth, sum(miss)))
}
```

::: {.ask}
From Module 1: what is a **counterfactual**?
:::

::: {.panel .fragment}
What would have happened to the same people, over the same period, without the
programme. We never observe it. Every evaluation method estimates it with a
**comparison**, and a result can only be read once you know what that
comparison was.
:::

::: {.panel .fragment}
Today we attach five words to that idea, using one programme and its data from
start to finish.
:::

::: {.notes}
About 3 minutes. Recall only; Module 1 taught this. Open the deck a couple of minutes before the session starts: R and its statistics package take about a minute to load in the browser, after which every live cell runs in a few seconds.
:::

# 1 · What are we comparing? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The traffic camera programme

:::::: {.columns}

::::: {.column width="54%"}
In 2020 a roads authority installed automated speed cameras in twelve sectors
of its road network. The other twelve sectors were due to get them two years
later.

Before extending the programme, someone has to decide whether the cameras
worked.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>  <div class="v">Automated speed cameras</div>
  <div class="k">Roads</div>      <div class="v">1,200 road segments in 24 sectors</div>
  <div class="k">Measured</div>   <div class="v">2019, before; 2021, after</div>
  <div class="k">Decision</div>   <div class="v">Extend to the other twelve sectors?</div>
</div>
```
:::::

::::::

::: {.warn}
**Illustrative case, with invented figures.** This is not a real programme; the
numbers are generated for this training.
:::

## The headline

```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%s%%</div>
  <div class="label">Fatal collisions on camera roads, 2019 to 2021</div>
</div>
<div class="stat-row">
  <div><span class="n">%d</span><span class="k">Before</span></div>
  <div><span class="n">%d</span><span class="k">After</span></div>
</div>', gfmt(fatal_pct), fatal_before, fatal_after))
```

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask}
**`[MENTI PLACEHOLDER]`** On this number alone, would you extend the cameras to
the other twelve sectors?

Go to **menti.com**, code **`r MENTI_CODE`**, or scan the code.
:::
:::::

::::: {.column width="38%"}
```{r out.width='230px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

## What if traffic had changed?

```{ojs}
//| echo: false
viewof traffic = Inputs.range([-40, 20], {value: 0, step: 2, label: "Change in driving on camera roads (%)"})
```

```{webr}
#| input:
#|   - traffic
#| autorun: true
#| edit: false
fatal <- tapply(cam$fatal_collisions, cam$year, sum)
km    <- sum(cam$vehicles_km_millions[cam$year == 2019]) * c(1, 1 + traffic / 100)
round(data.frame(fatal_change_pct = 100 * (fatal[2] / fatal[1] - 1),
                 risk_per_km_change_pct = 100 * ((fatal[2] / km[2]) / (fatal[1] / km[1]) - 1),
                 row.names = NULL))
```

::: {.ask}
The count fell 32%. Slide driving down to −32%. **What happens to the risk per
kilometre driven?**
:::

::: {.panel .fragment}
It stops moving. If people drove a third less, a third fewer deaths is what you
would expect with or without cameras. The headline alone cannot tell the two
apart.
:::

## What actually happened to traffic

```{webr}
fatal <- tapply(cam$fatal_collisions, cam$year, sum)
km    <- tapply(cam$vehicles_km_millions, cam$year, sum)
round(data.frame(fatal_change_pct = 100 * (fatal[2] / fatal[1] - 1),
                 driving_change_pct = 100 * (km[2] / km[1] - 1),
                 risk_per_km_change_pct = 100 * ((fatal[2] / km[2]) / (fatal[1] / km[1]) - 1),
                 row.names = NULL), 1)
```

::: {.panel}
Driving on camera roads went **up** by `r gfmt(vkm_pct, 1)`%. So the risk per
kilometre fell by `r gfmt(abs(rate_pct))`%, a little more than the headline
said.
:::

::: {.ask .fragment}
The comparison moved the answer, this time in the programme's favour. What
other comparison would you want before crediting the cameras?
:::

::: {.notes}
Take answers. If someone says "roads without cameras", say that is exactly
where Session 2 goes, and leave it there. Then a show of hands: who would now
change their vote from the headline slide? Take one reason from someone who
changed and one from someone who did not. Do not show the comparison roads'
collisions here: Session 2's last section depends on it.
:::

## "Compared to what?"

::: {.panel}
Every result is a comparison. A before-and-after number compares the programme
with **its own past**. What else changed over the same period (traffic,
vehicles, a national campaign) is inside that number, and it can push the
answer either way.
:::

::: {.ask}
Think of a figure you reported, or were sent, this year. **What was it compared
to?** If nobody said, what would you have picked?
:::

# 2 · Three questions for any number {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## One idea, three questions

::: {.panel}
The five words that follow each answer one of three questions you can ask of
any number an evaluator hands you.
:::

```{=html}
<div class="facts">
  <div class="k">Compared to what?</div>             <div class="v">Comparison group, counterfactual</div>
  <div class="k">How big, in units that matter?</div> <div class="v">Mean · treatment effect · coefficient</div>
  <div class="k">How sure are we?</div>              <div class="v">p-value · confidence interval</div>
</div>
```

::: {.panel}
We come back to this list at the end of the session, and on every day of the
week.
:::

# 3 · The five words {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Word 1 · Mean: ten roads, by hand

:::::: {.columns}

::::: {.column width="42%"}
```{r}
ten %>%
  transmute(Road = segment_id, Type = corridor_type,
            `Injury collisions` = injury_collisions) %>%
  kable(align = "llr", row.names = FALSE)
```
:::::

::::: {.column width="58%"}
::: {.ask}
Ten camera roads in 2019. With pen and paper: **what is the average number of
injury collisions per road?** Then: is a typical road close to it?
:::

::: {.panel .fragment}
The mean is **`r gfmt(mean(ten$injury_collisions), 1)`**.
`r sum(ten$injury_collisions < mean(ten$injury_collisions))` of the ten roads are
below it. One ring road with `r max(ten$injury_collisions)` collisions pulls
the mean up; the middle road (the **median**) has
`r median(ten$injury_collisions)`.
:::
:::::

::::::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

::: {.notes}
Paper (Oct12_session1_materials.docx): Hand out the ten-roads worksheet (pack page 1, one per participant; the speed table is on the back). Give two minutes before the reveal.
:::

## Word 1 · What the mean hides

::::::: {.columns}

:::::: {.column width="45%"}
```{ojs}
//| echo: false
viewof keep = Inputs.radio(["All ten roads", "Without the busiest road"], {value: "All ten roads"})
```
::::::

:::::: {.column width="55%"}
::: {.ask}
Remove one road. How far does the mean move, and how far does the median move?
:::
::::::

:::::::

```{webr}
#| input:
#|   - keep
#| autorun: true
#| fig-width: 10
#| fig-height: 1.9
roads <- ten$injury_collisions
if (keep == "Without the busiest road") roads <- roads[roads < max(roads)]
dots(roads)
```

::: {.panel .fragment}
The mean drops from 13.5 to 10.3; the median from 10 to 8. One road moved the
mean further than the median. When an evaluator reports an average, ask what
the spread looks like and whether a few cases drive it.
:::

## Word 2 · Treatment effect

```{r results='asis'}
html_out(sprintf('
<table>
  <thead><tr><th>Average speed (km/h)</th><th>2019</th><th>2021</th><th>Change</th></tr></thead>
  <tbody>
    <tr><td>Camera roads</td><td>%.2f</td><td>%.2f</td><td class="fragment" data-fragment-index="1">%.2f</td></tr>
    <tr><td>Roads without cameras</td><td>%.2f</td><td>%.2f</td><td class="fragment" data-fragment-index="1">%.2f</td></tr>
    <tr class="fragment" data-fragment-index="1"><td><strong>Difference</strong></td><td>%.2f</td><td>%.2f</td><td><strong>%.2f</strong></td></tr>
  </tbody>
</table>',
  r1(sp_c0), r1(sp_c1), r1(sp_c1) - r1(sp_c0), r1(sp_o0), r1(sp_o1), r1(sp_o1) - r1(sp_o0),
  r1(sp_c0) - r1(sp_o0), r1(sp_c1) - r1(sp_o1), sp_effect))
```

::: {.ask}
Three numbers describe this result: **`r sprintf("%.2f", r1(sp_c1) - r1(sp_c0))`**,
**`r sprintf("%.2f", r1(sp_c1) - r1(sp_o1))`** and **`r sprintf("%.2f", r1(sp_c0) - r1(sp_o0))`**.
Work out what each one is, then decide which is the effect of the cameras on speed.
:::

::: {.panel .fragment fragment-index=2}
**`r sprintf("%.2f", r1(sp_c0) - r1(sp_o0))`** is the head start: camera roads
were faster before anything happened, and **`r sprintf("%.2f", r1(sp_c1) - r1(sp_o1))`**
still carries it. The **treatment effect** is the change on camera roads minus
the change on the others: **`r sprintf("%.2f", sp_effect)`** km/h. It is close to
the camera roads' own change only because the other roads barely moved.
:::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

::: {.notes}
Paper (Oct12_session1_materials.docx): Participants turn the worksheet over: part 2, the speed table. They fill the blank cells before you reveal them.
:::

## Word 2 · The same number from 1,200 roads

```{webr}
cams   <- mean(wide$change[wide$treated == 1])
others <- mean(wide$change[wide$treated == 0])
round(data.frame(camera_roads = cams, other_roads = others, effect = cams - others), 2)
```

::: {.panel}
`wide` holds one row per road: its speed in 2019, in 2021, and the change. The
three lines are the arithmetic you just did, on every road.
:::

## Word 2 · The effect depends on the comparison

```{ojs}
//| echo: false
viewof other = Inputs.range([-4, 2], {value: 0, step: 0.2, label: "Speed change on roads without cameras (km/h)"})
```

```{webr}
#| input:
#|   - other
#| autorun: true
#| edit: false
#| fig-width: 7
#| fig-height: 2.6
cams <- mean(wide$change[wide$treated == 1])
bars(c("Camera roads" = cams, "Other roads" = other, "Effect" = cams - other))
```

::: {.ask}
Camera roads are fixed at −3.4. Move the other roads. **Which value makes the
effect zero, and what would that mean in plain words?**
:::

## Word 3 · Coefficient

```{webr}
fit <- lm(change ~ treated, data = wide)
round(data.frame(coefficient = coef(fit)), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
`treated` is 1 for camera roads, 0 for the rest. A **coefficient** says how much
the outcome moves per one unit of the row. Here one unit is "has cameras".
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Which of the two numbers is the effect? What is the other one?
:::

::: {.panel .fragment}
`treated` is **−3.36**, the effect you worked out by hand. `(Intercept)` is the
average change on roads without cameras. It is not an effect.
:::
:::::

::::::

## Word 3 · The other rows of a table

```{webr}
fit2 <- lm_robust(change ~ treated + school_within_500m + lanes, data = wide)
round(as.data.frame(summary(fit2)$coefficients[, 1:4]), 3)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Std. Error** is how much the estimate would move from one sample of roads to
another. Small relative to the estimate means a steady estimate.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The other rows adjust the comparison. `school_within_500m` says roads near
schools slowed a little more. That is not a recommendation to build schools
next to roads.
:::
:::::

::::::

## Word 4 · p-value: shuffle the labels

```{ojs}
//| echo: false
viewof k = Inputs.range([0, 1000], {value: 0, step: 50, label: "Times we shuffle the camera labels"})
```

```{webr}
#| input:
#|   - k
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.7
small <- study(20)          # a study of 20 roads: 10 with cameras, 10 without
shuffle_plot(small, k)
```

::: {.ask}
If the cameras did nothing, the labels would be arbitrary, so we shuffle them.
**How often does chance alone give a gap as large as the red line?** Guess first.
:::

::: {.panel .fragment}
That share is the **p-value**: here about 0.17, one shuffle in six.
:::

::: {.notes}
Start at 0 shuffles; the red line is this 20-road study's gap (-1.9). Slide up.
At 1,000 shuffles p is about 0.17: chance produces a gap this big about one time
in six. Keep that number for two slides from now.
:::

## Word 4 · Significant is not the same as big

```{ojs}
//| echo: false
viewof cut_by = Inputs.range([0.1, 3.4], {value: 3.4, step: 0.1, label: "Suppose cameras cut speed by only (km/h)"})
```

```{webr}
#| input:
#|   - cut_by
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.7
significance_demo(cut_by)
```

::: {.ask}
Slide the effect down to 0.3 km/h, a change no driver would notice. **Is it
still statistically significant?**
:::

::: {.warn .fragment}
Yes, p is about 0.03. With 1,200 roads even a trivial change passes the test.
Whether 0.3 km/h is worth paying for is a separate question, and the p-value
cannot answer it.
:::

## Word 4 · The warning cuts both ways

```{ojs}
//| echo: false
viewof roads_n = Inputs.range([6, 200], {value: 20, step: 2, label: "Roads in the study"})
```

```{webr}
#| input:
#|   - roads_n
#| autorun: true
#| edit: false
set.seed(3)
p <- replicate(200, t.test(change ~ treated, data = study(roads_n, seed = sample(1e6, 1)))$p.value)
data.frame(roads_in_study = roads_n, share_of_studies_that_detect_it = round(mean(p < 0.05), 2))
```

::: {.panel}
The real effect is −3.4 km/h on every run. What changes is the size of the
study. At 20 roads, about one study in five misses it, like the p = 0.17 study
two slides ago.
:::

::: {.ask .fragment}
An evaluator reports "no significant effect". Ask two things: **how many roads
(or people) were in the study, and what is the smallest effect it could have
detected?**
:::

## Word 5 · Confidence interval

```{ojs}
//| echo: false
viewof run = Inputs.button("Run another 100 studies")
```

```{webr}
#| input:
#|   - run
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 2.4
many_studies(n = 40, runs = 100, seed = run + 1)
```

::: {.ask}
Each line is one study of 40 roads and its **95% confidence interval**, the
range of values consistent with its data. **Out of 100, how many will miss the
grey line?** Guess before the next press.
:::

::: {.notes}
About 95 of 100 lines cross the grey line (the all-roads value), and about 5 miss
(red). That is what "95%" refers to: the method, over many studies, not the
probability for any one interval. Press the button two or three times.
:::

## Word 5 · Which one would you act on?

::: {.panel}
Suppose three evaluations of the cameras each reported **1.8 fewer injury
collisions per road**, and the authority's bar is **2.0**. Only the intervals
differ.
:::

```{=html}
<div class="facts">
  <div class="k">A</div> <div class="v">1.8 fewer, 95% interval 0.4 to 3.2</div>
  <div class="k">B</div> <div class="v">1.8 fewer, 95% interval 1.6 to 2.0</div>
  <div class="k">C</div> <div class="v">1.8 fewer, 95% interval -0.9 to 4.5</div>
</div>
```

::: {.ask}
Which would you act on? What is your answer for each?
:::

::: {.panel .fragment}
B is precise and below the bar: confidently not enough. A is too uncertain to
call, and C more so. All three have the same estimate, so the intervals make the
difference.
:::

::: {.notes}
These three results are invented for this exercise; they are not the
programme's result, which Session 2 and Day 2 estimate from the data.

Paper (Oct12_session1_materials.docx): Part 3 of the worksheet: everyone ticks act / do not act / cannot tell yet for A, B and C before the reveal.
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI prompt cards

:::::: {.columns}

::::: {.column width="45%"}
::: {.panel}
Each group takes one card. Paste its prompt into your AI tool, read the answer
together, and mark anything it gets wrong or overstates.
:::

```{=html}
<div class="facts">
  <div class="k">Card 1</div> <div class="v">"Explain a 95% confidence interval, in one paragraph, for someone who is not a statistician."</div>
  <div class="k">Card 2</div> <div class="v">"Explain what a p-value means."</div>
  <div class="k">Card 3</div> <div class="v">"Fatal collisions on our camera roads fell 32%. Is that a good result?"</div>
  <div class="k">Card 4</div> <div class="v">"The coefficient on school_within_500m is -0.37. What should I do about it?"</div>
</div>
```
:::::

::::: {.column width="55%"}
::: {.ai-output}
*Card 1. A response like this is possible:*

"A 95% confidence interval is the range in which the true value is likely to
fall. [There is a 95% probability that the true effect lies inside this
interval]{.flag}, so if the interval is −3.6 to −3.1 km/h the true effect is
probably in that range. A narrower interval means a more precise estimate."
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session1_materials.docx): Hand out one set of the four prompt cards per group (pack page 3, cut); each group takes one card.
:::

## Debrief

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Card 1.** The flagged sentence is the common error. "95%" describes the method:
over many studies, 95 of 100 intervals catch the true value, as the chart of 100
studies showed. It is not a probability for this one interval.
:::

::: {.warn}
**Card 3.** Did the AI ask what the 32% was compared with, or how much driving
changed? The prompt gave it neither.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Card 4.** Look for advice about schools. The row adjusts the comparison; it is
not a policy lever.
:::

::: {.panel}
Two habits: ask the AI to **ask you questions before it answers**, and put its
answer through a second chat to check it. Also note what it explained well.
:::
:::::

::::::

# 5 · Triage the claims {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Triage the claims

::: {.excerpt}
**1.** "Fatal collisions on camera roads fell 32% over two years." \
**2.** "The effect of cameras on speed is −3.4 km/h, 95% confidence interval −3.6 to −3.1." \
**3.** "The fall in speed is statistically significant (p < 0.001), so every road should get a camera." \
**4.** "Speed fell 3.4 km/h on camera roads and did not change on roads without them." \
**5.** "The average road in our sample had 13.5 injury collisions." \
**6.** "A study of 20 roads found no significant change in speed." \
**7.** "The coefficient on school_within_500m is −0.37, so schools near roads slow traffic." \
**8.** "Speed fell 3.4 km/h more than on roads without cameras; the target was at least 3 km/h."
:::

::: {.ask}
In groups: for each excerpt, **which of the three questions does it leave
unanswered, and would you act, not act, or ask first?** Six minutes. Then: which
one were you most tempted to act on?
:::

::: {.notes}
Target in excerpt 8 is invented for the exercise. Key: 1 compared to what (ask
first); 2 how big in units that matter, speed not collisions (ask first); 3
significance is not a decision rule (do not act); 4 how sure, no interval (ask
first); 5 one road pulls the mean (ask for the spread); 6 too small to see
anything (do not conclude "no effect"); 7 not a policy lever (do not act); 8
comparison, effect and bar (act). Groups usually split on 2 and 4: take one
group's reasoning for each.

Paper (Oct12_session1_materials.docx): Hand out the triage sheet (pack page 4, one per group) and one set of excerpt slips per group (page 5, cut).
:::

# 6 · Take this back to your desk {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The three questions

::: {.panel}
From memory, before you look: what are the three questions you can ask of any
number?
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what?</div>
  <div class="k">2</div> <div class="v">How big is it, in units that matter to the decision?</div>
  <div class="k">3</div> <div class="v">How sure are we?</div>
</div>
```

::: {.panel}
Next session opens the full camera data file and asks these three questions of
the collision numbers.
:::

::: {.notes}
If there is time: ask the room for a better way to ask AI about one of today's
words. For example: "I make policy decisions using impact evaluations. How
should I read a confidence interval when I compare an estimate with a decision
threshold? Ask me questions first." Saying who you are and why you ask usually
gets a more useful answer.

Paper (Oct12_session1_materials.docx): Hand out the take-away cards (pack page 6). Give a minute to fill in "On my desk" on the card, then take two or three examples from the room.
:::
