---
title: "Did the programme make the difference?"
subtitle: "Day 1 | Session 3: A pilot lottery, its uncertainty and the limits of expansion"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, session3_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session1_review_interactions.html, _session3_review_interactions.html]
webr:
  render-df: kable
  resources: [evaluation_data_GreenWaste_simple.csv]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('session3_hybrid.R', encoding='UTF-8')
```

## Who and what are we comparing? The pilot district. {#s3-map .reader-slide .s1-review .s3-review .case-map-slide}

```{r, include=FALSE}
map_highlight <- "pilot"
```

::: {.case-map}
{{< include _case_map.qmd >}}
:::

<div class="map-facts"><div><strong>Unit</strong>one business</div><div><strong>Group</strong>400 businesses in the pilot district: 200 treatment group, 200 control group, chosen by lottery</div><div><strong>Outcome</strong>annual waste management cost, in AED per business</div><div><strong>Period</strong>before GreenWaste and 12 months after</div></div>

::: {.notes}
1 minute. Point to the pilot district only. The lottery is what makes this study different from the city comparisons. The same map appears in every methods session so that participants know the population before the details start.
:::

## We have two city comparisons. Why would a lottery be better? {#s3-p01 .reader-slide .s1-review .s3-review}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
pilot <- subset(gw, setting == 'pilot')
ages <- pilot$manager_age[1:20]
draw <- 0
```

<div class="comparison-pair"><div><h3>Before and after: −669 AED</h3><p>The same participant businesses, measured over time.</p><p>What would have changed without GreenWaste?</p></div><div><h3>With and without: −1,638 AED</h3><p>Participant businesses and comparison businesses, after a year.</p><p>Were the groups already different?</p></div></div>

::: {.ask}
What if chance, rather than the score rule, decided which businesses took part?
:::

Use the summary of the two comparisons from Session 2. Now we move to the **pilot district**.

::: {.notes}
4 minutes. One minute recall, two minutes pairs explain the two missing comparisons, one minute bridge. No new city arithmetic or repeat baseline reveal. Pilot is a different population, not the city's known true effect. Start webR setup early for one optional trainer draw. Printed activity is independent of internet. Transition: there were 400 eligible businesses but only 200 places.
:::

## The pilot had 400 eligible businesses but only 200 places in the programme. {#s3-p02 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_flow())
```

**The pilot trial box on page 1 of your worksheet:** in one district, a lottery assigned 400 businesses to join or wait. The pilot assigned exactly 200 and 200. The 200 that receive GreenWaste are the treatment group (the “join” businesses on your worksheet). The 200 that do not are the control group (the “wait” businesses). The outcome is annual waste management cost 12 months after GreenWaste.

::: {.ask}
The lottery assigns **individual businesses**. It does not draw a sample of the country.
:::

::: {.notes}
5 minutes. Read the pilot trial box on page 1. The sheet says about half joined; the pilot assigned exactly 200 and 200. Exactly 200/200, not roughly half; each eligible business has a place in join or wait. All scores at/below 58; city rule has not allocated these pilot places. Distinguish choosing the pilot district from randomly assigning inside it. Ask who the comparison will describe. Transition: why does the method of allocation matter?
:::

## What does a lottery give us? {#s3-p03 .reader-slide .s1-review .s3-review}

<div class="comparison-pair"><div><h3>If managers choose</h3><p>More motivated managers might join first. That could also affect their costs.</p></div><div><h3>If the lottery chooses</h3><p>Managers' motivation does not decide which eligible business gets a place.</p></div></div>

The first example is **hypothetical**. Over many possible lotteries, random assignment does not favour one kind of business.

::: {.ask}
Discuss with your group; there is no written question for this slide. Why does the control group give a more reliable comparison than businesses that chose whether to join?
:::

::: {.notes}
7 minutes. Explain random assignment/RCT in plain language. Two minutes in pairs, then discuss. Chance balances observed and unobserved characteristics in expectation; one draw need not be identical. Design still needs implementation, comparable measurement/follow-up and no important effects on the control group. Source §3 explicitly states assignment matches participation and all 400 costs observed in this fictional pilot. Avoid a long validity checklist. Transition: experience the difference between chance and perfect balance.
:::

## Draw the lottery, then draw it again. {#s3-p04 .reader-slide .s1-review .s3-review}

This practice lottery has **20 businesses and 10 places**. The slips show **manager ages before the programme**.

Draw ten age slips for the treatment group. The other ten slips are the control group. Compare the mean manager age of the two groups. Put all the slips back, and draw again.

**R code**

```{webr}
#| echo: true
# Each run makes a new practice lottery.
draw <- draw + 1
set.seed(draw)
# Ten businesses go to the treatment group and ten to the control group.
group <- sample(rep(c('Treatment', 'Control'), each = 10))
# Compare the mean manager ages. These are not programme effects.
round(tapply(ages, group, mean), 1)
```

::: {.ask}
Did both lotteries give exactly the same mean ages?
:::

::: {.notes}
12 minutes. Change of activity, not an extra break: distribute one cut slip set per group, two minutes assign roles/read, three minutes draw/calculate, three minutes replace/redraw, two minutes reflection, two minutes optional R/debrief. Physical slips use the first 20 pilot manager-age records; this is not the allocation used in the pilot. A reader checks IDs and a recorder notes both means on scrap paper. The trainer run repeats the same pool with a different fixed seed; it need not reproduce a physical draw. Code comments explain operations; no participant code. A calculator or total/10 is enough. If runtime unavailable complete paper route within the same time. Transition: what might change with more businesses?
:::

## Small lotteries can look different {#s3-p05 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<div id="s3-sizes"><div data-variant>',s3_sizes(),'</div><div data-variant hidden>',s3_sizes(TRUE),'</div></div>'))
```

<button data-swap="s3-sizes" data-closed="Show 400-business practice lotteries" data-opened="Show small lotteries only" aria-pressed="false">Show 400-business practice lotteries</button>

In this illustration, we ran **200 practice lotteries of each size** using manager ages from the pilot records. The differences shown are not programme effects.

::: {.ask}
Predict first: would the larger groups usually be closer in mean age?
:::

::: {.notes}
6 minutes. Each dot is the treatment mean age minus the control mean age in one practice allocation; same horizontal scale. For N=20, take a random 20-record subset from the full age pool each time; for N=400 use full pool. This isolates sample-size variation; the physical slips used a fixed selected 20. Explain the display, not probability formulas. Reveal large draws and ask for a sample-size reflection. Larger groups usually closer, no guarantee for a specific draw; no p-value balance proof. Transition: inspect the pilot baseline rather than a redraw.
:::

## Were the treatment and control groups similar before GreenWaste? {#s3-p06 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_baseline())
```

The figure shows the two pilot groups as assigned: 200 businesses in the treatment group and 200 in the control group. The distributions are similar, but the groups are not identical.

::: {.ask}
What reassurance does this give? What does it not prove?
:::

::: {.notes}
6 minutes. Means 1,390/1,411; side-by-side bars in each 250 AED band on common axes. Listen for starting distributions similar and small mean difference plausible under chance. No balance p-value, no significance testing to prove randomisation. Similar costs alone cannot establish unobserved balance or correct allocation; check assignment procedure in a real report. Ask for a baseline reflection. Transition: the outcome comparison is after the programme.
:::

## The result is the difference between two averages. {#s3-p07 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_after())
```

::: {.ask}
**Worksheet, question 1 (Name the comparison):** Name the two groups, the outcome and the period. The treatment group's mean cost minus the control group's mean cost is 769 − 1,783 ≈ **−1,014 AED**.
:::

The treatment group's mean annual waste management cost was lower, **12 months after GreenWaste**, among the eligible pilot businesses.

::: {.notes}
8 minutes. The result table on page 1 of the worksheet reports this same comparison. First axes/population/period, then two minutes calculate/interpret, then three minutes pair check and feedback. Unrounded 769.125-1783.48=-1014.355; display -1,014. Negative cost difference means positive saving. It is an after-period randomised group comparison, not participants before/after. Assignment equals participation here. Transition: the simple regression expresses this same comparison.
:::

## We create the regression table in R with one line of code. {#s3-p07b .reader-slide .s1-review .s3-review}

We create the table in R. The box shows the **R code** and the numbers it produces.

```{r, echo=TRUE}
# pilot has one row per business (400 rows).
# took_part is 1 for the treatment group and 0 for the control group.
# cost_after is annual waste management cost 12 months after GreenWaste (AED).
fit <- lm(cost_after ~ took_part, data = pilot)
# Row 1 (Intercept): the control group's mean cost.
# Row 2 (took_part): treatment group minus control group.
round(cbind(estimate = coef(fit), confint(fit)))
```

::: {.ask}
Which line of the code gives the control group row, and which gives the comparison row?
:::

You do not need to write or run R. The pilot trial box on your worksheet shows the second row of this table.

::: {.notes}
4 minutes. Walk through the comments, not the syntax; participants do not code. Match each printed row to a row of the table on the next slide. Estimates: 1,783 and -1,014; interval -1,167 to -862. This is the same two-group comparison as the previous slide.
:::

## The regression table reports the same comparison. {#s3-p08 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(sub('<p class="caption">.*$', '', s3_regression()))
```

<div class="review-annotate"><button data-table-column="1" aria-pressed="false">Coefficient</button><button data-table-column="2" aria-pressed="false">Interval</button><button data-table-column="3" aria-pressed="false">P-value</button></div>

::: {.ask}
**Worksheet, question 1:** Your worksheet shows one row, “Took part (lottery)”. Which row of this table is that? What does the control group row show?
:::

1,783 + (−1,014) ≈ **769 AED**, which is the treatment group's mean cost. The lottery design supports reading the difference as an effect. The p-value alone does not.

::: {.notes}
6 minutes. Annotate the table. Control group intercept is the mean cost after 12 months for the reference group; treatment coefficient is the treatment minus control difference. CI -1,167 to -862; ordinary least squares model matches canonical pilot source. Point to code/model note for mechanics, no need fit models. Show each column with native highlights. A causal reading depends on assignment/implementation/measurement and interference assumptions, not regression magic. Transition: a convincing saving does not settle the exact policy minimum.
:::

## Is the whole interval at or above our rule of a 1,000 AED reduction? {#s3-p09 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_interval_svg('trial',TRUE))
html_out('<button data-open="interval-trial" data-closed="Show the 95% interval for the reduction" data-opened="Hide the interval" aria-controls="interval-trial" aria-expanded="false">Show the 95% interval for the reduction</button>')
```

::: {.ask}
**Worksheet, question 2 (Read the estimate and interval):** State the saving and explain whether it settles the decision rule. Does evidence of some reduction in cost show that the reduction is at least **1,000 AED**?
:::

Our rule is a reduction in annual waste management cost of at least 1,000 AED per business per year. GreenWaste is a fictional case, so the rule is fictional too.

::: {.notes}
8 minutes. Predict from point estimate before reveal: 1,014 just above 1,000; reveal positive saving interval 862-1,167, which spans rule. Ask pairs for two statements: evidence of positive saving, uncertain whether minimum met. CI describes mean-effect uncertainty under model, not a probability statement about this fixed interval or individual businesses' savings. Convert negative effect interval by reversing endpoints. No value-for-money conclusion because full costs absent here. Transition: uncertainty is one limit; coverage is another.
:::

## Who does this pilot describe? {#s3-p10 .reader-slide .s1-review .s3-review}

<div class="comparison-pair"><div><h3>What the pilot covers</h3><p>One pilot district.<br>Businesses with scores of 58 or below.<br>Support delivered under pilot conditions.</p></div><div><h3>What wider use could change</h3><p>Other businesses and districts.<br>Delivery capacity and take-up.<br>Programme costs.</p></div></div>

::: {.ask}
**Worksheet, question 3 (How far does this result reach):** What would you need to know before using this pilot result for the whole city?
:::

A lottery allocates places within the pilot. It does not make the pilot representative of the country.

::: {.notes}
6 minutes. Two minutes specific reach request, then compare. Accept profile/cost/outcome differences, delivery capacity or representative next-stage evidence with purpose. Do not invent observed national outcomes or understate strength of the pilot. No claim result applies only to sampled individuals under all conditions; identify the eligible pilot context and transport assumptions. Transition: place the three estimates beside their comparisons.
:::

## Each of the three numbers compares different groups. {#s3-p11 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_comparisons())
```

::: {.ask}
Which comparison is best designed to estimate an effect? Which businesses does it describe?
:::

The pilot result does **not automatically apply to the city**, because the city's businesses and delivery conditions are different.

::: {.notes}
6 minutes. This comparison table is on the slide only. No new arithmetic. Ask for comparison and population, not just ranking numbers. Stronger pilot design for eligible pilot businesses; do not use its estimate to prove city's before/after bias or city's true effect. All three costs measured after 12 months; city before/after uses its own baseline. Transition: decide what action the pilot evidence supports.
:::

## Would you extend GreenWaste now? {#s3-p12 .reader-slide .s1-review .s3-review}

<div data-choices><button data-choice aria-pressed="false">Extend</button><button data-choice aria-pressed="false">Do not extend</button><button data-choice aria-pressed="false">Ask first</button><p class="feedback" role="status"></p></div>

::: {.ask}
Name one strength of the pilot design, one limit that matters for the decision, and one condition that fits the size of the decision.
:::

Choose alone first. Then vote in the room poll (Extend, Do not extend or Ask first) and compare your reasons with your group. The right action depends on the decision, the costs and how much risk the authority will accept.

::: {.notes}
8 minutes. ROOM POLL (room_poll.py): open Before with Extend, Do not extend and Ask first and collect short reasons; close and discuss; the group discussion that follows replaces a second phase. Two minutes individual thinking, four minutes discuss, two minutes hear reasons. Native buttons record choice locally, not shared room voting. Paper labelled cards are equivalent. Do not award ask first by default or force consensus. Good reasons use randomised comparison, threshold uncertainty and reach; condition must address a consequential concern. Transition: turn this into advice the director can read.
:::

## Write a decision note with a partner. {#s3-p13 .reader-slide .s1-review .s3-review}

Use **page 2 of your worksheet, Write a decision note**. With a partner, write three or four sentences:

1. **Result and comparison:** what the pilot found, and which groups were compared.
2. **Provisional recommendation:** your recommendation, or the condition you set.
3. **Evidence still needed:** the most important limit, and the evidence you would request.

::: {.ask}
Swap notes with another pair. Can they find support in the pilot trial box on page 1 for each claim?
:::

Keep page 1 open. Different recommendations can be defensible.

::: {.notes}
13 minutes. Seven minutes write in pairs, four minutes peer source-check, two minutes feedback. The checking pair tests units and period, the lottery comparison, the interval against the rule, reach and a specific request; they circle a claim to strengthen, not grammar. Full costs are absent, so do not assert value for money. No predetermined scale-up answer. Transition: compare your own advice with a short AI draft.
:::

## An AI writes a paragraph for the minister. Which claims does the pilot support? {#s3-p14 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<p class="caption">Naive prompt (a prompt with no context, as a busy person might type it): “Write a short paragraph for the minister recommending that GreenWaste is expanded.” This response was written for the exercise. It is not a real AI answer. It is also printed on page 2 of your worksheet.</p><div class="ai-reading">',s3_paragraphs(hy3_ai),'</div>'))
```

::: {.ask}
**Worksheet, page 2, Check an illustrative AI draft:** Mark each unsupported conclusion. Rewrite it using the pilot trial result on page 1.
:::

::: {.notes}
8 minutes. AI SNAPSHOT, stage 1: four minutes individually, four minutes debrief; do not run the naive prompt. Supported: the pilot estimated an annual saving of 1,014 AED per business. Unsupported: that the trial proves the minimum saving is met (the interval, 862 to 1,167 AED, spans 1,000 AED, and a small p-value does not show that a threshold is met); that nationwide rollout is good value for money (one district, no programme cost evidence). Ask participants to quote the part of the pilot trial result that supports their rewrite. Transition: run a prompt constrained by the pilot information.
:::

## Ask the AI for a recommendation the pilot supports. {#s3-p15 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<p class="caption"><strong>Improved prompt</strong></p><div class="better-prompt">',s3_paragraphs(hy3_prompt),'</div>'))
```

Use the **AI handout** to copy the prompt into **Microsoft Copilot**. Keep your AI good-practice card beside you.

::: {.ask}
What improved compared with the draft you marked and with your own decision note? Which advice still needs checking against the pilot trial box on page 1?
:::

::: {.notes}
7 minutes. AI SNAPSHOT, stage 2: four minutes run, three minutes compare; the AI Snapshot totals 15 minutes. The prompt already contains the pilot facts, so no study summary needs to be pasted. Ask for clarifying questions first; no account setup in the session. If Copilot is unavailable or delayed, use the prepared response on the back of the AI handout. Compare with the marked draft, the decision note and the pilot trial result, especially the interval against the rule, reach and value for money. Agreement between answers is not proof.
:::

## Which evidence could change your decision? {#s3-p16 .reader-slide .s1-review .s3-review}

Use the last box on page 2 of your worksheet, **Your question from memory**. Close page 1, the AI responses and peer comments. Work alone.

::: {.ask}
What would you ask the evaluator, and how could the answer change your recommendation?
:::

Tomorrow: what can the city comparison of changes tell us?

::: {.notes}
10 minutes. Four minutes independent revision/request, three minutes partner checks purpose, three minutes feedback. Require evidence/document/result and consequence for advice; accept design verification, interval precision, expansion profile/delivery or costs when relevant. Keep source open but learner writes without AI/peer copy. Daily wrap-up/buffer is separate from this 120-minute session. No internal extra break. Total 120 minutes.
:::
