---
title: "Was It Worth It? Reading a Cost-Benefit Result"
subtitle: "Module 2 · Day 3, Session 1: The Assumption That Decides the Answer"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ===========================================================================
# The cost-benefit model for GreenWaste
# ===========================================================================
# Everything here is shown to the room on a slide. The brief is explicit that
# the trainer displays the calculations rather than asking participants to
# produce them, so the arithmetic is visible and checkable rather than asserted.
#
# The two inputs that come from earlier in the week:
#   the annual saving per business, which is the DiD estimate from Day 2 S1
#   the number of businesses, which is the enrolled count from Day 2 S3
# The one input that is a pure programme fact:
#   the cost per business of delivering it

COST_PER_BUSINESS <- 1800      # AED, one-off cost of putting a business through
BENEFIT_PER_YEAR  <- 1000      # AED per year, from the difference-in-differences
BENEFIT_LAG       <- 1         # no saving in year 1: rollout and adoption take time
ENROLLED          <- 2964      # businesses enrolled, from the matching session

DISCOUNT_BASE <- 0.05          # the rate the authority uses by default
HORIZON_BASE  <- 5             # years of benefit assumed by default

# Present value of 1 AED a year, from year (lag+1) to year (lag + years).
pv_annuity <- function(r, years, lag = BENEFIT_LAG) {
  if (r == 0) return(years)
  a <- function(n) (1 - (1 + r)^(-n)) / r
  a(lag + years) - a(lag)
}

# The benefit-cost ratio: present value of everything we get, over what we spend.
bcr <- function(r = DISCOUNT_BASE, years = HORIZON_BASE,
                cost = COST_PER_BUSINESS, benefit = BENEFIT_PER_YEAR) {
  benefit * pv_annuity(r, years) / cost
}

bcr_headline <- bcr()
pv_per_biz   <- BENEFIT_PER_YEAR * pv_annuity(DISCOUNT_BASE, HORIZON_BASE)

# Total programme figures, for the scale of the decision.
total_cost    <- COST_PER_BUSINESS * ENROLLED
total_benefit <- pv_per_biz * ENROLLED
total_net     <- total_benefit - total_cost

# ---- The two analyst choices, alone and together ---------------------------
# Choice 1: the discount rate. Choice 2: how long the benefit is assumed to last.
rate_grid  <- c(0.03, 0.05, 0.08, 0.12, 0.20)
years_grid <- c(2, 3, 5, 8, 10)

grid_tbl <- expand_grid(years = years_grid, r = rate_grid) %>%
  mutate(bcr = map2_dbl(r, years, ~ bcr(.x, .y)),
         cell = sprintf("%.2f", bcr),
         below = bcr < 1)

# A tidy long form for the heat table on the slide.
grid_wide <- grid_tbl %>%
  mutate(rate_lab = sprintf("%.0f%%", 100 * r),
         year_lab = sprintf("%d years", years)) %>%
  select(year_lab, rate_lab, cell) %>%
  pivot_wider(names_from = rate_lab, values_from = cell)

# Where the ratio crosses 1 for each horizon. This is the number that makes the
# session's point: the flip is not exotic, it happens inside normal practice.
flip_rate <- function(yrs) {
  r <- 0
  while (bcr(r, yrs) >= 1 && r < 2) r <- r + 0.001
  if (bcr(r, yrs) < 1) r * 100 else NA_real_
}
years_for_flip <- c(2, 3, 4, 5)
flip_tbl <- tibble(years = years_for_flip, flip = map_dbl(years_for_flip, flip_rate))

# The rate alone, over the base horizon. This shows the rate is NOT enough on its
# own: even at 30% the ratio is still above 1. That negative result is worth a
# slide, because a commissioner's instinct is to blame the discount rate.
rate_only <- tibble(
  r = c(0, 0.03, 0.05, 0.08, 0.10, 0.15, 0.20, 0.30, 0.40),
  bcr = map_dbl(c(0, 0.03, 0.05, 0.08, 0.10, 0.15, 0.20, 0.30, 0.40),
                ~ bcr(.x, HORIZON_BASE))
) %>% mutate(lab = sprintf("%.0f%%", 100 * r))

# ---- What the ratio would be if the benefit were smaller -------------------
# The Day 2 S1 interval ran from 962 to 1,038 AED. Carrying that forward shows
# the cost-benefit ratio inherits the uncertainty, rather than adding new.
benefit_range <- tibble(
  benefit = c(962, 1000, 1038),
  bcr = map_dbl(c(962, 1000, 1038), ~ bcr(DISCOUNT_BASE, HORIZON_BASE,
                                          benefit = .x))
)

# ---- The three-question framework, scored on this programme ---------------
framework <- tribble(
  ~Question, ~Asks,
  "Are the benefits plausible?",
  "The saving is the difference-in-differences estimate. It is measured, and it is local to the businesses compared.",
  "Are all the costs included?",
  "1,800 AED per business is the delivery cost. It does not include the authority's own set-up, or the businesses' time.",
  "What would flip the result?",
  "Assuming the saving lasts two years instead of five. That alone takes the ratio below break-even."
)
```

# 1 · The headline

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">three questions &middot; reading a number &middot; what a fair comparison is</div>
  <div class="k">Day 2</div>  <div class="v">four ways to estimate an effect: descriptives, DiD, RDD, matching</div>
  <div class="k">Today</div>  <div class="v">what the effect is <em>worth</em> &middot; cost, and the assumptions that decide it</div>
</div>
```

::: {.panel}
Everything so far has been about **how big the effect is**. Today is about
whether it is **worth the money** — and that is a different question with a
different set of judgement calls.
:::

::: {.ask}
A programme can have a real effect and still not be worth funding. You saw the
shape of that all week: real, and not enough.

Today the bar is a ratio instead of a threshold.
:::

## The decision, and the ratio

:::::: {.columns}

::::: {.column width="54%"}
GreenWaste cost the authority **`r gfmt(COST_PER_BUSINESS)` AED per business** to
deliver, and saved each participating business about **`r gfmt(BENEFIT_PER_YEAR)`
AED a year**.

The standard way to compare those is a **benefit-cost ratio**: what you get,
divided by what you spend, with future money counted as worth less than money
today.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Cost</div>     <div class="v">1,800 AED per business, one-off</div>
  <div class="k">Benefit</div>  <div class="v">1,000 AED per business per year</div>
  <div class="k">Decision rule</div> <div class="v">A ratio above 1 means it pays for itself</div>
  <div class="k">Ratio found</div> <div class="v">2.29</div>
</div>
```
:::::

::::::

## The headline number

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Benefit-cost ratio, GreenWaste</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">AED spent per business</span></div>
  <div><span class="n">%s</span><span class="k">AED returned per business</span></div>
  <div><span class="n">%s</span><span class="k">AED net, across the programme</span></div>
</div>', bcr_headline, gfmt(COST_PER_BUSINESS), gfmt(pv_per_biz), gfmt(total_net)))
```

::: {.panel}
**2.29 means: for every AED spent, about 2.29 AED comes back.** On that reading
this is not a borderline decision. It is clearly worth doing.
:::

::: {.warn}
The ratio is the answer to a question you did not choose. Somebody decided the
benefit lasts a certain number of years, and decided how much future money is
worth. Those two choices built this number.
:::

## `[MENTI PLACEHOLDER]` Would you approve it?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
On a ratio of **2.29**, what would you do?

Approve &nbsp;·&nbsp; Reject &nbsp;·&nbsp; I need to know something else first
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
Commit before we go further. If most of the room picks "I need to know
something else first", ask two people **what** — write the answers up, and
compare them with the three questions at the end of the session.
:::

## What went into that number

This is the arithmetic, on the screen.

```{r}
tibble(
  Step = c("Cost per business",
           "Annual saving",
           "Years of saving assumed",
           "Discount rate assumed",
           "Value of the savings, today's money",
           "Benefit-cost ratio"),
  Working = c("one-off delivery cost",
              "from the difference-in-differences estimate",
              "5, beginning one year after rollout",
              "5% a year",
              sprintf("1,000 x %.3f = %s AED", pv_annuity(DISCOUNT_BASE, HORIZON_BASE),
                      gfmt(pv_per_biz)),
              sprintf("%s / %s", gfmt(pv_per_biz), gfmt(COST_PER_BUSINESS))),
  Value = c(sprintf("%s AED", gfmt(COST_PER_BUSINESS)),
            sprintf("%s AED", gfmt(BENEFIT_PER_YEAR)),
            "5 years", "5%",
            sprintf("%s AED", gfmt(pv_per_biz)),
            sprintf("%.2f", bcr_headline))
) %>%
  kable(align = "llr", col.names = c("Step", "Working", "Value")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(3:4, bold = FALSE, background = "#F7F1E8") %>%
  row_spec(6, bold = TRUE, background = "#E6F1F5")
```

::: {.ask}
Two rows are shaded. They are not costs and they are not benefits. They are
**choices** — and they are the subject of the next twenty minutes.
:::

# 2 · The assumptions that decide it

## The saving does not arrive today

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">A saving next year is worth less than a saving today.<br>How much less is a choice.</div>
</div>')
```

::: {.panel}
Money you have now can be used, invested or saved. Money promised in three
years cannot. Discounting is the arithmetic of that difference, and the
**discount rate** is how fast you decide future money loses value.
:::

::: {.ask}
Nobody debates that future money is worth less. They debate **how much less** —
and that number goes straight into the ratio you are shown.
:::

## Choice 1: the discount rate, on its own

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.7, fig.width=5.6}
rate_only %>%
  plot_ly(x = ~lab, y = ~bcr, type = "bar",
          marker = list(color = ifelse(rate_only$bcr < 1, ALERT, ACCENT),
                        line = list(width = 0)),
          text = ~sprintf("%.2f", bcr), textposition = "outside",
          textfont = list(size = 12, color = INK), showlegend = FALSE,
          hovertemplate = "%{x} discount: ratio %{y:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Discount rate", zeroline = FALSE),
         yaxis = ax("Benefit-cost ratio", range = c(0, 3.3)),
         shapes = list(list(type = "line", x0 = -0.5, x1 = 8.5,
                            y0 = 1, y1 = 1, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 8.5, y = 1, yref = "y", xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "break-even", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_WC, height = 260)
```
:::::

::::: {.column width="48%"}
::: {.warn}
**The discount rate on its own does not flip this decision.**

Even at a 30% rate — far higher than any agency uses — the ratio is
`r sprintf("%.2f", rate_only$bcr[rate_only$r == 0.30])`. You have to reach
**40%** to push it below 1, and by then the assumption is not defensible.
:::

::: {.panel}
So if someone tells you the ratio is fragile because of the discount rate, that
is only half true. It is the *other* choice that does the damage.
:::
:::::

::::::

## Choice 2: how long the saving lasts

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.7, fig.width=5.6}
tibble(years = years_grid, bcr = map_dbl(years_grid, ~ bcr(DISCOUNT_BASE, .x))) %>%
  plot_ly(x = ~factor(years), y = ~bcr, type = "bar",
          marker = list(color = ifelse(map_dbl(years_grid, ~ bcr(DISCOUNT_BASE, .x)) < 1,
                                       ALERT, ACCENT), line = list(width = 0)),
          text = ~sprintf("%.2f", bcr), textposition = "outside",
          textfont = list(size = 13, color = INK), showlegend = FALSE,
          hovertemplate = "%{x} years: ratio %{y:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Years of saving assumed (at a 5% discount rate)", zeroline = FALSE),
         yaxis = ax("Benefit-cost ratio", range = c(0, 5)),
         shapes = list(list(type = "line", x0 = -0.5, x1 = 4.5,
                            y0 = 1, y1 = 1, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 4.5, y = 1, yref = "y", xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "break-even", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_WC, height = 260)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The Day 2 sessions measured the saving over **one period**. Nothing measured how
long it lasts.

If it lasts **two years**, the ratio is
`r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))` — **below break-even**.
:::

::: {.warn}
Five years is a reasonable assumption. So is two. Both are guesses about a
future nobody has observed, and they produce opposite recommendations from the
same evidence.
:::
:::::

::::::

## Both choices together

```{r}
grid_wide %>%
  kable(align = "lrrrrr", col.names = c("Years of saving", "3%", "5%", "8%", "12%", "20%")) %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  row_spec(1, bold = TRUE, background = "#F7E9E6") %>%
  row_spec(3, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = "Shaded: the two cells the session turns on. Red = below break-even. Every cell uses the same cost and the same measured annual saving; only the two analyst choices differ.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.ask}
**Red row:** two years of saving. **Blue row:** five years, the base case.

Read the two rows against each other. The recommendation reverses, and the
evidence never changed. Only the horizon assumption did.
:::

## Where the ratio crosses 1

```{r}
flip_tbl %>%
  mutate(`Discount rate at which it flips` =
           sprintf("%.1f%%", flip)) %>%
  select(`Years of saving` = years, `Discount rate at which it flips`) %>%
  kable(align = "rr") %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  footnote(general = "Below this rate the ratio exceeds 1; above it, the programme does not pay for itself.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.warn}
Read the first row. **If you assume the saving lasts two years, a discount rate
of 4.4% flips the conclusion.** A 3% rate is normal practice in some agencies; a
5% rate is normal in others. Both are defensible. They give opposite answers.
:::

::: {.ask}
So the honest statement is not "the ratio is 2.29". It is:

**"The ratio is 2.29 under assumptions that the evidence does not determine, and
plausible alternatives put it anywhere between 0.71 and 4.60."**
:::

## What the ratio is actually telling you

```{r fig.height=3.1, fig.width=11.5}
grid_tbl %>%
  mutate(r_lab = sprintf("%.0f%%", 100 * r),
         y_lab = sprintf("%d years", years),
         r_lab = factor(r_lab, levels = sprintf("%.0f%%", 100 * rate_grid)),
         y_lab = factor(y_lab, levels = sprintf("%d years", rev(years_grid)))) %>%
  plot_ly(x = ~r_lab, y = ~y_lab, z = ~bcr, type = "heatmap",
          colorscale = list(c(0, "#8E2413"), c(0.34, "#E3C9A8"),
                            c(0.36, "#CFE4E9"), c(1, "#0E6E80")),
          zmin = 0.6, zmax = 4.6,
          text = ~sprintf("%.2f", bcr), texttemplate = "%{text}",
          textfont = list(size = 15, color = INK),
          hovertemplate = "%{y}, discount %{x}: ratio %{z:.2f}<extra></extra>",
          colorbar = list(title = "ratio", thickness = 14, len = 0.8)) %>%
  layout(xaxis = ax("Discount rate", zeroline = FALSE),
         yaxis = ax("Years of saving assumed", zeroline = FALSE)) %>%
  pl_m2(height = 300)
```

::: {.ask}
Every cell is the same programme, the same cost, and the same measured saving.

**Only two judgement calls differ.** Find the cells below 1.0. Now ask yourself
which cell the authority would put in the press release, and whether anyone
would mention the others.
:::

# 3 · The three questions

## How to judge any cost-benefit claim

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Are the benefits plausible?</div>
  <div class="k">2</div> <div class="v">Are all the costs included?</div>
  <div class="k">3</div> <div class="v">What would flip the result?</div>
</div>
```

::: {.panel}
Three questions, and none of them needs a spreadsheet. They are the same shape
as the questions every other session has ended on: ask what the number is
compared to, ask what is missing, ask what it would take to change the answer.
:::

::: {.ask}
Question 3 is the one that separates a reader from a recipient. It is also the
one an evaluator is least often asked.
:::

## Question 1: are the benefits plausible?

:::::: {.columns}

::::: {.column width="52%"}
```{r}
benefit_range %>%
  mutate(`Annual saving assumed` = sprintf("%s AED", gfmt(benefit)),
         `Benefit-cost ratio` = sprintf("%.2f", bcr)) %>%
  select(-benefit, -bcr) %>%
  kable(align = "rr") %>%
  kable_styling(full_width = FALSE, font_size = 16)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The saving here is **not** assumed. It is the difference-in-differences estimate
from Day 2, and the three rows show its confidence interval carried forward.

The ratio moves from `r sprintf("%.2f", min(benefit_range$bcr))` to
`r sprintf("%.2f", max(benefit_range$bcr))` across the interval. So the cost-benefit
result inherits the uncertainty already measured.
:::

::: {.warn}
That is the **good** news about this number. The bad news is the horizon, which
has no interval because nobody measured it.
:::
:::::

::::::

## Question 2: are all the costs included?

```{=html}
<div class="facts">
  <div class="k">Included</div> <div class="v">1,800 AED per business to deliver the programme</div>
  <div class="k">Not included</div> <div class="v">the authority's own set-up and administration</div>
  <div class="k">Not included</div> <div class="v">the time the businesses spent on it</div>
  <div class="k">Not included</div> <div class="v">anything the programme displaced</div>
</div>
```

::: {.ask}
Every cost left out makes the ratio **look better**. None of them appears in the
table you were shown.

Ask an evaluator what is in the cost line. A ratio built on a partial cost is not
a conservative estimate of anything.
:::

## Question 3: what would flip it?

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
You already know the answer for this programme, and it is not the discount rate.

**Assume the saving lasts two years instead of five, and the ratio falls from
`r sprintf("%.2f", bcr_headline)` to `r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))`.**

The programme goes from clearly worth doing to not paying for itself.
:::

::: {.ask}
So the question to ask is not "what discount rate did you use?"

It is: **"How long do you assume the benefit lasts, and what evidence do you
have for that?"**
:::
:::::

::::: {.column width="46%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value" style="font-size:1.9em;">%.2f &nbsp;&rarr;&nbsp; %.2f</div>
  <div class="label">The same evidence, two horizons</div>
</div>
<div class="stat-row">
  <div><span class="n">5 yrs</span><span class="k">Assumption A</span></div>
  <div><span class="n">2 yrs</span><span class="k">Assumption B</span></div>
</div>', bcr_headline, bcr(DISCOUNT_BASE, 2)))
```
:::::

::::::

# 4 · Put AI to work

## AI Snapshot: find the scenario that breaks it

::: {.panel}
Groups: paste the cost-benefit table and this prompt.

**"Generate one plausible scenario in which this programme's benefit-cost ratio
falls below 1. Then tell me whether that scenario is realistic."**
:::

::: {.ai-output}
**AI response:**

The ratio falls below 1 if the discount rate is raised. [At a discount rate of
around 40%, the ratio drops below break-even]{.flag}, which would mean the
programme does not pay for itself and should not be funded.

[Alternatively, if the programme's benefits are delayed by a few years, the
ratio also falls below 1]{.flag}. Both scenarios are [realistic and commonly
observed in infrastructure appraisal]{.flag}.
:::

::: {.ask}
Check each claim against the table. Which is right, which is wrong, and which is
true but useless? Four minutes.
:::

## Debrief: right arithmetic, wrong lever

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The arithmetic is correct.** A 40% discount rate does push this ratio below 1.
The model checks out.
:::

::: {.warn}
**1 · Wrong lever.** 40% is not a rate any agency uses. It answered the question
it was asked — *name a scenario* — rather than the useful one: *name a plausible
one*. **Arithmetically true and practically irrelevant.**
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**2 · The second scenario is vague where it needed to be specific.** "Delayed by
a few years" is not the same as a longer lag. The lever that actually works here
is the **horizon**, and the model never named it.

**3 · "Commonly observed in infrastructure appraisal."** Confident, unsourced,
and about a different kind of programme. It reads as authority it does not have.
:::
:::::

::::::

::: {.panel}
**This is a new failure shape.** The model did not invent a number or misread a
table. It answered the literal question competently and missed the point.

**The lesson: ask for the plausible, not the possible.** A model will find the
scenario you asked for, and it will be technically true, and it may tell you
nothing.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Generate one plausible scenario in which this benefit-cost ratio falls below 1.

The word "plausible" does the work, and it is not defined. The model reaches for
whatever is easiest to compute.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Here is a cost-benefit model. The cost and annual saving are measured; the
> discount rate and the number of benefit years are assumptions. Tell me which
> **single assumption** has the largest effect on the ratio, quantify it, and
> say whether that assumption is one a government agency would normally make.
> **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both. The second asks the model to **rank** the assumptions before choosing
one, which is exactly the step the first was allowed to skip.
:::

# 5 · The Abu Dhabi case

## `[PLACEHOLDER]` The cost-benefit case

::: {.warn}
**Awaiting the real findings.** The Abu Dhabi cost-benefit case for the outline
has not been shared with this workspace yet, so nothing here is presented as its
result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">[PLACEHOLDER]</div>
  <div class="k">Ratio claimed</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Cost per unit</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Horizon assumed</div><div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed: the ratio, what went into it, and the two assumptions
that decide it. When the findings arrive they drop into these fields without
redesigning the session.
:::

## Applying the three questions

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Are the benefits plausible? Measured, modelled, or assumed?</div>
  <div class="k">2</div> <div class="v">Are all the costs included? What was left out, and does it flatter the ratio?</div>
  <div class="k">3</div> <div class="v">What would flip it? Which assumption, and by how much?</div>
</div>
```

::: {.ask}
Same three questions we just ran on GreenWaste, in the same order. The programme
changes; the interrogation does not.

**In groups:** run them on whatever the real case turns out to be, and present
your verdict in two minutes.
:::

# 6 · Close

## What today gives you

```{=html}
<div class="facts">
  <div class="k">A ratio is a model</div>  <div class="v">not a measurement &middot; it contains choices as well as evidence</div>
  <div class="k">The rate alone</div>       <div class="v">rarely flips anything &middot; do not be satisfied with that answer</div>
  <div class="k">The horizon</div>          <div class="v">often decides it &middot; and nobody measures it</div>
  <div class="k">Always ask</div>           <div class="v">what is excluded from the cost, and what would change the answer</div>
</div>
```

::: {.ask}
The single sentence to take to a meeting:

**"What assumption would have to change for this ratio to fall below 1?"**

An evaluator who can answer that has thought about their model. One who cannot
has run it.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">is the number answering your question, and is it big enough?</div>
  <div class="k">Day 2</div>  <div class="v">which method produced it, and what does that method assume?</div>
  <div class="k">Today</div>  <div class="v">what is it worth, and which assumption decides that?</div>
  <div class="k">This afternoon</div> <div class="v">reading the figures that carry these claims into a room</div>
</div>
```

::: {.panel}
Tomorrow morning you use all of it: a full credibility review of an evaluation
report, then a clinic where you put the questions to an analyst who has to answer
them.
:::
