---
title: "Reading a Regression Discontinuity Result"
subtitle: "Module 2 · Day 2, Session 2: What Happens at the Cut-off"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html.
#
# Built from Fiona's Oct13_session2.qmd (Sep 24), with one correction that
# changes the numbers: the RDD uses businesses in the OFFERED neighbourhoods
# only, as Module 1 did. In the other neighbourhoods an "eligible" business was
# offered nothing, so pooling them waters the jump down (to about -249 at +/-2)
# and makes a sharp design look fuzzy. Here the rule is sharp and take-up is
# complete. One example all session: the Abu Dhabi case slot is dropped.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  render-df: kable
  resources:
    - evaluation_data_GreenWaste.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(knitr)
library(estimatr)
library(qrcode)

gfmt  <- function(x) format(round(x), big.mark = ",")
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_d2s2.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0)); plot(qr_code(url, ecl = "H")); path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000
CUT     <- 58

# ---- The numbers the static slides quote (live cells recompute them) -------
gw  <- read.csv("./evaluation_data_GreenWaste.csv")
off <- subset(gw, round == 1 & treatment_neighborhood == 1)   # offered neighbourhoods
off$distance <- off$efficiency_index - CUT
off$below    <- as.integer(off$efficiency_index <= CUT)

at <- function(d, width) {
  d <- subset(d, abs(distance) <= width)
  f <- lm_robust(waste_management_costs ~ below * distance, data = d)
  list(est = coef(f)[["below"]], ci = confint(f)["below", ], n = nrow(d),
       slope = coef(f)[["distance"]])
}
w2 <- at(off, 2)                                   # the +/-2 window: about -791
w2_least <- min(abs(w2$ci))                        # interval end closest to zero

# Module 1's two numbers: every business in offered neighbourhoods, a straight
# line either side, without and with covariates.
covs <- c("age_manager", "age_deputy", "educ_manager", "educ_deputy", "female_manager",
          "foreign_owned", "staff_size", "advanced_filtration", "water_treatment_system",
          "business_area", "recycling_center_distance")
m1_plain <- coef(lm(waste_management_costs ~ below + distance, data = off))[["below"]]
m1_cov_fit <- lm_robust(reformulate(c("below", "distance", covs), "waste_management_costs"), data = off)
m1_cov    <- coef(m1_cov_fit)[["below"]]
m1_cov_ci <- confint(m1_cov_fit)["below", ]

# The pooled version (all neighbourhoods) that the earlier deck used.
all1 <- subset(gw, round == 1)
all1$distance <- all1$efficiency_index - CUT
all1$below    <- as.integer(all1$efficiency_index <= CUT)
pooled2 <- at(all1, 2)
```

# 0 · Where Module 1 left us {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What you saw in Module 1

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the live cells
# below use is defined here, so each cell on a slide stays a few lines long.
gw <- read.csv("evaluation_data_GreenWaste.csv")
library(estimatr)
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)
# Warm up estimatr now: its first call in the browser is slow, later ones fast.
invisible(lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1))))

CUT <- 58
prep <- function(d) {
  d$distance <- d$efficiency_index - CUT          # the score measured from the line
  d$below    <- as.integer(d$efficiency_index <= CUT)   # 1 = eligible side
  d
}
offered <- prep(subset(gw, round == 1 & treatment_neighborhood == 1))
everyone <- prep(subset(gw, round == 1))

clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.2)

# The jump at the line, inside a window of +/- width points, with its interval.
jump_at <- function(d, width, cut = CUT) {
  d$distance <- d$efficiency_index - cut
  d$below    <- as.integer(d$efficiency_index <= cut)
  d <- subset(d, abs(distance) <= width)
  f <- lm_robust(waste_management_costs ~ below * distance, data = d)
  ci <- confint(f)["below", ]
  data.frame(cut_off = cut, window = width, businesses = nrow(d),
             jump = round(coef(f)[["below"]]), lower = round(ci[1]), upper = round(ci[2]),
             row.names = NULL)
}

# Average cost in half-point bins, a fitted line on each side, and the jump.
jump_plot <- function(width = 4, view = 10) {
  d <- subset(offered, abs(distance) <= view)
  d$bin <- floor(d$distance * 2) / 2 + 0.25            # half-point bins
  bins <- aggregate(waste_management_costs ~ bin, data = d, FUN = mean)
  names(bins) <- c("bin", "cost")
  clean_par()
  plot(bins$bin + CUT, bins$cost, pch = 19, cex = 1.3, bty = "n",
       col = ifelse(bins$bin < 0, ACCENT, ACCENT2),
       xlab = "Efficiency index (eligible: 58 or below)", ylab = "Waste costs, AED")
  abline(v = CUT, col = MUTED, lty = 2)
  w <- subset(offered, abs(distance) <= width)
  f <- lm(waste_management_costs ~ below * distance, data = w)
  for (s in 0:1) {
    xs <- if (s == 1) seq(-width, 0, length.out = 20) else seq(0, width, length.out = 20)
    lines(xs + CUT, predict(f, data.frame(below = s, distance = xs)),
          col = if (s == 1) ACCENT else ACCENT2, lwd = 4)
  }
  top <- coef(f)[["(Intercept)"]]
  segments(CUT, top, CUT, top + coef(f)[["below"]], col = ALERT, lwd = 6)
}

# Does a characteristic jump at the line? Same regression, a covariate as outcome.
balance <- function(width) {
  vars <- c(manager_age = "age_manager", staff = "staff_size",
            area = "business_area", recycling_distance = "recycling_center_distance")
  d <- subset(offered, abs(distance) <= width)
  do.call(rbind, lapply(names(vars), function(v) {
    f  <- lm_robust(reformulate(c("below", "distance", "below:distance"), vars[[v]]), data = d)
    ci <- confint(f)["below", ]
    data.frame(characteristic = v, jump = round(coef(f)[["below"]], 2),
               lower = round(ci[1], 2), upper = round(ci[2], 2),
               differs = ci[1] > 0 | ci[2] < 0, row.names = NULL)
  }))
}

# Businesses per one-point slice of the index, 54 to 62.
density_plot <- function() {
  slice <- subset(offered, efficiency_index >= 54 & efficiency_index < 62)
  clean_par()
  hist(slice$efficiency_index, breaks = 54:62, col = ifelse(54:61 < 58, ACCENT, ACCENT2),
       border = "white", main = "", xlab = "Efficiency index", ylab = "Businesses")
  abline(v = 58, col = ALERT, lwd = 3, lty = 2)
}
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Module 1 compared businesses either side of the eligibility cut-off, in the
neighbourhoods where GreenWaste was offered, and reported two numbers.
:::

::: {.ask}
From memory: why does a comparison at a cut-off count as fair?
:::
:::::

::::: {.column width="50%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Module 1, straight lines</div> <div class="v">%s AED</div>
  <div class="k">Module 1, with covariates</div> <div class="v">%s AED</div>
  <div class="k">The rule</div>                  <div class="v">at least 1,000 AED</div>
</div>', gfmt(m1_plain), gfmt(m1_cov)))
```

::: {.panel .fragment}
Businesses just either side of the line are near-twins. Which side they land on
is close to chance, so the line creates the comparison.
:::
:::::

::::::

::: {.notes}
About 3 minutes, recall only. Open the deck a couple of minutes before the
session: R and its statistics package take about a minute to load in the
browser, after which every live cell runs in a few seconds.
:::

## Today we open the box

::: {.panel}
Why did adding covariates move the answer from `r gfmt(m1_plain)` to
`r gfmt(m1_cov)`? Today we rebuild the estimate from the data and run the four
checks an RDD result has to pass.
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">The rule and the data: who was offered, who took part</div>
  <div class="k">2</div> <div class="v">From the picture to a number</div>
  <div class="k">3</div> <div class="v">Four checks: the window, bending the rule, balance, who it applies to</div>
</div>
```

::: {.warn}
You will not write code. You call the choices; I type and press **Run**.
:::

# 1 · The rule and the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The rule that decided who got in

:::::: {.columns}

::::: {.column width="54%"}
In the offered neighbourhoods, GreenWaste went to businesses whose **efficiency
index** was **at or below 58**: the less efficient ones.

Two businesses either side of that line are almost identical. One got the
programme, one did not.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Running variable</div> <div class="v">Efficiency index</div>
  <div class="k">Cut-off</div>          <div class="v">58</div>
  <div class="k">Eligible</div>         <div class="v">at or below 58</div>
  <div class="k">Rule followed?</div>   <div class="v">we check it next</div>
</div>
```
:::::

::::::

::: {.warn}
**Read the direction carefully.** The rule offers the programme to the **less**
efficient businesses, so the eligible side is on the **left** of the chart.
Getting this backwards flips the sign of everything that follows.
:::

## `[MENTI PLACEHOLDER]` Predict the picture

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask}
Businesses at or below 58 got the programme, and it cut costs. On the chart of
costs against the index, **where is the eligible side, and at the line does it
sit above or below the other side?**

**A** Left, lower · **B** Left, higher · **C** Right, lower · **D** Right, higher
:::

Go to **menti.com**, code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='260px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel .fragment}
**A.** A low index sits on the left, and low is eligible; a programme that cut
costs puts the eligible side **below** at the line. Anyone who assumed the
programme went to the *better* businesses would read the chart backwards.
:::

::: {.notes}
Two predictions in one vote: the direction of the rule and the sign of the
effect. Keep the result on screen and point back to it on the next chart.
:::

## Did the rule hold?

```{webr}
data.frame(side = c("above 58", "58 or below"),
           businesses = as.vector(table(offered$below)),
           offered    = as.vector(tapply(offered$eligible, offered$below, mean)),
           took_part  = as.vector(tapply(offered$enrolled, offered$below, mean)))
```

::: {.ask}
Before I press Run: what share of eligible businesses do you expect took part?
:::

::: {.panel .fragment}
Every business at or below 58 was offered and took part; none above 58 did. The
rule is **sharp**, and the jump at the line is the effect of taking part.
:::

::: {.notes}
Read the two rows aloud. `offered` and `took_part` are shares: 1 means every
business on that side, 0 means none.
:::

## Which businesses belong in the comparison?

```{ojs}
//| echo: false
viewof sample = Inputs.radio(["Offered neighbourhoods only", "Every neighbourhood"],
  {value: "Offered neighbourhoods only"})
```

```{webr}
#| input:
#|   - sample
#| autorun: true
#| edit: false
d <- if (sample == "Every neighbourhood") everyone else offered
data.frame(jump_at_the_line = jump_at(d, width = 2)$jump,
           eligible_who_took_part = round(mean(d$enrolled[d$below == 1]), 2))
```

::: {.ask}
Switch to every neighbourhood. **Why does the jump shrink, and why does take-up
fall to about half?**
:::

::: {.warn .fragment}
In the other neighbourhoods an "eligible" business was offered nothing. Mixing
them in puts untreated businesses on the eligible side, so the jump falls from
about `r gfmt(w2$est)` to `r gfmt(pooled2$est)` AED. Ask the evaluator exactly
which businesses are in the comparison.
:::

::: {.notes}
An earlier version of this deck made this mistake and reported -249. Module 1
used the offered neighbourhoods only.
:::

## The picture the method is built on

```{webr}
#| autorun: true
#| edit: false
#| fig-width: 11
#| fig-height: 3.6
jump_plot(width = 4)
```

::: {.ask}
Dots are average costs in half-point bins. **Is there a step at the dashed line?**
:::

# 2 · From the picture to a number {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The idea behind the method

::: {.panel}
Compare businesses **just below** the cut-off, who got the programme, with
businesses **just above** it, who did not.
:::

::: {.panel}
Right at the line, scoring 57.9 or 58.1 is close to a matter of chance, so the
two sides are alike except for the programme. The **step in costs at the
cut-off** is the estimated effect.
:::

::: {.notes}
Point back at the previous chart: the step is the gap between the two clouds of
dots at the dashed line.
:::

## A first, rough number

```{webr}
near <- subset(offered, abs(distance) <= 2)          # within 2 points of the line
eligible_side <- mean(near$waste_management_costs[near$below == 1])
other_side    <- mean(near$waste_management_costs[near$below == 0])
round(data.frame(eligible_side, other_side, gap = eligible_side - other_side))
```

::: {.ask}
Each pair: call out the jump on your card. **Why are they so different?** Then
Run: the same subtraction on every business within 2 points.
:::

::: {.panel .fragment}
Six businesses a side is too few to trust. And even with many, an average across
the window mixes in the slope of costs along the index. The regression deals
with both.
:::

::: {.notes}
Paper (Oct13_session2_materials.docx): Before pressing Run, hand out the jump cards (pack pages 1-4, four different cards A-D; neighbouring pairs get different cards). Pairs average six businesses on each side and subtract. Write every pair's jump on the flipchart (they range from about -170 to -1,260), ask why they differ, then press Run: the regression uses the 773 businesses within two points.
:::

## Running the regression

```{webr}
near <- subset(offered, abs(distance) <= 2)
fit <- lm_robust(waste_management_costs ~ below * distance, data = near)
round(as.data.frame(summary(fit)$coefficients[, c(1, 2, 4)]), 1)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
`distance` is the score measured from the line (0 = 58); `below` is 1 on the
eligible side. `below * distance` fits a line on each side.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Only one row is the programme's effect. Which one?
:::

::: {.panel .fragment}
**`below`: the jump, `r gfmt(w2$est)` AED.** `distance`, about
`r sprintf("%+.0f", w2$slope)` AED per point, is the slope **along** the line,
not the step **at** it.
:::
:::::

::::::

## Seeing the jump: the red line is `below`

```{webr}
#| autorun: true
#| edit: false
#| fig-width: 11
#| fig-height: 3.4
jump_plot(width = 2, view = 4)
```

::: {.panel}
A line on each side, fitted to businesses within 2 points of 58. The red segment
at the line is the `below` row, `r gfmt(w2$est)` AED. The lines slope up, which
is why the raw average overstated the gap.
:::

## Does it clear the rule?

```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED at the cut-off, within 2 points</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">No</span><span class="k">Clears the bar</span></div>
</div>', gfmt(w2$est), gfmt(RULEBAR), gfmt(w2_least)))
```

::: {.ask}
A real effect, clearly not zero. The interval runs from `r gfmt(w2_least)` to
`r gfmt(max(abs(w2$ci)))`. **Would you scale the programme on this?**
:::

::: {.notes}
The interval is wide because only 773 businesses sit within 2 points. Check 1
shows how the answer moves with that choice.
:::

# 3 · Four checks {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## This method's one assumption

```{=html}
<div class="hero">
  <div class="value" style="font-size:1.9em;">Nothing else jumps at the cut-off.</div>
</div>
```

::: {.panel}
A business one point below 58 and one point above should be alike in every way
except whether it got the programme. If something else also changes at 58, the
jump is not only the programme.
:::

::: {.ask}
This morning's difference-in-differences assumed the groups moved in parallel;
this method assumes nothing else changes at the line. **If something else did
change at 58, where in the data would you see it?**
:::

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the four-checks scorecard (pack page 5, one per group). Groups mark each check as you run it, starting with check 0 (who is compared) from Section 1.
:::

## Check 1: does the jump survive a different window?

```{ojs}
//| echo: false
viewof width = Inputs.range([1, 10], {value: 2, step: 1, label: "Window either side of 58 (index points)"})
```

```{webr}
#| input:
#|   - width
#| autorun: true
#| edit: false
jump_at(offered, width)
```

::: {.ask}
The window is the analyst's choice. Slide from 1 to 10. **Does the verdict against
1,000 AED change?**
:::

::: {.warn .fragment}
The estimate moves between about 790 and 1,120 AED, above the bar from a window
of 5. The least generous end of the interval never reaches 1,000 (at most about
960). Ask which window
was used, and why.
:::

## Check 1, continued: a jump where there is no rule?

```{ojs}
//| echo: false
viewof placebo = Inputs.range([48, 68], {value: 62, step: 2, label: "Pretend the cut-off is at"})
```

```{webr}
#| input:
#|   - placebo
#| autorun: true
#| edit: false
jump_at(offered, width = 2, cut = placebo)
```

::: {.ask}
Move the cut-off to 50, 54, 62, 66, where no rule exists. **What should the jump
be?**
:::

::: {.panel .fragment}
Close to zero, and every interval there contains zero. At 58 the jump is far
outside that range. If the method found jumps where there is no programme, its
jump at 58 would mean nothing.
:::

## Check 2: did anyone bend the rule?

```{webr}
#| fig-width: 10
#| fig-height: 2.6
density_plot()
```

::: {.ask}
If businesses could nudge their score to 58 or below to qualify, what would the
bar just left of the line look like?
:::

::: {.panel .fragment}
Taller than its neighbours, with a dip just right. The bars at the line are not
out of pattern: no sign of a pile-up. Ask whether anyone drew this chart; a
pile-up only shows if someone looks.
:::

## Check 3: are the two sides alike?

```{ojs}
//| echo: false
viewof bwidth = Inputs.range([1, 8], {value: 2, step: 1, label: "Window (index points)"})
```

```{webr}
#| input:
#|   - bwidth
#| autorun: true
#| edit: false
balance(bwidth)
```

::: {.ask}
Each row runs the same regression with a business characteristic as the outcome.
If the sides are alike, no row should jump. Change the window. **Which row does
not behave?**
:::

::: {.warn .fragment}
**Manager age**: businesses just below the line have managers 5 to 9 years
younger, at every window. Something other than the programme changes at 58.
This check fails.
:::

## What that row means

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
If younger managers also run their waste costs differently, part of the jump is
the managers, not the programme. That is why Module 1's estimate with covariates
(`r gfmt(m1_cov)`) differs from the one without (`r gfmt(m1_plain)`).
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Adjusting for manager age only fixes what was measured. If manager age jumps at
58, something unmeasured may jump too.
:::

::: {.ask .fragment}
What would you ask the evaluator? *Why does manager age change at the cut-off,
and what else might?*
:::
:::::

::::::

## Check 4: who does the result apply to?

```{webr}
near <- abs(offered$distance) <= 2
data.frame(group = c("within 2 points of 58", "everyone else"),
           businesses = c(sum(near), sum(!near)),
           staff = round(c(mean(offered$staff_size[near]), mean(offered$staff_size[!near])), 1),
           area  = round(c(mean(offered$business_area[near]), mean(offered$business_area[!near])), 2))
```

::: {.panel}
The estimate describes the businesses near the line: somewhat smaller, with
smaller premises. It does not describe the businesses far from it.
:::

::: {.ask .fragment}
If the plan is to extend the programme to everyone, what have you learned about
the businesses the extension would reach?
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: who does this apply to?

:::::: {.columns}

::::: {.column width="42%"}
::: {.panel}
Paste the regression table and the four checks into an AI tool with this
prompt:

**"Explain, without jargon, who this result actually applies to."**
:::

::: {.ask}
Mark every claim the study supports. Four minutes.
:::
:::::

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

This study found that the GreenWaste programme reduced costs by around
**790 AED** for participating businesses. [Because the study uses a cut-off at
an efficiency index of 58, the findings apply to all businesses in the
programme]{.flag}. [The cut-off design means the result generalises to the wider
population of inefficient businesses]{.flag}. The estimate is therefore a
reasonable basis for [scaling the programme to every business below average
efficiency]{.flag}.
:::
:::::

::::::

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the AI handout (pack page 6, one per group).
:::

## Debrief: one error, three times

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "Applies to all businesses in the programme."** The estimate is local to
businesses near the line.
:::

::: {.warn}
**2 · "Generalises to the wider population."** Backwards: RDD uses only the
businesses closest to the line, so it generalises least.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "A basis for scaling to every business."** A scale-up needs what happens
away from the cut-off, which this design discards.
:::

::: {.panel}
And it never mentions the manager-age jump. The question it was asked, *who does
this apply to?*, is the one it answered least well.
:::
:::::

::::::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain, without jargon, who this result actually applies to.

Nothing about the design, so the model may reach for the most natural-sounding
story.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> This is a regression discontinuity result with the cut-off at an efficiency
> index of 58, estimated within 2 points either side. The businesses near the
> line are smaller than average, and manager age differs across the line. Name
> which businesses the estimate describes and what it cannot tell us. **Ask me
> questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both. Does error 1 survive? What did each answer explain well? The habit:
give the model the constraint it cannot infer, then ask a narrower question.
:::

# 5 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Your turn: ask AI first

::: {.ask}
Before we show our list, ask an AI tool: **"I am a commissioner about to hear an
evaluator present a regression discontinuity result. What are the key questions
I should ask?"** Three minutes, then compare with the next slide.
:::

::: {.notes}
Optional. If time is short, skip to the next slide. If you run it, ask what the
AI missed, what it added that is worth keeping, and whether anything was wrong.
:::

## Five questions to ask any evaluator presenting RDD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, which side is eligible, and exactly which units are compared?</div>
  <div class="k">2</div> <div class="v">Does the jump survive other windows, and is there none at placebo cut-offs?</div>
  <div class="k">3</div> <div class="v">Is the number of units smooth at the line, or did anyone bend the rule?</div>
  <div class="k">4</div> <div class="v">Does anything else jump at the line, such as manager age?</div>
  <div class="k">5</div> <div class="v">Who is near the line, and does the estimate clear our rule for them?</div>
</div>
```

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the take-away cards (pack page 7). Give a minute to fill in "On my desk" on the card, then take two or three examples from the room.
:::

## Where this sits

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons and a randomised trial: -1,014, interval spans 1,000</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences: -816 against a 1,000 rule</div>
  <div class="k">Today</div>    <div class="v">regression discontinuity: %s near the line (%s with covariates), against 1,000</div>
</div>', gfmt(w2$est), gfmt(m1_cov)))
```

::: {.panel}
DiD gets its comparison by assuming parallel trends. RDD gets it from the rule,
and the estimate describes only the businesses near the line. Next session,
matching covers more businesses by assuming everything that matters was
measured.
:::
