---
title: "Read a chart before trusting its story"
subtitle: "Module 2 | Day 3 | Session 2"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day3_case.R')
```

## Which picture would you use? {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
tp <- subset(gw, setting == 'city' & took_part == 1)
```
```{r, results='asis'}
html_out(paste0('<div class="chart-pair"><div><h3>A</h3>',bar_svg(0,FALSE),'</div><div><h3>B</h3>',bar_svg(700,FALSE),'</div></div><p class="caption">Two draft charts of the same records. Labels are missing.</p>',fiction))
```
::: {.ask}
Can you choose responsibly without the labels?
:::


::: {.notes}
4 minutes. Retrieve source and comparison questions. These are deliberately incomplete drafts; accepting either shape alone is premature. Commit a reason. No room poll needed; pairs or local cards. Next screen supplies values and scales.
:::

## Same values different axes {.reader-slide}

```{r, results='asis'}
html_out(paste0('<div id="axis-charts"><div data-variant>',bar_svg(700),'</div><div data-variant hidden>',bar_svg(0),'</div></div><button data-swap="axis-charts">Change the axis start</button>',fiction))
```
::: {.ask}
Which impression changes, and which recorded values stay the same?
:::


::: {.notes}
6 minutes. Predict before toggle. Truncated bar baseline starts at 700 versus 0; true mean values identical. Bars should ordinarily start at 0; a truncated axis needs explicit disclosure and a reason. This is descriptive before/after, not a causal effect chart. Keyboard button must not advance deck.
:::

## Mark up the chart {.reader-slide}

Use the printed chart on **Sheet 1**. Work alone first.

Circle the units. Mark where the axis starts.

Underline who is compared and when.

::: {.ask}
What can this chart support on its own?
:::


::: {.notes}
7 minutes. Chart is full axis, participant before/after. Ask for source, annual AED, period 12 months and no causal comparison. Distribution and uncertainty absent, so do not infer everyone gained. The worksheet also includes its descriptive source caption.
:::

## What an average hides {.reader-slide}

```{r, results='asis'}
html_out(sprintf('<p class="distribution-mean">Mean: %s tonnes</p><button data-open="distribution" data-closed="Open the individual values" data-opened="Hide the individual values" aria-controls="distribution" aria-expanded="false">Open the individual values</button><div id="distribution" hidden>%s</div>%s',fmt(mean(ten$landfill_before),1),distribution_svg(),fiction))
```
::: {.ask}
Does the mean describe most of these ten businesses?
:::


::: {.notes}
6 minutes. Predict before revealing Day 1 ten actual landfill records. Nine below the mean; one at 46 tonnes. Dot stacking makes repeated values visible. This is a small teaching selection, not a representative city distribution or a programme effect.
:::

## A percentage of what? {.reader-slide}

```{r, results='asis'}
html_out(sprintf('<div class="percentage-pair"><div><h3>Change in total tonnes</h3><p>%s%%</p></div><div><h3>Mean business percentage</h3><p>%s%%</p></div></div>%s',fmt(percent_total,1),fmt(percent_average,1),fiction))
```
```{webr}
#| echo: true
#| autorun: false
total_fall <- 100 * (1 - sum(tp$landfill_after) /
  sum(tp$landfill_before))
round(total_fall, 1)
```
::: {.ask}
Why can these percentages differ for the same businesses?
:::


::: {.notes}
7 minutes. One trainer demo after prediction. Total percentage weights businesses by their starting tonnes; mean business percentage gives each business equal weight. All before tonnes are positive. Two valid different summaries; neither alone establishes cause. Shorten variableline if editorclips. No school-zone or pie-chart case.
:::

## Read an estimate and its interval {.reader-slide}

```{r, results='asis'}
html_out(interval_svg('trial',reveal=TRUE))
```
```{=html}
<button data-open="interval-trial" data-closed="Open the interval" data-opened="Hide the interval" aria-controls="interval-trial" aria-expanded="false">Open the interval</button>
```
::: {.ask}
Does the pilot chart settle the minimum saving rule?
:::


::: {.notes}
6 minutes. Predict before interval opens. Pilot lottery after 12 months, annual AED saving. Interval crosses 1000. Teach units, compared population and reported uncertainty; intervals do not account for every bias. Use the existing computed SVG, not a new hand-coded result.
:::

## Choose and repair {.reader-slide}

Pairs work on **Sheet 2**.

Repair a chart caption that claims cause from a before/after fall.

Name the source, units, comparison and one limit.

::: {.ask}
Would a decision-maker understand what the chart actually shows?
:::


::: {.notes}
10 minutes. Protect repair task. Accept explicit source title and reference to worksheetfigure. No drawing software or code. Repairing caption alone cannot supply missing causal comparison; say so.
:::

## Check an AI caption {.reader-slide}

Illustrative AI draft:

"GreenWaste caused a fall of almost half in every business's landfill waste. The before/after chart proves the programme works."

::: {.ask}
Which words would you keep, qualify or remove?
:::
```{=html}
<button data-open="chart-ai" data-closed="Open feedback" data-opened="Hide feedback" aria-controls="chart-ai" aria-expanded="false">Open feedback</button><div id="chart-ai" class="feedback" hidden>The total fell by about half among participants. Every business is unsupported, and this descriptive comparison cannot establish cause.</div>
```


::: {.notes}
6 minutes. Authored example applies to the total landfill result, not the cost bars. Participants verify source and denominator. Optional AI prompt asks it to describe visible evidence and missing context before recommending. No account required.
:::

## Defend your chart choice {.reader-slide}

::: {.ask}
What can your repaired caption help a reader conclude?
:::
Give one supported statement and one limit.

Point to the chart evidence while explaining.

::: {.notes}
4 minutes. Pairs share a short defence. Require the actual values, comparison and denominator; accept different captions when faithful.
:::

## Your question from memory {.reader-slide}

::: {.ask}
What would you check first in a chart from an evaluation report?
:::
Close the sheet. Write the check and explain why it matters.

::: {.notes}
4 minutes. Independent close. Scales and labels, averages, comparison and uncertainty. Total 60minutes.
:::
