---
title: "GreenWaste: read the chart and its claim"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    include-in-header:
      text: |
        <style>
        body{font:16px/1.4 Arial,sans-serif;color:#111418;margin:0;background:#eef2f6}main{max-width:850px;margin:auto}h1,h2,h3{color:#063360}h1{font-size:26px}h2{font-size:22px;margin:0 0 12px}h3{font-size:18px;margin:14px 0 6px}p{margin:8px 0}.sheet-page{background:white;padding:28px 40px;margin:20px 0;box-sizing:border-box;min-height:950px}.reading-table{width:100%;border-collapse:collapse;font-size:13px}.reading-table th,.reading-table td{padding:6px;border-bottom:1px solid #D5DEE8;text-align:left}.reading-table thead{background:#063360;color:white}.writing{height:45px;border-bottom:1px solid #c8d6e5;margin:6px 0}.writing.tall{height:75px}.writing.caption-space{height:110px}.review-svg{display:block;width:100%;height:175px}.review-grid{display:grid;grid-template-columns:1fr 1fr;gap:15px}.review-grid h3{font-size:16px}.chart-pair .review-svg{height:180px}.caption{font-size:12px;color:#545860}.ai-source{border-left:3px solid #7da1c4;padding-left:16px}.prompt{border-top:2px solid #215a9e;padding-top:8px;font-size:14px}header button{font:inherit;padding:8px 15px;color:#063360;background:white;border:1px solid #215a9e;cursor:pointer}
        @page{size:A4;margin:14mm}@media print{body{background:white;font-size:11pt;line-height:1.3}main{max-width:none}.sheet-page{padding:0;margin:0;min-height:0;break-after:page}.sheet-page:last-child{break-after:auto}header,#title-block-header{display:none}h2{font-size:17pt}h3{font-size:12pt}.writing{height:10mm}.writing.tall{height:17mm}.writing.caption-space{height:24mm}.reading-table{font-size:9pt}.reading-table th,.reading-table td{padding:5px}.review-svg{height:165px}.chart-pair .review-svg{height:175px}.caption{font-size:9pt}.prompt{font-size:10pt}}
        </style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('day3_session2_review.R', encoding='UTF-8')
```

<header><p>Separate Day 3 Session 2 review worksheet. Work on pages 3-4 as concepts are taught. Page 7 holds numerical references: open after the two-business task. Page 6 is the AI fallback: hold until Stage 1 is finished.</p><button onclick="window.print()">Print worksheet</button></header>

<div class="sheet-page">

## Page 1 | H8-A §1: read the cost bars

**GreenWaste is a fictional training case.** This supplied source is the reference; there is no separate report to find.

### Figure C | Same participant means, two bar baselines

```{r, results='asis'}
html_out(d8_pair())
```

### Table C | Underlying values

```{r, results='asis'}
html_out(d8_cost_table())
```

**Source caption:** Mean annual waste cost in AED per business among the **same 4,794 city participants**, before and **12 months after** GreenWaste. Corrected city CSV. The bars compare two times for participants, not participants against nonparticipants. Values are rounded; the recorded mean fall is about **669 AED**.

### What is supported, and what is not?

The bars support a description of the participant mean before and after. They do not establish that GreenWaste caused the fall, that every participant's cost fell, or that the mean represents every business. Individual changes, a credible causal comparison and uncertainty about an estimated effect are not displayed.

### Why does the baseline matter?

A bar's length represents magnitude. A zero baseline normally preserves that visual comparison. Chart B starts at **700 AED**, so its bar lengths represent amounts above 700, not full costs. Labels retain the actual values, but do not necessarily undo a reader's first impression. A nonzero scale can be appropriate in other chart types or clearly justified contexts; it is not automatically dishonest.

The intended sentence and audience matter. A dramatic bar-height drop does not establish a programme effect. A repaired caption cannot create missing causal evidence.

<p class="caption">Source IDs: Figure C and Table C, H8-A §1. Calculations: day3_case.R and day3_session2_review.R. Both charts use the same upper bound, 1,600 AED; only the lower bound differs.</p>

</div>

<div class="sheet-page">

## Page 2 | H8-A §2: distributions, not uncertainty

### Figure D | All city participants' annual before-costs

```{r, results='asis'}
html_out(d8_distribution())
```

This histogram includes **all 4,794 city participants**. Bar heights count businesses within 300 AED bins on one cost scale. Mean **`r fmt(mean(tp$cost_before))` AED**, median **`r fmt(d8_quantiles[2])` AED**, middle 50% **`r paste(fmt(d8_quantiles[c(1,3)]),collapse='-')` AED**, range **`r paste(fmt(range(tp$cost_before)),collapse='-')` AED**. These describe pre-programme business variation, not an estimated effect or confidence interval.

### Figure E | Ten selected actual landfill records

```{r, results='asis'}
html_out(d8_selected())
```

Actual before-landfill values: **3, 4, 5, 5, 6, 7, 7, 8, 9, 46 tonnes** per business per year. Mean **10.0**, median **6.5**, range **3-46**. Nine values are below the mean. Repeated values are stacked as separate dots.

These ten records were selected for teaching and are **not representative of the city**. The outcome is now landfill tonnes, not the AED costs in Figure D. A large value can shift the mean; that alone is not a reason to delete it.

### Two different meanings of spread

A distribution shows variation **between businesses**. The pilot interval on page 7 shows uncertainty **about an estimated mean effect**. Neither the middle 50% here nor the range of these ten records is that confidence interval.

<p class="caption">Source IDs: H8-A §2, Figures D/E. Figure D uses every participant before-cost; Figure E uses the same ten actual records selected earlier in the course. Page 7 changes population and outcome again for the pilot.</p>

</div>

<div class="sheet-page">

## Page 3 | Q1-Q3: annotate and interpret

### Q1 | Mark Figure C, page 1

Circle the **units** and mark each **axis start**. Underline the population, time period and comparison in the caption. Use Table C to check the bars.

What changes when the baseline changes? What stays the same?

<div class="writing tall"></div>

### Q2 | What do the cost bars support?

One supported descriptive statement, with a source ID:

<div class="writing tall"></div>

One specific limit, and the evidence needed to address it:

<div class="writing tall"></div>

### Q3 | Read Figures D and E, page 2

What does the full-participant histogram add to its mean? Explain the middle 50% without calling it a confidence interval.

<div class="writing tall"></div>

Why does the mean of the selected ten landfill records differ from the median? Why must this not be called a representative city result?

<div class="writing tall"></div>

Which changes between Figures D and E: outcome, units or population?

<div class="writing"></div>

</div>

<div class="sheet-page">

## Page 4 | Q4-Q7: denominators, caption and judgement

### Q4 | Two invented businesses: calculate before page 7

| Business | Before tonnes | After tonnes | Business fall |
|---|---|---|---|
| Small starting amount | 10 | 2 | __________ % |
| Large starting amount | 90 | 54 | __________ % |

Total before: _____. Total after: _____. Fall in total tonnes: _____ %.

Mean of the two business percentage falls: _____ %. Why are they different?

<div class="writing"></div>

After the actual city summary is shown, explain what **49.5%** and **58.6%** measure and which denominator each uses:

<div class="writing"></div>

### Q5 | Repair the cost-chart caption for the director

Draft: **"GreenWaste substantially reduced every participant's costs."**

Use Figure C. Include population, outcome, units, time, comparison, source and causal limit. Choose a baseline suited to your sentence.

<div class="writing caption-space"></div>

### Q6 | Read the pilot interval [page 7, Figure P]

What uncertainty does the interval represent? Does it securely establish the 1,000 AED annual-saving minimum?

<div class="writing"></div>

### Q7 | Close the source: your first chart check

Name a check and the error it prevents. Give one supported statement and one limit from today's evidence.

<div class="writing tall"></div>

</div>

<div class="sheet-page">

## Page 5 | AI Snapshot: chart, claim and audience (15 minutes)

### Stage 1 | Mark the supplied response (8 minutes)

Naive prompt: **"Which of these two charts better represents the difference, and why?"** The model was supplied Figure C and the draft claim. This is an **illustrative prepared response**, not a live Copilot transcript.

<div class="ai-source">
```{r, results='asis'}
html_out(paste0('<p>',gsub('\n\n','</p><p>',d8_ai,fixed=TRUE),'</p>'))
```
</div>

Keep supported readings. Cross out or qualify unsupported advice and inferences. Check whether each sentence answers the actual claim/audience question.

<div class="writing"></div>

### Stage 2 | Use the better prompt in Microsoft Copilot (7 minutes)

<div class="prompt">
```{r, results='asis'}
html_out(paste0('<p>',d8_prompt,'</p>'))
```
</div>

Use fictional evidence only and the shared **AI good-practice card**. No device: open page 6 after Stage 1.

One verified correction and one useful missing-evidence question:

<div class="writing tall"></div>

</div>

<div class="sheet-page">

## Page 6 | Prepared paper alternative

**Open only after marking page 5.** This is an example, not the exact response Copilot must give.

### Questions before choosing the display

What decision should the director make from this figure? Is there credible causal evidence elsewhere, or must this sentence remain descriptive? Is the claim about an average, every participant, or programme-wide value for money?

### Check the numbers and the visual claim

H8-A §1, Table C and Figure C show the same before/after mean annual costs, about **1,432 and 763 AED**, among the same **4,794 city participants**. The recorded mean fall is about **669 AED**.

Chart B's 700 AED lower bound makes the after bar look small relative to the before bar. Its labels are numerically correct, but that does not make the dramatic shape appropriate beside a causal/every-participant claim for a non-technical reader.

For a bar comparison of these full cost magnitudes, Chart A's zero baseline is a safer choice. Other chart types may use a clearly justified nonzero axis; it is not necessary to prohibit all nonzero scales.

### Correct the sentence, not just the chart

"Among 4,794 city participants, mean annual waste cost was about 1,432 AED before and 763 AED 12 months after GreenWaste. Source: H8-A §1, corrected city records. This descriptive before/after comparison does not establish the programme's causal effect or a reduction for every participant."

H8-A §2 shows variation in before-costs, not individual changes or uncertainty about a programme effect. To assess an expansion decision, request the relevant causal comparison and its assumptions, target population and economic evidence. A clearer plot cannot provide those missing facts.

### Verify the alternative yourself

One factual reading checked against its source:

<div class="writing tall"></div>

One reason the chart choice fits the corrected claim, and one evidence request:

<div class="writing tall"></div>

</div>

<div class="sheet-page">

## Page 7 | H8-A §§3-4: numerical references

**Open after completing the two-business calculation in Q4.** These references are not the prepared AI response on page 6.

### §3 | Hypothetical percentage example and actual city summary

```{r, results='asis'}
html_out(d8_toy_table(TRUE))
```

Invented example: total tonnes fall from **100 to 56**, or **44%**. Mean business percentage fall is **(80% + 40%)/2 = 60%**. Total-tonne weighting gives **0.1 × 80% + 0.9 × 40% = 44%**. Every starting denominator here is positive.

```{r, results='asis'}
html_out(d8_land_table())
```

Actual **same 4,794 city participants**, landfill tonnes before and 12 months after. Total before **`r fmt(sum(tp$landfill_before),1)` tonnes**, total after **`r fmt(sum(tp$landfill_after),1)` tonnes**. Every before value is positive. Neither summary alone establishes cause, nor a fall for every business. One summary weights by starting tonnes; the other gives every business equal weight.

### §4 | Figure P: separate pilot, annual cost saving

```{r, results='asis'}
html_out(d8_pilot())
```

This is the separate **400-business pilot**, lottery-assigned 200 to join and 200 to wait, comparing annual waste costs after 12 months. Estimated mean reduction **1,014 AED**, 95% interval **862-1,167 AED**, crossing the fictional **1,000 AED annual-saving minimum**. The interval is uncertainty about the estimated mean effect, not the distribution of individual savings; it does not account for every source of bias.

<p class="caption">Source IDs: H8-A §3 (two-business illustration, city summaries) and §4 Figure P (pilot). Calculations from day3_case.R/day3_session2_review.R. Do not attach these landfill percentages or pilot intervals to the descriptive cost bars on page 1.</p>

</div>
