---
title: "How similar is similar enough?"
subtitle: "Day 2 | Session 3: Matching, measured balance and missing influences"
author: "3ie · Review version"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, day2_session3_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session1_review_interactions.html]
webr:
  render-df: kable
  resources: [evaluation_data_GreenWaste_simple.csv]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day2_session3_review.R', encoding='UTF-8')
```

## Which comparison would you trust? {#d2s3-p01 .reader-slide .s1-review .d2s3-review}

The city did **not** allocate GreenWaste by lottery. Participation followed the efficiency-score rule.

Matching selects comparison businesses using characteristics recorded **before** the programme.

::: {.ask}
If two businesses look similar in the records, what can still differ?
:::

Annual waste cost in **AED per business**, measured 12 months after. The full-city study is not the separate pilot or the cutoff study.

::: {.notes}
5 minutes. Retrieve comparison/population distinctions from DiD and RDD. Matching asks about participants compared with selected nonparticipants, not the local jump. Start webR warming. Programme is equipment subsidies, installation help and training. Participants interpret, never write code.
:::

## Similar on what we can see {#d2s3-p02 .reader-slide .s1-review .d2s3-review}

<div class="review-grid"><div><h3>Business A: took part</h3><p>Manager age: 45<br>Size: small</p></div><div><h3>Business 2: did not take part</h3><p>Manager age: 46<br>Size: small</p></div></div>

::: {.ask}
Would you pair these businesses? Name one reason and one unknown.
:::

**Invented profiles** for the activity. Do not open the cost slips.

::: {.notes}
6 minutes. Two minutes individual, two pair discussion, two collect. Similar recorded features do not mean identical businesses. These two cards teach the choice, not the actual four-feature metric. Keep outcomes withheld.
:::

## Find the closest recorded comparisons {#d2s3-p03 .reader-slide .s1-review .d2s3-review}

Lay out the **seven profile cards**. Pair participants **A, B and C** with comparison businesses **1-4**.

1. Use manager age and size; use each comparison card once.
2. Record your pairs and reasons in **Q1, worksheet page 3**.
3. Identify the unused comparison. Keep costs closed.

::: {.ask}
Where is the pairing convincing? Where would you ask for more information?
:::

::: {.notes}
14 minutes. Distribute only profile page 1 of the separate cards. Optional seven volunteers with eighth group observing. Intended A2 B1 C3, comparison 4 unused. Ten minutes pair/record, four compare reasoning. Do not rematch on outcomes. Toy uses no replacement; full study permits reuse. This is a physical change of activity, not an additional scheduled break.
:::

## Open the costs, not a new pairing {#d2s3-p04 .reader-slide .s1-review .d2s3-review}

The trainer now distributes the **separate cost slips**.

For each pair: **participant annual after-cost minus comparison annual after-cost**.

Record the three differences and their mean in **Q1, page 3**.

<button data-open="d6-toy" aria-controls="d6-toy" aria-expanded="false" data-closed="Check the arithmetic" data-opened="Hide the arithmetic">Check the arithmetic</button>
<div id="d6-toy" class="feedback" hidden>A-2: 900 - 1,900 = -1,000. B-1: 1,400 - 2,300 = -900. C-3: 700 - 1,800 = -1,100. Mean: -1,000 AED.</div>

::: {.ask}
Does the arithmetic establish what GreenWaste caused?
:::

::: {.notes}
8 minutes. Work one pair then allow group calculation and reveal. These are invented costs, not city results. Correct subtraction is not causal identification. Unequal unrecorded influences could explain the gap. Pair choices must remain fixed.
:::

## From seven cards to the full city {#d2s3-p05 .reader-slide .s1-review .d2s3-review}

1. Record **manager age, staff, premises area and filtration** before GreenWaste.
2. Put the four features on a common scale.
3. Select the nearest available nonparticipant for each participant.
4. Compare annual after-costs, allowing a control to be used again.

::: {.ask}
What does "nearest" guarantee? What does it not guarantee?
:::

**No maximum acceptable distance** was imposed; no participant was excluded. Before-cost was **not** a matching feature.

::: {.notes}
7 minutes. Explain standardisation as distance in city standard-deviation units, not four equal raw units. Squared distances choose a nearest neighbour; no propensity score is used. Matching with replacement differs from toy. Q2 page 3. Nearest is relative to available records; not necessarily adequate similarity. Transition to reuse.
:::

## One comparison business can be reused {#d2s3-p06 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_reuse())
```

**4,794 participants; 585 unique comparison businesses.** The most-used control appears **440 times**.

::: {.ask}
Why are 4,794 matched pairs not 9,588 independent businesses?
:::

::: {.notes}
5 minutes. Diagram is illustrative, actual counts are computed. There are 9,588 rows in stacked matched sample but 5,379 unique businesses. Repeated controls contribute repeatedly to the comparison mean and require appropriate uncertainty accounting. Do not confuse a reused control with 440 separate measurements. Q2 page 3.
:::

## What changed in measured balance? {#d2s3-p07 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_balance())
```

::: {.ask}
Q3, page 4: which gaps narrowed, and which concern remains?
:::

The matched-control column counts **reused records**. The last column is the signed mean gap divided by the full-city standard deviation.

::: {.notes}
7 minutes. Reference table is worksheet page 2. Full-city SD is descriptive scaling, not a universal pass/fail threshold or standard error. Filtration mean is share with filtration. Age still differs by about 2.97 years; before-cost differs by about 161 AED. Cost was not matched on. Similar means are not identical distributions or evidence that all bias is absent.
:::

## Closest available may still be too far {#d2s3-p08 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_support())
```

Actual **pre-programme age distributions**; bar heights are within-group proportions on the same scale.

::: {.ask}
What else would you need to judge whether matches are close enough?
:::

Hypothetical: age 45 matched to 46 versus 45 matched to 70. Both could be "nearest" in different pools.

::: {.notes}
7 minutes. Actual plot uses all available controls, not matched controls; see page 2. It is a one-feature marginal illustration, not a joint four-feature overlap diagnosis. The 45/46 and 45/70 cases are hypothetical, not claimed city pairs. Request distance distributions, weak-support regions and sensitivity to an acceptable-distance rule. No caliper was applied here. Q3 page 4.
:::

## Read the full-city difference {#d2s3-p09 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_table(data.frame(
  'Result row'='Participant minus matched control',
  'Estimate AED'=fmt(match_A$est),
  '95% interval AED'=paste(fmt(match_A$ci), collapse=' to '),
  check.names=FALSE)))
```

::: {.ask}
Q2: explain the sign, comparison, period and units in one sentence.
:::

This is an **after-cost difference**, not a before/after change or a lottery estimate.

::: {.notes}
6 minutes. H6-A section 2, page 1. About 1,029 AED lower annual after-cost among city participants than selected controls. Business-clustered robust interval accommodates reuse, treating matches as given. It does not fully account for match selection or unmeasured bias. Population is the participants, not every national business.
:::

## The trainer computes the paired difference {#d2s3-p10 .reader-slide .s1-review .d2s3-review}

```{webr}
#| context: setup
#| autorun: true
#| include: false
#| echo: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
city <- subset(gw, setting == 'city')
X <- scale(city[, c('manager_age','staff','area','filtration')])
participants <- which(city$took_part == 1)
controls <- which(city$took_part == 0)
Xc <- t(X[controls, , drop=FALSE])
paired_difference <- function() {
  twins <- controls[vapply(participants, function(i)
    which.min(colSums((Xc-X[i, ])^2)), 1L)]
  mean(city$cost_after[participants]-city$cost_after[twins])
}
```

```{webr}
#| echo: true
# Compare each participant with its selected control.
# Average annual after-cost differences, in AED.
round(paired_difference())
```

Paper result: **-1,029 AED**. Participants do not need to run R.

::: {.ask}
Does this compute a difference, or prove the causal assumption?
:::

::: {.notes}
5 minutes. Predict, then one trainer Run. Hidden setup uses same standardisation, nearest neighbour and replacement as shared source. Calculation reproduces coefficient, not displayed robust interval. If browser is slow use paper result and maintain allocation. No participant coding.
:::

## Leave age out. Change the comparison. {#d2s3-p11 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_results())
```

Full-feature age gap: **`r fmt(match_A$age_gap, 1)` years**. Without age: **`r fmt(match_B$age_gap, 1)` years**.

::: {.ask}
Q4, page 4: why is a larger estimated reduction not necessarily better evidence?
:::

::: {.notes}
6 minutes. Same city, same algorithm, three rather than four features. Controls change, not merely a row label. Baseline cost gaps 161 versus 540 AED. No-age result can balance retained features while missing age. Difference does not reveal true effect; fuller analysis also has limitations. Do not compare interval widths as overall credibility.
:::

## What if motivation is not recorded? {#d2s3-p12 .reader-slide .s1-review .d2s3-review}

**Hypothetical influence, not a measured GreenWaste variable:**

Motivation could affect **joining** and also **cost-saving practices**.

Matching on age, staff, area and filtration would not directly remove that path.

::: {.ask}
What would have to be true for the matched difference to represent a programme effect?
:::

After accounting for recorded features, there must be **no remaining important difference affecting both participation and costs**, with suitable comparisons available.

::: {.notes}
6 minutes. Q4 page 4. Spell out no important unrecorded confounding; not proof or a claim that motivation actually drove results. Consistent measurement and no important spillovers also matter. Better measured balance cannot certify this assumption. Ask for selection process, omitted predictors and sensitivity evidence.
:::

## Read the source before trusting the summary {#d2s3-p13 .reader-slide .s1-review .d2s3-review}

Read **H6-A, worksheet page 1**, and the balance reference on **page 2**.

In **Q5, page 4**:

1. Circle a finding and cite its paragraph/table.
2. Underline the causal assumption.
3. Request one missing check and explain its purpose.

::: {.ask}
What does "everyone matched" establish, and what remains unknown?
:::

::: {.notes}
8 minutes. Four independent reading, two partner comparison, two debrief. H6-A section 2 supports coverage/reuse/result; section 3 assumption; section 4 missing distances/sensitivity and conditional interval. No external full report required. Accept concrete consequential requests, not generic more data.
:::

## Above the rule. Certain enough? {#d2s3-p14 .reader-slide .s1-review .d2s3-review}

```{r, results='asis'}
html_out(d6_interval())
```

::: {.ask}
Q5: distinguish "the estimate exceeds 1,000" from "the minimum is securely met".
:::

The interval crosses the fictional **1,000 AED annual-saving minimum**. Causal credibility and full programme costs remain separate questions.

::: {.notes}
7 minutes. Negative difference interval -1,200 to -859 becomes reduction 859 to 1,200. Estimate 1,029 above rule but uncertainty straddles it. Interval is not guarantee, individual range, probability of passing rule or allowance for all bias. Allow act/conditions/request with defensible reason, not mandatory policy answer.
:::

## AI Snapshot: what should a commissioner flag? {#d2s3-p15 .reader-slide .s1-review .d2s3-review}

Naive prompt: **"Compare these two matching evaluations and advise the commissioner."** Source: H6-A and balance table.

<div class="ai-reading">
```{r, results='asis'}
html_out(paste0('<p>', gsub('\n\n', '</p><p>', d6_ai, fixed=TRUE), '</p>'))
```
</div>

::: {.ask}
Page 5: keep supported claims; cross out assurances the source cannot justify.
:::

::: {.notes}
8 minutes. Stage 1 of 15-minute Snapshot. Illustrative supplied response, not a live Copilot transcript. Three minutes individual, three discuss, two debrief. Keep features, coverage/reuse, estimates, costs caution. Cross out bias removed, closeness proved, rule securely achieved. Fuller model is not automatically definitive. Do not disclose prepared response yet.
:::

## Ask Copilot for a reading you can verify {#d2s3-p16 .reader-slide .s1-review .d2s3-review}

<div class="better-prompt">
```{r, results='asis'}
html_out(paste0('<p>', d6_prompt, '</p>'))
```
</div>

::: {.ask}
Use the full prompt on page 5. Compare Copilot's answer with H6-A, not with its confidence.
:::

No device: use the **prepared alternative on page 6** after marking the first response.

::: {.notes}
7 minutes. Stage 2. Two paste/run, three verify one corrected claim and one missing evidence request, two share. Participants use Microsoft Copilot; no setup required. Use fictional source only, shared AI good-practice card; do not upload real sensitive records. No second-chat task. Paper alternative is prepared, not an expected exact response. Check source citations, numeric interval and actual questions before recommendation.
:::

## Close the source. Make a defensible request. {#d2s3-p17 .reader-slide .s1-review .d2s3-review}

Write in **Q6, page 4**, without the source:

- One interpretation of the matched result.
- One strength and one consequential concern.
- Act, act with conditions, or ask first: state why.
- One matching-specific evidence request and how it could change your judgement.

::: {.ask}
Which method's comparison best answers your next decision?
:::

**Lottery:** assigned groups. **DiD:** extra change. **RDD:** fitted jump at the rule. **Matching:** selected recorded similarities.

::: {.notes}
8 minutes. Three individual, three pair challenge, two collect targeted feedback. Accept different proportionate actions. These methods answer different comparisons/populations; do not average their estimates or call pilot ground truth for city. Total 120 minutes; daily 14:45-15:00 wrap-up/buffer is separate and needs no additional material.
:::
