---
title: "Compared to what"
subtitle: "Module 2 · Day 1 · Session 1"
author: "3ie"
always_allow_html: true
format:
  revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---



```{r setup, include=FALSE}
source('day1_case.R')
```


## A report has reached your desk {.reader-slide}

::: {.lead}
“Landfill waste from businesses in GreenWaste fell `r fmt(landfill_fall)`% in a year.”
:::

::: {.ask}
Would this sentence justify extending the programme?
:::

Write your decision and one question on **Sheet 1**. Work alone first.

Fictional GreenWaste case · invented figures

::: {.notes}
7 minutes. This is the entry task, not a lecture. Do not define terms or correct answers yet. Collect a few decisions and questions. Keep the sheets so participants can compare their own reading at the end of the week. No AI for this task.
:::

## Meet GreenWaste {.reader-slide}

{{< include _case_map.qmd >}}

Fictional case · today we read **waste sent to landfill**, in tonnes per business per year.

::: {.notes}
3 minutes. Use the existing case picture/card. GreenWaste advises businesses and helps them improve waste handling. One pilot used a lottery; the rest of the city used a score rule. Do not teach either method yet or reveal cost findings. The map shows an illustrative arrangement, not actual counts. The corrected city records follow the displayed rule: score 58 or below took part.
:::

## Would you extend it? {.reader-slide}

::: {.big-number}
`r fmt(landfill_fall)`% lower
:::

Landfill among businesses that took part, one year later.

```{=html}
<div data-choices>
  <button data-choice aria-pressed="false">Extend</button>
  <button data-choice aria-pressed="false">Do not extend</button>
  <button data-choice aria-pressed="false">Ask first</button>
  <p role="status" class="feedback"></p>
</div>
```

Fictional GreenWaste case · invented figures

::: {.notes}
3 minutes. Individual choice, then a reason. Buttons record only the choice in this open page. For shared phone voting use room_poll.py and its facilitator page; the room poll is separate from this local interaction. Keep results hidden until voting closes. Paper fallback: three labelled cards or a show of hands. Revisit the same question after opening the comparison.
:::

## What is missing? {.reader-slide}

::: {.lead}
“Landfill fell after GreenWaste.”
:::

::: {.ask}
What would you need to know before saying GreenWaste caused the fall?
:::

Write one question. Compare it with a partner’s.

::: {.notes}
4 minutes. Invite concrete questions, not a list of technical terms. Prompt only if needed: what else changed, what happened elsewhere, how businesses were selected? Keep the next slide hidden until participants have committed.
:::

## Open the comparison {.reader-slide}

```{r landfill-comparison, results='asis'}
html_out(sprintf('<div class="evidence-pair"><div class="evidence"><h3>Took part</h3><p class="big-number">%s &rarr; %s</p><p>Before &rarr; one year later</p></div><div class="evidence"><h3>Did not take part</h3><button data-open="other-landfill" data-closed="Open their result" data-opened="Hide their result" aria-controls="other-landfill" aria-expanded="false">Open their result</button><div id="other-landfill" hidden><p class="big-number">%s &rarr; %s</p><p>Before &rarr; one year later</p></div></div></div><p class="caption">Landfill, tonnes per business per year</p>%s', fmt(landfill['before'],1), fmt(landfill['after'],1), fmt(landfill['other_before'],1), fmt(landfill['other_after'],1), fiction))
```

::: {.ask}
What can the first fall tell us on its own?
:::

::: {.notes}
5 minutes. Ask participants to predict the other group’s direction before opening it. The other group also improved. This weakens a causal reading of before/after; it does not prove that these groups are a valid causal comparison. They started at different levels. Ask the same extension question again using the room poll’s After phase or cards. Invite a changed reason, not a winning answer.
:::

## Compared to what? {.reader-slide}

::: {.lead}
“Waste was lower **than it had been before**.”
:::

The **counterfactual** is what would have happened without GreenWaste.

::: {.ask}
Does this sentence tell us what would have happened without it?
:::

::: {.fragment}
Ask how the evaluator estimated that missing outcome.
:::

::: {.notes}
5 minutes. Point to the comparison words. Explain counterfactual once in plain language. We cannot observe the same business, at the same time, both with and without the programme. A comparison can help estimate the missing outcome if its assumptions are credible. Avoid launching a methods survey. Have pairs add comparison words to one claim.
:::

## What does an average hide? {.reader-slide}

Landfill before GreenWaste, tonnes per year, for ten businesses:

```{r ten-businesses, results='asis'}
html_out(paste0('<div class="ten-businesses">', paste(sprintf('<span>%s</span>', fmt(ten$landfill_before)), collapse=''), '</div>', fiction))
```

::: {.ask}
Would one number describe these businesses well?
:::

::: {.fragment}
The **mean** is the total divided by ten: `r fmt(mean(ten$landfill_before),1)` tonnes. Nine businesses are below it.
:::

::: {.notes}
5 minutes. Participants point to the large business before revealing the mean. No long arithmetic exercise. A mean can answer a useful total-related question, but it does not describe every business. Ask which extra information they would want about the spread. Do not introduce quartiles or a second lesson on medians.
:::

## Try four claims {.reader-slide}

Read the four claim cards. For each, choose:

**Act · Ask first · Do not act**

Then name what is missing:

**Compared to what? · How big? · How sure?**

::: {.ask}
What question would move your decision forward?
:::

::: {.notes}
12 minutes. Preserve Fiona’s four cards and triage activity. Give one card set per group and use Sheet 2. Optional A1 board: one group-coloured marker per group; write card numbers into its question/decision grid. No extra AI wall board. Card 2’s causal wording is intentionally open to challenge. Technical terms on cards can be glossed briefly; the full uncertainty lesson comes next session. Protect time for groups to write their reasons.
:::

## Defend one choice {.reader-slide}

Choose a card that two groups placed differently.

::: {.ask}
Which words or numbers support your choice?
:::

Explain one missing piece of evidence that could change it.

::: {.notes}
6 minutes. Discuss two disagreements, not a tour through every group. Expected primary placements: card 1 compared to what/ask first; card 2 how big/ask first (it concerns landfill, not the cost rule; also ask about the causal comparison); card 3 do not act on the nationwide recommendation; card 4 how sure/ask first. More than one missing question can be justified. Do not treat grid placement as a single-answer test.
:::

## Check the AI sentence {.reader-slide data-correct="The second sentence claims a cause the headline has not established. Ask what happened without GreenWaste."}

An illustrative AI draft. Select the unsupported sentence.

```{=html}
<div class="claim-list">
<button data-claim="supported" aria-pressed="false">Landfill fell among businesses that took part.</button>
<button data-claim="unsupported" aria-pressed="false">GreenWaste caused the entire fall.</button>
</div>
<button data-check>Check against the evidence</button>
<p data-feedback class="feedback" role="status"></p>
```

::: {.ask}
What should the AI ask before writing that conclusion?
:::

::: {.notes}
6 minutes. This is an authored example of a possible AI error, not a transcript of a verified tool response. Participants must point to source evidence. Optional live AI: provide only the headline and ask it to list the missing information before offering a conclusion. Review its questions yourselves. No AI account is required to complete the activity.
:::

## Your next question {.reader-slide}

Close the cards and turn over the sheet.

::: {.ask}
What would you ask the evaluator before acting on a claim of improvement?
:::

Write one question and explain why the answer matters.

::: {.notes}
4 minutes. Individual retrieval, no list or AI. Look for a concrete comparison/counterfactual question, a decision-relevant size question or a useful uncertainty question, with a reason. Collect two answers. Session total 60 minutes. Do not add a summary lecture.
:::
