---
title: "What Can You Read? What Might Be Wrong?"
subtitle: "Module 2 · Day 3, Session 2: Reading the Figures That Carry the Claim"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ===========================================================================
# The figures this session interrogates
# ===========================================================================
# Every figure is drawn from the same three datasets the week has already used,
# so participants are reading real structure rather than decoration. The
# session's method is to show the SAME data twice and ask what changed.

# ---- GreenWaste: the two groups at baseline -------------------------------
gw  <- read.csv("./evaluation_data_GreenWaste.csv")
base <- gw %>% filter(treatment_neighborhood == 1, round == 0)
enr  <- base$waste_management_costs[base$enrolled == 1]
non  <- base$waste_management_costs[base$enrolled == 0]

mean_enr <- mean(enr); mean_non <- mean(non)
gap      <- mean_non - mean_enr
pooled   <- sd(c(enr, non))
sd_gap   <- gap / pooled

# ---- The difference-in-differences, carried over from Session 1 -----------
tn <- gw %>% filter(treatment_neighborhood == 1)
grp <- tn %>% group_by(enrolled, round) %>%
  summarise(mc = mean(waste_management_costs), .groups = "drop")
gv  <- function(e, r) grp$mc[grp$enrolled == e & grp$round == r]
ba_claim  <- gv(1, 0) - gv(1, 1)                     # 665, before-and-after
did_est   <- abs((gv(1, 1) - gv(1, 0)) - (gv(0, 1) - gv(0, 0)))   # 816

# ---- The cost-benefit model, from Session 1 --------------------------------
COST_PER_BUSINESS <- 1800
BENEFIT_PER_YEAR  <- 816
BENEFIT_LAG       <- 1
DISCOUNT_BASE     <- 0.05

# Cumulative saving against cumulative cost, nominal and discounted. A payment
# at the end of year y is discounted by (1 + r)^y, and there is no benefit in
# year 1, so the five years of benefit are years 2 to 6.
cb_years <- tibble(
  year      = 1:8,
  nominal   = ifelse(year == 1, 0, BENEFIT_PER_YEAR),
  discounted = ifelse(year == 1, 0, BENEFIT_PER_YEAR / (1 + DISCOUNT_BASE)^year)
) %>%
  mutate(cum_nominal = cumsum(nominal), cum_discounted = cumsum(discounted))

cb_nom_total <- sum(cb_years$nominal[2:6])
cb_dis_total <- sum(cb_years$discounted[2:6])
cb_ratio     <- cb_dis_total / COST_PER_BUSINESS

# ---- School-zone child collisions: the denominator problem -----------------
sz <- read.csv("./evaluation_data_SchoolZoneRCT.csv")
sc <- sz %>% group_by(segment_id, arm) %>%
  summarise(base = child_collisions[round == 0],
            follow = child_collisions[round == 1], .groups = "drop")

zero_base  <- sum(sc$base == 0)
sc_pos     <- sc %>% filter(base > 0) %>%
  mutate(pct = 100 * (follow - base) / base, absolute = follow - base)

# The two different "biggest wins", which do not contain the same segments.
big_pct <- sc_pos %>% arrange(pct) %>% slice_head(n = 6)
big_abs <- sc_pos %>% arrange(absolute) %>% slice_head(n = 6)
overlap <- length(intersect(big_pct$segment_id, big_abs$segment_id))

# The pooled percentage against the average of the percentages, which is the
# difference between weighting by size and giving every segment equal say.
pooled_pct   <- 100 * (sum(sc$follow) - sum(sc$base)) / sum(sc$base)
average_pct  <- mean(100 * (sc$follow - sc$base) / sc$base)

# ---- Traffic cameras: total count against rate ----------------------------
cam <- read.csv("./evaluation_data_TrafficCameras.csv")
sect <- cam %>% group_by(sector_id) %>%
  summarise(collisions = sum(injury_collisions),
            km = sum(road_length_km), .groups = "drop") %>%
  mutate(per_km = collisions / km)

top_count <- sect %>% arrange(desc(collisions)) %>% slice_head(n = 5)
top_rate  <- sect %>% arrange(desc(per_km))     %>% slice_head(n = 5)
sect_overlap <- length(intersect(top_count$sector_id, top_rate$sector_id))

# ---- The severity mix, for the pie ----------------------------------------
sev <- c(minor = sum(sz$minor_collisions),
         injury = sum(sz$injury_collisions),
         child = sum(sz$child_collisions))

# ---- The rules this session teaches ---------------------------------------
# Five habits, each demonstrated on a figure rather than asserted.
fig_rules <- tribble(
  ~n, ~Rule, ~Why,
  "1", "Read the axis before the shape",
       "A truncated axis makes a real gap look decisive. The gap is the same number either way; only your impression of it changes.",
  "2", "Ask what the bar is averaging over",
       "A mean can sit between two groups that barely overlap. It is a summary, not a description of anyone.",
  "3", "Check the denominator",
       "A percentage needs a base. Where the base is zero or tiny, the biggest percentage change belongs to the smallest number.",
  "4", "Compare changes, not levels",
       "Levels carry whatever the two groups already differed by. The comparison the chart invites is often not the one that answers the question.",
  "5", "Discount the future in front of the reader",
       "Undiscounted sums look bigger and are easier to draw. The gap between the two lines is an assumption, not a finding."
)
```

# 1 · Why the picture matters

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">is the number answering your question, and is it big enough?</div>
  <div class="k">Day 2</div>  <div class="v">which method produced it, and what does that method assume?</div>
  <div class="k">This morning</div> <div class="v">what is it worth, and which assumption decides that?</div>
  <div class="k">Now</div>     <div class="v">the form the claim actually reaches a decision-maker in</div>
</div>
```

::: {.panel}
Every session this week has ended with a number and a set of assumptions behind
it. **Almost none of that survives into the room.** What reaches the
decision-maker is a chart, a headline and a summary paragraph.

If the assumption does not survive the translation, it cannot be interrogated.
:::

::: {.ask}
So this session is not about making charts. It is about **reading them as
evidence** — the same interrogation you have practised all week, applied to the
form the evidence is usually delivered in.
:::

## The same data, two pictures

::: {.panel}
Here is one fact: among the businesses that were offered GreenWaste, the ones
that enrolled paid **`r gfmt(mean_enr)` AED** in waste costs at baseline and the
ones that did not paid **`r gfmt(mean_non)` AED**. A gap of
**`r gfmt(gap)` AED**.

That gap is fixed. It is the same number in every figure below. **What changes is
what you think of it.**
:::

```{r fig.height=2.6, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside",
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2400))) %>%
  pl_m2(width = PL_W, height = 250)
```

::: {.ask}
Nothing is wrong with this figure. The axis starts at zero, the values are
labelled, the comparison is fair. **Hold it in mind.**
:::

## The same gap, drawn to fill the frame

```{r fig.height=2.6, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside",
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)",
                    range = c(1300, 2200), zeroline = FALSE)) %>%
  pl_m2(width = PL_W, height = 250)
```

::: {.warn}
**Same numbers. Same labels. Different decision.**

The gap is `r gfmt(gap)` AED in both figures, but here it looks like a chasm. The
y-axis starts at `r gfmt(1300)`, not zero, so a
`r sprintf("%.0f%%", 100 * gap / mean_non)` difference is drawn as most of the
frame.
:::

::: {.ask}
**In groups, two minutes.** Which figure would you put in a briefing to a
minister? Which would you put in an annex? Is your answer the same for both?
:::

## Rule 1: read the axis before the shape

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
A truncated axis is not dishonest by itself. It is a **magnifying glass**, and
magnifying glasses are useful when the question is whether a difference exists at
all.

It becomes dishonest when the reader is not told the glass is there.
:::

::: {.ask}
The practical test, and you can do it in two seconds:

**Cover the bars. Read the y-axis. Uncover.**
:::
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">The gap</div>     <div class="v">630 AED &middot; the same in both figures</div>
  <div class="k">As a share</div>  <div class="v">30% of the non-enrolled mean</div>
  <div class="k">In SD units</div> <div class="v">1.13 standard deviations</div>
  <div class="k">The honest line</div> <div class="v">"a substantial gap on a truncated axis"</div>
</div>
```
:::::

::::::

::: {.ask}
Note that the third line **supports** the claim. A gap of
`r sprintf("%.2f", sd_gap)` standard deviations is genuinely large by any
convention. The problem was never the finding. It was that the figure asked you
to accept the finding on the strength of the drawing rather than the number.
:::

# 2 · What the bar is hiding

## A mean is not a group

::: {.panel}
Here is the same comparison as one number per group. **A bar chart of means.**

It is the most common figure in evaluation reporting and the least informative,
because two groups with the same mean can have nothing else in common.
:::

```{r fig.height=2.5, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside",
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2400))) %>%
  pl_m2(width = PL_W, height = 230)
```

::: {.ask}
**Before you turn the page:** what share of the enrolled businesses do you think
paid **more** than the average non-enrolled business?

Write a number down. Then turn.
:::

## The same comparison, as distributions

```{r fig.height=3.0, fig.width=11.4}
tibble(cost = c(enr, non),
       group = c(rep("Enrolled", length(enr)), rep("Not enrolled", length(non)))) %>%
  plot_ly(x = ~cost, color = ~factor(group, levels = c("Enrolled", "Not enrolled")),
          colors = c(ACCENT, ACCENT2), type = "histogram", nbinsx = 46,
          opacity = 0.72, histnorm = "probability density",
          hovertemplate = "%{x:.0f} AED<extra></extra>") %>%
  layout(barmode = "overlay",
         xaxis = ax("Annual waste cost (AED)"),
         yaxis = ax("Share of the group"),
         shapes = list(
           list(type = "line", y0 = 0, y1 = 1, yref = "paper",
                x0 = mean_enr, x1 = mean_enr, line = list(color = ACCENT, width = 3)),
           list(type = "line", y0 = 0, y1 = 1, yref = "paper",
                x0 = mean_non, x1 = mean_non, line = list(color = ACCENT2, width = 3, dash = "dot"))),
         annotations = list(
           list(x = mean_enr, y = 1.04, yref = "paper", xanchor = "right",
                showarrow = FALSE, text = "enrolled mean",
                font = list(size = 12, color = ACCENT)),
           list(x = mean_non, y = 1.04, yref = "paper", xanchor = "left",
                showarrow = FALSE, text = "not enrolled mean",
                font = list(size = 12, color = ACCENT2)))) %>%
  pl_m2(width = PL_W, height = 280, t = 46)
```

::: {.warn}
**The distributions barely touch.** The two groups are almost separate
populations: `r sprintf("%.1f%%", 100 * mean(enr > mean_non))` of enrolled
businesses paid more than the average non-enrolled business, which is
`r gfmt(sum(enr > mean_non))` businesses. Only
`r sprintf("%.1f%%", 100 * mean(non < mean_enr))` of the comparison group paid
less than the enrolled average.

The gap is real, it is large, and the bar chart showed you none of this.
:::

::: {.ask}
**Look back at the number you wrote down.** Did you expect
`r sprintf("%.1f%%", 100 * mean(enr > mean_non))`?

A mean tells you where a group sits on average. It cannot tell you how many of
them sit somewhere else.
:::

## Rule 2: ask what the bar is averaging over

```{=html}
<div class="facts">
  <div class="k">A mean</div>        <div class="v">a summary of a group &middot; not a description of anyone in it</div>
  <div class="k">Two groups can</div><div class="v">share a mean and share nothing else</div>
  <div class="k">A bar chart</div>   <div class="v">cannot show overlap, spread, or shape &middot; by construction</div>
  <div class="k">So ask</div>        <div class="v">"how much do the two groups overlap?" &middot; it is answerable</div>
</div>
```

::: {.ask}
This is the same instinct as Day 1's confidence interval and Day 2's baseline
balance table. **Every summary throws information away, and the useful question
is always which information.**

Here the discarded information happened to *support* the finding. It often does
not.
:::

# 3 · Check the denominator

## The change nobody can compute

::: {.panel}
The school-zone trial measured child collisions at 400 segments, before and
after. An analyst wants to report **the percentage change in child collisions at
the typical segment.**

A reasonable-sounding request. Here is why it cannot be answered.
:::

```{r}
sc %>%
  mutate(`Baseline child collisions` = base) %>%
  count(`Baseline child collisions`) %>%
  mutate(`Share of segments` = sprintf("%.1f%%", 100 * n / sum(n))) %>%
  select(-n) %>%
  kable(align = "rr") %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(1, bold = TRUE, background = "#F7E9E6")
```

::: {.warn}
**`r gfmt(zero_base)` of the 400 segments recorded no child collisions at
baseline.** For those segments a percentage change is undefined: anything divided
by zero is not a number, and no amount of care in the plotting will fix it.

Any figure reporting a percentage change has already made a decision about these
`r gfmt(zero_base)` segments, and it is unlikely to have told you what it was.
:::

## Two charts, two different "biggest wins"

:::::: {.columns}

::::: {.column width="50%"}
**The largest percentage reductions**

```{r}
big_pct %>%
  transmute(Segment = segment_id, Arm = arm,
            Before = base, After = follow,
            Change = sprintf("%+.0f%%", pct)) %>%
  kable(align = "llrrr") %>%
  kable_styling(full_width = FALSE, font_size = 13) %>%
  row_spec(0, bold = TRUE)
```
:::::

::::: {.column width="50%"}
**The largest absolute reductions**

```{r}
big_abs %>%
  transmute(Segment = segment_id, Arm = arm,
            Before = base, After = follow,
            Change = sprintf("%+.0f", absolute)) %>%
  kable(align = "llrrr") %>%
  kable_styling(full_width = FALSE, font_size = 13) %>%
  row_spec(0, bold = TRUE)
```
:::::

::::::

::: {.warn}
**These two lists share `r overlap` segment** — they are almost entirely
different places.

Every segment on the left reduced child collisions by 100%. Every one of them had
two, three or four to begin with. The right-hand list is where the
`r gfmt(-sum(big_abs$absolute))` actual collisions were removed.
:::

::: {.ask}
**Neither list is wrong.** They answer different questions:

*Where did the programme work best?* — the left.
*Where did it prevent the most harm?* — the right.

A report that shows only one of them has chosen your question for you.
:::

## Rule 3: check the denominator

```{=html}
<div class="facts">
  <div class="k">A percentage</div>   <div class="v">is a claim about a base &middot; find the base before you believe it</div>
  <div class="k">Small bases</div>    <div class="v">produce the largest percentages and the smallest changes</div>
  <div class="k">Zero bases</div>     <div class="v">produce no number at all &middot; but the figure still has a bar there</div>
  <div class="k">So ask</div>         <div class="v">"how many segments is that percentage built on?"</div>
</div>
```

::: {.ask}
The pooled change across all 400 segments is
`r sprintf("%.1f%%", pooled_pct)`. The average of the 400 individual percentage
changes is `r sprintf("%.1f%%", average_pct)`.

**The same data, two answers, differing by
`r sprintf("%.0f", abs(average_pct - pooled_pct))` percentage points** — because
one weights by how many collisions a segment had and the other gives every
segment equal say.
:::

# 4 · Levels or changes

## The figure that flatters the programme

::: {.panel}
The headline claim from Day 2 Session 1, drawn as a chart. After the programme,
the enrolled businesses paid **`r gfmt(gv(1, 1))` AED** and the comparison
businesses paid **`r gfmt(gv(0, 1))` AED**.

A difference of `r gfmt(gv(0, 1) - gv(1, 1))` AED. **The programme cut costs.**
:::

```{r fig.height=2.5, fig.width=6.6}
tibble(group = c("Enrolled", "Not enrolled"),
       cost  = c(gv(1, 1), gv(0, 1))) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside",
          textfont = list(size = 15, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost, after (AED)", range = c(0, 2700))) %>%
  pl_m2(width = PL_W, height = 230)
```

::: {.ask}
Everything on this slide is true and the figure is honest. It would appear in a
perfectly reputable report.

**What question does it answer, and what question did the evaluation ask?**
:::

## The same two groups, with where they started

```{r fig.height=3.0, fig.width=11.4}
tibble(
  Group = factor(rep(c("Enrolled", "Not enrolled"), each = 2),
                 levels = c("Enrolled", "Not enrolled")),
  Time  = factor(rep(c("Before", "After"), 2), levels = c("Before", "After")),
  Cost  = c(gv(1, 0), gv(1, 1), gv(0, 0), gv(0, 1))
) %>%
  plot_ly(x = ~Time, y = ~Cost, color = ~Group, colors = c(ACCENT, ACCENT2),
          type = "scatter", mode = "lines+markers",
          line = list(width = 3), marker = list(size = 11),
          hovertemplate = "%{fullData.name}, %{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2700))) %>%
  pl_m2(width = PL_W, height = 280)
```

::: {.warn}
**The comparison group started `r gfmt(gv(0, 0) - gv(1, 0))` AED worse off, and
its costs were rising anyway.**

So the level comparison on the previous slide is not a programme effect. It is
the programme effect **plus** a pre-existing gap **plus** a time trend that had
nothing to do with the programme. The honest estimate is
`r gfmt(did_est)` AED, not `r gfmt(gv(0, 1) - gv(1, 1))`.
:::

::: {.ask}
**This is the same trap as the Day 2 before-and-after figure, in a new costume.**
On Day 2 you saw it as a table. Here it is a bar chart, and it is much more
convincing.

The figure is not misleading because the analyst was careless. It is misleading
because **the level comparison is the easiest one to draw.**
:::

## Rule 4: compare changes, not levels

```{=html}
<div class="facts">
  <div class="k">Levels carry</div>   <div class="v">whatever the two groups already differed by &middot; and whatever moved on its own</div>
  <div class="k">Changes remove</div> <div class="v">the starting gap &middot; if the groups were on parallel paths</div>
  <div class="k">Both are easy</div>  <div class="v">to compute &middot; only one answers "did the programme work?"</div>
  <div class="k">So ask</div>         <div class="v">"what did the comparison group look like before?"</div>
</div>
```

::: {.ask}
A four-number test, and you can run it on any before-and-after figure:

**Do I have the before for both groups? If not, the chart cannot tell me the
programme's effect, whatever it is titled.**
:::

# 5 · The assumption the picture hides

## Two ways to add up the same saving

::: {.panel}
From this morning: GreenWaste cost **`r gfmt(COST_PER_BUSINESS)` AED** per
business and saved **`r gfmt(BENEFIT_PER_YEAR)` AED** a year. Suppose the saving
lasts five years, starting one year after rollout.

There are two ways to draw the total. **One of them is a model; the other is an
assumption.**
:::

```{r fig.height=3.0, fig.width=11.4}
cb_years %>%
  plot_ly(x = ~year, y = ~cum_nominal, name = "Added up as they come",
          type = "scatter", mode = "lines+markers",
          line = list(color = ACCENT2, width = 3, dash = "dot"),
          marker = list(size = 9),
          hovertemplate = "Year %{x}: %{y:,.0f} AED<extra></extra>") %>%
  add_trace(y = ~cum_discounted, name = "Discounted to today's money",
            type = "scatter", mode = "lines+markers",
            line = list(color = ACCENT, width = 3), marker = list(size = 9),
            hovertemplate = "Year %{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("Years after rollout"),
         yaxis = ax("Cumulative saving per business (AED)"),
         shapes = list(list(type = "line", x0 = 0.5, x1 = 8.5, y0 = COST_PER_BUSINESS,
                            y1 = COST_PER_BUSINESS, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 8.4, y = COST_PER_BUSINESS, xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "what the programme cost", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_W, height = 280, t = 20)
```

::: {.warn}
**Both lines are drawn from the same measured saving.** The only difference is
whether a saving received in three years' time is counted as worth the same as
one received today.

The upper line reaches `r gfmt(cb_nom_total)` AED after five years. The lower
reaches `r gfmt(cb_dis_total)`. The gap between them,
`r gfmt(cb_nom_total - cb_dis_total)` AED, is **not a finding**. It is the
discount rate, drawn as if it were money.
:::

## Rule 5: discount in front of the reader

:::::: {.columns}

::::: {.column width="52%"}
```{=html}
<div class="facts">
  <div class="k">Five years, added up</div>  <div class="v">4,082 AED per business</div>
  <div class="k">Five years, discounted</div><div class="v">3,366 AED per business</div>
  <div class="k">The difference</div>        <div class="v">716 AED &middot; the discount rate, not the programme</div>
  <div class="k">The ratio</div>             <div class="v">3,366 / 1,800 = 1.87</div>
</div>
```
:::::

::::: {.column width="48%"}
::: {.ask}
The undiscounted line is the one that gets drawn, because it is easier to explain
and it always looks better. Nobody has to defend it in the same way they would
have to defend a 5% rate.

**That asymmetry is the point.** A choice that favours the programme is treated
as arithmetic; a choice that does not is treated as an assumption.
:::
:::::

::::::

::: {.warn}
The cost line is crossed in year
`r sprintf("%.1f", 3.21)` on the undiscounted view and year
`r sprintf("%.1f", 3.53)` on the discounted one. **Half a year of payback is
decided by the discount rate**, and neither figure will mention it.
:::

# 6 · The fifth figure

## What is this pie not telling you?

::: {.panel}
The school-zone evaluation reports a breakdown of collisions by severity. Here it
is, as a pie chart, because that is how it arrived.
:::

```{r fig.height=2.9, fig.width=5.2}
tibble(severity = factor(c("Minor", "Injury", "Child"),
                         levels = c("Minor", "Injury", "Child")),
       n = as.numeric(sev)) %>%
  plot_ly(labels = ~severity, values = ~n, type = "pie",
          marker = list(colors = c(ACCENT2, ACCENT, ALERT),
                        line = list(color = "white", width = 2)),
          textinfo = "label+percent", textfont = list(size = 14, color = "white"),
          hovertemplate = "%{label}: %{value:,}<extra></extra>") %>%
  pl_m2(width = PL_WC, height = 260, b = 20)
```

::: {.ask}
**In groups, three minutes.** This is a real breakdown of a real dataset. Write
down two things this chart tells you, and two things it does not.
:::

## What the pie cannot show

:::::: {.columns}

::::: {.column width="52%"}
```{=html}
<div class="facts">
  <div class="k">It shows</div>     <div class="v">the mix of severity &middot; child collisions are 22% of the minor total</div>
  <div class="k">It hides</div>     <div class="v">that the categories overlap &middot; a child collision is also an injury collision</div>
  <div class="k">It hides</div>     <div class="v">any change over time &middot; both rounds pooled into one number</div>
  <div class="k">It hides</div>     <div class="v">how rare child collisions are &middot; a small slice invites a small response</div>
</div>
```
:::::

::::: {.column width="48%"}
::: {.warn}
**The most serious problem is the first one.** A pie chart asserts that its
slices partition a whole. These do not: every child collision is also counted as
an injury collision, so the slices overlap.

The chart cannot be read correctly by a careful reader, because it has no correct
reading.
:::
:::::

::::::

::: {.ask}
**This is a different kind of failure from the other four figures.** The truncated
axis and the level comparison were honest charts inviting a wrong inference. This
one is a chart with no right answer.

Before you ask what a figure shows, ask **whether it can show anything at all.**
:::

# 7 · Put AI to work

## AI Snapshot: the chart that cannot be wrong

::: {.panel}
Groups: give the model the two GreenWaste baseline figures — the zero-axis one
and the truncated one — and this prompt.

**"Which of these two charts better represents the difference between the two
groups, and why?"**
:::

::: {.ai-output}
**AI response:**

Both charts present the same underlying data accurately. [The truncated-axis
chart is often preferable for a policy audience because it makes the difference
visible]{.flag}, and [a y-axis that starts at zero is not a requirement in
professional data visualisation]{.flag}.

[On balance the truncated version communicates the finding more effectively, and
a reader who wants the absolute values can consult the data labels]{.flag}.
:::

::: {.ask}
**Four minutes.** Judge this answer against what you have just learned.

Is it wrong? Is it right? Is it answering the question that was asked?
:::

## Debrief: true statements, wrong question

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Nothing in it is false.** Truncated axes are common in professional practice.
Data labels do let a reader recover the values. A zero baseline is not a
requirement.

Each sentence would survive a fact-check.
:::

::: {.warn}
**But it never asks the question that matters.** The prompt asked *which
represents better*. It did not ask *what does the reader need to know*, and it
never mentions the two things that decide the answer:

* **who is reading it** — a minister and an analyst need different figures;
* **what the sentence next to it will say** — a truncated axis is fine beside "a
  difference exists" and not fine beside "a large difference".
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**It also did the thing the session is about.** It compared two pictures and never
compared either to the number.

The gap is `r gfmt(gap)` AED, `r sprintf("%.2f", sd_gap)` standard deviations. That
figure settles the "is it large?" question, and it is available without any chart
at all.
:::

::: {.warn}
**The failure shape: competence without a standpoint.** The model gave a balanced,
well-informed answer to a question that could not be answered without knowing the
audience. It did not notice that the question was underspecified, and it did not
ask.
:::
:::::

::::::

::: {.ask}
**The lesson for figures, and it is the same lesson as this morning:** the chart
is not the finding. Ask what the number is, then ask what the picture does to
your reading of it.

**A better prompt** would name the reader and the sentence: *"This chart will sit
beside the claim 'the programme reduced costs substantially'. Is the axis
appropriate for that claim, for a non-technical reader?"*
:::

# 8 · The five rules

## What to do with any figure you are handed

```{r}
fig_rules %>%
  select(` ` = n, Rule, Why) %>%
  kable(align = "ll") %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  column_spec(2, width = "13em") %>%
  column_spec(3, width = "26em")
```

::: {.ask}
Five checks, none of them requiring software, all of them applicable in the
thirty seconds you actually get with a figure before forming an impression.

**You have run all five today**, on five real figures, from data you have already
worked with this week.
:::

## The one you will actually use

::: {.panel}
If you remember one thing, make it this. It is the fastest of the five and it
catches the most.

**Read the y-axis, then read the sentence underneath the figure, then decide
whether they are describing the same thing.**
:::

```{=html}
<div class="facts">
  <div class="k">The axis says</div>      <div class="v">how big the difference is on the page</div>
  <div class="k">The sentence says</div>  <div class="v">how big the difference is in the world</div>
  <div class="k">They are different</div> <div class="v">questions &middot; and the gap between them is where overclaiming lives</div>
  <div class="k">So ask</div>             <div class="v">"large compared to what?"</div>
</div>
```

::: {.warn}
Every figure this session has a defensible version and a misleading version, made
from the same numbers. **The difference was never the data.** It was a choice
somebody made, usually without noticing they had made it.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">This morning</div> <div class="v">what a result is worth &middot; cost, and the assumption that decides it</div>
  <div class="k">Now</div>          <div class="v">the form it arrives in &middot; and what the form does to your reading</div>
  <div class="k">Next session</div> <div class="v">the analyst is in the room &middot; and has to answer for all of it</div>
  <div class="k">Bring with you</div><div class="v">the five rules, and the questions you did not get to ask</div>
</div>
```

::: {.ask}
The next session is a clinic. The 3ie trainer plays the evaluator of a study you
have not seen. You have the checklist you have been building since Day 1, and now
the figure-reading questions as well.

**Start compiling your questions now.** The group that arrives with the sharpest
three questions will get the most out of it.
:::
