---
title: "What Do the Numbers Say? Reading an Evaluation Output"
subtitle: "Module 2 · Day 1, Session 2: Illustrative Case Data"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)   # lm_robust, clustered standard errors

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / offered
ACCENT2 <- "#A9561F"   # series 2: comparison / not offered
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

# ---- Chart sizing ----------------------------------------------------------
# Every chart carries an explicit pixel size. Relying on plotly's autosize
# inside revealjs produces half-drawn charts: a widget initialised while its
# slide is hidden measures a narrow container and never corrects itself.
# These must stay in step with the chunk's fig.width / fig.height, which size
# the htmlwidgets wrapper (at dpi 96, one inch is 96 px).

PL_W  <- 1100   # full-width chart
PL_WC <- 520    # chart sitting in a half-width column
PL_WT <- 340    # chart sitting in a third-width column

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width  = width,
      height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))

# ---- Session 2: the traffic camera programme -------------------------------
# One row per road segment per round. 1,200 segments in 24 sectors, two rounds
# (2019 before, 2021 after). Twelve sectors got cameras in 2020.
#
# NOTE ON THE DATA, AND READ THIS BEFORE EDITING ANY SLIDE.
#
# evaluation_data_TrafficCameras.csv is FICTIONAL data generated by
# make_traffic_camera_data.R. It is not an Abu Dhabi Police evaluation and it is
# not any real programme. It was originally drafted as a stand-in for Police
# findings that do not exist in this workspace, and it named the force, which
# attributed fabricated numbers to a real organisation on a public site. The
# case now names no real body.
#
# Every slide that shows a number from it carries a visible disclaimer, per
# CLAUDE.md's placeholder convention.
pol <- read.csv("./evaluation_data_TrafficCameras.csv")

# The bar the authority set before committing to a national roll-out.
POL_RULE <- 2.0     # injury collisions avoided, per segment per round

pol_before <- pol %>% filter(round == 0)
pol_after  <- pol %>% filter(round == 1)

# The headline the authority would put in a press release: treated sectors only,
# before and after, no comparison.
pol_t_before <- mean(pol_before$injury_collisions[pol_before$treated == 1])
pol_t_after  <- mean(pol_after$injury_collisions[pol_after$treated == 1])
pol_headline_fall <- 100 * (pol_t_before - pol_t_after) / pol_t_before

# What happened in the sectors that did not get cameras over the same period.
pol_c_before <- mean(pol_before$injury_collisions[pol_before$treated == 0])
pol_c_after  <- mean(pol_after$injury_collisions[pol_after$treated == 0])
pol_comp_fall <- 100 * (pol_c_before - pol_c_after) / pol_c_before

# The honest estimate: the model the deck projects, with errors clustered by
# road segment because the two rounds of one segment are not independent.
pol_fit <- lm_robust(injury_collisions ~ treated * round,
                     data = pol, clusters = segment_id)

pol_did    <- coef(pol_fit)[["treated:round"]]
pol_did_ci <- confint(pol_fit)["treated:round", ]
pol_did_p  <- pol_fit$p.value[["treated:round"]]
pol_avoided_hi <- abs(pol_did_ci[1])   # most the interval will allow

# The table as a participant would receive it. Built here rather than on the
# slide so that the "reading the output" slide and the "annotate this output"
# exercise show the identical table.
pol_tbl <- pol_fit %>%
  tidy() %>%
  mutate(` ` = c("(Intercept)", "Programme sectors", "Round 2 (2021)",
                 "Programme x Round 2"),
         CI = sprintf("[%.2f, %.2f]", conf.low,
                      estimate + (estimate - conf.low))) %>%
  select(` `, estimate, std.error, p.value, CI)

# The two group means behind the four numbers on the comparison slide.
pol_means <- pol %>%
  mutate(Group = ifelse(treated == 1, "Programme sectors",
                        "Comparison sectors"),
         Time  = tlev(ifelse(round == 0, "Before", "After"))) %>%
  group_by(Group, Time) %>%
  summarise(m = mean(injury_collisions), .groups = "drop")

# Segment counts, for the "who is in the evaluation" slide.
pol_seg  <- n_distinct(pol$segment_id)
pol_sect <- n_distinct(pol$sector_id)
pol_t_seg <- n_distinct(pol$segment_id[pol$treated == 1])
```

# 1 · The case

## The traffic camera programme

:::::: {.columns}

::::: {.column width="52%"}
In 2020 a roads authority began installing automated speed cameras. Not
everywhere at once: twelve sectors first, the rest two years later.

It set a bar before committing further: the programme had to avoid
**`r sprintf("%.1f", POL_RULE)` injury collisions per road segment per round**.
Below that, the money goes elsewhere.

::: {.warn .note-ctr}
**Illustrative case, with invented figures.** This is not a real programme and
the numbers are generated for this training. Every number in this session comes
from this fictional dataset.
:::
:::::

::::: {.column width="48%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">Automated speed enforcement</div>
  <div class="k">Rolled out</div>  <div class="v">12 sectors in 2020, 12 in 2022</div>
  <div class="k">Outcome</div>     <div class="v">Injury collisions</div>
  <div class="k">Decision rule</div> <div class="v">2.0 fewer per segment per round</div>
</div>
```
:::::

::::::

## Who is in the evaluation

```{r results='asis'}
cat(sprintf('
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">Road segments</span></div>
  <div><span class="n">%d</span><span class="k">Sectors</span></div>
  <div><span class="n">%s</span><span class="k">Segments with cameras</span></div>
  <div><span class="n">2</span><span class="k">Rounds: 2019 and 2021</span></div>
</div>', format(pol_seg, big.mark = ","), pol_sect,
   format(pol_t_seg, big.mark = ",")))
```

::: {.panel}
Session 1 taught three questions. The first is **compared to what?** This
session is that question pointed at one programme.

Before the numbers: two things about the design. The programme went first to
the **worst roads**, so the sectors with cameras are not the same as the sectors
without. And traffic **grew** between the two rounds.
:::

::: {.ask}
Keep both in mind. One of them is why the headline number will turn out to
overstate what the programme did.
:::

## The decision on the table

::: {.panel}
The authority is deciding whether to extend cameras elsewhere. That is a
spending decision, and it turns on one figure: how many injury collisions the
cameras avoided, **per segment, per round**.

The rule is `r sprintf("%.1f", POL_RULE)`. A result below it does not justify the
money, however impressive it looks as a percentage.
:::

::: {.ask}
You will see three numbers that describe the same programme:

**-`r sprintf("%.0f", pol_headline_fall)`%** &nbsp;·&nbsp; **-`r format(abs(pol_did), nsmall = 2)`** &nbsp;·&nbsp; **`r sprintf("%.1f", POL_RULE)`**

Two of them are wrong answers to the question being asked. Work out which is the
real one, and whether it clears the rule.
:::

# 2 · What the numbers say

## The headline

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Injury collisions, sectors with cameras</div>
</div>
<div class="stat-row">
  <div><span class="n">%.1f</span><span class="k">Per segment, 2019</span></div>
  <div><span class="n">%.1f</span><span class="k">Per segment, 2021</span></div>
  <div><span class="n">-%d%%</span><span class="k">The headline</span></div>
</div>', pol_headline_fall, pol_t_before, pol_t_after,
   round(pol_headline_fall)))
```

::: {.warn}
This is the number that reaches the minister. It is **accurate**: in the sectors
where cameras went, injury collisions fell
`r sprintf("%.0f", pol_headline_fall)`% between 2019 and 2021.

It is also a before-and-after comparison. It compares those roads to their own
past, which is exactly the comparison Session 1 spent an hour on.
:::

## `[MENTI PLACEHOLDER]` Would you fund the extension?

::: {.ask}
On the number you have just seen: **would you commit the money?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Not on this alone
:::

::: {.panel}
Commit before we look further. This is the same poll shape as Session 1, and it
worked there for the same reason: the room has to hold a position before it can
feel it move.
:::

## What else happened in those two years

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.7, fig.width=5.6}
pol_means %>%
  plot_ly(x = ~Time, y = ~m, color = ~Group, type = "bar",
          colors = c("Programme sectors" = ACCENT,
                     "Comparison sectors" = ACCENT2),
          text = ~sprintf("%.1f", m), textposition = "outside",
          textfont = list(size = 14, color = INK),
          hovertemplate = "%{fullData.name}: %{y:.1f}<extra></extra>") %>%
  layout(barmode = "group",
         xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Injury collisions per segment", range = c(0, 13))) %>%
  pl_m2(width = PL_WC, height = 270, title = "Same two years, both groups")
```
:::::

::::: {.column width="50%"}
::: {.panel}
Cameras went to the twelve worst sectors. The twelve without cameras also saw
collisions fall, by
`r sprintf("%.0f", pol_comp_fall)`%, and nothing was installed there.

Some of that is safer vehicles and national campaigns. Whatever it is, it
happened to both groups, so it is not the programme.
:::

::: {.warn}
Read the two bars together. The gap between the two falls is the part that could
belong to the cameras.
:::
:::::

::::::

## The honest estimate

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Injury collisions avoided per segment, per round</div>
</div>
<div class="stat-row">
  <div><span class="n">%.1f%%</span><span class="k">Programme sectors fell</span></div>
  <div><span class="n">%.1f%%</span><span class="k">Comparison sectors fell</span></div>
  <div><span class="n">%.2f</span><span class="k">The difference</span></div>
</div>', pol_did, pol_headline_fall, pol_comp_fall, pol_did))
```

::: {.panel}
Subtract one fall from the other and
`r sprintf("%.2f", abs(pol_did))` injury collisions per segment per round is
left. That is the estimate. Note it is in **collisions**, not per cent:
percentages cannot be checked against a rule that is written in collisions.
:::

## Does it clear the rule?

:::::: {.columns}

::::: {.column width="48%"}
```{=html}
<div class="facts">
  <div class="k">Decision rule</div>      <div class="v">2.00 avoided</div>
  <div class="k">The estimate</div>       <div class="v">1.13 avoided</div>
  <div class="k">95% interval</div>       <div class="v">0.51 to 1.76</div>
  <div class="k">Best case in interval</div> <div class="v">1.76 avoided</div>
</div>
```
:::::

::::: {.column width="52%"}
::: {.panel .fragment}
The estimate is statistically clear: the interval sits well away from zero, so
the programme did something. **That is not the question being asked.**

The question is whether it cleared 2.00. It did not. Even the interval's most
generous end, `r sprintf("%.2f", pol_avoided_hi)`, falls short.
:::

::: {.warn}
**Statistically significant does not mean big enough.** Session 1 said this about
a school programme. Here is the same lesson with a real spending decision
attached to it.
:::
:::::

::::::

## Two reasons the headline overstates

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**1. The comparison moved too.** Collisions fell
`r sprintf("%.0f", pol_comp_fall)`% in sectors with no cameras. A before-and-after
reading gives the programme credit for all of it.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**2. Traffic grew.** Vehicles travelled 4% further in 2021. Collisions per
million vehicle-km is the fairer measure of **risk**, and it says something
slightly different from the raw count.
:::
:::::

::::::

::: {.ask .fragment}
Session 1's Al Falah case had exactly these two problems. Distance driven fell,
so fatalities fell. This time traffic **rose**, so part of the collision fall is
simply less exposure. Same trap, opposite direction.
:::

# 3 · Reading the output

## The table you would actually be handed

:::::: {.columns}

::::: {.column width="52%"}
```{r}
pol_tbl %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5")
```
:::::

::::: {.column width="48%"}
::: {.panel}
**Row 4 is the impact.** The interaction of programme and round: the extra change
in the sectors that got cameras, over and above the change everywhere else.

**Row 2** is the baseline difference. Cameras went to the worst roads, so they
started higher.
:::

::: {.panel}
**Row 3** is the national trend: what happened in the comparison sectors anyway.
**Std. error** is the wobble.

Most reports lead with their preferred number. **Find the row that is actually
the impact.**
:::
:::::

::::::

## The same estimate, three ways of saying it

::: {.panel}
A decision-maker will be offered three framings of this one number. They are not
equally honest.
:::

```{=html}
<div class="facts">
  <div class="k">Per cent</div>
  <div class="v">"Collisions fell 22% where cameras were installed" &middot; ignores the comparison</div>
  <div class="k">Point estimate</div>
  <div class="v">"Cameras avoided 1.13 collisions per segment per round" &middot; accurate, but a single number</div>
  <div class="k">With its interval</div>
  <div class="v">"Between 0.51 and 1.76, so somewhere short of the 2.0 required" &middot; this is the honest one</div>
</div>
```

::: {.ask}
Which of the three would you put in a briefing? Session 1's Word 5 answer was
that **the interval decides, not the point estimate.** Does it hold here?
:::

## An interval that does not reach the rule

```{r fig.height=2.9, fig.width=11.5}
tibble(
  what = factor(c("The estimate", "95% interval", "Decision rule"),
                levels = c("Decision rule", "95% interval", "The estimate")),
  lo   = c(pol_did, pol_did_ci[1], POL_RULE),
  hi   = c(pol_did, pol_did_ci[2], POL_RULE),
  pt   = c(pol_did, pol_did, POL_RULE),
  col  = c(ACCENT, ACCENT, ALERT)
) %>%
  plot_ly() %>%
  add_segments(x = ~abs(lo), xend = ~abs(hi), y = ~what, yend = ~what,
               line = list(color = ~col, width = 5), showlegend = FALSE,
               hovertemplate = "avoids %{x:.2f}<extra></extra>") %>%
  add_markers(x = ~abs(pt), y = ~what, marker = list(color = ~col, size = 11),
              showlegend = FALSE,
              hovertemplate = "avoids %{x:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Injury collisions avoided per segment, per round",
                    range = c(0, 2.4)),
         yaxis = ax("", zeroline = FALSE)) %>%
  pl_m2(height = 260)
```

::: {.warn}
The whole interval sits to the left of the red line. This is not "we cannot
tell": it is a clear result that falls short. **A precise answer below the bar
is still a no.**
:::

## What the table does not tell you

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Nothing about roads without cameras.** The comparison sectors are a
counterfactual, not a plan. The estimate says what the cameras did where they
went, not what they would do elsewhere.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**Nothing about compliance.** Cameras work by changing behaviour. If drivers
learned where they are and sped up in between, the effect fades after 2021.
:::
:::::

::::::

::: {.ask}
Two habits from Session 1, applied to this table:

Ask what is **not** in it. Then ask the evaluator what would have to be true for
the result to be wrong. Here, the answer is that the two groups were already
diverging before 2020. Nobody has shown you a pre-trend.
:::

# 4 · Put AI to work

## AI Snapshot: three sentences for a decision-maker

::: {.panel}
Paste the table on the previous slide into an AI tool with this prompt:

**"Explain this result to a decision-maker in three sentences. They have a rule:
the programme must avoid 2.0 collisions per segment per round to be funded."**
:::

::: {.ai-output}
**AI response:**

"The programme was associated with a **22% reduction** in injury collisions in
sectors where cameras were installed. This represents a **clear improvement in
road safety**, with the effect likely to be **even larger** once the remaining
sectors are covered. [The evidence supports extending the programme
nationally.]{.flag}"
:::

::: {.ask}
Mark every claim you can check against the table. How many can you falsify?
:::

## Debrief

::: {.warn}
**Three things went wrong, and only one is a fabrication.**

**The 22% is real but not the impact.** It is the before-and-after figure. The
table's fourth row says `r sprintf("%.2f", abs(pol_did))`. The AI reached for the
most impressive true number in the room.

**"Likely to be even larger" has no basis.** Nothing in the table says the
increasing-returns case. This is the invented-precision failure: a confident
direction supplied from nowhere.
:::

::: {.warn}
**The flagged sentence answers a different question.** It recommends funding.
The brief asked whether the programme cleared **`r sprintf("%.1f", POL_RULE)`**.
The AI never mentioned the rule it was given, so it never compared the estimate
to it.
:::

::: {.panel}
**The habit to take away.** The AI was not lying. It was fluent, and it picked
the flattering number, and it answered "is this good?" instead of "does it clear
the bar?" Ask it to **state the decision rule and compare against it** before it
recommends anything.
:::

# 5 · Annotate this output

## Your turn

:::::: {.columns}

::::: {.column width="54%"}
```{r}
pol_tbl %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5")
```

::: {.panel}
Outcome: injury collisions per road segment per round. Standard errors
clustered by road segment.
:::
:::::

::::: {.column width="46%"}
::: {.ask}
On the printed copy in front of you:

**1.** Circle the number that is the treatment effect.

**2.** Draw a box around its confidence interval.

**3.** Underline the baseline difference, and say in one line why it is not the
impact.

**4.** Write down one thing you cannot tell from this table, and the question
you would put to the evaluator.
:::

::: {.panel}
Four minutes. Question 4 matters most, and Session 1 prepared you for it.
:::
:::::

::::::

# 6 · Take this back to your desk

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Which group, which period, and did it move too?</div>
  <div class="k">2</div> <div class="v">Which row is the impact, and is it in units the decision is written in?</div>
  <div class="k">3</div> <div class="v">Does the interval reach the decision rule, not just clear zero?</div>
</div>
```

::: {.panel}
This result would survive question 2. It fails question 3, and that is the whole
decision.
:::

## Where this goes next

::: {.panel}
Today you read a **before-and-after with a comparison group**. That is the
simplest honest design, and it rests on one assumption: that the two groups were
heading the same way before the programme started.

Day 2 does three things this session did not:

**Difference-in-differences** on this same data, testing that assumption rather
than assuming it. **Regression discontinuity** on the speed cut-off that decided
which roads got cameras. **Matching**, because the twelve worst sectors were not
chosen at random.
:::

::: {.ask}
Same programme, same 1,200 road segments, four ways of asking what it did. The
answers will not be identical, and the differences are the lesson.
:::
