---
title: "Plan Your Own Evaluation"
subtitle: "Module 2 · Day 4, Session 3: Building on Module 1"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"

# ---- The template ----------------------------------------------------------
# Shared with the printable Evaluation Design Update template.
source("qa_design_update.R")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · The last session

## Four days. One design.

::: {.panel}
On Monday you learned what a comparison is. Tuesday, four ways of building one.
Wednesday, what a result costs and what a figure hides. Thursday, how to tell
whether evidence is credible and how to translate it.

Today you point all of it at your own evaluation.
:::

::: {.ask .fragment}
**How far along is your design?**
Not started / rough outline / drafted / ready to pitch.
:::

## What today is not

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**It is not new material.** Every technique today has already been taught. This
session assembles them.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**It is not a written exam.** You are not being graded on completeness. You are
being asked to make choices and to defend them in five minutes.
:::
:::::

::::::

::: {.ask .fragment}
The template has three parts: where your design stands, the four additions, and
your pitch. Parts 1 and 2 are working time. Part 3 is performance.
:::

# 1 · Where your design stands

## The eight conditions, from Module 1

```{r}
#| results: asis
tab <- data.frame(
  `#` = seq_len(nrow(design_conditions)),
  Condition = design_conditions$condition,
  check.names = FALSE
)
kable(tab, col.names = c("#", "Condition"),
      caption = "The eight conditions for a credible impact evaluation") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Nothing here has changed since Module 1, and nothing this week contradicted it.
Part 1 of the template asks you to write your design against these, so the new
work has something to attach to.
:::

## The one condition that decides the rest

::: {.ask}
Of the eight, which one, if you get it wrong, makes the other seven irrelevant?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**The counterfactual.**

Every method this week has been a different way of operationalising it:
randomisation, a cut-off, an instrument, a matched comparison, a parallel trend.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
And every one of them has a specific way of being false. If you cannot state
yours in a sentence, you have not chosen a design — you have chosen a method.
:::
:::::

::::::

::: {.ask .fragment}
Write your counterfactual sentence now, before we go on. One sentence. You will
read it out in the pitch.
:::

## Working time

::: {.panel}
**Twenty minutes.** Part 1 of the template: write your design as it stands,
against the eight conditions.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
If you have the Module 1 document, work from it. If not, the recap on the previous
slide is enough — write what you would do, not what you did.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Do not polish. This is the foundation for Part 2, and Part 2 is where the marks
are in the pitch.
:::
:::::

::::::

# 2 · The four additions

## What you are adding

```{r}
#| results: asis
tab <- data.frame(
  Addition = design_additions$addition,
  `From` = design_additions$from,
  check.names = FALSE
)
kable(tab, col.names = c("Addition", "From"),
      caption = "The four additions, and where each came from") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Each of these is something this week taught you to demand of someone else's
evaluation. Now you have to supply it for your own.
:::

::: {.ask .fragment}
Which of the four is hardest to do **before** you have any data? Why?
:::

## Addition 1: results presentation plan

::: {.ask}
What will you show, to whom, and in what order? Name the one number you will lead
with — and say what the reader must **not** be able to conclude from it.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Yesterday's lesson, applied**

The fictitious report led with its largest estimate and called a different one its
preferred specification. A reader who stopped after the summary concluded the
opposite of the truth.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
So the question is not "what is my headline number". It is "what will a reader
who reads only the headline come away believing — and is that true?"
:::
:::::

::::::

## Addition 2: cost-benefit framework

::: {.ask}
Four things to state: the **threshold**, the **benefit per unit per year**, the
**horizon**, and the **discount rate**.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

Of those four, the discount rate is the one everybody argues about and the one
that usually matters least. The horizon decides almost everything and is often
never stated.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Write your horizon down. If your benefit stream runs for five years and someone
assumes two, your ratio moves from 1.87 to 0.80 — the same programme, the same
evidence, the opposite decision.
:::
:::::

::::::

## Addition 3: data visualisation sketch

::: {.ask}
Sketch the figure that carries your headline result. Then sketch the **same data
presented a second way**, and say what the second version hides.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

Every pair of figures showed identical numbers and different impressions. The
truncated axis, the bar of means, the denominator, the levels-versus-changes, the
discounting.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second sketch is the one that matters. If you cannot say what your own
preferred figure hides, someone else will find it in the questions.
:::
:::::

::::::

## Addition 4: evidence translation plan

::: {.ask}
Write the **headline sentence** of your brief, with the number in it. Then name
the one thing you will state plainly that your evaluation did **not** establish.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**This morning's lesson**

A brief that reports the estimate and omits the threshold leaves the reader to
assume the target was met. Omission is the failure mode, not error.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second half of this addition is the harder one, and it is the one that will
distinguish your pitch. Naming your own limit is more persuasive, not less.
:::
:::::

::::::

## Working time

::: {.panel}
**Thirty minutes.** Part 2 of the template: all four additions.

Work as a team. Split the four if that is faster, but every member should be able
to speak to all four by the end.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Use the template's prompts. Each one is phrased as the question a commissioner
would ask, because that is who you are pitching to.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
You will not finish all four to the same standard. That is expected. Decide as a
team which two you will lead on.
:::
:::::

::::::

# 3 · The pitch

## Five minutes, five parts

```{r}
#| results: asis
tab <- data.frame(
  Part = pitch_structure$part,
  Time = pitch_structure$minutes,
  `What it says` = pitch_structure$says,
  check.names = FALSE
)
kable(tab, col.names = c("Part", "Time", "What it says"),
      caption = "The pitch structure") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 15)
```

::: {.warn}
Note the order. **The honest limit comes before the ask**, not after it. That is
deliberate and it is not how pitches are usually built.
:::

## Why the limit comes first

::: {.ask}
If you tell the room what your design cannot do, why would that make them more
likely to give you what you want?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
Because it tells them **which parts you have already thought about**. A pitch that
claims no limits is either naive or hiding something, and the room cannot tell
which.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
It also changes what you are asking for. If you know your design cannot establish
X, you can ask for what would — which is a more specific request than "fund the
evaluation".
:::
:::::

::::::

## Feedback from the room

::: {.panel}
Three minutes per team. The room gives feedback in this order, and the order
matters:

1. **What is the design?** Say it back in one sentence. If you cannot, that is the
   feedback.
2. **What would make it fail?** The one assumption that would have to break.
3. **What is missing?** One thing you would have wanted to hear.
:::

::: {.ask .fragment}
Feedback is not a verdict. The first question is a diagnostic: if the room cannot
say your design back to you, the pitch did not land, and that is worth knowing
before a real commissioner.
:::

## Peer feedback on the day

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Write your feedback down. Spoken feedback evaporates and the team cannot use it.</div>\n')
cat('<div class="panel"><b>2.</b> One thing that worked, one thing that did not. Not three of each.</div>\n')
cat('<div class="panel"><b>3.</b> Direct the feedback at the design, never at the team.</div>\n')
cat('<div class="panel"><b>4.</b> If you cannot say the design back in one sentence, say so. It is the most useful thing you can give them.</div>\n')
```

::: {.ask .fragment}
Which of those four is hardest to do well? Which do people usually skip?
:::

# 4 · The week in one slide

## What you can now do

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Say what a number is compared to, and why that comparison is the argument.</div>\n')
cat('<div class="panel"><b>2.</b> Read DiD, RDD, IV and matching results, and name the assumption each rests on.</div>\n')
cat('<div class="panel"><b>3.</b> Judge whether a benefit-cost ratio clears a threshold, and what decides it.</div>\n')
cat('<div class="panel"><b>4.</b> See the same data drawn two ways, and say what each version hides.</div>\n')
cat('<div class="panel"><b>5.</b> Interrogate an evaluation with a checklist, and rate a report section by section.</div>\n')
cat('<div class="panel"><b>6.</b> Write a brief a decision-maker can act on without being misled.</div>\n')
```

::: {.ask .fragment}
Which of the six would you have said you could do on Monday morning? Which would
you now add?
:::

## The one question

::: {.panel}
If you take one thing from four days, take this. It has been the first question in
every session, under every method:

> **What is being compared to what, and why should those two things be
> comparable?**
:::

::: {.ask .fragment}
Every method this week was a different answer to it. Every method had a different
way of being wrong about it. And every evaluation you commission from now on will
stand or fall on it.
:::

## Thank you

::: {.panel}
Four days, twelve sessions, one question.

The slides stay available online, and every dataset and script is in the
repository if you want to reproduce any of the analyses on your own programme.
:::

::: {.ask}
What is the first evaluation decision you will make when you get back to your
desk?
:::
