---
title: "Interrogate the Analyst"
subtitle: "Module 2 · Day 3, Session 3: The Quality Assurance Clinic"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html.
#
# One example all session: GreenWaste. New in this version: section 4, "The
# analyst answers with the data", where the trainer (playing the analyst)
# answers the strongest questions by running the analysis live.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  render-df: kable
  resources:
    - evaluation_data_GreenWaste.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(knitr)
library(qrcode)

FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- The checklist itself -------------------------------------------------
# Defined once in qa_checklist.R and shared with the printable handout, so the
# slide and the handout cannot drift apart.
source("qa_checklist.R")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_d3s3.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0)); plot(qr_code(url, ecl = "H")); path
}
MENTI_QR <- menti_qr()
```

# 0 · Sign-off {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Three days of reading. Today you ask the questions.

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. The analyst's answers in
# section 4 use what is defined here.
gw <- read.csv("evaluation_data_GreenWaste.csv")
library(estimatr)
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Warm up estimatr now, while the opening slides are talked through: its first
# call in the browser takes about 10 seconds, every later one is instant.
invisible(estimatr::lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1)),
                              clusters = c(1, 1, 2, 2)))
clean_par <- function() par(mar = c(4.2, 12, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.2)

offered  <- subset(gw, treatment_neighborhood == 1)          # DiD sample, both rounds
rdd_data <- subset(offered, round == 1)
rdd_data$distance <- rdd_data$efficiency_index - 58
rct_data <- subset(gw, round == 1 & eligible == 1)
covs <- c("age_manager", "age_deputy", "educ_manager", "educ_deputy", "female_manager",
          "foreign_owned", "staff_size", "advanced_filtration", "water_treatment_system",
          "business_area", "recycling_center_distance")

row_of <- function(method, fit, term, n) {
  ci <- confint(fit)[term, ]
  data.frame(method, estimate = coef(fit)[[term]], lower = ci[1], upper = ci[2],
             businesses = n, row.names = NULL)
}

# The three estimates we can rerun here, each with its 95% interval.
estimates <- function() {
  # se_type "stata" (CR1): the default CR2 with 4,959 clusters is too slow in
  # the browser, and the interval differs from CR2 by about 1 AED.
  did <- lm_robust(waste_management_costs ~ enrolled * round, data = offered,
                   clusters = business_identifier, se_type = "stata")
  rdd <- lm_robust(reformulate(c("eligible", "distance", covs), "waste_management_costs"),
                   data = rdd_data)
  rct <- lm_robust(waste_management_costs ~ treatment_neighborhood, data = rct_data,
                   clusters = neighborhood_identifier)
  rbind(row_of("Difference-in-differences", did, "enrolled:round", nrow(offered) / 2),
        row_of("Regression discontinuity", rdd, "eligible", nrow(rdd_data)),
        row_of("Randomised offering", rct, "treatment_neighborhood", nrow(rct_data)))
}

# Each estimate as AED avoided, with its interval, against the 1,000 AED bar.
interval_plot <- function(tab) {
  clean_par()
  lo <- pmin(-tab$lower, -tab$upper); hi <- pmax(-tab$lower, -tab$upper)
  y  <- rev(seq_len(nrow(tab)))
  plot(NA, xlim = c(400, 1300), ylim = c(0.5, nrow(tab) + 0.5), yaxt = "n", bty = "n",
       xlab = "AED avoided per business per year", ylab = "")
  axis(2, at = y, labels = tab$method, tick = FALSE)
  abline(v = 1000, col = ALERT, lwd = 3, lty = 2)
  segments(lo, y, hi, y, col = ACCENT, lwd = 8)
  points(-tab$estimate, y, pch = 19, cex = 2.2, col = ACCENT)
}

# The discontinuity at a chosen window around the cut-off.
rdd_window <- function(width) {
  d <- subset(rdd_data, abs(distance) <= width)
  f <- lm_robust(waste_management_costs ~ eligible * distance, data = d)
  row_of(sprintf("Discontinuity, +/- %s points", width), f, "eligible", nrow(d))
}
```

::: {.panel}
On Monday you learned what a comparison is. On Tuesday you read three more ways
of building one. This morning you costed a programme.

Until now you have read someone else's analysis. Today you question the analyst
and decide whether you would act on the evaluation.
:::

::: {.ask .fragment}
The analyst is in this room, played by the trainer. You have the checklist. The
analyst has to answer, and today the analyst has the data open.
:::

## Would you sign off?

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">AED per business per year, the bar to clear</div>
</div>
```

::: {.ask}
**`[MENTI PLACEHOLDER]`** Would you sign off on this evaluation for national
scale-up? Yes / No / Not without more answers.
:::
:::::

::::: {.column width="42%"}
```{r out.width='280px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.warn}
Vote now. You vote again at the end, and the two votes are compared.
:::

::: {.notes}
Open the deck a couple of minutes before the session starts: R and its statistics package take about a minute to load in the browser, after which every live cell runs in a few seconds. A show of hands is enough if the Menti is not running. What matters is that the
room is on record before the analyst speaks.
:::

# 1 · The claim on the table {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What is being offered for sign-off

::: {.panel}
The GreenWaste evaluation. The published claim: the programme **reduced
waste-management costs by about 1,000 AED per business per year**, clearing the
rule for national scale-up.
:::

:::::: {.columns}

::::: {.column width="50%"}
**The evidence**

- A difference-in-differences estimate
- A regression discontinuity at the eligibility cut-off
- A randomised comparison of neighbourhoods
- A matched comparison
- A benefit-cost ratio over five years
:::::

::::: {.column width="50%"}
**What the analyst says it shows**

- Four methods, same direction, similar size
- The result is statistically significant
- The programme pays for itself
- Nothing depends on a single fragile choice
:::::

::::::

::: {.ask .fragment}
You already know one of those four statements is not true. Which one, and how
would you show it?
:::

## The numbers, as presented

```{r}
kable(data.frame(
  Method = c("Difference-in-differences", "Regression discontinuity",
             "Randomised offering", "Matched comparison"),
  `Estimate (AED)` = c("-816", "-905", "-1,014", "-1,000"),
  `What it covers` = c("Enrolled and not, offered neighbourhoods", "Businesses near the cut-off",
                       "All eligible businesses", "Matched enrolled businesses"),
  check.names = FALSE), align = "lrl")
```

::: {.warn .fragment}
Every method lands between 816 and 1,014 AED, and the rule is 1,000. The
randomised estimate clears it on its point estimate, the matched one sits on it,
and the other two fall short. None of this table shows an interval.
:::

## The rule, and the room to manoeuvre inside it

:::::: {.columns}

::::: {.column width="44%"}
```{=html}
<div class="facts">
  <div class="k">The rule</div>    <div class="v">At least 1,000 AED per business per year</div>
  <div class="k">The range</div>   <div class="v">816 to 1,014 AED</div>
  <div class="k">The claim</div>   <div class="v">"About 1,000"</div>
  <div class="k">The decision</div><div class="v">Scale up nationally</div>
</div>
```
:::::

::::: {.column width="56%"}
::: {.panel}
The claim reports a number no method produced, summarises a range that
straddles the threshold, and treats the threshold as crossed.
:::

::: {.ask .fragment}
Is "about 1,000" a rounding convention, or is the decision being made by the
rounding?
:::
:::::

::::::

## First question of the day

::: {.ask}
**Which of the estimates should the decision rest on, and why not the others?**
:::

::: {.panel .fragment}
Each estimate describes a different group of businesses: all eligible
businesses, those near a cut-off, the matched enrolled businesses, or the
enrolled and not enrolled in offered neighbourhoods. An evaluator who cannot say
which group the decision concerns cannot say which estimate supports it.
:::

# 2 · The checklist {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Four areas, in the order they break

```{r}
kable(data.frame(Area = qa_domains,
                 `What you are testing` = unname(qa_purpose[qa_domains]),
                 check.names = FALSE), align = "ll")
```

::: {.panel}
Methodology comes first: a design that cannot answer the question cannot be
rescued by good data. Conclusions come last: a conclusion is only as strong as
the three areas under it.
:::

## The five to ask every time

```{r results='asis'}
html_out(paste(sprintf('<div class="panel"><b>%d.</b> %s</div>', seq_along(qa_headline),
                       qa_headline), collapse = "\n"))
```

::: {.warn}
Five are easy to hold in your head in a meeting. The printed handout has all
twenty, five under each area.
:::

## How to tell a strong question from a weak one

:::::: {.columns}

::::: {.column width="50%"}
**Weak**

- Was the study rigorous?
- Are the results significant?
- Was the sample size adequate?
- Did you control for confounders?
- Is the data reliable?
:::::

::::: {.column width="50%"}
**Strong**

- Which assumption would reverse the sign?
- What is the estimate and its confidence interval?
- How many units were dropped, and why?
- What changed at the same time?
- Who is missing from the data?
:::::

::::::

::: {.ask .fragment}
Anyone can answer a weak question with "yes". A strong question needs a fact to
answer it.
:::

## Your turn: sort these

```{r results='asis'}
cands <- c(
  "Did the programme work?",
  "What would have happened to these businesses without the programme?",
  "Was the sample large enough?",
  "How many businesses dropped out between the two survey rounds?",
  "Were the results statistically significant?",
  "Which assumption, if wrong, would flip the finding?",
  "Is this good quality data?",
  "If the effect were half this size, would you still recommend scale-up?"
)
html_out(paste0('<div class="excerpt">',
                paste(sprintf('<p><b>%d.</b> %s</p>', seq_along(cands), cands), collapse = ""),
                '</div>'))
```

::: {.ask}
Mark each one **S** or **W**. Then say what makes the S ones answerable.
:::

::: {.notes}
Paper (Oct14_session3_materials.docx): Hand out the eight question slips (pack page 1, cut, one set per group) to sort into strong and weak.
:::

## Which of those survived

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Weak, and why**

- *Did the programme work?* One word answers it.
- *Was the sample large enough?* For what? Precision is not the problem here.
- *Were the results significant?* Yes or no, where the number is what matters.
- *Is this good quality data?* Invites an adjective, not a fact.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**Strong, and why**

- *What would have happened without it?* Asks for the counterfactual.
- *How many dropped out?* A count the analyst has or does not have.
- *Which assumption would flip it?* Brings the weak point into the open.
- *Half the size?* Tests whether the estimate is carrying the decision.
:::
:::::

::::::

::: {.notes}
Point out that the strong questions are no harder to ask; the difference is
that they need a fact to answer. Section 4 shows the analyst producing those
facts.
:::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the clinic, with the analyst in the room.
:::

# 3 · The clinic {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## How this runs

:::::: {.columns}

::::: {.column width="58%"}
::: {.panel}
The analyst answers as themselves. Groups ask questions, one area at a time.

1. One question at a time; the answer comes before the next question
2. Any group may follow up once on an answer
3. The analyst may say *"I don't know"* or *"we didn't check that"*
4. An admission is useful information, not a failure
5. Nothing is scored
:::
:::::

::::: {.column width="42%"}
::: {.warn}
**Four groups**, one per area: methodology, assumptions, data quality,
conclusions.

Five minutes to agree your three sharpest questions before the clinic opens.
:::
:::::

::::::

::: {.notes}
When a group asks one of the questions in section 4, jump to that slide and run
it: the analyst answers with the data. Questions without a prepared slide get a
spoken answer, or an honest "we did not check".

Paper (Oct14_session3_materials.docx): Hand out the QA checklist (Oct14_session3_qa_checklist.docx) and one clinic record sheet per group (pack page 2).
:::

## The answers that cost the analyst something

::: {.panel}
Some answers make the analysis look better. Some make it look worse. You are
looking for the second kind.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Listen for these phrases**

- "That's a good question…"
- "In the main specification, yes"
- "We discussed that"
- "That was outside the scope"
- "The data doesn't let us"
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**And follow up with**

- *Which other specification?*
- *What happened in the others?*
- *What did you decide, and why?*
- *Whose scope? Who set it?*
- *What would you need to answer it?*
:::
:::::

::::::

## The one you must not let go

::: {.ask}
**What is the comparison, and how was it chosen?**
:::

::: {.panel .fragment}
Every method this week rests on a comparison group that matched the
participants before the programme began: because it was drawn by lot, sat just
past a cut-off, was matched on what was measured, or was trending in parallel.
Those are four different reasons, and each can fail in its own way. The analyst
should be able to say, in one sentence, which one applies here.
:::

# 4 · The analyst answers with the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## "What is the estimate, and how sure are you?"

```{webr}
#| fig-width: 7.5
#| fig-height: 2.2
tab <- estimates()
round(tab[, -1])
interval_plot(tab)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Three estimates rerun live, each with its 95% interval. The matched −1,000 is as
reported.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
No interval clears the 1,000 line at its least generous end.
:::
:::::

::::::

## "Does anything depend on a single choice?"

```{ojs}
//| echo: false
viewof width = Inputs.range([2, 12], {value: 12, step: 1, label: "Window around the cut-off (index points)"})
```

```{webr}
#| input:
#|   - width
#| autorun: true
#| edit: false
round(rdd_window(width)[, -1])
```

::: {.ask}
The analyst said nothing depends on a fragile choice. Narrow the window the
discontinuity uses. **What happens to the estimate?**
:::

::: {.warn .fragment}
At ±2 points it falls to about 790 AED; at ±6 it is about 1,080. The same data
gives an answer on either side of the rule depending on one analyst choice. The
fourth statement on the claim slide is the false one.
:::

## "Who is each estimate about?"

```{webr}
data.frame(
  estimate = c("Difference-in-differences", "Discontinuity", "Randomised"),
  businesses = c(length(unique(offered$business_identifier)), nrow(rdd_data), nrow(rct_data)),
  who = c("offered neighbourhoods, enrolled and not", "offered neighbourhoods, 2021",
          "eligible businesses, all neighbourhoods"))
```

::: {.panel}
Different groups of businesses, so different questions. A national roll-out
would reach businesses like the eligible ones in every neighbourhood.
:::

::: {.ask .fragment}
Which estimate speaks to the businesses a national roll-out would reach?
:::

## "Who is missing from the data?"

```{webr}
data.frame(rows_per_business = names(table(table(gw$business_identifier))),
           businesses = as.vector(table(table(gw$business_identifier))))
missing <- colSums(is.na(gw))
missing[missing > 0]
with(gw, table(compliance_missing = is.na(recycling_compliance), enrolled))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Every business appears in both rounds, so no one dropped out. But one column is
empty for 8,570 rows.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Recycling compliance was recorded only for eligible businesses. Any analysis
that controls for it silently drops every ineligible business. Ask what was
dropped, not only what was measured.
:::
:::::

::::::

## "Were the groups on the same path before?"

```{webr}
table(round = gw$round)
```

::: {.ask}
The parallel-trends question from Day 2. What can the analyst show?
:::

::: {.panel .fragment}
Nothing: there are only two rounds, one before and one after. The honest answer
is "the data does not let us check." That is an admission, and it belongs in the
decision: the difference-in-differences estimate rests on an assumption nobody
tested.
:::

# 5 · The debrief {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The unanswered questions are the finding

::: {.panel}
For each area, how many of your questions got a definite answer, and how many
got an admission?
:::

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Answered</div>     <div class="v">The fact exists and was produced</div>
  <div class="k">Admitted</div>     <div class="v">"We didn't check that"</div>
  <div class="k">Deflected</div>    <div class="v">Answered a different question</div>
  <div class="k">Unanswerable</div> <div class="v">The design cannot settle it</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.ask .fragment}
Which of the four is not a failure, and which one tells you the analysis is
weaker than it looked?
:::
:::::

::::::

::: {.warn .fragment}
A deflection is worse than an admission. An admission shows where the evidence
stops. A deflection suggests the analyst may not know either.
:::

## Which unanswered question changes the decision

::: {.ask}
Of everything the analyst could not answer, which **one** would change your
sign-off if it came back the wrong way?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Most unanswered questions are interesting and not decisive. They belong in the
limitations section.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
One or two are decisive. Those are worth commissioning more work on, or writing
into the decision as a condition.
:::
:::::

::::::

## Would you sign off now?

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">The same bar, after the clinic</div>
</div>
```

::: {.ask}
**`[MENTI PLACEHOLDER]`** Same three options. If your vote moved, say which
answer moved it.
:::
:::::

::::: {.column width="42%"}
```{r out.width='280px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

# 6 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The prompt

::: {.excerpt}
I have an evaluation of a programme that subsidised waste-management technology
for businesses. It used difference-in-differences, regression discontinuity,
a randomised comparison and matching. All four methods found a cost reduction of
roughly 1,000 AED per business per year. List the critical questions a
commissioner should ask before acting on this result.
:::

::: {.panel}
A common kind of prompt. Run it yourselves, then compare what you get with the
example on the next slide and with your own questions.
:::

::: {.notes}
Paper (Oct14_session3_materials.docx): Hand out the AI prompt card (pack page 3, one per group).
:::

## What came back

:::::: {.columns}

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**1. Was the study well designed?** It used a comparison group and a pre-post
design. **2. Was the sample adequate?** About 9,900 observations, so power is
[not a concern here]{.flag} and the results can be trusted. **3. Were the
results significant?** Significance is the main sign a result is real.
**4. Were confounders controlled for?** Four methods suggest robustness.
**5. Is the data reliable?** Check for missing values. **6. Is it policy
relevant?** The reduction appears to exceed the threshold.

**Overall:** the convergence of four methods on a similar estimate is [strong
evidence of causality]{.flag}.
:::
:::::

::::: {.column width="42%"}
::: {.ask}
Find the two flagged errors. Then find the bigger problem, which is not flagged.
How does your AI's list compare?
:::

::: {.panel .fragment}
The bigger problem: none of the six questions needs a fact to answer. Every one
would pass the sorting exercise as **weak**.
:::
:::::

::::::

## What AI got, and what it could not get

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- Produced the shape of a critique
- Covered the right broad topics
- Read fluently and organised itself
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it missed**

- A large sample does not remove selection bias; both flagged errors confuse the two.
- It cannot know which assumption this design rests on: it was never told.
- It never asks what the comparison was.
- It never asks what was not checked, which is where the clinic found most.
:::
:::::

::::::

## The same task, a better prompt

::: {.excerpt}
I am advising on whether to scale up a programme nationally. An evaluator reports
a cost reduction of about 1,000 AED per business per year from
difference-in-differences, with a decision threshold of 1,000 AED. First, tell me
which single assumption this design rests on. Second, tell me what evidence would
show that assumption holds, and what it would look like if it failed. Third, tell
me what you cannot determine from the information I have given you.
:::

::: {.panel .fragment}
Three changes: ask for **one** assumption, not a list; ask what would **show it
failing**; and ask what it **cannot tell you**, which lets the model admit a
limit instead of writing around it.
:::

# 7 · Take this back to your desk {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The three you will use

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**From memory, no notes**

1. What is the comparison, and how was it chosen?
2. What has to be true for this to be causal, and did you test it?
3. Does the conclusion follow, or does it go beyond the evidence?
:::
:::::

::::: {.column width="50%"}
::: {.warn}
If you remember one, remember the first. Every method this week is a different
answer to it.
:::
:::::

::::::

::: {.ask .fragment}
Which of the three will you ask first next time, and does it depend on the
method?
:::

::: {.notes}
Paper (Oct14_session3_materials.docx): Hand out the take-away cards (pack page 4).
:::

## Tomorrow

::: {.panel}
The full checklist, applied to a written report section by section, with a
traffic-light rating for each. Bring the questions you wrote down today.
:::
