---
title: "Introduction to Impact Evaluation: Key Concepts and Terminology"
subtitle: "Module 2 · Day 1, Session 1: Building a Common Language for Causal Analysis"
author: "Dr. Lucas Sempé"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(qrcode)      # live poll QR, generated below so it cannot drift

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / offered
ACCENT2 <- "#A9561F"   # series 2: comparison / not offered
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

# ---- Chart sizing ----------------------------------------------------------
# Every chart carries an explicit pixel size. Relying on plotly's autosize
# inside revealjs produces half-drawn charts: a widget initialised while its
# slide is hidden measures a narrow container and never corrects itself.
# These must stay in step with the chunk's fig.width / fig.height, which size
# the htmlwidgets wrapper (at dpi 96, one inch is 96 px).

PL_W  <- 1100   # full-width chart
PL_WC <- 520    # chart sitting in a half-width column
PL_WT <- 340    # chart sitting in a third-width column

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width  = width,
      height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. Everything below reads from here: the QR on
# the opening slide, the printed link, and the results iframe on the reveal
# slide. Change MENTI_CODE only, and all three move together.
#
# This is the short JOIN code participants type at menti.com (usually 8 digits),
# NOT the long alphanumeric ID in the editor URL. Confirm it while the Menti is
# live: a code that is not currently running sends the room to a "no longer
# active" page.
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

# The results embed. This is the SLIDES view for a webpage, not the participant
# voting view: Mentimeter does not expose voting to embeds, so the room still
# votes on their own phones. Set this to the long ID from the editor URL.
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/al68xcnp6hm6zmc9k94rhbxwimzq6m29/embed"

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))   # ecl H = highest error correction
  path
}
MENTI_QR <- menti_qr()

# ---- Section 1: the Al Falah campaign, in two worlds ------------------------
# One headline, two sets of driving data, opposite verdicts. The death rate is
# what breaks the tie, and the headline alone cannot tell you which world you
# are in.
af_fatal_before <- 420
af_fatal_after  <- 294
af_vkm_before   <- 8.0
af_vkm_world_a  <- 5.6      # fuel prices spike, people drive less
af_vkm_world_b  <- 8.0      # driving unchanged

af_rate_before   <- af_fatal_before / af_vkm_before
af_rate_world_a  <- af_fatal_after  / af_vkm_world_a
af_rate_world_b  <- af_fatal_after  / af_vkm_world_b

af_headline_fall  <- 100 * (af_fatal_before - af_fatal_after) / af_fatal_before
af_drive_fall     <- 100 * (af_vkm_before - af_vkm_world_a) / af_vkm_before
af_rate_fall_a    <- 100 * (af_rate_before - af_rate_world_a) / af_rate_before
af_rate_fall_b    <- 100 * (af_rate_before - af_rate_world_b) / af_rate_before

# ---- Section 1 coda: the trap runs the other way too ------------------------
mc_prog <- c(62, 60)        # youth employment rate, programme areas
mc_comp <- c(61, 55)        # youth employment rate, comparison areas
mc_naive  <- mc_prog[2] - mc_prog[1]
mc_change <- mc_comp[2] - mc_comp[1]
mc_effect <- mc_naive - mc_change

# ---- Word 1: mean hides spread. Two programmes, same average ---------------
set.seed(7)
mean_dat <- bind_rows(
  tibble(Programme = "Programme A", cost = rgamma(120, shape = 40, rate = 40 / 1400)),
  tibble(Programme = "Programme B", cost = rgamma(120, shape = 2, rate = 2 / 1400))
) %>%
  group_by(Programme) %>%
  # Scale each programme so both averages are exactly 1,400 AED. Sampling
  # alone left them 1,462 and 1,53x, which would make the single average line
  # on the chart a small lie and the caption imprecise. Scaling, not shifting,
  # so no cost can go negative.
  mutate(cost = cost * (1400 / mean(cost))) %>%
  ungroup() %>%
  mutate(Programme = factor(Programme, levels = c("Programme A", "Programme B")),
         id  = row_number(),
         row = as.integer(Programme),
         x   = row + runif(n(), -0.34, 0.34))

mean_summary <- mean_dat %>%
  group_by(Programme) %>%
  summarise(avg  = mean(cost),
            near = 100 * mean(abs(cost - 1400) <= 140),
            .groups = "drop")

# ---- Word 2: the 2x2 the room does by hand ---------------------------------
te_prog_before <- 54; te_prog_after <- 71
te_comp_before <- 50; te_comp_after <- 65

# Named individually so the reveal table can show all four in a row, rather
# than making the reader hold two of them in their head while a third is
# explained. The two "change" columns are what line up to give the effect.
te_prog_change <- te_prog_after - te_prog_before      # 17
te_comp_change <- te_comp_after - te_comp_before      # 15

te_change      <- te_prog_after - te_prog_before      # 17 (kept: used by the chart)
te_end_gap     <- te_prog_after - te_comp_after       # 6
te_base_gap    <- te_prog_before - te_comp_before     # 4
te_effect      <- te_prog_change - te_comp_change     # 2

# ---- Word 3: a coefficient is a gap between groups -------------------------
set.seed(11)
reg_n <- 240
coef_dat <- tibble(
  id    = 1:reg_n,
  group = factor(rep(c("Comparison", "Programme"), each = reg_n / 2),
                 levels = c("Comparison", "Programme"))
) %>%
  mutate(outcome = 40.5 + 2.1 * (group == "Programme") + rnorm(n(), 0, 6),
         x       = as.integer(group) + runif(n(), -0.30, 0.30))

coef_means <- coef_dat %>%
  group_by(group) %>%
  summarise(m = mean(outcome), .groups = "drop")

coef_gap <- coef_means$m[coef_means$group == "Programme"] -
            coef_means$m[coef_means$group == "Comparison"]

reg_tbl <- tribble(
  ~Variable,                     ~Coef,  ~SE,   ~p,      ~CI_lo, ~CI_hi,
  "(Intercept)",                  40.5,  3.2,   "<0.001", 34.2,   46.8,
  "Programme (enrolled = 1)",      2.1,  0.9,   "0.021",   0.3,    3.9,
  "Baseline pass rate",            0.58,  0.04,  "<0.001",  0.50,   0.66,
  "Class size (students)",        -0.32,  0.14,  "0.024",  -0.59,  -0.05,
  "Teacher experience (years)",    0.15,  0.08,  "0.06",   -0.01,   0.31
)
te_threshold <- 3   # the effect must reach 3 points to justify scale-up

# ---- Words 4 and 5: the same effect of 2.1 points, seen two ways -----------
# perm_null() builds a study that measures exactly `effect`, then breaks the
# link between the label and the outcome 1,000 times. What survives is the
# distribution of differences a world with no effect produces.
perm_null <- function(n_total, effect = 2.1, sd_y = 7, n_perm = 1000, seed = 2026) {
  set.seed(seed)
  half <- n_total %/% 2
  g    <- rep(c("Comparison", "Programme"), each = half)
  y    <- 60 + rnorm(2 * half, 0, sd_y)
  y    <- y + (effect - (mean(y[g == "Programme"]) - mean(y[g == "Comparison"]))) *
         (g == "Programme")
  replicate(n_perm, {
    i <- sample(2 * half)
    mean(y[i][1:half]) - mean(y[i][(half + 1):(2 * half)])
  })
}

perm_240 <- perm_null(240)
p_240    <- mean(abs(perm_240) >= 2.1)

pval_bins <- seq(floor(min(perm_240) * 4) / 4, ceiling(max(perm_240) * 4) / 4,
                 by = 0.25)
pval_h0   <- hist(perm_240, breaks = pval_bins, plot = FALSE)

pval_hist <- tibble(mid = pval_h0$mids, n = pval_h0$counts)
pval_lo   <- min(perm_240)
pval_hi   <- max(perm_240)
pval_ymax <- max(pval_h0$counts) * 1.18

# The sample-size slider: identical effect, seven studies of growing size.
n_grid     <- c(40, 80, 160, 240, 480, 960, 2400)
n_perms    <- map(n_grid, perm_null)
names(n_perms) <- as.character(n_grid)

slider_bins <- seq(-9, 9, by = 0.25)
slider_mids <- slider_bins[-length(slider_bins)] + 0.125

null_frames <- map_dfr(n_grid, function(nn) {
  h <- hist(n_perms[[as.character(nn)]], breaks = slider_bins, plot = FALSE)$counts
  tibble(n = nn, mid = slider_mids, share = h / sum(h))
})
slider_ymax <- max(null_frames$share) * 1.15

null_p <- tibble(
  n = n_grid,
  p = map_dbl(n_grid, function(nn) mean(abs(n_perms[[as.character(nn)]]) >= 2.1))
)

# The interval, 100 repeat studies of the same programme
set.seed(4)
mu_ci <- 2.1; se_ci <- 0.45
ci_dat <- tibble(run = 1:100, est = rnorm(100, mu_ci, se_ci)) %>%
  mutate(lo     = est - 1.96 * se_ci,
         hi     = est + 1.96 * se_ci,
         covers = lo <= mu_ci & hi >= mu_ci,
         txt    = sprintf("Study %d: %.1f points [%.1f, %.1f]", run, est, lo, hi)) %>%
  arrange(est) %>%
  mutate(y = row_number())

ci_cov <- sum(ci_dat$covers)
```

# 1 · What are we comparing?

## The Al Falah Road Safety Campaign

:::::: {.columns}

::::: {.column width="50%"}
A national government runs a one-year awareness campaign: billboards, TV
spots, and school talks promoting seatbelt use and safe speeds.

At the end of the year, someone has to decide: did it work, and should it
run again?

::: {.warn}
*This is an illustrative example, not a real Abu Dhabi case study.*
:::
:::::

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Ran</div>        <div class="v">One year, nationwide</div>
  <div class="k">Delivered</div>  <div class="v">Billboards, TV spots, school talks</div>
  <div class="k">Outcome</div>    <div class="v">Traffic fatalities</div>
  <div class="k">The question</div> <div class="v">Did the campaign save lives?</div>
</div>
```
:::::

::::::

## The headline number

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Traffic fatalities, year on year</div>
</div>
<div class="stat-row">
  <div><span class="n">%d</span><span class="k">Deaths, year before</span></div>
  <div><span class="n">%d</span><span class="k">Deaths, year after</span></div>
  <div><span class="n">%d</span><span class="k">Fewer deaths</span></div>
</div>', af_headline_fall, af_fatal_before, af_fatal_after,
   af_fatal_before - af_fatal_after))
```

## Many times we make decisions based on numbers

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
*Your first decision.* The programme reported that traffic fatalities fell by
30% in one year.

**Would you fund a second year?**
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right. Please, give an answer.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.warn .fragment}
Every number you will be shown this week is **correct**. Almost none of them will
be *wrong*.

BUT... sometimes an accurate
number answers a question you did not ask, so we might still be wrong.
:::

## What else happened that year

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.6, fig.width=5.4}
tibble(Time = tlev(c("Before", "After")),
       Deaths = c(af_fatal_before, af_fatal_after)) %>%
  plot_ly(x = ~Time, y = ~Deaths, type = "bar",
          marker = list(color = ACCENT, line = list(width = 0)),
          text = ~Deaths, textposition = "outside",
          textfont = list(size = 15, color = INK),
          showlegend = FALSE,
          hovertemplate = "%{y} deaths<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Traffic fatalities", range = c(0, 520))) %>%
  pl_m2(width = PL_WC, height = 250, title = "Fatalities, nationwide")
```
:::::

::::: {.column width="50%"}
```{r fig.height=2.6, fig.width=5.4}
tibble(Time = tlev(c("Before", "After")),
       VKM = c(af_vkm_before, af_vkm_world_a)) %>%
  plot_ly(x = ~Time, y = ~VKM, type = "bar",
          marker = list(color = ACCENT2, line = list(width = 0)),
          text = ~VKM, textposition = "outside",
          textfont = list(size = 15, color = INK),
          showlegend = FALSE,
          hovertemplate = "%{y} billion km<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Billion km driven", range = c(0, 9.5))) %>%
  pl_m2(width = PL_WC, height = 250, title = "Distance driven, nationwide")
```
:::::

::::::

::: {.panel}
Fuel prices rose sharply that year, and drivers cut back: distance driven fell
by `r sprintf("%.0f", af_drive_fall)`%, almost exactly as much as fatalities
did.
:::

## World B · the same headline, a real effect

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Deaths per billion km driven</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f</span><span class="k">Rate before</span></div>
  <div><span class="n">%.0f</span><span class="k">Rate after</span></div>
  <div><span class="n">-%d%%</span><span class="k">The campaign worked</span></div>
</div>', af_rate_fall_b, af_rate_before, af_rate_world_b, af_rate_fall_b))
```

::: {.warn .fragment}
World B is not what happened. It is what the **same headline** would have meant
if driving had not changed at all: every one of those 126 fewer deaths would be
the campaign's doing, and the death rate would have fallen by 30%.
:::



## World A · the death rate did not move

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.0f%%</div>
  <div class="label">Deaths per billion km driven</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f</span><span class="k">Rate before</span></div>
  <div><span class="n">%.0f</span><span class="k">Rate after</span></div>
  <div><span class="n">%d</span><span class="k">Lives saved by the campaign</span></div>
</div>', af_rate_fall_a, af_rate_before, af_rate_world_a,
   round(af_rate_before - af_rate_world_a)))
```

::: {.warn .fragment}
If the roads emptied by 30% and deaths fell by 30%, the risk of dying on them
did not change. The campaign saved nobody. The headline was the fuel price.
:::

::: {.panel .fragment}
Two worlds, one headline, opposite verdicts. Nothing in the headline, or in the
number 30%, tells you which world you are in.
:::

## Did it work?

```{r results='asis'}
# data-external="1" is required. Without it, embed-resources: true inlines the
# page as a base64 data: URI with role="img", and the embed becomes a frozen
# picture instead of a live results view.
cat(sprintf('
<iframe data-external="1" allowfullscreen="true" allowtransparency="true"
        frameborder="0" src="%s"
        style="width:100%%; height:74%%; border:none;"></iframe>', MENTI_EMBED))
```

## The trap also runs the other way

```{r fig.height=3, fig.width=11.5}
tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Programme districts", "Comparison districts"), each = 2),
  Rate  = c(mc_prog[1], mc_prog[2], mc_comp[1], mc_comp[2])) %>%
  plot_ly(x = ~Time, y = ~Rate, color = ~Group,
          type = "scatter", mode = "lines+markers",
          colors = c("Programme districts" = ACCENT,
                     "Comparison districts" = ACCENT2),
          line = list(width = 3), marker = list(size = 11),
          hovertemplate = "%{fullData.name}: %{y}%<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Youth employment rate (%)", range = c(50, 66))) %>%
  pl_m2(height = 290)
```

::: {.warn}
A youth employment programme: employment fell `r abs(mc_naive)` points in the
districts it ran in. On a before-and-after reading, it failed. But employment
fell `r abs(mc_change)` points in the comparison districts, so the programme
held up `r mc_effect` points that would otherwise have been lost.
:::

## "Compared to what?"

::: {.panel}
Every evaluation result is a comparison. A before-and-after number compares the
programme to *the past*, not to what would have happened anyway. The comparison
can move the answer down, as in World A, or up, as in the mirror case. The
direction is not knowable in advance. If you do not know what a number is being
compared to, you do not yet know what it means.
:::

# 2 · Three questions for any number

## One idea, three questions

::: {.panel}
Five words follow. They are easier to hold if you attach them to the three
questions you are entitled to ask of any number you are handed.
:::

```{=html}
<div class="facts">
  <div class="k">Compared to what?</div>          <div class="v">Comparison group · the question nobody puts in the summary</div>
  <div class="k">How big, in units that matter?</div><div class="v">Mean · treatment effect · coefficient</div>
  <div class="k">How sure are we?</div>           <div class="v">p-value · confidence interval</div>
</div>
```

::: {.ask}
Keep these three questions. You will use them on every slide for the rest of the week,
and on the next evaluation report that lands on your desk.
:::

# 3 · The five words

## Word 1 · Mean

```{r fig.height=3, fig.width=11.5}
plot_ly() %>%
  add_trace(data = filter(mean_dat, Programme == "Programme A"),
            x = ~x, y = ~cost, type = "scatter", mode = "markers",
            marker = list(size = 7, color = ACCENT, opacity = 0.5),
            customdata = ~id, name = "Programme A",
            hovertemplate = "Programme A: %{y:,.0f} AED<extra></extra>") %>%
  add_trace(data = filter(mean_dat, Programme == "Programme B"),
            x = ~x, y = ~cost, type = "scatter", mode = "markers",
            marker = list(size = 7, color = ACCENT2, opacity = 0.5),
            customdata = ~id, name = "Programme B",
            hovertemplate = "Programme B: %{y:,.0f} AED<extra></extra>") %>%
  add_trace(x = c(0.4, 2.6), y = c(1400, 1400), type = "scatter", mode = "lines",
            line = list(color = INK, width = 3, dash = "dot"),
            name = "Average of each programme", hoverinfo = "name") %>%
  layout(xaxis = ax("", tickvals = c(1, 2),
                    ticktext = c("Programme A", "Programme B"),
                    range = c(0.4, 2.6), zeroline = FALSE),
         yaxis = ax("Waste management cost (AED)", range = c(0, 6200))) %>%
  pl_m2(height = 290)
```

::: {.ask}
Each dot is one business. Both programmes average exactly 1,400 AED

In Programme A,
`r sprintf("%.0f", mean_summary$near[mean_summary$Programme == "Programme A"])`%
of businesses are within 10% of the average. In Programme B, only
`r sprintf("%.0f", mean_summary$near[mean_summary$Programme == "Programme B"])`%.

**Which would you fund, and what else do you need to know?**
:::

## Word 2 · Treatment effect

:::::: {.columns}

::::: {.column width="44%"}
```{=html}
<div class="facts">
  <div class="k">Programme schools</div>   <div class="v">54% &rarr; 71%</div>
  <div class="k">Comparison schools</div>  <div class="v">50% &rarr; 65%</div>
  <div class="k">Decision rule</div>       <div class="v">Must add 3 points to justify scale-up</div>
</div>
```

::: {.ask}
Grade 6 maths pass rates.

**Three numbers describe this result: 17, 6 and 2.**

Work out what each one is, then decide which is the treatment effect.
:::
:::::

::::: {.column width="56%"}
::: {.fragment}
```{r results='asis'}
# The reveal. Every number the reader needs sits in the grid, and the bottom
# row is simply the two above it subtracted, so the effect can be read off
# the Change column (17 - 15) or the After column (71 - 65) and checked
# against the Before column (54 - 50). No number is ever held in the head.
te_tbl <- tribble(
  ~Group,               ~Before,          ~After,          ~Change,
  "Programme schools",  te_prog_before,   te_prog_after,   te_prog_change,
  "Comparison schools", te_comp_before,   te_comp_after,   te_comp_change,
  "Difference",         te_base_gap,      te_end_gap,      te_effect
)

te_tbl %>%
  kable(align = "lrrr", col.names = c("", "Before", "After", "Change")) %>%
  kable_styling(full_width = FALSE, font_size = 19) %>%
  row_spec(2, hline_after = TRUE) %>%
  row_spec(3, bold = TRUE)
```

::: {.panel}
The bottom row is the two above it subtracted.

**17** is not the answer. The national reform was lifting both groups.

**6** is not the answer either. It carries the 4-point head start the
programme schools already had.

**2** is the treatment effect: 17 minus 15, and equally 6 minus 4.
:::
:::
:::::

::::::

## Word 2 · The effect depends on the comparison

::: {.panel}
The programme schools are fixed at 54% then 71%. Only the comparison schools
move. The red bar is the treatment effect, and it is the only thing on the
chart that responds.
:::

```{r fig.height=3.35, fig.width=11.5}
cf_seq <- seq(48, 72, by = 2)

te_lines <- expand_grid(cf = cf_seq, Time = tlev(c("Before", "After"))) %>%
  mutate(Programme  = ifelse(Time == "After", te_prog_after, te_prog_before),
         Comparison = ifelse(Time == "After", cf, te_comp_before)) %>%
  pivot_longer(c(Programme, Comparison), names_to = "Group", values_to = "Rate")

te_gap <- expand_grid(cf = cf_seq, pt = 1:2) %>%
  mutate(Time  = tlev("After"),
         Group = "Effect",
         Rate  = ifelse(pt == 1, cf, te_prog_after)) %>%
  select(cf, Time, Group, Rate)

bind_rows(te_lines, te_gap) %>%
  plot_ly(x = ~Time, y = ~Rate, color = ~Group, frame = ~cf,
          type = "scatter", mode = "lines+markers",
          colors = c(Programme = ACCENT, Comparison = ACCENT2, Effect = ALERT),
          line = list(width = 3), marker = list(size = 9),
          hovertemplate = "%{fullData.name}: %{y}%<extra></extra>") %>%
  animation_opts(frame = 700, transition = 0) %>%
  animation_slider(currentvalue = list(prefix = "Comparison schools end on "),
                   font = list(size = 11)) %>%
  layout(xaxis = ax("", zeroline = FALSE,
                    type = "category", categoryorder = "array",
                    categoryarray = c("Before", "After")),
         yaxis = ax("Grade 6 maths pass rate (%)", range = c(40, 80))) %>%
  pl_m2(height = 320, t = 34, b = 92)
```

::: {.ask}
Press play, or drag the slider. Programme schools never move. Which comparison
value makes the effect zero, and what does that value mean in plain words?
:::

## Word 3 · Coefficient

::: {.panel}
A coefficient reports the size of a relationship as a **rate**: how much the
outcome moves per one unit of whatever sits on the row. On an evaluation table,
the coefficient on the programme row is often the treatment effect itself.
:::

```{r fig.height=3, fig.width=11.5}
plot_ly() %>%
  add_trace(data = coef_dat, x = ~x, y = ~outcome, type = "scatter", mode = "markers",
            marker = list(size = 7, color = ACCENT, opacity = 0.45),
            showlegend = FALSE, customdata = ~id,
            hovertemplate = "School %{customdata}: %{y:.1f}%<extra></extra>") %>%
  add_trace(data = coef_means, x = ~as.integer(group), y = ~m,
            type = "scatter", mode = "markers",
            marker = list(size = 17, color = INK, symbol = "diamond"),
            name = "Group mean", hovertemplate = "Mean: %{y:.1f}%<extra></extra>") %>%
  add_segments(x = 2.38, xend = 2.38,
               y = coef_means$m[coef_means$group == "Comparison"],
               yend = coef_means$m[coef_means$group == "Programme"],
               line = list(color = ALERT, width = 4),
               name = sprintf("Coefficient = %.1f points", coef_gap),
               hoverinfo = "name") %>%
  layout(xaxis = ax("", tickvals = c(1, 2),
                    ticktext = c("Comparison", "Programme"),
                    range = c(0.5, 2.75), zeroline = FALSE),
         yaxis = ax("Grade 6 maths pass rate (%)", range = c(30, 55))) %>%
  pl_m2(height = 290)
```

::: {.panel}
Hover the diamonds, then the red bar. Here the coefficient is the vertical gap
between the two group means, `r sprintf("%.1f", coef_gap)` points. That gap is
what the model reports on the programme row of the next slide.
:::

## Reading the table column by column

```{r}
reg_tbl %>%
  mutate(CI = sprintf("[%.2f, %.2f]", CI_lo, CI_hi)) %>%
  select(Variable, Coef, SE, p, CI) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("Variable", "Coefficient", "Std. error", "p-value", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 22) %>%
  row_spec(2, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = paste("N =", reg_n, "schools. Outcome: Grade 6 maths pass rate."),
           general_title = "", footnote_as_chunk = TRUE)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Programme row:** the one number that answers the decision. `2.1` points,
interval `[0.3, 3.9]`. The rule needs 3 points. The interval contains 3, so we
cannot say whether the programme clears the bar.

**Intercept:** the value when everything on the table is zero. Nobody is a
school with zero students and zero teachers. Ignore it.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**Covariate rows:** class size, teacher experience, baseline pass rate. These
adjust the comparison. They are not recommendations: the coefficient on teacher
experience is `0.15`, and its interval `[-0.01, 0.31]` crosses zero, so it
cannot rule out no effect at all.

**Std. error:** the wobble. Divide the coefficient by it to see how steady the
estimate is.
:::
:::::

::::::

## Word 4 · p-value

::: {.panel}
**Plain language:** roughly, how surprising this result would be if the
programme had truly done nothing at all. A small p-value means "this would be a
strange coincidence if there were no effect."
:::

```{r fig.height=3, fig.width=11.5}
plot_ly(pval_hist, x = ~mid, y = ~n, type = "bar",
        marker = list(color = ACCENT, line = list(width = 0)),
        showlegend = FALSE,
        hovertemplate = paste0("A shuffled difference of %{x:.1f} points<br>",
                               "happened in %{y} of 1,000 shuffles<extra></extra>")) %>%
  layout(bargap = 0.15,
         xaxis = ax("Difference between the two groups (points)"),
         yaxis = ax("Number of shuffles", range = c(0, pval_ymax)),
         shapes = list(
           list(type = "rect", x0 = pval_lo, x1 = -abs(2.1),
                y0 = 0, y1 = 1, yref = "paper",
                fillcolor = "rgba(142, 36, 19, 0.10)", line = list(width = 0)),
           list(type = "rect", x0 = abs(2.1), x1 = pval_hi,
                y0 = 0, y1 = 1, yref = "paper",
                fillcolor = "rgba(142, 36, 19, 0.10)", line = list(width = 0)),
           list(type = "line", x0 = 2.1, x1 = 2.1,
                y0 = 0, y1 = 1, yref = "paper",
                line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 2.1, y = 1, yref = "paper",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "what we measured (2.1 points)",
                                 font = list(size = 12, color = ALERT)))) %>%
  pl_m2(height = 290)
```

::: {.ask}
We took the 240 schools, broke the link between the programme and the outcome,
and did it 1,000 times. The red tails are the shuffles that beat what we
measured: `r sprintf("%.3f", p_240)`, or about `r round(p_240 * 1000)` of 1,000.
That fraction is the p-value.
:::

## Word 4 · Why significance is not importance

::: {.panel}
Same programme. Same effect of 2.1 points. Three studies of different sizes.
Press play, or drag the slider, and watch the shuffled distribution tighten
around zero. The effect never changes.
:::

```{r fig.height=3.35, fig.width=11.5}
plot_ly(null_frames, x = ~mid, y = ~share, frame = ~n, type = "bar",
        marker = list(color = ACCENT, line = list(width = 0)),
        showlegend = FALSE,
        hovertemplate = paste0("A shuffled difference of %{x:.2f} points<br>",
                               "in %{y:.1%} of the shuffles<extra></extra>")) %>%
  animation_opts(frame = 800, transition = 0, redraw = FALSE) %>%
  animation_slider(currentvalue = list(prefix = "Schools in the study: "),
                   font = list(size = 11)) %>%
  layout(bargap = 0.05,
         xaxis = ax("Difference between the two groups (points)",
                    range = c(-9, 9)),
         yaxis = ax("Share of the 1,000 shuffles", range = c(0, slider_ymax)),
         shapes = list(list(type = "line", x0 = 2.1, x1 = 2.1,
                            y0 = 0, y1 = 1, yref = "paper",
                            line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 2.1, y = 1, yref = "paper",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "the effect we measured (2.1 points)",
                                 font = list(size = 12, color = ALERT)))) %>%
  pl_m2(height = 320, t = 34, b = 92)
```

::: {.warn}
At 40 schools the same effect gives p = `r sprintf("%.2f", null_p$p[null_p$n == 40])`.
At 240 schools, p = `r sprintf("%.2f", null_p$p[null_p$n == 240])`. At 2,400
schools it is below 0.001. The programme did not improve by a single point.
**Statistically significant does not mean big, and it does not mean worth
funding.**
:::

## Word 4 · The warning cuts both ways

```{r fig.height=2.5, fig.width=11.5}
p_small <- perm_null(40)
sm_bins <- seq(-9, 9, by = 0.25)
sm_h0   <- hist(p_small, breaks = sm_bins, plot = FALSE)

plot_ly(tibble(mid = sm_h0$mids, n = sm_h0$counts),
        x = ~mid, y = ~n, type = "bar",
        marker = list(color = ACCENT, line = list(width = 0)),
        showlegend = FALSE,
        hovertemplate = "A shuffled difference of %{x:.2f} points<br>in %{y} of 1,000 shuffles<extra></extra>") %>%
  layout(bargap = 0.1,
         xaxis = ax("Difference between the two groups (points)", range = c(-9, 9)),
         yaxis = ax("Number of shuffles"),
         shapes = list(list(type = "line", x0 = 2.1, x1 = 2.1,
                            y0 = 0, y1 = 1, yref = "paper",
                            line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 2.1, y = 1, yref = "paper",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "the effect we measured (2.1 points)",
                                 font = list(size = 12, color = ALERT)))) %>%
  pl_m2(height = 240, t = 34, b = 56)
```

::: {.warn}
At 40 schools the same study would have found nothing, and nothing is exactly
what it would have reported. **Absence of evidence is not evidence of
absence.**
:::

::: {.ask}
An evaluator reports "no significant effect". Before you accept it, ask two
things. How many businesses were in the study? And what is the smallest effect
the study could have detected?
:::

## Word 5 · Confidence interval

::: {.panel}
**Plain language:** a range of plausible values for the true effect, not one
precise number. A wider range means less certainty, not a bigger effect.
:::

```{r fig.height=3, fig.width=11.5}
plot_ly() %>%
  add_segments(data = filter(ci_dat, covers),
               x = ~lo, xend = ~hi, y = ~y, yend = ~y,
               line = list(color = ACCENT, width = 2.5),
               text = ~txt, hovertemplate = "%{text}<extra></extra>",
               name = sprintf("Covers the true effect (%d of 100)", ci_cov)) %>%
  add_segments(data = filter(ci_dat, !covers),
               x = ~lo, xend = ~hi, y = ~y, yend = ~y,
               line = list(color = ALERT, width = 3),
               text = ~txt, hovertemplate = "%{text}<extra></extra>",
               name = sprintf("Misses the true effect (%d of 100)", 100 - ci_cov)) %>%
  add_segments(x = mu_ci, xend = mu_ci, y = 0, yend = 101,
               line = list(color = INK, width = 2, dash = "dot"),
               name = sprintf("The true effect (%.1f points)", mu_ci),
               hoverinfo = "name") %>%
  layout(xaxis = ax("Programme effect (points), with its 95% confidence interval"),
         yaxis = ax("100 repeat studies of the same programme",
                    showticklabels = FALSE, zeroline = FALSE)) %>%
  pl_m2(height = 290)
```

::: {.ask}
Hover any bar for one study's estimate. Click a legend entry to isolate the
misses. Now the decision. All three programmes report an effect of 2.1 points,
and the bar to clear is 3 points.

**A** `[0.3, 3.9]` &nbsp; **B** `[1.9, 2.3]` &nbsp; **C** `[-0.6, 4.8]`

Which one would you act on, and what is your answer for each?
:::

::: {.panel}
B is precise and below the bar: confidently not enough. A is too uncertain to
call. C is worse still. **The interval decides, not the point estimate.** An
interval that spans the decision threshold means the honest answer is that we
do not yet know.
:::

# 4 · Put AI to work

## AI prompt cards

::: {.panel}
Each group takes one card. Paste its prompt into your AI tool, read the answer
together, and mark anything you think it gets wrong or overstates. Then compare
with the answer on the next slide.
:::

::: {.ai-output}
**Prompt:** "Explain what a 95% confidence interval means, in one paragraph, for
someone who is not a statistician."

**AI:** "A 95% confidence interval is the range in which the true value is
likely to fall. [There is a 95% probability that the true effect lies inside
this interval]{.flag}, so if the interval is [0.3, 3.9] the true effect is
probably between 0.3 and 3.9. A narrower interval means the estimate is more
precise."
:::

```{=html}
<div class="facts">
  <div class="k">Card 2</div> <div class="v">"Explain what a p-value means."</div>
  <div class="k">Card 3</div> <div class="v">"School pass rates rose 5 percentage points after our programme. Is that a good result?"</div>
  <div class="k">Card 4</div> <div class="v">"The coefficient on class size is -0.32. What should I do about it?"</div>
</div>
```

## Debrief

::: {.warn}
**Card 1.** The flagged sentence is the classic error. A 95% interval does not
carry a 95% probability for *this* interval. It means that if you ran the study
100 times, 95 of the intervals would catch the true value, as the previous
chart shows. The AI stated a probability the method does not deliver.
:::

::: {.warn}
**Card 3.** The AI almost certainly answered without asking what the pass rate
was compared to, and without asking what bar the programme has to clear. Both
are missing from the prompt, and a fluent answer hides that.
:::

::: {.panel}
**Card 4.** Look for a recommendation with no interval. If the AI tells you to
reduce class sizes without noting that the interval decides whether the
relationship is real, it has answered a different question.


Two habits, every time. Ask the AI to **ask you questions before it answers**.
Then put its answer through a second chat, or a second model, and compare.
:::

# 5 · Terms bingo

## Terms bingo

:::::: {.columns}

::::: {.column width="50%"}
::: {.excerpt}
**1.** "Waste costs in participating businesses fell 30% over the programme period."
:::

::: {.excerpt}
**2.** "The programme effect is 1,240 AED, with a 95% confidence interval of 210 to 2,270."
:::

::: {.excerpt}
**3.** "The improvement is statistically significant (p = 0.008), so the programme should be scaled nationally."
:::

::: {.excerpt}
**4.** "Costs fell 420 AED in the programme group while rising 90 AED in the comparison group."
:::
:::::

::::: {.column width="50%"}
::: {.excerpt}
**5.** "Across the sample, the average business saved 1,449 AED."
:::

::: {.excerpt}
**6.** "With 40 businesses in the study, no effect was detected."
:::

::: {.excerpt}
**7.** "The coefficient on class size is -0.32, so reducing class size raises pass rates."
:::

::: {.excerpt}
**8.** "Costs fell 420 AED against a rise of 90 AED in the comparison group. The programme must cut costs by at least 300 AED to justify scale-up."
:::
:::::

::::::

::: {.ask}
Mark every term you hear used. Then make two calls, using the three questions.

**Excerpt 3**: would you act on it? 

**Excerpt 8**: would you act on it? They do
not have the same answer.
:::

# 6 · Take this back to your desk

## The three questions

::: {.panel}
From memory, before you look: what are the three questions you can ask of any
number?
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what?</div>
  <div class="k">2</div> <div class="v">How big is it, in units that matter to the decision?</div>
  <div class="k">3</div> <div class="v">How sure are we?</div>
</div>
```

