---
title: "Is This Evidence Credible?"
subtitle: "Module 2 · Day 4, Session 1: A Quality-Assurance Deep Dive"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_dge.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"

# ---- The report and the rating scale --------------------------------------
# Shared with the printable rating sheet and the report itself, so the deck
# cannot disagree with the paper the room is holding.
source("qa_checklist.R")
source("qa_rating.R")

# Traffic-light colours for the on-screen reveal.
LIGHT_COL <- c(Green = "#2E7D4F", Amber = "#A9561F", Red = ALERT)

# Plain HTML tables (no kableExtra, whose JavaScript needs jQuery), left-aligned
# and styled by the theme.
tbl <- function(df, ...) {
  h <- as.character(knitr::kable(df, format = "html", ...))
  knitr::asis_output(sub("<table", '<table style="margin-left:0;"', h, fixed = TRUE))
}

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · Sign-off {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 5 minutes (0:00 to 0:05). Hook and first vote.
:::

## From the analyst to the report

::: {.panel}
Yesterday the analyst admitted things out loud. This is the written report that
goes to the minister.
:::

::: {.ask .fragment}
**Does it say what the analyst admitted?**
:::

## Before you open it

:::::: {.columns}

::::: {.column width="60%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">AED per business per year &middot; the authority's bar</div>
</div>
```
:::::

::::: {.column width="40%"}
```{r out.width='210px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.ask}
**Would you sign off on this report, as written, for national scale-up?**
Yes / No / Not without more answers. Vote now, before reading.
:::

::: {.panel}
Four pages, confident and well written, with real strengths. Does it support its
own claim?
:::

::: {.notes}
Paper (Oct15_session1_materials.docx): Hand out the rating sheet (pack pages 1 and 2, printed back to back, one per participant). Everyone circles a "Before reading" vote on the back as well as voting on Menti.
:::

# 1 · The report {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 10 minutes (0:05 to 0:15). Everyone reads and rates the executive summary alone.
:::

## What is in front of you

```{r}
tbl(data.frame(Section = c("1. Executive summary", "2. The programme", "3. Evaluation design",
                           "4. Data", "5. Results", "6. Limitations",
                           "7. Conclusions and recommendations", "Annexes A and B")),
    col.names = "Section", caption = "The report, section by section", align = "l")
```

::: {.ask}
Read only the executive summary for now. Do not read ahead.
:::

::: {.notes}
Paper: Hand out the report (Oct15_session1_report.docx, one per participant). Ask everyone to keep it face down until you say go.
:::

## What the report claims

::: {.excerpt}
This evaluation finds that the programme reduced waste-management costs by
**approximately 1,000 AED per business per year**, and that the estimate is
statistically significant. The finding is robust across four independent
estimation methods… The programme therefore meets the threshold for scale-up.
:::

::: {.panel .fragment}
Four claims in three sentences. Note each one, because you will rate them
separately:

- a **number**, described as approximately 1,000
- a claim of **statistical significance**
- a claim of **robustness** across four methods
- a claim that the **threshold is met**
:::

## Read the summary and rate it

::: {.ask}
On your rating sheet: **the executive summary**.

Circle a colour. Then write your reason for selecting this colour. If you rate
the summary amber or red, write what would be needed to shift the colour to green.
Then put a sticky note with your reason on the wall grid, in your colour's row.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Five minutes.** Work alone. The point is your own reading, before anyone
else's.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
If you found nothing wrong, say so. Hold your colour and we will come back to it.
:::
:::::

::::::

# 2 · The four areas, applied {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 5 minutes (0:15 to 0:20). Recap of yesterday's checklist and the rating scale; keep it brisk.
:::

## The checklist, on paper this time

```{r}
tbl(data.frame(Area = qa_domains,
               `What you are testing` = unname(qa_purpose[qa_domains]),
               check.names = FALSE),
    caption = "The four areas from yesterday", align = "ll")
```

::: {.panel}
Yesterday the analyst answered these out loud. Today the report has to answer
them. Your rating sheet names the area to check for each section.
:::

::: {.ask .fragment}
Which of the four areas can a report make look strong without actually being
strong?
:::

## The five questions, and what a report can hide

```{r}
#| results: asis
for (i in seq_along(qa_headline)) {
  cat(sprintf('<div class="panel"><b>%d.</b> %s</div>\n', i, qa_headline[i]))
}
```

::: {.warn .fragment}
None of the five asks whether the report sounds convincing.
:::

## What the rating means

```{r}
#| results: asis
for (i in seq_len(nrow(rating_scale))) {
  cat(sprintf('<div class="panel"><b style="color:%s">%s.</b> %s</div>\n',
              LIGHT_COL[[rating_scale$light[i]]], rating_scale$light[i],
              rating_scale$meaning[i]))
}
```

::: {.ask .fragment}
What single fact would move your amber to green or red?
:::

# 3 · Section by section {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 40 minutes (0:20 to 1:00). 20 minutes of reading in pairs, then 20 minutes of debrief, about 3 minutes per slide. In the debrief, read the section's column on the wall grid before each reveal (no show of colours needed). Grid setup is in the facilitator key: six section columns, rows Green, Amber, Red.
:::

## Your turn: rate the rest

::: {.ask}
**20 minutes in pairs:** rate Sections 3 to 7 from the report itself, on the
rating sheet. Do not wait for the slides.
:::

::: {.panel}
For each section: a colour, your reason for selecting this colour, and, if amber
or red, what would be needed to shift the colour to green. The Area to check
column tells you which part of yesterday's QA checklist to use. **One sticky note
per pair per section on the wall grid.**
:::

::: {.notes}
Walk the room. At 15 minutes, give a five-minute warning. Pairs who finish early compare their executive-summary colour with their Results colour.
:::

## Evaluation design

::: {.excerpt}
The parallel trends assumption underlying the difference-in-differences design is
the standard assumption in this literature and is satisfied by construction in
programmes of this type, since treatment and control neighbourhoods were selected
from the same administrative area and are subject to the same economic conditions.
:::

::: {.ask}
Read the Section 3 column on the wall. What does parallel trends actually require?
:::

::: {.warn .fragment}
That the two groups **would have moved in parallel without the programme**.
Sharing an area makes a common shock likely. It does not show the groups were on
the same path, and no pre-trend test is reported.

The groups are also the wrong ones: the DiD compares enrolled with non-enrolled
businesses inside treatment neighbourhoods.
:::

## Data

::: {.excerpt}
Recycling compliance… is missing for 8,570 of the 19,826 observations. These gaps
reflect the inspection schedule…

…the survey instrument separated expenditure on collection from expenditure on
treatment… This change improves the precision of the measure…
:::

::: {.ask}
Read the Section 4 column on the wall. Two disclosures: which is handled well?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**Handled well:** the instrument change is *disclosed*. Many reports would not
mention it.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**Not handled:** the change is never tested for its effect on the outcome, and
the report never says which businesses lack compliance data.
:::
:::::

::::::

## Results

::: {.excerpt}
All four estimates are negative and statistically significant… The largest
estimate, from the randomised comparison, indicates a cost reduction of
1,014 AED per business per year. The randomised comparison is our headline
result.
:::

::: {.ask}
Read the Section 5 column on the wall. Four estimates are given. Why is this one the
headline?
:::

::: {.warn .fragment}
The report's reason: the randomised comparison covers all eligible businesses,
the population that would be scaled up. That reason is defensible. It is also
the largest estimate, and the only one above 1,000.
:::

## The table you should look at twice

```{r}
tbl(data.frame(
  Method = c("Difference-in-differences", "Regression discontinuity",
             "Randomised comparison", "Matched comparison"),
  Estimate = c("-816", "-905", "-1,014", "-1,000"),
  CI = c("[-879, -753]", "[-992, -818]", "[-1,093, -935]", "[-1,072, -927]"),
  Point = c("No", "No", "Yes", "Yes, exactly"),
  Whole = c("No", "No", "No", "No")),
  col.names = c("Method", "Estimate (AED)", "95% CI",
                "Point estimate reaches 1,000?", "Whole interval reaches 1,000?"),
  caption = "Table 1 of the report, with two columns added", align = "lrrrr")
```

::: {.warn .fragment}
At the point estimate, two of the four reach 1,000. At the cautious end of the interval, none does. The preferred specification, named in Section 3, is 816.
:::

::: {.notes}
Before the click, ask the room to read each row with its interval. Point out that 1,014 and 1,000 reach the bar only at the point estimate; the RCT interval runs down to 935. Then reveal. Ask: which rule should the authority use, the point estimate or the whole interval?
:::

## Limitations

::: {.excerpt}
We note three limitations. First, the randomised comparison estimates the effect
of being offered the programme, not of enrolling… Second, the regression
discontinuity estimate is local to the eligibility cut-off… Third,
self-reported expenditure may be subject to recall error…
:::

::: {.ask}
Read the Section 6 column on the wall. Are these the right three?
:::

::: {.warn .fragment}
All three are correct, and none of them threatens the conclusion.

The problems raised in Section 4 (the change in the survey instrument and the
unexplained gap in the compliance data) are never mentioned again.
:::

## Conclusions

::: {.excerpt}
**Recommendation 2.** The subsidy rate should be maintained at its current level,
since the evaluation provides no evidence that a lower rate would be less
effective.

**Recommendation 3.** The eligibility cut-off should be reviewed, as the
regression discontinuity results suggest that businesses just above the threshold
may also benefit.
:::

::: {.ask}
Read the Section 7 column on the wall. What was estimated about subsidy rates, and about
businesses above the cut-off?
:::

::: {.warn .fragment}
Nothing on either. Recommendation 2 turns an absence of evidence into a
recommendation to do nothing. Recommendation 3 rests on a result about businesses
*near* the cut-off, and even there the estimate was 905, below the threshold.
:::

# 4 · The verdict {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 10 minutes (1:00 to 1:10). Second vote, the decisive problem, the strengths.
:::

## The same question

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">The bar, and the one estimate the summary chose to echo</div>
</div>
```

Would you sign off on this report, as written? Vote again.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.ask .fragment}
If your vote moved, name the section that moved it.
:::

::: {.notes}
Paper (Oct15_session1_materials.docx): Participants circle their "At the end" vote and answer the three questions on the back of the rating sheet (pack page 2).

Wall grid: point to the executive-summary column and the Results column side by side. If the summary leans green and Results leans amber or red, ask what the summary led people to believe.
:::

## The most important problem

```{r}
#| results: asis
cat(sprintf('<div class="warn">%s</div>\n', rating_decisive))
```

::: {.ask .fragment}
Which sections did you rate green, and why do they still hold?
:::

## The strengths are real

::: {.panel}
Before we close, the report's genuine merits:

- It **states the threshold up front**, so the reader knows the standard being applied.
- It **reports four methods** with their different populations named, rather than one number.
- It **discloses** the survey instrument change, which most reports would bury.
- It gives **confidence intervals for every estimate**, and an annex of robustness specifications.
- Its **limitations are technically correct**, even though they are the wrong ones.
:::

::: {.ask .fragment}
Does any of that rescue the recommendation?
:::

# 5 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 15 minutes (1:10 to 1:25). Groups run the prompt, compare with the example, then see the better prompt.
:::

## The prompt

::: {.excerpt}
Here is an evaluation report on a programme that subsidised waste-management
technology for businesses. It found a cost reduction of approximately 1,000 AED
per business per year and recommends national scale-up. Assess the credibility of
this evidence and tell me whether the recommendation is sound.
:::

::: {.panel}
You have just done this by hand, section by section. Run the prompt on the
report yourselves, then compare the AI's verdict, and the example on the next
slides, with yours. The report is on the course website: download
Oct15_session1_report.docx from the landing page.
:::

::: {.notes}
Paper (Oct15_session1_materials.docx): Hand out the AI Snapshot (pack page 3, one per group). Groups paste the downloaded report into their AI tool with the prompt.
:::

## What came back

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**Assessment: the evidence is credible and the recommendation is sound.**

The report is unusually thorough for this kind of evaluation. Several points
support its credibility:

**Multiple methods.** The use of four estimation approaches, all producing
consistent results, is a strong indicator of robustness. When independent methods
agree, the finding is unlikely to be an artefact of any single specification.

**Large sample.** With roughly 9,900 observations, the study is well powered and
the estimates are precise.
:::

## What came back, continued

::: {.ai-output}
**Transparency.** The report discloses a change in the survey instrument between
rounds and notes the gaps in the recycling compliance data. This level
of disclosure suggests a careful and honest evaluation team.

**Appropriate limitations.** The limitations section correctly identifies that the
randomised comparison estimates the offer effect and that the RDD estimate is local to the cut-off.

**Recommendation.** The headline estimate of 1,014 AED exceeds the 1,000 AED
threshold, so [the programme meets the criterion for scale-up]{.fragment .highlight-red fragment-index=2}. The
recommendation is [supported by the evidence]{.fragment .highlight-red fragment-index=2}.
:::

::: {.ask .fragment fragment-index=1}
Underline the two claims in the AI's response that the report does not support.
Then name the bigger problem that the response never mentions. What did your own
AI get right, and what did your AI miss?
:::

## What AI got, and what it could not get

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- Found the real disclosures and called them transparency
- Correctly described the offer-effect and locality limitations
- Reached a verdict, which is what was asked
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it missed**

- It read "approximately 1,000" as if all four methods agreed. Only the matched comparison gives 1,000.
- It never noticed that the preferred specification, 816, fails the threshold.
- It treated the compliance gap as evidence of honesty.
- It did not ask why 1,014 is the headline.
:::
:::::

::::::

::: {.ask .fragment}
Which of the four would slip past you in practice?
:::

## The same task, a better prompt

::: {.excerpt}
The same report claims a cost reduction of approximately 1,000 AED against a
decision threshold of 1,000 AED. Do three things.

First, list every estimate the report gives, with its confidence interval, and say
for each one whether it clears the threshold.

Second, identify any claim in the executive summary that is not supported by the
results section, and quote both.

Third, tell me what the report does not tell me that I would need in order to
decide.
:::

::: {.ask .fragment}
What do the three instructions have in common?
:::

::: {.panel .fragment}
Each asks for a comparison: estimates against the threshold, the summary against
the results, the report against what a decision needs.
:::

## Two checks on any AI answer

::: {.panel}
**1.** Ask the AI to ask you questions before it answers. It then has to find out
what it does not know, such as which estimate is the preferred specification.
:::

::: {.panel}
**2.** Paste its answer into a fresh chat and ask what is wrong with it. A new chat
has not committed to the first verdict.
:::

# 6 · Take this back to your desk {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Timing: 5 minutes (1:25 to 1:30). Take-away card and the "On my desk" minute.
:::

## Three habits for reading a report

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Tabulate the estimates yourself, with the threshold in its own column. Do not let the report choose which number you see first.</div>\n')
cat('<div class="panel"><b>2.</b> Read the limitations section, then check whether the limitations section names the problems in the data section. If those problems are missing, ask why they were left out.</div>\n')
cat('<div class="panel"><b>3.</b> Find the preferred specification (the estimate the evaluators trust most). Check whether the headline number is that estimate. When the two differ, ask why.</div>\n')
```

::: {.ask .fragment}
Which of the three would have caught the most important problem in this report?
:::

::: {.notes}
Paper (Oct15_session1_materials.docx): Hand out the take-away cards (pack page 4, cut). Give a minute to fill in "Back at work" on the card, then take two or three examples from the room.
:::

## Next session

::: {.panel}
This week you have practised reading other people's evidence critically. In the
next session you write your own: a policy brief, from findings to
recommendation.
:::

::: {.panel .fragment}
Bring the ratings you gave today. You will apply the same judgements to your own
draft.
:::
