---
title: "What Do the Numbers Say? Reading an Evaluation Output"
subtitle: "Module 2 · Day 1, Session 2: Descriptive Analysis, Live"
author: "3ie"
always_allow_html: yes
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR, the same set-up as Oct12_session3_live.qmd. The live
# cells need an internet connection on the day (webR loads from a CDN), and the
# CSV listed under webr: resources must sit next to the published .html.
# Static knitr chunks (the CI chart, the Menti QR, the printed table) work as
# before. revealjs only: pandoc cannot carry live cells or htmlwidgets into pptx.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  render-df: kable
  resources:
    - evaluation_data_TrafficCameras.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate: knitr sizes htmlwidgets as fig.width * dpi.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(estimatr)   # lm_robust, errors clustered by road segment
library(qrcode)     # live poll QR, generated below so it cannot drift

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

# ---- Tables: plain knitr::kable, styled by theme_dge.scss -------------------
# No kableExtra: its JavaScript needs jQuery, which live-revealjs does not load.
# font_size in px; hl = rows to shade (hl_bold to bold them); note = a line
# under the table. The table is centred, as kable_styling(full_width = FALSE) did.
HL_BLUE <- "#E4EBF3"; HL_RED <- "#F8E9EA"
tbl <- function(df, ..., font_size = NULL, hl = NULL, hl_bg = HL_BLUE,
                hl_bold = FALSE, note = NULL) {
  h <- as.character(knitr::kable(df, format = "html", ...))
  fs <- if (is.null(font_size)) "" else sprintf(" font-size:%dpx;", font_size)
  h <- sub("<table", sprintf('<table style="margin-left:auto; margin-right:auto;%s"', fs),
           h, fixed = TRUE)
  if (length(hl)) {
    parts <- strsplit(h, "<tbody>", fixed = TRUE)[[1]]
    rows  <- strsplit(parts[2], "<tr>", fixed = TRUE)[[1]]
    st <- sprintf("background:%s;%s", hl_bg, if (hl_bold) " font-weight:700;" else "")
    for (i in hl) {
      rows[i + 1] <- gsub('<td style="', paste0('<td style="', st, " "), rows[i + 1], fixed = TRUE)
      rows[i + 1] <- gsub("<td>", paste0('<td style="', st, '">'), rows[i + 1], fixed = TRUE)
    }
    h <- paste0(parts[1], "<tbody>", paste(rows, collapse = "<tr>"))
  }
  if (!is.null(note))
    h <- paste0(h, sprintf('\n<p style="text-align:center; font-size:0.75em; color:%s; margin-top:0.3em;">%s</p>',
                           MUTED, note))
  fence <- strrep("`", 3)
  knitr::asis_output(paste0("\n", fence, "{=html}\n", h, "\n", fence, "\n"))
}

PL_W <- 1100; PL_WC <- 520

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54) {
  p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    ) %>%
    config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

# Emit an HTML snippet as a raw block. Plain cat() leaves it to pandoc, which
# reads indented lines inside a multi-line <div> as a code block.
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. The QR and printed link on the vote slide
# and the results iframe on the look-back slide all read from here.
# MENTI_CODE is the short JOIN code participants type at menti.com, NOT the
# long ID in the editor URL.
MENTI_CODE  <- "1234 5678"   # <!-- PLACEHOLDER: this session's join code -->
MENTI_JOIN  <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
# The results embed: the SLIDES view for a webpage. Long ID from the editor URL.
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/PLACEHOLDER/embed"  # <!-- PLACEHOLDER -->

# Distinct filename so this QR never overwrites Session 1's menti_qr.png.
menti_qr <- function(path = "menti_qr_s2.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- The traffic camera data ------------------------------------------------
# evaluation_data_TrafficCameras.csv is FICTIONAL data generated by
# make_traffic_camera_data.R. It is not an Abu Dhabi Police evaluation and not
# any real programme, so the case names no real body. Every slide with a number
# from it sits under the invented-figures disclaimer on the case slide.
#
# This session is DESCRIPTIVE: it reads the data, graphs it and measures the
# before-and-after change in the camera sectors. It is not a DiD, and this case
# is not reused in the method sessions (those use GreenWaste). The comparison
# sectors appear once, at the end, to raise "compared to what?".
#
# The live cells recompute every number below in the browser. They are computed
# here too so the static slides (rule chart, printed table) cannot drift.
pol  <- read.csv("./evaluation_data_TrafficCameras.csv")
cams <- pol %>% filter(treated == 1)

POL_RULE <- 2.0     # injury collisions avoided, per segment per round

n_rows <- nrow(pol)
n_seg  <- n_distinct(pol$segment_id)

cam_before <- mean(cams$injury_collisions[cams$round == 0])
cam_after  <- mean(cams$injury_collisions[cams$round == 1])
cam_fall   <- cam_before - cam_after
cam_pct    <- 100 * cam_fall / cam_before

oth_before <- mean(pol$injury_collisions[pol$treated == 0 & pol$round == 0])
oth_after  <- mean(pol$injury_collisions[pol$treated == 0 & pol$round == 1])
oth_fall   <- oth_before - oth_after
oth_pct    <- 100 * oth_fall / oth_before

# Share of camera segments whose count fell.
seg_chg <- cams %>%
  select(segment_id, round, injury_collisions) %>%
  pivot_wider(names_from = round, values_from = injury_collisions,
              names_prefix = "r") %>%
  mutate(change = r1 - r0)
share_fell <- 100 * mean(seg_chg$change < 0)

# The before-and-after regression, errors clustered by road segment because
# each segment appears twice.
ba_fit <- lm_robust(injury_collisions ~ round, data = cams, clusters = segment_id)
ba_ci  <- sort(abs(confint(ba_fit)["round", ]))   # [1] = end closest to zero

# Estimate, interval and rule on one axis.
rule_plot <- function(height = 200, b = 46) {
  est <- cam_fall; lo <- ba_ci[1]; hi <- ba_ci[2]
  plot_ly() %>%
    add_segments(x = lo, xend = hi, y = 1, yend = 1, showlegend = FALSE,
                 line = list(color = ACCENT, width = 6), hoverinfo = "skip") %>%
    add_markers(x = c(lo, hi), y = c(1, 1), showlegend = FALSE, hoverinfo = "skip",
                marker = list(color = ACCENT, size = 18, symbol = "line-ns-open",
                              line = list(color = ACCENT, width = 3))) %>%
    add_markers(x = est, y = 1, showlegend = FALSE,
                marker = list(color = ACCENT, size = 16),
                hovertemplate = "estimate %{x:.2f} fewer<extra></extra>") %>%
    layout(
      shapes = list(list(type = "line", x0 = POL_RULE, x1 = POL_RULE,
                         y0 = 0.4, y1 = 1.6, xref = "x", yref = "y",
                         line = list(color = ALERT, width = 3, dash = "dash"))),
      annotations = list(
        list(x = POL_RULE, y = 1.72, xref = "x", yref = "y", showarrow = FALSE,
             text = sprintf("Decision rule: %.1f", POL_RULE),
             font = list(color = ALERT, size = 14)),
        list(x = est, y = 0.62, xref = "x", yref = "y", showarrow = FALSE,
             text = sprintf("estimate %.2f", est),
             font = list(color = INK, size = 13))),
      xaxis = ax("Fewer injury collisions per segment, 2019 to 2021",
                 range = c(0, 3.6)),
      yaxis = list(range = c(0.4, 1.9), showticklabels = FALSE,
                   showgrid = FALSE, zeroline = FALSE)) %>%
    pl_m2(height = height, b = b)
}
```

# 1 · The case {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 8 minutes.
:::

## The traffic camera programme

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the live cells
# below use is defined here, so each cell on a slide stays a few lines long.
pol  <- read.csv("evaluation_data_TrafficCameras.csv")
library(estimatr)
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Layout only: margins, font size and muted axes for a projector.
clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)

# Two lines, before and after: sectors with cameras and sectors without.
plot_two_groups <- function(means) {
  clean_par()
  plot(0:1, means[, "1"], type = "b", pch = 19, lwd = 4, col = ACCENT, bty = "n",
       ylim = c(0, 13), xlim = c(-0.1, 1.7), xaxt = "n",
       xlab = "", ylab = "Collisions per segment")
  lines(0:1, means[, "0"], type = "b", pch = 19, lwd = 4, col = ACCENT2)
  axis(1, at = 0:1, labels = rownames(means))
  text(1.06, means[2, "1"], "With cameras", col = ACCENT, adj = 0)
  text(1.06, means[2, "0"], "Without cameras", col = ACCENT2, adj = 0)
}
```

:::::: {.columns}

::::: {.column width="52%"}
In 2020 a roads authority began installing automated speed cameras. It started
with twelve sectors and planned the rest for two years later.

Before extending the programme, it set a bar: cameras had to cut injury
collisions by at least **`r sprintf("%.1f", POL_RULE)` per road segment per
round**. Below that, the money goes to other road safety work.
:::::

::::: {.column width="48%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">Automated speed enforcement</div>
  <div class="k">Rolled out</div>  <div class="v">12 sectors in 2020, 12 in 2022</div>
  <div class="k">Outcome</div>     <div class="v">Injury collisions</div>
  <div class="k">Decision rule</div> <div class="v">2.0 fewer per segment per round</div>
</div>
```
:::::

::: {.warn .note-ctr .fragment}
**Illustrative case, with invented figures.** This is not a real programme and
the numbers are generated for this training. Every number in this session comes
from this fictional dataset.
:::

::::::

::: {.notes}
The live cells start loading in the background while you talk (about 10 s).
:::

## Today we answer two of Session 1's questions

```{=html}
<div class="facts">
  <div class="k">How big?</div>          <div class="v">In the units the decision is written in: collisions per segment</div>
  <div class="k">How sure?</div>         <div class="v">The p-value and the confidence interval</div>
  <div class="k">Compared to what?</div> <div class="v">We come back to this at the end</div>
</div>
```

::: {.panel}
We start from the raw data file, as an evaluator would, and build the numbers on
screen: what is in the file, what it looks like on a chart, and how much
collisions changed where cameras went.
:::

::: {.warn}
You will not write any code. You call the numbers; I type and press **Run**. The
grey boxes are real R, running in this page.
:::

# 2 · Meet the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 15 minutes.
:::

## One row, one road segment, one year

```{webr}
cols <- c("segment_id", "sector_id", "corridor_type", "year",
          "treated", "injury_collisions")
head(pol[, cols], 4)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**treated** is 1 if the segment's sector got cameras in 2020.
**injury_collisions** is the number of collisions with an injury on that segment
in that year.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
The file has `r format(n_rows, big.mark = ",")` rows. How many **road
segments**? Say a number before I show the next slide.
:::

::: {.panel .fragment}
`r format(n_seg, big.mark = ",")`: each segment appears twice, once in 2019 and
once in 2021.
:::
:::::

::::::

::: {.notes}
`head()` shows the first rows of a data file. Read one row aloud in plain words:
"segment 1 in sector 1, a suburban arterial, in 2019, in a camera sector, had 10
injury collisions."
:::

## Who is in the data?

```{webr}
base <- subset(pol, year == 2019)
data.frame(
  segments = tapply(base$segment_id, base$treated, length),
  sectors  = tapply(base$sector_id, base$treated, function(x) length(unique(x))))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Row **1** is the sectors that got cameras in 2020; row **0** is the sectors still
waiting. Half the road network in each.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Before we look at any collisions: what would you want to know about **how** the
first twelve sectors were chosen?
:::
:::::

::::::

::: {.notes}
Take two or three answers. The design document says cameras went first to the
sectors with the worst safety record. Note it and move on: it matters for the
"compared to what?" question at the end, and Session 3 takes it further.
:::

## Word 1 · The average

```{webr}
summary(pol$injury_collisions)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
`summary()` gives six numbers for one column: the smallest, the largest, the
middle road (**median**) and the **mean**.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
The mean is about 9.5 collisions. Is a typical road segment close to 9.5?
What do the minimum and maximum suggest?
:::
:::::

::::::

## Word 1 · What the average hides

```{webr}
#| fig-width: 10
#| fig-height: 3.1
before <- subset(pol, year == 2019)
clean_par()
hist(before$injury_collisions, breaks = 40, col = ACCENT, border = "white",
     main = "", xlab = "Injury collisions per segment, 2019")
abline(v = mean(before$injury_collisions), col = ALERT, lwd = 3, lty = 2)
```

::: {.warn .fragment}
About seven in ten segments had between 5 and 15 collisions in 2019, and nine
had more than 30. Those few pull the mean (red line) above the median. Session 1 showed the
same thing with ten camera roads: one busy ring road pulled the mean to 13.5 while the
median stayed at 10.
:::

<!-- Lucas, 2026-09-25: this line used to refer to "Programmes A and B", an example
     from the old Session 1. Oct12_session1_live.qmd now makes the mean-vs-median
     point with ten camera roads (mean 13.5, median 10), so the reference was
     updated to match. -->

# 3 · Graph it {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 12 minutes.
:::

## Before and after, where cameras went

```{webr}
#| fig-width: 7
#| fig-height: 3.3
cams  <- subset(pol, treated == 1)
means <- tapply(cams$injury_collisions, cams$year, mean)
clean_par()
b <- barplot(means, col = ACCENT, border = NA, ylim = c(0, 14),
             ylab = "Collisions per segment")
text(b, means + 0.8, round(means, 1), cex = 1.3)
```

::: {.ask}
This is the chart a press release would use. What does it show, and what does
it not show?
:::

## Did every road fall?

```{webr}
#| fig-width: 10
#| fig-height: 2.8
wide <- reshape(cams[, c("segment_id", "year", "injury_collisions")],
                idvar = "segment_id", timevar = "year", direction = "wide")
change <- wide$injury_collisions.2021 - wide$injury_collisions.2019
clean_par()
hist(change, breaks = 30, col = ACCENT, border = "white", main = "",
     xlab = "Change in collisions per segment, 2019 to 2021")
abline(v = 0, col = ALERT, lwd = 3, lty = 2)
```

::: {.panel .fragment}
The average fell, but not on every road. About
`r sprintf("%.0f", share_fell)`% of camera segments had fewer collisions in 2021;
the rest stayed the same or went up. This is Word 1 again: an average change
hides how much roads differ.
:::

## Would you fund the extension?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
Collisions fell from `r sprintf("%.1f", cam_before)` to
`r sprintf("%.1f", cam_after)` per segment where cameras went.
**Would you commit the money to extend them?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Not on this alone
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.notes}
Leave the vote open. We look at the results again at the end of the session.
:::

# 4 · How big, and how sure? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 20 minutes.
:::

## How big? Work it out from the averages

```{webr}
before <- mean(cams$injury_collisions[cams$year == 2019])
after  <- mean(cams$injury_collisions[cams$year == 2021])
round(data.frame(before, after, change = after - before,
                 percent = 100 * (after - before) / before), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
The rule asks for **`r sprintf("%.1f", POL_RULE)` fewer collisions per
segment**. Which of these numbers do you check against it?
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
The change, `r sprintf("%.2f", -cam_fall)`, which is in the same units as the
rule. It clears 2.0. The `r sprintf("%.0f", cam_pct)`% is in different units
and cannot be checked against the rule.
:::
:::::

::::::

## Word 3 · The regression gives the same number

```{webr}
fit <- lm(injury_collisions ~ round, data = cams)
coef(fit)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**round** is 0 for 2019 and 1 for 2021. The table an evaluator hands you comes
from a line like this one.
:::

::: {.ask}
Which number is the 2019 average, and which is the change?
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**(Intercept)** is the 2019 average, `r sprintf("%.2f", cam_before)`.
**round** is the change, `r sprintf("%.2f", -cam_fall)`: the same number we
got by subtracting two averages. Session 1's Word 3: a coefficient is a gap
between two groups. Here the two groups are the same roads in two years.
:::
:::::

::::::

## Words 4 and 5 · How sure?

```{webr}
fit <- lm_robust(injury_collisions ~ round, data = cams, clusters = segment_id)
round(summary(fit)$coefficients[, c(1, 2, 4, 5, 6)], 3)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Pr(>|t|)** is the p-value. Printed as 0, it means smaller than 0.001: a fall
this size would be very unlikely if collisions had not really changed.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**CI Lower** and **CI Upper** are the 95% confidence interval: the fall is
somewhere between `r sprintf("%.2f", ba_ci[1])` and
`r sprintf("%.2f", ba_ci[2])` collisions per segment.
:::
:::::

::::::

::: {.notes}
`lm_robust` with `clusters = segment_id` counts each road once rather than
twice, because the same segment appears in 2019 and 2021. Session 3 returns to
why this matters. Vocabulary recap from Session 1: the **standard error** is how
much the estimate would move from sample to sample; the **p-value** is how
surprising the result would be if nothing had changed; the **confidence
interval** is the range of plausible values.
:::

## Does it clear the rule?

```{r fig.height=2.2, fig.width=11.5}
rule_plot(height = 200)
```

:::::: {.columns}

::::: {.column width="50%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Decision rule</div>           <div class="v">%.2f fewer</div>
  <div class="k">The change</div>              <div class="v">%.2f fewer</div>
  <div class="k">95%% confidence interval</div> <div class="v">%.2f to %.2f</div>
</div>', POL_RULE, cam_fall, ba_ci[1], ba_ci[2]))
```
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
The estimate clears the rule, and so does the whole interval, just. On
**how big** and **how sure**, the programme passes.
:::

::: {.ask .fragment}
Which of Session 1's three questions have we not asked yet?
:::
:::::

::::::

# 5 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 15 minutes.
:::

## AI Snapshot: ask AI to read the chart

:::::: {.columns}

::::: {.column width="55%"}
::: {.panel}
Give a picture of the before-and-after bar chart to an AI tool with the prompt
most people would type: **"What does this chart show? Did the cameras work?"**
:::
:::::

::::: {.column width="45%"}
::: {.ask}
In pairs: mark every claim you can check against the chart. Which ones can the
chart support?
:::
:::::

::::::

::: {.ai-output}
*Illustrative response. A response like this is possible, especially with a short or leading prompt:*

"The chart shows a 22% reduction in injury collisions after speed cameras were
installed, from 11.5 to 9.0 per road segment. This reduction is statistically
significant (p < 0.001), confirming that the cameras caused the decline. The
fall was consistent across all road types, which suggests the effect is robust.
Given the size of the effect, the programme should be extended to the remaining
sectors."
:::

## Debrief: what the AI added

```{=html}
<div class="facts">
  <div class="k">"22% reduction, 11.5 to 9.0"</div>
  <div class="v">Correct. This is what the chart shows.</div>
  <div class="k">"p &lt; 0.001"</div>
  <div class="v">Not on the chart. The model supplied a p-value it was never given.</div>
  <div class="k">"confirming that the cameras caused"</div>
  <div class="v">A before-and-after chart cannot show cause. A small p-value says the fall is unlikely to be noise, not what caused it.</div>
  <div class="k">"consistent across all road types"</div>
  <div class="v">Invented. The chart has no road types, and more than a third of camera roads did not fall.</div>
  <div class="k">"should be extended"</div>
  <div class="v">A recommendation with no rule and no comparison.</div>
</div>
```

::: {.warn .fragment}
Only the first sentence comes from the chart. Everything after it is the model filling gaps with what often follows a chart
like this. Check whether yours did the same.
:::

## A better prompt

::: {.panel}
**"This chart shows average injury collisions per road segment in the sectors
that got speed cameras, in 2019 and 2021. Describe only what the chart shows.
Then list what else could explain the change, and what comparison you would need
before saying the cameras caused it. Before you answer, ask me any questions you
need."**
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Run both prompts on the same chart. What does the second answer include that the first left out? What did either
explain especially well?
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Two habits for every session: ask the AI to **ask you questions first**, and have
a second chat review what the first one wrote.
:::
:::::

::::::

::: {.notes}
~7 min for groups to run both prompts, ~3 min to take answers. Listen for: the
second answer asks what happened on roads without cameras, and whether traffic
changed.
:::

# 6 · Compared to what? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 10 minutes.
:::

## What else happened between 2019 and 2021?

::: {.panel}
So far we have compared the camera roads with **their own past**. The fall of
`r sprintf("%.2f", cam_fall)` includes anything else that changed over those two
years.
:::

::: {.ask}
In your table, list two or three things apart from cameras that could have cut
collisions in these sectors. Then say what you would compare the camera roads
with to rule them out.
:::

::: {.notes}
Likely answers: a national safety campaign, new vehicles, fewer trips during
2020, road works, police patrols. For comparisons, steer the room towards "roads
that did not get cameras over the same years".
:::

## One comparison: roads without cameras

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 4
means <- tapply(pol$injury_collisions,
                list(pol$year, pol$treated), mean)
plot_two_groups(means)
```
:::::

::::: {.column width="44%"}
::: {.ask}
The twelve sectors still waiting for cameras. Which way did their collisions go?
:::

::: {.warn .fragment}
Down too, by `r sprintf("%.2f", oth_fall)` per segment
(`r sprintf("%.0f", oth_pct)`%), with no cameras installed. Some of the
`r sprintf("%.2f", cam_fall)` would have happened anyway.
:::

::: {.panel .fragment}
How much, and whether these roads are a fair comparison, is the subject of
Session 3.
:::
:::::

::::::

## Look back at your vote

```{r results='asis'}
# Shows the live Menti results once this session's Menti exists. Until
# MENTI_EMBED is set, a marked placeholder box renders instead.
if (!grepl("PLACEHOLDER", MENTI_EMBED)) {
  html_out(sprintf('
<iframe data-external="1" allowfullscreen="true" allowtransparency="true"
        frameborder="0" src="%s"
        style="width:100%%; height:66%%; border:none;"></iframe>', MENTI_EMBED))
} else {
  html_out('<div class="panel note-ctr"><strong>[PLACEHOLDER] Live Menti results.</strong> Set <code>MENTI_EMBED</code> in the setup chunk to show the room vote here.</div>')
}
```

::: {.warn}
You voted on a before-and-after chart. It answers "how big?" and "how sure?"
well, and it still clears the rule. It cannot answer "compared to what?", and
roads without cameras also improved.
:::

# 7 · Take this back to your desk {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
Section timing: about 10 minutes.
:::

## Annotate this output

:::::: {.columns}

::::: {.column width="58%"}
```{r}
tibble(
  ` `       = c("(Intercept)", "round"),
  Estimate  = coef(ba_fit),
  `Std. Error` = ba_fit$std.error,
  `Pr(>|t|)`   = ba_fit$p.value,
  `CI Lower`   = ba_fit$conf.low,
  `CI Upper`   = ba_fit$conf.high
) %>%
  tbl(align = "lrrrrr", digits = 3,
        col.names = c("", "Estimate", "Std. Error", "p-value",
                      "CI Lower", "CI Upper"), font_size = 16)
```

::: {.panel}
Outcome: injury collisions per road segment, camera sectors only, 2019 (round 0)
and 2021 (round 1). Standard errors clustered by road segment. The p-value is
the column R printed as Pr(>|t|).
:::
:::::

::::: {.column width="42%"}
::: {.ask}
On the printed copy in front of you:

**1.** Circle the before-and-after change.

**2.** Draw a box around its confidence interval.

**3.** Underline the 2019 average, and say in one line why it is not an effect.

**4.** Write the comparison you would ask the evaluator for.
:::

::: {.notes}
Four minutes. Question 4 carries into Session 3.
:::
:::::

::::::

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">How big?</div>          <div class="v">In the units the decision is written in, not a percentage</div>
  <div class="k">How sure?</div>         <div class="v">Does the confidence interval clear the rule, not just zero?</div>
  <div class="k">Compared to what?</div> <div class="v">Which group, which period, and did it move too?</div>
</div>
```

::: {.panel}
The camera result passes the first two questions. The third is still open: the
roads without cameras improved as well, and Session 3 looks at how to account
for that.
:::
