---
title: "Read the extra change"
subtitle: "Module 2 | Day 2 | Session 1"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day2_case.R')
```

## A question from yesterday {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
city <- subset(gw, setting == 'city')
change <- city$cost_after - city$cost_before
joined <- city$took_part == 1
```
::: {.ask}
What comparison would help us read a fall in costs?
:::
Tell a partner. Today we compare **changes in two groups**.

::: {.notes}
3 minutes. Retrieve yesterday's question. No AI or definitions first. GreenWaste is fictional; city participants score 58 or below. Warm up webR before class. No participant coding.
:::

## Two groups over time {.reader-slide}

```{r, results='asis'}
html_out(sprintf('<div class="evidence-pair"><div class="evidence"><h3>Took part</h3><p class="big-number">%s &rarr; %s</p></div><div class="evidence"><h3>Did not take part</h3><p class="big-number">%s &rarr; %s</p></div></div><p class="caption">Mean annual waste cost, AED. Before &rarr; 12 months later.</p>%s',fmt(means[1]),fmt(means[2]),fmt(means[3]),fmt(means[4]),fiction))
```
::: {.ask}
Which change should we subtract from which?
:::


::: {.notes}
4 minutes. Take predictions before showing the calculation. Different starting levels alone do not rule out DiD; different untreated trends do.
:::

## Complete the four numbers {.reader-slide}

Use **Sheet 1**. Work alone, then compare with your group.

1. Change among participants: after minus before.
2. Change in the comparison group: after minus before.
3. Extra change: first answer minus second answer.

```{=html}
<button data-open="did-arithmetic" data-closed="Open the calculation" data-opened="Hide feedback" aria-controls="did-arithmetic" aria-expanded="false">Open the calculation</button><div id="did-arithmetic" class="feedback" hidden>Extra change = participants' change minus comparison change. Negative means lower costs.</div>
```


::: {.notes}
8 minutes. Fiona's four-number task remains. Collect signs and units before feedback. Printed means are rounded; a one-AED discrepancy is acceptable. Use actual values for the final estimate.
:::

## Read the result row {.reader-slide}

```{r, results='asis'}
html_out(paste0(as.character(knitr::kable(report_row('did'),format='html',row.names=FALSE,table.attr='class="reading-table"')),fiction))
```
::: {.ask}
Does this show an annual saving of at least `r fmt(decision_rule)` AED?
:::
```{webr}
#| echo: true
#| autorun: false
round(mean(change[joined]) - mean(change[!joined]))
```


::: {.notes}
5 minutes. One trainer calculation, prediction before Run. Output is the same coefficient as the regression in the printed extract. 95% interval uses business-clustered robust standard errors. The estimate and interval are short of the rule. This does not yet settle causal credibility.
:::

## What has to be true? {.reader-slide}

<div class="lead">Without GreenWaste, the two groups would have changed in similar ways.</div>

Reports call this **parallel trends**.

::: {.ask}
What could make their changes differ even without GreenWaste?
:::
Suggest one concrete reason.

::: {.notes}
6 minutes. Examples: different price shocks, business mix changing, another policy affecting one group. The assumption concerns unobserved change without the programme; it does not require identical levels.
:::

## Would you trust these pictures? {.reader-slide}

```{r, results='asis'}
html_out(paste0('<div class="trend-options"><div><h3>A</h3>',trend_svg(1),'</div><div><h3>B</h3>',trend_svg(2),'</div></div><p class="caption">Hypothetical pictures before a programme. These are not extra GreenWaste data.</p>'))
```
::: {.ask}
Which picture raises more concern? What if the comparison group changes?
:::
```{=html}
<button data-open="trend-feedback" data-closed="Open feedback" data-opened="Hide feedback" aria-controls="trend-feedback" aria-expanded="false">Open feedback</button><div id="trend-feedback" class="feedback" hidden>B was already changing differently. A is reassuring evidence, though it cannot prove future parallel trends. A changing mix of businesses can also mislead.</div>
```


::: {.notes}
8 minutes. Give each pair a picture to explain. Both axes use the same invented scale. Picture A shows similar changes at different levels; B shows diverging changes. Ask about a comparator losing its larger businesses. No significance test establishes the assumption.
:::

## What cannot be checked here? {.reader-slide}

<div class="big-number">One before period</div>

The GreenWaste file has one measure before and one after.

::: {.ask}
Can this file show whether earlier trends were similar?
:::
```{=html}
<button data-open="pretrend-feedback" data-closed="Open the limit" data-opened="Hide feedback" aria-controls="pretrend-feedback" aria-expanded="false">Open the limit</button><div id="pretrend-feedback" class="feedback" hidden>No. Ask for several earlier measures and an explanation of other changes affecting the groups.</div>
```


::: {.notes}
5 minutes. The earlier pre-trend test was misleading with only one before period. State the data limit explicitly. Even several earlier periods cannot prove the unobserved post-programme trend.
:::

## Mark up the report extract {.reader-slide}

Open **Sheet 2**. Read alone before comparing with a partner.

Circle the result and its units.

Underline the assumption.

Write one question that could change your decision.

::: {.ask}
What does the extract support, and what remains uncertain?
:::


::: {.notes}
10 minutes. Protect the reading time. Ask participants to distinguish the estimated extra fall from an established causal saving. The printed extract names two periods and the parallel-trends assumption, and admits earlier trends cannot be checked.
:::

## Repair an AI claim {.reader-slide}

<div class="lead">Illustrative AI draft</div>

"The extra fall proves GreenWaste caused the saving. Parallel trends is satisfied. Scale-up meets the decision rule."

::: {.ask}
Which claims can you support from the extract?
:::
```{=html}
<button data-open="did-ai" data-closed="Open feedback" data-opened="Hide feedback" aria-controls="did-ai" aria-expanded="false">Open feedback</button><div id="did-ai" class="feedback" hidden>The estimated extra fall is supported. Cause requires a credible assumption; the file cannot check earlier trends. The result is below the saving rule.</div>
```


::: {.notes}
7 minutes. Authored example, not a captured live AI response. Participants cross out unsupported claims on Sheet 2 and rewrite. Optional prompt: identify the comparison, assumption and missing evidence; ask me questions before concluding. Check any AI response against the source yourselves.
:::

## Your question for the evaluator {.reader-slide}

::: {.ask}
What would you ask, and how could the answer change your decision?
:::
Close the sheets. Write one question and its purpose. Work alone.

**Difference-in-differences:** compare the changes, then question the assumption.

::: {.notes}
4 minutes. Independent close, no AI. Collect a few questions about earlier trends, competing changes or stable group composition. Total 60 minutes.
:::
