---
title: "Good news. Enough to act?"
subtitle: "Day 1 | Session 2: Descriptive statistics and the comparison behind a cost claim"
author: "3ie · Review version"
always_allow_html: true
format:
  revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, session2_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session2_review_interactions.html]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r setup, include=FALSE}
source('session2_review.R', encoding='UTF-8')
```

## We read -669. Was it the programme's effect? {#s2-p01 .reader-slide .s1-review .s2-review}

GreenWaste offers equipment subsidies, installation help and staff training.

::: {.lead}
Participants' **mean annual waste cost fell by 669 AED**, over 12 months.
:::

::: {.ask}
Does that tell us how much GreenWaste caused costs to fall?
:::

Record your reason on **worksheet page 2**. Keep it to compare with your final reading.

::: {.notes}
4 minutes. One-minute reminder: city score rule, annual waste cost in AED, same businesses measured twice. Do not repeat the full case introduction. Opening judgement: one minute alone, one minute compare, one minute hear reasons without settling them. Ask for what is missing, rather than a technical label. Transition: we will inspect the records behind that mean.
:::

## One row. One business. Two measurements. {#s2-p02 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(s2_table(rows=TRUE))
```

::: {.ask}
Describe business **B00402** in one sentence: participation, costs and period.
:::

**Q1 on page 2.** These are four selected real records, not a representative sample.

::: {.notes}
5 minutes. Use H2-A1 on page 1. Each row is one business, not one measurement; before and after are columns in the same row. Annual costs measured before and 12 months after. Expected B00402: did not take part, score 88.3, costs rose from 2,425 to 2,897 AED. Explain identifiers and score briefly, not the later RDD method. Ask where the learner sees participation rather than guessing from costs. Transition: four rows are an illustration; the averages use all city records in their groups.
:::

## Who is included in these averages? {#s2-p03 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(s2_counts())
```

::: {.ask}
The participant average describes **4,794 businesses**, not all 10,000.
:::

Add the population to your **Q1** sentence. The pilot lottery is a separate dataset.

::: {.notes}
3 minutes. Point to denominator of each group average. The score rule may make participants systematically different from others. This is not random assignment in the city. Do not teach the cutoff method now. Check that no one uses 10,000 as participant denominator or merges the pilot with city data. Transition: a group mean still hides the distribution inside that group.
:::

## What do costs look like before the programme? {#s2-p04 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(paste0('<div id="s2-hist"><div data-variant>',s2_hist(),'</div><div data-variant hidden>',s2_hist(markers=TRUE),'</div></div>'))
```

<button data-swap="s2-hist" data-closed="Show mean and median" data-opened="Hide mean and median" aria-pressed="false">Show mean and median</button>

All **10,000 city businesses**; each bar counts businesses in a 250 AED cost band.

::: {.ask}
**Q2:** What is common? What would a single average hide?
:::

::: {.notes}
6 minutes. Read axes first: horizontal costs, vertical number of businesses, not cost over time. All city baseline records, not H1's selected ten. Give two minutes to describe shape/spread before revealing mean 1,862 and median 1,801. The mean is a little higher; high-cost businesses stretch the right tail. Do not imply a dramatic mean/median difference or introduce a quartiles lesson. Learners write one supported description and one limit of the mean in Q2. Transition: now follow the participant group over time.
:::

## Participants: what changed after GreenWaste? {#s2-p05 .reader-slide .s1-review .s2-review .code-example}

```{r, results='asis'}
html_out(s2_lines('participants'))
```
**How the trainer obtains the two means**

```{r, echo=TRUE}
# Same participants, two annual cost columns.
# Average first; round the displayed results.
round(colMeans(tp[c('cost_before', 'cost_after')]))
```

::: {.ask}
**Q3:** After minus before: 763 − 1,432 ≈ **−669 AED**. What can we say?
:::

::: {.notes}
6 minutes. Walk through code comments, not syntax; participants need not read or run R. Source H2-A2 on page 1. Actual difference uses unrounded means: -669.138, displayed -669. This is a mean fall among participants over a year, not yet caused saving. Learners calculate/interpret Q3 as taught. Common vertical scale begins at zero. Ask who and when, then sign and units. Transition: does a falling average require every cost to fall?
:::

## Did every business's cost fall? {#s2-p06 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(s2_hist(change=TRUE))
```

**4,794 participants.** Left of zero: costs fell. Right of zero: costs rose.

<button data-open="s2-rises" data-closed="Check the records" data-opened="Hide the record check" aria-controls="s2-rises" aria-expanded="false">Check the records</button>
<p id="s2-rises" hidden><strong>485 businesses had rising costs</strong>; one was unchanged. The mean fall does not describe everybody.</p>

::: {.notes}
5 minutes. Use H2-A3 on page 3. Read change axis, not before-cost axis from P04. Ask Q3 variation before opening record count: any values right of zero disprove everybody fell. Count 485 rising, 1 unchanged, 4,308 falling. These are recorded changes, not individual causal effects. Learner writes one supported statement in Q3; do not equate a histogram of changes with CI of the mean. Transition: even genuine falls can have more than one cause.
:::

## What else could have changed costs? {#s2-p07 .reader-slide .s1-review .s2-review}

<div class="comparison-pair"><div><h3>What we observe</h3><p>Participants before: <strong>1,432 AED</strong><br>Participants after: <strong>763 AED</strong></p></div><div><h3>What is missing</h3><p>The same participants' costs after a year <strong>without GreenWaste</strong>.</p></div></div>

Suppose waste-service prices changed during that year. **This is a hypothetical possibility, not a case fact.**

::: {.ask}
**Q3:** What evidence would help separate the programme from other changes?
:::

::: {.notes}
5 minutes. Two minutes pairs propose a concrete request, then hear two. Accept service-price histories, other concurrent support, or a credible untreated comparison; ask how each would help. Do not tell learners a shock actually happened. Before/after cannot observe the counterfactual. Do not claim we know direction/size of bias. Transition: other city businesses may help us ask that question, but are they comparable?
:::

## Other city businesses: what changed? {#s2-p08 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(s2_lines('others'))
```

::: {.ask}
**Q4:** 2,401 − 2,258 ≈ **+143 AED**. Does that prove what would have happened to participants?
:::

5,206 businesses above the participation score cutoff. Same period and cost units.

::: {.notes}
6 minutes. Learners calculate other-group change Q4 then compare descriptions with Q3. A rise in the other group rules out describing these two group means as a common city-wide fall. It does not validate a counterfactual for participants: score selection and starting costs differ. Means rounded for display, change computed from unrounded values. Do not subtract changes as an unexplained causal DiD estimate. Transition: another tempting number is the gap after a year.
:::

## Lower than without. Already lower before. {#s2-p09 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(paste0('<div id="s2-levels"><div data-variant>',s2_lines(baseline=FALSE),'</div><div data-variant hidden>',s2_lines(),'</div></div>'))
```

<button data-swap="s2-levels" data-closed="Reveal their starting costs" data-opened="Show after only" aria-pressed="false">Reveal their starting costs</button>

**After-only gap:** 763 − 2,401 ≈ **−1,638 AED**.

::: {.ask}
**Q4:** Would you credit GreenWaste with that whole gap? Why?
:::

::: {.notes}
7 minutes. Predict judgement before reveal, then show participants already had lower mean costs before (1,432 versus 2,258). Actual rounded gap from unrounded means is -1,638. The rule selects groups; baseline difference is direct evidence that they were not identical in mean cost. It does not prove that all selection concerns are observed or establish true bias direction. Learners state one concern and one evidence request in Q4. Transition: both numbers are real; neither alone answers the causal question.
:::

## Two numbers. Two weak comparisons. {#s2-p10 .reader-slide .s1-review .s2-review}

<div class="comparison-pair"><div><h3>Before/after: −669 AED</h3><p><strong>Same participant group</strong><br>After minus before</p><p>Misses what would have changed without the programme.</p></div><div><h3>With/without: −1,638 AED</h3><p><strong>Different groups, after a year</strong><br>Participants minus nonparticipants</p><p>Misses differences the groups already had.</p></div></div>

::: {.ask}
Which comparison sits behind the claim? What makes it fair?
:::

Use the **page 3 summary** to check Q3 and Q4. Neither number alone is a caused saving.

::: {.notes}
5 minutes. Have pairs point to group/time in both comparisons and explain one missing piece. Counterfactual means participants' outcome without the programme at the same follow-up time. These descriptives do not establish true causal effect or actual bias directions. Preview that a fair design, not a preferred number, is needed; pilot next session. Transition: use this reading to triage the four cards.
:::

## Four cost claims. Which would you act on? {#s2-p11 .reader-slide .s1-review .s2-review}

Read the **four separate cost cards**. Use **Q5 on worksheet page 4**.

<div class="claim-steps"><strong>Act</strong><strong>Ask first</strong><strong>Do not act</strong></div>

For each card, write the **source support**, your **missing question**, and why its answer matters.

::: {.ask}
Compared to what? How big? How sure?
:::

::: {.notes}
12 minutes. One set per group; preserve Fiona's triage grid. Two minutes read alone, seven minutes discuss/record, three minutes place card numbers and prepare a reason. Optional board has decision columns and question rows with words, not colour alone. Native board choices stay in this browser, are not a shared room poll; paper/card placement is the equivalent. Assign a reader and recorder but everyone states one reason. No extra AI wall. Cards use actual cost descriptives; card 2 deliberately mislabels before/after as effect. Do not introduce unexplained DiD. Transition: compare two disagreements rather than all presentations.
:::

## Which evidence supports your choice? {#s2-p12 .reader-slide .s1-review .s2-review}

Choose **two cards** that groups placed differently.

::: {.ask}
Which words or numbers support the choice? Which do not?
:::

What answer could change your advice? **Correct or strengthen your Q5 reason.**

::: {.notes}
5 minutes. Card 1: correct descriptive fall, ask causal comparison before expansion. Card 2: actual before/after interval but effect is unsupported; 669 below rule, statistical precision not attribution. Card 3: reject nationwide recommendation from p; reach, size and causal design matter. Card 4: both changes correct; ask comparability and uncertainty of the relevant causal estimate. How sure is a defensible primary placement, not the only one. Act on a descriptive finding may mean investigate, not expand; require scope of action. No single-answer placement test. Transition: see how AI can mix these same facts with unjustified advice.
:::

## Would you accept this AI chart reading? {#s2-p13 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(paste0('<div class="ai-reading">',s2_paragraphs(s2_ai),'</div>'))
```

::: {.ask}
On **page 5**, underline support; cross out overclaims. Check **H2-A2 on page 1**.
:::

::: {.notes}
8 minutes. AI Snapshot A: five minutes individual/pair markup, three minutes source-check debrief. Naive prompt on worksheet: explain chart and advise expansion. Authored exercise, not verified AI transcript. Supported means and changes, sensible caveat and monitoring. Unsupported attribution of whole after gap, other group automatically counterfactual, every participant improved, national transfer. The observation that these two means do not share a city-wide fall is correct; it does not by itself exclude other explanations. Ask for both a supported passage and an error. Transition: make the comparison explicit in the better prompt.
:::

## Ask for a comparison-aware explanation {#s2-p14 .reader-slide .s1-review .s2-review}

```{r, results='asis'}
html_out(paste0('<div class="better-prompt">',s2_paragraphs(s2_prompt),'</div>'))
```

Copy **page 5's prompt and H2-A2** into **Microsoft Copilot**. Keep your AI good-practice card beside you.

::: {.ask}
Which claim improved? Which still needs checking against the source?
:::

::: {.notes}
7 minutes. AI Snapshot B: four minutes run then three minutes compare. No account setup during session. If delayed/unavailable, use prepared page 6 response, same task and timing. Compare recorded versus caused change, after gap versus starting levels, every-business and national claims. Correctness comes from source support, not agreement of AI answers. Do not add a second mandatory chat review; point to the shared card. Transition: close AI responses and give your own advice.
:::

## What comparison would you ask for next? {#s2-p15 .reader-slide .s1-review .s2-review}

Keep **H2-A2 open**. Close the AI responses and turn over the cards.

::: {.ask}
Write one supported conclusion and one evidence request that would help decide whether to expand.
:::

On **page 4**, compare your final reason with your opening reason. What changed?

Next: the pilot used a lottery. Could that give us a fairer comparison?

::: {.notes}
6 minutes. Three minutes individual, one minute compare opening, two minutes feedback. Accept descriptive cost fall/after gap with explicit comparison and limit; request must identify evidence and decision purpose. Do not accept more data without explaining what it resolves. Bridge to pilot without teaching result or design here. Total 90 minutes. This is a separate review prototype; existing teaching source and handout unchanged.
:::
