---
title: "From Findings to Policy"
subtitle: "Module 2 · Day 4, Session 2: Evidence Translation"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"

# ---- The findings set and the brief template ------------------------------
# Shared with the printable pack and template, so the deck cannot disagree with
# the paper in the room's hands.
source("qa_translation.R")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · The decision

## You are not reading any more. You are writing.

::: {.panel}
Three days of reading other people's evidence. Yesterday you decided whether a
report was credible.

Today someone hands you the findings and asks what the authority should do. That
is a different job, and it is the one most of you actually have.
:::

::: {.ask .fragment}
**Would you recommend this programme for rollout?**
Yes / No / Only with changes.
:::

## The programme

```{r}
#| results: asis
cat(sprintf('
<div class="facts">
  <div class="k">Programme</div> <div class="v">%s</div>
  <div class="k">What it is</div> <div class="v">%s</div>
  <div class="k">Where</div>     <div class="v">%s</div>
  <div class="k">When</div>      <div class="v">%s</div>
</div>', transl_programme$name, transl_programme$what,
    transl_programme$where, transl_programme$when))
```

::: {.warn}
This is **not** the GreenWaste programme you have spent the week on. It is a
different scheme, and you have not seen these findings before.
:::

## The number the decision turns on

```{r}
#| results: asis
cat('
<div class="hero">
  <div class="value">900</div>
  <div class="label">AED per vehicle per year &middot; the threshold to clear</div>
</div>
')
```

::: {.ask .fragment}
Vote now, before you have read the findings. Then we will read them together, and
you will vote again before you write.
:::

:::::: {.columns}

::::: {.column width="58%"}
The evaluation was randomised, well executed, and reported transparently. Nothing in
it fails yesterday's checklist.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

# 1 · The findings

## Six findings, all of them sound

```{r}
#| results: asis
tab <- data.frame(
  `#` = seq_len(nrow(transl_findings)),
  Finding = transl_findings$finding,
  check.names = FALSE
)
kable(tab, col.names = c("#", "Finding"),
      caption = "The six findings, in the order the evaluation reports them") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Every one of these survives yesterday's checklist. There is no methodological
problem to find, and no number to distrust.
:::

::: {.ask .fragment}
So the difficulty is not whether to believe the evidence. What is it?
:::

## Finding 1: the saving

::: {.excerpt}
Fuel costs fell for participating operators. Net saving of **743 AED per vehicle
per year** (95% CI: 690 to 796). Statistically significant at conventional levels.
:::

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Estimate</div>  <div class="v">743 AED per vehicle per year</div>
  <div class="k">Interval</div>  <div class="v">690 to 796</div>
  <div class="k">Threshold</div> <div class="v">900</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
This is a clean, well-measured effect on the primary outcome, from a randomised
allocation. It is the finding the evaluation was designed to produce.
:::

::: {.warn .fragment}
The whole confidence interval lies below 900. On the commissioning authority's own rule, the
programme does not clear the bar.
:::
:::::

::::::

## Finding 2: the same finding, read against the rule

::: {.ask}
The evaluation says the saving is **743 AED**. The threshold is **900 AED**.

Is the scheme a success or a failure?
:::

::: {.panel .fragment}
The evaluation states the 157 AED shortfall plainly.

But notice what the shortfall does **not** tell you: whether the threshold was the
right threshold. Nothing in the evaluation was designed to answer that.
:::

## Finding 3: the pass-through

::: {.excerpt}
Freight rates charged by participating operators fell by an average of **1.8%**,
against **0.4%** among non-participants. The difference is **not statistically
significant** (p = 0.11).
:::

::: {.ask}
This is the finding most likely to end up in a brief. What would you write?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**What the evidence supports**

A fall in freight rates among participants, in the expected direction, larger
than among non-participants.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it does not**

That customers benefited. The estimate is not distinguishable from zero. This is
a hypothesis the evaluation could not settle, not a result.
:::
:::::

::::::

## Finding 4: who did not join

::: {.excerpt}
**34%** of eligible operators enrolled. Non-participants were more likely to be
small fleets and to operate on fixed public-sector contracts.
:::

::: {.ask .fragment}
The evaluation reports a pattern. Does it explain it?
:::

::: {.panel .fragment}
No. It measured **who** did not join, not **why**. The pattern is a fact; the
reason is a question the evaluation left open.

Briefs often blur this distinction. "Small operators did not take it
up" is reportable. "Small operators could not take it up" is an inference the
evidence does not support, and it points at a completely different
recommendation.
:::

## Finding 5: the cost of running it

::: {.excerpt}
Administration ran **22% over budget**. The overrun is concentrated in
verification of fuel purchase records, which required manual inspection in
**41%** of cases.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
This is not an estimate. It is a figure from the scheme's own accounts.

It is also the finding that bears most directly on the question the authority is
actually asking: whether to do this again.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
And it is the finding most likely to be **left out** of a brief, because it is
about delivery rather than impact.

If your brief recommends rollout and does not mention this, the authority will
discover it after deciding.
:::
:::::

::::::

## Finding 6: the null result

::: {.excerpt}
No measurable effect on vehicle maintenance or fleet renewal. Point estimates for
both outcomes are close to zero with **wide confidence intervals**. The evaluation
cannot rule out effects of either sign.
:::

::: {.ask .fragment}
How do you write a finding that says nothing was found, and is that a finding?
:::

::: {.panel .fragment}
Yes, and it is a useful one, but only if you say **which kind of nothing** this is.

This is *absence of evidence*: the confidence intervals are wide, so a meaningful effect
could be hiding in them. It is not *evidence of absence*. A brief that turns this
into "the scheme had no wider benefits" has overstated a null.
:::

# 2 · The brief

## Four elements, one page

```{r}
#| results: asis
tab <- data.frame(
  Element = brief_elements$element,
  `Length` = brief_elements$length_guide,
  `What it has to do` = brief_elements$asks,
  check.names = FALSE
)
kable(tab, col.names = c("Element", "Length", "What it has to do"),
      caption = "The four elements of the brief") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 15)
```

::: {.warn}
One side of A4. Decision-makers rarely read past the first page.
:::

## The headline sentence

::: {.ask}
Write your headline finding as **one sentence**, with the number in it.

Then read it back and ask: if this were the only sentence anyone read, would they
conclude the opposite of what you found?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The test**

A reader who stops after one sentence should not be able to reach a conclusion
the evidence does not support.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
"Fuel costs fell for participating operators" passes a readability test and fails
this one. It is true, and it lets the reader assume the threshold was cleared.

Including the number prevents that misreading.
:::
:::::

::::::

## The recommendation

::: {.ask}
Your recommendation must be **one specific action** the authority can actually
decide.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Decidable**

- Roll out to the three regions where the scheme was piloted
- Extend the pilot for two years and re-run the costing
- Do not roll out, and re-tender the verification process
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Not decidable**

- The scheme should be improved
- More evidence is needed
- Stakeholders should be engaged
:::
:::::

::::::

::: {.ask .fragment}
Which of the three decidable options is defensible on these findings, and does it depend on how you weigh the 743 against the 900?
:::

## The risk

::: {.ask}
Name **one thing** that would make your recommendation wrong, stated so that it
can be watched.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**A risk you can watch**

"If the 22% cost overrun is repeated at scale, the administration cost per vehicle
would exceed the saving. The first two quarters of spend would show this."
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**A caveat dressed as a risk**

"Results may not generalise." True of almost any evaluation. It tells
the authority nothing and can be neither confirmed nor falsified.
:::
:::::

::::::

# 3 · Pairs and critique

## Swap

::: {.panel}
Pairs, swap briefs. Read your partner's as though you were the person deciding.

**Three questions, in this order:**

1. What does this brief want me to do?
2. What is the strongest thing it claims, and does the findings pack support it?
3. What has it left out that I would have wanted to know?
:::

::: {.ask .fragment}
Question 3 is the one most often skipped. You are looking for
the omission, not the error.
:::

## What to look for

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Does the headline sentence contain the number?</div>\n')
cat('<div class="panel"><b>2.</b> Does it mention the 900 AED threshold at all?</div>\n')
cat('<div class="panel"><b>3.</b> Does it claim a customer benefit from the pass-through finding?</div>\n')
cat('<div class="panel"><b>4.</b> Does it explain why two-thirds did not enrol?</div>\n')
cat('<div class="panel"><b>5.</b> Does it mention the 22% cost overrun?</div>\n')
cat('<div class="panel"><b>6.</b> Does it turn the null finding into a statement of no effect?</div>\n')
```

::: {.warn .fragment}
Six checks. Every one of them is a way to write a brief that is accurate in every
sentence and misleading as a whole.
:::

## The two failures that matter most

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Omitting the threshold**

A brief that reports 743 and never mentions 900 leaves the reader to conclude the
scheme met its target. It is a common and consequential failure.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Omitting the cost overrun**

The authority is deciding whether to spend money. A brief that discusses impact
and not cost has answered a different question from the one asked.
:::
:::::

::::::

::: {.ask .fragment}
Both are omissions rather than errors. Which is harder to catch in your own draft?
:::

# 4 · Put AI to work

## The prompt

::: {.excerpt}
Here are six findings from an evaluation of a fuel-price stabilisation scheme for
small freight operators. Write a one-page policy brief for the commissioning
authority with a headline finding, key implications, a recommendation, and one
risk to flag.
:::

::: {.panel}
A reasonable prompt: it asks for exactly what the room was asked for. Read the
draft, then edit it as a team. The editing is the exercise.
:::

## What came back

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**Policy Brief: Fuel-Price Stabilisation Scheme**

**Headline finding.** The scheme reduced fuel costs for participating operators,
[delivering a net saving of 743 AED per vehicle per year]{.flag}. The result is
statistically significant and the evaluation was rigorously conducted.

**Key implications.** The scheme achieved its primary objective of reducing the
fuel cost burden on small freight operators. Operators [passed part of the saving
on to customers, with freight rates falling by 1.8% among participants]{.flag},
suggesting a wider economic benefit. [Some operators chose not to enrol]{.flag},
which may indicate a need for better communication about the scheme.

**Recommendation.** The scheme should be rolled out more widely, with
[targeted outreach to operators who did not participate]{.flag}.

**Risk to flag.** Results from a pilot may not generalise to a full rollout at
scale.
:::

::: {.ask .fragment}
There are four flagged problems. Find them. Then find the two things the draft
leaves out entirely, which are not flagged.
:::

## What AI did, and what it could not do

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- Produced the right four elements, in the right order
- Wrote in the register of a brief
- Kept to a page
- Was fluent enough to be sent without editing
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it did not**

- Never mentions the 900 threshold
- Calls a non-significant estimate a wider economic benefit
- Turns 34% enrolment into "some operators"
- Omits the 22% cost overrun entirely
- Names a risk that is a caveat, not a risk
:::

::::::

::: {.ask .fragment}
Notice the pattern: **it overstated the good news and softened the bad**. Check your own drafts, and any AI draft, for this.
:::

## The same task, a better prompt

::: {.excerpt}
The same six findings. Before you write anything, do two things.

First, list the findings that argue **against** rollout, and state for each one
how strongly it argues. Then list those that argue for it.

Second, tell me which of the six findings would have to be different for your
recommendation to change, and in which direction.

Then write the brief, and include a sentence that states plainly what the
evaluation did not establish.
:::

::: {.panel .fragment}
Three changes. **Ask for the counter-case first**, so the case against is not an
afterthought. **Ask what would change the recommendation**, which forces the
threshold into the analysis. And **require a statement of what was not
established**, which makes omission harder.
:::

::: {.ask .fragment}
The first instruction often makes the biggest difference, and you can use it on
any writing task.
:::

# 5 · Close

## Vote again

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">900</div>
  <div class="label">The bar, and what 743 sits below</div>
</div>
```

Same three options: Yes / No / Only with changes.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.ask .fragment}
Compare the two votes. If they moved, was it the threshold that moved them?
:::

## What makes a brief trustworthy

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> The number is in the headline sentence, not buried in the body.</div>\n')
cat('<div class="panel"><b>2.</b> The threshold the number is judged against is stated, even when it is missed.</div>\n')
cat('<div class="panel"><b>3.</b> Findings that argue against the recommendation are named, not omitted.</div>\n')
cat('<div class="panel"><b>4.</b> The recommendation is an action the reader can take or refuse.</div>\n')
cat('<div class="panel"><b>5.</b> The risk is something that could happen and could be watched.</div>\n')
cat('<div class="panel"><b>6.</b> What the evaluation did not establish is stated as plainly as what it did.</div>\n')
```

::: {.ask .fragment}
Which of the six is the hardest to write when you already know what you want the
recommendation to be?
:::

## Tomorrow

::: {.panel}
The last session. You take the evaluation design you started in Module 1 and
finish it: how you would present results, how you would cost the programme, what
your figures would look like, and how you would translate the findings.

Then you pitch it. Five minutes, with feedback.
:::

::: {.panel .fragment}
Bring everything from this week. The pitch is where it all has to fit together.
:::
