---
title: "Reading a Difference-in-Differences Result"
subtitle: "Module 2 · Day 2, Session 1: The Extra Change in the Programme Group"
author: "3ie"
always_allow_html: yes
# live-revealjs (quarto-live): the baked plotly/kable charts still render at knit time
# exactly as before; the three teaching code cells (Running the regression, Visualising
# the DiD, Testing for parallel pre-trends) are {webr} cells that run live in the browser
# via WebAssembly. Needs internet at presentation time to load webR. Export to PDF for a
# portable copy (live cells then show their code but do not run).
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    # Single-file deck. The live (webR) cells work fine with this over http —
    # via `quarto preview` or a published URL — but NOT from a file:// path.
    # Do not open the rendered .html directly from disk; serve it or publish it.
    embed-resources: true
engine: knitr
resources:
  - evaluation_data_GreenWaste_simple.csv
  - gw_lab.R
webr:
  packages:
    - dplyr
    - ggplot2
    - estimatr
    - broom
  cell-options:
    autorun: false
    edit: true
    warning: false
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(estimatr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

# ---- Tables: plain knitr::kable, styled by theme_dge.scss -------------------
# No kableExtra: its JavaScript needs jQuery, which live-revealjs does not load.
# font_size in px; hl = rows to shade (hl_bold to bold them); note = a line
# under the table. The table is centred, as kable_styling(full_width = FALSE) did.
HL_BLUE <- "#E4EBF3"; HL_RED <- "#F8E9EA"
tbl <- function(df, ..., font_size = NULL, hl = NULL, hl_bg = HL_BLUE,
                hl_bold = FALSE, note = NULL) {
  h <- as.character(knitr::kable(df, format = "html", ...))
  fs <- if (is.null(font_size)) "" else sprintf(" font-size:%dpx;", font_size)
  h <- sub("<table", sprintf('<table style="margin-left:auto; margin-right:auto;%s"', fs),
           h, fixed = TRUE)
  if (length(hl)) {
    parts <- strsplit(h, "<tbody>", fixed = TRUE)[[1]]
    rows  <- strsplit(parts[2], "<tr>", fixed = TRUE)[[1]]
    st <- sprintf("background:%s;%s", hl_bg, if (hl_bold) " font-weight:700;" else "")
    for (i in hl) {
      rows[i + 1] <- gsub('<td style="', paste0('<td style="', st, " "), rows[i + 1], fixed = TRUE)
      rows[i + 1] <- gsub("<td>", paste0('<td style="', st, '">'), rows[i + 1], fixed = TRUE)
    }
    h <- paste0(parts[1], "<tbody>", paste(rows, collapse = "<tr>"))
  }
  if (!is.null(note))
    h <- paste0(h, sprintf('\n<p style="text-align:center; font-size:0.75em; color:%s; margin-top:0.3em;">%s</p>',
                           MUTED, note))
  fence <- strrep("`", 3)
  knitr::asis_output(paste0("\n", fence, "{=html}\n", h, "\n", fence, "\n"))
}

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))
gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. Everything below reads from here.
# This is the short JOIN code participants type at menti.com, NOT the long
# alphanumeric ID in the editor URL. Confirm it while the Menti is live.
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/al68xcnp6hm6zmc9k94rhbxwimzq6m29/embed"

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- The decision rule, fixed for the whole week ---------------------------
RULEBAR <- 1000    # must cut costs by at least this many AED

# ---- GreenWaste: the rest of the city --------------------------------------
# The simple case (SIMPLE_GREENWASTE_PLAN.md). In the rest of the city every
# business scoring 58 or below took part and none above did; DiD compares the
# change in costs of the two groups. greenwaste_case.R is sourced into its own
# environment because it defines RULE (here a colour).
case_env <- new.env()
case_env$gw <- read.csv("./evaluation_data_GreenWaste_simple.csv")
sys.source("greenwaste_case.R", envir = case_env)
city  <- case_env$city
pilot <- case_env$pilot
took  <- city$took_part == 1
rct_est <- case_env$case$rct

# One row per business per period: the shape an evaluator's regression uses.
long <- data.frame(business = rep(city$business, 2), after = rep(0:1, each = nrow(city)),
                   took_part = rep(city$took_part, 2), cost = c(city$cost_before, city$cost_after))

off_b <- mean(city$cost_before[took]);  off_a <- mean(city$cost_after[took])    # took part
ctl_b <- mean(city$cost_before[!took]); ctl_a <- mean(city$cost_after[!took])   # did not

off_chg  <- off_a - off_b               # the before-and-after change
ctl_chg  <- ctl_a - ctl_b               # the comparison group's change
did_hand <- off_chg - ctl_chg           # the four numbers done by hand

# ---- The regression an evaluator would actually report ---------------------
fit_did <- lm_robust(cost ~ after * took_part, data = long,
                     clusters = business, se_type = "stata")
did_est <- coef(fit_did)[["after:took_part"]]
did_ci  <- confint(fit_did)["after:took_part", ]
did_p   <- fit_did$p.value[["after:took_part"]]

# Sign trap: the coefficient is negative, so conf.low is the LARGER effect and
# abs(conf.low) is the OPTIMISTIC end. Sorting the absolute values makes the
# first element the end closest to zero, which is the one the decision turns on.
did_avoided       <- sort(abs(did_ci))
did_avoided_least <- did_avoided[1]
did_avoided_most  <- did_avoided[2]
did_clears        <- did_avoided_least >= RULEBAR

# ---- The two traps, both computed rather than asserted ---------------------
trap_gap  <- off_b - ctl_b              # the baseline gap: the groups did not start level
trap_time <- ctl_a - ctl_b              # the time trend

# ---- The two naive answers, which bracket the truth ------------------------
naive_ba <- off_a - off_b               # before-and-after only
naive_ww <- off_a - ctl_a               # with-and-without at follow-up

# ---- The counterfactual line, for the chart -------------------------------
cf_after <- off_b + ctl_chg

did_tbl_viz <- tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Took part", "Did not take part"), each = 2),
  Cost  = c(off_b, off_a, ctl_b, ctl_a)
)

# The table as a participant would receive it.
did_tbl <- fit_did %>% tidy() %>%
  mutate(` ` = c("(Intercept)", "After", "Took part", "After x Took part"),
         CI = sprintf("[%s, %s]", gfmt(conf.low), gfmt(conf.high))) %>%
  select(` `, estimate, std.error, p.value, CI)

# ---- A check this data does allow ------------------------------------------
# The lottery in the pilot district made 200 businesses wait. They score 58 or
# below, like the city businesses that took part, and got nothing: their change
# is what businesses of that kind would have done without the programme.
wait     <- pilot[pilot$took_part == 0, ]
wait_chg <- mean(wait$cost_after - wait$cost_before)
did_wait <- off_chg - wait_chg

# ---- How far apart the groups started ---------------------------------------
bal_vars <- c("manager_age", "staff", "area", "filtration", "cost_before")
bal_labs <- c(manager_age = "Manager age (years)", staff = "Staff",
              area = "Premises (100 m2)", filtration = "Advanced filtration (share)",
              cost_before = "Waste costs before (AED)")
balance_tbl <- map_dfr(bal_vars, function(v) {
  m <- lm_robust(city[[v]] ~ city$took_part)
  tibble(Characteristic = bal_labs[[v]],
         Programme   = mean(city[[v]][took]),
         Comparison  = mean(city[[v]][!took]),
         Difference  = coef(m)[2],
         p = tidy(m)$p.value[2])
})
```


# 1 · Compared to what {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

<!-- Hidden setup (webR priming, deck CSS). Kept under the first section heading, not before it: content before the first heading makes Quarto create an empty slide after the title. -->
:::: {style="display:none;"}
```{webr}
#| autorun: true
#| edit: false
#| echo: false
#| output: false
#| warning: false
#| message: false
# ---- Live (webR) setup ----------------------------------------------------
# Runs once in the browser when the deck opens, priming the shared webR session
# so the three teaching cells (Running the regression, Visualising the DiD,
# Testing for parallel pre-trends) have their data and helpers ready. This is
# separate from the knitr `setup` chunk above, which prepares the baked charts.
library(dplyr)
library(ggplot2)
library(estimatr)
library(broom)

gw    <- read.csv("evaluation_data_GreenWaste_simple.csv")
city  <- subset(gw, setting == "city")     # the score rule: 58 or below took part
source("gw_lab.R")   # the GreenWaste lab: an invented world where the true effect is known

# One row per business per period (after = 0 before the programme, 1 after).
long <- data.frame(business = rep(city$business, 2), after = rep(0:1, each = nrow(city)),
                   took_part = rep(city$took_part, 2), cost = c(city$cost_before, city$cost_after))

t1 <- city$took_part == 1
off_b <- mean(city$cost_before[t1]);  off_a <- mean(city$cost_after[t1])
ctl_b <- mean(city$cost_before[!t1]); ctl_a <- mean(city$cost_after[!t1])
off_chg <- off_a - off_b
ctl_chg <- ctl_a - ctl_b
did_hand <- off_chg - ctl_chg

tlev <- function(x) factor(x, levels = c("Before", "After"))
viz <- data.frame(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Took part", "Did not take part"), each = 2),
  Cost  = c(off_b, off_a, ctl_b, ctl_a))

ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
grp_cols <- c("Took part" = ACCENT, "Did not take part" = ACCENT2)

theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(text = element_text(colour = INK),
          axis.title = element_text(colour = MUTED, size = rel(0.85)),
          axis.text = element_text(colour = MUTED),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
          legend.position = "bottom", legend.title = element_blank(),
          strip.text = element_text(colour = INK, size = rel(0.95)),
          plot.margin = margin(4, 10, 4, 4))
}

CUTOFF <- 58; THRESHOLD <- 1000

# Plots for "Visualising the DiD" and "Testing for parallel pre-trends", kept here
# so each slide shows one line of code and the chart fits the slide.
did_plot <- function() {
  cf <- off_b + ctl_chg
  ggplot(viz, aes(Time, Cost, colour = Group, group = Group)) +
    geom_line(linewidth = 1.4) +
    geom_point(size = 3.5) +
    annotate("segment", x = 1, xend = 2, y = off_b, yend = cf,
             linetype = "22", colour = ACCENT) +
    annotate("label", x = 2, y = (off_a + cf) / 2, colour = ALERT,
             label = sprintf("DiD = %.0f AED", did_hand), fill = "white") +
    scale_colour_manual(values = grp_cols) +
    labs(x = NULL, y = "Waste costs (AED)") +
    theme_m2()
}
pretrend_plot <- function() {
  set.seed(1)
  pre <- expand.grid(period = 1:4, unit = 1:50,
                     group = c("Comparison", "Programme")) %>%
    mutate(y = 10 + 1.5 * period +
      ifelse(group == "Programme" & period > 2, -3 * (period - 2), 0) + rnorm(n()))

  ggplot(pre, aes(period, y, colour = group)) +
    stat_summary(fun = mean, geom = "line", linewidth = 1.3) +
    geom_vline(xintercept = 2.5, linetype = "dashed", colour = ALERT) +
    scale_colour_manual(values = c(Programme = ACCENT, Comparison = ACCENT2)) +
    labs(x = "Period", y = "Outcome") +
    theme_m2()
}
```
::::

## The GreenWaste case

{{< include _case_map.qmd >}}

::: {.notes}
About 1 minute: a reminder of the case from Day 1. Point at the pilot district
(a lottery decided) and the rest of the city (the score decided). This session
works in the rest of the city, and uses the pilot district once, in Section 4.
:::

## The GreenWaste programme, one more time

:::::: {.columns}

::::: {.column width="54%"}
**GreenWaste** cut waste-management costs for businesses that took it up. You
met it on Day 1.

Today the question changes from *did costs fall* to **how much of the fall
was the programme**, and how much was going to happen anyway.

The bar is unchanged: **cut costs by at least `r gfmt(RULEBAR)` AED**.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>  <div class="v">GreenWaste, waste-cost reduction</div>
  <div class="k">Outcome</div>    <div class="v">Waste-management costs (AED)</div>
  <div class="k">Method</div>     <div class="v">Difference-in-differences</div>
  <div class="k">The rule</div>   <div class="v">Cut at least 1,000 AED</div>
</div>
```
:::::

::::::

::: {.panel}
Day 1 gave you three questions. Today is question one, **compared to what?**,
with a method that answers it properly.
:::

::: {.notes}
Sections 1 and 2 are the "remember" recap: aim for 5 to 10 minutes total. The
room already met GreenWaste, the before-and-after and with-and-without numbers,
and the idea of subtracting one group's change from the other's (on landfill in
Day 1 Session 1). Today's data is the rest of the city: businesses scoring 58 or
below took part, the rest did not. Move through the recall quickly. The new material
today is reading the table (Section 3) and the one assumption DiD rests on
(Section 4): that is where the time should go.
:::

## What happened to the businesses that took it up

```{r fig.height=4.2, fig.width=11.5}
tibble(Time = tlev(c("Before", "After")), Cost = c(off_b, off_a)) %>%
  plot_ly(x = ~Time, y = ~Cost, type = "scatter",
          mode = "lines+markers+text",
          line = list(color = ACCENT, width = 4),
          marker = list(color = ACCENT, size = 15),
          text = ~paste0(gfmt(Cost), " AED"), textposition = "top center",
          textfont = list(size = 17, color = INK), showlegend = FALSE,
          hovertemplate = "%{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 1900))) %>%
  pl_m2(height = 400, title = "Businesses that took part")
```

::: {.ask}
Costs fell by **`r gfmt(abs(off_chg))` AED**, a drop of
**`r sprintf("%.0f", 100 * abs(off_chg) / off_b)`%**. Is the programme's true
effect bigger or smaller than that?
:::

::: {.notes}
Do not editorialise. Present it the way an evaluator might present a headline
finding, and let it look good. The full picture only appears once the comparison group is on the chart, which is the next move after the Menti.
:::

## `[MENTI PLACEHOLDER]` Did the programme work?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
On the before-and-after reading alone: **did GreenWaste work, and would you
fund it?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Can't tell yet
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
Commit now. "Can't tell yet" is a legitimate answer and I want to see how many
choose it; we will come back to it after the next chart.
:::

## Now we add the businesses that did not receive the program

```{r fig.height=4, fig.width=11.5}
did_tbl_viz %>%
  plot_ly(x = ~Time, y = ~Cost, color = ~Group, type = "scatter",
          mode = "lines+markers",
          colors = c("Took part" = ACCENT, "Did not take part" = ACCENT2),
          line = list(width = 4), marker = list(size = 13),
          hovertemplate = "%{fullData.name}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2800))) %>%
  pl_m2(height = 380)
```

::: {.ask}
`[MENTI PLACEHOLDER]` Same question, now with the comparison group.
:::

::: {.panel .fragment}
The comparison group's costs **rose** over the same period. So the programme's effect
is *larger* than the headline drop.
:::

::: {.notes}
Silence for a beat. Let them see the comparison line rising, then ask the Menti
question again and watch the hands change. Show the fragment after the vote. We
compute the effect exactly next.
:::

# 2 · The idea, in four numbers {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Remember from Day 1: two comparisons, two biases

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Before and after only.** Credits the programme with the **time trend**: whatever
would have moved costs anyway.

On Day 1 this gave **-`r gfmt(abs(naive_ba))` AED**.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Took part vs did not, afterwards only.** Credits the programme with the
**baseline gap**: differences that existed before it started.

On Day 1 this gave **-`r gfmt(abs(naive_ww))` AED**.
:::
:::::

::::::

::: {.ask}
You met both of these yesterday: **`r gfmt(abs(naive_ba))`** and
**`r gfmt(abs(naive_ww))` AED**. Which one is too small, which is too big, and
why?
:::

::: {.notes}
Recap: these are the two naive comparisons from Day 1 Sessions 2 and 3, on this
same GreenWaste data. Take one answer for each, then say that difference-in-differences
removes both problems at once. The new work today is the assumption DiD rests on.
:::

## One sentence to take to a minister

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Compare <strong>the change over time</strong> in the group that got it,<br>with <strong>the change over time</strong> in a group that did not.</div>
</div>')
```

::: {.panel}
The difference between those two changes is the estimated effect.
:::

::: {.ask}
Which two changes would you subtract here?
:::

## Your turn: fill in the blanks

Difference-in-differences = (change for "took part") − (change for "did not")

:::::: {.columns}

::::: {.column width="44%"}
::: {.panel}
**First difference:** each group's change over time. This nets out fixed
differences between the groups, but still contains the general time trend.

**Second difference:** the difference between those two changes. This nets out
the time trend and leaves the estimated effect.
:::
:::::

::::: {.column width="56%"}
```{r}
tibble(
  Group  = c("Took part", "Did not take part", "Difference of the changes"),
  Before = c(gfmt(off_b), gfmt(ctl_b), ""),
  After  = c(gfmt(off_a), gfmt(ctl_a), ""),
  Change = c("?", "?", "?")
) %>%
  tbl(align = "lrrr", col.names = c("", "Before", "After", "Change"), font_size = 20)
```
:::::

::::::



::: {.panel .fragment}
**Watch the sign.** "Did not take part" went **up**, so the DiD is bigger than the top row's fall.
:::

::: {.notes}
Paper (Oct13_session1_materials.docx): Hand out "By hand: four numbers" (pack
page 1, one per participant) before this slide.

Give the room under two minutes for the two subtractions.
:::

## The answer

```{r}
tibble(
  Group  = c("Took part", "Did not take part", "Difference of the changes"),
  Before = c(gfmt(off_b), gfmt(ctl_b), ""),
  After  = c(gfmt(off_a), gfmt(ctl_a), ""),
  Change = c(sprintf("%+.0f", off_chg), sprintf("%+.0f", ctl_chg),
             sprintf("%.0f", did_hand))
) %>%
  tbl(align = "lrrr", col.names = c("", "Before", "After", "Change"),
      font_size = 24, hl = 3, hl_bold = TRUE)
```

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the difference-in-differences estimate</div>
</div>', sprintf("%.0f", did_hand)))
```

::: {.ask}
You have just estimated a causal effect with two subtractions. What would a
regression add?
:::

::: {.notes}
Answer to the ask: the same two subtractions, plus a confidence interval.
:::

## Visualising the DiD

::: {.ask}
Before you press **Run**: where would "took part" have ended up without the
programme?
:::

```{webr}
#| fig-width: 9
#| fig-height: 3.05
did_plot()     # the two groups, and the dotted line: where "took part" would have gone
```

::: {.panel .fragment}
The dotted line's rise (`r sprintf("%+.0f", ctl_chg)`) is the time trend. The gap
between its end and where "took part" actually finished is the programme:
`r gfmt(abs(did_hand))` AED.
:::

## The GreenWaste lab: add DiD

:::::: {.columns}

::::: {.column width="46%"}
```{ojs}
//| echo: false
viewof did_trend = Inputs.range([0, 600], {value: 0, step: 50, label: "Costs rise for everyone anyway (AED)"})
viewof did_size_gap = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners are smaller businesses (AED)"})
viewof did_hidden = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners differ in something nobody measured (AED)"})
viewof did_diverge = Inputs.range([0, 600], {value: 0, step: 50, label: "Joiners' costs were already falling faster (AED)"})
```
:::::

::::: {.column width="54%"}
::: {.ask}
DiD joins the lab, with one new problem: the last slider. **Which sliders will
DiD survive, and which one breaks it?**
:::

::: {.panel .fragment}
It survives the first three. The fourth breaks it: DiD assumes the two groups
would have moved in parallel.
:::
:::::

::::::

```{webr}
#| input:
#|   - did_trend
#|   - did_size_gap
#|   - did_hidden
#|   - did_diverge
#| autorun: true
#| edit: false
#| echo: false
#| fig-width: 10
#| fig-height: 2.9
lab(c("before_after", "with_without", "randomised", "did"),
    trend = did_trend, size_gap = did_size_gap, hidden = did_hidden, diverge = did_diverge)
```

::: {.notes}
About 5 minutes. All sliders start at zero, where every method sits on the
truth. Take a prediction for each slider before moving it. Costs rising anyway
moves only before-and-after; the two "joiners differ" sliders move only
with-and-without; DiD stays on -900 for all three. The last slider (joiners'
costs already falling faster) pulls DiD away: that is the parallel-trends
assumption, in one slider. The randomised comparison survives all four.
:::

# 3 · Implementing DiD in R {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Remember: what DiD needs from the data

::: {.panel}
1. **An outcome**, measured the same way each time
2. **A before and an after**: a time variable
3. **A group that got the programme and one that did not**: a group variable
4. **The same units followed over time**, or comparable samples in each period
5. **Ideally, more than one "before" period**, so we can check the groups were moving together before the programme
:::

::: {.panel}
Before any DiD, ask AI: *what data do I need, and does my dataset have it?* Ask
it to question you about your data before it answers.
:::

::: {.notes}
Tie each item back to the minister sentence: "the change over time" needs items
2 and 4; "in a group that did not" needs item 3. Item 5 is a deliberate
foreshadow of Section 4. Do not resolve it here.
:::

## Does GreenWaste have what we need?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
long[order(long$business, long$after), ] |> head(6)

table(took_part = long$took_part, after = long$after)
```
:::::

::::: {.column width="44%"}
::: {.panel}
✔ **Outcome:** `cost`, waste costs in AED

✔ **Time:** `after`, 0 before and 1 after

✔ **Group:** `took_part`, 1 took part and 0 did not

✔ **Same businesses:** each one appears in both periods
:::

::: {.warn}
⚠ Only **one** before period. Hold that thought for Section 4.
:::
:::::

::::::

::: {.notes}
Press Run. Read one business across its two rows so the room sees "one row per
business per period". The counts show `r format(sum(took), big.mark = ",")`
businesses that took part and `r format(sum(!took), big.mark = ",")` that did
not, each observed twice.

If asked about the groups: these are the businesses in the rest of the city.
Every business scoring 58 or below took part and none above did, so "did not take
part" means the businesses above the cut-off. That is our comparison.

The warning sets up "What this data can and cannot show" in Section 4: with a single
pre-period, parallel pre-trends cannot be tested.
:::

## Running the regression

```{webr}
out_did <- lm_robust(cost ~ after * took_part, data = long,
                     clusters = business, se_type = "stata")

res <- tidy(out_did)[, c("term", "estimate", "std.error", "p.value")]
res[, -1] <- round(res[, -1], 1)
res
```

::: {.ask}
Press **Run**. An intercept and three coefficients, and only one of them is the
programme's effect. Which row?
:::

::: {.notes}
Paper (Oct13_session1_materials.docx): Hand out "Which row is the impact?"
(pack page 2, one per participant) now, after the hand calculation. Tell the
room to leave question 4 until Section 4 ("What this data can and cannot show").

Before running, point at the formula: `after * took_part` gives the time
indicator, the group indicator and their interaction (`after:took_part`), and
the clusters line counts each business once, because it appears twice.

Run the chunk live, then take a vote across the three coefficients. Many will
pick `took_part` because it is large and has the words "took part" in it; that is
the trap the next slides unpack. The number in `after:took_part` is the same
`r gfmt(did_hand)` they got by hand.
:::

## Reading the table

:::::: {.columns}

::::: {.column width="50%"}
```{r}
did_tbl %>%
  mutate(across(c(estimate, std.error), ~ round(.x, 0)),
         p.value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))) %>%
  tbl(align = "lrrrr", digits = 2,
      col.names = c("", "Coefficient", "Std. error", "p", "95% CI"),
      font_size = 15, hl = 4, hl_bold = TRUE)
```
:::::

::::: {.column width="50%"}
::: {.panel}
One regression produces all four numbers from your hand calculation.

**Row 4 is the impact.** The interaction of after and took part: the extra
change in the group that took part, over and above the change everyone
experienced.
:::

::: {.warn}
Rows 2 and 3 are the **two things you must not report as the effect**:
`r gfmt(abs(coef(fit_did)[["took_part"]]))` AED is the baseline gap, and
`r gfmt(coef(fit_did)[["after"]])` AED is the time trend.
:::
:::::

::::::

::: {.notes}
Reminder for the room (vocab from Day 1 Session 1): the **coefficient** is the
effect size, here in AED; the **95% confidence interval** is the range of
plausible true values, not a margin of certainty; the **p-value** is how
surprising this result would be if the programme had truly done nothing. Walk row
4 first, then name rows 2 and 3 as the two numbers a careless report would quote
instead. Facilitation question worth asking here: which row would you want on the
front page of the briefing, and which two would a careless evaluator report?
:::

## The number that answers your question

::: {.ask}
The rule is `r gfmt(RULEBAR)` AED. Does the estimate clear it?
:::

:::: {.fragment}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the estimated cut in costs (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">%s</span><span class="k">Clears the bar</span></div>
</div>', gfmt(abs(did_est)), gfmt(did_ci[1]), gfmt(did_ci[2]),
   gfmt(RULEBAR), gfmt(did_avoided_least), if (did_clears) "Yes" else "No"))
```

::: {.warn}
The estimate is statistically clear: the interval sits far from zero. The rule
asks something else. Even the interval's most generous end reaches only
`r gfmt(did_avoided_most)` AED. **`r if (did_clears) "It clears the bar." else "It does not clear the bar."`**
:::
::::

::: {.ask .fragment}
Day 1: *statistically significant does not mean big enough.* Where have you seen
a significant result that was too small to act on?
:::

# 4 · The one question {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Everything rests on one assumption

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Without the programme, the two groups<br>would have moved in parallel.</div>
</div>')
```

::: {.panel}
The dotted counterfactual line from Section 2 exists only if this is true. It
cannot be **proven** from the result. It can only be made more or less
plausible.
:::

::: {.ask}
Before you look at an estimate, what would you ask an evaluator to show you
that makes this assumption plausible?
:::

## Which of these would you trust?

```{r fig.height=3.2, fig.width=11.5}
mk <- function(lab, tr, ct) {
  tibble(panel = lab, t = rep(1:4, 2),
         grp = rep(c("Programme", "Comparison"), each = 4), y = c(tr, ct))
}
bind_rows(
  mk("A. Moving in parallel",  c(40, 42, 44, 46), c(50, 52, 54, 56)),
  mk("B. Already diverging",   c(40, 41, 42, 43), c(50, 53, 56, 59)),
  mk("C. Already converging",  c(40, 44, 48, 52), c(50, 52, 54, 56))
) %>%
  plot_ly(x = ~t, y = ~y, color = ~grp, frame = ~panel,
          type = "scatter", mode = "lines+markers",
          colors = c(Programme = ACCENT, Comparison = ACCENT2),
          line = list(width = 3), marker = list(size = 9),
          hovertemplate = "%{fullData.name}: %{y}<extra></extra>") %>%
  animation_opts(frame = 900, transition = 0) %>%
  animation_slider(currentvalue = list(prefix = "Panel: "),
                   font = list(size = 11)) %>%
  layout(xaxis = ax("Periods before the programme started", zeroline = FALSE,
                    dtick = 1),
         yaxis = ax("Outcome", range = c(35, 62))) %>%
  pl_m2(height = 310, t = 34, b = 88)
```

::: {.ask}
Press play, or drag the slider. In which panel would difference-in-differences
give you a trustworthy number?
:::

::: {.warn .fragment}
Only **A**. In B the gap was already widening and in C already closing, so part
of what the method calls an effect was going to happen anyway.

Note the directions: **B would overstate** the effect, **C would understate**
it.
:::

## Testing for parallel pre-trends

::: {.ask}
Several periods *before* the programme: if the groups are comparable and the
programme worked, what should the lines do before the dashed line, and after?
:::

```{webr}
#| fig-width: 9
#| fig-height: 2.65
pretrend_plot()   # illustrative data: four periods, programme starts after period 2
```

::: {.warn .fragment}
Parallel **before** the dashed line, split only after: a credible pre-trend.
GreenWaste has only **one** period before the programme, so it cannot run this
check.
:::

::: {.notes}
Illustrative data, not GreenWaste. Point out that the groups move together before
the dashed cut-off and only diverge after. Then contrast: GreenWaste has a single
before-period, so the check cannot be run, which sets up the next slide.
:::

## What this data can and cannot show

:::::: {.columns}

::::: {.column width="54%"}
::: {.warn}
We measured costs at **two points**: once before the programme, once after.

Testing whether the groups were already diverging needs **at least two
observations before it started**. We have one.
:::

::: {.panel}
So the parallel trends assumption **cannot be tested with this data**.
:::
:::::

::::: {.column width="46%"}
::: {.ask}
Most evaluations you receive have this limitation. So what do you do instead?
:::
:::::

::::::

::: {.notes}
Paper (Oct13_session1_materials.docx): the room now answers question 4 on
"Which row is the impact?" (pack page 2). Take two or three answers to the ask
before moving on.
:::

## A check this data does allow

```{webr}
waiting <- subset(gw, setting == "pilot" & took_part == 0)   # the lottery said "not yet"
did_not <- subset(gw, setting == "city" & took_part == 0)
round(data.frame(waiting_businesses = mean(waiting$cost_after - waiting$cost_before),
                 city_did_not_take_part = mean(did_not$cost_after - did_not$cost_before)))
```

::: {.ask}
The pilot lottery made `r nrow(wait)` businesses wait. They score 58 or below,
like the city businesses that took part. **Before you press Run: did their costs
rise as much as those of city businesses that did not take part?**
:::

::: {.warn .fragment}
They rose `r gfmt(wait_chg)` AED, not `r gfmt(ctl_chg)`. The two city groups were
**not** moving in parallel, so the DiD **understates** the effect. With the
waiting businesses as the comparison: `r gfmt(did_wait)` AED, close to the
lottery's `r gfmt(rct_est)`.
:::

::: {.notes}
About 4 minutes. Most evaluations have no group like this; here the pilot's
waiting list provides one. The point to land: the assumption could not be tested
with the DiD data alone, and when another group finally allows a check, it fails.
:::


## How far apart did they start?

:::::: {.columns}

::::: {.column width="56%"}
```{r}
balance_tbl %>%
  mutate(across(c(Programme, Comparison), ~ round(.x, 2)),
         Difference = round(Difference, 2), p = round(p, 2)) %>%
  tbl(align = "lrrrr", digits = 2,
      col.names = c("Characteristic", "Took part", "Did not",
                    "Difference", "p-value"), font_size = 13)
```
:::::

::::: {.column width="44%"}
::: {.panel}
The comparison group is businesses above 58: older managers, more staff,
larger premises, and costlier from the start (`r gfmt(ctl_b)` vs `r gfmt(off_b)` AED).
:::

::: {.warn}
Not a pre-trend test, but the closest this data allows. Are the differences
**small**, and are the p-values **not tiny**?
:::
:::::

::::::

::: {.ask}
These groups were split by an eligibility score, not by chance. How much does
the result now lean on the parallel trends assumption?
:::

::: {.notes}
Recap: they built a balance table on Day 1 Session 3. It makes the
parallel-trends point concrete, because the pre-trend test cannot be run on
two-wave data. Land the ask: the result leans heavily on the assumption, and a
recommendation should say so. Reminder for the room: the
**p-value** answers "how surprising is this baseline gap if the two groups were
really the same?", so a tiny p-value is the worry here, not the reassurance.
:::

## What to ask when it cannot be tested

```{=html}
<div class="facts">
  <div class="k">Ask</div> <div class="v">"Which comparison group did you use, and why that one?"</div>
  <div class="k">Ask</div> <div class="v">"What differed between the groups before the programme started?"</div>
  <div class="k">Ask</div> <div class="v">"Is there any earlier data, even one extra year, that could show the trend?"</div>
  <div class="k">Ask</div> <div class="v">"Is there a group like the programme group that did not get it yet, such as a waiting list?"</div>
  <div class="k">Then</div> <div class="v">treat the estimate as a plausible range, not a precise number</div>
</div>
```

::: {.panel}
None of these questions requires you to run anything. A commissioner is
entitled to ask them. Expect an evaluator to answer all four.
:::

# 5 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: explaining the table {.dense}

::: {.panel}
Paste the Section 3 table into an AI tool with: **"Explain this regression
table to a non-technical decision-maker, and say what should be checked before
believing the result."**
:::

::: {.ai-output}
*A response like this is possible:*

The table reports a difference-in-differences estimate. **The programme reduced
costs by `r gfmt(abs(did_est))` AED**, and the effect is highly statistically significant
(p < 0.001), [so the programme was a success]{.flag}. [The confidence interval does
not include zero, which confirms the finding is robust]{.flag}.

[The parallel trends assumption has been satisfied]{.flag}. [The result clears
the 1,000 AED threshold required for scale-up]{.flag}.
:::

::: {.ask}
Underline the sentences that the table supports. Cross out the sentences that the table does **not** support.
:::

::: {.notes}
Paper (Oct13_session1_materials.docx): Hand out "AI Snapshot: explaining the
table" (pack page 3, one per pair).
:::

## Debrief: four claims the table does not support

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "So the programme was a success."** It answers *is the effect non-zero*,
not *does it clear the 1,000 AED bar*. Day 1 warned about the same
substitution.
:::

::: {.warn}
**2 · "The interval does not include zero, so the finding is robust."**
"Not zero" and "big enough" are different claims. The confidence interval is
[`r gfmt(did_ci[1])`, `r gfmt(did_ci[2])`].
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "The parallel trends assumption has been satisfied."** The serious one.
**The two-wave data cannot test it, and the pilot's waiting businesses suggest
it fails.** The model reported a test that was never run.
:::

::: {.warn}
**4 · "The result clears the 1,000 AED threshold."** False. Even the
interval's most generous end is `r gfmt(did_avoided_most)` AED.
:::
:::::

::::::

::: {.panel}
Error 3 is Day 1 Session 2's fabricated-robustness failure, in a new place.
**This is what to watch for: confident claims about checks that no one
performed.**
:::

## The same question, a better prompt {.dense}

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people usually type**

> Explain this regression table to a non-technical decision-maker and say what
> should be checked before believing the result.

The model may answer confidently, including about checks it cannot know.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> We have a difference-in-differences table. The data has **two waves only, one
> before and one after**, so no pre-trend test is possible. The decision rule is
> 1,000 AED. Explain what the table supports and, separately, list what it
> cannot tell us. **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Once the AI knows there are only two waves, does it still claim that parallel
trends were satisfied? Paste the second answer into a new chat and ask the AI to
check that answer for claims about tests that were never run.
:::

::: {.panel .fragment}
**If you have time, a prompt that pushes the AI towards an answer:** *"Our DiD
shows an `r gfmt(abs(did_est))` AED cut in costs (p<0.001). Write two sentences for the minister
recommending scale-up."* Does the AI mention the 1,000 AED rule or parallel
trends without being asked?
:::

# 6 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Your turn: ask AI first

::: {.ask style="font-size: 110%;"}
Before we show our list of questions to ask someone presenting DiD, ask an AI tool:

**"I am a commissioner about to hear an evaluator present a difference-in-differences (DiD) result. What are the key questions I should ask?"**
:::

::: {.panel}
Three minutes. Then compare with our list on the next slide: what did it miss, what did it add, and is anything incorrect?
:::

::: {.notes}
OPTIONAL. If time is short, skip this slide and go straight to the questions
on the next one. If you run it: give three minutes, then reveal the list and
ask what the AI missed, what it added that is worth keeping, and whether
anything it suggested was incorrect.
:::

## Three questions to ask any evaluator presenting DiD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Which row is the estimate? Does its whole confidence interval meet our decision rule, not just exclude zero?</div>
  <div class="k">2</div> <div class="v">How many periods of data before the programme? Were the two groups changing in the same way before the programme?</div>
  <div class="k">3</div> <div class="v">Which comparison group was used, how was it chosen, and how did it differ before the programme?</div>
</div>
```

::: {.panel}
Question 1 you learned on Day 1. Questions 2 and 3 are the ones specific to DiD.
:::

::: {.notes}
Paper (Oct13_session1_materials.docx): Hand out the take-away cards (pack page 4, cut into cards). Give a minute to fill in "Back at work", then take two or
three examples.
:::

## Where this sits

```{r results='asis'}
cat(sprintf('
<div class="facts">
  <div class="k">Day 1 S1</div> <div class="v">compared to what? &middot; the three questions, on landfill</div>
  <div class="k">Day 1 S2</div> <div class="v">before and after &middot; %s AED, short of the rule, with no comparison group</div>
  <div class="k">Day 1 S3</div> <div class="v">with and without, and the lottery &middot; %s AED</div>
  <div class="k">Today</div>  <div class="v">DiD &middot; %s AED, with an assumption the waiting list shows failing</div>
</div>', gfmt(naive_ba), gfmt(rct_est), gfmt(did_hand)))
```

::: {.ask}
Every estimate so far has been statistically clear. Only the lottery came close
to the 1,000 AED bar, and today's DiD fell short partly because its comparison
group was on a different path.

What does that mean for the way evaluations get commissioned? Bring your answer
to the next session, where the comparison comes from a **cut-off**.
:::

::: {.notes}
If there is time at the very end, ask the room: is there a different
question you would ask AI to help you understand a concept from today? For
example: "I am learning about impact evaluation to help me make policy
decisions. How should I think about a confidence interval, and how is it
relevant when I compare an estimate to a decision threshold?" Adding context
about who you are and why you are asking usually gets a more useful answer.
:::
