---
title: "Compared to What? Know the Language First"
subtitle: "Module 2 · Day 1, Session 1: Five Words for Reading Any Result"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html. Static knitr chunks work as before.
#
# The simple GreenWaste case (SIMPLE_GREENWASTE_PLAN.md), the one case of the
# whole module. This session introduces it and practises the five words on
# LANDFILL (tonnes per business per year, the rest of the city), so it does not
# give away the cost results of Sessions 2 and 3. Costs are not shown here.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(knitr)
library(qrcode)

gfmt  <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = ",")
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_s1.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- The numbers the static slides quote -----------------------------------
# greenwaste_case.R is sourced into its own environment (it defines RULE).
# This session uses only the rest of the city and its landfill columns.
case_env <- new.env()
case_env$gw <- read.csv("./evaluation_data_GreenWaste_simple.csv")
sys.source("greenwaste_case.R", envir = case_env)
city <- case_env$city
took <- case_env$took
city$change <- city$landfill_after - city$landfill_before   # landfill, tonnes (case_env keeps costs)

# The headline: businesses that took part, before and after.
lf_before <- mean(city$landfill_before[took])
lf_after  <- mean(city$landfill_after[took])
lf_pct    <- 100 * (lf_after / lf_before - 1)
rest_pct  <- 100 * (mean(city$landfill_after[!took]) / mean(city$landfill_before[!took]) - 1)

# Ten businesses from the rest of the city, chosen once by ID so the room can
# do the arithmetic by hand (whole tonnes; mean 10 with the large one, 6
# without). The last one is a large business that pulls the mean above the
# median. The same ten are on the worksheet (make_session_materials.R).
TEN <- c("B02136", "B00755", "B03791", "B05600", "B01274",
         "B07507", "B01673", "B04954", "B08013", "B03836")
ten <- subset(city, business %in% TEN)
ten <- ten[order(ten$landfill_before), ]
ten_x <- ten$landfill_before
stopifnot(nrow(ten) == 10, ten$business[10] == "B03836", mean(ten_x) > median(ten_x))

# The 2x2 for Word 2: average landfill, took part or not, before and after.
r2 <- function(x) round(x, 2)   # two decimals keep the hand arithmetic exact
lf_t0 <- r2(mean(city$landfill_before[took]));  lf_t1 <- r2(mean(city$landfill_after[took]))
lf_o0 <- r2(mean(city$landfill_before[!took])); lf_o1 <- r2(mean(city$landfill_after[!took]))
lf_effect <- (lf_t1 - lf_t0) - (lf_o1 - lf_o0)

# Word 3: the regression on every business.
fit   <- lm(change ~ took_part, data = city)
fit_b <- coef(fit)[["took_part"]]
fit_ci <- confint(fit)["took_part", ]
fit2  <- lm(change ~ took_part + staff + area, data = city)
area_b <- coef(fit2)[["area"]]

# Word 4: the same draws as the live cells (same seeds, same code).
study <- function(n, seed = 2) {
  set.seed(seed)
  city[c(sample(which(city$took_part == 1), n / 2), sample(which(city$took_part == 0), n / 2)), ]
}
gap <- function(d) mean(d$change[d$took_part == 1]) - mean(d$change[d$took_part == 0])
small <- study(20)
set.seed(1)
small_sh <- replicate(1000, { e <- small; e$took_part <- sample(small$took_part); gap(e) })
small_p  <- mean(abs(small_sh) >= abs(gap(small)))

tiny <- city
tiny$change <- tiny$change + tiny$took_part * (-gap(city) - 0.1)
tiny_p <- format.pval(summary(lm(change ~ took_part, data = tiny))$coefficients["took_part", 4],
                      digits = 2, eps = 0.001)

detect_share <- function(n) {
  set.seed(3)
  mean(replicate(200, t.test(change ~ took_part, data = study(n, seed = sample(1e6, 1)))$p.value) < 0.05)
}
det20 <- detect_share(20); det6 <- detect_share(6)

# ---- The case introduction: one illustrative city -----------------------------
# Shop position, efficiency score, and whether it joined. Inside the pilot
# district every score is below 58 and a lottery decided; everywhere else a
# score of 58 or below meant joining. Illustrative, not the real counts.
SHOP_BLUE <- "#215a9e"; SHOP_GREY <- "#c9d6e6"
pilot_shops <- data.frame(x = c(70, 165, 260, 70, 165, 260), y = c(98, 98, 98, 186, 186, 186),
                          score = c(41, 52, 37, 48, 33, 55), joined = c(1, 0, 1, 0, 0, 1))
city_shops  <- data.frame(
  x = c(415, 525, 635, 745, 415, 525, 635, 745, 70, 180, 290, 415, 525, 635, 745),
  y = c(80, 80, 80, 80, 195, 195, 195, 195, 330, 330, 330, 330, 330, 330, 330),
  score = c(72, 45, 88, 39, 56, 63, 29, 91, 47, 77, 52, 68, 35, 84, 58))
city_shops$joined <- as.integer(city_shops$score <= 58)
shop_svg <- function(d) paste(sprintf(
  '<use href="#shop" x="%d" y="%d" width="62" height="58" color="%s"/><text x="%d" y="%d" font-size="19" text-anchor="middle" fill="%s" font-weight="%s">%d</text>',
  d$x, d$y, ifelse(d$joined == 1, SHOP_BLUE, SHOP_GREY), d$x + 31, d$y + 80,
  ifelse(d$joined == 1, "#063360", "#545860"), ifelse(d$joined == 1, "700", "400"), d$score), collapse = "\n")
```

# 0 · Where Module 1 left us {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The word Module 1 was built on

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Every live cell below uses
# what is defined here, so each one on a slide stays two or three lines long.
gw   <- read.csv("evaluation_data_GreenWaste_simple.csv")
city <- subset(gw, setting == "city")                        # the rest of the city
city$change <- city$landfill_after - city$landfill_before    # landfill, tonnes per year
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)

TEN <- c("B02136", "B00755", "B03791", "B05600", "B01274",
         "B07507", "B01673", "B04954", "B08013", "B03836")
ten <- subset(city, business %in% TEN)

# Bars with the value printed on each (handles negative values).
bars <- function(v, col = c(ACCENT, ACCENT2, ALERT)) {
  clean_par(); par(mar = c(2.5, 1, 1.5, 1))
  b <- barplot(v, col = col, border = NA, axes = FALSE,
               ylim = range(c(0, v)) + c(-1.2, 1.2))
  abline(h = 0, col = MUTED)
  text(b, v, sprintf("%.2f t", v), pos = ifelse(v < 0, 1, 3), cex = 1.2, xpd = NA)
}

# Dots for a handful of values, with the mean (red) and the median (dashed).
dots <- function(x) {
  clean_par()
  stripchart(x, method = "stack", pch = 19, cex = 2.2, col = ACCENT, offset = 0.7,
             xlim = c(0, 50), xlab = "Landfill before, tonnes per year", at = 0.2, ylim = c(0, 1.2))
  abline(v = mean(x), col = ALERT, lwd = 3)
  abline(v = median(x), col = MUTED, lwd = 3, lty = 2)
  text(mean(x), 0.95, paste("mean", round(mean(x), 1)), col = ALERT, pos = 4)
  text(median(x), 0.95, paste("median", median(x)), col = MUTED, pos = 2)
}

# A study of n businesses (half took part), always the same draw for a given n.
study <- function(n, seed = 2) {
  set.seed(seed)
  city[c(sample(which(city$took_part == 1), n / 2), sample(which(city$took_part == 0), n / 2)), ]
}
gap <- function(d) mean(d$change[d$took_part == 1]) - mean(d$change[d$took_part == 0])

# Shuffle the took_part labels k times and count how often chance alone beats
# the gap we measured.
shuffle_plot <- function(d, k) {
  real <- gap(d)
  set.seed(1)
  sh <- replicate(max(k, 1), { e <- d; e$took_part <- sample(d$took_part); gap(e) })
  if (k == 0) sh <- numeric(0)
  h <- hist(sh, breaks = seq(-8, 8, by = 0.2), plot = FALSE)
  clean_par()
  hist(sh, breaks = seq(-8, 8, by = 0.2), col = ACCENT2, border = "white",
       main = "", xlab = "Gap in landfill change, shuffled labels (tonnes)",
       ylab = "Shuffles", xlim = c(-3, 3), ylim = c(0, max(12, h$counts)))
  abline(v = c(real, -real), col = ALERT, lwd = 3, lty = c(1, 2))
  cat(sprintf("Measured gap: %.2f tonnes.  Shuffles: %d.  As far from zero by chance: %d  ->  p = %s\n",
              real, k, sum(abs(sh) >= abs(real)),
              if (k == 0) "?" else format(round(mean(abs(sh) >= abs(real)), 3))))
}

# Estimate and 95% interval against zero, one line.
one_interval <- function(est, lo, hi, xlim = c(-3.5, 0.5)) {
  clean_par()
  plot(NA, xlim = xlim, ylim = c(0.5, 1.5), yaxt = "n", bty = "n",
       xlab = "Effect on landfill (tonnes per business per year)", ylab = "")
  abline(v = 0, col = MUTED, lwd = 2, lty = 2)
  segments(lo, 1, hi, 1, col = ACCENT, lwd = 8)
  points(est, 1, pch = 19, cex = 2.5, col = ACCENT)
}

# "Suppose GreenWaste had cut landfill by only `cut_by` tonnes": shift the
# changes of the businesses that took part so the effect is exactly that, keep
# every business, re-test.
significance_demo <- function(cut_by) {
  what_if <- city
  what_if$change <- what_if$change + what_if$took_part * (-gap(city) - cut_by)
  f  <- lm(change ~ took_part, data = what_if)
  ci <- confint(f)["took_part", ]
  p  <- summary(f)$coefficients["took_part", 4]
  cat(sprintf("All %s businesses. Effect %.2f tonnes, 95%% interval %.2f to %.2f, p = %s\n",
              format(nrow(city), big.mark = ","), coef(f)[["took_part"]], ci[1], ci[2],
              format.pval(p, digits = 2, eps = 0.001)))
  one_interval(coef(f)[["took_part"]], ci[1], ci[2])
}

# 100 studies of n businesses each, with their intervals. Red: misses the value
# we get from every business in the rest of the city.
many_studies <- function(n = 40, runs = 100, seed = 1) {
  truth <- gap(city)
  set.seed(seed)
  res <- t(replicate(runs, {
    d <- city[c(sample(which(city$took_part == 1), n / 2), sample(which(city$took_part == 0), n / 2)), ]
    f <- lm(change ~ took_part, data = d)
    c(coef(f)[["took_part"]], confint(f)["took_part", ])
  }))
  res  <- res[order(res[, 1]), ]
  miss <- res[, 2] > truth | res[, 3] < truth
  clean_par()
  plot(NA, xlim = c(-6, 1), ylim = c(1, runs), yaxt = "n", bty = "n",
       xlab = "Effect on landfill (tonnes per business), one line per study", ylab = "")
  segments(res[, 2], 1:runs, res[, 3], 1:runs, col = ifelse(miss, ALERT, ACCENT2), lwd = 2)
  points(res[, 1], 1:runs, pch = 19, cex = 0.5, col = ifelse(miss, ALERT, ACCENT))
  abline(v = truth, col = MUTED, lwd = 2)
  cat(sprintf("%d studies of %d businesses. %d intervals contain the all-businesses value (%.2f); %d miss it.\n",
              runs, n, sum(!miss), truth, sum(miss)))
}
```

::: {.ask}
From Module 1: what is a **counterfactual**?
:::

::: {.panel .fragment}
What would have happened to the same people, over the same period, without the
programme. We never observe it. Every evaluation method estimates it with a
**comparison**, and a result can only be read once you know what that
comparison was.
:::

::: {.panel .fragment}
Today we attach five words to that idea, using one programme and its data from
start to finish.
:::

::: {.notes}
About 3 minutes. Recall only; Module 1 taught this. Open the deck a couple of minutes before the session starts: R takes about a minute to load in the browser, after which every live cell runs in a few seconds.
:::

# 1 · What are we comparing? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The GreenWaste programme

<p style="font-size:0.85em; margin: 0 0 4px; max-width: none;">GreenWaste helps businesses cut their waste costs. Here is how it reached them.</p>

```{r city-map, results='asis'}
html_out(paste0('
<svg viewBox="0 0 1200 470" width="100%" role="img"
     aria-label="A city map. In the pilot district a lottery chose which shops joined. In the rest of the city, shops with an efficiency score of 58 or below joined."
     style="font-family: \'Noto Kufi Arabic\', \'Segoe UI\', Arial, sans-serif;">
  <defs>
    <symbol id="shop" viewBox="0 0 60 56">
      <path d="M4 14 L10 2 H50 L56 14 Z" fill="currentColor"/>
      <path d="M4 14 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 12 0" fill="currentColor"/>
      <rect x="8" y="20" width="44" height="34" rx="2" fill="currentColor"/>
      <rect x="14" y="30" width="12" height="24" fill="#fff" opacity="0.85"/>
      <rect x="32" y="28" width="14" height="12" fill="#fff" opacity="0.85"/>
    </symbol>
  </defs>
  <rect x="20" y="20" width="820" height="440" rx="18" fill="#f4f7fb" stroke="#D5DEE8" stroke-width="2"/>
  <path d="M370 20 V460 M20 300 H840" stroke="#e3e9f1" stroke-width="10"/>
  <text x="820" y="450" font-size="22" font-weight="700" fill="#7f97b5" text-anchor="end">THE CITY</text>
  <rect x="45" y="45" width="305" height="245" rx="14" fill="#e3ebf5" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
  <text x="62" y="80" font-size="22" font-weight="700" fill="#063360">Pilot district</text>
  <g transform="translate(300,55)">
    <rect width="38" height="38" rx="7" fill="#fff" stroke="#063360" stroke-width="2.5"/>
    <circle cx="11" cy="11" r="3.5" fill="#063360"/><circle cx="27" cy="11" r="3.5" fill="#063360"/>
    <circle cx="19" cy="19" r="3.5" fill="#063360"/>
    <circle cx="11" cy="27" r="3.5" fill="#063360"/><circle cx="27" cy="27" r="3.5" fill="#063360"/>
  </g>
', shop_svg(pilot_shops), '
', shop_svg(city_shops), '
  <g font-size="21" fill="#111418">
    <rect x="875" y="40" width="300" height="150" rx="12" fill="#fff" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
    <text x="895" y="82" font-weight="700" fill="#063360">In the pilot district</text>
    <text x="895" y="118">a <tspan font-weight="700">lottery</tspan> chose which</text>
    <text x="895" y="148">shops joined</text>
    <rect x="875" y="215" width="300" height="178" rx="12" fill="#fff" stroke="#D5DEE8" stroke-width="2"/>
    <text x="895" y="257" font-weight="700" fill="#063360">Everywhere else</text>
    <text x="895" y="293">shops scoring <tspan font-weight="700">58 or below</tspan></text>
    <text x="895" y="323">joined</text>
    <text x="895" y="356" font-size="16" fill="#545860">the number under each shop</text>
    <text x="895" y="377" font-size="16" fill="#545860">is its efficiency score</text>
  </g>
  <g font-size="19" fill="#111418">
    <use href="#shop" x="878" y="404" width="30" height="28" color="', SHOP_BLUE, '"/><text x="918" y="425">joined GreenWaste</text>
    <use href="#shop" x="878" y="436" width="30" height="28" color="', SHOP_GREY, '"/><text x="918" y="457">did not join</text>
  </g>
</svg>'))
```

::: {.notes}
About 3 minutes. One city; the pilot district is part of it. Point at the pilot
district: every shop there scores below 58, yet only some joined, because a
lottery decided. Then the rest of the city: the number under each shop decided.
Ask: "Where would it be fairer to compare shops that joined with shops that did
not: inside the pilot district or in the rest of the city? Why?" Take answers;
do not resolve it yet. Session 3 comes back to it. The map is illustrative, and
the case is invented for this training.

Today stays in the rest of the city and looks at one thing GreenWaste changed:
the tonnes of waste each business sends to landfill in a year. The waste costs,
on which the decision rests, open in Session 2.

Paper: hand out the GreenWaste picture card (GreenWaste_case_card.docx, one per participant).
:::

## The headline

```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%s%%</div>
  <div class="label">Landfill waste, businesses that took part, before to 12 months after</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">Tonnes per business, before</span></div>
  <div><span class="n">%s</span><span class="k">Tonnes per business, after</span></div>
</div>', gfmt(lf_pct), gfmt(lf_before, 1), gfmt(lf_after, 1)))
```

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask}
**`[MENTI PLACEHOLDER]`** On this number alone, would you extend GreenWaste?

Go to **menti.com**, code **`r MENTI_CODE`**, or scan the code.
:::
:::::

::::: {.column width="38%"}
```{r out.width='230px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

## What if every business cut its landfill anyway?

```{ojs}
//| echo: false
viewof anyway = Inputs.range([0, 60], {value: 0, step: 5, label: "Every business cut its landfill anyway (%)"})
```

```{webr}
#| input:
#|   - anyway
#| autorun: true
#| edit: false
before <- mean(city$landfill_before[city$took_part == 1])
after  <- mean(city$landfill_after[city$took_part == 1])
without <- before * (1 - anyway / 100)    # where they would have ended without GreenWaste
round(data.frame(before, without, after, fall_left_for_GreenWaste_pct = 100 * (after / without - 1)), 1)
```

::: {.ask}
Landfill fell `r gfmt(abs(lf_pct))`%. Slide "anyway" up to `r gfmt(round(abs(lf_pct) / 5) * 5)`%.
**How much of the fall is left for GreenWaste?**
:::

::: {.panel .fragment}
Almost none. If every business cut its landfill that much anyway, the headline
would look the same with or without GreenWaste. The headline alone cannot tell
the two apart.
:::

## The businesses that did not take part

```{webr}
lf <- aggregate(cbind(landfill_before, landfill_after) ~ took_part, data = city, FUN = mean)
lf$change_pct <- 100 * (lf$landfill_after / lf$landfill_before - 1)
round(lf, 1)
```

::: {.panel}
Businesses in the rest of the city that did **not** take part cut their landfill
by `r gfmt(abs(rest_pct))`% too, without GreenWaste. Part of the fall would have
happened anyway.
:::

::: {.ask .fragment}
Those businesses all scored above 58. Is their `r gfmt(abs(rest_pct))`% what
would have happened to the businesses that took part? What would you want to
know first?
:::

::: {.notes}
Ask for a prediction before pressing Run: did landfill fall for the businesses
that did not take part, and by how much? Then take answers to the question. If
someone says they are different businesses (more efficient, larger), say that
is exactly where Session 3 goes, and leave it there. Then a show of hands: who
would now change their vote from the headline slide? Take one reason from
someone who changed and one from someone who did not.
:::

## "Compared to what?"

::: {.panel}
Every result is a comparison. A before-and-after number compares the programme
with **its own past**. What else changed over the same period (prices, rules on
waste, a city campaign) is inside that number, and it can push the answer
either way.
:::

::: {.ask}
Think of a figure you reported, or were sent, this year. **What was it compared
to?** If nobody said, what would you have picked?
:::

# 2 · Three questions for any number {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## One idea, three questions

::: {.panel}
The five words that follow each answer one of three questions you can ask of
any number an evaluator hands you.
:::

```{=html}
<div class="facts">
  <div class="k">Compared to what?</div>             <div class="v">Comparison group, counterfactual</div>
  <div class="k">How big, in units that matter?</div> <div class="v">Mean · treatment effect · coefficient</div>
  <div class="k">How sure are we?</div>              <div class="v">p-value · confidence interval</div>
</div>
```

::: {.panel}
We come back to this list at the end of the session, and on every day of the
week.
:::

# 3 · The five words {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Word 1 · Mean: ten businesses, by hand

:::::: {.columns}

::::: {.column width="42%"}
```{r}
data.frame(Business = ten$business, Staff = ten$staff,
           `Landfill (tonnes)` = ten_x, check.names = FALSE) |>
  kable(align = "lrr", row.names = FALSE)
```
:::::

::::: {.column width="58%"}
::: {.ask}
Ten businesses in the rest of the city, before the programme. With pen and
paper: **what is the average landfill per business?** Then: is a typical
business close to it?
:::

::: {.panel .fragment}
The mean is **`r gfmt(mean(ten_x), 1)`** tonnes.
`r sum(ten_x < mean(ten_x))` of the ten businesses are below it. One business
with `r ten$staff[10]` staff sends `r max(ten_x)` tonnes and pulls the mean up;
the middle business (the **median**) sends `r median(ten_x)`.
:::
:::::

::::::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

::: {.notes}
Paper (Oct12_session1_materials.docx): Hand out the ten-businesses worksheet (pack page 1, one per participant; the landfill table is on the back). Give two minutes before the reveal.
:::

## Word 1 · What the mean hides

::::::: {.columns}

:::::: {.column width="45%"}
```{ojs}
//| echo: false
viewof keep = Inputs.radio(["All ten businesses", "Without the largest"], {value: "All ten businesses"})
```
::::::

:::::: {.column width="55%"}
::: {.ask}
Remove one business. How far does the mean move, and how far does the median
move?
:::
::::::

:::::::

```{webr}
#| input:
#|   - keep
#| autorun: true
#| fig-width: 10
#| fig-height: 1.9
tonnes <- ten$landfill_before
if (keep == "Without the largest") tonnes <- tonnes[tonnes < max(tonnes)]
dots(tonnes)
```

::: {.panel .fragment}
The mean drops from `r gfmt(mean(ten_x), 1)` to `r gfmt(mean(ten_x[-10]), 1)`;
the median from `r median(ten_x)` to `r median(ten_x[-10])`. One business moved
the mean further than the median. When an evaluator reports an average, ask what
the spread looks like and whether a few cases drive it.
:::

## Word 2 · Treatment effect

```{r results='asis'}
html_out(sprintf('
<table>
  <thead><tr><th>Average landfill (tonnes per year)</th><th>Before</th><th>After</th><th>Change</th></tr></thead>
  <tbody>
    <tr><td>Took part</td><td>%.2f</td><td>%.2f</td><td class="fragment" data-fragment-index="1">%.2f</td></tr>
    <tr><td>Did not take part</td><td>%.2f</td><td>%.2f</td><td class="fragment" data-fragment-index="1">%.2f</td></tr>
    <tr class="fragment" data-fragment-index="1"><td><strong>Difference</strong></td><td>%.2f</td><td>%.2f</td><td><strong>%.2f</strong></td></tr>
  </tbody>
</table>',
  lf_t0, lf_t1, lf_t1 - lf_t0, lf_o0, lf_o1, lf_o1 - lf_o0,
  lf_t0 - lf_o0, lf_t1 - lf_o1, lf_effect))
```

::: {.ask}
Three numbers describe this result: **`r sprintf("%.2f", lf_t1 - lf_t0)`**,
**`r sprintf("%.2f", lf_t1 - lf_o1)`** and **`r sprintf("%.2f", lf_t0 - lf_o0)`**.
Work out what each one is, then decide which is the effect of GreenWaste on landfill.
:::

::: {.panel .fragment fragment-index=2}
**`r sprintf("%.2f", lf_t0 - lf_o0)`** is the head start: the businesses that took
part sent less to landfill before anything happened, and
**`r sprintf("%.2f", lf_t1 - lf_o1)`** still carries it. The **treatment effect**
is the change for those that took part minus the change for the rest:
**`r sprintf("%.2f", lf_effect)`** tonnes. It is smaller than their own fall
because the rest fell too.
:::

::: {.warn .note-ctr}
**Illustrative case, with invented figures.**
:::

::: {.notes}
Paper (Oct12_session1_materials.docx): Participants turn the worksheet over: part 2, the landfill table. They fill the blank cells before you reveal them.
:::

## Word 2 · The same number from `r gfmt(nrow(city))` businesses

```{webr}
took   <- mean(city$change[city$took_part == 1])
others <- mean(city$change[city$took_part == 0])
round(data.frame(took_part = took, did_not = others, effect = took - others), 2)
```

::: {.panel}
`city` holds one row per business in the rest of the city; `change` is its
landfill after minus before. The three lines are the arithmetic you just did,
on every business.
:::

## Word 2 · The effect depends on the comparison

```{ojs}
//| echo: false
viewof other = Inputs.range([-4, 1], {value: 0, step: 0.1, label: "Landfill change, businesses that did not take part (tonnes)"})
```

```{webr}
#| input:
#|   - other
#| autorun: true
#| edit: false
#| fig-width: 7
#| fig-height: 2.6
took <- mean(city$change[city$took_part == 1])
bars(c("Took part" = took, "Did not" = other, "Effect" = took - other))
```

::: {.ask}
The businesses that took part are fixed at `r sprintf("%.1f", lf_t1 - lf_t0)`.
Move the others. **Which value makes the effect zero, and what would that mean
in plain words?**
:::

## Word 3 · Coefficient

```{webr}
fit <- lm(change ~ took_part, data = city)
round(data.frame(coefficient = coef(fit)), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
`took_part` is 1 for businesses that took part, 0 for the rest. A
**coefficient** says how much the outcome moves per one unit of the row. Here
one unit is "took part".
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Which of the two numbers is the effect? What is the other one?
:::

::: {.panel .fragment}
`took_part` is **`r sprintf("%.2f", fit_b)`**, the effect you worked out by hand.
`(Intercept)` is the average change for businesses that did not take part. It is
not an effect.
:::
:::::

::::::

## Word 3 · The other rows of a table

```{webr}
fit2 <- lm(change ~ took_part + staff + area, data = city)
round(as.data.frame(summary(fit2)$coefficients), 3)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Std. Error** is how much the estimate would move from one sample of
businesses to another. Small relative to the estimate means a steady estimate.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The other rows adjust the comparison. `area` is premises in 100 m²: each extra
100 m² goes with a change of `r sprintf("%.2f", area_b)` tonnes. That is not
advice to move businesses into different premises.
:::
:::::

::::::

## Word 4 · p-value: shuffle the labels

```{ojs}
//| echo: false
viewof k = Inputs.range([0, 1000], {value: 0, step: 50, label: "Times we shuffle the took_part labels"})
```

```{webr}
#| input:
#|   - k
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.7
small <- study(20)          # a study of 20 businesses: 10 took part, 10 did not
shuffle_plot(small, k)
```

::: {.ask}
If GreenWaste did nothing, the labels would be arbitrary, so we shuffle them.
**How often does chance alone give a gap as large as the red line?** Guess first.
:::

::: {.panel .fragment}
That share is the **p-value**: here `r small_p`, `r round(small_p * 1000)` shuffles
in a thousand.
:::

::: {.notes}
Start at 0 shuffles; the red line is this 20-business study's gap
(`r sprintf("%.2f", gap(small))` tonnes; all `r gfmt(nrow(city))` businesses give
`r sprintf("%.2f", fit_b)`). Slide up. At 1,000 shuffles p is `r small_p`. The
next slide asks whether a small p means a big effect.
:::

## Word 4 · Significant is not the same as big

```{ojs}
//| echo: false
viewof cut_by = Inputs.range([0.05, 2.7], {value: 2.7, step: 0.05, label: "Suppose GreenWaste cut landfill by only (tonnes)"})
```

```{webr}
#| input:
#|   - cut_by
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.7
significance_demo(cut_by)
```

::: {.ask}
Slide the effect down to 0.1 tonnes: 100 kg a year, about 2 kg a week. **Is it
still statistically significant?**
:::

::: {.warn .fragment}
Yes, p is `r tiny_p`. With `r gfmt(nrow(city))` businesses even a trivial change
passes the test. Whether 100 kg a year is worth paying for is a separate
question, and the p-value cannot answer it.
:::

## Word 4 · The warning cuts both ways

```{ojs}
//| echo: false
viewof biz_n = Inputs.range([6, 200], {value: 20, step: 2, label: "Businesses in the study"})
```

```{webr}
#| input:
#|   - biz_n
#| autorun: true
#| edit: false
set.seed(3)
p <- replicate(200, t.test(change ~ took_part, data = study(biz_n, seed = sample(1e6, 1)))$p.value)
data.frame(businesses_in_study = biz_n, share_of_studies_that_detect_it = round(mean(p < 0.05), 2))
```

::: {.ask}
The real effect never changes. Slide down to 6 businesses: **what share detect it?**
:::

::: {.panel .fragment}
At 20 businesses, `r round(100 * det20)` studies in 100 detect it; at 6, only `r round(100 * det6)`.
:::

::: {.ask .fragment}
An evaluator reports "no significant effect". Ask: **how many businesses were in
the study, and what is the smallest effect it could detect?**
:::

## Word 5 · Confidence interval

```{ojs}
//| echo: false
viewof run = Inputs.button("Run another 100 studies")
```

```{webr}
#| input:
#|   - run
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 2.4
many_studies(n = 40, runs = 100, seed = run + 1)
```

::: {.ask}
Each line is one study of 40 businesses and its **95% confidence interval**, the
range of values consistent with its data. **Out of 100, how many will miss the
grey line?** Guess before the next press.
:::

::: {.notes}
About 95 of 100 lines cross the grey line (the value from every business in the
rest of the city), and about 5 miss (red). That is what "95%" refers to: the
method, over many studies, not the probability for any one interval. Press the
button two or three times.
:::

## Word 5 · Which one would you act on?

::: {.panel}
Suppose three evaluations of GreenWaste each reported **1.8 tonnes less landfill
per business per year**, and the city's target is **2.0 tonnes**. Only the
intervals differ.
:::

```{=html}
<div class="facts">
  <div class="k">A</div> <div class="v">1.8 tonnes less, 95% interval 0.4 to 3.2</div>
  <div class="k">B</div> <div class="v">1.8 tonnes less, 95% interval 1.6 to 2.0</div>
  <div class="k">C</div> <div class="v">1.8 tonnes less, 95% interval -0.9 to 4.5</div>
</div>
```

::: {.ask}
Which would you act on? What is your answer for each?
:::

::: {.panel .fragment}
B is precise and below the target: confidently not enough. A is too uncertain
to call, and C more so. All three have the same estimate, so the intervals make
the difference.
:::

::: {.notes}
These three results and the 2.0-tonne target are invented for this exercise;
they are not the programme's result. GreenWaste's own decision rests on waste
costs, which the rest of the week estimates from the data.

Paper (Oct12_session1_materials.docx): Part 3 of the worksheet: everyone ticks act / do not act / cannot tell yet for A, B and C before the reveal.
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI prompt cards

:::::: {.columns}

::::: {.column width="45%"}
::: {.panel}
Each group takes one card. Paste its prompt into your AI tool, read the answer
together, and mark anything it gets wrong or overstates.
:::

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Card 1</div> <div class="v">"Explain a 95%% confidence interval, in one paragraph, for someone who is not a statistician."</div>
  <div class="k">Card 2</div> <div class="v">"Explain what a p-value means."</div>
  <div class="k">Card 3</div> <div class="v">"Landfill waste from businesses in our programme fell %s%%. Is that a good result?"</div>
  <div class="k">Card 4</div> <div class="v">"The coefficient on area is %s. What should I do about it?"</div>
</div>', gfmt(abs(lf_pct)), sprintf("%.2f", area_b)))
```
:::::

::::: {.column width="55%"}
::: {.ai-output}
*Card 1. A response like this is possible:*

"A 95% confidence interval is the range in which the true value is likely to
fall. [There is a 95% probability that the true effect lies inside this
interval]{.flag}, so if the interval is `r sprintf("%.2f", fit_ci[1])` to
`r sprintf("%.2f", fit_ci[2])` tonnes the true effect is probably in that range.
A narrower interval means a more precise estimate."
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session1_materials.docx): Hand out one set of the four prompt cards per group (pack page 3, cut); each group takes one card.
:::

## Debrief

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Card 1.** The flagged sentence is the common error. "95%" describes the method:
over many studies, 95 of 100 intervals catch the true value, as the chart of 100
studies showed. It is not a probability for this one interval.
:::

::: {.warn}
**Card 3.** Did the AI ask what the `r gfmt(abs(lf_pct))`% was compared with, or
what happened to businesses outside the programme? The prompt gave it neither.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Card 4.** Look for advice about premises. The row adjusts the comparison; it is
not a policy lever.
:::

::: {.panel}
Two habits: ask the AI to **ask you questions before it answers**, and put its
answer through a second chat to check it. Also note what it explained well.
:::
:::::

::::::

# 5 · Triage the claims {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Triage the claims

::: {.excerpt}
**1.** "Landfill waste from businesses in GreenWaste fell `r gfmt(abs(lf_pct))`% in a year." \
**2.** "The effect of GreenWaste on landfill is `r sprintf("%.2f", fit_b)` tonnes per business, 95% confidence interval `r sprintf("%.2f", fit_ci[1])` to `r sprintf("%.2f", fit_ci[2])`." \
**3.** "The fall in landfill is statistically significant (p < 0.001), so every business in the country should join GreenWaste." \
**4.** "Landfill fell `r sprintf("%.1f", abs(lf_t1 - lf_t0))` tonnes per business among those that took part and `r sprintf("%.1f", abs(lf_o1 - lf_o0))` tonnes among those that did not."
:::

::: {.ask}
In groups: for each excerpt, **which of the three questions does the excerpt not
answer? Would you act, not act, or ask a question first?** Four minutes. Then:
which excerpt would you most want to act on?
:::

::: {.notes}
Key: 1 compared to what (ask first; the businesses that did not take part fell
`r gfmt(abs(rest_pct))`% too); 2 how big in units that matter, landfill not the
waste costs the decision rests on (ask first); 3 significance is not a decision
rule (do not act); 4 how sure, no interval (ask first). Groups usually split on
2 and 4: take one group's reasoning for each. None of the four earns "act": ask
what an excerpt would need to get there (a comparison, an effect and a target,
as in "Landfill fell `r sprintf("%.1f", abs(lf_effect))` tonnes per business
more than among businesses that did not take part; the target was at least 2
tonnes").

Paper (Oct12_session1_materials.docx): Hand out the triage sheet (pack page 4, one per group) and one set of the four excerpt slips per group (page 5, cut).

Optional wall matrix (set up beforehand; signs at the end of the pack): columns Act / Ask first / Do not act, rows Compared to what? / How big? / How sure? / None. Each group has its own marker colour, draws a stripe across its slips and tapes them into the matrix after filling in the sheet. Ask two groups whose slips for the same excerpt sit in different cells to explain.
:::

# 6 · Take this back to your desk {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The three questions

::: {.panel}
From memory, before you look: what are the three questions you can ask of any
number?
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what?</div>
  <div class="k">2</div> <div class="v">How big is the number, in units that matter to the decision?</div>
  <div class="k">3</div> <div class="v">How sure are we?</div>
</div>
```

::: {.panel}
Next session opens the waste-cost data and asks these three questions of the
costs.
:::

::: {.notes}
If there is time: ask the room for a better way to ask AI about one of today's
words. For example: "I make policy decisions using impact evaluations. How
should I read a confidence interval when I compare an estimate with a decision
threshold? Ask me questions first." Saying who you are and why you ask usually
gets a more useful answer.

Paper (Oct12_session1_materials.docx): Hand out the take-away cards (pack page 6). Give a minute to fill in "Back at work" on the card, then take two or three examples from the room.
:::
