---
title: "The analyst is in the room"
subtitle: "Day 3 | Session 3: Ask for evidence that could change your answer"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day3_case.R')
```

## Would you sign this note? {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
pilot <- subset(gw, setting == 'pilot')
```
```{r, results='asis'}
html_out(paste0('<p class="lead">',qa_note,'</p>',fiction))
```
```{=html}
<div data-choices><button data-choice aria-pressed="false">Sign off</button><button data-choice aria-pressed="false">With conditions</button><button data-choice aria-pressed="false">Request evidence</button><p role="status" class="feedback"></p></div>
```


::: {.notes}
4 minutes. The fictional note intentionally overclaims. Record the opening decision and reason before feedback. Optional course room poll with Before phase and hidden totals; no AI for this task.
:::

## Follow each claim back to its evidence. {.reader-slide}

```{r, results='asis'}
html_out(html_table(method_rows))
```
::: {.ask}
Which result did the note choose, and who does each result describe?
:::


::: {.notes}
5 minutes. Savings are positive and interval endpoints sorted. Pilot and matching intervals span the rule; DiD and RDD fall short. City methods share records and describe different comparisons. Four estimates are not four votes.
:::

## Only three questions. Which ones? {.reader-slide}

```{r, results='asis'}
html_out(paste0('<ol>',paste(sprintf('<li>%s</li>',read_questions),collapse=''),'</ol>'))
```
::: {.ask}
Which three questions would most affect your sign-off decision?
:::


::: {.notes}
6 minutes. Use the five-question QA sheet. Each question must request specific evidence. Individuals choose first, then groups. The full historical question battery is outside the teaching route.
:::

## Will 'is it reliable?' get a useful answer? {.reader-slide}

A reviewer asks: **"Is the evidence robust?"**

Rewrite it to request a specific result or check.

::: {.ask}
What answer would help you decide, and what would count as evidence?
:::
```{=html}
<button data-open="qa-question" data-closed="Open an example" data-opened="Hide feedback" aria-controls="qa-question" aria-expanded="false">Open an example</button><div id="qa-question" class="feedback" hidden>Show the estimate and interval for the analysis you planned to rely on. Does its conclusion change under reasonable alternative choices?</div>
```


::: {.notes}
5 minutes. Ask for a result or check that could change the decision. The example requests the planned analysis; it does not establish that a documented plan exists here.
:::

## Choose before the analyst arrives. {.reader-slide}

Groups rank their questions. Choose a spokesperson.

Write the evidence requested, before asking.

Keep one follow-up question ready.

::: {.ask}
Which answer would you keep pressing for?
:::


::: {.notes}
7 minutes. Eight groups rank questions. Assign comparison, assumptions, data and conclusions as focus areas if helpful. Use the fixed answer bank; say evidence is not supplied when it is absent.
:::

## Take a break {.reader-slide}

Keep your questions ready.

The trainer returns as the analyst.

::: {.notes}
5 minutes. Protect the five-minute break. Keep the questions ready and do not add extra analyses.
:::

## The analyst is here. What will you ask? {.reader-slide}

Ask one specific question per group.

Record the answer and its source on **Sheet 2**.

Follow up when a claim has no supporting evidence.

::: {.ask}
Did the answer address the question you asked?
:::


::: {.notes}
14 minutes. Allow one question and follow-up per group, about ninety seconds each. Answer briefly from the source bank. Ask which evidence is still missing; never invent fieldwork documentation.
:::

## Pilot estimate: 1,014. Enough to act? {.reader-slide}

A reviewer asks whether the pilot settles the saving rule.

Predict where the two interval limits sit.

```{webr}
#| echo: true
#| autorun: false
# Compare annual waste costs in the pilot lottery groups.
fit <- lm(cost_after ~ took_part, data = pilot)
# Find the 95% interval for their cost difference (AED).
limits <- confint(fit)['took_part', ]
# Turn the cost difference into saving; show lower then upper limit.
sort(round(-unname(limits)))
```
::: {.ask}
How does this answer change the recommendation?
:::


::: {.notes}
6 minutes. One trainer demonstration. Output gives the lower and upper 95% interval limits for annual saving in AED. This answers the threshold question; paper answers serve other questions. The pilot cannot establish nationwide reach or full value for money.
:::

## An answer, a partial answer or no answer? {.reader-slide}

For each answer choose:

**Supported / Partial / Unanswered**

Write the source or the evidence still missing.

::: {.ask}
Which answer deserves a supported rating?
:::


::: {.notes}
6 minutes. Recognise strengths as well as missing evidence. The stated pilot comparison and interval are supported; fieldwork quality remains unanswered. Transparent disclosure does not itself resolve the concern.
:::

## What still stands between you and approval? {.reader-slide}

::: {.ask}
Which unresolved issue could most change your decision?
:::
Write the missing evidence and what you would do with it.

Avoid an undirected list of concerns.

::: {.notes}
5 minutes. Work alone then in pairs. Choose a concrete evidence request and explain its decision relevance. Complete generated rows do not establish complete real-world follow-up.
:::

## Now would you sign? {.reader-slide}

```{=html}
<div data-choices><button data-choice aria-pressed="false">Sign off</button><button data-choice aria-pressed="false">With conditions</button><button data-choice aria-pressed="false">Request evidence</button><p role="status" class="feedback"></p></div>
```
::: {.ask}
Which evidence changed your reason, or kept it the same?
:::
Write your final decision before comparing with your opening choice.

::: {.notes}
5 minutes. Optional room poll After phase: close before showing totals. Compare reasons, not a winning vote. Assess written evidence of reading. Different provisional decisions need clear limits.
:::

## The AI says four studies agree. Do they? {.reader-slide}

Illustrative AI draft:

"There are four independent studies. All intervals clear the saving rule. No businesses are missing, so data quality is established."

::: {.ask}
Which statements would you correct using the source?
:::
```{=html}
<button data-open="qa-ai" data-closed="Open feedback" data-opened="Hide feedback" aria-controls="qa-ai" aria-expanded="false">Open feedback</button><div id="qa-ai" class="feedback" hidden>City methods share records and differ in reach. Two intervals span the rule and two fall short. A complete synthetic file does not establish fieldwork quality.</div>
```


::: {.notes}
4 minutes. Authored AI draft, not a captured response. Correct claims using the result table and answer sources. No external AI account is required.
:::

## Ask for evidence you would actually use. {.reader-slide}

::: {.ask}
What specific evidence should the evaluator send next?
:::
Write a short request and explain how it would affect your decision. Work alone.

::: {.notes}
3 minutes. Independent close without AI. Look for a specific source request, its purpose and how it could affect the recommendation. Total 75 minutes.
:::
