---
title: "What Do the Numbers Say? Reading an Evaluation Output"
subtitle: "Module 2 · Day 1, Session 2: Illustrative Case Data"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)   # lm_robust, clustered standard errors
library(qrcode)     # live poll QR, generated below so it cannot drift

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / offered
ACCENT2 <- "#A9561F"   # series 2: comparison / not offered
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

# ---- Chart sizing ----------------------------------------------------------
# Every chart carries an explicit pixel size. Relying on plotly's autosize
# inside revealjs produces half-drawn charts: a widget initialised while its
# slide is hidden measures a narrow container and never corrects itself.
# These must stay in step with the chunk's fig.width / fig.height, which size
# the htmlwidgets wrapper (at dpi 96, one inch is 96 px).

PL_W  <- 1100   # full-width chart
PL_WC <- 520    # chart sitting in a half-width column
PL_WT <- 340    # chart sitting in a third-width column

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width  = width,
      height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. The QR and printed link on the vote slide
# and the results iframe on the reveal slide all read from here: change
# MENTI_CODE only and all three move together.
#
# PLACEHOLDER until this session's Menti is built. MENTI_CODE is the short
# JOIN code participants type at menti.com (usually 8 digits), NOT the long
# ID in the editor URL. Confirm it while the Menti is live: a code that is
# not running sends the room to a "no longer active" page.
MENTI_CODE  <- "1234 5678"   # <!-- PLACEHOLDER: this session's join code -->
MENTI_JOIN  <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

# The results embed: the SLIDES view for a webpage, not the voting view (the
# room still votes on their phones). Set to the long ID from the editor URL.
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/PLACEHOLDER/embed"  # <!-- PLACEHOLDER -->

# Distinct filename so this QR never overwrites Session 1's menti_qr.png.
menti_qr <- function(path = "menti_qr_s2.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))   # ecl H = highest error correction
  path
}
MENTI_QR <- menti_qr()

# ---- Session 2: the traffic camera programme -------------------------------
# One row per road segment per round. 1,200 segments in 24 sectors, two rounds
# (2019 before, 2021 after). Twelve sectors got cameras in 2020.
#
# NOTE ON THE DATA, AND READ THIS BEFORE EDITING ANY SLIDE.
#
# evaluation_data_TrafficCameras.csv is FICTIONAL data generated by
# make_traffic_camera_data.R. It is not an Abu Dhabi Police evaluation and it is
# not any real programme. It was originally drafted as a stand-in for Police
# findings that do not exist in this workspace, and it named the force, which
# attributed fabricated numbers to a real organisation on a public site. The
# case now names no real body.
#
# Every slide that shows a number from it carries a visible disclaimer, per
# CLAUDE.md's placeholder convention.
pol <- read.csv("./evaluation_data_TrafficCameras.csv")

# The bar the authority set before committing to a national roll-out.
POL_RULE <- 2.0     # injury collisions avoided, per segment per round

pol_before <- pol %>% filter(round == 0)
pol_after  <- pol %>% filter(round == 1)

# The headline the authority would put in a press release: treated sectors only,
# before and after, no comparison.
pol_t_before <- mean(pol_before$injury_collisions[pol_before$treated == 1])
pol_t_after  <- mean(pol_after$injury_collisions[pol_after$treated == 1])
pol_headline_fall <- 100 * (pol_t_before - pol_t_after) / pol_t_before

# What happened in the sectors that did not get cameras over the same period.
pol_c_before <- mean(pol_before$injury_collisions[pol_before$treated == 0])
pol_c_after  <- mean(pol_after$injury_collisions[pol_after$treated == 0])

# The honest estimate: the model the deck projects, with errors clustered by
# road segment because the two rounds of one segment are not independent.
pol_fit <- lm_robust(injury_collisions ~ treated * round,
                     data = pol, clusters = segment_id)

pol_did    <- coef(pol_fit)[["treated:round"]]
pol_did_ci <- confint(pol_fit)["treated:round", ]
pol_did_p  <- pol_fit$p.value[["treated:round"]]
pol_avoided_hi <- abs(pol_did_ci[1])   # most the interval will allow

# The table as a participant would receive it. Built here rather than on the
# slide so that the "reading the output" slide and the "annotate this output"
# exercise show the identical table.
pol_tbl <- pol_fit %>%
  tidy() %>%
  mutate(` ` = c("(Intercept)", "Programme sectors", "Round 2 (2021)",
                 "Programme x Round 2"),
         CI = sprintf("[%.2f, %.2f]", conf.low,
                      estimate + (estimate - conf.low))) %>%
  select(` `, estimate, std.error, p.value, CI)

# The two group means behind the four numbers on the comparison slide.
pol_means <- pol %>%
  mutate(Group = ifelse(treated == 1, "Programme sectors",
                        "Comparison sectors"),
         Time  = tlev(ifelse(round == 0, "Before", "After"))) %>%
  group_by(Group, Time) %>%
  summarise(m = mean(injury_collisions), .groups = "drop")

# Segment counts, for the "who is in the evaluation" slide.
pol_seg  <- n_distinct(pol$segment_id)
pol_sect <- n_distinct(pol$sector_id)
pol_t_seg <- n_distinct(pol$segment_id[pol$treated == 1])

# ---- Exposure, and why the count and the rate disagree ----------------------
# The CSV carries a per-segment ratio, injuries_per_million_vkm. Averaging that
# column is NOT the rate for the group: a mean of ratios gives a 0.6 km slip
# road the same weight as a 3 km arterial, and the two disagree by a percentage
# point here. For "did risk fall", take the totals first and divide once.
pol_agg <- pol %>%
  mutate(Group = ifelse(treated == 1, "Programme sectors", "Comparison sectors")) %>%
  group_by(Group, round) %>%
  summarise(coll = sum(injury_collisions),
            vkm  = sum(vehicles_km_millions),
            .groups = "drop") %>%
  mutate(rate = coll / vkm)

pol_exp <- pol_agg %>%
  group_by(Group) %>%
  summarise(count_fall = 100 * (coll[round == 0] - coll[round == 1]) / coll[round == 0],
            vkm_rise   = 100 * (vkm[round == 1]  - vkm[round == 0])  / vkm[round == 0],
            rate_fall  = 100 * (rate[round == 0] - rate[round == 1]) / rate[round == 0],
            .groups = "drop")

pol_count_fall_t <- pol_exp$count_fall[pol_exp$Group == "Programme sectors"]
pol_count_fall_c <- pol_exp$count_fall[pol_exp$Group == "Comparison sectors"]
pol_vkm_rise_t   <- pol_exp$vkm_rise[pol_exp$Group == "Programme sectors"]
pol_vkm_rise_c   <- pol_exp$vkm_rise[pol_exp$Group == "Comparison sectors"]
pol_rate_fall_t  <- pol_exp$rate_fall[pol_exp$Group == "Programme sectors"]
pol_rate_fall_c  <- pol_exp$rate_fall[pol_exp$Group == "Comparison sectors"]

# The two falls the waterfall subtracts, in collisions per segment.
pol_fall_t <- pol_t_before - pol_t_after
pol_fall_c <- pol_c_before - pol_c_after

# Baseline imbalance, the measured version of "cameras went to the worst roads".
# This is the gap a treated-only before-and-after has no way to remove.
pol_base_gap <- pol_t_before - pol_c_before
pol_prior_t  <- mean(pol$prior_collisions_3yr[pol$treated == 1])
pol_prior_c  <- mean(pol$prior_collisions_3yr[pol$treated == 0])

# ---- The estimate against the rule, as one picture --------------------------
# A horizontal bar: the estimate (dot) with its 95% confidence interval
# (whiskers) and the decision rule as a vertical red line, so "the interval does
# not reach the rule" is something you can see rather than read.
#
# abs() flips the negative coefficient into "collisions avoided", so the
# interval's lower (more negative) end becomes the HIGHER avoided value: lo is
# the least avoided, hi is the best case the interval allows.
#
# A function, not a chunk, because two slides show this chart. Defined once so
# they cannot disagree about where the rule sits.
pol_ci_plot <- function(width = PL_W, height = 260, t = 30, b = 54) {
  est <- abs(pol_did)
  lo  <- abs(pol_did_ci[2])
  hi  <- abs(pol_did_ci[1])
  plot_ly() %>%
    add_segments(x = lo, xend = hi, y = 1, yend = 1, showlegend = FALSE,
                 line = list(color = ACCENT, width = 6), hoverinfo = "skip") %>%
    add_markers(x = c(lo, hi), y = c(1, 1), showlegend = FALSE,
                hoverinfo = "skip",
                marker = list(color = ACCENT, size = 18, symbol = "line-ns-open",
                              line = list(color = ACCENT, width = 3))) %>%
    add_markers(x = est, y = 1, showlegend = FALSE,
                marker = list(color = ACCENT, size = 16),
                hovertemplate = "estimate %{x:.2f} avoided<extra></extra>") %>%
    layout(
      shapes = list(list(type = "line", x0 = POL_RULE, x1 = POL_RULE,
                         y0 = 0.4, y1 = 1.6, xref = "x", yref = "y",
                         line = list(color = ALERT, width = 3, dash = "dash"))),
      annotations = list(
        list(x = POL_RULE, y = 1.68, xref = "x", yref = "y", showarrow = FALSE,
             text = sprintf("Decision rule: %.1f", POL_RULE),
             font = list(color = ALERT, size = 14), xanchor = "middle"),
        list(x = est, y = 0.66, xref = "x", yref = "y", showarrow = FALSE,
             text = sprintf("estimate %.2f", est),
             font = list(color = INK, size = 13), xanchor = "middle")),
      xaxis = ax("Injury collisions avoided per segment, per round",
                 range = c(0, 2.4)),
      yaxis = list(range = c(0.4, 1.9), showticklabels = FALSE,
                   showgrid = FALSE, zeroline = FALSE)) %>%
    pl_m2(width = width, height = height, t = t, b = b)
}
```

# 1 · The case

## The traffic camera programme

:::::: {.columns}

::::: {.column width="52%"}
In 2020 a roads authority began installing automated speed cameras. Not
everywhere at once: twelve sectors first, the rest two years later.

It set a bar before committing further: the programme had to avoid
**`r sprintf("%.1f", POL_RULE)` injury collisions per road segment per round**.
Below that, the money goes elsewhere.
:::::

::::: {.column width="48%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">Automated speed enforcement</div>
  <div class="k">Rolled out</div>  <div class="v">12 sectors in 2020, 12 in 2022</div>
  <div class="k">Outcome</div>     <div class="v">Injury collisions</div>
  <div class="k">Decision rule</div> <div class="v">2.0 fewer per segment per round</div>
</div>
```
:::::

::: {.warn .note-ctr .fragment}
**Illustrative case, with invented figures.** This is not a real programme and
the numbers are generated for this training. Every number in this session comes
from this fictional dataset.
:::

::::::

## Who is in the evaluation

```{r results='asis'}
cat(sprintf('
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">Road segments</span></div>
  <div><span class="n">%d</span><span class="k">Sectors</span></div>
  <div><span class="n">%s</span><span class="k">Segments with cameras</span></div>
  <div><span class="n">2</span><span class="k">Rounds: 2019 and 2021</span></div>
</div>', format(pol_seg, big.mark = ","), pol_sect,
   format(pol_t_seg, big.mark = ",")))
```

::: {.panel}
Session 1 taught three questions.

The first is **compared to what?** This session is that question pointed at one programme.

Two things about the design, before the numbers:

-  The programme went first to the
**worst roads**, so the sectors with cameras are not the same as the sectors
without. 

- Traffic **grew** between the two rounds.
:::

::: {.ask}
One of those two is why the headline number will turn out to overstate what the
programme did. **It is not the traffic.**
:::

## The decision on the table

::: {.panel}
The authority is deciding whether to extend cameras elsewhere. That is a
spending decision, and it turns on one figure: how many injury collisions the
cameras avoided, **per segment, per round**.

The rule is `r sprintf("%.1f", POL_RULE)`. A result below it does not justify the
money, however impressive it looks as a percentage.
:::

::: {.ask}
Three numbers describe the same programme:

**-`r sprintf("%.0f", pol_headline_fall)`%** &nbsp;·&nbsp; **-`r format(abs(pol_did), nsmall = 2)`** &nbsp;·&nbsp; **`r sprintf("%.1f", POL_RULE)`**

Two are wrong answers to the question being asked. Work out which is the real
one, and whether it clears the rule.
:::

# 2 · What the numbers say

## The headline

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Injury collisions, sectors with cameras</div>
</div>
<div class="stat-row">
  <div><span class="n">%.1f</span><span class="k">Per segment, 2019</span></div>
  <div><span class="n">%.1f</span><span class="k">Per segment, 2021</span></div>
  <div><span class="n">-%d%%</span><span class="k">The headline</span></div>
</div>', pol_headline_fall, pol_t_before, pol_t_after,
   round(pol_headline_fall)))
```

::: {.warn}
This is the number that reaches the minister. It is **accurate**: in the sectors
where cameras went, injury collisions fell
`r sprintf("%.0f", pol_headline_fall)`% between 2019 and 2021.

It is also a before-and-after comparison. It compares those roads to their own
past, which is exactly the comparison Session 1 spent an hour on.
:::

```{r fig.height=2.2, fig.width=11.5}
# The headline has to be a picture before the room votes on it. One group, two
# rounds, nothing else: this is deliberately the same shape as the Al Falah
# slide in Session 1, so the room recognises the trap.
pol_means %>%
  filter(Group == "Programme sectors") %>%
  plot_ly(x = ~Time, y = ~m, type = "bar",
          marker = list(color = ACCENT, line = list(width = 0)),
          text = ~sprintf("%.1f", m), textposition = "outside",
          textfont = list(size = 17, color = INK), cliponaxis = FALSE,
          showlegend = FALSE,
          hovertemplate = "%{y:.1f} collisions per segment<extra></extra>") %>%
  layout(barmode = "group",
         xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 16, color = INK)),
         yaxis = ax("Injury collisions per segment", range = c(0, 14.5))) %>%
  pl_m2(height = 200, b = 40)
```

## Would you fund the extension?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
On the number you have just seen, a `r sprintf("%.0f", pol_headline_fall)`% fall
where cameras went: **would you commit the money to extend them?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Not on this alone
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right. Commit before we look further.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

## What else happened in those two years

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.6, fig.width=5.4}
# Levels, both groups. The comparison bars are the counterfactual the headline
# ignores: collisions fell there too, with no cameras installed.
pol_means %>%
  plot_ly(x = ~Time, y = ~m, color = ~Group, type = "bar",
          colors = c("Programme sectors" = ACCENT,
                     "Comparison sectors" = ACCENT2),
          text = ~sprintf("%.1f", m), textposition = "outside",
          textfont = list(size = 16, color = INK), cliponaxis = FALSE,
          hovertemplate = "%{fullData.name}: %{y:.1f}<extra></extra>") %>%
  layout(barmode = "group",
         xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 15, color = INK)),
         yaxis = ax("Collisions per segment", range = c(0, 15))) %>%
  pl_m2(width = PL_WC, height = 250)
```

::: {.panel}
Cameras went to the twelve worst sectors. The twelve without cameras also saw
collisions fall, and nothing was installed there.
:::
:::::

::::: {.column width="50%"}
```{r fig.height=2.6, fig.width=5.4}
# The same data as the two FALLS. Turning levels into falls is the whole trick:
# once both groups are measured as falls, the comparison's fall is visibly not
# the programme's, and the gap between the bars is what is left.
pol_exp %>%
  plot_ly(x = ~Group, y = ~count_fall, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~sprintf("%.0f%%", count_fall), textposition = "outside",
          textfont = list(size = 16, color = INK), cliponaxis = FALSE,
          showlegend = FALSE,
          hovertemplate = "%{x}: %{y:.1f}%% fewer<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 14, color = INK)),
         yaxis = ax("Fall in injury collisions (%)", range = c(0, 30)),
         annotations = list(list(
           x = 0.5, y = 27, xref = "paper", yref = "y", showarrow = FALSE,
           text = sprintf("gap = %.0f points", pol_count_fall_t - pol_count_fall_c),
           font = list(size = 13, color = ALERT)))) %>%
  pl_m2(width = PL_WC, height = 250)
```

::: {.warn}
The gap between the two falls is the part that could belong to the cameras.
Nothing else on the slide is.
:::
:::::

::::::

## The honest estimate

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Injury collisions avoided per segment, per round</div>
</div>
<div class="stat-row">
  <div><span class="n">%.2f</span><span class="k">Programme sectors fell (collisions)</span></div>
  <div><span class="n">%.2f</span><span class="k">Comparison sectors fell (collisions)</span></div>
  <div><span class="n">%.2f</span><span class="k">The difference</span></div>
</div>', abs(pol_did), pol_fall_t, pol_fall_c, abs(pol_did)))
```

```{r fig.height=2.5, fig.width=11.5}
# The subtraction the slide used to describe in words. The comparison's fall is
# not a gain, so it is drawn as a step DOWN from the programme's fall; what
# survives to the right of it is the estimate. Reads left to right as the
# arithmetic the room is being asked to trust.
wf <- tibble(
  step = factor(c("Programme\nsectors fell", "Less the comparison's\nown fall",
                  "What is left"),
                levels = c("Programme\nsectors fell", "Less the comparison's\nown fall",
                           "What is left")),
  base = c(0, abs(pol_did), 0),
  val  = c(pol_fall_t, pol_fall_t - abs(pol_did), abs(pol_did)),
  # The minus sign is load-bearing: without it the middle bar reads as a gain.
  lab  = c(sprintf("%.2f", pol_fall_t), sprintf("\u2212%.2f", pol_fall_c),
           sprintf("%.2f", abs(pol_did))),
  col  = c(ACCENT, ACCENT2, ALERT)
)
plot_ly() %>%
  add_bars(data = wf, x = ~step, y = ~base, marker = list(color = "rgba(0,0,0,0)"),
           showlegend = FALSE, hoverinfo = "skip") %>%
  add_bars(data = wf, x = ~step, y = ~val, marker = list(color = ~col),
           text = ~lab, textposition = "outside",
           textfont = list(size = 15, color = INK), cliponaxis = FALSE,
           showlegend = FALSE,
           hovertemplate = "%{x}: %{text} collisions<extra></extra>") %>%
  layout(bargap = 0.45,
         xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 14, color = INK)),
         yaxis = ax("Collisions per segment", range = c(0, 3.4))) %>%
  pl_m2(height = 230, b = 46)
```

::: {.panel}
Note the units: **collisions**, not per cent. A percentage cannot be checked
against a rule that is written in collisions.
:::

## Does it clear the rule?

```{r fig.height=1.9, fig.width=11.5}
# Same chart as the later slide, from the shared function, so the rule cannot
# sit in one place here and another place there.
pol_ci_plot(height = 170, b = 38)
```

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Decision rule</div>      <div class="v">2.00 avoided</div>
  <div class="k">The estimate</div>       <div class="v">1.13 avoided</div>
  <div class="k">95% confidence interval</div>       <div class="v">0.51 to 1.76</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
Statistically clear, so the programme did something. **That is not the question
being asked.** It did not clear 2.00, and even the interval's best case,
`r sprintf("%.2f", pol_avoided_hi)`, falls short.
:::
:::::

::::::

::: {.warn .fragment}
**Statistically significant does not mean big enough.** Session 1 said this about
a school programme. Here is the same lesson with a real spending decision
attached to it.
:::

## Look back at your vote

```{r results='asis'}
# Shows the live Menti results once this session's Menti exists. Until
# MENTI_EMBED is set (still a placeholder), a marked placeholder box renders
# instead, so the deck never shows a broken iframe. data-external="1" keeps
# embed-resources from freezing the live view into a static image.
if (!grepl("PLACEHOLDER", MENTI_EMBED)) {
  cat(sprintf('
<iframe data-external="1" allowfullscreen="true" allowtransparency="true"
        frameborder="0" src="%s"
        style="width:100%%; height:66%%; border:none;"></iframe>', MENTI_EMBED))
} else {
  cat('<div class="panel note-ctr"><strong>[PLACEHOLDER] Live Menti results.</strong> Set <code>MENTI_EMBED</code> in the setup chunk to show the room vote here.</div>')
}
```

::: {.warn}
You voted on the `r sprintf("%.0f", pol_headline_fall)`% headline. The honest
estimate is `r sprintf("%.2f", abs(pol_did))` avoided, short of the
`r sprintf("%.1f", POL_RULE)` the rule needs. If the room leaned "yes", that is
the headline doing the work, not the programme.
:::

## Two reasons the headline overstates

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**1. The comparison moved too.** Collisions fell
`r sprintf("%.0f", pol_count_fall_c)`% in sectors with no cameras. A
before-and-after reading gives the programme credit for all of it.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**2. The groups started different.** Cameras went to the worst roads:
`r sprintf("%.1f", pol_t_before)` collisions per segment against
`r sprintf("%.1f", pol_c_before)`. A treated-only before-and-after has no way to
separate the programme from that head start.
:::
:::::

::::::

## The two groups were never the same

```{r fig.height=2.4, fig.width=11.5}
# The measured version of "cameras went to the worst roads": collisions in the
# round BEFORE anything was installed. This gap is the reason a treated-only
# before-and-after cannot be read as the programme's effect.
tibble(
  Group = factor(c("Programme sectors", "Comparison sectors"),
                 levels = c("Programme sectors", "Comparison sectors")),
  Base  = c(pol_t_before, pol_c_before)
) %>%
  plot_ly(x = ~Group, y = ~Base, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~sprintf("%.1f", Base), textposition = "outside",
          textfont = list(size = 16, color = INK), cliponaxis = FALSE,
          showlegend = FALSE,
          hovertemplate = "%{x}: %{y:.2f} collisions per segment in 2019<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 15, color = INK)),
         yaxis = ax("Injury collisions per segment, 2019", range = c(0, 15)),
         annotations = list(list(
           x = 0.5, y = 13.2, xref = "paper", yref = "y", showarrow = FALSE,
           text = sprintf("gap at the start = %.1f", pol_base_gap),
           font = list(size = 13, color = ALERT)))) %>%
  pl_m2(height = 230, b = 46)
```

::: {.warn}
In 2019, before a single camera was installed, the programme sectors were already
`r sprintf("%.1f", pol_base_gap)` collisions per segment worse. Nothing in the
headline accounts for that.
:::

## And one thing that cuts the other way

```{r fig.height=2.3, fig.width=11.5}
# Exposure rose, so collisions per vehicle-km fell MORE than the raw count did.
# This does NOT rescue the programme: the rule is written in collisions, so the
# count is still the number the decision turns on. But it does mean the headline
# is wrong in two directions at once, which is the point of the slide.
pol_exp %>%
  pivot_longer(c(count_fall, rate_fall), names_to = "Measure", values_to = "fall") %>%
  mutate(Measure = factor(
    ifelse(Measure == "count_fall", "Raw collision count", "Per million vehicle-km"),
    levels = c("Raw collision count", "Per million vehicle-km"))) %>%
  plot_ly(x = ~Group, y = ~fall, color = ~Measure, type = "bar",
          colors = c("Raw collision count" = ACCENT,
                     "Per million vehicle-km" = ACCENT2),
          text = ~sprintf("%.0f%%", fall), textposition = "outside",
          textfont = list(size = 15, color = INK), cliponaxis = FALSE,
          hovertemplate = "%{fullData.name}: %{y:.1f}%% fewer<extra></extra>") %>%
  layout(barmode = "group",
         xaxis = ax("", zeroline = FALSE,
                    tickfont = list(size = 15, color = INK)),
         yaxis = ax("Fall, 2019 to 2021 (%)", range = c(0, 33))) %>%
  pl_m2(height = 220, b = 46)
```

::: {.panel}
Exposure rose in **both** groups, so the fall in collisions per vehicle-km is
larger than the fall in the raw count:
`r sprintf("%.1f", pol_rate_fall_t)`% against
`r sprintf("%.1f", pol_count_fall_t)`% in programme sectors. On the risk question
the count understates what happened.
:::

::: {.ask .fragment}
Session 1's Al Falah case ran the same trap the other way. There, distance driven
**fell**, so fatalities fell with it and the campaign looked better than it was.
Here traffic **rose**, so the count looks worse than the risk.

The rule is written in collisions, so the count is still the number the decision
turns on. This changes what you can say about risk, not whether the programme
cleared the bar.
:::

# 3 · Reading the output

## The table you would actually be handed

```{r}
pol_tbl %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 22) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = "Outcome: injury collisions per road segment per round. Standard errors clustered by road segment.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.panel}
**Row 4 is the impact.** The interaction of programme and round: the extra change
in the sectors that got cameras, over and above the change everywhere else. It is
the one number the decision turns on.

Most reports lead with their preferred number. **Find the row that is actually
the impact**, not the one printed largest or first.
:::
::: {.notes}
Reminder for the room (vocab from Day 1 Session 1): the **coefficient** is the effect size, here in injury collisions per segment; the **standard error** is how much the estimate would move from sample to sample; the **95% confidence interval** is the range of plausible true values, not a margin of certainty; the **p-value** is how surprising this result would be if the programme had truly done nothing.
:::

## The same table: the other rows

```{r}
pol_tbl %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 22) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = "Outcome: injury collisions per road segment per round. Standard errors clustered by road segment.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.panel}
**Row 2** is the baseline difference. Cameras went to the worst roads, so the
programme sectors started higher. This gap is not the impact.

**Row 3** is the national trend: what happened in the comparison sectors anyway.
**Std. err.** is the wobble on each estimate: divide the coefficient by it to see
how steady the number is.
:::

::: {.notes}
Reminder for the room (vocab from Day 1 Session 1): the **coefficient** is the effect size, here in injury collisions per segment; the **standard error** is how much the estimate would move from sample to sample; the **95% confidence interval** is the range of plausible true values, not a margin of certainty; the **p-value** is how surprising this result would be if the programme had truly done nothing.
:::

## The same estimate, three ways of saying it

::: {.panel}
A decision-maker will be offered three framings of this one number. They are not
equally honest.
:::

```{=html}
<div class="facts">
  <div class="k">Per cent</div>
  <div class="v">"Collisions fell 22% where cameras were installed" &middot; ignores the comparison</div>
  <div class="k">Point estimate</div>
  <div class="v">"Cameras avoided 1.13 collisions per segment per round" &middot; accurate, but a single number</div>
  <div class="k">With its confidence interval</div>
  <div class="v">"Between 0.51 and 1.76, so somewhere short of the 2.0 required" &middot; this is the honest one</div>
</div>
```

::: {.ask}
Which would you put in a briefing? Session 1's Word 5 answer was that **the
confidence interval decides, not the point estimate.** Does it hold here?
:::

## A confidence interval that does not reach the rule

```{r fig.height=2.6, fig.width=11.5}
# The same chart as the "does it clear the rule?" slide, from the shared
# function. Here the room has the vocabulary, so it is redrawn bigger and the
# callout below names each part.
pol_ci_plot(height = 250)
```

::: {.warn}
**How to read it:** the blue dot is the estimate, `r sprintf("%.2f", abs(pol_did))`
collisions avoided, and the blue bar is its 95% confidence interval,
`r sprintf("%.2f", abs(pol_did_ci[2]))` to `r sprintf("%.2f", abs(pol_did_ci[1]))`.
The red dashed line is the rule: `r sprintf("%.1f", POL_RULE)`. The whole bar sits
to its left, so even the best case falls short. This is not "we cannot tell": it
is a clear result below the bar, and **a precise answer below the bar is still a
no.**
:::

## What the table does not tell you

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Nothing about roads without cameras.** The comparison sectors are a
counterfactual, not a plan. The estimate says what the cameras did where they
went, not what they would do elsewhere.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**Nothing about compliance.** Cameras work by changing behaviour. If drivers
learned where they are and sped up in between, the effect fades after 2021.
:::
:::::

::::::

::: {.ask}
Two habits from Session 1, applied to this table:

Ask what is **not** in it. Then ask the evaluator what would have to be true for
the result to be wrong. Here, the answer is that the two groups were already
diverging before 2020. Nobody has shown you a pre-trend.
:::

# 4 · Put AI to work

## AI Snapshot: three sentences for a decision-maker

::: {.panel}
Paste or take a picture of the table on the previous slide and add it to an AI tool with this prompt:

**"Explain this result to a decision-maker in three sentences. They have a rule:
the programme must avoid 2.0 collisions per segment per round to be funded."**
:::

::: {.ai-output}
**Illustrative AI response:**

"The programme was associated with a **22% reduction** in injury collisions in
sectors where cameras were installed. This represents a **clear improvement in
road safety**, with the effect likely to be **even larger** once the remaining
sectors are covered. [The evidence supports extending the programme
nationally.]{.flag}"
:::

::: {.ask}
Mark every claim you can check against the table. How many are incorrect?
:::

## Debrief: what the AI got wrong

```{=html}
<div class="facts">
  <div class="k">Claim 1</div>
  <div class="v">"22% reduction in injury collisions" &middot; true, but it is the before-and-after figure, not the impact</div>
  <div class="k">Claim 2</div>
  <div class="v">"effect likely to be even larger" &middot; nothing in the table supports a direction, let alone a bigger one</div>
  <div class="k">Claim 3</div>
  <div class="v">"evidence supports extending nationally" &middot; the rule it was given never appears in the answer</div>
</div>
```

::: {.warn .fragment}
**Claim 2 is the invention.** It supplies a confident direction from nowhere. The
table's fourth row says `r sprintf("%.2f", abs(pol_did))`, and the interval's best
case is `r sprintf("%.2f", pol_avoided_hi)`. Neither is "larger".
:::

## Debrief: the habit to take away

::: {.panel}
The AI was not lying. It picked the flattering number, and it answered
"is this good?" instead of "does it clear the bar?"
:::

```{=html}
<div class="facts">
  <div class="k">What it did</div>      <div class="v">Reached for the most impressive true number in the room</div>
  <div class="k">What it skipped</div>  <div class="v">Compared the estimate to the 2.0 rule it was handed</div>
  <div class="k">What to ask next time</div> <div class="v">"Compare this against the decision rule before you recommend anything."</div>
</div>
```

::: {.ask}
Session 1's habit, applied here: ask the AI to **ask you questions before it
answers**. It had the rule and still did not use it.
:::

# 5 · Annotate this output

## Your turn

:::::: {.columns}

::::: {.column width="54%"}
```{r}
pol_tbl %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5")
```

::: {.panel}
Outcome: injury collisions per road segment per round. Standard errors
clustered by road segment.
:::
:::::

::::: {.column width="46%"}
::: {.ask}
On the printed copy in front of you:

**1.** Circle the number that is the treatment effect.

**2.** Draw a box around its confidence interval.

**3.** Underline the baseline difference, and say in one line why it is not the
impact.

**4.** Write down one thing you cannot tell from this table, and the question
you would put to the evaluator.
:::

::: {.panel}
Four minutes. Question 4 matters most, and Session 1 prepared you for it.
:::
:::::

::::::

# 6 · Take this back to your desk

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Which group, which period, and did it move too?</div>
  <div class="k">2</div> <div class="v">Which row is the impact, and is it in units the decision is written in?</div>
  <div class="k">3</div> <div class="v">Does the confidence interval reach the decision rule, not just clear zero?</div>
</div>
```

::: {.panel}
This result would survive question 2. It fails question 3, and that is the whole
decision. And question 1 is why the headline number was never the answer: the
two groups started `r sprintf("%.1f", pol_base_gap)` collisions apart.
:::

## Where this goes next

::: {.panel}
Today you read a **before-and-after with a comparison group**. That is the
simplest honest design, and it rests on one assumption: that the two groups were
heading the same way before the programme started.

Day 2 does three things this session did not:

**Difference-in-differences** on this same data, testing that assumption rather
than assuming it. **Regression discontinuity** on the speed cut-off that decided
which roads got cameras. **Matching**, because the twelve worst sectors were not
chosen at random.
:::

::: {.ask}
Same programme, same 1,200 road segments, four ways of asking what it did. The
answers will not be identical, and the differences are the lesson.
:::
