---
title: "From Findings to Policy"
subtitle: "Module 2 · Day 4, Session 2: Evidence Translation"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN). No data file: the findings are in qa_translation.R
# (static slides) and repeated in the hidden setup cell (live slides).
#
# One example all session: Tariff Shield, a fictional scheme, deliberately not
# GreenWaste (see qa_translation.R for why). New in this version: a brief
# checker the trainer runs live on the room's sentences and on the AI draft.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: []
  render-df: kable
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(knitr)
library(qrcode)

FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- The findings set and the brief template ------------------------------
# Shared with the printable pack and template, so the deck cannot disagree with
# the paper in the room's hands.
source("qa_translation.R")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_d4s2.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0)); plot(qr_code(url, ecl = "H")); path
}
MENTI_QR <- menti_qr()
```

# 0 · The decision {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## From reading evidence to writing it

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. The numbers repeat
# qa_translation.R; keep the two in step.
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)
saving <- 743; ci_low <- 690; ci_high <- 796; threshold <- 900

# One line: the estimate and its interval against a threshold.
threshold_plot <- function(rule) {
  par(mar = c(4.2, 1, 1, 1), family = "sans", las = 1, col.axis = MUTED,
      col.lab = MUTED, fg = MUTED, cex = 1.25)
  plot(NA, xlim = c(500, 1100), ylim = c(0.5, 1.5), yaxt = "n", bty = "n",
       xlab = "Net saving, AED per vehicle per year", ylab = "")
  abline(v = rule, col = ALERT, lwd = 3, lty = 2)
  segments(ci_low, 1, ci_high, 1, col = ACCENT, lwd = 10)
  points(saving, 1, pch = 19, cex = 2.8, col = ACCENT)
  text(rule, 1.3, paste("threshold", rule), col = ALERT, pos = 4)
}

# The verdict at a given threshold: a small table, then the chart.
threshold_view <- function(rule) {
  threshold_plot(rule)
  data.frame(threshold = rule, shortfall = rule - saving, interval_clears = ci_low >= rule)
}

# The six checks from "What to look for", run on any piece of text. Each is a
# plain word search, so it can be fooled; it is a prompt for the room, not a
# verdict.
has <- function(text, words) any(sapply(words, grepl, x = tolower(text), fixed = TRUE))
check_brief <- function(text) {
  data.frame(
    check = c("Gives the saving (743)",
              "States the 900 AED threshold",
              "Does not claim customers benefited",
              "Gives the enrolment figure (34%)",
              "Mentions the 22% cost overrun",
              "Does not turn the null into 'no effect'"),
    result = ifelse(c(
      has(text, "743"),
      has(text, c("900", "threshold")),
      !has(text, c("customers benefit", "benefit to customers", "passed part of the saving",
                   "passed on to customers", "wider economic benefit", "consumers benefit")),
      has(text, c("34%", "34 per", "two-thirds", "two thirds", "a third")),
      has(text, c("22%", "overrun", "over budget")),
      !has(text, c("no effect", "no wider benefit", "no impact", "had no benefit"))),
      "yes", "NO"))
}
```

::: {.panel}
This morning you judged whether a report was credible. Now someone hands you
findings you did not produce and asks what the authority should do. For many of
you that is the usual request.
:::

::: {.ask .fragment}
**The last time you wrote up evidence for a decision, what was the hardest
part?**
:::

::: {.notes}
Open the deck a couple of minutes before the session starts: R takes about a
minute to load in the browser, after which every live cell runs in a few seconds.
:::

## The programme

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Programme</div>  <div class="v">%s</div>
  <div class="k">What it is</div> <div class="v">%s</div>
  <div class="k">Where</div>      <div class="v">%s</div>
  <div class="k">When</div>       <div class="v">%s</div>
</div>', transl_programme$name, transl_programme$what,
    transl_programme$where, transl_programme$when))
```

::: {.warn}
This is **not** GreenWaste. It is a different scheme, and you have not seen these
findings before. **Illustrative case, with invented figures.**
:::

## The number the decision turns on

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">900</div>
  <div class="label">AED per vehicle per year, the threshold to clear</div>
</div>
```

The evaluation was randomised, well executed and reported transparently. Nothing
in it fails this morning's checklist.

::: {.ask}
**`[MENTI PLACEHOLDER]`** Would you recommend this scheme for rollout? Yes / No
/ Only with changes. Vote before reading the findings; you vote again at the end.
:::
:::::

::::: {.column width="42%"}
```{r out.width='280px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

# 1 · The findings {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Six findings, all of them sound

```{r}
kable(data.frame(`#` = seq_len(nrow(transl_findings)), Finding = transl_findings$finding,
                 check.names = FALSE), align = "rl")
```

::: {.panel}
Every one survives this morning's checklist.
:::

::: {.ask .fragment}
**If the evidence is sound, what makes the brief hard to write?**
:::

::: {.notes}
Paper (Oct15_session2_materials.docx): Hand out the findings pack (Oct15_session2_findings.docx), one per participant.
:::

## Finding 1: the saving

::: {.excerpt}
Fuel costs fell for participating operators. Net saving of **743 AED per vehicle
per year** (95% CI: 690 to 796). Statistically significant at conventional levels.
:::

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Estimate</div>  <div class="v">743 AED per vehicle per year</div>
  <div class="k">Interval</div>  <div class="v">690 to 796</div>
  <div class="k">Threshold</div> <div class="v">900</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
A clean, well-measured effect on the main outcome, from a randomised
allocation. It is the finding the evaluation was designed to produce.
:::

::: {.warn .fragment}
The whole interval lies below 900. On the authority's own rule, the scheme does
not clear the bar.
:::
:::::

::::::

## Finding 2: the same finding, read against the rule

```{ojs}
//| echo: false
viewof rule = Inputs.range([600, 1000], {value: 900, step: 10, label: "Threshold (AED per vehicle)"})
```

```{webr}
#| input:
#|   - rule
#| autorun: true
#| edit: false
#| fig-width: 6
#| fig-height: 1.9
threshold_view(rule)
```

::: {.ask}
743 against 900: success or failure? Move the threshold: **where does the
verdict flip, and who chose 900?**
:::

::: {.panel .fragment}
The evaluation states the 157 AED shortfall plainly. It does not say whether 900
was the right threshold; nothing in it was designed to answer that.
:::

## Finding 3: the pass-through

::: {.excerpt}
Freight rates charged by participating operators fell by an average of **1.8%**,
against **0.4%** among non-participants. The difference is **not statistically
significant** (p = 0.11).
:::

::: {.ask}
This is the finding most likely to end up in a brief. What would you write?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**What the evidence supports**

Freight rates fell among participants, in the expected direction, and more than
among non-participants.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it does not**

That customers benefited. The estimate cannot be told apart from zero: it is a
hypothesis the evaluation could not settle.
:::
:::::

::::::

## Finding 4: who did not join

::: {.excerpt}
**34%** of eligible operators enrolled. Non-participants were more likely to be
small fleets and to operate on fixed public-sector contracts.
:::

::: {.ask .fragment}
The evaluation reports a pattern. Does it explain it?
:::

::: {.panel .fragment}
No. It measured **who** did not join, not **why**. "Small operators did not take
it up" is reportable. "Small operators could not take it up" is an inference the
evidence does not support, and it points to a different recommendation.
:::

## Finding 5: the cost of running it

::: {.excerpt}
Administration ran **22% over budget**. The overrun is concentrated in
verification of fuel purchase records, which required manual inspection in
**41%** of cases.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
This figure comes from the scheme's own accounts, not from the impact
estimate, and it bears most directly on whether to do this again.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
It is also the finding most often left out of a brief, because it is about
delivery rather than impact. If a brief recommends rollout without it, the
authority finds out after deciding.
:::
:::::

::::::

## Finding 6: the null result

::: {.excerpt}
No measurable effect on vehicle maintenance or fleet renewal. Point estimates for
both outcomes are close to zero with **wide confidence intervals**. The evaluation
cannot rule out effects of either sign.
:::

::: {.ask .fragment}
How do you write a finding that says nothing was found, and is that a finding?
:::

::: {.panel .fragment}
Yes, if you say what kind of result it is. The intervals are wide, so the study
could have missed an effect that matters. "The evaluation could not detect an
effect" is accurate; "the scheme had no wider benefits" overstates it.
:::

# 2 · The brief {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Four elements, one page

```{r}
kable(data.frame(Element = brief_elements$element, Length = brief_elements$length_guide,
                 `What it has to do` = brief_elements$asks, check.names = FALSE),
      align = "lll")
```

::: {.warn}
One side of A4. Decision-makers rarely read past the first page.
:::

::: {.notes}
Paper (Oct15_session2_materials.docx): Hand out the brief template (Oct15_session2_brief_template.docx), one per pair.
:::

## The headline sentence

::: {.ask}
Write your headline finding as **one sentence**, with the number in it. Then ask:
if this were the only sentence anyone read, could they conclude the opposite of
what you found?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The test**

A reader who stops after one sentence should not be able to reach a conclusion
the evidence does not support.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
"Fuel costs fell for participating operators" is true and easy to read, and it
lets the reader assume the threshold was cleared. The number prevents that.
:::
:::::

::::::

## Test your headline

```{webr}
headline <- "Fuel costs fell for participating operators."
check_brief(headline)[1:2, ]
```

::: {.ask}
Read me your headline. I type it between the quotes and press Run. A headline
needs the first two checks; the rest apply to the whole brief.
:::

::: {.notes}
Take three or four headlines from the room. The check is a word search, so it
can be fooled: a sentence can contain "900" and still mislead. Use it to start
the discussion, not to settle it.
:::

## The recommendation

::: {.ask}
Your recommendation must be **one specific action** the authority can decide.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Decidable**

- Roll out to the three regions where the scheme was piloted
- Extend the pilot for two years and re-run the costing
- Do not roll out, and re-tender the verification process
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Not decidable**

- The scheme should be improved
- More evidence is needed
- Stakeholders should be engaged
:::
:::::

::::::

::: {.ask .fragment}
Which option can you defend, and does it depend on how you weigh 743 against 900?
:::

## The risk

::: {.ask}
Name **one thing** that would make your recommendation wrong, stated so that it
can be watched.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**A risk you can watch**

"If the 22% cost overrun is repeated at scale, administration cost per vehicle
would exceed the saving. The first two quarters of spending would show this."
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**A caveat presented as a risk**

"Results may not generalise." True of almost any evaluation; it can be neither
confirmed nor watched.
:::
:::::

::::::

## Write the brief

::: {.panel}
Pairs, twelve minutes, on the template: the headline with its number, two or
three implications, one recommendation, one risk you can watch.
:::

::: {.ask}
Keep the findings pack open. Every sentence in your brief should point to a
finding.
:::

::: {.notes}
Walk the room. If a pair is stuck, ask them to write the recommendation first
and work back to the headline.
:::

# 3 · Pairs and critique {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Swap

::: {.panel}
Pairs, swap briefs. Read your partner's as the person deciding.

1. What does this brief want me to do?
2. What is its strongest claim, and does the findings pack support it?
3. What has it left out that I would have wanted to know?
:::

::: {.ask .fragment}
Spend most of your time on question 3. What is missing is harder to spot than
what is wrong.
:::

::: {.notes}
Paper (Oct15_session2_materials.docx): Hand out the peer-critique sheet (pack page 1), one per pair.
:::

## What to look for

```{r results='asis'}
checks <- c("Does the headline sentence contain the number?",
            "Does it mention the 900 AED threshold at all?",
            "Does it claim a customer benefit from the pass-through finding?",
            "Does it explain why two-thirds did not enrol?",
            "Does it mention the 22% cost overrun?",
            "Does it turn the null finding into a statement of no effect?")
html_out(paste(sprintf('<div class="panel"><b>%d.</b> %s</div>', seq_along(checks), checks),
               collapse = "\n"))
```

::: {.warn .fragment}
Each is a way to write a brief that is accurate sentence by sentence and
misleading as a whole.
:::

## Run the checks on a draft

```{webr}
draft <- "Tariff Shield saved operators 743 AED per vehicle per year, short of the
900 AED threshold. Freight rates fell slightly, but the evaluation could not show
customers gained. Only 34% of eligible operators joined."
check_brief(draft)
```

::: {.ask}
A volunteer pair reads out their brief; I paste it in and press Run. Which
checks does it pass, and is the check right?
:::

::: {.notes}
The default draft passes five checks and misses the cost overrun, which is the
point to draw out. The word search misreads some phrasing (for example
"customers did not benefit" contains "customers benefit"): say so if it happens,
and let the room judge.
:::

## The two failures that matter most

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Omitting the threshold**

A brief that reports 743 and never mentions 900 lets the reader conclude the
scheme met its target.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Omitting the cost overrun**

The authority is deciding whether to spend money. A brief about impact and not
cost answers a different question from the one asked.
:::
:::::

::::::

::: {.ask .fragment}
Both are omissions rather than errors. Which is harder to catch in your own
draft?
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The prompt

::: {.excerpt}
Here are six findings from an evaluation of a fuel-price stabilisation scheme for
small freight operators. Write a one-page policy brief for the commissioning
authority with a headline finding, key implications, a recommendation, and one
risk to flag.
:::

::: {.panel}
It asks for what you were just asked to write. Read the draft, then edit it as
a team.
:::

## What came back

:::::: {.columns}

::::: {.column width="60%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**Headline finding.** The scheme reduced fuel costs for participating operators,
[delivering a net saving of 743 AED per vehicle per year]{.flag}. The result is
statistically significant and the evaluation was rigorously conducted.
**Key implications.** Operators [passed part of the saving on to customers, with
freight rates falling by 1.8%]{.flag}, suggesting a wider economic benefit.
[Some operators chose not to enrol]{.flag}, which may indicate a need for better
communication. **Recommendation.** Roll the scheme out more widely, with
[targeted outreach to operators who did not participate]{.flag}.
**Risk to flag.** Results from a pilot may not generalise to a full rollout.
:::
:::::

::::: {.column width="40%"}
::: {.ask}
Find the four flagged problems. Then find the two things the draft leaves out
entirely, which are not flagged.
:::
:::::

::::::

::: {.notes}
Paper (Oct15_session2_materials.docx): Hand out the AI draft (pack page 2), one per group; they edit it on paper before you run the checker.
:::

## The checks, run on the AI draft

```{webr}
ai_draft <- "The scheme reduced fuel costs for participating operators, delivering a
net saving of 743 AED per vehicle per year. Operators passed part of the saving on
to customers, suggesting a wider economic benefit. Some operators chose not to
enrol. The scheme should be rolled out more widely. Results may not generalise."
check_brief(ai_draft)
```

::: {.panel .fragment}
It fails four of six: no threshold, a customer benefit it cannot show, no
enrolment figure, no cost overrun. It overstated the good news and softened the
bad. Check your own drafts, and any AI draft, for the same pattern.
:::

## What AI did, and what it could not do

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- The right four elements, in the right order
- The register of a brief
- One page, fluent enough to send unedited
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it did not**

- Never mentions the 900 threshold
- Calls a non-significant estimate a wider benefit
- Turns 34% enrolment into "some operators"
- Leaves out the 22% cost overrun
- Names a caveat, not a risk
:::
:::::

::::::

## The same task, a better prompt

::: {.excerpt}
The same six findings. Before you write anything: first, list the findings that
argue **against** rollout and how strongly each one argues, then those that argue
for it. Second, tell me which finding would have to be different for your
recommendation to change, and in which direction. Then write the brief, including
a sentence that states plainly what the evaluation did not establish.
:::

::: {.panel .fragment}
Three changes: ask for **the case against first**, so it is not an afterthought;
ask **what would change the recommendation**, which brings the threshold in; and
require a statement of **what was not established**, which makes omission harder.
:::

# 5 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Vote again

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">900</div>
  <div class="label">The bar, and 743 sits below it</div>
</div>
```

::: {.ask}
**`[MENTI PLACEHOLDER]`** Same three options: Yes / No / Only with changes.
Compare the two votes. If they moved, was it the threshold that moved them?
:::
:::::

::::: {.column width="42%"}
```{r out.width='280px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

## What makes a brief trustworthy

```{r results='asis'}
rules <- c("The number is in the headline sentence, not buried in the body.",
           "The threshold the number is judged against is stated, even when it is missed.",
           "Findings that argue against the recommendation are named, not left out.",
           "The recommendation is an action the reader can take or refuse.",
           "The risk is something that could happen and could be watched.",
           "What the evaluation did not establish is stated as plainly as what it did.")
html_out(paste(sprintf('<div class="panel"><b>%d.</b> %s</div>', seq_along(rules), rules),
               collapse = "\n"))
```

::: {.ask .fragment}
Which of the six is hardest to write when you already know the recommendation
you want?
:::

::: {.notes}
Paper (Oct15_session2_materials.docx): Hand out the take-away cards (pack page 3). Give a minute to fill in "On my desk" on the card, then take two or three examples from the room.
:::

## Next session

::: {.panel}
The last session of the module. You take the evaluation design you started in
Module 1 and add to it: how you would present results, cost the programme, show
the figures and translate the findings. Then a five-minute pitch, with feedback.
Bring your notes from the week.
:::
