---
title: "Worth the money?"
subtitle: "Day 3 | Session 1: Read the model, question its assumptions"
author: "3ie · Review version"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, day3_session1_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session1_review_interactions.html]
webr:
  render-df: kable
  resources: [evaluation_data_GreenWaste_simple.csv]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day3_session1_review.R', encoding='UTF-8')
```

## A saving is only half the question {#d3s1-p01 .reader-slide .s1-review .d3s1-review}

City DiD: an extra recorded annual cost reduction of **`r fmt(model$saving)` AED per participant**.

This is today's **provisional benefit input**, not a guaranteed causal saving. Parallel trends remains an assumption.

::: {.ask}
What else would you need before approving spending?
:::

The **1,000 AED annual-saving minimum** and **benefits exceeding costs** are different questions.

::: {.notes}
4 minutes. Retrieve city DiD without another methods lecture. Pilot, cutoff and matching concern different comparisons/populations; do not swap or average them as model inputs. Programme remains equipment subsidies, help and training. Future savings, duration and all economic cost/rate inputs are fictional assumptions. Start webR warming.
:::

## A ratio of 1.86. Approve? {#d3s1-p02 .reader-slide .s1-review .d3s1-review}

<p class="big-number">`r fmt(ratio(), 2)`</p>

**Modelled benefits divided by included costs**, expressed at today's value.

<div data-choices><button data-choice aria-pressed="false">Approve</button><button data-choice aria-pressed="false">Ask first</button><button data-choice aria-pressed="false">Do not approve</button><p class="feedback" role="status"></p></div>

::: {.ask}
Give one reason before opening the inputs.
:::

::: {.notes}
4 minutes. Commit locally or use physical cards for room responses; buttons do not collect room votes. Above 1 means included modelled benefits exceed included modelled costs, not proof of programme effect or complete spending approval. Revisit judgement after scenarios.
:::

## What went into the number? {#d3s1-p03 .reader-slide .s1-review .d3s1-review}

```{r, results='asis'}
html_out(d7_inputs())
```

::: {.ask}
Q1, worksheet page 3: which inputs are estimates, and which are assumptions?
:::

Source: **H7-A §1, Table I, page 1**. Costs, persistence and the rate need evidence.

::: {.notes}
6 minutes. Explicitly distinguish estimated observed extra change from projected benefits. 1,800 cost is assumed for equipment delivery per participant at time zero, not measured expenditure. Five saving years follow a one-year lag. Future causal benefit depends on both attribution and persistence. No attrition/fieldwork evidence or full cost audit supplied.
:::

## When do the costs and savings happen? {#d3s1-p04 .reader-slide .s1-review .d3s1-review}

```{r, results='asis'}
html_out(d7_timeline())
```

::: {.ask}
Q1: how many saving years are there? Is there a saving in year 1?
:::

The cash-flow schedule is **assumed**, not five years of follow-up data.

::: {.notes}
5 minutes. Cost now, no benefit year 1, five equal saving amounts in years 2-6. Distinguish programme before/after observation from economic projection. Timeline is not another observed outcome series. Worksheet page 1 reference. Learners mark start and end years before calculating.
:::

## A later saving is worth less today {#d3s1-p05 .reader-slide .s1-review .d3s1-review}

```{r, results='asis'}
html_out(d7_discount())
```

::: {.ask}
At an assumed 5% rate, why is next year's 812 AED not counted as 812 AED today?
:::

Discounting expresses future amounts at a **common date**. It does not verify that savings will occur.

::: {.notes}
5 minutes. One-year worked illustration approximately 812/1.05=774. Explain opportunity cost/time value without finance lecture. Use unrounded saving in calculations. Rate is assumed, not claimed official guidance. First model saving is two years away and divides by 1.05 squared. Distinguish discounting from an explicit annual decay of benefits.
:::

## From five future savings to today's total {#d3s1-p06 .reader-slide .s1-review .d3s1-review}

<div class="cash-table">
```{r, results='asis'}
html_out(d7_cash_table())
```
</div>

**Year 2:** `r fmt(model$saving)` / 1.05² ≈ **`r fmt(model$saving/1.05^2)` AED**.

**Sum years 2-6:** present-value benefits ≈ **`r fmt(d7_benefits)` AED per participant**.

::: {.ask}
Q2, page 3: why not simply multiply 812 by five?
:::

::: {.notes}
6 minutes. Read each discounted saving, not all rows aloud. Year 0 cost needs no discount; year 1 saving zero. Table F page 2. Q2 year2 amount and total. Printed rounded rows may not sum exactly; total uses unrounded records. Discounting reduces the sum, while duration and decay are separate model assumptions.
:::

## Ratio and net present value answer different questions {#d3s1-p07 .reader-slide .s1-review .d3s1-review}

```{r, results='asis'}
html_out(d7_output())
```

::: {.ask}
Q2: explain 1.86 and 1,550 AED without treating either as a proven outcome.
:::

**Ratio:** 3,350 / 1,800 ≈ 1.86. **NPV:** 3,350 - 1,800 ≈ 1,550 AED.

::: {.notes}
5 minutes. H7-A Table O page 2. Ratio unitless; NPV AED per participant at year0. Same model, same boundary. They do not rank programmes identically when scale/cost differs; no new comparison task. A positive model NPV/ratio above1 is not the separate 1,000 annual-saving rule or proof of value for money.
:::

## The trainer runs the model arithmetic {#d3s1-p08 .reader-slide .s1-review .d3s1-review}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
city <- subset(gw, setting == 'city')
change <- city$cost_after - city$cost_before
joined <- city$took_part == 1
saving <- -(mean(change[joined])-mean(change[!joined]))
```

```{webr}
#| echo: true
# Discount five assumed savings in years 2-6.
benefits <- saving * sum(1 / 1.05^(2:6))
# Compare with assumed year-0 equipment delivery cost.
cost <- 1800
round(c(ratio = benefits / cost, NPV_AED = benefits - cost), 2)
```

Paper result: **ratio 1.86; NPV 1,549.97 AED**.

::: {.ask}
Does running the calculation verify its assumptions?
:::

::: {.notes}
5 minutes. Predict then one trainer Run; participants never code. Same input unrounded city saving, rate, lag, duration and cost as canonical source. Output has ratio and AED labelled. If browser is slow use paper result. No repeated live scenario calculations; subsequent scenarios prepared from same model.
:::

## Whose costs and benefits are counted? {#d3s1-p09 .reader-slide .s1-review .d3s1-review}

<div class="review-grid"><div><h3>Authority</h3><p>Equipment delivery: included assumed cost.<br>Administration and staff time: coverage unknown.</p></div><div><h3>Businesses and others</h3><p>Business waste-cost savings: modelled benefit.<br>Own time, maintenance, wider impacts: coverage unknown.</p></div></div>

::: {.ask}
Q3, page 3: whose perspective is needed for this decision, and what is already included?
:::

The headline is a **limited mixed-actor model**, not a complete authority-budget or social appraisal. Check boundaries before adding costs.

::: {.notes}
6 minutes. Source §3 Table P page1. Authority cost versus business financial saving: cannot call ratio a return to government budget or complete social benefit. Waste-cost savings are not automatically net social gains; resource costs, charges/transfers, environmental benefits need appropriate treatment. Check amounts/timing/duplicates. Card2 assumes 600 additional non-duplicated year0 cost for exercise, not measured missing expenditure.
:::

## Change an assumption. Predict before the reveal. {#d3s1-p10 .reader-slide .s1-review .d3s1-review}

Eight groups receive cards **1, 2, 3, 4, 5, 1, 2, 3**.

Use **Q4, worksheet page 4**:

1. Identify exactly what changes and what stays fixed.
2. Predict the ratio and post your card number **above** the A1 line.
3. Request evidence about whether the change is plausible.

::: {.ask}
Which changed input might bring modelled benefits below included costs?
:::

::: {.notes}
9 minutes. Five preserved cards: decay30% after first saving year; extra600 year0 costs; rate8%; three saving years2-4; decay plus extra600. Card5 deliberately combines two inputs. Use existing A1 board or separate review board page6 of cards. Do not hand out results before predictions. Allocation eight exact. Six group work, three post reasons; no walking break.
:::

## Where do the five scenarios land? {#d3s1-p11 .reader-slide .s1-review .d3s1-review}

<button data-open="d7-results" aria-controls="d7-results" aria-expanded="false" data-closed="Open scenario results" data-opened="Hide scenario results">Open scenario results</button>
<div id="d7-results" hidden>
```{r, results='asis'}
html_out(d7_scenario_table())
```
</div>

::: {.ask}
Place results below the line. Which scenario crosses 1, and what makes it different?
:::

Base ratio **1.86**. Board: **cards page 6**. Results reference: **cards page 7**, distributed after predictions.

::: {.notes}
7 minutes. Results cards1-5 1.07,1.40,1.67,1.17,.80; only5 below1, it combines costs and decay. Card1 comes closest among single-input changes. Distribute Table S on cards page7 only after predictions recorded. Three reports, two compare predictions, two explain. Results under hypothetical settings, not measured failures. Q4 record result and evidence question. Existing A1 route unchanged.
:::

## Which change matters, over which range? {#d3s1-p12 .reader-slide .s1-review .d3s1-review}

Among the **specified single-input Cards 1-4**, the 30% decay setting moves the ratio most: **1.86 → 1.07**.

::: {.ask}
Does this prove decay is always the most important assumption?
:::

No. The ranking depends on the **ranges tested**. Card 5 is not a single-input comparison.

Choose an input, the evidence needed and a possible decision consequence in **Q4, page 4**.

::: {.notes}
5 minutes. Different scale/size perturbations cannot establish general global importance. No distributions or empirical ranges given. Do not claim largest slope/standardised sensitivity. Decay plausibility could use persistence/maintenance evidence; costs invoices/time records; rate applicable guidance. Attribution of benefit remains separate and potentially consequential.
:::

## Read the model note, not just the headline {#d3s1-p13 .reader-slide .s1-review .d3s1-review}

Read **H7-A, worksheet pages 1-2**, independently.

In **Q5, page 4**:

- Identify one estimated input and one assumption, citing their source IDs.
- State what the ratio supports **under those assumptions**.
- Identify one missing fact that could change your recommendation.

::: {.ask}
Which part of this value-for-money claim is not established by the calculation?
:::

::: {.notes}
4 minutes. Two independent source reading, one partner comparison, one collect. Tables I/F/O/P and §§1-4 clearly supplied; no external report. Look for provisional DiD, assumed persistence/cost/rate, limited boundary. Five-year observed data are not present. Use judgement to target later feedback, not compulsory single answer.
:::

## AI Snapshot: find a scenario that breaks it {#d3s1-p14 .reader-slide .s1-review .d3s1-review}

Naive prompt: **"Generate one plausible scenario where this programme's benefit-cost ratio falls below 1. Tell me whether it is realistic."**

<div class="ai-reading">
```{r, results='asis'}
html_out(paste0('<p>', gsub('\n\n', '</p><p>', d7_ai, fixed=TRUE), '</p>'))
```
</div>

::: {.ask}
Page 5: check arithmetic, timing and claims of realism against H7-A.
:::

::: {.notes}
8 minutes. Stage1 of15. Adapt older v1 response, illustrative not live transcript. Keep baseline and correct high-rate result. Reject unsupported rate guidance, unconditional no-funding, claim few years delay breaks ratio and five years established. Source §4 gives40% result and three-extra-year delay check, neither is plausibility evidence. No external rate lookup needed; no universal claim no agency uses40%. Three mark, three discuss, two feedback.
:::

## Ask Copilot to identify and quantify assumptions {#d3s1-p15 .reader-slide .s1-review .d3s1-review}

<div class="better-prompt">
```{r, results='asis'}
html_out(paste0('<p>', d7_prompt, '</p>'))
```
</div>

::: {.ask}
Run the full prompt on page 5. Check one figure and one requested missing fact against the source.
:::

No device: open the **prepared response on page 6**, after marking Stage 1. Use the shared **AI good-practice card**.

::: {.notes}
7 minutes. Stage2. Adapt old v2 while correcting false measured-cost assertion; finite scenario ranges make ranking meaningful. Two prompt/run, three compare, two share. Microsoft Copilot expected, no setup. Fictional data only. Response should ask questions before recommendation. No second-chat exercise. Paper is example not exact expected wording.
:::

## Close the source. Make a conditional recommendation. {#d3s1-p16 .reader-slide .s1-review .d3s1-review}

In **Q6, worksheet page 4**, write:

- What **1.86** means under the stated model.
- Your provisional action and its condition.
- One evidence request, and how the answer could change that action.

::: {.ask}
Which missing fact matters most for your decision?
:::

**Arithmetic is necessary. Credible benefits, suitable boundaries and defensible assumptions are necessary too.**

::: {.notes}
4 minutes. Two independent, one pair check, one targeted feedback. No mandatory approve/reject answer. Costs, persistence, rate or benefit attribution requests acceptable if specific and consequential. Ratio break-even1 versus annual-saving1000 different. Total90 minutes, 09:00-10:30 proposed schedule; break follows session, not inside it.
:::
