---
title: "Reading a Difference-in-Differences Result"
subtitle: "Module 2 · Day 2, Session 1: The Extra Change in the Programme Group"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / offered
ACCENT2 <- "#A9561F"   # series 2: comparison / not offered
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))
gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. Everything below reads from here.
# This is the short JOIN code participants type at menti.com, NOT the long
# alphanumeric ID in the editor URL. Confirm it while the Menti is live.
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/al68xcnp6hm6zmc9k94rhbxwimzq6m29/embed"

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- The decision rule, fixed for the whole week ---------------------------
RULEBAR <- 1000    # must cut costs by at least this many AED

# ---- GreenWaste, among the neighbourhoods that were offered the programme ---
# DiD needs an offered and a not-offered group inside the same comparison. The
# distinction here is ENROLMENT: in the treated neighbourhoods, some businesses
# took the programme up and some did not. That is the comparison the outline
# asks for, and it is the one the draft already used.
gw  <- read.csv("./evaluation_data_GreenWaste.csv")
tn  <- gw %>% filter(treatment_neighborhood == 1)

gw_means <- tn %>%
  group_by(enrolled, round) %>%
  summarise(mc = mean(waste_management_costs), .groups = "drop")
gv <- function(e, r) gw_means$mc[gw_means$enrolled == e & gw_means$round == r]

off_b <- gv(1, 0); off_a <- gv(1, 1)   # offered / enrolled
ctl_b <- gv(0, 0); ctl_a <- gv(0, 1)   # not offered / not enrolled

off_chg  <- off_a - off_b               # -665
ctl_chg  <- ctl_a - ctl_b               # +151
did_hand <- off_chg - ctl_chg           # -816, the four numbers done by hand

# ---- The regression an evaluator would actually report ---------------------
fit_did <- lm_robust(waste_management_costs ~ round * enrolled, data = tn,
                     clusters = neighborhood_identifier)
did_est <- coef(fit_did)[["round:enrolled"]]
did_ci  <- confint(fit_did)["round:enrolled", ]
did_p   <- fit_did$p.value[["round:enrolled"]]

# Sign trap: the coefficient is negative, so conf.low is the LARGER effect and
# abs(conf.low) is the OPTIMISTIC end. Sorting the absolute values makes the
# first element the end closest to zero, which is the one the decision turns on.
did_avoided       <- sort(abs(did_ci))
did_avoided_least <- did_avoided[1]
did_avoided_most  <- did_avoided[2]
did_clears        <- did_avoided_least >= RULEBAR

# ---- The two traps, both computed rather than asserted ---------------------
# Trap 1: the baseline gap. The two groups did not start level.
trap_gap  <- off_b - ctl_b              # -630 (offered started lower)
# Trap 2: the time trend. Both groups would have moved with it.
trap_time <- ctl_a - ctl_b              # +151

# ---- The two naive answers, which bracket the truth ------------------------
naive_ba <- off_a - off_b               # -665  before-and-after only
naive_ww <- off_a - ctl_a               # -1447 with-and-without at follow-up

# ---- The counterfactual line, for the chart -------------------------------
cf_after <- off_b + ctl_chg             # what the offered group would have
                                        # paid, had it merely followed the
                                        # not-offered group's path

did_tbl_viz <- tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Took part", "Did not take part"), each = 2),
  Cost  = c(off_b, off_a, ctl_b, ctl_a)
)

# The table as a participant would receive it.
did_tbl <- fit_did %>% tidy() %>%
  mutate(` ` = c("(Intercept)", "Round 2", "Enrolled", "Round 2 x Enrolled"),
         CI = sprintf("[%s, %s]", gfmt(conf.low), gfmt(conf.high))) %>%
  select(` `, estimate, std.error, p.value, CI)

# ---- The balance table, which is the honest substitute for a pre-trend test -
# GreenWaste has ONE period before the programme: round 0. A pre-trend test
# needs at least TWO. So it cannot be run, and the deck says so rather than
# implying it was checked. What CAN be shown is how the two groups differed at
# baseline, which is the other half of the same worry.
bal_vars <- c("age_manager", "educ_manager", "staff_size", "foreign_owned",
              "recycling_center_distance", "business_area")
bal_labs <- c(age_manager = "Manager age (years)",
              educ_manager = "Manager education (level)",
              staff_size = "Staff size",
              foreign_owned = "Foreign owned (share)",
              recycling_center_distance = "Distance to recycling centre (m)",
              business_area = "Business area (units)")
bal0 <- tn %>% filter(round == 0)

balance_tbl <- map_dfr(bal_vars, function(v) {
  m <- lm_robust(bal0[[v]] ~ bal0$enrolled)
  tibble(Characteristic = bal_labs[[v]],
         Programme   = mean(bal0[[v]][bal0$enrolled == 1]),
         Comparison  = mean(bal0[[v]][bal0$enrolled == 0]),
         Difference  = coef(m)[2],
         p = tidy(m)$p.value[2])
})
```

# 1 · Compared to what, again

## The GreenWaste programme, one more time

:::::: {.columns}

::::: {.column width="54%"}
**GreenWaste** cut waste-management costs for businesses that took it up. You
met it on Day 1.

Today the question changes. Not *did costs fall* — but **how much of the fall
was the programme**, and how much was going to happen anyway.

The bar is unchanged: **cut costs by at least `r gfmt(RULEBAR)` AED**.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>  <div class="v">GreenWaste, waste-cost reduction</div>
  <div class="k">Outcome</div>    <div class="v">Waste-management costs (AED)</div>
  <div class="k">Method</div>     <div class="v">Difference-in-differences</div>
  <div class="k">The rule</div>   <div class="v">Cut at least 1,000 AED</div>
</div>
```
:::::

::::::

::: {.ask}
Day 1 gave you three questions. Today is question one, **compared to what?**,
with a method that answers it properly.

Before we do the method, what do you already expect the answer to be?
:::

::: {.notes}
Sections 1 and 2 are the "remember" recap: aim for 5 to 10 minutes total. The
room already met GreenWaste, the before-and-after and with-and-without numbers,
and the idea of subtracting one group's change from the other's (on the traffic
cameras in Day 1 Session 2). Move through the recall quickly. The new material
today is reading the table (Section 3) and the one assumption DiD rests on
(Section 4): that is where the time should go.
:::

## What happened to the businesses that took it up

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.9, fig.width=5.6}
tibble(Time = tlev(c("Before", "After")), Cost = c(off_b, off_a)) %>%
  plot_ly(x = ~Time, y = ~Cost, type = "bar",
          marker = list(color = ACCENT, line = list(width = 0)),
          text = ~gfmt(Cost), textposition = "outside",
          textfont = list(size = 15, color = INK), showlegend = FALSE,
          hovertemplate = "%{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 1900))) %>%
  pl_m2(width = PL_WC, height = 275, title = "Businesses that took part")
```
:::::

::::: {.column width="50%"}
::: {.panel}
Businesses that took part paid **`r gfmt(off_b)` AED** before and
**`r gfmt(off_a)` AED** after.
:::

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, before to after</div>
</div>', gfmt(abs(off_chg))))
```

::: {.warn}
A real fall, in the right direction. Below the `r gfmt(RULEBAR)` bar, so on this
reading the programme **fails**.
:::
:::::

::::::

::: {.ask}
Hold that judgement, and hold the number: **`r gfmt(abs(off_chg))` AED**.

Day 1 taught you not to trust a before-and-after. Today you find out what the
honest number is instead.
:::

## `[MENTI PLACEHOLDER]` Did the programme work?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
On the before-and-after reading alone: **did GreenWaste work, and would you
fund it?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Can't tell yet
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
Commit now. "Can't tell yet" is a legitimate answer and I want to see how many
choose it — that number is the point of the next twenty minutes.
:::

# 2 · The idea, in four numbers

## Remember from Day 1: two comparisons, two biases

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Before and after only.** Credits the programme with the **time trend**: whatever
would have moved costs anyway.

On Day 1 this gave **-`r gfmt(abs(naive_ba))` AED**.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Took part vs did not, afterwards only.** Credits the programme with the
**baseline gap**: differences that existed before it started.

On Day 1 this gave **-`r gfmt(abs(naive_ww))` AED**.
:::
:::::

::::::

::: {.ask}
You met both of these yesterday. Two honest-looking methods, two very different
answers: **`r gfmt(abs(naive_ba))`** and **`r gfmt(abs(naive_ww))` AED**.

One is too small, one is too big. Today's method, difference-in-differences,
removes both problems at once. That is its entire job.
:::

::: {.notes}
Recap, not a re-teach: these are the two naive comparisons from Day 1 Session 3,
on this same GreenWaste data. Spend a minute reminding the room why each is
biased, then move on. The new work today is the assumption DiD rests on.
:::

## One sentence to take to a minister

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Compare the change over time in the group that got it,<br>with the change over time in a group that did not.</div>
</div>')
```

::: {.panel}
The difference between those two changes is the estimated effect.

Everything else today is detail on top of that sentence.
:::

::: {.ask}
Say it back to me before we go on. If you can say this sentence, you can read
every DiD table you will ever be handed.
:::

## Your turn: four numbers, two minutes

Difference-in-differences = (change for "took part") − (change for "did not")

```{r}
tibble(
  Group  = c("Took part", "Did not take part", "Difference of the changes"),
  Before = c(gfmt(off_b), gfmt(ctl_b), ""),
  After  = c(gfmt(off_a), gfmt(ctl_a), ""),
  Change = c("?", "?", "?")
) %>%
  kable(align = "lrrr", col.names = c("", "Before", "After", "Change")) %>%
  kable_styling(full_width = FALSE, font_size = 22)
```

::: {.ask .fragment}
Two subtractions. Work out the three missing numbers **before** I show the
answer.

The top two are easy. The bottom one is the method.
:::

::: {.panel .fragment}
**Watch the sign.** "Did not take part" went **up**, not down. So the difference
between the changes is bigger than the fall in the top row alone — which is why
the honest answer will not match your first guess.
:::

## The answer

```{r}
tibble(
  Group  = c("Took part", "Did not take part", "Difference of the changes"),
  Before = c(gfmt(off_b), gfmt(ctl_b), ""),
  After  = c(gfmt(off_a), gfmt(ctl_a), ""),
  Change = c(sprintf("%+.0f", off_chg), sprintf("%+.0f", ctl_chg),
             sprintf("%.0f", did_hand))
) %>%
  kable(align = "lrrr", col.names = c("", "Before", "After", "Change")) %>%
  kable_styling(full_width = FALSE, font_size = 24) %>%
  row_spec(3, bold = TRUE, background = "#E6F1F5")
```

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the difference-in-differences estimate</div>
</div>
<div class="stat-row">
  <div><span class="n">%+.0f</span><span class="k">Took part</span></div>
  <div><span class="n">%+.0f</span><span class="k">Did not take part</span></div>
  <div><span class="n">%.0f</span><span class="k">The difference</span></div>
</div>', sprintf("%.0f", did_hand), off_chg, ctl_chg, did_hand))
```

::: {.ask}
You have just estimated a causal effect with two subtractions. Every DiD table
you are ever handed is this, wrapped in a regression.
:::

## Why that number is bigger than the headline

```{r fig.height=3.1, fig.width=11.5}
bind_rows(
  did_tbl_viz,
  tibble(Time = tlev(c("Before", "After")), Group = "What would have happened",
         Cost = c(off_b, cf_after))
) %>%
  plot_ly(x = ~Time, y = ~Cost, color = ~Group, type = "scatter",
          mode = "lines+markers",
          colors = c("Took part" = ACCENT, "Did not take part" = ACCENT2,
                     "What would have happened" = MUTED),
          line = list(width = 3, dash = c("solid", "solid", "dot")),
          marker = list(size = 12),
          hovertemplate = "%{fullData.name}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2800))) %>%
  pl_m2(height = 300)
```

::: {.panel}
The dotted line is the counterfactual: what the businesses that took part would
have paid, **had they merely followed the other group's path**.

The gap between the dotted line and where "took part" actually ended up is the
DiD estimate, `r gfmt(abs(did_hand))` AED. Everything above the dotted line is
the time trend, not the programme.
:::

::: {.ask}
Hover the dotted line. Compare it with where "took part" finished.

Session 1's World B asked whether a headline was the programme or the tide.
Same question, and now you can answer it.
:::

# 3 · The table an evaluator hands you

## Running it as a regression

:::::: {.columns}

::::: {.column width="50%"}
```{r}
did_tbl %>%
  mutate(across(c(estimate, std.error), ~ round(.x, 0)),
         p.value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coefficient", "Std. error", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(4, bold = TRUE, background = "#E6F1F5")
```
:::::

::::: {.column width="50%"}
::: {.panel}
One regression produces all four numbers from the previous slide.

**Row 4 is the impact.** The interaction of round and enrolment: the extra
change in the group that took part, over and above the change everyone
experienced.
:::

::: {.warn}
Rows 2 and 3 are the **two things you must not report as the effect**:
`r gfmt(abs(coef(fit_did)[["enrolled"]]))` AED is the baseline gap, and
`r gfmt(coef(fit_did)[["round"]])` AED is the time trend.
:::
:::::

::::::

::: {.notes}
Reminder for the room (vocab from Day 1 Session 1): the **coefficient** is the
effect size, here in AED; the **95% confidence interval** is the range of
plausible true values, not a margin of certainty; the **p-value** is how
surprising this result would be if the programme had truly done nothing. Walk row
4 first, then name rows 2 and 3 as the two numbers a careless report would quote
instead. Facilitation question worth asking here: which row would you want on the
front page of the briefing, and which two would a careless evaluator report?
:::

## The number that answers your question

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the estimated cut in costs (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">No</span><span class="k">Clears the bar</span></div>
</div>', gfmt(abs(did_est)), gfmt(did_ci[1]), gfmt(did_ci[2]),
   gfmt(RULEBAR), gfmt(did_avoided_least)))
```

::: {.warn}
The estimate is statistically clear: the confidence interval sits far from zero.
**That is not the question.**

The rule asks for `r gfmt(RULEBAR)` AED. The estimate is
`r gfmt(abs(did_est))`, and even the confidence interval's most generous end
reaches only `r gfmt(did_avoided_most)`. **It does not clear the bar.**
:::

::: {.ask .fragment}
Session 1: *statistically significant does not mean big enough.* Session 2: same
verdict. Here it is a third time.

Three sessions, three results that are real and not sufficient. This is one of
the most useful things you will take home from this week.
:::

# 4 · The one question

## Everything rests on one assumption

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Without the programme, the two groups<br>would have moved in parallel.</div>
</div>')
```

::: {.panel}
The dotted counterfactual line from Section 2 exists only if this is true. It
cannot be **proven** from the result. It can only be made more or less
plausible.
:::

::: {.ask}
This is the question to ask any evaluator presenting DiD, and most reports do
not raise it. Learn to ask it before you look at the estimate.
:::

## Which of these would you trust?

```{r fig.height=3.2, fig.width=11.5}
mk <- function(lab, tr, ct) {
  tibble(panel = lab, t = rep(1:4, 2),
         grp = rep(c("Programme", "Comparison"), each = 4), y = c(tr, ct))
}
bind_rows(
  mk("A. Moving in parallel",  c(40, 42, 44, 46), c(50, 52, 54, 56)),
  mk("B. Already diverging",   c(40, 41, 42, 43), c(50, 53, 56, 59)),
  mk("C. Already converging",  c(40, 44, 48, 52), c(50, 52, 54, 56))
) %>%
  plot_ly(x = ~t, y = ~y, color = ~grp, frame = ~panel,
          type = "scatter", mode = "lines+markers",
          colors = c(Programme = ACCENT, Comparison = ACCENT2),
          line = list(width = 3), marker = list(size = 9),
          hovertemplate = "%{fullData.name}: %{y}<extra></extra>") %>%
  animation_opts(frame = 900, transition = 0) %>%
  animation_slider(currentvalue = list(prefix = "Panel: "),
                   font = list(size = 11)) %>%
  layout(xaxis = ax("Periods before the programme started", zeroline = FALSE,
                    dtick = 1),
         yaxis = ax("Outcome", range = c(35, 62))) %>%
  pl_m2(height = 310, t = 34, b = 88)
```

::: {.ask}
Press play, or drag the slider. In which panel would difference-in-differences
give you a trustworthy number?
:::

::: {.warn .fragment}
Only **A**. In B the gap was already widening and in C already closing, so part
of what the method calls an effect was going to happen anyway.

Note the directions: **B would overstate** the effect, **C would understate**
it. The bias is not always in the flattering direction.
:::

## The honest answer for this data

:::::: {.columns}

::::: {.column width="54%"}
::: {.warn}
We measured costs at **two points**: once before the programme, once after.

Testing whether the groups were already diverging needs **at least two
observations before it started**. We have one.
:::

::: {.panel}
So the parallel trends assumption **cannot be tested with this data**.

Not "was not tested". **Cannot be.**
:::
:::::

::::: {.column width="46%"}
::: {.ask}
This is the normal situation, not an unusual one. Most evaluations you will be
handed have exactly this limitation. So what do you do instead?
:::

::: {.panel .fragment}
Ask for the things that stand in for the test:

**Which comparison group did you choose, and why?**

**What differed between the groups at baseline?**
:::
:::::

::::::

## How far apart did they start?

:::::: {.columns}

::::: {.column width="56%"}
```{r}
balance_tbl %>%
  mutate(across(c(Programme, Comparison), ~ round(.x, 2)),
         Difference = round(Difference, 2), p = round(p, 2)) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("Characteristic", "Took part", "Did not",
                      "Difference", "p-value")) %>%
  kable_styling(full_width = FALSE, font_size = 14)
```
:::::
:::::

::::: {.column width="44%"}
::: {.panel}
You read one of these on Day 1: a balance table shows how the two groups
differed **before** anything happened. It is not a pre-trend test (it says
nothing about direction of travel), but it is the closest thing this data allows.
:::

::: {.warn}
Read it for two things: are the differences **small**, and are the p-values
**not tiny**?
:::
:::::

::::::

::: {.ask}
If these two groups differ systematically at baseline **and** were free to
choose whether to enrol, the parallel trends assumption is doing a lot of
unpaid work. Say so when you write your recommendation.
:::

::: {.notes}
Recap, not new teaching: they built a balance table on Day 1 Session 3. Here it
earns its place by making the parallel-trends point concrete, because the
pre-trend test itself cannot be run on two-wave data. Reminder for the room: the
**p-value** answers "how surprising is this baseline gap if the two groups were
really the same?", so a tiny p-value is the worry here, not the reassurance.
:::

## What to ask when it cannot be tested

```{=html}
<div class="facts">
  <div class="k">Ask</div> <div class="v">"Which comparison group did you use, and why that one?"</div>
  <div class="k">Ask</div> <div class="v">"What differed between the groups before the programme started?"</div>
  <div class="k">Ask</div> <div class="v">"Is there any earlier data, even one extra year, that could show the trend?"</div>
  <div class="k">Then</div> <div class="v">treat the estimate as a plausible range, not a precise number</div>
</div>
```

::: {.panel}
None of these questions requires you to run anything. They are the questions a
commissioner is entitled to ask, and an evaluator who cannot answer them has
not finished the job.
:::

# 5 · Put AI to work

## AI Snapshot: explaining the table

::: {.panel}
Paste the table from Section 3 into an AI tool with this prompt:

**"Explain this regression table to a non-technical decision-maker, and say what
should be checked before believing the result."**
:::

::: {.ai-output}
The table reports a difference-in-differences estimate. **The programme reduced
costs by 816 AED**, and the effect is highly statistically significant
(p < 0.001), **so the programme was a success**. [The confidence interval does
not include zero, which confirms the finding is robust]{.flag}.

[The parallel trends assumption has been satisfied]{.flag}. [The result clears
the 1,000 AED threshold required for scale-up]{.flag}.
:::

::: {.ask}
Mark every sentence you can check against the table, and every sentence the
table does **not** support. Four minutes.
:::

## Debrief: three errors, one of them serious

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "So the programme was a success."** It answers *is the effect non-zero*,
not *does it clear the 1,000 AED bar*. Session 1, Session 2, and now here: the
same substitution, every time.
:::

::: {.warn}
**2 · "The interval does not include zero, so the finding is robust."**
"Not zero" and "big enough" are different claims. The confidence interval is
[`r gfmt(did_ci[1])`, `r gfmt(did_ci[2])`].
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "The parallel trends assumption has been satisfied."** The serious one.

Nothing in this table can show that. **We established two slides ago that it
cannot be tested with this data at all.** The model asserted a test result for a
test that was never run and could not be.
:::
:::::

::::::

::: {.panel}
That third error is Session 2's fabricated-robustness failure, in a new place.
**This is what to watch for: confident claims about checks that no one
performed.**
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain this regression table to a non-technical decision-maker and say what
> should be checked before believing the result.

The model answers, confidently, on all three points — including the one it
cannot know.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> We have a difference-in-differences table. The data has **two waves only, one
> before and one after**, so no pre-trend test is possible. The decision rule is
> 1,000 AED. Explain what the table supports and, separately, list what it
> cannot tell us. **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Give the model the constraint — two waves, no pre-trend test — and see whether
error 3 survives. Supply what it cannot know, and it stops inventing.
:::

# 6 · The Abu Dhabi case

## `[PLACEHOLDER]` The DOH evaluation

::: {.warn}
**Awaiting the real findings.** The Abu Dhabi Department of Health evaluation
has not been shared with this workspace yet, so nothing here is presented as its
result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>     <div class="v">[PLACEHOLDER]</div>
  <div class="k">Design</div>        <div class="v">Difference-in-differences</div>
  <div class="k">Outcome</div>       <div class="v">[PLACEHOLDER]</div>
  <div class="k">Decision rule</div><div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure above is fixed. When the DOH findings arrive, they drop into these
four fields and the three slides that follow, without redesigning the session.
:::

## Reading a DOH result: the same three questions

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Which row is the DiD estimate, and does it clear the rule, not just zero?</div>
  <div class="k">2</div> <div class="v">How many waves before the programme? If one, can parallel trends be tested at all?</div>
  <div class="k">3</div> <div class="v">What differed between the groups before the programme started?</div>
</div>
```

::: {.ask}
Same three questions as the GreenWaste table. The programme changes; the
interrogation does not.

That is the point of reading the two side by side.
:::

# 7 · Close

## Three questions to ask any evaluator presenting DiD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Which row is the estimate, and does its interval clear our rule, not just zero?</div>
  <div class="k">2</div> <div class="v">How many periods before the programme, and were the groups already moving together?</div>
  <div class="k">3</div> <div class="v">Which comparison group, chosen how, and how did it differ at baseline?</div>
</div>
```

::: {.panel}
Question 1 you learned on Day 1. Questions 2 and 3 are the ones specific to this
method, and they are the ones that decide whether the number means anything.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1 S1</div> <div class="v">compared to what? &middot; the three questions</div>
  <div class="k">Day 1 S2</div> <div class="v">reading a descriptive output &middot; -1.13 against a 2.0 rule</div>
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons and an RCT &middot; the design that removes the difference</div>
  <div class="k">Today</div>  <div class="v">DiD &middot; -816 against a 1,000 rule, with the assumption that cannot be tested</div>
</div>
```

::: {.ask}
Notice the pattern across the week so far: every result has been real, and none
of them has cleared its bar.

Ask yourself what that means for the way evaluations get commissioned. Then
bring it to the next session, where the identification comes from a **cut-off**
rather than a comparison group.
:::
