---
title: "What Can You Read? What Might Be Wrong?"
subtitle: "Module 2 · Day 3, Session 2: Reading the Figures That Carry the Claim"
author: "3ie"
always_allow_html: yes
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR, the same set-up as Oct12_session2.qmd. Three slides
# carry live cells ("Try it"); the CSVs under webr: resources must sit next to
# the published .html, and webR needs an internet connection on the day. The
# plotly charts are baked at render time as before.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: []
  render-df: kable
  resources:
    - evaluation_data_GreenWaste.csv
    - evaluation_data_SchoolZoneRCT.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

# ---- Tables: plain knitr::kable, styled by theme_dge.scss -------------------
# No kableExtra: its JavaScript needs jQuery, which live-revealjs does not load.
# font_size in px; hl = rows to shade (hl_bold to bold them); widths = CSS widths
# per column; note = a line under the table. The table is centred.
HL_BLUE <- "#E4EBF3"; HL_RED <- "#F8E9EA"
tbl <- function(df, ..., font_size = NULL, hl = NULL, hl_bg = HL_BLUE,
                hl_bold = FALSE, widths = NULL, note = NULL) {
  h <- as.character(knitr::kable(df, format = "html", ...))
  fs <- if (is.null(font_size)) "" else sprintf(" font-size:%dpx;", font_size)
  h <- sub("<table", sprintf('<table style="margin-left:auto; margin-right:auto;%s"', fs),
           h, fixed = TRUE)
  if (!is.null(widths))
    h <- sub("(<table[^>]*>)", paste0("\\1<colgroup>",
             paste0('<col style="width:', widths, ';">', collapse = ""), "</colgroup>"), h)
  if (length(hl)) {
    parts <- strsplit(h, "<tbody>", fixed = TRUE)[[1]]
    rows  <- strsplit(parts[2], "<tr>", fixed = TRUE)[[1]]
    st <- sprintf("background:%s;%s", hl_bg, if (hl_bold) " font-weight:700;" else "")
    for (i in hl) {
      rows[i + 1] <- gsub('<td style="', paste0('<td style="', st, " "), rows[i + 1], fixed = TRUE)
      rows[i + 1] <- gsub("<td>", paste0('<td style="', st, '">'), rows[i + 1], fixed = TRUE)
    }
    h <- paste0(parts[1], "<tbody>", paste(rows, collapse = "<tr>"))
  }
  if (!is.null(note))
    h <- paste0(h, sprintf('\n<p style="text-align:center; font-size:0.75em; color:%s; margin-top:0.3em;">%s</p>',
                           MUTED, note))
  fence <- strrep("`", 3)
  knitr::asis_output(paste0("\n", fence, "{=html}\n", h, "\n", fence, "\n"))
}

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ===========================================================================
# The figures this session interrogates
# ===========================================================================
# Every figure is drawn from the same three datasets the week has already used,
# so participants are reading real structure rather than decoration. The
# session's method is to show the SAME data twice and ask what changed.

# ---- GreenWaste: the two groups at baseline -------------------------------
gw  <- read.csv("./evaluation_data_GreenWaste.csv")
base <- gw %>% filter(treatment_neighborhood == 1, round == 0)
enr  <- base$waste_management_costs[base$enrolled == 1]
non  <- base$waste_management_costs[base$enrolled == 0]

mean_enr <- mean(enr); mean_non <- mean(non)
gap      <- mean_non - mean_enr
pooled   <- sd(c(enr, non))
sd_gap   <- gap / pooled

# ---- The difference-in-differences, carried over from Session 1 -----------
tn <- gw %>% filter(treatment_neighborhood == 1)
grp <- tn %>% group_by(enrolled, round) %>%
  summarise(mc = mean(waste_management_costs), .groups = "drop")
gv  <- function(e, r) grp$mc[grp$enrolled == e & grp$round == r]
ba_claim  <- gv(1, 0) - gv(1, 1)                     # 665, before-and-after
did_est   <- abs((gv(1, 1) - gv(1, 0)) - (gv(0, 1) - gv(0, 0)))   # 816

# ---- The cost-benefit model, from Session 1 --------------------------------
COST_PER_BUSINESS <- 1800
BENEFIT_PER_YEAR  <- 816
BENEFIT_LAG       <- 1
DISCOUNT_BASE     <- 0.05

# Cumulative saving against cumulative cost, nominal and discounted. A payment
# at the end of year y is discounted by (1 + r)^y, and there is no benefit in
# year 1, so the five years of benefit are years 2 to 6.
cb_years <- tibble(
  year      = 1:8,
  nominal   = ifelse(year == 1, 0, BENEFIT_PER_YEAR),
  discounted = ifelse(year == 1, 0, BENEFIT_PER_YEAR / (1 + DISCOUNT_BASE)^year)
) %>%
  mutate(cum_nominal = cumsum(nominal), cum_discounted = cumsum(discounted))

cb_nom_total <- sum(cb_years$nominal[2:6])
cb_dis_total <- sum(cb_years$discounted[2:6])
cb_ratio     <- cb_dis_total / COST_PER_BUSINESS

# ---- School-zone child collisions: the denominator problem -----------------
sz <- read.csv("./evaluation_data_SchoolZoneRCT.csv")
sc <- sz %>% group_by(segment_id, arm) %>%
  summarise(base = child_collisions[round == 0],
            follow = child_collisions[round == 1], .groups = "drop")

zero_base  <- sum(sc$base == 0)
sc_pos     <- sc %>% filter(base > 0) %>%
  mutate(pct = 100 * (follow - base) / base, absolute = follow - base)

# The two different "biggest wins", which do not contain the same segments.
big_pct <- sc_pos %>% arrange(pct) %>% slice_head(n = 6)
big_abs <- sc_pos %>% arrange(absolute) %>% slice_head(n = 6)
overlap <- length(intersect(big_pct$segment_id, big_abs$segment_id))

# The pooled percentage against the average of the percentages, which is the
# difference between weighting by size and giving every segment equal say.
pooled_pct   <- 100 * (sum(sc$follow) - sum(sc$base)) / sum(sc$base)
# Only segments with a non-zero baseline have a percentage change to average.
average_pct  <- mean(sc_pos$pct)
n_pct        <- nrow(sc_pos)

# ---- Traffic cameras: total count against rate ----------------------------
cam <- read.csv("./evaluation_data_TrafficCameras.csv")
sect <- cam %>% group_by(sector_id) %>%
  summarise(collisions = sum(injury_collisions),
            km = sum(road_length_km), .groups = "drop") %>%
  mutate(per_km = collisions / km)

top_count <- sect %>% arrange(desc(collisions)) %>% slice_head(n = 5)
top_rate  <- sect %>% arrange(desc(per_km))     %>% slice_head(n = 5)
sect_overlap <- length(intersect(top_count$sector_id, top_rate$sector_id))

# ---- The severity mix, for the pie ----------------------------------------
sev <- c(minor = sum(sz$minor_collisions),
         injury = sum(sz$injury_collisions),
         child = sum(sz$child_collisions))

# ---- The rules this session teaches ---------------------------------------
# Five habits, each demonstrated on a figure rather than asserted.
fig_rules <- tribble(
  ~n, ~Rule, ~Why,
  "1", "Read the axis before the shape",
       "A truncated axis makes a real gap look decisive. The gap is the same number either way; only your impression of it changes.",
  "2", "Ask what the bar is averaging over",
       "A mean can sit between two groups that barely overlap. It summarises the group and may describe nobody in it.",
  "3", "Check the denominator",
       "A percentage needs a base. Where the base is zero or tiny, the biggest percentage change belongs to the smallest number.",
  "4", "Compare changes, not levels",
       "Levels carry whatever the two groups already differed by. The comparison the chart invites is often not the one that answers the question.",
  "5", "Discount the future in front of the reader",
       "Undiscounted sums look bigger and are easier to draw. The gap between the two lines comes from the discount rate."
)
```

# 1 · Why the picture matters {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Where we are

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the three "Try it"
# cells use is defined here, so each live cell stays a few lines long.
gw <- read.csv("evaluation_data_GreenWaste.csv")
b0 <- subset(gw, treatment_neighborhood == 1 & round == 0)
costs <- c(Enrolled       = mean(b0$waste_management_costs[b0$enrolled == 1]),
           "Not enrolled" = mean(b0$waste_management_costs[b0$enrolled == 0]))
sz <- read.csv("evaluation_data_SchoolZoneRCT.csv")
sc <- merge(subset(sz, round == 0, c(segment_id, child_collisions)),
            subset(sz, round == 1, c(segment_id, child_collisions)), by = "segment_id")
names(sc) <- c("segment_id", "before", "after")
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Layout only: margins, font size and muted axes for a projector.
clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)

# Cumulative saving, year by year, against the one-off cost.
plot_payback <- function(today, cost = 1800) {
  clean_par()
  plot(seq_along(today), cumsum(today), type = "b", pch = 19, lwd = 4, col = ACCENT,
       bty = "n", ylim = c(0, 4500), xlab = "Years after rollout",
       ylab = "Saving so far (AED)")
  abline(h = cost, col = ALERT, lwd = 3)
  text(1, cost, "what it cost", col = ALERT, pos = 3, adj = 0)
}
```

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">is the number answering your question, and is it big enough?</div>
  <div class="k">Day 2</div>  <div class="v">which method produced it, and what does that method assume?</div>
  <div class="k">This morning</div> <div class="v">what is it worth, and which assumption decides that?</div>
  <div class="k">Now</div>     <div class="v">the form the claim actually reaches a decision-maker in</div>
</div>
```

::: {.panel}
Every session this week has ended with a number and the assumptions behind it.
A decision-maker usually sees neither directly: they see a chart and a sentence
about it.

An assumption that is left out of the chart is hard to question.
:::

::: {.ask}
This session is about **reading charts as evidence**: the same interrogation
you have practised all week, applied to the form evidence usually arrives in.
:::

## Six pairs, two questions

::: {.panel}
Each section of this session shows the same data drawn two ways. One version is
fair; the other invites a wrong reading, or cannot be read correctly at all.
:::

::: {.ask}
**For every pair, in groups:**

1. What is this figure telling you?
2. What might be wrong or misleading?
:::

## The same data, two pictures

::: {.panel}
Among the businesses that were offered GreenWaste, the ones that enrolled paid **`r gfmt(mean_enr)` AED** in waste costs at baseline and the
ones that did not paid **`r gfmt(mean_non)` AED**. A gap of
**`r gfmt(gap)` AED**.

The next two figures draw the same gap.
:::

```{r fig.height=2.6, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside", cliponaxis = FALSE,
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2400))) %>%
  pl_m2(width = PL_W, height = 250)
```

::: {.ask}
This figure is fair: the axis starts at zero and the values are labelled.
Keep it in mind for the next slide.
:::

## The same gap, drawn to fill the frame

```{r fig.height=2.6, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside", cliponaxis = FALSE,
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)",
                    range = c(1300, 2200), zeroline = FALSE)) %>%
  pl_m2(width = PL_W, height = 250, t = 40)
```

::: {.warn}
**Same numbers and labels, different impression.**

The gap is `r gfmt(gap)` AED in both figures. Here the y-axis starts at
`r gfmt(1300)`, so a `r sprintf("%.0f%%", 100 * gap / mean_non)` difference fills
most of the frame.
:::

::: {.ask}
**In groups, two minutes.** Which figure would you put in a briefing to a
minister? Which would you put in an annex? Is your answer the same for both?
:::

## Try it: where should the axis start?

```{webr}
start <- 1300    # where the y-axis begins: try 0, then 1400
clean_par()
barplot(costs, ylim = c(start, 2300), xpd = FALSE, border = NA,
        col = c(ACCENT, ACCENT2), ylab = "Mean annual waste cost (AED)")
```

::: {.ask}
Before each run, predict: what happens to the gap when the axis starts at 0? At
1,400?
:::

::: {.notes}
Run it as written first: 1300 reproduces the previous slide. Then change `start`
to 0 and run, then 1400. The costs never change; only the frame around them does.
:::

## Rule 1: read the axis before the shape

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
A truncated axis works like a **magnifying glass**. That is useful when the
question is whether a difference exists at all.

It misleads when the reader does not notice that the axis has been cut.
:::

::: {.ask}
The practical test, and you can do it in two seconds:

**Cover the bars. Read the y-axis. Uncover.**
:::
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">The gap</div>     <div class="v">630 AED &middot; the same in both figures</div>
  <div class="k">As a share</div>  <div class="v">30% of the non-enrolled mean</div>
  <div class="k">In SD units</div> <div class="v">1.13 standard deviations</div>
  <div class="k">The accurate line</div> <div class="v">"a substantial gap on a truncated axis"</div>
</div>
```
:::::

::::::

::: {.ask}
The third line **supports** the claim: a gap of `r sprintf("%.2f", sd_gap)`
standard deviations is large by the usual conventions. The problem with the second
figure is that it asks you to judge the gap from the drawing instead of the number.
:::

# 2 · What the bar is hiding {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## A mean is not a group

::: {.panel}
The same comparison as a **bar chart of means**, one number per group. It is one
of the most common figures in evaluation reports.

Two groups with the same mean can have little else in common.
:::

```{r fig.height=2.5, fig.width=5.6}
tibble(group = c("Enrolled", "Not enrolled"), cost = c(mean_enr, mean_non)) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside", cliponaxis = FALSE,
          textfont = list(size = 16, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2400))) %>%
  pl_m2(width = PL_W, height = 230)
```

::: {.ask}
**Before the next slide:** what share of the enrolled businesses do you think
paid **more** than the average non-enrolled business? Write a number down.
:::

## The same comparison, as distributions

```{r fig.height=2.4, fig.width=11.4}
tibble(cost = c(enr, non),
       group = c(rep("Enrolled", length(enr)), rep("Not enrolled", length(non)))) %>%
  plot_ly(x = ~cost, color = ~factor(group, levels = c("Enrolled", "Not enrolled")),
          colors = c(ACCENT, ACCENT2), type = "histogram", nbinsx = 46,
          opacity = 0.72, histnorm = "probability density",
          hovertemplate = "%{x:.0f} AED<extra></extra>") %>%
  layout(barmode = "overlay",
         xaxis = ax("Annual waste cost (AED)"),
         yaxis = ax("Share of the group"),
         shapes = list(
           list(type = "line", y0 = 0, y1 = 1, yref = "paper",
                x0 = mean_enr, x1 = mean_enr, line = list(color = ACCENT, width = 3)),
           list(type = "line", y0 = 0, y1 = 1, yref = "paper",
                x0 = mean_non, x1 = mean_non, line = list(color = ACCENT2, width = 3, dash = "dot"))),
         annotations = list(
           list(x = mean_enr, y = 1.04, yref = "paper", xanchor = "right",
                showarrow = FALSE, text = "enrolled mean",
                font = list(size = 12, color = ACCENT)),
           list(x = mean_non, y = 1.04, yref = "paper", xanchor = "left",
                showarrow = FALSE, text = "not enrolled mean",
                font = list(size = 12, color = ACCENT2)))) %>%
  pl_m2(width = PL_W, height = 230, t = 40)
```

::: {.warn}
**The two groups overlap very little.** Only
`r sprintf("%.1f%%", 100 * mean(enr > mean_non))` of enrolled businesses
(`r gfmt(sum(enr > mean_non))`) paid more than the average non-enrolled business.
A bar chart of means cannot show this.
:::

::: {.ask .fragment}
**Look back at the number you wrote down.** Did you expect
`r sprintf("%.1f%%", 100 * mean(enr > mean_non))`?
:::

::: {.notes}
A mean tells you where a group sits on average. It cannot tell you how many of
them sit somewhere else. The reverse figure: `r sprintf("%.1f%%", 100 * mean(non < mean_enr))`
of the comparison group paid less than the enrolled average.
:::

## Rule 2: ask what the bar is averaging over

```{=html}
<div class="facts">
  <div class="k">A mean</div>        <div class="v">a summary of a group &middot; not a description of anyone in it</div>
  <div class="k">Two groups can</div><div class="v">share a mean and share nothing else</div>
  <div class="k">A bar chart</div>   <div class="v">cannot show how much two groups overlap</div>
  <div class="k">So ask</div>        <div class="v">"how much do the two groups overlap?" &middot; it is answerable</div>
</div>
```

::: {.ask}
This is the same instinct as Day 1's confidence interval and Day 2's baseline
balance table. **Every summary leaves something out, so ask what.**

Here the missing information supported the finding. That will not always be the
case.
:::

# 3 · Check the denominator {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The change nobody can compute

::: {.panel}
The school-zone trial measured child collisions at 400 segments, before and
after. An analyst wants to report **the percentage change in child collisions at
the typical segment.**

For some segments it cannot be calculated.
:::

```{r}
sc %>%
  mutate(`Baseline child collisions` = cut(base, c(-Inf, 0, 1, 2, 5, 9, Inf),
           labels = c("0", "1", "2", "3 to 5", "6 to 9", "10 or more"))) %>%
  count(`Baseline child collisions`) %>%
  mutate(`Share of segments` = sprintf("%.1f%%", 100 * n / sum(n))) %>%
  select(-n) %>%
  tbl(align = "rr", font_size = 15, hl = 1, hl_bg = HL_RED, hl_bold = TRUE)
```

::: {.warn}
**`r gfmt(zero_base)` of the 400 segments recorded no child collisions at
baseline.** For those segments a percentage change is undefined, because it
would mean dividing by zero.

Any figure that reports a percentage change has dropped these
`r gfmt(zero_base)` segments or handled them somehow. Check whether it says which.
:::

## Two charts, two different "biggest wins"

:::::: {.columns}

::::: {.column width="50%"}
**The largest percentage reductions**

```{r fig.height=2.3, fig.width=5.4}
big_pct %>%
  mutate(lab = sprintf("%s: %d to %d", segment_id, base, follow)) %>%
  plot_ly(x = ~pct, y = ~factor(lab, levels = rev(lab)), type = "bar",
          orientation = "h", marker = list(color = ACCENT),
          text = ~sprintf("%.0f%%", pct), textposition = "inside",
          insidetextanchor = "start", textfont = list(color = "white", size = 13),
          hovertemplate = "%{y}<extra></extra>") %>%
  layout(xaxis = ax("Change in child collisions (%)", range = c(-100, 0)),
         yaxis = ax("")) %>%
  pl_m2(width = PL_WC, height = 220, t = 10, b = 46) %>%
  layout(margin = list(l = 130))
```
:::::

::::: {.column width="50%"}
**The largest absolute reductions**

```{r fig.height=2.3, fig.width=5.4}
big_abs %>%
  mutate(lab = sprintf("%s: %d to %d", segment_id, base, follow)) %>%
  plot_ly(x = ~absolute, y = ~factor(lab, levels = rev(lab)), type = "bar",
          orientation = "h", marker = list(color = ACCENT2),
          text = ~sprintf("%+d", as.integer(absolute)), textposition = "inside",
          insidetextanchor = "start", textfont = list(color = "white", size = 13),
          hovertemplate = "%{y}<extra></extra>") %>%
  layout(xaxis = ax("Change in child collisions (number)", range = c(-13, 0)),
         yaxis = ax("")) %>%
  pl_m2(width = PL_WC, height = 220, t = 10, b = 46) %>%
  layout(margin = list(l = 130))
```
:::::

::::::

::: {.warn .fragment}
**These two lists have `r if (overlap == 0) "no segments" else if (overlap == 1) "one segment" else paste(overlap, "segments")` in common.**
Every segment on the left started with two to four child collisions. The segments
on the right removed `r gfmt(-sum(big_abs$absolute))` between them.
:::

::: {.ask .fragment}
**Neither is wrong.** The left answers *where did it work best?*; the right,
*where did it prevent the most harm?* Ask which question a report's chart answers.
:::

## Try it: one change, two averages

```{webr}
# 1. Pooled: all collisions after, against all before
100 * (sum(sc$after) - sum(sc$before)) / sum(sc$before)

# 2. The average of each segment's own percentage change
#    (only segments that had a collision to begin with)
has_base <- sc$before > 0
mean(100 * (sc$after - sc$before)[has_base] / sc$before[has_base])
```

::: {.ask}
Predict before I run it: will the two answers be close?
:::

::: {.notes}
Expected: about -32.7% pooled and about -10.8% averaged. The next slide explains
the gap: the pooled figure weights each segment by its collisions; the average
gives every segment equal weight, so small bases pull it around.
:::

## Rule 3: check the denominator

```{=html}
<div class="facts">
  <div class="k">A percentage</div>   <div class="v">is a claim about a base &middot; find the base before you believe it</div>
  <div class="k">Small bases</div>    <div class="v">produce the largest percentages and the smallest changes</div>
  <div class="k">Zero bases</div>     <div class="v">produce no number at all &middot; but the figure still has a bar there</div>
  <div class="k">So ask</div>         <div class="v">"how many segments is that percentage built on?"</div>
</div>
```

::: {.ask}
The pooled change across all 400 segments is
`r sprintf("%.1f%%", pooled_pct)`. The average of the `r n_pct` percentage changes
that can be computed is `r sprintf("%.1f%%", average_pct)`.

**They differ by `r sprintf("%.0f", abs(average_pct - pooled_pct))` percentage
points** because the pooled figure weights each segment by its collisions and the
average gives every segment equal weight.
:::

# 4 · Levels or changes {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The figure that flatters the programme

::: {.panel}
The headline claim from Day 2 Session 1, drawn as a chart. After the programme,
the enrolled businesses paid **`r gfmt(gv(1, 1))` AED** and the comparison
businesses paid **`r gfmt(gv(0, 1))` AED**.

A difference of `r gfmt(gv(0, 1) - gv(1, 1))` AED. **The programme cut costs.**
:::

```{r fig.height=2.5, fig.width=6.6}
tibble(group = c("Enrolled", "Not enrolled"),
       cost  = c(gv(1, 1), gv(0, 1))) %>%
  plot_ly(x = ~group, y = ~cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(cost), textposition = "outside", cliponaxis = FALSE,
          textfont = list(size = 15, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost, after (AED)", range = c(0, 2700))) %>%
  pl_m2(width = PL_W, height = 230)
```

::: {.ask}
Every number here is correct, and a figure like this could appear in a
reputable report.

**What question does it answer, and what question did the evaluation ask?**
:::

## The same two groups, with where they started

```{r fig.height=2.4, fig.width=11.4}
tibble(
  Group = factor(rep(c("Enrolled", "Not enrolled"), each = 2),
                 levels = c("Enrolled", "Not enrolled")),
  Time  = factor(rep(c("Before", "After"), 2), levels = c("Before", "After")),
  Cost  = c(gv(1, 0), gv(1, 1), gv(0, 0), gv(0, 1))
) %>%
  plot_ly(x = ~Time, y = ~Cost, color = ~Group, colors = c(ACCENT, ACCENT2),
          type = "scatter", mode = "lines+markers",
          line = list(width = 3), marker = list(size = 11),
          hovertemplate = "%{fullData.name}, %{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Mean annual waste cost (AED)", range = c(0, 2700))) %>%
  pl_m2(width = PL_W, height = 230)
```

::: {.warn .fragment}
**The comparison group started `r gfmt(gv(0, 0) - gv(1, 0))` AED worse off, and its
costs were rising anyway.** The gap on the previous slide mixes the programme
effect with that starting gap and with a trend unrelated to the programme. The
DiD estimate is `r gfmt(did_est)` AED, not `r gfmt(gv(0, 1) - gv(1, 1))`.
:::

::: {.ask .fragment}
Day 1 showed this problem as a with-and-without comparison. Drawn as a bar chart,
it looks more convincing.
:::

::: {.notes}
The analyst need not have been careless: the level comparison is the easiest one
to draw.
:::

## Rule 4: compare changes, not levels

```{=html}
<div class="facts">
  <div class="k">Levels carry</div>   <div class="v">whatever the two groups already differed by &middot; and whatever moved on its own</div>
  <div class="k">Changes remove</div> <div class="v">the starting gap &middot; if the groups were on parallel paths</div>
  <div class="k">Both are easy</div>  <div class="v">to compute &middot; only one answers "did the programme work?"</div>
  <div class="k">So ask</div>         <div class="v">"what did the comparison group look like before?"</div>
</div>
```

::: {.ask}
A four-number test, and you can run it on any before-and-after figure:

**Do I have the before for both groups? If not, the chart cannot tell me the
programme's effect, whatever it is titled.**
:::

# 5 · The assumption the picture hides {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Two ways to add up the same saving

::: {.panel}
From this morning: GreenWaste cost **`r gfmt(COST_PER_BUSINESS)` AED** per
business and saved **`r gfmt(BENEFIT_PER_YEAR)` AED** a year, for five years
starting one year after rollout. The two lines add up that saving in two ways.
:::

```{r fig.height=2.4, fig.width=11.4}
cb_years %>%
  plot_ly(x = ~year, y = ~cum_nominal, name = "Added up as they come",
          type = "scatter", mode = "lines+markers",
          line = list(color = ACCENT2, width = 3, dash = "dot"),
          marker = list(size = 9),
          hovertemplate = "Year %{x}: %{y:,.0f} AED<extra></extra>") %>%
  add_trace(y = ~cum_discounted, name = "Discounted to today's money",
            type = "scatter", mode = "lines+markers",
            line = list(color = ACCENT, width = 3), marker = list(size = 9),
            hovertemplate = "Year %{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("Years after rollout"),
         yaxis = ax("Cumulative saving per business (AED)"),
         shapes = list(list(type = "line", x0 = 0.5, x1 = 8.5, y0 = COST_PER_BUSINESS,
                            y1 = COST_PER_BUSINESS, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 8.4, y = COST_PER_BUSINESS, xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "what the programme cost", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_W, height = 230, t = 20)
```

::: {.warn}
**Both lines use the same measured saving.** The upper line counts a future
saving as worth the same as one today and reaches `r gfmt(cb_nom_total)` AED. The
lower line discounts at 5% and reaches `r gfmt(cb_dis_total)`. The
`r gfmt(cb_nom_total - cb_dis_total)` AED gap is the **discount rate** at work.
:::

## Try it: change the discount rate

```{webr}
rate   <- 0.05                        # try 0, then 0.10
saving <- c(0, rep(816, 5))           # nothing in year 1, then five years
today  <- saving / (1 + rate)^(1:6)   # each year's saving in today's money
round(sum(today))                     # the total, against a cost of 1,800
plot_payback(today)
```

::: {.ask}
Predict: what discount rate would it take for the total to fall below the
1,800 AED cost?
:::

::: {.notes}
Run at 0.05 (3,365), then 0 (4,080: the undiscounted line), then 0.10. The
answer to the question is about 24%, the flip rate from this morning's session.
:::

## Rule 5: discount in front of the reader

:::::: {.columns}

::::: {.column width="52%"}
```{=html}
<div class="facts">
  <div class="k">Five years, added up</div>  <div class="v">4,080 AED per business</div>
  <div class="k">Five years, discounted</div><div class="v">3,365 AED per business</div>
  <div class="k">The difference</div>        <div class="v">715 AED &middot; the discount rate, not the programme</div>
  <div class="k">The ratio</div>             <div class="v">3,365 / 1,800 = 1.87</div>
</div>
```
:::::

::::: {.column width="48%"}
::: {.ask}
The undiscounted line is easier to explain and looks better. Leaving out the
discount is still a choice, and it favours the programme, but nobody is asked to
defend it the way they would defend a 5% rate.
:::
:::::

::::::

::: {.warn}
The cost line is crossed in year
`r sprintf("%.1f", 3.21)` on the undiscounted view and year
`r sprintf("%.1f", 3.53)` on the discounted one. **The discount rate alone moves
payback by about four months.**
:::

# 6 · The pie {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What is this pie not telling you?

::: {.panel}
The school-zone evaluation reports its collisions by severity, as a pie chart.
:::

```{r fig.height=2.9, fig.width=5.2}
tibble(severity = factor(c("Minor", "Injury", "Child"),
                         levels = c("Minor", "Injury", "Child")),
       n = as.numeric(sev)) %>%
  plot_ly(labels = ~severity, values = ~n, type = "pie",
          marker = list(colors = c(ACCENT2, ACCENT, ALERT),
                        line = list(color = "white", width = 2)),
          textinfo = "label+percent", textfont = list(size = 14, color = "white"),
          hovertemplate = "%{label}: %{value:,}<extra></extra>") %>%
  pl_m2(width = PL_WC, height = 260, b = 20)
```

::: {.ask}
**In groups, three minutes.** Write down two things this chart tells you and two
things it does not.
:::

## What the pie cannot show

:::::: {.columns}

::::: {.column width="52%"}
```{=html}
<div class="facts">
  <div class="k">It shows</div>     <div class="v">which kind of collision is most common &middot; minor ones</div>
  <div class="k">It hides</div>     <div class="v">that the categories overlap &middot; a child collision is also an injury collision</div>
  <div class="k">It hides</div>     <div class="v">any change over time &middot; both rounds pooled into one number</div>
  <div class="k">It hides</div>     <div class="v">the counts &middot; a pie shows shares, not how many collisions</div>
</div>
```
:::::

::::: {.column width="48%"}
::: {.warn}
**The first problem is the most serious.** A pie chart says its slices add up to
one whole. These do not: child collisions are also counted as injury collisions,
so the slices overlap and the percentages are wrong.
:::
:::::

::::::

::: {.ask}
The truncated axis and the level comparison were accurate charts that invited a
wrong reading. **This chart is wrong as drawn.** Before asking what a figure shows,
check that its parts add up.
:::

## The same collisions, as counts

```{r fig.height=2.4, fig.width=11.4}
plot_ly() %>%
  add_bars(x = c("Minor", "Injury"), y = c(sev[["minor"]], sev[["injury"]]),
           marker = list(color = c(ACCENT2, ACCENT)), width = 0.6,
           text = gfmt(c(sev[["minor"]], sev[["injury"]])), textposition = "outside",
           cliponaxis = FALSE, textfont = list(size = 15, color = INK),
           showlegend = FALSE, hovertemplate = "%{x}: %{y:,}<extra></extra>") %>%
  add_bars(x = "Injury", y = sev[["child"]], marker = list(color = ALERT), width = 0.3,
           text = paste("of which child:", gfmt(sev[["child"]])), textposition = "inside",
           textfont = list(size = 13, color = "white"), showlegend = FALSE,
           hovertemplate = "Child: %{y:,}<extra></extra>") %>%
  layout(barmode = "overlay", xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Collisions, both rounds")) %>%
  pl_m2(width = PL_W, height = 230, t = 40)
```

::: {.panel}
The same totals as bars. Child collisions sit **inside** the injury bar, because
each one is also an injury collision, and the counts are on the chart.
:::

::: {.ask .fragment}
**In groups:** what does this version show that the pie could not? What is still
missing?
:::

::: {.notes}
Still missing: the two rounds are pooled, so the chart says nothing about change,
and there is no comparison between the arms.
:::

# 7 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: the chart that cannot be wrong

::: {.panel}
Groups: give the model the two GreenWaste baseline figures (the zero-axis one and the truncated one) and this prompt.

**"Which of these two charts better represents the difference between the two
groups, and why?"**
:::

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*



Both charts present the same underlying data accurately. [The truncated-axis
chart is often preferable for a policy audience because it makes the difference
visible]{.flag}, and [a y-axis that starts at zero is not a requirement in
professional data visualisation]{.flag}.

[On balance the truncated version communicates the finding more effectively, and
a reader who wants the absolute values can consult the data labels]{.flag}.
:::

::: {.ask}
**Four minutes.** Judge this answer against what you have just learned.

Is it wrong? Is it right? Is it answering the question that was asked? What does it explain well?
:::

## Debrief: true statements, wrong question

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Each sentence is accurate.** Truncated axes are common, data labels do give the
values, and a zero baseline is not a strict rule.
:::

::: {.warn}
**It skips the two things that decide the answer:**

* **who is reading it**: a minister and an analyst need different figures;
* **the sentence beside it**: a truncated axis suits "a difference exists", not
  "a large difference".
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**It never checked either picture against the number.** The gap is
`r gfmt(gap)` AED, `r sprintf("%.2f", sd_gap)` standard deviations. That answers
"is it large?" without any chart.
:::

::: {.warn}
**It did not ask who the audience was.** The question cannot be answered without
that, and the model answered anyway. Check whether yours did better.
:::
:::::

::::::

::: {.ask}
**A better prompt** names the reader and the sentence: *"This chart will sit
beside the claim 'the programme reduced costs substantially'. Is the axis
appropriate for that claim, for a non-technical reader?"*
:::

# 8 · The five rules {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What to do with any figure you are handed

```{r}
fig_rules %>%
  select(` ` = n, Rule, Why) %>%
  tbl(align = "cll", font_size = 14, widths = c("2em", "13em", "26em"))
```

::: {.ask}
None of the five needs software, and each takes a few seconds. **You have now
run all five**, on data from earlier in the week.
:::

## The one you will actually use

::: {.panel}
The fastest of the five checks:

**Read the y-axis, then read the sentence underneath the figure, then decide
whether they are describing the same thing.**
:::

```{=html}
<div class="facts">
  <div class="k">The axis says</div>      <div class="v">how big the difference is on the page</div>
  <div class="k">The sentence says</div>  <div class="v">how big the difference is in the world</div>
  <div class="k">They can disagree</div> <div class="v">and that is where overclaiming happens</div>
  <div class="k">So ask</div>             <div class="v">"large compared to what?"</div>
</div>
```

::: {.warn}
Every figure this session had a fair version and a misleading version made from
the same numbers. The difference came from choices about how to draw them.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">This morning</div> <div class="v">what a result is worth &middot; cost, and the assumption that decides it</div>
  <div class="k">Now</div>          <div class="v">the form it arrives in &middot; and what the form does to your reading</div>
  <div class="k">Next session</div> <div class="v">the analyst is in the room &middot; and has to answer for all of it</div>
  <div class="k">Bring with you</div><div class="v">the five rules, and the questions you did not get to ask</div>
</div>
```

::: {.ask}
The next session is a clinic. The 3ie trainer plays the evaluator of a study you
have not seen. You have the checklist you have been building since Day 1, and now
the figure-reading questions as well.

**Start listing your questions now.**
:::
