---
title: "Spot the Problem: Naive Comparisons and Reading an RCT"
subtitle: "Module 2 · Day 1, Session 3: Two Comparisons That Both Look Convincing"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)   # lm_robust, clustered standard errors
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / offered
ACCENT2 <- "#A9561F"   # series 2: comparison / not offered
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

# ---- Chart sizing ----------------------------------------------------------
PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

tlev <- function(x) factor(x, levels = c("Before", "After"))
gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
# ONE place to set the join code. Everything below reads from here.
# This is the short JOIN code participants type at menti.com, NOT the long
# alphanumeric ID in the editor URL. Confirm it while the Menti is live.
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
MENTI_EMBED <- "https://www.mentimeter.com/app/presentation/al68xcnp6hm6zmc9k94rhbxwimzq6m29/embed"

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ---- The decision rule, fixed for the whole week ---------------------------
RULEBAR <- 1000    # must cut costs by at least this many AED

# ---- GreenWaste: the three comparisons -------------------------------------
gw <- read.csv("./evaluation_data_GreenWaste.csv")

# A. Before-and-after, among businesses that enrolled.
ba_before <- mean(gw$waste_management_costs[gw$enrolled == 1 & gw$round == 0])
ba_after  <- mean(gw$waste_management_costs[gw$enrolled == 1 & gw$round == 1])
ba_change <- ba_after - ba_before

# The drift among businesses that did NOT enrol, over the same two rounds. This
# is what "time was not standing still" looks like in this dataset.
comp_before <- mean(gw$waste_management_costs[gw$treatment_neighborhood == 1 &
                                              gw$eligible == 0 & gw$round == 0])
comp_after  <- mean(gw$waste_management_costs[gw$treatment_neighborhood == 1 &
                                              gw$eligible == 0 & gw$round == 1])

# B. With-and-without at follow-up: enrolled vs not enrolled, among eligible
# businesses in treated neighbourhoods.
gw1 <- gw %>% filter(round == 1, eligible == 1)
ww_prog <- mean(gw1$waste_management_costs[gw1$enrolled == 1])
ww_comp <- mean(gw1$waste_management_costs[gw1$enrolled == 0])
ww_gap  <- ww_prog - ww_comp

# C. The randomised estimate: eligible businesses, treated vs control
# neighbourhoods, at follow-up, errors clustered by neighbourhood.
rct <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                 data = gw1, clusters = neighborhood_identifier)
rct_est    <- coef(rct)[["treatment_neighborhood"]]
rct_ci     <- confint(rct)["treatment_neighborhood", ]
rct_p      <- rct$p.value[["treatment_neighborhood"]]
rct_control <- coef(rct)[["(Intercept)"]]
rct_treated <- rct_control + rct_est

# The decision question, done properly. The rule asks the programme to AVOID at
# least RULEBAR. It clears the bar only if even the interval's LEAST generous
# end reaches RULEBAR. Here it does not: the -935 end falls short, so the honest
# answer is that the estimate clears the bar and the interval spans it.
rct_avoided_lo <- abs(rct_ci[1])   # least the interval will allow
rct_avoided_hi <- abs(rct_ci[2])   # most the interval will allow
rct_clears     <- rct_avoided_lo >= RULEBAR

# ---- Balance table values --------------------------------------------------
# Built from the data so the slide cannot drift from the numbers.
bal_vars <- c("age_manager", "educ_manager", "staff_size",
              "foreign_owned", "recycling_center_distance")
bal_labs <- c(age_manager = "Manager age (years)",
              educ_manager = "Manager education (level)",
              staff_size = "Staff size",
              foreign_owned = "Foreign owned (share)",
              recycling_center_distance = "Distance to recycling centre (m)")
bal0 <- gw %>% filter(round == 0, eligible == 1)

balance_tbl <- map_dfr(bal_vars, function(v) {
  m <- lm_robust(bal0[[v]] ~ bal0$treatment_neighborhood)
  tibble(Characteristic = bal_labs[[v]],
         Treated   = mean(bal0[[v]][bal0$treatment_neighborhood == 1]),
         Control   = mean(bal0[[v]][bal0$treatment_neighborhood == 0]),
         Difference = coef(m)[2],
         p = tidy(m)$p.value[2])
})

# ---- The RCT result table --------------------------------------------------
rct_tbl <- rct %>% tidy() %>%
  mutate(` ` = c("Control group average", "Package vs. control"),
         CI = sprintf("[%s, %s]", gfmt(conf.low), gfmt(conf.high))) %>%
  select(` `, estimate, std.error, p.value, CI)

# ---- Police school-zone trial (the randomised Abu Dhabi case) --------------
pr <- read.csv("./evaluation_data_PoliceRCT.csv") %>%
  mutate(arm = factor(arm, levels = c("No package", "Package")))

pr0 <- pr %>% filter(round == 0)
pr1 <- pr %>% filter(round == 1)

pr_ctl_before <- mean(pr0$injury_collisions[pr0$arm == "No package"])
pr_ctl_after  <- mean(pr1$injury_collisions[pr1$arm == "No package"])
pr_trt_before <- mean(pr0$injury_collisions[pr0$arm == "Package"])
pr_trt_after  <- mean(pr1$injury_collisions[pr1$arm == "Package"])

pr_fit <- lm_robust(injury_collisions ~ arm, data = pr1, clusters = segment_id)
pr_est <- coef(pr_fit)[["armPackage"]]
pr_ci  <- confint(pr_fit)["armPackage", ]
pr_p   <- pr_fit$p.value[["armPackage"]]
pr_avoided_lo <- abs(pr_ci[1])
pr_avoided_hi <- abs(pr_ci[2])

# Police balance, as a check that the draw worked.
pr_bal <- map_dfr(c("lanes", "road_length_km", "prior_collisions_3yr"),
                  function(v) {
  a <- pr0[[v]][pr0$arm == "Package"]; b <- pr0[[v]][pr0$arm == "No package"]
  tibble(Characteristic = c(lanes = "Lanes",
                            road_length_km = "Length (km)",
                            prior_collisions_3yr = "Collisions, prior 3 yrs")[[v]],
         Package = mean(a), Control = mean(b), p = t.test(a, b)$p.value)
})
```

# 1 · Two comparisons, both convincing

## The GreenWaste programme

:::::: {.columns}

::::: {.column width="54%"}
The **GreenWaste Program** helped businesses cut their waste-management costs.
A pilot ran, data was collected before and after, and someone now has to decide
whether to take it national.

The bar it must clear, fixed for the whole week: **cut costs by at least
`r gfmt(RULEBAR)` AED**.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">GreenWaste, waste-cost reduction</div>
  <div class="k">Outcome</div>     <div class="v">Waste-management costs (AED)</div>
  <div class="k">The decision</div><div class="v">Scale up nationally, or not</div>
  <div class="k">The rule</div>    <div class="v">Cut at least 1,000 AED</div>
</div>
```
:::::

::::::

::: {.ask}
Session 1 taught you to ask **compared to what?** This session gives you two
answers to that question, and both look convincing.

You have to spot which is wrong, and why.
:::

## Comparison A: before and after

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.8, fig.width=5.6}
tibble(Time = tlev(c("Before", "After")),
       Cost = c(ba_before, ba_after)) %>%
  plot_ly(x = ~Time, y = ~Cost, type = "bar",
          marker = list(color = ACCENT, line = list(width = 0)),
          text = ~gfmt(Cost), textposition = "outside",
          textfont = list(size = 15, color = INK), showlegend = FALSE,
          hovertemplate = "%{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 1900))) %>%
  pl_m2(width = PL_WC, height = 275, title = "Businesses that took part")
```
:::::

::::: {.column width="50%"}
::: {.panel}
Businesses that took part paid **`r gfmt(ba_before)` AED** before and
**`r gfmt(ba_after)` AED** after.
:::

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, before to after</div>
</div>', gfmt(abs(ba_change))))
```

::: {.warn}
Read plainly: the programme cut costs by `r gfmt(abs(ba_change))` AED. That is
clear of the `r gfmt(RULEBAR)` bar. **Hold that thought.**
:::
:::::

::::::

## Comparison B: with and without

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.8, fig.width=5.6}
tibble(Group = factor(c("Took part", "Did not"),
                      levels = c("Took part", "Did not")),
       Cost  = c(ww_prog, ww_comp)) %>%
  plot_ly(x = ~Group, y = ~Cost, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~gfmt(Cost), textposition = "outside",
          textfont = list(size = 15, color = INK), showlegend = FALSE,
          hovertemplate = "%{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2300))) %>%
  pl_m2(width = PL_WC, height = 275,
        title = "At follow-up, same neighbourhoods")
```
:::::

::::: {.column width="50%"}
::: {.panel}
At the end, businesses that took part paid **`r gfmt(ww_prog)` AED**; those
that did not paid **`r gfmt(ww_comp)` AED**.
:::

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, the gap between them</div>
</div>', gfmt(abs(ww_gap))))
```

::: {.warn}
The **same programme**, a bigger number. Two honest-looking readings:
**-`r gfmt(abs(ba_change))`** and **-`r gfmt(abs(ww_gap))`**. They cannot both
be right.
:::
:::::

::::::

## `[MENTI PLACEHOLDER]` Which would you trust?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
Both comparisons say GreenWaste worked.

**Which would you trust more, and why?**

Before-and-after &nbsp;·&nbsp; With-and-without &nbsp;·&nbsp; Neither on its own
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right. Then give a reason.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
A split room is the ideal starting point. Take two or three "why" answers out
loud, and **do not correct them yet**.
:::

# 2 · The hidden problem in each

## Before-and-after compares to the past

:::::: {.columns}

::::: {.column width="52%"}
::: {.panel}
A before-and-after number compares the programme to **what came before**, not to
what would have happened anyway. Anything else that changed over the same period
is baked into it.
:::

::: {.warn}
Over these same two rounds, comparable businesses that did **not** take part
moved from `r gfmt(comp_before)` to `r gfmt(comp_after)` AED on their own.

Time was not standing still. So `r gfmt(abs(ba_change))` is not what the
programme alone did.
:::
:::::

::::: {.column width="48%"}
```{r fig.height=3, fig.width=5.6}
tibble(Time  = tlev(rep(c("Before", "After"), 2)),
       Group = rep(c("Took part", "Did not take part"), each = 2),
       Cost  = c(ba_before, ba_after, comp_before, comp_after)) %>%
  plot_ly(x = ~Time, y = ~Cost, color = ~Group,
          type = "scatter", mode = "lines+markers",
          colors = c("Took part" = ACCENT, "Did not take part" = ACCENT2),
          line = list(width = 3), marker = list(size = 11),
          hovertemplate = "%{fullData.name}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2700))) %>%
  pl_m2(width = PL_WC, height = 300)
```
:::::

::::::

::: {.ask}
Which way does this bias the `r gfmt(abs(ba_change))` figure: too big, or too
small? Note which way the line for businesses that did not take part went.
:::

## With-and-without compares different groups

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
A with-and-without number compares people who took part to people who did not.
But those two groups were **already different before the programme started**.

The businesses that did not take part were not eligible. They were larger, less
efficient, and more costly to run from the outset.
:::

::: {.warn}
So the `r gfmt(abs(ww_gap))` gap mixes two things: any real effect of the
programme, **and** the head start the eligible businesses already had.

It is not measuring the programme. It is measuring a pre-existing difference.
:::
:::::

::::: {.column width="50%"}
```{r fig.height=3, fig.width=5.6}
tibble(Group = factor(c("Control\nneighbourhoods", "Treated\nneighbourhoods",
                        "Took part", "Did not take part"),
                      levels = c("Control\nneighbourhoods",
                                 "Treated\nneighbourhoods",
                                 "Took part", "Did not take part")),
       Cost  = c(mean(gw$waste_management_costs[gw$treatment_neighborhood == 0 &
                                                 gw$round == 1]),
                 mean(gw$waste_management_costs[gw$treatment_neighborhood == 1 &
                                                 gw$round == 1]),
                 ww_prog, ww_comp),
       Kind  = c("Randomised", "Randomised", "Self-selected", "Self-selected")) %>%
  plot_ly(x = ~Group, y = ~Cost, color = ~Kind, type = "bar",
          colors = c("Randomised" = ACCENT, "Self-selected" = ACCENT2),
          text = ~gfmt(Cost), textposition = "outside",
          textfont = list(size = 12, color = INK),
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Waste costs at follow-up (AED)", range = c(0, 2600))) %>%
  pl_m2(width = PL_WC, height = 300)
```
:::::

::::::

## Two biases, opposite directions

```{=html}
<div class="facts">
  <div class="k">Before-and-after</div>
  <div class="v">Folds in every other change over the same period &middot; here it UNDERSTATES the fall</div>
  <div class="k">With-and-without</div>
  <div class="v">Compares groups that already differed &middot; here it OVERSTATES it</div>
  <div class="k">So the truth</div>
  <div class="v">Sits between the two, and neither comparison can find it</div>
</div>
```

::: {.ask}
The two numbers **bracket** something. `r gfmt(abs(ba_change))` is too small,
`r gfmt(abs(ww_gap))` is too big, and the honest answer is in between.

What would it take to land in the middle, deliberately?
:::

::: {.warn}
Both problems share one root: **the group that took part and the group that did
not were not comparable.** Section 3 is the design that fixes that.
:::

# Break

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the one design that removes the difference
between the groups, instead of trying to adjust for it afterwards.
:::

# 3 · How randomisation fixes it

## Random draw

::: {.panel}
Decide who is offered the programme by **random draw**, not by who is eligible,
who is keen, or who is easy to reach.

That single change is the whole idea.
:::

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.1em;">Same on average,<br>even on what we cannot measure</div>
</div>')
```

::: {.panel}
Random assignment balances the things we **can** see: size, age, location,
previous costs.

It also balances the things we **cannot** see and did not think to collect:
management quality, how motivated the staff are, how much luck the business has
had. On average. Which means they stop being an explanation.
:::

## Why it works: two people versus two crowds

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Pick **two people** at random from this room.

They could differ in almost every way. Neither is a fair stand-in for the other.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
Pick **two large groups** at random.

On average they look alike: same mix of ages, sizes, experience, and the same
mix of the things nobody measured.
:::
:::::

::::::

::: {.ask}
So when two large randomly-formed groups differ on the outcome, the difference
has nowhere else to come from. Chance is small, and everything else balances.
:::

::: {.panel .fragment}
**The caveat to hold onto.** This works *because* the draw was random. It says
nothing about whether people took up the programme, and nothing about whether
the effect would repeat somewhere else.

It answers one question cleanly: what did this programme do, here?
:::

## `[MENTI PLACEHOLDER]` True or false?

::: {.panel}
**1.** Randomisation makes the groups similar only on things we can measure.

**2.** With enough units, random assignment balances unmeasured things too.

**3.** A randomised comparison still needs the groups to have started out alike.
:::

::: {.ask}
Live quiz, revealed one at a time. Commit before I show each answer.
:::

::: {.panel .fragment}
**1. False.** Measured *and* unmeasured, on average. This is the one most people
get wrong.

**2. True.** That is what "on average" is doing.

**3. It creates that.** It does not need it to be true beforehand, which is the
entire point.
:::

# 4 · Reading the randomised result

## First check: did the random draw work?

:::::: {.columns}

::::: {.column width="56%"}
```{r}
balance_tbl %>%
  mutate(across(c(Treated, Control), ~ round(.x, 2)),
         Difference = round(Difference, 2), p = round(p, 2)) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("Characteristic", "Treated", "Control",
                      "Difference", "p-value")) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="44%"}
::: {.panel}
A **balance table** compares the two groups on their starting characteristics,
measured *before* anything happened.

Read it for two things: are the differences **small**, and are the p-values
**not tiny**?
:::

::: {.warn}
A tiny p-value here would mean the draw failed: the groups differed at the
start, and the comparison is compromised before it begins.
:::
:::::

::::::

## The estimate

:::::: {.columns}

::::: {.column width="52%"}
```{r}
rct_tbl %>%
  mutate(across(c(estimate, std.error), ~ round(.x, 0)),
         p.value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coefficient", "Std. error", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 17) %>%
  row_spec(2, bold = TRUE, background = "#E6F1F5")
```
:::::

::::: {.column width="48%"}
::: {.panel}
**Row 2 is the impact.** The difference in costs at follow-up between the two
randomly-formed groups.

`r gfmt(rct_est)` AED, interval
[`r gfmt(rct_ci[1])`, `r gfmt(rct_ci[2])`].
:::

::: {.warn}
**Not** the control group average. That is where the comparison group happened
to sit. It is not an effect, and it is not attributable to anything.
:::
:::::

::::::

## Which number is the impact?

```{=html}
<div class="facts">
  <div class="k">The impact</div>
  <div class="v">the Package vs. control row: -1,014 AED &middot; the difference the programme made</div>
  <div class="k">Not the impact</div>
  <div class="v">the control average, 1,798 AED &middot; where the comparison group sat, not an effect</div>
  <div class="k">Not sufficient either</div>
  <div class="v">the p-value &middot; it says the effect is unlikely to be zero, not that it is big enough</div>
</div>
```

::: {.ask}
Three ways to misread this one table, and only the first row is the answer.

Session 1 met the third one. **Statistically significant does not mean big
enough.** Here it is again, on a clean randomised result.
:::

## Does it clear the rule?

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.5, fig.width=6}
tibble(
  what = factor(c("The estimate", "95% interval", "Decision rule"),
                levels = c("Decision rule", "95% interval", "The estimate")),
  lo = c(abs(rct_est), rct_avoided_lo, RULEBAR),
  hi = c(abs(rct_est), rct_avoided_hi, RULEBAR),
  col = c(ACCENT, ACCENT, ALERT)
) %>%
  plot_ly() %>%
  add_segments(x = ~lo, xend = ~hi, y = ~what, yend = ~what,
               line = list(color = ~col, width = 5), showlegend = FALSE,
               hovertemplate = "avoids %{x:,.0f} AED<extra></extra>") %>%
  add_markers(x = ~lo, y = ~what, marker = list(color = ~col, size = 11),
              showlegend = FALSE,
              hovertemplate = "avoids %{x:,.0f}<extra></extra>") %>%
  layout(xaxis = ax("AED avoided per business", range = c(0, 1350)),
         yaxis = ax("", zeroline = FALSE)) %>%
  pl_m2(width = PL_WC, height = 250)
```
:::::

::::: {.column width="48%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, the randomised estimate</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
</div>', gfmt(abs(rct_est)), gfmt(RULEBAR), gfmt(rct_avoided_lo)))
```

::: {.warn}
The estimate clears the bar. The interval **spans** it: at its least generous,
the programme avoids `r gfmt(rct_avoided_lo)` AED, short of
`r gfmt(RULEBAR)`.
:::
:::::

::::::

::: {.ask .fragment}
Session 1's Word 5: **the interval decides, not the point estimate.**

Compare the two ends. Would you sign a national roll-out on this? Commit, then
we look at the same question on a real Abu Dhabi programme.
:::

## The picture

```{r fig.height=3, fig.width=11.5}
tibble(Group = factor(c("Control", "Package"), levels = c("Control", "Package")),
       Cost  = c(rct_control, rct_treated)) %>%
  plot_ly(x = ~Group, y = ~Cost, type = "scatter", mode = "markers",
          marker = list(size = 19, color = c(ACCENT2, ACCENT)),
          text = ~gfmt(Cost), textposition = "top center",
          textfont = list(size = 14, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y:,.0f} AED<extra></extra>") %>%
  add_segments(x = 2.4, xend = 2.4, y = rct_control, yend = rct_treated,
               line = list(color = ALERT, width = 4), hoverinfo = "skip") %>%
  add_annotations(x = 2.45, y = mean(c(rct_control, rct_treated)),
                  text = sprintf("Effect<br>%s AED", gfmt(abs(rct_est))),
                  showarrow = FALSE, xanchor = "left",
                  font = list(size = 13, color = ALERT)) %>%
  layout(xaxis = ax("", zeroline = FALSE, tickvals = c(1, 2),
                    ticktext = c("Control", "Package"), range = c(0.5, 3.2)),
         yaxis = ax("Waste costs at follow-up (AED)", range = c(0, 2200))) %>%
  pl_m2(height = 290)
```

::: {.panel}
Same programme as Section 1, but now the comparison is fair. The gap between two
groups formed by a draw, and nothing else, is the effect.
:::

# 5 · The real case: Abu Dhabi Police

## The school-zone package, decided by lottery

:::::: {.columns}

::::: {.column width="54%"}
Abu Dhabi Police had a road-safety package for school zones: raised crossings,
lower limits, crossing patrols. Four hundred zones needed it and only half could
be done in the first year.

Rather than pick the worst, the force **drew lots**: 200 zones chosen at random
for the first year, 200 held back.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>  <div class="v">School-zone safety package</div>
  <div class="k">Allocation</div> <div class="v">Random draw, 200 vs 200</div>
  <div class="k">Outcome</div>    <div class="v">Injury collisions</div>
  <div class="k">The rule</div>   <div class="v">Avoid 2.0 per zone per round</div>
</div>
```
:::::

::::::

::: {.warn}
*Illustrative case. The findings shown are synthetic, generated for this
training, not the evaluation's actual results.*
:::

::: {.ask}
Before the numbers: **why is this a stronger design than GreenWaste's?**
:::

## Did the draw work? Police zones

:::::: {.columns}

::::: {.column width="54%"}
```{r}
pr_bal %>%
  mutate(across(c(Package, Control), ~ round(.x, 2)), p = round(p, 2)) %>%
  kable(align = "lrrr", digits = 2,
        col.names = c("Characteristic", "Package", "No package", "p-value")) %>%
  kable_styling(full_width = FALSE, font_size = 16)
```
:::::

::::: {.column width="46%"}
::: {.panel}
Same check as before, on the real case.

Differences are small, p-values are not tiny. The draw did its job: whatever you
can see here, and whatever you cannot, the two sets of zones are alike on
average.
:::

::: {.warn}
This is the difference from Comparison B. There, the two groups **chose**
themselves. Here, a lottery placed them.
:::
:::::

::::::

## The randomised result

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Injury collisions avoided per zone per round</div>
</div>
<div class="stat-row">
  <div><span class="n">%.1f</span><span class="k">No package, before</span></div>
  <div><span class="n">%.1f</span><span class="k">No package, after</span></div>
  <div><span class="n">%.1f</span><span class="k">Package, after</span></div>
</div>', pr_est, pr_ctl_before, pr_ctl_after, pr_trt_after))
```

::: {.panel}
Both arms fell: the national trend reached the school zones too. The package
zones fell further.

The difference is `r sprintf("%.2f", abs(pr_est))` per zone per round, interval
[`r sprintf("%.2f", abs(pr_ci[1]))`, `r sprintf("%.2f", abs(pr_ci[2]))`].
:::

## Does it clear the rule?

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=2.5, fig.width=5.6}
tibble(
  what = factor(c("The estimate", "95% interval", "Decision rule"),
                levels = c("Decision rule", "95% interval", "The estimate")),
  lo = c(abs(pr_est), pr_avoided_lo, 2.0),
  hi = c(abs(pr_est), pr_avoided_hi, 2.0),
  col = c(ACCENT, ACCENT, ALERT)
) %>%
  plot_ly() %>%
  add_segments(x = ~lo, xend = ~hi, y = ~what, yend = ~what,
               line = list(color = ~col, width = 5), showlegend = FALSE,
               hovertemplate = "avoids %{x:.2f}<extra></extra>") %>%
  add_markers(x = ~lo, y = ~what, marker = list(color = ~col, size = 11),
              showlegend = FALSE,
              hovertemplate = "avoids %{x:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Collisions avoided per zone per round", range = c(0, 4.2)),
         yaxis = ax("", zeroline = FALSE)) %>%
  pl_m2(width = PL_WC, height = 250)
```
:::::

::::: {.column width="50%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.2f</div>
  <div class="label">Collisions avoided, per zone per round</div>
</div>
<div class="stat-row">
  <div><span class="n">%.2f</span><span class="k">The rule</span></div>
  <div><span class="n">%.2f</span><span class="k">Least it could be</span></div>
</div>', abs(pr_est), 2.0, pr_avoided_lo))
```

::: {.warn}
Same shape as GreenWaste: the estimate clears the bar, the interval straddles
it. The least it could be is `r sprintf("%.2f", pr_avoided_lo)`, under the
`r sprintf("%.1f", 2.0)` required.
:::
:::::

::::::

# 6 · Put AI to work

## AI Snapshot: explaining the pitfall

::: {.panel}
Groups: paste this prompt.

**"Explain in plain language why a before-and-after comparison can be
misleading."**
:::

::: {.ai-output}
**AI response:**

A before-and-after comparison looks at the outcome before a programme and after
it, and treats the difference as the programme's effect. It can mislead because
[if the numbers changed, the programme is what changed them]{.flag}.

The fix is simple: [as long as you measure the same group before and after, a
before-and-after comparison gives you the true effect]{.flag}. To be safe, you
can also [run a statistical test, and if the change is significant then the
programme worked]{.flag}.

In this case costs fell, so [we can conclude the programme caused the full fall
of 665 AED]{.flag}.
:::

::: {.ask}
Mark anything that looks wrong. Three minutes, in your groups.
:::

## Debrief: four errors

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · It assumed away the whole problem.** "If the numbers changed, the
programme changed them" ignores everything else that moved over the same period.
:::

::: {.warn}
**2 · It called before-and-after a true-effect method.** It is exactly the
comparison we just spent twenty minutes showing is biased. There is no
comparison group in it at all.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · It confused significant with correct.** A significant before-and-after
change is still a biased number. Significance does not repair a design.
:::

::: {.warn}
**4 · It quoted the 665 as the causal effect.** That is the naive figure. The
randomised estimate is `r gfmt(abs(rct_est))` AED, and the model had no way to
know that from the question it was asked.
:::
:::::

::::::

::: {.panel}
**Nothing in the tone flagged the wrong parts.** No hedge, no hesitation. Error 2
is the dangerous one: the model recommended the exact design this session exists
to warn against, in fluent prose.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain why a before-and-after comparison can be misleading.

The model answers the question it was asked, in general terms, and recommends
its own fix without knowing your design.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> We have a programme with a randomised control group and costs measured before
> and after. Explain what a before-and-after comparison of our data would get
> wrong, and what the randomised comparison adds. **Ask me questions before you
> answer.**
:::
:::::

::::::

::: {.ask}
Run both. Which one asks about *your* data before asserting anything?

Two habits, every session: make the AI **ask you questions first**, and have a
**second chat** review the first one's answer.
:::

# 7 · Close

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Before-and-after, with-and-without, or a fair comparison?</div>
  <div class="k">2</div> <div class="v">If the groups were not randomised, how do you know they started out alike?</div>
  <div class="k">3</div> <div class="v">Does the effect, and its interval, clear our decision rule, not just zero?</div>
</div>
```

::: {.panel}
Question 2 separates a convincing number from a trustworthy one. Question 3
separates a real effect from a decision.
:::

## What today gives you

```{=html}
<div class="facts">
  <div class="k">Before-and-after</div> <div class="v">folds in everything else that changed &middot; can understate</div>
  <div class="k">With-and-without</div>  <div class="v">compares groups that already differed &middot; can overstate</div>
  <div class="k">Randomisation</div>     <div class="v">makes the groups alike on average, including on what nobody measured</div>
  <div class="k">Reading the result</div><div class="v">find the impact row, then judge it against the rule and its interval</div>
</div>
```

::: {.panel}
Next session we return to the GreenWaste data and ask what today's two
comparisons were missing: **was the difference between the groups stable
before the programme started?**
:::
