---
title: "A practice lottery: 20 manager-age slips"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    header-includes: |
      <style>
      body{font:17px/1.4 Arial,sans-serif;color:#111418;max-width:850px;margin:25px auto;padding:0 25px}h1{font-size:28px;color:#063360}.slips{display:grid;grid-template-columns:repeat(4,1fr);gap:0;margin:18px 0}.slip{border:1px dashed #7f8995;padding:13px 10px;text-align:center}.number{font-size:24px;font-weight:bold;color:#215a9e;margin:0}.age{font-size:23px;margin:5px 0}.id{font-size:12px;color:#545860;margin:0}.caption{font-size:14px;color:#545860}button{font:inherit;padding:8px;color:#063360;background:white;border:1px solid #7da1c4}@page{size:A4;margin:14mm}@media print{body{margin:0;padding:0;font-size:11pt;max-width:none}h1{font-size:19pt}.number{font-size:17pt}.age{font-size:16pt}.slip{padding:10px}.caption{font-size:9pt}button{display:none}}</style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('session3_review.R', encoding='UTF-8')
```

**20 businesses, 10 places.** Cut one set per group along the dashed lines. Draw ten slips for join; the other ten wait. Calculate each group's mean age. Replace all slips before drawing again.

<button onclick="window.print()">Print age slips</button>

```{r, results='asis'}
html_out(paste0('<div class="slips">',paste(vapply(1:20,function(i)paste0('<div class="slip"><p class="number">',i,'</p><p class="age">Age ',s3_ages[i],'</p><p class="id">',pilot$business[i],'</p></div>'),''),collapse=''),'</div>'))
```

This practice uses the first 20 actual pilot manager-age records. A draw compares **pre-programme ages**, not effects on costs. It is not the actual pilot assignment: that involved **400 businesses, 200 places**. Record both draws on worksheet page 3.

::: {.caption}
GreenWaste is a fictional training case. Source: corrected pilot data. Ages are in years.
:::
