---
title: "Was It Worth It? Reading a Cost-Benefit Result"
subtitle: "Module 2 · Day 3, Session 1: The Assumption That Decides the Answer"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

# ===========================================================================
# The cost-benefit model for GreenWaste
# ===========================================================================
# Everything here is shown to the room on a slide. The brief is explicit that
# the trainer displays the calculations rather than asking participants to
# produce them, so the arithmetic is visible and checkable rather than asserted.
#
# The two inputs that come from earlier in the week:
#   the annual saving per business, which is the DiD estimate from Day 2 S1
#   the number of businesses, which is the enrolled count from Day 2 S3
# The one input that is a pure programme fact:
#   the cost per business of delivering it
#
# The saving is 816 AED, which is the estimate the room computed by hand in the
# Day 2 S1 session and which missed that session's 1,000 AED rule. It is not the
# rule. Using the rule as though it were the finding would have inflated the
# ratio from 1.87 to 2.29 and invented a programme that performed better than
# the one that was measured.

COST_PER_BUSINESS <- 1800      # AED, one-off cost of putting a business through
BENEFIT_PER_YEAR  <- 816       # AED per year, the Day 2 S1 difference-in-differences
BENEFIT_LO        <- 752       # the interval end closest to zero: least saving
BENEFIT_HI        <- 881       # the other end: most saving
BENEFIT_LAG       <- 1         # no saving in year 1: rollout and adoption take time
ENROLLED          <- 2964      # businesses enrolled, from the matching session
RULEBAR           <- 1000      # the Day 1-2 decision rule, for the contrast below

DISCOUNT_BASE <- 0.05          # the rate the authority uses by default
HORIZON_BASE  <- 5             # years of benefit assumed by default

# Present value of 1 AED a year, from year (lag+1) to year (lag + years).
pv_annuity <- function(r, years, lag = BENEFIT_LAG) {
  if (r == 0) return(years)
  a <- function(n) (1 - (1 + r)^(-n)) / r
  a(lag + years) - a(lag)
}

# The benefit-cost ratio: present value of everything we get, over what we spend.
bcr <- function(r = DISCOUNT_BASE, years = HORIZON_BASE,
                cost = COST_PER_BUSINESS, benefit = BENEFIT_PER_YEAR) {
  benefit * pv_annuity(r, years) / cost
}

bcr_headline <- bcr()
pv_per_biz   <- BENEFIT_PER_YEAR * pv_annuity(DISCOUNT_BASE, HORIZON_BASE)

# Total programme figures, for the scale of the decision.
total_cost    <- COST_PER_BUSINESS * ENROLLED
total_benefit <- pv_per_biz * ENROLLED
total_net     <- total_benefit - total_cost

# ---- The two analyst choices, alone and together ---------------------------
# Choice 1: the discount rate. Choice 2: how long the benefit is assumed to last.
rate_grid  <- c(0.03, 0.05, 0.08, 0.12, 0.20)
years_grid <- c(2, 3, 5, 8, 10)

grid_tbl <- expand_grid(years = years_grid, r = rate_grid) %>%
  mutate(bcr = map2_dbl(r, years, ~ bcr(.x, .y)),
         cell = sprintf("%.2f", bcr),
         below = bcr < 1)

# A tidy long form for the heat table on the slide.
grid_wide <- grid_tbl %>%
  mutate(rate_lab = sprintf("%.0f%%", 100 * r),
         year_lab = sprintf("%d years", years)) %>%
  select(year_lab, rate_lab, cell) %>%
  pivot_wider(names_from = rate_lab, values_from = cell)

# Where the ratio crosses 1 for each horizon. This is the number that makes the
# session's point: the flip is not exotic, it happens inside normal practice.
flip_rate <- function(yrs) {
  # If it is already below 1 with no discounting at all, no rate rescues it and
  # the honest answer is that there is no flip point to report.
  if (bcr(0, yrs) < 1) return(NA_real_)
  r <- 0
  while (bcr(r, yrs) >= 1 && r < 3) r <- r + 0.001
  if (bcr(r, yrs) < 1) r * 100 else NA_real_
}
years_for_flip <- c(2, 3, 4, 5)
flip_tbl <- tibble(years = years_for_flip, flip = map_dbl(years_for_flip, flip_rate))

# The rate alone, over the base horizon. This shows the rate is NOT enough on its
# own across the range agencies actually use: 3% to 8% leaves the ratio between
# 2.02 and 1.68. That negative result is worth a slide, because a commissioner's
# instinct is to blame the discount rate.
rate_only <- tibble(
  r = c(0, 0.03, 0.05, 0.08, 0.10, 0.15, 0.20, 0.30, 0.40),
  bcr = map_dbl(c(0, 0.03, 0.05, 0.08, 0.10, 0.15, 0.20, 0.30, 0.40),
                ~ bcr(.x, HORIZON_BASE))
) %>% mutate(lab = sprintf("%.0f%%", 100 * r))

# ---- What the ratio would be if the benefit were smaller -------------------
# The Day 2 S1 interval ran from 962 to 1,038 AED. Carrying that forward shows
# the cost-benefit ratio inherits the uncertainty, rather than adding new.
benefit_range <- tibble(
  benefit = c(BENEFIT_LO, BENEFIT_PER_YEAR, BENEFIT_HI),
  bcr = map_dbl(c(BENEFIT_LO, BENEFIT_PER_YEAR, BENEFIT_HI),
                ~ bcr(DISCOUNT_BASE, HORIZON_BASE, benefit = .x))
)

# ---- The horizon at which the ratio is exactly 1 ---------------------------
# Solved rather than eyeballed, so the slide can state it precisely. This is the
# session's single most useful number.
breakeven_horizon <- function(r = DISCOUNT_BASE) {
  lo <- 1; hi <- 20
  for (i in 1:60) {
    mid <- (lo + hi) / 2
    if (bcr(r, mid) < 1) lo <- mid else hi <- mid
  }
  (lo + hi) / 2
}

# ---- Why a programme that missed the rule still clears a ratio of 1 --------
# The rule asks for 1,000 AED in a single period. The ratio compares a one-off
# cost against several periods of saving. Different questions, different answers.
ratio_if_rule_met <- bcr(benefit = RULEBAR)

# ---- The three-question framework, scored on this programme ---------------
framework <- tribble(
  ~Question, ~Asks,
  "Are the benefits plausible?",
  "The saving is the difference-in-differences estimate. It is measured, and it is local to the businesses compared.",
  "Are all the costs included?",
  "1,800 AED per business is the delivery cost. It does not include the authority's own set-up, or the businesses' time.",
  "What would flip the result?",
  "Assuming the saving lasts two years instead of five. That alone takes the ratio below break-even."
)
```

# 1 · The headline

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">three questions &middot; reading a number &middot; what a fair comparison is</div>
  <div class="k">Day 2</div>  <div class="v">four ways to estimate an effect: descriptives, DiD, RDD, matching</div>
  <div class="k">Today</div>  <div class="v">what the effect is <em>worth</em> &middot; cost, and the assumptions that decide it</div>
</div>
```

::: {.panel}
Everything so far has been about **how big the effect is**. Today is about
whether it is **worth the money** — and that is a different question with a
different set of judgement calls.
:::

::: {.ask}
A programme can have a real effect and still not be worth funding. You saw the
shape of that all week: real, and not enough.

Today the bar is a ratio instead of a threshold.
:::

## The decision, and the ratio

:::::: {.columns}

::::: {.column width="54%"}
GreenWaste cost the authority **`r gfmt(COST_PER_BUSINESS)` AED per business** to
deliver, and saved each participating business about **`r gfmt(BENEFIT_PER_YEAR)`
AED a year** — the estimate the room computed by hand on Day 2, and the one that
missed that session's 1,000 AED rule.

The standard way to compare those is a **benefit-cost ratio**: what you get,
divided by what you spend, with future money counted as worth less than money
today.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Cost</div>     <div class="v">1,800 AED per business, one-off</div>
  <div class="k">Benefit</div>  <div class="v">816 AED per business per year</div>
  <div class="k">Decision rule</div> <div class="v">A ratio above 1 means it pays for itself</div>
  <div class="k">Ratio found</div> <div class="v">`r sprintf("%.2f", bcr_headline)`</div>
</div>
```
:::::

::::::

## The headline number

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Benefit-cost ratio, GreenWaste</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">AED spent per business</span></div>
  <div><span class="n">%s</span><span class="k">AED returned per business</span></div>
  <div><span class="n">%s</span><span class="k">AED net, across the programme</span></div>
</div>', bcr_headline, gfmt(COST_PER_BUSINESS), gfmt(pv_per_biz), gfmt(total_net)))
```

::: {.panel}
**`r sprintf("%.2f", bcr_headline)` means: for every AED spent, about
`r sprintf("%.2f", bcr_headline)` AED comes back.** On that reading this is not
a borderline decision. It is clearly worth doing.

**And that is the puzzle.** This is the same programme that failed the Day 2
rule. It cut costs by `r gfmt(BENEFIT_PER_YEAR)` AED against a `r gfmt(RULEBAR)`
AED bar, and here it returns nearly twice what it cost. Both statements are
right.
:::

::: {.warn}
A ratio and a rule are not the same test. The rule asks what happens **in one
period**. The ratio asks what happens **over the whole life of the investment**.
A programme can be too small to scale on the first test and comfortably worth
funding on the second.

Keep that gap in view. It is the reason this session exists.
:::

## `[MENTI PLACEHOLDER]` Would you approve it?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
On a ratio of **`r sprintf("%.2f", bcr_headline)`**, what would you do?

Approve &nbsp;·&nbsp; Reject &nbsp;·&nbsp; I need to know something else first
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code on the
right.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
Commit before we go further. If most of the room picks "I need to know
something else first", ask two people **what** — write the answers up, and
compare them with the three questions at the end of the session.

If anyone notices that this is the same programme that missed the 1,000 AED
rule on Day 2, stop and let them make the point to the room. That observation is
the session's hinge.
:::

## What went into that number

This is the arithmetic, on the screen.

```{r}
tibble(
  Step = c("Cost per business",
           "Annual saving",
           "Years of saving assumed",
           "Discount rate assumed",
           "Value of the savings, today's money",
           "Benefit-cost ratio"),
  Working = c("one-off delivery cost",
              "the Day 2 difference-in-differences estimate",
              "5, beginning one year after rollout",
              "5% a year",
              sprintf("%s x %.3f = %s AED", gfmt(BENEFIT_PER_YEAR),
                      pv_annuity(DISCOUNT_BASE, HORIZON_BASE), gfmt(pv_per_biz)),
              sprintf("%s / %s", gfmt(pv_per_biz), gfmt(COST_PER_BUSINESS))),
  Value = c(sprintf("%s AED", gfmt(COST_PER_BUSINESS)),
            sprintf("%s AED", gfmt(BENEFIT_PER_YEAR)),
            "5 years", "5%",
            sprintf("%s AED", gfmt(pv_per_biz)),
            sprintf("%.2f", bcr_headline))
) %>%
  kable(align = "llr", col.names = c("Step", "Working", "Value")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(3:4, bold = FALSE, background = "#F7F1E8") %>%
  row_spec(6, bold = TRUE, background = "#E6F1F5")
```

::: {.ask}
Two rows are shaded. They are not costs and they are not benefits. They are
**choices** — and they are the subject of the next twenty minutes.
:::

# 2 · The assumptions that decide it

## The saving does not arrive today

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">A saving next year is worth less than a saving today.<br>How much less is a choice.</div>
</div>')
```

::: {.panel}
Money you have now can be used, invested or saved. Money promised in three
years cannot. Discounting is the arithmetic of that difference, and the
**discount rate** is how fast you decide future money loses value.
:::

::: {.ask}
Nobody debates that future money is worth less. They debate **how much less** —
and that number goes straight into the ratio you are shown.
:::

## Choice 1: the discount rate, on its own

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.7, fig.width=5.6}
rate_only %>%
  plot_ly(x = ~lab, y = ~bcr, type = "bar",
          marker = list(color = ifelse(rate_only$bcr < 1, ALERT, ACCENT),
                        line = list(width = 0)),
          text = ~sprintf("%.2f", bcr), textposition = "outside",
          textfont = list(size = 12, color = INK), showlegend = FALSE,
          hovertemplate = "%{x} discount: ratio %{y:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Discount rate", zeroline = FALSE),
         yaxis = ax("Benefit-cost ratio", range = c(0, 3.3)),
         shapes = list(list(type = "line", x0 = -0.5, x1 = 8.5,
                            y0 = 1, y1 = 1, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 8.5, y = 1, yref = "y", xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "break-even", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_WC, height = 260)
```
:::::

::::: {.column width="48%"}
::: {.warn}
**The discount rate on its own does not flip this decision.**

Across the whole range an agency might actually use — 3% to 8% — the ratio moves
only from `r sprintf("%.2f", bcr(0.03, HORIZON_BASE))` to
`r sprintf("%.2f", bcr(0.08, HORIZON_BASE))`. You have to reach
**`r sprintf("%.0f%%", 100 * min(rate_only$r[rate_only$bcr < 1]))`** to push it
below 1, and by then the assumption is not defensible.
:::

::: {.panel}
So if someone tells you the ratio is fragile because of the discount rate, that
is only half true. It is the *other* choice that does the damage.
:::
:::::

::::::

## Choice 2: how long the saving lasts

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.7, fig.width=5.6}
tibble(years = years_grid, bcr = map_dbl(years_grid, ~ bcr(DISCOUNT_BASE, .x))) %>%
  plot_ly(x = ~factor(years), y = ~bcr, type = "bar",
          marker = list(color = ifelse(map_dbl(years_grid, ~ bcr(DISCOUNT_BASE, .x)) < 1,
                                       ALERT, ACCENT), line = list(width = 0)),
          text = ~sprintf("%.2f", bcr), textposition = "outside",
          textfont = list(size = 13, color = INK), showlegend = FALSE,
          hovertemplate = "%{x} years: ratio %{y:.2f}<extra></extra>") %>%
  layout(xaxis = ax("Years of saving assumed (at a 5% discount rate)", zeroline = FALSE),
         yaxis = ax("Benefit-cost ratio", range = c(0, 5)),
         shapes = list(list(type = "line", x0 = -0.5, x1 = 4.5,
                            y0 = 1, y1 = 1, line = list(color = ALERT, width = 3))),
         annotations = list(list(x = 4.5, y = 1, yref = "y", xanchor = "right",
                                 yanchor = "bottom", showarrow = FALSE,
                                 text = "break-even", font = list(size = 12, color = ALERT)))) %>%
  pl_m2(width = PL_WC, height = 260)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The Day 2 sessions measured the saving over **one period**. Nothing measured how
long it lasts.

If it lasts **two years**, the ratio is
`r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))` — **below break-even**.
:::

::: {.warn}
**This is the one that decides it.**

Below about **`r sprintf("%.1f", breakeven_horizon())` years** the programme does
not pay for itself. At two years the ratio is
`r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))`; at three it is
`r sprintf("%.2f", bcr(DISCOUNT_BASE, 3))`.

Nothing in this evaluation measured how long the saving lasts.
:::
:::::

::::::

## The horizon decides it, not the rate

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">The rate, 3% to 8%</div>      <div class="v">moves the ratio from 2.02 to 1.68 &middot; never close to 1</div>
  <div class="k">The horizon, 2 to 5 years</div> <div class="v">moves it from 0.80 to 1.87 &middot; crosses 1</div>
  <div class="k">Break-even horizon</div>      <div class="v">2.5 years at a 5% rate &middot; 2.2 years with no discounting at all</div>
  <div class="k">So the question is</div>    <div class="v">how long does a saving of this kind actually last?</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.ask}
The break-even horizon barely moves when you change the discount rate. At 5% it
is 2.5 years; at 3% it is 2.5 years as well. Even with no discounting at all the
programme needs 2.2 years of saving.

**That is what a dominant assumption looks like.** One choice can be argued about
indefinitely and hardly matters. The other is barely discussed and decides the
answer.
:::
:::::

::::::

## Both choices together

```{r}
grid_wide %>%
  kable(align = "lrrrrr", col.names = c("Years of saving", "3%", "5%", "8%", "12%", "20%")) %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  row_spec(1, bold = TRUE, background = "#F7E9E6") %>%
  row_spec(3, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = "Shaded: the two rows the session turns on. Every cell uses the same cost and the same measured annual saving; only the two analyst choices differ. Values below 1.00 mean the programme does not pay for itself.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.ask}
**Red row:** two years of saving. **Blue row:** five years, the base case.

Read the two rows against each other. The recommendation reverses, and the
evidence never changed. Only the horizon assumption did.
:::

## Where the ratio crosses 1

```{r}
flip_tbl %>%
  mutate(`Discount rate at which it flips` =
           ifelse(is.na(flip), "already below 1", sprintf("%.1f%%", flip))) %>%
  select(`Years of saving` = years, `Discount rate at which it flips`) %>%
  kable(align = "rr") %>%
  kable_styling(full_width = FALSE, font_size = 16) %>%
  footnote(general = "Below this rate the ratio exceeds 1; above it, the programme does not pay for itself. Read the top rows: at a two-year horizon there is no discount rate that rescues it.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.warn}
Read the top two rows. **If you assume the saving lasts only two years, no
discount rate at all makes this worth doing.** The programme is already below
break-even before the discount rate is applied.

At three years, a rate of
`r sprintf("%.1f%%", flip_tbl$flip[flip_tbl$years == 3])` flips it, which is at
the edge of normal practice. At the five-year base case you would need
`r sprintf("%.1f%%", flip_tbl$flip[flip_tbl$years == 5])`, which nobody uses.

**Read down the column, not across the row.** The rate only matters once you
have already assumed the saving persists for years.
:::

::: {.ask}
So the honest statement is not "the ratio is `r sprintf("%.2f", bcr_headline)`".
It is:

**"The ratio is `r sprintf("%.2f", bcr_headline)` under assumptions the evidence
does not determine, and plausible alternatives put it anywhere between
`r sprintf("%.2f", min(grid_tbl$bcr))` and
`r sprintf("%.2f", max(grid_tbl$bcr))`."**
:::

## What the ratio is actually telling you

```{r fig.height=3.1, fig.width=11.5}
grid_tbl %>%
  mutate(r_lab = sprintf("%.0f%%", 100 * r),
         y_lab = sprintf("%d years", years),
         r_lab = factor(r_lab, levels = sprintf("%.0f%%", 100 * rate_grid)),
         y_lab = factor(y_lab, levels = sprintf("%d years", rev(years_grid)))) %>%
  plot_ly(x = ~r_lab, y = ~y_lab, z = ~bcr, type = "heatmap",
          colorscale = list(c(0, "#8E2413"), c(0.19, "#C98F72"),
                            c(0.21, "#CFE4E9"), c(1, "#0E6E80")),
          zmin = 0.55, zmax = 3.8,
          text = ~sprintf("%.2f", bcr), texttemplate = "%{text}",
          textfont = list(size = 15, color = INK),
          hovertemplate = "%{y}, discount %{x}: ratio %{z:.2f}<extra></extra>",
          colorbar = list(title = "ratio", thickness = 14, len = 0.8)) %>%
  layout(xaxis = ax("Discount rate", zeroline = FALSE),
         yaxis = ax("Years of saving assumed", zeroline = FALSE)) %>%
  pl_m2(height = 300)
```

::: {.ask}
Every cell is the same programme, the same cost, and the same measured saving.

**Only two judgement calls differ.** The whole grid is one decision, drawn nine
ways. Find the cells below 1.0, then find the row that is below 1 everywhere.
:::

# 3 · The three questions

## How to judge any cost-benefit claim

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Are the benefits plausible?</div>
  <div class="k">2</div> <div class="v">Are all the costs included?</div>
  <div class="k">3</div> <div class="v">What would flip the result?</div>
</div>
```

::: {.panel}
Three questions, and none of them needs a spreadsheet. They are the same shape
as the questions every other session has ended on: ask what the number is
compared to, ask what is missing, ask what it would take to change the answer.
:::

::: {.ask}
Question 3 is the one that separates a reader from a recipient. It is also the
one an evaluator is least often asked.
:::

## Question 1: are the benefits plausible?

:::::: {.columns}

::::: {.column width="52%"}
```{r}
benefit_range %>%
  mutate(`Annual saving assumed` = sprintf("%s AED", gfmt(benefit)),
         `Benefit-cost ratio` = sprintf("%.2f", bcr)) %>%
  select(-benefit, -bcr) %>%
  kable(align = "rr") %>%
  kable_styling(full_width = FALSE, font_size = 16)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The saving here is **not** assumed. It is the difference-in-differences estimate
from Day 2, and the three rows carry its confidence interval forward.

The ratio moves from `r sprintf("%.2f", min(benefit_range$bcr))` to
`r sprintf("%.2f", max(benefit_range$bcr))` across the interval.
:::

::: {.warn}
**Look at how small that is.** The measured uncertainty in the saving barely
moves the ratio, because the interval is tight: the Day 2 estimate was
`r gfmt(BENEFIT_PER_YEAR)` with bounds of `r gfmt(BENEFIT_LO)` and
`r gfmt(BENEFIT_HI)`.

The horizon, which moves the ratio from `r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))`
to `r sprintf("%.2f", bcr_headline)`, has no interval at all, because nobody
measured it.
:::
:::::

::::::

## The measured part and the assumed part

```{r results='asis'}
cat(sprintf('
<div class="facts">
  <div class="k">Measured</div>  <div class="v">the annual saving &middot; %s AED &middot; interval %s to %s &middot; moves the ratio by %.2f</div>
  <div class="k">Measured</div>  <div class="v">the cost &middot; %s AED &middot; a programme fact</div>
  <div class="k">Assumed</div>   <div class="v">how many years the saving lasts &middot; moves the ratio by %.2f</div>
  <div class="k">Assumed</div>   <div class="v">the discount rate &middot; at 3%% to 8%% it moves the ratio by only %.2f</div>
</div>',
  gfmt(BENEFIT_PER_YEAR), gfmt(BENEFIT_LO), gfmt(BENEFIT_HI),
  max(benefit_range$bcr) - min(benefit_range$bcr),
  gfmt(COST_PER_BUSINESS),
  bcr_headline - bcr(DISCOUNT_BASE, 2),
  abs(bcr(0.03, HORIZON_BASE) - bcr(0.08, HORIZON_BASE))))
```

::: {.ask}
Two lines are measured and two are assumed. **The assumed lines move the answer
more than the measured ones do.**

That is the whole session in one table. A confident-looking ratio can rest almost
entirely on choices nobody has evidence for, and the numbers that were carefully
estimated are not the ones deciding it.
:::

## Question 2: are all the costs included?

```{=html}
<div class="facts">
  <div class="k">Included</div> <div class="v">1,800 AED per business to deliver the programme</div>
  <div class="k">Not included</div> <div class="v">the authority's own set-up and administration</div>
  <div class="k">Not included</div> <div class="v">the time the businesses spent on it</div>
  <div class="k">Not included</div> <div class="v">anything the programme displaced</div>
</div>
```

::: {.ask}
Every cost left out makes the ratio **look better**. None of them appears in the
table you were shown.

Ask an evaluator what is in the cost line. A ratio built on a partial cost is not
a conservative estimate of anything.
:::

## Question 3: what would flip it?

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
You already know the answer for this programme, and it is not the discount rate.

**Assume the saving lasts two years instead of five, and the ratio falls from
`r sprintf("%.2f", bcr_headline)` to `r sprintf("%.2f", bcr(DISCOUNT_BASE, 2))`.**
:::

::: {.ask}
So the question to ask is not "what discount rate did you use?"

It is: **"How long do you assume the benefit lasts, and what evidence do you
have for that?"**
:::
:::::

::::: {.column width="46%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value" style="font-size:1.9em;">%.2f &nbsp;&rarr;&nbsp; %.2f</div>
  <div class="label">The same evidence, two horizons</div>
</div>
<div class="stat-row">
  <div><span class="n">5 yrs</span><span class="k">Assumption A</span></div>
  <div><span class="n">2 yrs</span><span class="k">Assumption B</span></div>
</div>', bcr_headline, bcr(DISCOUNT_BASE, 2)))
```
:::::

::::::

# 4 · Put AI to work

## AI Snapshot: find the scenario that breaks it

::: {.panel}
Groups: paste the cost-benefit table and this prompt.

**"Generate one plausible scenario in which this programme's benefit-cost ratio
falls below 1. Then tell me whether that scenario is realistic."**
:::

::: {.ai-output}
**AI response:**

The ratio falls below 1 if the discount rate is raised. [At a discount rate of
around 40%, the ratio drops below break-even]{.flag}, which would mean the
programme does not pay for itself and should not be funded.

[Alternatively, if the programme's benefits are delayed by a few years, the
ratio also falls below 1]{.flag}. Both scenarios are [realistic and commonly
observed in infrastructure appraisal]{.flag}.
:::

::: {.ask}
Check each claim against the table. Which is right, which is wrong, and which is
true but useless? Four minutes.
:::

## Debrief: right arithmetic, right lever, no numbers

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The arithmetic is correct.** A 40% discount rate does push this ratio below 1.
The model checks out.
:::

::: {.warn}
**1 · Wrong lever, and it took the easy one.** The rate does flip the ratio, but
only at
`r sprintf("%.0f%%", 100 * min(rate_only$r[rate_only$bcr < 1]))`, which no
agency uses. It answered *name a scenario* rather than *name the one that
matters*. **Arithmetically true and practically irrelevant.**
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**2 · It pointed at the right lever and then did not use it.** "Benefits delayed
by a few years" is the horizon, which is the assumption that actually decides
this programme. But the model never says how many years, never gives a ratio,
and never mentions that the saving is measured for one period only.

**Right area, no number.** A direction without a magnitude cannot be acted on.
:::

::: {.warn}
**3 · "Commonly observed in infrastructure appraisal."** Confident, unsourced,
and about a different kind of programme. It reads as authority it does not have.
:::
:::::

::::::

::: {.panel}
**This is a new failure shape, and a harder one to catch than the errors you
found on Days 1 and 2.** The model invented no number and misread no table. It
reached for the lever that is easiest to compute, and when it stumbled onto the
right one it stopped at a direction and never produced a figure.

**The lesson: ask for the assumption that matters, and ask for the number.**
A model will find the scenario you asked for, and it may even name the true
cause, and it can still leave you with nothing you can act on.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Generate one plausible scenario in which this benefit-cost ratio falls below 1.

The word "plausible" does the work, and it is not defined. The model reaches for
whatever is easiest to compute.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Here is a cost-benefit model. The cost and annual saving are measured; the
> discount rate and the number of benefit years are assumptions. Tell me which
> **single assumption** has the largest effect on the ratio, quantify it, and
> say whether that assumption is one a government agency would normally make.
> **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both. The second asks the model to **rank** the assumptions before choosing
one, which is exactly the step the first was allowed to skip.
:::

# 5 · The Abu Dhabi case

## `[PLACEHOLDER]` The cost-benefit case

::: {.warn}
**Awaiting the real findings.** The Abu Dhabi cost-benefit case for the outline
has not been shared with this workspace yet, so nothing here is presented as its
result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>   <div class="v">[PLACEHOLDER]</div>
  <div class="k">Ratio claimed</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Cost per unit</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Horizon assumed</div><div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed: the ratio, what went into it, and the two assumptions
that decide it. When the findings arrive they drop into these fields without
redesigning the session.
:::

## Applying the three questions

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Are the benefits plausible? Measured, modelled, or assumed?</div>
  <div class="k">2</div> <div class="v">Are all the costs included? What was left out, and does it flatter the ratio?</div>
  <div class="k">3</div> <div class="v">What would flip it? Which assumption, and by how much?</div>
</div>
```

::: {.ask}
Same three questions we just ran on GreenWaste, in the same order. The programme
changes; the interrogation does not.

**In groups:** run them on whatever the real case turns out to be, and present
your verdict in two minutes.
:::

# 6 · Close

## Two tests, one programme

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Day 2 rule</div>      <div class="v">cut costs by 1,000 AED in one period</div>
  <div class="k">What was found</div>  <div class="v">816 AED &middot; real, significant, and short of the bar</div>
  <div class="k">Verdict</div>         <div class="v">do not scale it on that evidence</div>
</div>
```
:::::

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Today's ratio</div>   <div class="v">1.87 AED returned per AED spent</div>
  <div class="k">What it assumes</div> <div class="v">the saving persists for five years</div>
  <div class="k">Verdict</div>         <div class="v">clearly worth funding</div>
</div>
```
:::::

::::::

::: {.ask}
These are the same programme, the same estimate and the same cost.

**One test says no and the other says yes.** That is not a contradiction. The
rule asks whether the effect is big enough to be worth scaling; the ratio asks
whether the investment pays for itself over its life. An evaluator who reports
only one of them has answered half the question the commissioner asked.
:::

::: {.warn}
Notice which test is doing the work. The rule is measured. The ratio is mostly
assumed.
:::

## What today gives you

```{=html}
<div class="facts">
  <div class="k">A ratio is a model</div>  <div class="v">not a measurement &middot; it contains choices as well as evidence</div>
  <div class="k">The rate</div>            <div class="v">3% to 8% barely moves it &middot; do not be satisfied with that answer</div>
  <div class="k">The horizon</div>         <div class="v">decides it &middot; and nothing in the evaluation measured it</div>
  <div class="k">Always ask</div>          <div class="v">what is excluded from the cost, and what would change the answer</div>
</div>
```

::: {.ask}
The single sentence to take to a meeting:

**"What assumption would have to change for this ratio to fall below 1, and what
is your evidence for that assumption?"**

An evaluator who can answer that has thought about their model. One who cannot
has run it.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1</div>  <div class="v">is the number answering your question, and is it big enough?</div>
  <div class="k">Day 2</div>  <div class="v">which method produced it, and what does that method assume?</div>
  <div class="k">Today</div>  <div class="v">what is it worth, and which assumption decides that?</div>
  <div class="k">Next session</div> <div class="v">reading the figures that carry these claims into a room</div>
</div>
```

::: {.panel}
Next: the same claims, in the form they actually reach a decision-maker — a
chart. Every assumption you have interrogated this morning is invisible in the
picture, and the picture is what gets remembered.
:::
