---
title: "Is This Evidence Credible?"
subtitle: "Module 2 · Day 4, Session 1: A Quality-Assurance Deep Dive"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"

# ---- The report and the rating scale --------------------------------------
# Shared with the printable rating sheet and the report itself, so the deck
# cannot disagree with the paper the room is holding.
source("qa_checklist.R")
source("qa_rating.R")

# Traffic-light colours for the on-screen reveal.
LIGHT_COL <- c(Green = "#2E7D4F", Amber = ACCENT2, Red = ALERT)

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · Sign-off

## The same question, on paper

::: {.panel}
Yesterday you interrogated an analyst who was in the room, and you did it with
questions you wrote yourselves.

Today the analyst is not here. In their place is a report — and the report cannot
answer you back.
:::

::: {.ask .fragment}
**Would you sign off on this evaluation for national scale-up?**
Yes / No / Not without more answers.
:::

## Before you open it

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">AED per business per year &middot; the bar to clear</div>
</div>
```

That is the threshold the commissioning authority set. Everything in the report
is supposed to be measured against it.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.ask}
Vote now, before reading. You will vote again at the end of the session.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
The report is four pages. It is well written, it is confident, and it contains
several real strengths. It is not a bad report.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
That is what makes it hard. You are not looking for errors in a bad document.
You are deciding whether a good document supports the claim it makes.
:::
:::::

::::::

# 1 · The report

## What is in front of you

```{r}
#| results: asis
tab <- data.frame(
  Section = c("1. Executive summary", "2. The programme", "3. Evaluation design",
              "4. Data", "5. Results", "6. Limitations",
              "7. Conclusions and recommendations", "Annexes A and B"),
  `Pages` = c("1", "1", "1", "1", "1", "1", "1", "1"),
  check.names = FALSE
)
kable(tab, col.names = c("Section", "Pages"),
      caption = "The report, section by section") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.ask}
Read the executive summary first, on its own. Then turn the page. Do not read
ahead.
:::

## What the report claims

::: {.excerpt}
This evaluation finds that the programme reduced waste-management costs by
**approximately 1,000 AED per business per year**, and that the estimate is
statistically significant. The finding is robust across four independent
estimation methods… The programme therefore meets the threshold for scale-up.
:::

::: {.panel .fragment}
Four claims in three sentences. Note each one, because you will rate them
separately:

- a **number**, described as approximately 1,000
- a claim of **statistical significance**
- a claim of **robustness** across four methods
- a claim that the **threshold is met**
:::

## Read the summary and rate it

::: {.ask}
On your rating sheet: **the executive summary**.

Give it a colour. Then write the one reason that decided it — and if you rate it
amber or red, what would settle it.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Five minutes.** Work alone, not in groups. The point is your own reading, before
anyone else's.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
If you found nothing wrong, that is a useful result and you should say so. Hold
your colour and we will come back to it.
:::
:::::

::::::

# 2 · The four areas, applied

## The checklist, on paper this time

```{r}
#| results: asis
dom <- data.frame(
  Area = qa_domains,
  `What you are testing` = unname(qa_purpose[qa_domains]),
  check.names = FALSE
)
kable(dom, col.names = c("Area", "What you are testing"),
      caption = "The four areas from yesterday") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Yesterday you asked these out loud and the analyst could answer. Today the report
has to answer, and it answers only what it chose to write down.
:::

::: {.ask .fragment}
Which of the four areas can a report make look strong without actually being
strong?
:::

## The five questions, and what a report can hide

```{r}
#| results: asis
for (i in seq_along(qa_headline)) {
  cat(sprintf('<div class="panel"><b>%d.</b> %s</div>\n', i, qa_headline[i]))
}
```

::: {.warn .fragment}
Notice what is missing from that list: **the colour of the conclusion**. None of
the five asks whether the report sounds convincing. A confident report and a
credible report are different objects, and this one is confident.
:::

## What the rating means

```{r}
#| results: asis
for (i in seq_len(nrow(rating_scale))) {
  cat(sprintf('<div class="panel"><b style="color:%s">%s.</b> %s</div>\n',
              LIGHT_COL[[rating_scale$light[i]]], rating_scale$light[i],
              rating_scale$meaning[i]))
}
```

::: {.ask .fragment}
Amber is not a soft red. It is the most useful rating on the scale, because it
forces you to name the one thing that would change your mind.
:::

# 3 · Section by section

## Evaluation design

::: {.excerpt}
The parallel trends assumption underlying the difference-in-differences design is
the standard assumption in this literature and is satisfied by construction in
programmes of this type, since treatment and control neighbourhoods were selected
from the same administrative area and are subject to the same economic conditions.
:::

::: {.ask}
Rate **Section 3, Evaluation design**. Then answer this: what does parallel trends
actually require?
:::

::: {.warn .fragment}
Parallel trends requires that the two groups **would have moved in parallel
without the programme**. Sharing an administrative area and facing the same
economic conditions is about the *level* of a shared shock, not about whether the
two groups were on the same path.

And no pre-trend test is reported anywhere in the document. The assumption is
asserted, in the passive voice, and the word "satisfied" is doing all the work.
:::

## Data

::: {.excerpt}
Of the businesses present in the baseline round, 12 per cent were not present in
the follow-up round. This reflects the normal churn of the business register.

…the survey instrument separated expenditure on collection from expenditure on
treatment… This change improves the precision of the measure and was applied
consistently across treatment and control groups.
:::

::: {.ask}
Rate **Section 4, Data**. There are two disclosures here. Which is handled well,
and which is not?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**Handled well:** the instrument change is *disclosed*. Most reports would not
mention it.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**Not handled:** disclosure is not the same as analysis. Neither change is tested
for its effect on the outcome, and "applied consistently" does not help if it
shifted the level of the measure.
:::
:::::

::::::

## Results

::: {.excerpt}
All four estimates are negative and statistically significant… The largest
estimate, from the randomised comparison, indicates a cost reduction of
1,014 AED per business per year. The randomised comparison is our headline
result.
:::

::: {.ask}
Rate **Section 5, Results**. Four estimates are given. Why is this one the
headline?
:::

::: {.panel .fragment}
The report gives a reason: the randomised comparison covers all eligible
businesses, the population that would actually be scaled up.

That reason is defensible. Now check whether it is the *only* reason, because
1,014 is also the largest number in the table, and the one that most clearly
clears 1,000.
:::

## The table you should look at twice

```{r}
#| results: asis
tab <- data.frame(
  Method = c("Difference-in-differences", "Regression discontinuity",
             "Randomised comparison", "Matched comparison"),
  Estimate = c("-816", "-905", "-1,014", "-1,000"),
  `Clears 1,000?` = c("No", "No", "Yes", "Exactly at it"),
  check.names = FALSE
)
kable(tab, col.names = c("Method", "Estimate (AED)", "Clears the 1,000 threshold?"),
      caption = "Table 1 of the report, with one column added") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.warn .fragment}
One of four clears the bar. The report calls the one that fails its **preferred
specification** — and says so in Section 3, four pages before the recommendation.
:::


::: {.notes}
Reminder for the room: read each estimate with its **95% confidence interval**, and remember that clearing the 1,000 threshold means the whole confidence interval clears it, not just the point estimate. Tabulate the numbers yourself before accepting the report's framing.
:::

## Limitations

::: {.excerpt}
We note three limitations. First, the randomised comparison estimates the effect
of being offered the programme, not of enrolling… Second, the regression
discontinuity estimate is local to the eligibility cut-off… Third,
self-reported expenditure may be subject to recall error…
:::

::: {.ask}
Rate **Section 6, Limitations**. Are these the right three?
:::

::: {.warn .fragment}
All three are correct and all three are harmless. They are textbook limitations
that any reviewer would accept.

The two limitations that actually threaten this conclusion — the 12 per cent
attrition and the change in the survey instrument — appear in Section 4 and are
never mentioned again.
:::

## Conclusions

::: {.excerpt}
**Recommendation 2.** The subsidy rate should be maintained at its current level,
since the evaluation provides no evidence that a lower rate would be less
effective.

**Recommendation 3.** The eligibility cut-off should be reviewed, as the
regression discontinuity results suggest that businesses just below the threshold
may also benefit.
:::

::: {.ask}
Rate **Section 7, Conclusions**. What was estimated about subsidy rates, and about
businesses below the cut-off?
:::

::: {.warn .fragment}
Nothing, and something different.

Recommendation 2 turns an absence of evidence into a recommendation to do nothing.
Recommendation 3 rests on a result that describes businesses *near* the cut-off,
not those *below* it, and even then, the estimate there was 905, not 1,014.
:::

# 4 · The verdict

## The same question

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">The bar, and the number the report never produced</div>
</div>
```

Vote again. Same three options.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.ask .fragment}
If your vote moved, name the section that moved it. That section is the finding of
this session.
:::

## The most important problem

```{r}
#| results: asis
cat(sprintf('<div class="warn">%s</div>\n', rating_decisive))
```

::: {.ask .fragment}
Which of the six sections did you rate **green**? Keep those — a report that is
wrong about one thing is not wrong about everything, and saying which parts hold
is part of the job.
:::

## The strengths are real

::: {.panel}
Before we close, the report's genuine merits, because a QA exercise that only
collects defects teaches you to read for defects:

- It **states the threshold up front**, so the reader knows the standard being applied.
- It **reports four methods** with their different populations named, rather than one number.
- It **discloses** the survey instrument change, which most reports would bury.
- It gives **confidence intervals for every estimate**, and an annex of robustness specifications.
- Its **limitations are technically correct**, even though they are the wrong ones.
:::

::: {.ask .fragment}
Now the hard question: does any of that rescue the recommendation?
:::

# 5 · Put AI to work

## The prompt

::: {.excerpt}
Here is an evaluation report on a programme that subsidised waste-management
technology for businesses. It found a cost reduction of approximately 1,000 AED
per business per year and recommends national scale-up. Assess the credibility of
this evidence and tell me whether the recommendation is sound.
:::

::: {.panel}
You have just done this by hand, section by section, with a rating sheet. Now watch
a model do it in one pass, and compare its verdict with yours.
:::

## What came back

::: {.ai-output}
**Assessment: the evidence is credible and the recommendation is sound.**

The report is unusually thorough for this kind of evaluation. Several points
support its credibility:

**Multiple methods.** The use of four estimation approaches, all producing
consistent results, is a strong indicator of robustness. When independent methods
agree, the finding is unlikely to be an artefact of any single specification.

**Large sample.** With roughly 9,900 observations, the study is well powered and
the estimates are precise.

**Transparency.** The report discloses a change in the survey instrument between
rounds and notes that 12 per cent of businesses were lost to follow-up. This level
of disclosure suggests a careful and honest evaluation team.

**Appropriate limitations.** The limitations section correctly identifies that the
randomised comparison estimates the offer effect and that the RDD estimate is local to the cut-off.

**Recommendation.** The headline estimate of 1,014 AED exceeds the 1,000 AED
threshold, so [the programme meets the criterion for scale-up]{.flag}. The
recommendation is [supported by the evidence]{.flag}.
:::

::: {.ask .fragment}
Find the two flagged claims. Then find the larger problem, which is not flagged.
:::

## What AI got, and what it could not get

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- Found the real disclosures and called them transparency
- Correctly described the offer-effect and locality limitations
- Produced a structured, readable assessment
- Reached a verdict, which is what was asked
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it could not do**

- It read "approximately 1,000" as if a method had produced 1,000. None did.
- It never noticed that the preferred specification is 816, and fails the threshold.
- It treated disclosure of attrition as evidence of quality, not as an open question.
- It has no way to ask whether 1,014 was *chosen* as the headline.
:::

::::::

::: {.ask .fragment}
Which of the four failures is the most dangerous in practice? Which would you
notice, and which would slip through?
:::

## The same task, a better prompt

::: {.excerpt}
The same report claims a cost reduction of approximately 1,000 AED against a
decision threshold of 1,000 AED. Do three things.

First, list every estimate the report gives, with its confidence interval, and say
for each one whether it clears the threshold.

Second, identify any claim in the executive summary that is not supported by the
results section, and quote both.

Third, tell me what the report does not tell me that I would need in order to
decide.
:::

::: {.panel .fragment}
The first instruction makes the model tabulate rather than summarise, and the
816 becomes impossible to miss. The second forces a comparison the model would
otherwise never make. The third gives it permission to find an absence.
:::

::: {.ask .fragment}
Note what all three have in common: **they ask for a comparison, not a verdict.**
The model is good at verdicts. It needs help with comparisons.
:::

# 6 · Take this back to your desk

## Three habits for reading a report

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Tabulate the estimates yourself, with the threshold in its own column. Do not let the report choose which number you see first.</div>\n')
cat('<div class="panel"><b>2.</b> Read the limitations section, then check whether it names the limitations in the data section. If it does not, the analysis stopped early.</div>\n')
cat('<div class="panel"><b>3.</b> Find the preferred specification and check whether the headline is it. When they differ, ask why.</div>\n')
```

::: {.ask .fragment}
Which of the three would have caught the most important problem in this report?
:::

## Tomorrow

::: {.panel}
You have spent two days learning to read other people\'s evidence critically. In
the next session you write your own: a policy brief, from findings to
recommendation.
:::

::: {.panel .fragment}
Bring the ratings you gave today. You will be making the same judgements about
your own draft, and it is harder when it is yours.
:::
