---
title: "Interrogate the Analyst"
subtitle: "Module 2 · Day 3, Session 3: The Quality Assurance Clinic"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the decks are HTML so that fragments, the Menti
# QR and speaker notes all work as one file. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"

# ---- The checklist itself -------------------------------------------------
# Defined once in qa_checklist.R and shared with the printable handout, so the
# slide and the handout cannot drift apart.
source("qa_checklist.R")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · Sign-off

## Three days of reading. Today you ask the questions.

::: {.panel}
On Monday you learned what a comparison is. On Tuesday you read three more ways of building one. On Wednesday morning you costed a programme and took five figures
apart.

Until now you have read someone else's analysis. Today you get an evaluation
and decide whether you would act on it.
:::

::: {.ask .fragment}
The analyst who ran it is in this room, playing themselves. You have the
checklist. They have to answer.
:::

## Would you sign off?

:::: {.columns}

::: {.column width="58%"}
The question for the next two hours:

```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">AED per business per year &middot; the bar to clear</div>
</div>
```

Vote before we start. You will vote again at the end, and the two votes are
compared.
:::

::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
::::

::: {.ask}
Menti: **Would you sign off on this evaluation for national scale-up?**
Yes / No / Not without more answers.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Hold your position.** Tomorrow morning you rate the same evidence
section by section.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
A show of hands is enough if the Menti is not running. What matters is that the
room is on record *before* the evaluator speaks.
:::
:::::

::::::

# 1 · The claim on the table

## What is being offered for sign-off

::: {.panel}
An evaluation of a programme that subsidises waste-management technology for
businesses. It ran in selected neighbourhoods. The published claim is that the
programme **reduced waste-management costs by about 1,000 AED per business per
year**, clearing the rule for national scale-up.
:::

:::::: {.columns}

::::: {.column width="50%"}
**What the evidence consists of**

- A difference-in-differences estimate
- A regression discontinuity at an eligibility cut-off
- A randomised comparison of treatment and control neighbourhoods
- A matched comparison
- A benefit-cost ratio over five years
:::::

::::: {.column width="50%"}
**What the analyst says it shows**

- Four methods, same direction, similar size
- The result is statistically significant
- The programme pays for itself
- Nothing in the analysis depends on a single fragile choice
:::::

::::::

::: {.ask .fragment}
You already know one of those four statements is not true. Which one, and how
would you show it?
:::

## The numbers, as presented

```{r}
#| echo: false
#| results: asis
tab <- data.frame(
  Method = c("Difference-in-differences", "Regression discontinuity",
             "Randomised offering", "Matched comparison"),
  Estimate = c("-816", "-905", "-1,014", "-1,000"),
  `What it covers` = c("All enrolled businesses", "Businesses near the cut-off",
                       "All eligible businesses", "Matched enrolled businesses"),
  check.names = FALSE
)
kable(tab, col.names = c("Method", "Estimate (AED)", "What it covers"),
      caption = "Cost reduction per business per year, by method") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 17)
```

::: {.warn .fragment}
Every method lands between 816 and 1,014 AED, and the rule is 1,000. **Only the
randomised estimate clears the bar outright; the matched estimate sits exactly on
it, and the difference-in-differences and discontinuity estimates fall short.**
:::

## The rule, and the wiggle room inside it

:::::: {.columns}

::::: {.column width="44%"}
```{=html}
<div class="facts">
  <div class="k">The rule</div>    <div class="v">At least 1,000 AED per business per year</div>
  <div class="k">The range</div>   <div class="v">816 to 1,014 AED</div>
  <div class="k">The claim</div>   <div class="v">"About 1,000"</div>
  <div class="k">The decision</div><div class="v">Scale up nationally</div>
</div>
```
:::::

::::: {.column width="56%"}
::: {.panel}
Notice what the claim does. It reports a number the evidence never produced
exactly, brackets a range that straddles the threshold, and then treats the
threshold as crossed.
:::

::: {.ask .fragment}
Is that a rounding convention, or is it the decision being made by the choice of
where to put the decimal point?
:::
:::::

::::::

## First question of the day

::: {.ask}
**Which of the five estimates should the decision rest on, and why not the
others?**
:::

::: {.panel .fragment}
This is the hardest question in the session. Each estimate
describes a different population: all eligible businesses, those near a cut-off,
the matched enrolled businesses, and the enrolled in treatment neighbourhoods.

An evaluator who cannot say which population the decision concerns cannot say
which estimate supports it.
:::

# 2 · The checklist

## Four areas, in the order they break

```{r}
#| echo: false
#| results: asis
dom <- data.frame(
  Area = qa_domains,
  `What you are testing` = unname(qa_purpose[qa_domains]),
  check.names = FALSE
)
kable(dom, col.names = c("Area", "What you are testing"),
      caption = "The four areas of the checklist") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 17)
```

::: {.panel}
The order matters. Methodology comes first because a design that cannot answer
the question cannot be rescued by good data. Conclusions come last because a
conclusion is only as strong as the three areas underneath it.
:::

## The five to ask every time

```{r}
#| echo: false
#| results: asis
for (i in seq_along(qa_headline)) {
  cat(sprintf('<div class="panel"><b>%d.</b> %s</div>\n', i, qa_headline[i]))
}
```

::: {.warn}
Five questions are easy to hold in your head in a meeting. The printed handout
has all twenty, five under each area.
:::

## How to tell a strong question from a weak one

:::::: {.columns}

::::: {.column width="50%"}
**Weak**

- Was the study rigorous?
- Are the results significant?
- Was the sample size adequate?
- Did you control for confounders?
- Is the data reliable?
:::::

::::: {.column width="50%"}
**Strong**

- Which assumption would reverse the sign?
- What is the estimate and its confidence interval?
- How many units were dropped, and why?
- What changed at the same time?
- Who is missing from the data?
:::::

::::::

::: {.ask .fragment}
A weak question can be answered
"yes" by anyone. A strong question requires a fact.
:::

## Your turn: sort these

```{r}
#| echo: false
#| results: asis
cands <- c(
  "Did the programme work?",
  "What would have happened to these businesses without the programme?",
  "Was the sample large enough?",
  "How many businesses dropped out between the two survey rounds?",
  "Were the results statistically significant?",
  "Which assumption, if wrong, would flip the finding?",
  "Is this good quality data?",
  "If the effect were half this size, would you still recommend scale-up?"
)
cat('<div class="excerpt">\n')
for (i in seq_along(cands)) cat(sprintf('<p><b>%d.</b> %s</p>\n', i, cands[i]))
cat('</div>\n')
```

::: {.ask}
Mark each one **S** or **W**. Then tell me what makes the S ones answerable.
:::

## Which of those survived

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Weak, and why**

- *Did the programme work?* Answerable with one word.
- *Was the sample large enough?* For what? Precision is not the binding constraint here.
- *Were the results significant?* A yes/no where the number is what matters.
- *Is this good quality data?* Invites an adjective, not a fact.
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**Strong, and why**

- *What would have happened without the programme?* Asks for the counterfactual, which the whole week has been about.
- *How many dropped out?* A count the analyst either has or does not.
- *Which assumption would flip it?* Forces the fragility into the open.
- *Half the size?* Tests whether the estimate is carrying the decision.
:::

::::::

::: {.ask .fragment}
Notice that none of the strong questions is harder to *ask*. They are harder to *answer*.
:::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the clinic, with the analyst in the room.
:::

# 3 · The clinic

## How this runs

:::::: {.columns}

::::: {.column width="58%"}
::: {.panel}
The analyst answers as themselves. Groups fire questions, one area at a time.

**The rules of engagement**

1. One question at a time, and the answer is given before the next question
2. Any group may follow up once on an answer
3. The analyst may answer *"I don't know"* or *"we didn't check that"*
4. An admission is useful information, not a failure
5. Nothing is scored. The point is the questions, not the stump
:::
:::::

::::: {.column width="42%"}
::: {.warn}
**Groups:** four, one per area. Methodology, assumptions, data quality,
conclusions.

You have five minutes to agree your three sharpest questions before the clinic
opens.
:::
:::::

::::::

::: {.ask .fragment}
Write your questions down. You will want them for the debrief, and for tomorrow.
:::

## The questions that cost the analyst something

::: {.panel}
Some questions have answers that make the analysis look better. Some have answers
that make it look worse. You are looking for the second kind.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**Listen for these phrases**

- "That's a good question…"
- "In the main specification, yes"
- "We discussed that"
- "That was outside the scope"
- "The data doesn't let us"
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
**Translate**

- *Which other specification?*
- *What happened in the ones that aren't main?*
- *And what did you decide, and why?*
- *Whose scope? Who set it?*
- *What would you need to answer it?*
:::

::::::

::: {.ask .fragment}
An answer that is true but narrower than the claim still leaves a gap. Follow
it up.
:::

## The one you must not let go

::: {.ask}
**What is the comparison, and how was it chosen?**
:::

::: {.panel .fragment}
Every method on Monday's and Tuesday's slides rests on a comparison group
that was equivalent to the participants before the programme began: because it
was randomly assigned, or because it was just below a cut-off, or because it was
matched on observables, or because it was trending in parallel. Those are four
different reasons, and each has its own way of being false.

An evaluator who cannot state which one applies here, in a sentence, has not
finished the analysis.
:::

# 4 · The debrief

## The unanswered questions are the finding

::: {.panel}
Now we count. For each area, how many of your questions got a definite answer,
and how many got an admission?
:::

:::::: {.columns}

::::: {.column width="50%"}
```{=html}
<div class="facts">
  <div class="k">Answered</div>     <div class="v">The fact exists and was produced</div>
  <div class="k">Admitted</div>     <div class="v">"We didn't check that"</div>
  <div class="k">Deflected</div>    <div class="v">Answered a different question</div>
  <div class="k">Unanswerable</div> <div class="v">The design cannot settle it</div>
</div>
```
:::::

::::: {.column width="50%"}
::: {.ask .fragment}
Which of the four is *not* a failure, and which one tells you the analysis is
weaker than it looked?
:::
:::::

::::::

::: {.warn .fragment}
A deflection is worse than an admission. An admission tells you where the edge of
the evidence is. A deflection tells you the analyst may not know either.
:::

## Which unanswered question changes the decision

::: {.ask}
Of everything the analyst could not answer, which **one** would move your
sign-off if it came back the wrong way?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Most unanswered questions are interesting and not decisive. They belong in the
limitations section and nowhere else.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
One or two are decisive. Those are the ones worth commissioning more work on, or
writing into the decision as a condition.
:::
:::::

::::::

::: {.ask .fragment}
If nothing you raised would change the decision, either the evaluation is strong
or your questions were not pointed at the decision. Which is it?
:::

## Would you sign off now?

:::::: {.columns}

::::: {.column width="58%"}
```{=html}
<div class="hero">
  <div class="value">1,000</div>
  <div class="label">The same bar, after the clinic</div>
</div>
```

Same Menti, same three options. The room has now heard the analyst answer.
:::::

::::: {.column width="42%"}
```{r}
#| fig.height: 3.4
#| fig.width: 3.4
#| fig.align: center
knitr::include_graphics(MENTI_QR)
```
::::

::: {.ask .fragment}
If your vote moved, say which answer moved it.
:::

# 5 · Put AI to work

## The prompt

::: {.excerpt}
I have an evaluation of a programme that subsidised waste-management technology
for businesses. It used difference-in-differences, regression discontinuity,
a randomised comparison and matching. All four methods found a cost reduction of
roughly 1,000 AED per business per year. List the critical questions a
commissioner should ask before acting on this result.
:::

::: {.panel}
A reasonable prompt, and a common one. Run it yourselves, then compare what you
get with the example on the next slide and with your own questions.
:::

## What came back

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

Here are the critical questions a commissioner should ask:

**1. Was the study well designed?** Confirm that the evaluation used a
comparison group and a pre-post design, which this one did.

**2. Was the sample size adequate?** A sample of roughly 9,900 observations is
large, so statistical power is [not a concern here]{.flag}, and the results can be
trusted.

**3. Were the results statistically significant?** This should be confirmed, as
significance is the main indicator that a result is real.

**4. Were confounders controlled for?** The mention of four methods suggests
robustness.

**5. Is the data reliable?** Check for missing values and obvious errors.

**6. Is the result policy relevant?** The cost reduction appears to exceed the
stated threshold.

**Overall:** the evaluation appears methodologically strong. The convergence of
four methods on a similar estimate is [strong evidence of causality]{.flag}.
:::

::: {.ask .fragment}
Find the two errors flagged above. Then find the bigger problem, which is not flagged.
How does the list your AI gave you compare?
:::

## What AI got, and what it could not get

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it did well**

- Produced the *shape* of a critique
- Covered the right broad topics
- Read fluently and organised itself
- Sounded confident, which is easy to mistake for competence
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
**What it missed**

- Sample size does not address selection bias. The two flagged errors are both that confusion.
- It cannot know *which* assumption this design rests on, because that depends on facts it was never given.
- It never asks what the comparison was.
- It has no way to say "we did not check that".
:::

::::::

::: {.ask .fragment}
Which of its six questions would have survived your sorting exercise as **strong**?
:::

## The same task, a better prompt

::: {.excerpt}
I am advising on whether to scale up a programme nationally. An evaluator reports
a cost reduction of about 1,000 AED per business per year from
difference-in-differences, with a decision threshold of 1,000 AED.

Do three things. First, tell me which single assumption this design rests on.
Second, tell me exactly what evidence would show that assumption holds, and what
it would look like if it failed. Third, tell me what you cannot determine from
the information I have given you.
:::

::: {.panel .fragment}
Three changes.

**Ask for the one thing**, not a list. **Ask what would falsify it**, which forces
a testable answer. And **ask what it cannot tell you**, which gives the model
permission to admit a limit instead of writing around it.
:::

::: {.ask .fragment}
The third instruction often makes the biggest difference. Try it on your own
material this week.
:::

# 6 · Take this back to your desk

## The three you will actually use

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**From memory, no notes**

1. What is the comparison, and how was it chosen?
2. What has to be true for this to be causal, and did you test it?
3. Does the conclusion follow, or does it outrun the evidence?
:::
:::::

::::: {.column width="50%"}
::: {.warn}
If you remember only one, remember the first. Every method this week has been a
different answer to it, and every method has a different way of being wrong about
it.
:::

::::::

::: {.ask .fragment}
Next time someone brings you a result, which of the three will you ask first, and does it depend on the method they used?
:::

## Tomorrow

::: {.panel}
The full checklist, applied to a written report, section by section, with a
traffic-light rating for each.

You have just done the hard version. Tomorrow you do the structured one.
:::

::: {.panel .fragment}
Bring the questions you wrote down today.
:::
