---
title: "Spot the Problem: Naive Comparisons and Reading an RCT"
subtitle: "Module 2 · Day 1, Session 3: Two Numbers Someone Might Report, and a Lottery"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources
# must sit next to the published .html. Static knitr chunks work as before.
#
# The simple GreenWaste case (SIMPLE_GREENWASTE_PLAN.md): one row per business;
# a lottery in the pilot district, a score rule at 58 in the rest of the city.
# Every number the static slides quote comes from greenwaste_case.R.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  # Data frames print as HTML tables (styled in theme_live.scss) instead of
  # console text. knitr is already loaded by quarto-live, so this costs nothing.
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
    - gw_lab.R
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate: knitr sizes htmlwidgets as fig.width * dpi.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(estimatr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"   # Tech / Light Blue
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

PL_W <- 1100; PL_WC <- 520; PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# Emit an HTML snippet as a raw block. Plain cat() leaves it to pandoc, which
# reads indented lines inside a multi-line <div> as a code block and prints the
# tags on the slide.
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000    # must cut costs by at least this many AED

# ---- The numbers the static slides quote -----------------------------------
# greenwaste_case.R computes every headline number once. It is sourced into its
# own environment because it defines RULE (1,000), which here is a colour.
case_env <- new.env()
case_env$gw <- read.csv("./evaluation_data_GreenWaste_simple.csv")
sys.source("greenwaste_case.R", envir = case_env)
case   <- case_env$case
pilot  <- case_env$pilot
city   <- case_env$city
took   <- case_env$took

ba_change <- case$before_after                  # city, took part: after minus before
ww_gap    <- case$with_without                  # city, took part vs not, after
ww_gap0   <- case$gap_before                    # the same gap, before the programme
city_rise <- mean(city$change[!took])           # costs of city businesses that did not take part
age_gap   <- mean(city$manager_age[!took]) - mean(city$manager_age[took])

rct_fit     <- lm(cost_after ~ took_part, data = pilot)
rct_est     <- coef(rct_fit)[["took_part"]]
rct_control <- coef(rct_fit)[["(Intercept)"]]
rct_ci      <- confint(rct_fit)["took_part", ]
# SIGN TRAP: sort the absolute values so [1] is always the end closest to zero.
rct_avoided       <- sort(abs(rct_ci))
rct_avoided_least <- rct_avoided[1]
rct_avoided_most  <- rct_avoided[2]

# The same smaller-trial draw as the live slider (same seed, same rule), so the
# text quotes what the slider shows.
small_trial_ci <- function(k) {
  set.seed(11)
  keep <- c(sample(which(pilot$took_part == 1), k / 2), sample(which(pilot$took_part == 0), k / 2))
  f <- lm_robust(cost_after ~ took_part, data = pilot[keep, ])
  sort(-confint(f)["took_part", ])
}
st20 <- small_trial_ci(20)

# How far apart a 4-business lottery can start: 95% of 1,000 draws fall within this.
crowd_q <- function(k, draws = 1000) {
  set.seed(7)
  quantile(abs(replicate(draws, { s <- sample(nrow(pilot), k)
    mean(pilot$cost_before[s[seq_len(k / 2)]]) - mean(pilot$cost_before[s[-seq_len(k / 2)]]) })), 0.95)
}
q4 <- crowd_q(4)

# Estimate, interval and rule on one axis.
rule_chart <- function(est, least, most, rule, xlab, xmax, digits = 0) {
  fmt <- paste0("avoids %{x:,.", digits, "f}<extra></extra>")
  tibble(
    what = factor(c("The estimate", "95% interval", "Decision rule"),
                  levels = c("Decision rule", "95% interval", "The estimate")),
    lo = c(abs(est), least, rule), hi = c(abs(est), most, rule),
    col = c(ACCENT, ACCENT, ALERT)
  ) %>%
    plot_ly() %>%
    add_segments(x = ~lo, xend = ~hi, y = ~what, yend = ~what,
                 line = list(color = ~col, width = 5), showlegend = FALSE,
                 hovertemplate = fmt) %>%
    add_markers(x = ~lo, y = ~what, marker = list(color = ~col, size = 11),
                showlegend = FALSE, hovertemplate = fmt) %>%
    layout(xaxis = ax(xlab, range = c(0, xmax)),
           yaxis = ax("", zeroline = FALSE)) %>%
    pl_m2(width = PL_WC, height = 250)
}

# ---- The case introduction: one illustrative city -----------------------------
# Shop position, efficiency score, and whether it joined. Inside the pilot
# district every score is below 58 and a lottery decided; everywhere else a
# score of 58 or below meant joining. Illustrative, not the real counts.
SHOP_BLUE <- "#215a9e"; SHOP_GREY <- "#c9d6e6"
pilot_shops <- data.frame(x = c(70, 165, 260, 70, 165, 260), y = c(98, 98, 98, 186, 186, 186),
                          score = c(41, 52, 37, 48, 33, 55), joined = c(1, 0, 1, 0, 0, 1))
city_shops  <- data.frame(
  x = c(415, 525, 635, 745, 415, 525, 635, 745, 70, 180, 290, 415, 525, 635, 745),
  y = c(80, 80, 80, 80, 195, 195, 195, 195, 330, 330, 330, 330, 330, 330, 330),
  score = c(72, 45, 88, 39, 56, 63, 29, 91, 47, 77, 52, 68, 35, 84, 58))
city_shops$joined <- as.integer(city_shops$score <= 58)
shop_svg <- function(d) paste(sprintf(
  '<use href="#shop" x="%d" y="%d" width="62" height="58" color="%s"/><text x="%d" y="%d" font-size="19" text-anchor="middle" fill="%s" font-weight="%s">%d</text>',
  d$x, d$y, ifelse(d$joined == 1, SHOP_BLUE, SHOP_GREY), d$x + 31, d$y + 80,
  ifelse(d$joined == 1, "#063360", "#545860"), ifelse(d$joined == 1, "700", "400"), d$score), collapse = "\n")
```

# 0 · The GreenWaste case {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The GreenWaste programme

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the live cells
# below use is defined here, so each cell on a slide stays two or three lines.
gw    <- read.csv("evaluation_data_GreenWaste_simple.csv")
pilot <- subset(gw, setting == "pilot")    # the lottery
city  <- subset(gw, setting == "city")     # the score rule
source("gw_lab.R")   # the GreenWaste lab: an invented world where the true effect is known
library(estimatr)
RULE <- 1000
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Warm up estimatr now, while the opening slides are talked through: its first
# call in the browser takes about 10 seconds, every later one is instant.
invisible(estimatr::lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1))))

# Defined here as well as on the slide that teaches it, so a later slide still
# runs if an earlier one was skipped or the page was reloaded mid-deck.
fit <- lm(cost_after ~ took_part, data = pilot)

clean_par <- function() par(mar = c(3, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)
aed <- function(x) format(round(x), big.mark = ",")

# Bar chart of a few averages, value printed on each bar:
#   bars(Before = before, After = after)
bars <- function(..., col = ACCENT) {
  v <- c(...)
  clean_par()
  par(mar = c(2.5, 1, 2, 1))               # no y axis: the value sits on each bar
  b <- barplot(v, col = col, border = NA, ylim = c(0, max(v) * 1.15), axes = FALSE)
  text(b, v, paste(aed(v), "AED"), pos = 3, cex = 1.2, xpd = NA)
}

# Before-and-after lines for the two groups in the rest of the city.
plot_two_lines <- function() {
  t1 <- city$took_part == 1
  took <- c(mean(city$cost_before[t1]),  mean(city$cost_after[t1]))
  oth  <- c(mean(city$cost_before[!t1]), mean(city$cost_after[!t1]))
  clean_par()
  plot(0:1, took, type = "b", pch = 19, lwd = 4, col = ACCENT, bty = "n",
       ylim = c(0, 2700), xlim = c(-0.1, 1.6), xaxt = "n",
       xlab = "", ylab = "Waste costs, AED")
  lines(0:1, oth, type = "b", pch = 19, lwd = 4, col = ACCENT2)
  axis(1, at = 0:1, labels = c("Before", "After"))
  text(1.06, took[2], paste("Took part", aed(took[2])), col = ACCENT, adj = 0)
  text(1.06, oth[2],  paste("Did not", aed(oth[2])),    col = ACCENT2, adj = 0)
}

# Draw the pilot lottery again (or let officials pick the cheapest-to-run 200)
# and compare the two groups' costs BEFORE the programme.
draw_lottery <- function(how, seed) {
  set.seed(seed)
  picked <- if (how == "Random draw") sample(nrow(pilot), nrow(pilot) / 2) else
            order(pilot$cost_before)[seq_len(nrow(pilot) / 2)]
  m <- c(Chosen = mean(pilot$cost_before[picked]),
         `Not chosen` = mean(pilot$cost_before[-picked]))
  clean_par()
  b <- barplot(m, col = c(ACCENT, ACCENT2), border = NA, ylim = c(0, 2000),
               ylab = "Costs before, AED")
  text(b, m + 110, aed(m), cex = 1.3)
  cat(sprintf("Gap BEFORE the programme started: %+.0f AED\n", m[1] - m[2]))
}

# One thousand lotteries among k businesses: how far apart can the halves start?
crowd_gaps <- function(k, draws = 1000) {
  set.seed(7)
  gaps <- replicate(draws, {
    s <- sample(nrow(pilot), k)
    mean(pilot$cost_before[s[seq_len(k / 2)]]) - mean(pilot$cost_before[s[-seq_len(k / 2)]])
  })
  clean_par()
  hist(gaps, breaks = seq(-2000, 2000, by = 50), col = ACCENT, border = "white",
       main = "", xlab = "Gap in costs before the programme, AED", ylab = "Draws",
       xlim = c(-1200, 1200))
  abline(v = 0, col = ALERT, lwd = 2, lty = 2)
  cat(sprintf("%d businesses: in 95%% of %d lotteries the groups start within +/- %.0f AED\n",
              k, draws, quantile(abs(gaps), 0.95)))
}

# Re-estimate on k pilot businesses (half chosen, half not): a smaller trial.
small_trial <- function(k) {
  set.seed(11)
  keep <- c(sample(which(pilot$took_part == 1), k / 2), sample(which(pilot$took_part == 0), k / 2))
  f   <- lm_robust(cost_after ~ took_part, data = pilot[keep, ])
  est <- -coef(f)[["took_part"]]
  ci  <- sort(-confint(f)["took_part", ])
  clean_par()
  plot(NA, xlim = c(0, 2000), ylim = c(0.5, 1.5), yaxt = "n", bty = "n",
       xlab = "AED avoided per business", ylab = "")
  abline(v = RULE, col = ALERT, lwd = 3, lty = 2)
  segments(ci[1], 1, ci[2], 1, col = ACCENT, lwd = 8)
  points(est, 1, pch = 19, cex = 2.5, col = ACCENT)
  text(RULE, 1.4, "the rule", col = ALERT, pos = 4)
  cat(sprintf("%d businesses: avoids %.0f AED, interval %.0f to %.0f\n", k, est, ci[1], ci[2]))
}
```

<p style="font-size:0.85em; margin: 0 0 4px; max-width: none;">GreenWaste helps businesses cut their waste costs. Here is how it reached them.</p>

```{r city-map, results='asis'}
html_out(paste0('
<svg viewBox="0 0 1200 470" width="100%" role="img"
     aria-label="A city map. In the pilot district a lottery chose which shops joined. In the rest of the city, shops with an efficiency score of 58 or below joined."
     style="font-family: \'Noto Kufi Arabic\', \'Segoe UI\', Arial, sans-serif;">
  <defs>
    <symbol id="shop" viewBox="0 0 60 56">
      <path d="M4 14 L10 2 H50 L56 14 Z" fill="currentColor"/>
      <path d="M4 14 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 12 0" fill="currentColor"/>
      <rect x="8" y="20" width="44" height="34" rx="2" fill="currentColor"/>
      <rect x="14" y="30" width="12" height="24" fill="#fff" opacity="0.85"/>
      <rect x="32" y="28" width="14" height="12" fill="#fff" opacity="0.85"/>
    </symbol>
  </defs>
  <rect x="20" y="20" width="820" height="440" rx="18" fill="#f4f7fb" stroke="#D5DEE8" stroke-width="2"/>
  <path d="M370 20 V460 M20 300 H840" stroke="#e3e9f1" stroke-width="10"/>
  <text x="820" y="450" font-size="22" font-weight="700" fill="#7f97b5" text-anchor="end">THE CITY</text>
  <rect x="45" y="45" width="305" height="245" rx="14" fill="#e3ebf5" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
  <text x="62" y="80" font-size="22" font-weight="700" fill="#063360">Pilot district</text>
  <g transform="translate(300,55)">
    <rect width="38" height="38" rx="7" fill="#fff" stroke="#063360" stroke-width="2.5"/>
    <circle cx="11" cy="11" r="3.5" fill="#063360"/><circle cx="27" cy="11" r="3.5" fill="#063360"/>
    <circle cx="19" cy="19" r="3.5" fill="#063360"/>
    <circle cx="11" cy="27" r="3.5" fill="#063360"/><circle cx="27" cy="27" r="3.5" fill="#063360"/>
  </g>
', shop_svg(pilot_shops), '
', shop_svg(city_shops), '
  <g font-size="21" fill="#111418">
    <rect x="875" y="40" width="300" height="150" rx="12" fill="#fff" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
    <text x="895" y="82" font-weight="700" fill="#063360">In the pilot district</text>
    <text x="895" y="118">a <tspan font-weight="700">lottery</tspan> chose which</text>
    <text x="895" y="148">shops joined</text>
    <rect x="875" y="215" width="300" height="178" rx="12" fill="#fff" stroke="#D5DEE8" stroke-width="2"/>
    <text x="895" y="257" font-weight="700" fill="#063360">Everywhere else</text>
    <text x="895" y="293">shops scoring <tspan font-weight="700">58 or below</tspan></text>
    <text x="895" y="323">joined</text>
    <text x="895" y="356" font-size="16" fill="#545860">the number under each shop</text>
    <text x="895" y="377" font-size="16" fill="#545860">is its efficiency score</text>
  </g>
  <g font-size="19" fill="#111418">
    <use href="#shop" x="878" y="404" width="30" height="28" color="', SHOP_BLUE, '"/><text x="918" y="425">joined GreenWaste</text>
    <use href="#shop" x="878" y="436" width="30" height="28" color="', SHOP_GREY, '"/><text x="918" y="457">did not join</text>
  </g>
</svg>'))
```

::: {.notes}
About 3 minutes. One city; the pilot district is part of it. Point at the pilot
district: every shop there scores below 58, yet only some joined, because a
lottery decided. Then the rest of the city: the number under each shop decided.
Ask: "Where would it be fairer to compare shops that joined with shops that did
not: inside the pilot district or in the rest of the city? Why?" Take answers;
do not resolve it yet. The map is illustrative; the real counts come next.

Open the deck a couple of minutes before the session starts: R and its statistics package take about a minute to load in the browser, after which every live cell runs in a few seconds.

Paper: hand out the GreenWaste picture card (GreenWaste_case_card.docx, one per participant) and the session tracker (Oct12_session3_materials.docx, pack page 1, one per participant); they fill in each row as the numbers appear.
:::

## Today

::: {.panel}
The authority will scale GreenWaste up nationally only if it cuts waste costs by
at least **`r gfmt(RULEBAR)` AED** per business per year. Every business's costs
were measured before the programme and 12 months after.
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Meet the data: one row per business, and who took part where</div>
  <div class="k">2</div> <div class="v">Two numbers someone might report, and what is wrong with each</div>
  <div class="k">3</div> <div class="v">Run the lottery ourselves, and see why it works</div>
  <div class="k">4</div> <div class="v">Read the randomised result against the 1,000 AED rule</div>
</div>
```

::: {.warn}
You will not write any code. You will call the numbers; I will type them in and
press **Run**. The grey boxes are real R, running in this page.
:::

# 1 · Meet the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## One row, one business

```{webr}
head(gw[, c("business", "setting", "score", "took_part", "cost_before", "cost_after")], 4)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**setting**: the pilot district or the rest of the city. **score**: efficiency,
0 to 100. **took_part** 1: the business took part. **cost_before**,
**cost_after**: waste costs in AED.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Which column tells you **how** a business came to take part?
:::

::: {.panel .fragment}
**setting**. In the pilot district a lottery decided; in the rest of the city the
score did. **took_part** records the result.
:::
:::::

::::::

## Who took part, where?

```{webr}
groups <- aggregate(score ~ setting + took_part, data = gw, FUN = length)
names(groups)[3] <- "businesses"
groups$highest_score <- aggregate(score ~ setting + took_part, data = gw, FUN = max)$score
groups <- groups[order(groups$setting), ]; rownames(groups) <- NULL
groups
```

::: {.ask}
**Who took part in the rest of the city?** And in the pilot district?
:::

::: {.warn .fragment}
In the rest of the city, **every business scoring 58 or below took part, and no
other did.** So there "did not take part" means "scored above 58": a different
kind of business. In the pilot district, half took part, all scoring below 58:
the lottery decided.
:::

# 2 · Two numbers someone might report {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Comparison A: before and after

::: {.ask}
A report says waste costs fell for the city businesses that took part.
**Before I press Run: by how much?**
:::

```{webr}
#| fig-width: 6
#| fig-height: 2.5
before <- mean(city$cost_before[city$took_part == 1])
after  <- mean(city$cost_after[city$took_part == 1])
round(data.frame(before, after, change = after - before))
bars(Before = before, After = after)
```

::: {.notes}
Take two or three guesses, then press Run. The room should see
`r gfmt(ba_change)` appear. Point at each line as you read it aloud in plain words.
:::

## Comparison B: with and without

```{webr}
#| fig-width: 6
#| fig-height: 2.5
took  <- mean(city$cost_after[city$took_part == 1])
didnt <- mean(city$cost_after[city$took_part == 0])
round(data.frame(took_part = took, did_not = didnt, gap = took - didnt))
bars("Took part" = took, "Did not" = didnt, col = c(ACCENT, ACCENT2))
```

::: {.warn}
Same city, after the programme. Two readings of one programme,
**`r gfmt(ba_change)`** and **`r gfmt(ww_gap)`**: one misses the 1,000 AED bar,
one clears it.
:::

## `[MENTI PLACEHOLDER]` Which way is each one wrong?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
Both numbers come from real comparisons, and both are biased. The harder question:

**Is `r gfmt(abs(ba_change))` too big or too small?**

**Is `r gfmt(abs(ww_gap))` too big or too small?**
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.notes}
~4 min. Do not settle it. The next three slides answer it from the data.
:::

# 3 · Catch each one out, in the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## A: what happened to everyone else?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 4.2
plot_two_lines()
```
:::::

::::: {.column width="44%"}
::: {.panel}
Before-and-after compares the programme to **the past**. If costs were moving
anyway, that movement is baked into the `r gfmt(abs(ba_change))`.
:::

::: {.ask}
Look at the businesses that did **not** take part. Which way did their costs go
with no programme at all?
:::

::: {.warn .fragment}
Up, by about `r gfmt(city_rise)` AED. Costs were rising anyway, so the
`r gfmt(abs(ba_change))` fall **understates** what the programme did.
:::
:::::

::::::

## B: the same gap, before the programme existed

:::::: {.columns}

::::: {.column width="42%"}
```{ojs}
//| echo: false
viewof when = Inputs.radio(["After the programme", "Before the programme"],
  {value: "After the programme"})
```
:::::

::::: {.column width="58%"}
::: {.ask}
Before the programme, nobody had received anything. **What gap should we see?**
:::
:::::

::::::

```{webr}
#| input:
#|   - when
#| autorun: true
#| fig-width: 6
#| fig-height: 1.8
cost  <- if (when == "Before the programme") city$cost_before else city$cost_after
took  <- mean(cost[city$took_part == 1])
didnt <- mean(cost[city$took_part == 0])
round(data.frame(took_part = took, did_not = didnt, gap = took - didnt))
bars("Took part" = took, "Did not" = didnt, col = c(ACCENT, ACCENT2))
```


::: {.warn .fragment}
It should be zero. It is **`r gfmt(ww_gap0)`** AED: a head start, so the
`r gfmt(abs(ww_gap))` gap **overstates** the effect.
:::

## Why they were already different

```{webr}
round(aggregate(cbind(manager_age, staff, area, filtration) ~ took_part,
                data = city, FUN = mean), 1)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
The rest of the city, **before** the programme. Took part (1) vs not (0) on four
things we measured.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Managers about `r round(age_gap)` years younger, fewer staff, smaller premises.
The two groups were different kinds of business, and in a real evaluation **some
of the ways they differ would not be measured at all.**
:::
:::::

::::::

## Two biases, opposite directions

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Before and after</div>
  <div class="v">%s AED &middot; folds in the rising costs everyone faced &middot; UNDERSTATES</div>
  <div class="k">With and without</div>
  <div class="v">%s AED &middot; folds in a %s AED head start &middot; OVERSTATES</div>
  <div class="k">So the truth</div>
  <div class="v">sits somewhere between the two, and neither comparison can say where</div>
</div>', gfmt(ba_change), gfmt(ww_gap), gfmt(abs(ww_gap0))))
```

::: {.ask}
Check your Menti answer. What would it take to land on the truth
**deliberately**, instead of guessing between two wrong numbers?
:::

::: {.warn}
In both cases **the two groups were not comparable.** After the break we look
at a design that makes them comparable by construction.
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker check: every row should now say too big or too small, with a reason.
:::

## The GreenWaste lab

:::::: {.columns}

::::: {.column width="46%"}
```{ojs}
//| echo: false
viewof lab_trend  = Inputs.range([0, 600], {value: 0, step: 50, label: "Costs rise for everyone anyway (AED)"})
viewof lab_size   = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners are smaller businesses (AED)"})
viewof lab_hidden = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners differ in something nobody measured (AED)"})
```
:::::

::::: {.column width="54%"}
::: {.panel}
An invented GreenWaste of 20,000 businesses where we know the truth: taking
part cuts costs by **900 AED**. Each slider switches on one problem.
:::

::: {.ask}
Before each move: **which estimate will that slider pull away from the truth,
and which way?**
:::
:::::

::::::

```{webr}
#| input:
#|   - lab_trend
#|   - lab_size
#|   - lab_hidden
#| autorun: true
#| edit: false
#| echo: false
#| fig-width: 10
#| fig-height: 2.5
lab(c("before_after", "with_without"), lab_trend, lab_size, lab_hidden)
```

::: {.notes}
About 5 minutes. With every slider at zero both estimates sit on the truth: a
naive comparison is only wrong when the world has these problems, and GreenWaste
had them. Move one slider at a time and take a call first. "Costs rise anyway"
moves only before-and-after (towards zero, as in the GreenWaste data, where the
other businesses' costs rose about `r gfmt(city_rise)` AED). The two "joiners
differ" sliders move only with-and-without, and make it look bigger than the
truth, as the `r gfmt(abs(ww_gap))` did. Leave the sliders where the room ends
up; the lab comes back after the lottery.
:::

# Break {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Five-minute break

::: {.panel}
Back in five minutes. When we return, we run the lottery ourselves.
:::

# 4 · Run the lottery yourself {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What the lottery buys you

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
In the **pilot district** every business scored below 58, so every one could have
joined. A **lottery** picked `r sum(pilot$took_part)` of the `r nrow(pilot)`.

So the randomised comparison is: the businesses the lottery picked vs the ones it
did not.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
The claim is that a draw makes the two groups **alike on average**, on what we
measured *and* on what we did not.
:::

::: {.ask}
If the claim holds, what should the gap in costs **before the programme** between
the two groups look like?
:::
:::::

::::::

## Draw the lottery again

```{ojs}
//| echo: false
viewof how = Inputs.radio(["Random draw", "Officials pick"],
  {value: "Random draw", label: "How the 200 are chosen"})
viewof draw = Inputs.button("Draw again")
```

:::::: {.columns}

::::: {.column width="55%"}
```{webr}
#| input:
#|   - how
#|   - draw
#| autorun: true
#| edit: false
#| fig-width: 5.6
#| fig-height: 3.4
draw_lottery(how, seed = draw)
```
:::::

::::: {.column width="45%"}
::: {.panel}
Bars show costs **before** the programme, for the 400 pilot businesses.
:::

::: {.ask}
Press **Draw again** five times. How big does the gap get? Then let officials
pick.
:::

::: {.warn .fragment}
When officials pick the cheapest to run, the chosen businesses start hundreds of
AED ahead, so the programme looks successful before it has done anything.
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker: record the largest gap after five draws, then the gap when officials pick.
:::

## Four businesses or four hundred?

```{ojs}
//| echo: false
viewof k = Inputs.range([4, 400], {value: 4, step: 2, label: "Businesses in the lottery"})
```

```{webr}
#| input:
#|   - k
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 3.3
crowd_gaps(k)
```

::: {.ask}
One thousand lotteries at each size. Start at 4 businesses, then slide up.
**At what size would you trust a single draw?**
:::

::: {.notes}
At 4 the gaps run to hundreds of AED either way; at 400 almost every draw starts
within a few dozen. This is "two people versus two crowds" from Module 1, seen in
the data. The same holds for characteristics nobody measured.
:::

## `[MENTI PLACEHOLDER]` Would you trust this draw?

::: {.panel}
Suppose GreenWaste had run its lottery among only **6 businesses**, 3 picked and
3 not. Before the programme the picked three already had costs **300 AED lower**.
Afterwards they are **900 AED lower**.
:::

::: {.ask}
What do you conclude? **A** The effect is 900: it was randomised. **B** About 600:
take off the head start. **C** Too few businesses to trust either number.
**D** I need to see something else first. Vote, then say what you would need.
:::

::: {.panel .fragment}
C, with D's instinct. A draw makes groups alike **on average over many draws**,
not in every draw: with a handful of units a 300 AED head start is ordinary (the
slider showed about ±`r gfmt(q4)` AED at 4 businesses). B looks careful but assumes the
head start would have stayed the same. Ask how many units were randomised, and
for the balance table.
:::

::: {.notes}
This asks the room to apply what the two lottery slides showed to a result
they might be handed. Expect a split between
A and B; let one person from each argue before revealing.
:::

## The lab again: add the lottery

:::::: {.columns}

::::: {.column width="46%"}
```{ojs}
//| echo: false
viewof lab2_trend  = Inputs.range([0, 600], {value: 300, step: 50, label: "Costs rise for everyone anyway (AED)"})
viewof lab2_size   = Inputs.range([0, 800], {value: 400, step: 100, label: "Joiners are smaller businesses (AED)"})
viewof lab2_hidden = Inputs.range([0, 800], {value: 400, step: 100, label: "Joiners differ in something nobody measured (AED)"})
```
:::::

::::: {.column width="54%"}
::: {.ask}
Move any slider as far as you like. **Which estimate never moves, and why
not?**
:::

::: {.panel .fragment}
The randomised one. The lottery decides who takes part, so every problem lands
equally on both sides of the comparison and cancels out.
:::
:::::

::::::

```{webr}
#| input:
#|   - lab2_trend
#|   - lab2_size
#|   - lab2_hidden
#| autorun: true
#| edit: false
#| echo: false
#| fig-width: 10
#| fig-height: 2.8
lab(c("before_after", "with_without", "randomised"), lab2_trend, lab2_size, lab2_hidden)
```

::: {.notes}
About 3 minutes. The sliders start with all three problems switched on. This
chart comes back in every method session with one more method added:
difference-in-differences tomorrow morning, then regression discontinuity, then
matching. Each new method will survive some sliders and not others; only the
lottery survives them all.
:::

# 5 · Reading the randomised result {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## First check: did the real draw work?

```{webr}
round(aggregate(cbind(cost_before, manager_age, staff, area) ~ took_part,
                data = pilot, FUN = mean), 1)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Same check as three slides ago, but now the two groups are the businesses the
lottery picked (1) and the ones it did not (0), in the pilot district.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Compare with the rest of the city: there, managers differed by about
`r round(age_gap)` years. Here, the groups look alike **before** the programme.
That is what a balance table is for.
:::
:::::

::::::

## The effect is a difference of two averages

::: {.ask}
Three lines of averages, in the pilot district. **Before I press Run: will the
gap be bigger or smaller than 1,000 AED?**
:::

```{webr}
#| fig-width: 6
#| fig-height: 1.8
picked     <- mean(pilot$cost_after[pilot$took_part == 1])
not_picked <- mean(pilot$cost_after[pilot$took_part == 0])
round(data.frame(picked, not_picked, effect = picked - not_picked))
bars(Picked = picked, "Not picked" = not_picked, col = c(ACCENT, ACCENT2))
```

::: {.panel .fragment}
`r gfmt(rct_est)` AED. The draw made the groups alike, so the gap between the two
averages **is the effect**, with no adjustment.
:::

## The regression gives the same number

```{webr}
fit <- lm(cost_after ~ took_part, data = pilot)
round(data.frame(coefficient = coef(fit)))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
The table an evaluator hands you comes from a line like this one.

**(Intercept)** is the average for the businesses the lottery did not pick.
**took_part** is the gap between the groups.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Which of these two numbers is the impact? And what is the other one?
:::

::: {.warn .fragment}
The coefficient is the difference you just computed by hand. The intercept is
where the comparison group sits: **not an effect**. Reading it as one is a
common mistake, by people and by AI.
:::
:::::

::::::

## Which number is the impact?

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">The impact</div>
  <div class="v">the took_part coefficient: %s AED &middot; the difference the programme made</div>
  <div class="k">Not the impact</div>
  <div class="v">the intercept, %s AED &middot; where the comparison group sat</div>
  <div class="k">Not sufficient either</div>
  <div class="v">the p-value &middot; it says the effect is unlikely to be zero, not that it is big enough</div>
</div>', gfmt(rct_est), gfmt(rct_control)))
```

::: {.ask}
A summary reads: *"highly significant, so the programme works and should be
scaled."* Which line of this table does it misuse?
:::

## Does it clear the rule?

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.5, fig.width=6}
rule_chart(rct_est, rct_avoided_least, rct_avoided_most, RULEBAR,
           "AED avoided per business", 1350)
```
:::::

::::: {.column width="48%"}
```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, the randomised estimate</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
</div>', gfmt(abs(rct_est)), gfmt(RULEBAR), gfmt(rct_avoided_least)))
```

::: {.warn}
The estimate clears the bar. The 95% interval **spans** it: at its least generous
the programme avoids `r gfmt(rct_avoided_least)` AED, short of
`r gfmt(RULEBAR)`.
:::
:::::

::::::

::: {.ask .fragment}
Would you sign a national roll-out on this? Commit, then we ask what a smaller
trial would have told us.
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker: record the interval and whether all of it clears 1,000 AED.
:::

## What if the trial had been smaller?

```{ojs}
//| echo: false
viewof kt = Inputs.range([20, 400], {value: 400, step: 10, label: "Businesses in the trial"})
```

```{webr}
#| input:
#|   - kt
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.8
small_trial(kt)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Slide down to 20. **What happens to the estimate, and to the interval?**
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The estimate moves a little, but the interval widens to about
`r gfmt(st20[1])` to `r gfmt(st20[2])` AED. At that size the trial cannot tell you
whether the bar is cleared.
:::
:::::

::::::

::: {.notes}
Same seed each time, so the businesses kept at a given size are stable while you
talk. Link back to Session 1: a study that is too small may fail to find a real
effect.
:::

# 6 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: a paragraph for the minister

:::::: {.columns}

::::: {.column width="42%"}
::: {.panel}
Groups, paste this prompt, written the way a busy official might write it:

**"Our pilot cut waste costs by `r gfmt(abs(ba_change))` AED per business (costs before vs after,
p < 0.001). Write a short paragraph for the minister on what this shows."**
:::

::: {.ask}
Mark anything the evidence does not support. Three minutes, in your groups. Then
run the same prompt yourselves: did your AI write something similar?
:::
:::::

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

The GreenWaste pilot produced a clear result: waste-management costs fell by `r gfmt(abs(ba_change))`
AED per business, and [with p < 0.001 we can be confident the programme caused
this reduction]{.flag}. [Because the same businesses were measured before and
after, differences between businesses are already accounted for]{.flag}, so
[the `r gfmt(abs(ba_change))` AED saving can be attributed to the programme]{.flag}. [This makes a
strong case for national scale-up]{.flag}. As with any pilot, results should be
monitored as the programme expands.
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session3_materials.docx): Hand out the AI handout (pack page 2, one per pair). The better prompt is printed below the flawed paragraph.
:::

## Debrief: four errors

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · Significant is not causal.** p < 0.001 says the fall is unlikely to be
chance. It says nothing about *what* caused it.
:::

::: {.warn}
**2 · Half right.** Following the same businesses
removes fixed differences between them. It does not remove what changed for
everyone: costs for businesses that did not take part rose by about `r gfmt(city_rise)` AED.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · It reported `r gfmt(abs(ba_change))` as the effect.** That is the naive figure. The randomised
estimate is `r gfmt(abs(rct_est))` AED, and the prompt gave the model no way to
know that.
:::

::: {.warn}
**4 · It recommended scale-up without asking** what bar the programme has to
clear, or what `r gfmt(abs(ba_change))` was compared to. The closing caveat is generic.
:::
:::::

::::::

::: {.panel}
**The prompt steered the answer.** It supplied a p-value, the word "cut" and
a request for the minister, so a model may write a confident summary instead of
questioning the design. People asked for a quick paragraph for a minister often
do the same.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Our pilot cut waste costs by `r gfmt(abs(ba_change))` AED per business (before vs after,
> p < 0.001). Write a short paragraph for the minister.

The prompt frames the fall as the programme's doing, and the model may follow
that framing.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Costs for participating businesses were `r gfmt(abs(ba_change))` AED lower after the pilot than
> before (p < 0.001). We also have businesses that did not take part and a
> randomised comparison. The scale-up rule is at least 1,000 AED. Before writing
> anything, tell me what this before-and-after number can and cannot show.
> **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both prompts. Which answer checks whether `r gfmt(abs(ba_change))` AED is a fair measure of the
effect before writing the paragraph? What did each answer explain well?

Two habits: make the AI **ask you questions first**, and have a **second chat**
review the first one's answer.
:::

::: {.notes}
If someone asks why we did not simply ask the AI to explain before-and-after
bias: current models answer that general question well. The short prompt is
closer to what people type when they are busy.
:::

# 7 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Your turn: ask AI first

::: {.ask style="font-size: 110%;"}
Before we show our list of questions to ask someone presenting an RCT, ask an AI tool:

**"I am a commissioner about to hear an evaluator present a randomised controlled trial (RCT) result. What are the key questions I should ask?"**
:::

::: {.panel}
Three minutes. Then compare with our list on the next slide: what did it miss, what did it add, and is anything incorrect?
:::

::: {.notes}
OPTIONAL. If time is short, skip this slide and go straight to the questions
on the next one. If you run it: give three minutes, then reveal the list and
ask what the AI missed, what it added that is worth keeping, and whether
anything it suggested was incorrect.
:::

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Before-and-after, with-and-without, or a fair comparison?</div>
  <div class="k">2</div> <div class="v">Show me the balance table. How do you know the groups started out alike?</div>
  <div class="k">3</div> <div class="v">How many units were randomised? Does the whole confidence interval meet our decision rule?</div>
</div>
```

::: {.panel}
Question 2 tells you whether the comparison is fair. Question 3 tells you
whether the result is enough to act on.
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Hand out the take-away cards (pack page 3). Give a minute to fill in "Back at work" on the card, then take two or three examples from the room.

Paper: hand out the GreenWaste case sheet (GreenWaste_case_brief.docx, one per participant) to keep for the rest of the week. It summarises the case and the four groups; hand it out now, not earlier, because it shows the averages the session asked the room to predict.
:::

## What today gives you

```{=html}
<div class="facts">
  <div class="k">Before-and-after</div> <div class="v">folds in everything else that changed &middot; here it understated</div>
  <div class="k">With-and-without</div>  <div class="v">compares groups that already differed &middot; here it overstated</div>
  <div class="k">A lottery</div>         <div class="v">makes groups alike on average, and more so the more units it draws</div>
  <div class="k">Reading the result</div><div class="v">a difference of two averages; judge its interval against the rule</div>
</div>
```

::: {.panel}
Tomorrow: the rest of the city, where no lottery decided. If the groups were not
drawn, **were their costs at least moving in parallel before the programme
started?**
:::
