---
title: "Reading Matching Results"
subtitle: "Module 2 · Day 2, Session 3: Two Evaluators, One Dataset"
author: "3ie"
always_allow_html: yes
# live-revealjs (quarto-live): the baked plotly/kable charts still render at knit time;
# the teaching code cells (checking the data, finding the twins, the overlap check,
# running the comparison, and the "never measured" check) are {webr}
# cells that run live in the browser via WebAssembly. Needs internet at presentation
# time to load webR. Export to PDF for a portable copy (live cells then show their
# code but do not run).
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    # Single-file deck. The live (webR) cells work fine with this over http,
    # via `quarto preview` or a published URL, but NOT from a file:// path.
    # Do not open the rendered .html directly from disk; serve it or publish it.
    embed-resources: true
engine: knitr
resources:
  - evaluation_data_GreenWaste_simple.csv
  - gw_lab.R
webr:
  packages:
    - dplyr
    - ggplot2
    - estimatr
    - broom
  cell-options:
    autorun: false
    edit: true
    warning: false
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(estimatr)
library(broom)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

# ---- Tables: plain knitr::kable, styled by theme_dge.scss -------------------
# No kableExtra: its JavaScript needs jQuery, which live-revealjs does not load.
# font_size in px; hl = rows to shade (hl_bold to bold them); note = a line
# under the table. The table is centred, as kable_styling(full_width = FALSE) did.
HL_BLUE <- "#E4EBF3"; HL_RED <- "#F8E9EA"
tbl <- function(df, ..., font_size = NULL, hl = NULL, hl_bg = HL_BLUE,
                hl_bold = FALSE, note = NULL) {
  h <- as.character(knitr::kable(df, format = "html", ...))
  fs <- if (is.null(font_size)) "" else sprintf(" font-size:%dpx;", font_size)
  h <- sub("<table", sprintf('<table style="margin-left:auto; margin-right:auto;%s"', fs),
           h, fixed = TRUE)
  if (length(hl)) {
    parts <- strsplit(h, "<tbody>", fixed = TRUE)[[1]]
    rows  <- strsplit(parts[2], "<tr>", fixed = TRUE)[[1]]
    st <- sprintf("background:%s;%s", hl_bg, if (hl_bold) " font-weight:700;" else "")
    for (i in hl) {
      rows[i + 1] <- gsub('<td style="', paste0('<td style="', st, " "), rows[i + 1], fixed = TRUE)
      rows[i + 1] <- gsub("<td>", paste0('<td style="', st, '">'), rows[i + 1], fixed = TRUE)
    }
    h <- paste0(parts[1], "<tbody>", paste(rows, collapse = "<tr>"))
  }
  if (!is.null(note))
    h <- paste0(h, sprintf('\n<p style="text-align:center; font-size:0.75em; color:%s; margin-top:0.3em;">%s</p>',
                           MUTED, note))
  fence <- strrep("`", 3)
  knitr::asis_output(paste0("\n", fence, "{=html}\n", h, "\n", fence, "\n"))
}

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  # Regenerate the QR when the qrcode package is available; otherwise keep the
  # menti_qr.png already in the folder.
  if (!requireNamespace("qrcode", quietly = TRUE)) return(path)
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qrcode::qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000    # GreenWaste bar, in AED

# ===========================================================================
# The simple GreenWaste case (SIMPLE_GREENWASTE_PLAN.md): the rest of the city,
# where every business scoring 58 or below took part. Matching pairs each
# business that took part with the business that did not that looks most like
# it on four characteristics measured before the programme. The efficiency
# score is not on the list: it decided who took part completely, so no
# business above 58 could be a twin on it. greenwaste_case.R is sourced into
# its own environment because it defines RULE (here a colour).
# ===========================================================================
case_env <- new.env()
case_env$gw <- read.csv("./evaluation_data_GreenWaste_simple.csv")
sys.source("greenwaste_case.R", envir = case_env)
case <- case_env$case
city <- case_env$city
took <- city$took_part == 1

VARS <- c("manager_age", "staff", "area", "filtration")
lab_vars <- c(manager_age = "Manager age", staff = "Staff", area = "Premises",
              filtration = "Advanced filtration",
              cost_before = "Waste costs before (the outcome)")

# ---- The matcher -----------------------------------------------------------
# Nearest neighbour on the standardised characteristics, with replacement: the
# same rule as greenwaste_case.R's match_on(), and the SAME function runs in
# the live (webR) cells, so the baked slides and the live cells agree. Written
# out rather than called from a package, because the session's job is to show
# what matching does: for each business that took part, take the business that
# did not with the closest characteristics, then compare.
twins_of <- function(data, vars) {
  X  <- scale(data[, vars, drop = FALSE])
  t  <- which(data$took_part == 1); co <- which(data$took_part == 0)
  Xc <- t(X[co, , drop = FALSE])
  co[vapply(t, function(i) which.min(colSums((Xc - X[i, ])^2)), 1L)]
}
std_gap <- function(v, tw) (mean(city[[v]][took]) - mean(city[[v]][tw])) / sd(city[[v]])
match_fit <- function(vars) {
  tw <- twins_of(city, vars)
  m  <- rbind(city[took, ], city[tw, ])
  r  <- lm_robust(cost_after ~ took_part, data = m, clusters = business)
  list(tw = tw, fit = r, est = coef(r)[["took_part"]],
       ci = c(r$conf.low[["took_part"]], r$conf.high[["took_part"]]),
       used = length(unique(tw)),
       sd = sapply(c(VARS, "cost_before"), std_gap, tw = tw),
       age_gap  = mean(city$manager_age[took]) - mean(city$manager_age[tw]),
       base_gap = mean(city$cost_before[took]) - mean(city$cost_before[tw]))
}

# Evaluator A matches on all four; evaluator B never measured manager age.
A <- match_fit(VARS)
B <- match_fit(setdiff(VARS, "manager_age"))
stopifnot(abs(A$est - case$matching) < 1e-6, abs(B$est - case$matching_no_age) < 1e-6)
s_age <- match_fit("manager_age")
s_sa  <- match_fit(c("staff", "area"))

# Before matching: the raw comparison and the baseline gap.
n_took  <- sum(took); n_rest <- sum(!took)
raw_t1  <- mean(city$cost_after[took]);  raw_c1 <- mean(city$cost_after[!took])
raw_t0  <- mean(city$cost_before[took]); raw_c0 <- mean(city$cost_before[!took])
raw_gap <- raw_t1 - raw_c1

est1 <- A$est; ci1 <- A$ci; fit1 <- A$fit
# Sign trap: the estimate is negative, so the end of the interval closest to
# zero is the LEAST generous one. Sort the absolute values: [1] is that end.
avoid1 <- sort(abs(ci1)); avoidB <- sort(abs(B$ci))

# Balance, before and after, as standardised differences.
before_sd <- sapply(c(VARS, "cost_before"), function(v)
  (mean(city[[v]][took]) - mean(city[[v]][!took])) / sd(city[[v]]))
bal1 <- tibble(var = c(VARS, "cost_before"),
               Characteristic = unname(lab_vars[c(VARS, "cost_before")]),
               Before = before_sd, After = A$sd)

# How hard the reuse leans on a few twins.
max_use <- max(table(A$tw))
top10   <- sum(sort(table(A$tw), decreasing = TRUE)[1:10]) / length(A$tw)

# ---- Sensitivity: change what you match on --------------------------------
sens_tbl <- tibble(
  spec = c("All four characteristics", "Manager age only",
           "Staff and premises only", "Without manager age"),
  est = c(A$est, s_age$est, s_sa$est, B$est),
  lo  = c(A$ci[1], s_age$ci[1], s_sa$ci[1], B$ci[1]),
  hi  = c(A$ci[2], s_age$ci[2], s_sa$ci[2], B$ci[2]))

# ---- The picture of what matching does (illustration on real businesses) --
# A small sample, matched on TWO characteristics only so that the pairs can be
# drawn on a flat chart. The real analysis uses four.
set.seed(2026)
pic <- bind_rows(
  city %>% filter(took_part == 1) %>% slice_sample(n = 16),
  city %>% filter(took_part == 0) %>% slice_sample(n = 55)) %>%
  mutate(zx = (manager_age - mean(city$manager_age)) / sd(city$manager_age),
         zy = (area - mean(city$area)) / sd(city$area))
pic_t <- pic %>% filter(took_part == 1)
pic_c <- pic %>% filter(took_part == 0)
pic_j <- sapply(seq_len(nrow(pic_t)), function(k)
  which.min((pic_c$zx - pic_t$zx[k])^2 + (pic_c$zy - pic_t$zy[k])^2))
pic_c$used <- seq_len(nrow(pic_c)) %in% pic_j

# ---- The two evaluators, side by side ---------------------------------------
# Honest balance cell: "yes" only when every matched characteristic is within 0.1.
bal_cell <- function(sds) {
  sds <- abs(sds)
  if (max(sds) <= 0.1) sprintf("yes (largest difference %.2f)", max(sds))
  else sprintf("partly: %d of %d over 0.1 (largest difference %.2f)", sum(sds > 0.1), length(sds), max(sds))
}
compare_tbl <- tibble(
  Question = c("Matched on",
               "Comparison businesses used as twins (some more than once)",
               "Balanced on what they matched on?",
               "Manager age after matching (not in B's table)",
               "Was the outcome similar before the programme?",
               "Estimate, costs after the programme",
               "Least generous end of the 95% interval, against the rule"),
  A = c("manager age, staff, premises, filtration",
        sprintf("%s twins for %s businesses", gfmt(A$used), gfmt(n_took)),
        bal_cell(A$sd[VARS]),
        sprintf("%.0f years apart", abs(A$age_gap)),
        sprintf("close (%s AED apart)", gfmt(abs(A$base_gap))),
        sprintf("%s AED", gfmt(A$est)),
        sprintf("%s against %s: %s", gfmt(avoid1[1]), gfmt(RULEBAR),
                if (avoid1[1] >= RULEBAR) "clears" else "below the rule")),
  B = c("staff, premises, filtration",
        sprintf("%s twins for %s businesses", gfmt(B$used), gfmt(n_took)),
        bal_cell(B$sd[setdiff(VARS, "manager_age")]),
        sprintf("%.0f years apart", abs(B$age_gap)),
        sprintf("no (%s AED apart)", gfmt(abs(B$base_gap))),
        sprintf("%s AED", gfmt(B$est)),
        sprintf("%s against %s: %s", gfmt(avoidB[1]), gfmt(RULEBAR),
                if (avoidB[1] >= RULEBAR) "clears" else "below the rule"))
)
```


# 1 · The third way to build a comparison {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

<!-- Hidden setup (webR priming, deck CSS). Kept under the first section heading, not before it: content before the first heading makes Quarto create an empty slide after the title. -->
:::: {style="display:none;"}
```{webr}
#| autorun: true
#| edit: false
#| echo: false
#| output: false
#| warning: false
#| message: false
# ---- Live (webR) setup ----------------------------------------------------
# Runs once in the browser when the deck opens, priming the shared webR session
# so every teaching cell has its data and helpers ready even if the trainer
# skips one. Separate from the knitr `setup` chunk above, which prepares the
# baked charts and tables. The matcher is the SAME function as in knitr.
library(dplyr)
library(ggplot2)
library(estimatr)
library(broom)

gw <- read.csv("evaluation_data_GreenWaste_simple.csv")
source("gw_lab.R")   # the GreenWaste lab: an invented world where the true effect is known
city <- subset(gw, setting == "city")   # the score rule: 58 or below took part

twins_of <- function(data, vars) {
  X  <- scale(data[, vars, drop = FALSE])
  t  <- which(data$took_part == 1); co <- which(data$took_part == 0)
  Xc <- t(X[co, , drop = FALSE])
  co[vapply(t, function(i) which.min(colSums((Xc - X[i, ])^2)), 1L)]
}
# Remembers each list it has matched on, so a slide that re-runs a match it
# has already done answers at once. Both lists are matched here, on opening.
find_twins <- local({
  done <- list()
  function(data, vars) {
    key <- paste(vars, collapse = " + ")
    if (is.null(done[[key]])) {
      tw <- twins_of(data, vars)
      m  <- rbind(data[data$took_part == 1, ], data[tw, ])
      attr(m, "twins") <- tw
      done[[key]] <<- m
    }
    done[[key]]
  }
})
matched  <- find_twins(city, c("manager_age", "staff", "area", "filtration"))
matched2 <- find_twins(city, c("staff", "area", "filtration"))

ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
grp_cols <- c("Took part" = ACCENT, "Did not take part" = ACCENT2)

theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(text = element_text(colour = INK),
          axis.title = element_text(colour = MUTED, size = rel(0.85)),
          axis.text = element_text(colour = MUTED),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
          legend.position = "bottom", legend.title = element_blank(),
          plot.margin = margin(4, 10, 4, 4))
}
```
::::

```{=html}
<style>
/* This deck: smaller live-code editors and output, so code plus result fits
   on a 1280x720 slide. The extension default is 1.3rem. */
.reveal .card.exercise-editor { --exercise-font-size: 1.05rem; margin: 0.3rem 0; }
.reveal .card.exercise-editor .card-header { padding: 0.25rem 0.75rem; }
.reveal .cell-output-container pre code,
.reveal .exercise-cell-output pre code { font-size: 0.85rem; line-height: 1.3; }
.reveal .cell-output-container pre { margin: 0.2rem 0; }
.reveal div.sourceCode pre code { font-size: 1.0rem; line-height: 1.35; }
.reveal .cell-output-display img,
.reveal .cell-output-display canvas { max-height: 280px; width: auto !important; }
</style>
```

## The GreenWaste case

{{< include _case_map.qmd >}}

::: {.notes}
About 1 minute: a reminder of the case. Point at the rest of the city, where the
score decided. This session works there: matching looks for businesses that did
not take part but look like the ones that did.
:::

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1 S3</div> <div class="v">randomisation &middot; the comparison chosen by a lottery</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences &middot; and an assumption the waiting list showed failing</div>
  <div class="k">Day 2 S2</div> <div class="v">regression discontinuity &middot; the comparison the rule creates</div>
  <div class="k">Today</div>     <div class="v">matching &middot; the comparison you build yourself</div>
</div>
```

::: {.panel}
Four methods, four ways of answering **compared to what?**

- Randomisation **creates** a comparison group.
- DiD uses the same groups' **change over time** (`r gfmt(case$did)` AED this morning).
- A cut-off **creates** one at the line (`r gfmt(case$rdd2[["estimate"]])` AED).
- Matching **constructs** one, by finding lookalikes.
:::

::: {.ask}
DiD said `r gfmt(case$did)`, RDD `r gfmt(case$rdd2[["estimate"]])`, both short of
1,000. Matching uses every business. Does it change the verdict?
:::

::: {.notes}
Section 1 is the recap and the idea: aim for about 15 minutes. The room already
knows GreenWaste, the 1,000 AED rule and the selection problem from Day 1
Session 3, so do not re-teach why a raw comparison misleads. The new material
is how matching builds a comparison (Section 2) and what it cannot see
(Section 3).
:::

## `[MENTI PLACEHOLDER]` Would that be fair?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
Two businesses: one took part and one did not. They have the same number of
staff, the same size of premises and the same equipment.

**Is comparing them fair?**
:::

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Only if nothing else matters

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel .fragment}
**"Only if nothing else matters" is the best answer. Most of this session
explains why.** Matching handles what you measured. Everything you did not measure is
left in the comparison, and you will not be able to see it in the table.
:::

## Matching, in one sentence

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">For each business that got the programme,<br>find one that did not but looks just like it.<br>Then compare the two.</div>
</div>')
```

::: {.panel}
There is no lottery and no cut-off. You build the comparison group yourself,
one matched pair at a time, using the characteristics you measured **before**
the programme started.
:::

::: {.ask}
Say it back to your neighbour.
:::

## Your turn: find the twins

:::::: {.columns}

::::: {.column width="50%"}
```{r}
tibble(
  ` ` = c("A", "B", "C"),
  `Manager age` = c(45, 38, 52),
  Size = c("small", "large", "small"),
  `Cost after` = c("900", "1,400", "700")
) %>%
  tbl(align = "lccr", caption = "Took part", font_size = 19)
```
:::::

::::: {.column width="50%"}
```{r}
tibble(
  ` ` = c("1", "2", "3", "4"),
  `Manager age` = c(39, 46, 51, 29),
  Size = c("large", "small", "small", "large"),
  `Cost after` = c("2,300", "1,900", "1,800", "2,600")
) %>%
  tbl(align = "lccr", caption = "Did not take part", font_size = 19)
```
:::::

::::::

::: {.ask}
In your group: pair each business that took part with its closest twin,
matching on **who they are**, not on their cost. Then take the average cost
difference across the three pairs. Two minutes, one answer per group.
:::

::: {.panel .fragment}
**A with 2, B with 1, C with 3.** Differences of -1,000, -900 and -1,100:
an average of **-1,000 AED**. Business 4 is nobody's twin, so it is set aside.
:::

::: {.notes}
Paper (Oct13_session3_materials.docx): Hand out the find-the-twins sheet (pack
page 1, one per group) and a set of profile cards per group (page 2, cut).
Decide on the day: groups use the sheet, pair the cards on the table, or both.
Walking alternative: one volunteer per group takes a card (seven groups; the
eighth checks the matches). A, B and C walk to their twin and their groups
explain the match; business 4 is left standing alone. Ask what happens to it.

The point they should feel: matching is intuitive, they already know how to
do it, and it quietly sets aside units that resemble nobody. That is
the overlap idea, which comes back in Section 2.

Circulate and listen for "but what if none of them really match?". Also listen
for anyone matching on cost: that is matching on the outcome, which would build
the answer into the comparison.
:::

## What matching does, on real businesses

```{r fig.height=3.5, fig.width=11.5}
# Three frames. Each trace is present in every frame, with the same number of
# points, so the slider moves cleanly in both directions.
steps <- c("1 · Everyone", "2 · Find each twin", "3 · Keep the pairs")
off <- -100   # a parked point, off the fixed axis range
fr_t <- map_dfr(steps, ~ pic_t %>% mutate(step = .x))
fr_c <- map_dfr(steps, function(s) pic_c %>% mutate(step = s,
  col = case_when(s == steps[3] & !used ~ "#D3D7DA", TRUE ~ ACCENT2)))
links <- map_dfr(seq_len(nrow(pic_t)), function(k) tibble(
  x = c(pic_t$manager_age[k], pic_c$manager_age[pic_j[k]], NA),
  y = c(pic_t$area[k], pic_c$area[pic_j[k]], NA)))
fr_l <- bind_rows(
  tibble(x = off, y = off, step = steps[1]),
  links %>% mutate(step = steps[2]),
  links %>% mutate(step = steps[3]))

plot_ly() %>%
  add_trace(data = fr_l, x = ~x, y = ~y, frame = ~step, type = "scatter",
            mode = "lines", line = list(color = MUTED, width = 1.5),
            hoverinfo = "skip", showlegend = FALSE) %>%
  add_trace(data = fr_c, x = ~manager_age, y = ~area, frame = ~step,
            type = "scatter", mode = "markers",
            marker = list(color = ~col, size = 11, symbol = "circle",
                          line = list(color = "white", width = 1)),
            hovertemplate = "Did not take part<br>manager %{x}, premises %{y}<extra></extra>",
            showlegend = FALSE) %>%
  add_trace(data = fr_t, x = ~manager_age, y = ~area, frame = ~step,
            type = "scatter", mode = "markers",
            marker = list(color = ACCENT, size = 14, symbol = "triangle-up",
                          line = list(color = "white", width = 1)),
            hovertemplate = "Took part<br>manager %{x}, premises %{y}<extra></extra>",
            showlegend = FALSE) %>%
  animation_opts(frame = 1200, transition = 0, redraw = FALSE) %>%
  animation_slider(currentvalue = list(visible = FALSE),
                   font = list(size = 12, color = INK), x = 0.08, len = 0.84) %>%
  layout(xaxis = ax("Manager age", range = c(12, 82), zeroline = FALSE),
         yaxis = ax("Premises (100 m2)", range = c(0, 6.5), zeroline = FALSE)) %>%
  pl_m2(height = 338, t = 34, b = 90,
        title = "&#9650; took part &nbsp;&nbsp; &#9679; did not take part &nbsp;&nbsp; (71 GreenWaste businesses, matched on two characteristics)")
```

::: {.ask}
Drag the slider. Where do the triangles sit, and which circles are never
anyone's twin?
:::

::: {.notes}
GreenWaste businesses from the rest of the city, a random 16 that took part
and 55 that did not, matched here on just two characteristics so the pairs can
be drawn on a flat chart. The real analysis uses four, and does the same thing
in four dimensions.

Frame 1: the triangles sit mostly on the left (younger managers) and lower
(smaller premises); the circles spread further right and up. The two groups
are different crowds. Frame 2: each triangle is joined to its nearest circle.
Some circles are the twin of more than one triangle: that is allowed. Frame 3:
circles that nobody picked fade out. They are set aside, and the comparison is
now only between the pairs.
:::

## The GreenWaste lab: add matching

:::::: {.columns}

::::: {.column width="46%"}
```{ojs}
//| echo: false
viewof mat_size_gap = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners are smaller businesses (AED)"})
viewof mat_hidden = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners differ in something nobody measured (AED)"})
```
:::::

::::: {.column width="54%"}
::: {.ask}
Matching joins the lab. Move the measured slider, then the unmeasured one.
**What does matching fix, and what can it not see?**
:::

::: {.panel .fragment}
It removes the measured difference. The unmeasured one moves it as far as
with-and-without, while its balance table still looks perfect.
:::
:::::

::::::

```{webr}
#| input:
#|   - mat_size_gap
#|   - mat_hidden
#| autorun: true
#| edit: false
#| echo: false
#| fig-width: 10
#| fig-height: 3.5
lab(c("before_after", "with_without", "randomised", "did", "rdd", "matching"),
    size_gap = mat_size_gap, hidden = mat_hidden)
```

::: {.notes}
About 5 minutes. The first slider (joiners smaller, which we measured) breaks
with-and-without; matching stays on -900 because it pairs on size. The second
(joiners differ in something nobody measured) moves matching by the same amount
as with-and-without. DiD, RDD and the randomised comparison survive it, because
none of them relies on measuring the difference. Section 3 returns to this as
"the one question".
:::

# 2 · Implementing matching in R {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 25 minutes. Four live cells; keep syntax talk short.
:::

## Remember: what matching needs from the data

::: {.panel}
1. **An outcome**, measured the same way for everyone
2. **A group that got the programme and a pool that did not**, and a variable
   saying which is which
3. **Characteristics measured before the programme**, never after it
4. **The characteristics that drove who took part**, and that also affect the
   outcome
5. **Enough untreated units that look like the treated ones**, so every
   treated unit has a real twin
:::

::: {.ask}
Which of the five would you check first in your own data?
:::

Take home: ask AI *what data do I need for matching, and does my dataset have
it?*, and have it question you about who took part before it answers.

::: {.notes}
Items 3 and 4 are the ones that matter most. Item 3: a characteristic measured
after the programme may itself have been changed by the programme, so matching
on it builds part of the effect into the comparison. Item 4 is the one Section 3
turns on: matching can only use what is on the list. Item 5 is the overlap
check two slides on.
:::

## Does GreenWaste have what we need?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
city %>%
  group_by(took_part) %>%
  summarise(businesses  = n(),
            cost_before = round(mean(cost_before)),
            cost_after  = round(mean(cost_after)),
            manager_age = round(mean(manager_age), 1),
            staff       = round(mean(staff), 1))
```

:::::

::::: {.column width="44%"}
::: {.panel}
✔ **Outcome, group:** `cost_after`; `took_part` 1 or 0

✔ **Before the programme:** manager age, staff, premises, filtration

✔ **A big pool:** `r gfmt(n_rest)` possible twins for `r gfmt(n_took)`
businesses that took part
:::

::: {.warn .fragment}
The businesses that took part were already cheaper **before** the programme.
The raw gap after, **`r gfmt(raw_gap)` AED**, mixes the programme with who they
already were.
:::
:::::

::::::

::: {.notes}
Press Run. `city` has one row per business: its characteristics before the
programme and its costs before and after.

Read the two rows. Took part: `r gfmt(raw_t0)` AED before and `r gfmt(raw_t1)`
after. Did not: `r gfmt(raw_c0)` before and `r gfmt(raw_c1)` after. Do not
dwell on the raw gap: the room met this problem on Day 1 Session 3. The point
is only that the two groups started in different places, so there is work for
matching to do.

If asked why the efficiency score is not on the list: in the rest of the city
it decided who took part completely. Every business at 58 or below took part,
so no business that did not take part could be a twin on it.
:::

## Implementing matching: four steps

:::::: {.columns}

::::: {.column width="56%"}
```{r}
#| echo: true
#| eval: false
# Step 1. Put each characteristic on one scale
#         (distance from the average, in
#         standard deviations).
X <- scale(city[, c("manager_age", "staff",
                    "area", "filtration")])

# Step 2. Each business that took part takes the
#         business that did not with the
#         smallest distance on X.
# Step 3. Check the pairs are alike (balance).
# Step 4. Compare their costs after the programme.
lm_robust(cost_after ~ took_part, data = matched,
          clusters = business)
```
:::::

::::: {.column width="44%"}
::: {.panel}
- **Step 1** makes a year of manager age and a member of staff comparable
- **Step 2** is your twins exercise, done `r gfmt(n_took)` times
- **Step 3** is the only evidence the pairs are alike
- **Step 4** is the comparison you wanted all along
:::

::: {.warn}
The **list in Step 1** is the analyst's choice. Section 3 tests it.
:::
:::::

::::::

::: {.notes}
Do not dwell on syntax. The one idea that matters: matching is preparation. It
builds the comparison group first, and only then do we compare. The regression
in Step 4 is deliberately plain: took part against not, on the matched pairs.

If asked about a "propensity score": many evaluations squeeze the
characteristics into one number, the likelihood of taking part, and match on
that. The logic is the same: pair each business with its closest lookalike.
:::

## Steps 1 and 2, live

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
# Steps 1 and 2. Each business that took part
# takes the closest business that did not,
# on these four characteristics
vars <- c("manager_age", "staff", "area",
          "filtration")
matched <- find_twins(city, vars)

c(took_part  = sum(city$took_part),
  twins_used = length(unique(attr(matched, "twins"))))
```
:::::

::::: {.column width="42%"}
::: {.ask}
Guess first: `r gfmt(n_took)` different twins, or far fewer? Then press
**Run**.
:::

::: {.panel .fragment}
**`r gfmt(A$used)` twins for `r gfmt(n_took)` businesses.** Some businesses
that did not take part are the closest match for several that did, and are
used more than once.
:::

::: {.warn .fragment}
The other `r gfmt(n_rest - A$used)` are set aside. Business 4,
`r gfmt(n_rest - A$used)` times over.
:::
:::::

::::::

::: {.notes}
`find_twins()` does Steps 1 and 2: it scales the four characteristics and, for
each business that took part, finds the closest business that did not. It ran
once when the deck opened, so Run answers at once.

Reusing a twin is allowed (analysts call it matching "with replacement"). It
keeps every pair as close as possible, at the cost of leaning harder on the
businesses that get reused. Worth knowing the phrase if an evaluator uses it.
:::

## Did every business find a real lookalike?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 2.8
city %>%
  mutate(group = ifelse(took_part == 1,
                        "Took part", "Did not take part")) %>%
  ggplot(aes(manager_age, fill = group)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = grp_cols) +
  labs(x = "Manager age", y = "Businesses") +
  theme_m2()
```
:::::

::::: {.column width="44%"}
::: {.panel}
A twin can only come from where the curves **overlap**. Young managers are
common among businesses that took part and rare among the rest.
:::

::: {.panel .fragment}
One business that did not take part is the twin for
**`r gfmt(max_use)`** that did. The ten most-used twins stand in for
`r round(100 * top10)`% of all pairs.
:::

::: {.warn .fragment}
Where there is little overlap, the result leans on a handful of businesses.
Ask how many twins were used, and how often.
:::
:::::

::::::

::: {.notes}
Press Run. This is the picture that decides whether matching is even possible.
The two curves overlap in the middle, but the youngest managers have few
counterparts among the businesses that did not take part, so the same few twins
are used again and again.

Ask: if a business had no counterpart and we dropped it, who would the result
still be about? Only the ones that had a match.
:::

## Did matching work? Balance

```{r fig.height=3.5, fig.width=11.5}
bal_long <- bal1 %>%
  mutate(Characteristic = fct_reorder(Characteristic, abs(Before))) %>%
  pivot_longer(c(Before, After), names_to = "stage", values_to = "smd") %>%
  mutate(stage = factor(stage, levels = c("Before", "After")))

plot_ly() %>%
  add_trace(data = filter(bal_long, stage == "Before"),
            x = ~abs(smd), y = ~Characteristic, type = "scatter", mode = "markers",
            name = "Before matching",
            marker = list(color = ACCENT2, size = 12),
            hovertemplate = "%{y}: %{x:.2f} before<extra></extra>") %>%
  add_trace(data = filter(bal_long, stage == "After"),
            x = ~abs(smd), y = ~Characteristic, type = "scatter", mode = "markers",
            name = "After matching",
            marker = list(color = ACCENT, size = 12, symbol = "diamond"),
            hovertemplate = "%{y}: %{x:.2f} after<extra></extra>") %>%
  layout(xaxis = ax("How far apart the groups are (standardised difference)",
                    range = c(0, 1.8), zeroline = FALSE),
         yaxis = ax("", tickfont = list(size = 12, color = INK)),
         shapes = list(list(type = "line", x0 = 0.1, x1 = 0.1, y0 = 0, y1 = 1,
                            yref = "paper", line = list(color = ALERT, dash = "dash", width = 1.5))),
         annotations = list(list(x = 0.1, y = 1.02, yref = "paper", xanchor = "left",
                                 text = " 0.1: the usual line", showarrow = FALSE,
                                 font = list(size = 11, color = ALERT)))) %>%
  pl_m2(height = 350, t = 36, b = 80, legend_y = -0.42)
```

::: {.ask}
Each row is one characteristic. Hover to read the values. Which ones are still
apart after matching?
:::

::: {.notes}
Light blue dots: how far apart the two groups were before matching. Dark blue
diamonds: how far apart the matched pairs are. Beyond about 0.1 is usually
treated as a problem.

Manager age was the furthest apart (`r sprintf("%.2f", abs(bal1$Before[bal1$var == "manager_age"]))`).
After matching, staff, premises and filtration are close; manager age is still
`r sprintf("%.2f", abs(bal1$After[bal1$var == "manager_age"]))` apart, about
`r sprintf("%.0f", abs(A$age_gap))` years, and costs before the programme
`r sprintf("%.2f", abs(bal1$After[bal1$var == "cost_before"]))`. Few young
managers have a lookalike, which is the overlap problem from the last slide.

Say it plainly: balance means alike **on the list**. It says nothing about
anything that is not on the list. That is Section 3.
:::

## Step 4: running the comparison

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
out <- lm_robust(cost_after ~ took_part, data = matched,
                 clusters = business)

res <- tidy(out)[, c("term", "estimate", "std.error",
                     "conf.low", "conf.high")]
res[, -1] <- round(res[, -1])
res
```
:::::

::::: {.column width="42%"}
::: {.ask}
Two rows will appear. Guess first: which one is the programme's effect? Then
press **Run**.
:::

::: {.panel .fragment}
**Row 2, `took_part`: `r gfmt(est1)` AED.** The cost difference between each
business that took part and its twin, averaged.
:::

::: {.warn .fragment}
**Trap:** the intercept, `r gfmt(coef(fit1)[["(Intercept)"]])` AED, is the
twins' average cost, not an effect.
:::
:::::

::::::

::: {.notes}
Take a vote before pressing Run. Most will get it; the point is the habit of
naming the row.

Reminder of the vocabulary (Day 1 Session 1): the **coefficient** is the effect
size in AED; the **confidence interval** (conf.low to conf.high) is the range of
plausible true values. Clustering by business means a twin used many times
counts as one business, not as many independent ones.

Close the loop: this is the twins card, done for every business at once, with
a margin of error attached.
:::

## Does it clear the rule?

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the matched estimate (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">Before matching</span></div>
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
</div>', gfmt(est1), gfmt(ci1[1]), gfmt(ci1[2]),
   gfmt(raw_gap), gfmt(RULEBAR), gfmt(avoid1[1])))
```

::: {.panel}
Matching took `r gfmt(abs(raw_gap) - abs(est1))` AED off the raw gap: that
part was **who the businesses already were**.
:::

::: {.ask}
Does the whole interval clear `r gfmt(RULEBAR)` AED?
:::

::: {.panel .fragment}
**No.** The estimate sits right on the bar, and the interval runs from
`r gfmt(avoid1[1])` to `r gfmt(avoid1[2])` AED. A result this close to the line
turns on the analyst's choices, which we test next.
:::

::: {.notes}
Same reading as the last two sessions: does my decision change across this
range? It does: the optimistic end clears the bar, the pessimistic end does
not.

Worth noting if asked: the lottery in the pilot district gave
`r gfmt(case$rct)`. Matching has landed close to it, which is reassuring about
the method here. But the commissioner deciding from a matching study alone
would not know that.
:::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: what matching cannot see, and how to
check how much that matters.
:::

# 3 · The one question, and how to check it {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 20 minutes.
:::

## What matching can and cannot see

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">Matching can only account for what you measured.</div>
</div>')
```

::: {.panel}
Every characteristic in that balance chart was **collected**. Matching cannot
balance on management quality, on how motivated the owner was, on whether the
business was about to invest anyway, unless somebody wrote it down.
:::

::: {.warn}
So the accurate claim for this method is: **we have made the two groups alike in
every way we can observe.**
:::

::: {.notes}
DiD rested on parallel trends; RDD on a clean cut-off. Matching rests on
something plainer and harder: everything that drove who took part and also
affects costs was measured and matched on. This cannot be tested from the data,
by definition. But we can see how much it would matter, which is the next slide.
:::

## Check 1: what if manager age had never been measured?

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
# Re-run Steps 1 and 2, pretending manager
# age was never collected
matched2 <- find_twins(city, c("staff", "area",
                               "filtration"))
matched2 %>%
  group_by(took_part) %>%
  summarise(cost_after  = round(mean(cost_after)),
            staff       = round(mean(staff), 1),
            area        = round(mean(area), 1),
            manager_age = round(mean(manager_age), 1))
```
:::::

::::: {.column width="42%"}
::: {.ask}
Press **Run**. Manager age is no longer on the list. Before you look: will the
pairs still match on it?
:::

::: {.panel .fragment}
The pairs' managers are now **`r sprintf("%.0f", abs(B$age_gap))` years
apart**, and the estimate moves to **`r gfmt(B$est)` AED**, across the bar.
:::

::: {.warn .fragment}
The balance check on the three would still pass (worst
`r sprintf("%.2f", max(abs(B$sd[c("staff", "area", "filtration")])))`).
**Nothing in the output would warn you.**
:::
:::::

::::::

::: {.notes}
This is the session's key demonstration. Manager age goes with the efficiency
score that decided who took part, and it also drives costs, so it is the kind
of characteristic that matters. Here we have it, so we can see what leaving it
out does. In a real evaluation, the missing characteristic is one nobody
collected, so you never get to see this comparison.

The estimate moves from `r gfmt(est1)` to `r gfmt(B$est)`, far enough to clear
the bar. The direction flatters the programme.
:::

## Check 2: change the list, see what moves

```{r fig.height=3.1, fig.width=11.5}
sens_plot <- sens_tbl %>% mutate(spec = factor(spec, levels = rev(spec)))
plot_ly(sens_plot, x = ~est, y = ~spec, type = "scatter", mode = "markers+text",
        error_x = list(type = "data", symmetric = FALSE,
                       array = ~hi - est, arrayminus = ~est - lo,
                       color = ACCENT, thickness = 2.5, width = 0),
        marker = list(color = ACCENT, size = 13),
        text = ~gfmt(est), textposition = "bottom center",
        textfont = list(size = 13, color = INK),
        hovertemplate = "%{y}: %{x:,.0f} AED<extra></extra>", showlegend = FALSE) %>%
  layout(xaxis = ax("Estimated effect on waste costs (AED)", range = c(-1500, -400),
                    zeroline = FALSE),
         yaxis = ax("", tickfont = list(size = 13, color = INK)),
         shapes = list(list(type = "line", x0 = -RULEBAR, x1 = -RULEBAR, y0 = 0, y1 = 1,
                            yref = "paper", line = list(color = ALERT, dash = "dash", width = 2))),
         annotations = list(list(x = -RULEBAR, y = 1.0, yref = "paper",
                                 yanchor = "bottom", yshift = 14,
                                 text = "the bar: -1,000", showarrow = FALSE,
                                 font = list(size = 12, color = ALERT)))) %>%
  pl_m2(height = 320, t = 46, b = 60)
```

::: {.warn}
Same businesses, same method, four choices of list. The estimates run from
about `r gfmt(min(abs(sens_tbl$est)))` to `r gfmt(max(abs(sens_tbl$est)))` AED,
on **both sides of the bar**.
:::

::: {.ask .fragment}
Would you sign off a national scale-up on any single one of these rows?
:::

::: {.notes}
Row 1 is the main estimate. Row 2 matches on manager age alone: the interval is
wide because only `r s_age$used` businesses end up as twins. Rows 3 and 4 leave
manager age out, and both clear the bar.

The takeaway for a commissioner: with matching, ask what was matched on, and
whether the answer holds when the list changes. If one choice of list flips the
recommendation, the result is not settled. Here, it flips.
:::

## The question to put to any evaluator

:::::: {.columns}

::::: {.column width="52%"}
::: {.ask style="font-size: 118%;"}
For each business that took part, you found one that did not but looked the
same.

**What else might be different about those two businesses that you could not
measure?**
:::

It cannot be answered from the table.
:::::

::::: {.column width="48%"}
::: {.panel}
For GreenWaste, the plausible answers are things like:

**How badly did this business want to cut costs?** Businesses that take part
are often the ones already looking for a solution.

**Was the owner about to invest anyway?** If so, we are crediting the programme
with a decision already made.
:::

::: {.warn}
Neither of these is in the data. Both would make the programme look better than
it was, just as leaving out manager age did.
:::
:::::

::::::

::: {.notes}
Get the room to generate the hidden differences before revealing the panel.
Owner motivation always comes up, and it is a good example: it is unmeasured
and it flatters the programme.
:::

# 4 · Two evaluators, one dataset {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 20 minutes.
:::

## Two evaluators, the same businesses

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Evaluator A** matched on manager age, staff, premises and filtration.

**Evaluator B** used a business survey that never asked the manager's age, and
matched on the other three.
:::

::: {.ask}
Each sends you a balance table of what they matched on, and an estimate. Which
check do you expect to tell them apart?
:::
:::::

::::: {.column width="50%"}
```{r}
tibble(
  Characteristic = c("Manager age", "Staff", "Premises", "Advanced filtration"),
  A = sprintf("%.2f", abs(A$sd[VARS])),
  B = c("not measured", sprintf("%.2f", abs(B$sd[c("staff", "area", "filtration")])))
) %>%
  tbl(align = "lrr", col.names = c("After matching", "Evaluator A", "Evaluator B"),
      font_size = 17, note = "Standardised differences. Beyond about 0.1 is usually a problem.")
```
:::::

::::::

::: {.notes}
Both evaluators did everything correctly with what they had. On their own
tables, B looks better: every row is close to zero. A has one row over the line.
:::

## What B's table did not show

:::::: {.columns}

::::: {.column width="50%"}
```{r}
tibble(
  ` ` = c("Manager age", "Waste costs before"),
  A = c(sprintf("%.0f years", abs(A$age_gap)), sprintf("%s AED", gfmt(abs(A$base_gap)))),
  B = c(sprintf("%.0f years", abs(B$age_gap)), sprintf("%s AED", gfmt(abs(B$base_gap))))
) %>%
  tbl(align = "lrr", col.names = c("How far apart the pairs are", "A", "B"),
      font_size = 18, hl = 1:2, hl_bg = HL_RED)
```
:::::

::::: {.column width="50%"}
::: {.panel}
B's pairs have managers `r sprintf("%.0f", abs(B$age_gap))` years apart, and
their costs were already `r gfmt(abs(B$base_gap))` AED apart **before** the
programme.
:::

::: {.warn .fragment}
B reports `r gfmt(B$est)` AED, and its whole interval clears the rule. A reports
`r gfmt(A$est)`. The lottery in the pilot district found `r gfmt(case$rct)`.
:::
:::::

::::::

::: {.notes}
The outcome's own starting point, costs before the programme, is the check B
could have run without measuring manager age. Ask to see it in any balance
table: it is often left out.
:::

## The two evaluators, side by side

```{r}
compare_tbl %>%
  tbl(align = "l", col.names = c("", "Evaluator A", "Evaluator B"),
      font_size = 15, hl = c(4, 5), hl_bg = HL_RED)
```

::: {.ask}
Read down the table. Which rows separate the two studies, and would you see
them in a typical results table?
:::

::: {.notes}
B passes the checks people usually report and fails the two they usually do
not. Rows 4 and 5 separate the studies: a characteristic that was not on the
list, and the outcome's own baseline. Neither appears in a typical matching
results table, which is why they are on the question card at the end.
:::

# 5 · Group exercise: which would you trust? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 15 minutes.
:::

## Which evaluator would you trust?

::: {.ask}
In your groups, ten minutes:

1. Which evaluator gives you **more confidence**, and why?
2. What would you ask **evaluator B** that A does not raise?
3. B's whole interval clears the rule; A's straddles it. Does that change your
   answer?
:::

::: {.panel .fragment}
Our read: **A** is the more trustworthy study, and it cannot show it clears the
rule. **B** gives the more attractive answer because it left out something that
drove who took part.
:::

::: {.notes}
Paper (Oct13_session3_materials.docx): Hand out the "Which evaluator would you
trust?" sheet (pack page 3, one per group). Groups circle a verdict for each
study before the reveal.

Verdict round at the end: hands, not discussion. Push them past "A is cleaner"
to why: the list, the outcome's baseline, and how few twins carry the young
managers. Ask whoever says "send it back" for the single question behind it.
:::

## What to ask evaluator B

```{=html}
<div class="facts">
  <div class="k">Ask</div> <div class="v">"What decided who took part, and is all of it on your list?"</div>
  <div class="k">Ask</div> <div class="v">"Were the pairs' costs similar before the programme?"</div>
</div>
```

::: {.notes}
These are the two questions A's study does not raise as sharply. The full list
of five is on the take-away card at the end.
:::

# 6 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 10 minutes. The snapshot and its debrief only; the better prompt is a take-home.
:::

## AI Snapshot: comparing two studies

::: {.panel}
**Prompt, with the side-by-side table and both balance tables pasted in:**
"Summarise the key differences between these two studies and identify what a
commissioner should flag."
:::

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**Evaluator A:** Manager age remains somewhat imbalanced. The programme reduced
costs by around
`r gfmt(round(abs(A$est), -1))` AED, with an interval that straddles the
1,000 AED threshold.

**Evaluator B:** [Matching was even stronger, with near-perfect balance on every
characteristic]{.flag}. [The matched estimate shows a cut of
`r gfmt(abs(B$est))` AED, comfortably exceeding the 1,000 AED threshold]{.flag},
**[so the programme should be scaled]{.flag}**.
:::

::: {.ask}
In groups (four minutes): underline each sentence you agree is correct. Cross
out each sentence that is wrong or claims too much.
:::

::: {.notes}
Paper (Oct13_session3_materials.docx): Hand out the AI Snapshot sheet (pack
page 4, one per group). Groups underline what they would sign and cross out
the rest. Keep the better-prompt strip (page 5) back until after the debrief.
:::

## Debrief: one clean read, three false assurances

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it got right.** A is reported accurately, including the awkward part:
the interval straddles the rule.
:::

::: {.warn}
**1 · "Near-perfect balance" for B.** Only on the three things B matched on.
The pairs' managers are `r sprintf("%.0f", abs(B$age_gap))` years apart.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**2 · "Comfortably exceeding."** B's pairs were already
`r gfmt(abs(B$base_gap))` AED apart before the programme, so part of the cut is
who they already were.
:::

::: {.warn}
**3 · "So the programme should be scaled."** The recommendation rests entirely
on sentences 1 and 2, and it is stated most confidently of all.
:::
:::::

::::::

::: {.panel}
The model did not invent a number. It had the details in front of it and
still went with the headline. Check its summary against the table you gave it.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Summarise the key differences between these two studies and identify what a
> commissioner should flag.

The model may summarise the tables it was given and repeat their framing.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Two matching studies of the same businesses. For each: what was **left off
> the list**; were the pairs' costs **similar before** the programme; how far
> apart are characteristics that were not matched on? Our decision rule is
> below. **Ask me questions before you answer.**
:::
:::::

::::::

**Take home:** run both after the session. What did each get wrong, and what
did each explain well?

::: {.notes}
Paper (Oct13_session3_materials.docx): Hand out the better-prompt strip now
(pack page 5, one per group, cut) as a take-home. It names the checks, so it
goes out only after the debrief. Do not run the prompts in the session.
:::

# 7 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 10 minutes. Skip the optional AI slide if short of time.
:::

## Your turn: ask AI first

::: {.ask style="font-size: 110%;"}
Before we show our list of questions to ask someone presenting matching results, ask an AI tool:

**"I am a commissioner about to hear an evaluator present a matching result. What are the key questions I should ask?"**
:::

::: {.panel}
Three minutes. Then compare with our list on the next slide: what did it miss, what did it add, and is anything incorrect?
:::

::: {.notes}
Optional: skip if short of time, and go straight to the questions on the next
one. If you run it: give three minutes, then reveal the list and
ask what the AI missed, what it added that is worth keeping, and whether
anything it suggested was incorrect.
:::

## Five questions to ask any evaluator presenting matching

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">What did you match on, and what was left out?</div>
  <div class="k">2</div> <div class="v">How many treated units had no close match, and how often was each twin reused?</div>
  <div class="k">3</div> <div class="v">Are the matched groups similar on their characteristics? Was the outcome itself similar before the programme?</div>
  <div class="k">4</div> <div class="v">Does the answer hold if you change what you matched on?</div>
  <div class="k">5</div> <div class="v">What else might differ that you could not measure?</div>
</div>
```

::: {.panel}
Question 5 is the one that matters most, and it is the only one the evaluator
cannot answer from their own output.

So a matching study rarely settles a decision on its own.
:::

::: {.notes}
Paper (Oct13_session3_materials.docx): Hand out the take-away cards (pack
page 6, one per participant). Give a minute to fill in "Back at work" on the
card, then take two or three examples from the room.
:::

## The week so far

```{r results='asis'}
cat(sprintf('
<div class="facts">
  <div class="k">Day 1 S2</div> <div class="v">before and after &middot; %s, with no comparison group</div>
  <div class="k">Day 1 S3</div> <div class="v">the lottery &middot; %s, interval spans 1,000</div>
  <div class="k">Day 2 S1</div> <div class="v">DiD &middot; %s, and an assumption the waiting list showed failing</div>
  <div class="k">Day 2 S2</div> <div class="v">RDD &middot; %s, and local to the line</div>
  <div class="k">Today</div>  <div class="v">matching &middot; %s, and only as good as the list</div>
</div>', gfmt(case$before_after), gfmt(case$rct), gfmt(case$did),
   gfmt(case$rdd2[["estimate"]]), gfmt(est1)))
```

::: {.ask}
Each method rests on one assumption:

**DiD** assumes parallel trends. **RDD** assumes nothing else jumps at the
cut-off, and answers only near it. **Matching** assumes everything that
matters was measured.

Which of the three would you trust least, and why?
:::

::: {.notes}
If there is time at the very end, ask the room: is there a different
question you would ask AI to help you understand a concept from today? For
example: "I am learning about impact evaluation to help me make policy
decisions. How should I think about a confidence interval, and how is it
relevant when I compare an estimate to a decision threshold?" Adding context
about who you are and why you are asking usually gets a more useful answer.
:::
