---
title: "Matching activity: profiles and separate cost slips"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    include-in-header:
      text: |
        <style>
        body{font:18px/1.4 Arial,sans-serif;color:#111418;background:#eef2f6;margin:0}main{max-width:850px;margin:auto}.sheet-page{background:white;padding:30px;margin:20px 0}h2,h3{color:#063360}h2{font-size:24px}.cards{display:grid;grid-template-columns:1fr 1fr;gap:14px}.card{border:1px dashed #7da1c4;padding:16px;min-height:145px;break-inside:avoid}.card h3{font-size:25px;margin:0 0 8px}.card p{margin:4px 0}header button{font:inherit;padding:8px 15px}#title-block-header{display:none}@page{size:A4;margin:16mm}@media print{body{background:white;font-size:12pt}.sheet-page{padding:0;margin:0;break-after:page}.sheet-page:last-child{break-after:auto}header{display:none}.card{min-height:36mm;padding:4mm}.cards{gap:4mm}}
        </style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('day2_session3_review.R', encoding='UTF-8')
```

<header><button onclick="window.print()">Print both pages</button><p>Trainer: cut along dashed borders. Distribute page 1 only; retain page 2 until pairs are recorded.</p></header>

<div class="sheet-page">

## Page 1 | Seven invented profile cards

Pair A, B and C with comparisons 1-4 using **manager age and size**. Use each comparison once. Record choices before receiving costs. This differs from full-city matching with replacement.

```{r, results='asis'}
html_out(paste0('<div class="cards">', paste(vapply(seq_len(nrow(profiles)), function(i)
  paste0('<div class="card"><h3>Business ', profiles$id[i], '</h3><p>',
    profiles$group[i], '</p><p>Manager age: ', profiles$age[i],
    '</p><p>Size: ', profiles$size[i], '</p></div>'), ''), collapse=''), '</div>'))
```

</div>

<div class="sheet-page">

## Page 2 | Cost slips: trainer holds until reveal

**Invented annual after-costs**, AED per business. Keep original pair choices. Compute participant minus comparison, then average the three differences.

```{r, results='asis'}
html_out(paste0('<div class="cards">', paste(vapply(seq_len(nrow(profiles)), function(i)
  paste0('<div class="card"><h3>Business ', profiles$id[i],
    '</h3><p>Annual after-cost</p><p><strong>', fmt(profiles$cost[i]),
    ' AED</strong></p><p>Invented activity value</p></div>'), ''), collapse=''), '</div>'))
```

</div>
