---
title: "Reading Matching Results"
subtitle: "Module 2 · Day 2, Session 3: Two Case Studies, One Clean and One Not"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000     # GreenWaste bar, in AED
CAM_RULE <- 2.0     # camera bar, in collisions per segment per round

# ---- The matcher -----------------------------------------------------------
# Nearest neighbour on the propensity score, with replacement. Written out in
# full rather than called from a package, because the session's job is to show
# what matching actually does: for each treated unit, find the untreated unit
# with the closest score, then compare. Nothing here is hidden behind a
# function name.
nn_match <- function(d, treat, ps, caliper = Inf) {
  ti <- which(d[[treat]] == 1); ca <- which(d[[treat]] == 0)
  mo <- integer(length(ti)); dd <- numeric(length(ti))
  for (k in seq_along(ti)) {
    dist <- abs(d[[ps]][ti[k]] - d[[ps]][ca])
    j <- which.min(dist); mo[k] <- ca[j]; dd[k] <- dist[j]
  }
  keep <- dd <= caliper
  list(ti = ti[keep], ci = mo[keep], dist = dd[keep], dropped = sum(!keep))
}

# ===========================================================================
# CASE 1: GreenWaste. Among eligible businesses, who enrolled and who did not.
# Matching should work cleanly here: the two groups overlap almost perfectly
# and every enrolled business finds a close comparison.
# ===========================================================================
gw  <- read.csv("./evaluation_data_GreenWaste.csv")
gw0 <- gw %>% filter(round == 0, eligible == 1)
gw1 <- gw %>% filter(round == 1)

cov1  <- c("age_manager", "educ_manager", "staff_size", "foreign_owned",
           "recycling_center_distance", "business_area")
lab1  <- c(age_manager = "Manager age (years)",
           educ_manager = "Manager education (level)",
           staff_size = "Staff size",
           foreign_owned = "Foreign owned (share)",
           recycling_center_distance = "Distance to recycling centre (m)",
           business_area = "Business area (units)")

d1 <- gw0 %>% select(business_identifier, enrolled, all_of(cov1))
d1$ps <- fitted(glm(as.formula(paste("enrolled ~", paste(cov1, collapse = " + "))),
                    data = d1, family = binomial))

m1 <- nn_match(d1, "enrolled", "ps")

y1t <- gw1$waste_management_costs[match(d1$business_identifier[m1$ti], gw1$business_identifier)]
y1c <- gw1$waste_management_costs[match(d1$business_identifier[m1$ci], gw1$business_identifier)]
c1_naive   <- mean(y1t) - mean(y1c)
c1_matched <- mean(y1t - y1c)
c1_se      <- sd(y1t - y1c) / sqrt(length(y1t))
c1_ci      <- c1_matched + c(-1.96, 1.96) * c1_se
c1_pairs   <- length(m1$ti)
c1_pct     <- 100 * c1_pairs / sum(d1$enrolled)

# Balance, before and after, for the same six characteristics.
c1_bal <- map_dfr(cov1, function(v) {
  tibble(Characteristic = lab1[[v]],
         Before  = (mean(d1[[v]][d1$enrolled == 1]) - mean(d1[[v]][d1$enrolled == 0])) / sd(d1[[v]]),
         After   = (mean(d1[[v]][m1$ti]) - mean(d1[[v]][m1$ci])) / sd(d1[[v]]))
})

# Propensity overlap for case 1.
c1_ps_t <- range(d1$ps[d1$enrolled == 1])
c1_ps_c <- range(d1$ps[d1$enrolled == 0])

# ===========================================================================
# CASE 2: traffic cameras. Cameras went where the roads were worst, not at
# random. Here matching is harder, and the interesting failure is that the two
# groups match on the covariates while their OUTCOMES still start apart.
# ===========================================================================
tc  <- read.csv("./evaluation_data_TrafficCameras.csv")
tc0 <- tc %>% filter(round == 0)
tc1 <- tc %>% filter(round == 1)

cov2 <- c("baseline_speed_85th", "lanes", "road_length_km",
          "school_within_500m", "prior_collisions_3yr")
lab2 <- c(baseline_speed_85th = "85th percentile speed (km/h)",
          lanes = "Lanes",
          road_length_km = "Segment length (km)",
          school_within_500m = "School within 500 m",
          prior_collisions_3yr = "Collisions, prior 3 years")

d2 <- tc0 %>%
  mutate(lighting_poor = as.integer(lighting_quality == "poor")) %>%
  select(segment_id, cameras_installed, y0 = injury_collisions,
         all_of(cov2), lighting_poor) %>%
  left_join(tc1 %>% select(segment_id, y1 = injury_collisions), by = "segment_id")
d2$ps <- fitted(glm(cameras_installed ~ .,
                    data = d2 %>% select(-segment_id, -y0, -y1),
                    family = binomial))

CAL2 <- 0.02
m2 <- nn_match(d2, "cameras_installed", "ps", caliper = CAL2)

# The three numbers that make this case what it is.
c2_levels   <- -(mean(d2$y1[m2$ti]) - mean(d2$y1[m2$ci]))          # naive on levels
c2_changes  <- mean((d2$y1 - d2$y0)[m2$ti]) - mean((d2$y1 - d2$y0)[m2$ci])
c2_se       <- sd((d2$y1 - d2$y0)[m2$ti] - (d2$y1 - d2$y0)[m2$ci]) / sqrt(length(m2$ti))
c2_ci       <- c2_changes + c(-1.96, 1.96) * c2_se

# The plain DiD on the same data, as a benchmark.
c2_did <- lm_robust(injury_collisions ~ cameras_installed * round, data = tc,
                    clusters = segment_id)
c2_did_b <- coef(c2_did)[["cameras_installed:round"]]

# Overlap: how much of the programme sits outside the comparison's range?
c2_ps_c   <- range(d2$ps[d2$cameras_installed == 0])
c2_outside <- sum(d2$ps[d2$cameras_installed == 1] < c2_ps_c[1] |
                  d2$ps[d2$cameras_installed == 1] > c2_ps_c[2])
c2_pct_out <- 100 * c2_outside / sum(d2$cameras_installed)

# The baseline gap that matching does NOT remove.
c2_base_t <- mean(d2$y0[m2$ti]); c2_base_c <- mean(d2$y0[m2$ci])

# Balance on covariates AND on the outcome's own baseline.
c2_bal <- map_dfr(c(cov2, "y0"), function(v) {
  nm <- if (v == "y0") "Collisions at baseline (the outcome)" else lab2[[v]]
  tibble(Characteristic = nm,
         Treated = mean(d2[[v]][m2$ti]),
         Control = mean(d2[[v]][m2$ci]),
         `Std. diff` = (mean(d2[[v]][m2$ti]) - mean(d2[[v]][m2$ci])) / sd(d2[[v]]))
})

# Bandwidth-style sensitivity for case 2: tighter calipers.
c2_cal_tbl <- map_dfr(c(0.02, 0.05, 0.10, Inf), function(cal) {
  mm <- nn_match(d2, "cameras_installed", "ps", caliper = cal)
  tibble(caliper = if (is.infinite(cal)) "none" else sprintf("%.2f", cal),
         kept = length(mm$ti),
         pct  = 100 * length(mm$ti) / sum(d2$cameras_installed),
         levels = -(mean(d2$y1[mm$ti]) - mean(d2$y1[mm$ci])),
         changes = mean((d2$y1 - d2$y0)[mm$ti]) - mean((d2$y1 - d2$y0)[mm$ci]))
})

# ---- The two-case comparison table ----------------------------------------
compare_tbl <- tibble(
  Question = c("Units that found a good match",
               "Overlap of the two groups",
               "Units outside the comparison's range",
               "Do covariates balance?",
               "Estimate on levels",
               "Estimate on changes",
               "Do the two agree?"),
  GreenWaste = c(sprintf("%s of %s (%d%%)", gfmt(c1_pairs), gfmt(sum(d1$enrolled)), round(c1_pct)),
                 "near perfect (0.44 to 0.61)",
                 "0",
                 "yes",
                 sprintf("%+.0f AED", c1_matched),
                 sprintf("%+.0f AED", c1_matched),
                 "yes"),
  Cameras = c(sprintf("%s of %s (%d%%)", gfmt(length(m2$ti)), gfmt(sum(d2$cameras_installed)),
                      round(100 * length(m2$ti) / sum(d2$cameras_installed))),
              "wide (0.00 to 0.98)",
              sprintf("%d (%d%%)", c2_outside, round(c2_pct_out)),
              "yes, but the outcome still differs",
              sprintf("%+.2f", c2_levels),
              sprintf("%+.2f", c2_changes),
              "no, opposite signs")
)
```

# 1 · The third way to build a comparison

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1 S3</div> <div class="v">randomisation &middot; the comparison chosen by a lottery</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences &middot; and an assumption that cannot be tested</div>
  <div class="k">Day 2 S2</div> <div class="v">regression discontinuity &middot; the comparison the rule creates</div>
  <div class="k">Today</div>     <div class="v">matching &middot; the comparison you build yourself</div>
</div>
```

::: {.panel}
Three methods, three ways of answering **compared to what?**

- Randomisation **creates** a comparison group.
- A cut-off **creates** one at the line.
- Matching **constructs** one, by finding lookalikes.
:::

::: {.ask}
Today is the most intuitive of the three, and the one that is easiest to get
wrong without noticing. Keep that tension in mind.
:::

## Matching, in one sentence

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">For each business that got the programme,<br>find one that did not but looks just like it.<br>Then compare the two.</div>
</div>')
```

::: {.panel}
There is no lottery and no cut-off. You build the comparison group yourself,
one matched pair at a time, using the characteristics you measured **before**
the programme started.
:::

::: {.ask}
Say it back before we go on. If you can say this sentence, you can interrogate
any matching study you are handed.
:::

## `[MENTI PLACEHOLDER]` Would that be fair?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
Two businesses, one enrolled and one did not. They are the same size, the same
age, the same area, in the same kind of neighbourhood.

**Is comparing them fair?**
:::

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Only if nothing else matters

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
**"Only if nothing else matters" is the honest answer, and it is the rest of this
session.** Matching handles what you measured. Everything you did not measure is
left in the comparison, and you will not be able to see it in the table.
:::

# 2 · Case 1: matching that works

## GreenWaste: who enrolled, and who did not

:::::: {.columns}

::::: {.column width="54%"}
Among the eligible businesses, some enrolled and some did not. That is the
comparison.

We match each enrolled business to the most similar non-enrolled business, using
six characteristics measured **before** the programme:
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Treated</div>  <div class="v">2,964 enrolled businesses</div>
  <div class="k">Comparison</div> <div class="v">2,664 not enrolled</div>
  <div class="k">Matched on</div> <div class="v">6 things measured at baseline</div>
  <div class="k">The rule</div> <div class="v">Cut costs by at least 1,000 AED</div>
</div>
```
:::::

::::::

::: {.ask}
Notice the order of that sentence. We match on what was measured **before** the
programme. If we matched on anything measured afterwards, we would be matching
on the very thing we are trying to measure.
:::

## Did every business find a match?

:::::: {.columns}

::::: {.column width="52%"}
```{r}
tibble(
  `The two groups` = c("Enrolled", "Not enrolled"),
  `Score range` = c(sprintf("%.3f to %.3f", c1_ps_t[1], c1_ps_t[2]),
                    sprintf("%.3f to %.3f", c1_ps_c[1], c1_ps_c[2])),
  `Matched` = c(sprintf("%s (100%%)", gfmt(c1_pairs)), ""),
  `Dropped for no match` = c("0", "")
) %>%
  kable(align = "lrrr", col.names = c("", "Score range", "Matched", "Dropped")) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="48%"}
::: {.panel}
Each of the `r gfmt(c1_pairs)` enrolled businesses found a comparison. **None
were dropped** for having no match.

The two score ranges sit almost on top of one another, which is what lets that
happen.
:::

::: {.warn}
**Remember what a dropped unit means.** A treated business with no comparable
untreated business cannot be matched. Drop enough of them and the study stops
being about the programme — it becomes about the easy-to-match part of it.
:::
:::::

::::::

## Balancing the two groups

:::::: {.columns}

::::: {.column width="54%"}
```{r}
c1_bal %>%
  mutate(across(c(Before, After), ~ round(.x, 3))) %>%
  kable(align = "lrr", digits = 3,
        col.names = c("Characteristic", "Before", "After")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  footnote(general = "Standardised differences. Anything beyond roughly +/-0.1 is usually considered a problem.",
           general_title = "", footnote_as_chunk = TRUE)
```
:::::

::::: {.column width="46%"}
::: {.panel}
**Before** shows how far apart the groups were on each characteristic.
**After** shows them once matched.

The worst difference falls from
`r sprintf("%.3f", max(abs(c1_bal$Before)))` to
`r sprintf("%.3f", max(abs(c1_bal$After)))`. Everything is close to zero.
:::

::: {.warn}
This table is the **only** evidence that the two groups are alike. If it looked
bad, nothing that follows would be worth reading.
:::
:::::

::::::

## The estimate

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the matched estimate (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">No matching</span></div>
  <div><span class="n">%s</span><span class="k">Matched</span></div>
  <div><span class="n">%s</span><span class="k">The rule</span></div>
</div>', gfmt(abs(c1_matched)), gfmt(c1_ci[1]), gfmt(c1_ci[2]),
   gfmt(abs(c1_naive)), gfmt(abs(c1_matched)), gfmt(RULEBAR)))
```

::: {.panel}
Matching barely moved the number: `r gfmt(abs(c1_naive))` AED became
`r gfmt(abs(c1_matched))` AED.

That is not a disappointment. It means the two groups were already very
similar, so there was little for matching to fix.
:::

::: {.ask}
And the rule? The interval runs
[`r gfmt(c1_ci[1])`, `r gfmt(c1_ci[2])`], and the least generous end is
`r gfmt(abs(c1_ci[2]))`. **Still short of `r gfmt(RULEBAR)`.**

Fifth result this week, fifth one that is real and not enough. But this case's
job is to show matching working cleanly, so hold that thought.
:::

## Why Case 1 is the clean one

```{=html}
<div class="facts">
  <div class="k">Overlap</div>   <div class="v">the two groups' scores sit almost on top of one another</div>
  <div class="k">Everyone matched</div> <div class="v">no treated business was dropped for want of a comparison</div>
  <div class="k">Balance</div>   <div class="v">worst standardised difference under 0.05 after matching</div>
  <div class="k">Matching changed little</div> <div class="v">which is itself informative: the groups were already alike</div>
</div>
```

::: {.ask}
So this case tells us the machinery works. Now take everything you just watched
— the match rate, the balance table, the agreement between matched and unmatched
— and apply it to a case where the programme went where the problem was.

Watch what breaks.
:::

# 3 · Case 2: matching that only half works

## The cameras went where the roads were worst

:::::: {.columns}

::::: {.column width="52%"}
```{r}
d2 %>%
  group_by(`Has a camera` = ifelse(cameras_installed == 1, "Yes", "No")) %>%
  summarise(`85th pct speed` = round(mean(baseline_speed_85th), 1),
            `Baseline collisions` = round(mean(y0), 2),
            `Prior 3-yr collisions` = round(mean(prior_collisions_3yr), 1),
            .groups = "drop") %>%
  kable(align = "lrrr", digits = 2) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The camera segments were **faster** (78.7 against 65.1 km/h) and had **more
collisions** before anything was installed.

That is the opposite of Case 1. There, the two groups chose themselves. Here,
the authority chose the worst roads on purpose.
:::

::: {.warn}
Matching has more work to do, and a harder job. The groups differ on things
that matter, and only some of those things were written down.
:::
:::::

::::::

## Finding a lookalike is harder here

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.8, fig.width=5.8}
plot_ly() %>%
  add_trace(data = d2 %>% filter(cameras_installed == 0),
            x = ~ps, type = "histogram", name = "No camera",
            marker = list(color = ACCENT2, line = list(width = 0)),
            opacity = 0.65, nbinsx = 40,
            hovertemplate = "score %{x:.2f}: %{y} segments<extra></extra>") %>%
  add_trace(data = d2 %>% filter(cameras_installed == 1),
            x = ~ps, type = "histogram", name = "Has a camera",
            marker = list(color = ACCENT, line = list(width = 0)),
            opacity = 0.65, nbinsx = 40,
            hovertemplate = "score %{x:.2f}: %{y} segments<extra></extra>") %>%
  layout(barmode = "overlay",
         xaxis = ax("Chance of looking like a camera segment", zeroline = FALSE),
         yaxis = ax("Segments")) %>%
  pl_m2(width = PL_WC, height = 265)
```
:::::

::::: {.column width="48%"}
::: {.panel}
The camera segments cluster at the **high** end of the score: they look like
camera segments because they are fast and busy.

The comparison group's range is wider, but thinner right where the camera
segments sit.
:::

::: {.warn}
**`r c2_outside` camera segments (`r round(c2_pct_out)`%) sit outside the
comparison group's range entirely.** For those, there is no lookalike to find.
:::
:::::

::::::

## The three numbers that do not agree

This is the slide the session exists for.

```{r}
tibble(
  `How the estimate is computed` = c(
    "Follow-up levels only (naive)",
    "Change within matched pairs",
    "Plain difference-in-differences, no matching"),
  `Estimate` = c(sprintf("%+.2f", c2_levels),
                 sprintf("%+.2f", c2_changes),
                 sprintf("%+.2f", c2_did_b)),
  `Says` = c("cameras made things WORSE",
             "cameras avoided collisions",
             "cameras avoided collisions")
) %>%
  kable(align = "lrrr", col.names = c("How the estimate is computed", "Estimate", "")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(1, bold = FALSE, background = "#F7E9E6")
```

::: {.ask}
Three ways of using the same matched pairs, and the first one has the **wrong
sign**.

Which is right, and what is the first one doing wrong?
:::

## Why the signs disagree

:::::: {.columns}

::::: {.column width="52%"}
::: {.panel}
The matched comparison segments were chosen because they had the same **speed**
as the camera segments. Look at what else came with that:

**Baseline collisions: `r sprintf("%.2f", c2_base_t)` on camera segments,
`r sprintf("%.2f", c2_base_c)` on their matches.**

Still `r sprintf("%.2f", c2_base_t - c2_base_c)` collisions apart, even after
matching.
:::

::: {.warn}
Matching balanced the things the model **contained**. The outcome's own starting
level was not one of them, so it was free to differ.
:::
:::::

::::: {.column width="48%"}
```{r fig.height=2.8, fig.width=5.6}
tibble(Group = factor(c("Camera segments", "Their matches"),
                      levels = c("Camera segments", "Their matches")),
       y0 = c(c2_base_t, c2_base_c)) %>%
  plot_ly(x = ~Group, y = ~y0, type = "bar",
          marker = list(color = c(ACCENT, ACCENT2), line = list(width = 0)),
          text = ~sprintf("%.2f", y0), textposition = "outside",
          textfont = list(size = 14, color = INK), showlegend = FALSE,
          hovertemplate = "%{y:.2f} collisions<extra></extra>") %>%
  layout(xaxis = ax("", zeroline = FALSE),
         yaxis = ax("Collisions at baseline", range = c(0, 16))) %>%
  pl_m2(width = PL_WC, height = 265,
        title = "Matched, and still not level")
```
:::::

::::::

## The fix is to compare changes, not levels

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">Collisions avoided per segment per round (95%% CI: %.2f to %.2f)</div>
</div>
<div class="stat-row">
  <div><span class="n">%.2f</span><span class="k">On levels (wrong sign)</span></div>
  <div><span class="n">-%.2f</span><span class="k">On changes</span></div>
  <div><span class="n">%.2f</span><span class="k">The rule</span></div>
</div>', c2_changes, c2_ci[1], c2_ci[2], c2_levels, abs(c2_changes), CAM_RULE))
```

::: {.panel}
Subtract each segment's own baseline before comparing, and the estimate becomes
`r sprintf("%.2f", abs(c2_changes))` avoided — close to the plain
difference-in-differences figure of `r sprintf("%.2f", abs(c2_did_b))`, which
used no matching at all. **Two methods that should agree, agreeing, is the
reassurance.** The levels comparison was the odd one out, and it was wrong.
:::

::: {.ask}
Does it clear the `r sprintf("%.1f", CAM_RULE)` rule? The interval runs
[`r sprintf("%.2f", c2_ci[1])`, `r sprintf("%.2f", c2_ci[2])`] in avoided
collisions — so yes, this one finally does.

**But read the rest of this session before you believe it.**
:::

## The bottom line of the two cases

```{r}
compare_tbl %>%
  kable(align = "l", col.names = c("", "Case 1 · GreenWaste", "Case 2 · Cameras")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  row_spec(6, bold = TRUE, background = "#E6F1F5") %>%
  row_spec(7, bold = TRUE, background = "#F7E9E6")
```

::: {.ask}
Read down the table. Case 1 passes every check. Case 2 passes most of them — and
still needed rescuing by a second method.

**Which of these two would you stake a decision on?**
:::

# 4 · The question matching cannot answer

## What matching can and cannot see

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">Matching can only account for what you measured.</div>
</div>')
```

::: {.panel}
Every characteristic in that balance table was **collected**. Matching cannot
balance on management quality, on how motivated the staff were, on whether the
business was about to be sold — unless somebody wrote it down.
:::

::: {.warn}
So the honest version of this method is: **we have made the two groups alike in
every way we can observe.** The words doing the work are the last four.
:::

## The question to put to any evaluator

:::::: {.columns}

::::: {.column width="52%"}
::: {.ask style="font-size: 118%;"}
For each business that enrolled, you found one that did not but looked the same.

**What else might be different about those two businesses that you could not
measure?**
:::

That is the whole of it. It is one sentence, and it cannot be answered from the
table.
:::::

::::: {.column width="48%"}
::: {.panel}
For GreenWaste, the plausible answers are things like:

**How badly did this business want to cut costs?** Businesses that enrol are
often the ones already looking for a solution.

**Was the owner about to invest anyway?** If so, we are crediting the programme
with a decision already made.
:::

::: {.warn}
Neither of these is in the data. Both would make the programme look better than
it was.
:::
:::::

::::::

## Which case would you trust?

::: {.ask}
Case 1 matched 100% of businesses, balanced every characteristic, and moved
almost nothing from the naive figure.

Case 2 matched 100% too, but left `r c2_outside` segments with no real
comparison, needed a change-based estimate to get a sensible sign, and came out
comfortably above the rule.

**Which would you stake a decision on, and why?**
:::

::: {.panel .fragment}
The uncomfortable answer: **Case 1** is the more trustworthy study, and its
result fails the rule. **Case 2** gives the attractive answer, and it is the one
carrying unquantified risk.

If your instinct was to prefer Case 2 because it says the programme worked, that
instinct is exactly what this session is here to interrupt.
:::

## What to ask the evaluator of Case 2

```{=html}
<div class="facts">
  <div class="k">Ask</div> <div class="v">"What happened to the 13 segments you could not match, and does their absence change the answer?"</div>
  <div class="k">Ask</div> <div class="v">"Why did you compare follow-up levels at all, when the groups still differed at baseline?"</div>
  <div class="k">Ask</div> <div class="v">"What did you decide to match on, and what was left out?"</div>
  <div class="k">Ask</div> <div class="v">"What else might differ between matched pairs that you did not measure?"</div>
</div>
```

::: {.ask}
None of these needs you to run anything. Each of them would change how much
weight you give the headline.

**Question two is the sharpest**, because it catches a real mistake that a
reported balance table would not reveal.
:::

# 5 · Put AI to work

## AI Snapshot: comparing two studies

::: {.panel}
Paste both sets of results — the two-case table and the two balance tables —
with this prompt:

**"Summarise the key differences between these two studies and identify what a
commissioner should flag."**
:::

::: {.ai-output}
**Case 1:** Matching was highly effective, with 100% of businesses matched and
excellent covariate balance. **The programme appears to have delivered a
reduction of around 1,000 AED in costs.** [It narrowly fails the 1,000 AED
threshold, so it should not be scaled]{.flag}.

**Case 2:** [Matching was similarly strong, with 100% of segments matched and
good balance]{.flag}. [The matched estimate shows cameras avoided 2.10
collisions per segment, comfortably exceeding the 2.0 threshold]{.flag},
**[so this programme should be scaled]{.flag}**.
:::

::: {.ask}
Which of those sentences would you sign your name to? Four minutes, in groups.
:::

## Debrief: one clean read, three false assurances

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it got right.** Case 1's numbers are reported accurately, including the
awkward part — it *did* notice the rule was not met.
:::

::: {.warn}
**1 · "Matching was similarly strong" in Case 2.** It was not. `r c2_outside`
segments had no comparable unit, and the levels-based estimate had the opposite
sign.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**2 · "Good balance."** The reported balance concealed that the outcome's own
baseline differed by over a collision per segment.
:::

::: {.warn}
**3 · "So this programme should be scaled."** The recommendation rests entirely
on sentence 2 being true. It is the conclusion, built on the false premise, and
stated most confidently of all.
:::
:::::

::::::

::: {.panel}
**The shape of this failure is new.** The model did not invent a number. It
repeated the study's **own summary of itself** — and the summary was the
misleading part. That is harder to catch: you have to compare the study's claims
against the study's details, not against the world.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Summarise the key differences between these two studies and identify what a
> commissioner should flag.

The model summarises the tables it was given, and repeats their framing.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Two matching studies. For each, tell me: how many treated units were dropped
> for want of a match; whether the **outcome's baseline** balanced, not just the
> covariates; and whether the levels-based and change-based estimates agree.
> **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both, then compare. The second prompt names the three diagnostics that
separate the cases — and notice that none of them is in either balance table.
:::

# 6 · The Abu Dhabi case

## `[PLACEHOLDER]` The DCD matching evaluation

::: {.warn}
**Awaiting the real findings.** The second case study in the outline is an Abu
Dhabi Department of Community Development matching study, on real-world data
where some participants had no good match. It has not been shared with this
workspace, so nothing here is presented as its result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>      <div class="v">[PLACEHOLDER]</div>
  <div class="k">Design</div>         <div class="v">Matching on observables</div>
  <div class="k">Match rate</div>     <div class="v">[PLACEHOLDER]</div>
  <div class="k">Outcome</div>        <div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed, and it is the same structure as Case 2 above. When the
DCD findings arrive they drop into these fields without redesigning the session.
:::

## The five questions for the DCD result

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">How many treated units found a match, and how many were dropped?</div>
  <div class="k">2</div> <div class="v">Do the covariates balance after matching — and does the outcome's own baseline balance?</div>
  <div class="k">3</div> <div class="v">What was matched on, and what was left out of the model?</div>
  <div class="k">4</div> <div class="v">Do the levels-based and change-based estimates agree?</div>
  <div class="k">5</div> <div class="v">What else might differ between matched pairs that was never measured?</div>
</div>
```

::: {.ask}
Same five questions we just worked through on the camera case. The programme
changes; the interrogation does not.
:::

# 7 · Close

## Five questions to ask any evaluator presenting matching

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">What did you match on, and what was left out?</div>
  <div class="k">2</div> <div class="v">How many treated units had no good match, and what happened to them?</div>
  <div class="k">3</div> <div class="v">Do the covariates balance — and does the outcome's baseline balance?</div>
  <div class="k">4</div> <div class="v">Did you compare levels or changes, and do the two agree?</div>
  <div class="k">5</div> <div class="v">What else might differ that you could not measure?</div>
</div>
```

::: {.panel}
Question 5 is the one that matters most, and it is the only one the evaluator
cannot answer from their own output.

That should tell you something about how much weight a matching study can carry
on its own.
:::

## The week so far

```{=html}
<div class="facts">
  <div class="k">Day 1 S2</div> <div class="v">descriptive reading &middot; -1.13 against a 2.0 rule</div>
  <div class="k">Day 1 S3</div> <div class="v">randomisation &middot; the comparison created by a lottery</div>
  <div class="k">Day 2 S1</div> <div class="v">DiD &middot; -816 against a 1,000 rule, one assumption untestable</div>
  <div class="k">Day 2 S2</div> <div class="v">RDD &middot; -249 against a 1,000 rule, and local to the line</div>
  <div class="k">Today</div>  <div class="v">matching &middot; the widest reach, and the weakest guarantee</div>
</div>
```

::: {.ask}
Each method bought credibility, and each paid for it:

**DiD** assumes parallel trends. **RDD** narrows the answer to a group near a
line. **Matching** assumes everything that matters was measured.

Those three sentences are the whole of Day 2. Tomorrow turns from reading
results to **questioning them**: cost analysis, data visualisation, and then a
clinic where you interrogate an analyst.
:::
