---
title: "Read the result"
subtitle: "Module 2 · Day 1 · Session 2"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day1_case.R')
```


## One question from memory {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
tp <- subset(gw, setting == 'city' & took_part == 1)
```

::: {.ask}
What would you ask before turning a fall into a claim of impact?
:::

Tell a partner. Keep the cards closed.

::: {.notes}
3 minutes. Retrieve the previous session’s question. Do not reteach the whole session. Hidden webR setup belongs beneath this heading to avoid a blank slide. Open the deck early for webR; its trainer demo is optional, with the same static result on paper.
:::

## What decision is required? {.reader-slide}

::: {.big-number}
`r fmt(decision_rule)` AED
:::

The course’s decision rule: an annual saving of at least this amount **per business**, measured **12 months later**.

::: {.ask}
What units and time period must the result use?
:::

Fictional GreenWaste case · invented decision rule

::: {.notes}
3 minutes. Distinguish a minimum saving from a value-for-money judgement. Programme costs, reach and implementation matter later. This rule alone is not a complete scale-up decision. It is not a real government policy.
:::

## Read this extract {.reader-slide}

```{r extract, results='asis'}
html_out(reading_table())
```

::: {.ask}
Which number answers “how much did costs change”?
:::

Read **Sheet 1** alone before discussing it.

::: {.notes}
5 minutes. Wait for reading. Ask participants to circle a number and underline the comparison words. Do not explain every column immediately. The extract is a before/after regression with business-clustered standard errors. The statistical method is in the trainer key, not a participant prerequisite.
:::

## Circle the change {.reader-slide}

```{r cost-pair, results='asis'}
html_out(sprintf('<div class="evidence-pair"><div class="evidence"><h3>Before</h3><p class="big-number">%s</p></div><div class="evidence"><h3>12 months later</h3><p class="big-number">%s</p></div></div><p class="caption">Annual waste cost, AED per city business that took part</p>%s',fmt(ba_mean['before']),fmt(ba_mean['after']),fiction))
```

::: {.ask}
What does the negative sign on `r fmt(case$before_after)` mean here?
:::

::: {.fragment}
Costs were `r fmt(abs(case$before_after))` AED lower. This is a change over time.
:::

::: {.notes}
5 minutes. The difference uses unrounded values, so subtracting displayed rounded means may differ by one AED. Distinguish lower costs from a harmful outcome. Ask for the time period and group in the spoken answer. Do not call the change the programme’s causal effect.
:::

## Find its row {.reader-slide}

**Coefficient**: the number beside a named result row.

::: {.ask}
Will this calculation give a rise or a fall?
:::

:::: {.columns}
::: {.column width="48%"}
```{webr}
#| echo: true
change <- tp$cost_after - tp$cost_before
data.frame(Result = 'After minus before',
           AED = round(mean(change)))
```
:::
::: {.column width="52%"}
```{r highlighted, results='asis'}
html_out(sprintf('<div class="evidence"><h3>After GreenWaste</h3><p class="big-number">%s</p><p>Coefficient, AED</p></div>%s',fmt(case$before_after),fiction))
```
:::
::::

::: {.notes}
5 minutes. Predict, then trainer Run. Participants read the output, not the code. The arithmetic produces the same coefficient as the before/after regression shown in the extract. The row’s name is essential: a different coefficient answers a different question. Use the printed coefficient if webR is unavailable. Never ask participants to fix code.
:::

## What does the interval add? {.reader-slide}

::: {.lead}
95% confidence interval: `r fmt(ba_ci[1])` to `r fmt(ba_ci[2])` AED
:::

A range compatible with the data and statistical model. It shows uncertainty around the estimate.

::: {.ask}
Does a narrow interval fix a poor comparison?
:::

::: {.fragment}
It does not account for every source of bias.
:::

Fictional GreenWaste case · invented figures

::: {.notes}
6 minutes. Box the interval on the worksheet. It is not a range of individual business savings or a 95% chance statement about this fixed interval. If needed: repeating the procedure under its assumptions yields intervals covering the true model quantity about 95% of the time. Do not teach repeated-sampling machinery. Precision cannot tell us what else changed over time.
:::

## Compare it with the rule {.reader-slide}

```{r before-rule, results='asis'}
html_out(interval_svg())
```

::: {.ask}
Does this reported change reach the rule?
:::

::: {.fragment}
The interval is below the rule. The comparison still does not establish what GreenWaste caused.
:::

::: {.notes}
6 minutes. Orient to the saving scale, which is positive here: it is the magnitude of the negative cost change in the table. Participants trace the estimate, interval and rule with a finger. Even a precise before/after change cannot settle a causal decision. Avoid concluding that the programme does not work.
:::

## Significant enough to act? {.reader-slide}

The **p-value** for this change is **less than 0.001**.

“Statistically significant” asks how unusual the result would be under a no-change model and its assumptions.

::: {.ask}
Does that tell us GreenWaste caused enough saving to extend it?
:::

::: {.fragment}
A small p-value establishes neither cause nor a large enough saving. A non-significant result does not prove no effect.
:::

Fictional GreenWaste case · invented figures

::: {.notes}
5 minutes. Recognise the term when reading reports. The p-value concerns the null tested by the analysis; here the before/after change, not a causal effect. It is not the probability that the programme works or that the result is due to chance. Do not introduce a hypothesis-testing lecture.
:::

## Finish the mark up {.reader-slide}

On **Sheet 1**:

Circle the estimate. Box its interval. Underline the comparison.

Write the units and period, then a decision sentence.

::: {.ask}
Which part of the scale-up decision does this extract leave unanswered?
:::

::: {.notes}
10 minutes. Five minutes alone, three in pairs, two for feedback. Every reader locates the result before discussing. Use the answer space on Sheet 1. The causal comparison is missing. Invite a threshold reading plus the limitation; do not accept “significant therefore extend”.
:::

## Repair an AI summary {.reader-slide data-correct="The second sentence is unsupported. The extract measures a change over time; it does not isolate what GreenWaste caused."}

Select every unsupported claim in this illustrative AI draft.

```{r ai-summary, results='asis'}
html_out(sprintf('<div class="claim-list"><button data-claim="supported" aria-pressed="false">City participants&#39; annual costs fell by %s AED.</button><button data-claim="unsupported" aria-pressed="false">GreenWaste caused that saving, so scale-up is justified.</button></div><button data-check>Check against the extract</button><p data-feedback class="feedback" role="status"></p>%s',fmt(abs(case$before_after)),fiction))
```

::: {.ask}
How would you rewrite the second sentence using only this source?
:::

::: {.notes}
7 minutes. Use Sheet 2. Example rewrite: the observed fall is below the rule, but the extract cannot isolate the programme’s effect. Ask an AI to identify the comparison, quote the supporting result row and ask for missing information before recommending. Check its answer against the source yourselves. A second AI review is assistance, not verification. The draft is an authored example.
:::

## Write the decision sentence {.reader-slide}

::: {.ask}
What can you say confidently, and what would you ask next?
:::

Write a result with its units, one limit, and your next question. Use the extract.

::: {.notes}
5 minutes. Individual close. Accept plain wording: participants’ annual costs fell about the estimated amount after 12 months; the change is below the 1,000 AED rule; we need a credible estimate of what happened without the programme. Do not turn this into a definitive rejection of GreenWaste. Total 60 minutes.
:::
