---
title: "Plan Your Own Evaluation"
subtitle: "Module 2 · Day 4, Session 3: Building on Module 1"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_dge.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(kableExtra)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"

# ---- The template ----------------------------------------------------------
# Shared with the printable Evaluation Design Update template.
source("qa_design_update.R")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()
```

# 0 · The last session {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Four days. One design.

::: {.panel}
On Monday you learned what a comparison is. Tuesday, three more ways of building one.
Wednesday, what a result costs and what a figure hides. Thursday, how to tell
whether evidence is credible and how to translate it.

Today you point all of it at your own evaluation. Nothing new is taught: the
template has three parts (where your design stands, the four additions, and your
pitch), and most of the session is working time.
:::

::: {.ask .fragment}
**How far along is your design?**
Not started / rough outline / drafted / ready to pitch.
:::

# 1 · Where your design stands {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The eight conditions, from Module 1

```{r}
#| results: asis
tab <- data.frame(
  `#` = seq_len(nrow(design_conditions)),
  Condition = design_conditions$condition,
  check.names = FALSE
)
kable(tab, col.names = c("#", "Condition"),
      caption = "The eight conditions for a credible impact evaluation") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Nothing here has changed since Module 1, and nothing this week contradicted it.
Part 1 of the template asks you to write your design against these, so the new
work has something to attach to.
:::

## The one condition that decides the rest

::: {.ask}
Of the eight, which one, if you get it wrong, makes the other seven irrelevant?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**The counterfactual.**

Every method this week has been a different way of operationalising it:
randomisation, a cut-off, a matched comparison, a parallel trend.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Each one has its own way of failing. If you cannot state yours in one sentence,
the design is not finished.
:::
:::::

::::::

::: {.ask .fragment}
Write your counterfactual sentence now, before we go on. One sentence. You will
read it out in the pitch.
:::

## Working time

::: {.panel}
**Twenty minutes.** Part 1 of the template: write your design as it stands,
against the eight conditions.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
If you have the Module 1 document, work from it. If not, the recap on the previous
slide is enough: write what you would do, not what you did.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Do not polish. This is the foundation for Part 2, and Part 2 is what the pitch focuses on.
:::
:::::

::::::

# 2 · The four additions {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What you are adding

```{r}
#| results: asis
tab <- data.frame(
  Addition = design_additions$addition,
  `From` = design_additions$from,
  check.names = FALSE
)
kable(tab, col.names = c("Addition", "From"),
      caption = "The four additions, and where each came from") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 16)
```

::: {.panel}
Each of these is something this week taught you to demand of someone else's
evaluation. Now you have to supply it for your own.
:::

::: {.ask .fragment}
Which of the four is hardest to do **before** you have any data? Why?
:::

## Addition 1: results presentation plan

::: {.ask}
What will you show, to whom, and in what order? Name the one number you will lead
with, and say what the reader must **not** be able to conclude from it.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**This morning's lesson, applied**

The fictitious report led with its largest estimate and called a different one its
preferred specification. A reader who stopped after the summary would have
believed the threshold was met, when the preferred estimate did not meet it.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
So ask: "What will a reader who reads only the headline come away believing,
and is that true?"
:::
:::::

::::::

## Addition 2: cost-benefit framework

::: {.ask}
Four things to state: the **threshold**, the **benefit per unit per year**, the
**horizon**, and the **discount rate**.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

In the GreenWaste example the horizon moved the ratio more than the discount
rate did, and reports often leave the horizon unstated.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Write your horizon down. If your benefit stream runs for five years and someone
assumes two, your ratio moves from 1.87 to 0.80, which changes the decision.
:::
:::::

::::::

## Addition 3: data visualisation sketch

::: {.ask}
Sketch the figure that carries your headline result. Then sketch the **same data
presented a second way**, and say what the second version hides.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

Every pair of figures showed identical numbers and different impressions. The
truncated axis, the bar of means, the denominator, the levels-versus-changes, the
discounting.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second sketch matters more. If you cannot say what your preferred figure
hides, expect it to come up in questions.
:::
:::::

::::::

## Addition 4: evidence translation plan

::: {.ask}
Write the **headline sentence** of your brief, with the number in it. Then name
the one thing you will state plainly that your evaluation did **not** establish.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**This morning's lesson**

A brief that reports the estimate and omits the threshold leaves the reader to
assume the target was met. The problem is what it leaves out.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second half is harder. Naming your own limit makes the pitch more
credible.
:::
:::::

::::::

## Working time

::: {.panel}
**Thirty minutes.** Part 2 of the template: all four additions.

Work as a team. Split the four if that is faster, but every member should be able
to speak to all four by the end.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Use the template's prompts. Each one is phrased as the question a commissioner
would ask, because that is who you are pitching to.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
You will not finish all four to the same standard. That is expected. Decide as a
team which two you will lead on.
:::
:::::

::::::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the pitches.
:::

# 3 · The pitch {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Five minutes, five parts

```{r}
#| results: asis
tab <- data.frame(
  Part = pitch_structure$part,
  Time = pitch_structure$minutes,
  `What it says` = pitch_structure$says,
  check.names = FALSE
)
kable(tab, col.names = c("Part", "Time", "What it says"),
      caption = "The pitch structure") |>
  kable_styling(c("striped", "condensed"), full_width = FALSE, position = "left",
                font_size = 15)
```

::: {.warn}
Note the order. **The limit comes before the ask.** Most pitches put it after,
if at all.
:::

## Why the limit comes first

::: {.ask}
If you tell the room what your design cannot do, why would that make them more
likely to give you what you want?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
Because it tells them **which parts you have already thought about**. A pitch that
claims no limits is either naive or hiding something, and the room cannot tell
which.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
It also changes what you are asking for. If you know your design cannot establish
X, you can ask for what would, which is a more specific request than "fund the
evaluation".
:::
:::::

::::::

## Feedback from the room

::: {.panel}
Three minutes per team. The room gives feedback in this order, and the order
matters:

1. **What is the design?** Say it back in one sentence. If you cannot, that is the
   feedback.
2. **What would make it fail?** The one assumption that would have to break.
3. **What is missing?** One thing you would have wanted to hear.
:::

::: {.warn}
**Write it down**, so the team can use it. Aim it at the design, never at the team.
:::

::: {.notes}
The first question is a check: if the room cannot say the design back in one
sentence, the pitch did not land, and saying so is the most useful feedback a
team can get. Better to learn that here than from a real commissioner.
:::

# 4 · The week in one slide {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What you can now do

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Say what a number is compared to, and why that comparison is the argument.</div>\n')
cat('<div class="panel"><b>2.</b> Read DiD, RDD and matching results, and name the assumption each rests on.</div>\n')
cat('<div class="panel"><b>3.</b> Judge whether a benefit-cost ratio clears a threshold, and what decides it.</div>\n')
cat('<div class="panel"><b>4.</b> See the same data drawn two ways, and say what each version hides.</div>\n')
cat('<div class="panel"><b>5.</b> Interrogate an evaluation with a checklist, and rate a report section by section.</div>\n')
cat('<div class="panel"><b>6.</b> Write a brief a decision-maker can act on without being misled.</div>\n')
```

::: {.ask .fragment}
Which of the six would you have said you could do on Monday morning? Which would
you now add?
:::

## The one question

::: {.panel}
The question behind every session this week:

> **What is being compared to what, and why should those two things be
> comparable?**
:::

::: {.ask .fragment}
Each method this week answered it differently, and each can be wrong about it
in its own way. The evaluations you commission will depend on it too.
:::

## Thank you

::: {.panel}
The slides stay available online, and every dataset and script is in the
repository if you want to reproduce any of the analyses on your own programme.
:::

::: {.ask}
What is the first evaluation decision you will make when you get back to your
desk?
:::
