---
title: "Reading Matching Results"
subtitle: "Module 2 · Day 2, Session 3: Two Case Studies, One Clean and One Not"
author: "3ie"
always_allow_html: yes
# live-revealjs (quarto-live): the baked plotly/kable charts still render at knit time;
# the teaching code cells (checking the data, building the score and finding the twins,
# common support, running the comparison, and the "never measured" check) are {webr}
# cells that run live in the browser via WebAssembly. Needs internet at presentation
# time to load webR. Export to PDF for a portable copy (live cells then show their
# code but do not run).
format:
  live-revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    # Single-file deck. The live (webR) cells work fine with this over http,
    # via `quarto preview` or a published URL, but NOT from a file:// path.
    # Do not open the rendered .html directly from disk; serve it or publish it.
    embed-resources: true
engine: knitr
resources:
  - evaluation_data_GreenWaste.csv
webr:
  packages:
    - dplyr
    - ggplot2
    - estimatr
    - broom
  cell-options:
    autorun: false
    edit: true
    warning: false
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(broom)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"
ACCENT2 <- "#A9561F"
ALERT   <- "#8E2413"
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12)))
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  # Regenerate the QR when the qrcode package is available; otherwise keep the
  # menti_qr.png already in the folder.
  if (!requireNamespace("qrcode", quietly = TRUE)) return(path)
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qrcode::qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR  <- 1000    # GreenWaste bar, in AED
CAM_RULE <- 2.0     # camera bar, in collisions avoided per segment per round

# ---- The matcher -----------------------------------------------------------
# Nearest neighbour on the likelihood score, with replacement. The SAME
# function runs in the live (webR) cells, so the numbers on the baked slides
# and the numbers the trainer produces live are identical. Written out rather
# than called from a package, because the session's job is to show what
# matching actually does: for each treated unit, take the untreated unit with
# the closest score, then compare.
find_twins <- function(data, score, treat = "enrolled") {
  took <- data[data[[treat]] == 1, ]
  rest <- data[data[[treat]] == 0, ]
  twin <- sapply(took[[score]], function(s) which.min(abs(s - rest[[score]])))
  out  <- bind_rows(took, rest[twin, ])
  attr(out, "twin_rows") <- twin
  out
}

# ===========================================================================
# CASE 1: GreenWaste. Every business, enrolled against not enrolled.
#
# Why the whole sample and not just the eligible businesses: within programme
# neighbourhoods EVERY eligible business enrolled, so the "not enrolled"
# eligible businesses are all in neighbourhoods that were never offered the
# programme. Restricting to eligible businesses therefore reproduces the
# randomised comparison (raw gap -1,014, the RCT number) and leaves matching
# nothing to do. The whole-sample comparison has real selection in it (the
# enrolled businesses had lower costs before the programme started, and all
# sit below the efficiency cut-off), which is what matching is for.
# ===========================================================================
gw  <- read.csv("./evaluation_data_GreenWaste.csv")

biz <- gw %>% filter(round == 0) %>%
  select(business_identifier, neighborhood_identifier, enrolled,
         efficiency_index, age_manager, educ_manager, female_manager,
         foreign_owned, staff_size, advanced_filtration,
         water_treatment_system, business_area, recycling_center_distance,
         y0 = waste_management_costs) %>%
  left_join(gw %>% filter(round == 1) %>%
              select(business_identifier, y1 = waste_management_costs),
            by = "business_identifier") %>%
  mutate(dy = y1 - y0)

chars <- c("efficiency_index", "age_manager", "educ_manager", "female_manager",
           "foreign_owned", "staff_size", "advanced_filtration",
           "water_treatment_system", "business_area",
           "recycling_center_distance")
lab_chars <- c(efficiency_index = "Efficiency index",
               age_manager = "Manager age",
               educ_manager = "Manager education",
               female_manager = "Female manager",
               foreign_owned = "Foreign owned",
               staff_size = "Staff size",
               advanced_filtration = "Advanced filtration",
               water_treatment_system = "Water treatment system",
               business_area = "Business area",
               recycling_center_distance = "Distance to recycling centre")

f_all      <- reformulate(chars, "enrolled")
f_no_index <- reformulate(setdiff(chars, "efficiency_index"), "enrolled")
f_thin     <- enrolled ~ age_manager + educ_manager

# Before matching: the raw comparison and the baseline gap.
n_took   <- sum(biz$enrolled == 1)
n_rest   <- sum(biz$enrolled == 0)
raw_t1   <- mean(biz$y1[biz$enrolled == 1]); raw_c1 <- mean(biz$y1[biz$enrolled == 0])
raw_t0   <- mean(biz$y0[biz$enrolled == 1]); raw_c0 <- mean(biz$y0[biz$enrolled == 0])
raw_gap  <- raw_t1 - raw_c1

# Steps 1 and 2: the score, then the twins.
biz$score <- fitted(glm(f_all, data = biz, family = binomial))
m1        <- find_twins(biz, "score")
twins_used <- length(unique(attr(m1, "twin_rows")))

# Common support: enrolled businesses scoring beyond anything in the pool.
ps_t <- range(biz$score[biz$enrolled == 1])
ps_c <- range(biz$score[biz$enrolled == 0])
n_beyond <- sum(biz$score[biz$enrolled == 1] > ps_c[2] |
                biz$score[biz$enrolled == 1] < ps_c[1])

# Step 3: balance, before and after, as standardised differences.
std_diff <- function(d, v) {
  (mean(d[[v]][d$enrolled == 1]) - mean(d[[v]][d$enrolled == 0])) / sd(biz[[v]])
}
bal1 <- tibble(var = chars,
               Characteristic = unname(lab_chars[chars]),
               Before = map_dbl(chars, ~ std_diff(biz, .x)),
               After  = map_dbl(chars, ~ std_diff(m1,  .x)))

# Step 4: the comparison on the matched sample.
fit1   <- lm_robust(y1 ~ enrolled, data = m1, clusters = neighborhood_identifier)
est1   <- coef(fit1)[["enrolled"]]
ci1    <- c(fit1$conf.low[["enrolled"]], fit1$conf.high[["enrolled"]])
# Sign trap: the estimate is negative, so the end of the interval closest to
# zero is the LEAST generous one. Sort the absolute values: [1] is that end.
avoid1 <- sort(abs(ci1))
tbl1   <- tidy(fit1) %>% select(term, estimate, std.error, conf.low, conf.high)

# ---- Sensitivity: change what you match on, and how you compare -----------
sens_fit <- function(f, outcome = "y1") {
  d <- biz; d$s <- fitted(glm(f, data = d, family = binomial))
  m <- find_twins(d, "s")
  r <- lm_robust(reformulate("enrolled", outcome), data = m,
                 clusters = neighborhood_identifier)
  list(est = coef(r)[["enrolled"]],
       lo = r$conf.low[["enrolled"]], hi = r$conf.high[["enrolled"]],
       index_gap = mean(m$efficiency_index[m$enrolled == 1]) -
                   mean(m$efficiency_index[m$enrolled == 0]),
       worst_other = max(abs(map_dbl(setdiff(chars, "efficiency_index"),
                                     ~ std_diff(m, .x)))))
}
s_full    <- sens_fit(f_all)
s_change  <- sens_fit(f_all, "dy")
s_noindex <- sens_fit(f_no_index)
s_thin    <- sens_fit(f_thin)

sens_tbl <- tibble(
  spec = c("All 10 characteristics",
           "All 10, compared as changes",
           "Without the efficiency index",
           "Manager age and education only"),
  est = c(s_full$est, s_change$est, s_noindex$est, s_thin$est),
  lo  = c(s_full$lo,  s_change$lo,  s_noindex$lo,  s_thin$lo),
  hi  = c(s_full$hi,  s_change$hi,  s_noindex$hi,  s_thin$hi))

# ---- The picture of what matching does (illustration on real businesses) --
# A small sample, matched on TWO characteristics only so that the pairs can be
# drawn on a flat chart. The real analysis above uses all ten.
set.seed(2026)
pic <- bind_rows(
  biz %>% filter(enrolled == 1) %>% slice_sample(n = 16),
  biz %>% filter(enrolled == 0) %>% slice_sample(n = 55)) %>%
  mutate(zx = (efficiency_index - mean(biz$efficiency_index)) / sd(biz$efficiency_index),
         zy = (age_manager - mean(biz$age_manager)) / sd(biz$age_manager))
pic_t <- pic %>% filter(enrolled == 1)
pic_c <- pic %>% filter(enrolled == 0)
pic_j <- sapply(seq_len(nrow(pic_t)), function(k)
  which.min((pic_c$zx - pic_t$zx[k])^2 + (pic_c$zy - pic_t$zy[k])^2))
pic_c$used <- seq_len(nrow(pic_c)) %in% pic_j

# ===========================================================================
# CASE 2: traffic cameras (the roads-authority data from Day 1 Session 2,
# FICTIONAL, generated for this training). Cameras went where the roads were
# worst, not at random. Here matching is harder: some camera segments have no
# lookalike at all, and the matched pairs still start apart on the outcome.
# All quantities below are in "collisions AVOIDED", so a bigger number is a
# bigger benefit and the rule (2.0) reads the same way for every row.
# ===========================================================================
tc  <- read.csv("./evaluation_data_TrafficCameras.csv")
tc0 <- tc %>% filter(round == 0)
tc1 <- tc %>% filter(round == 1)

cov2 <- c("baseline_speed_85th", "lanes", "road_length_km",
          "school_within_500m", "prior_collisions_3yr")
lab2 <- c(baseline_speed_85th = "85th percentile speed (km/h)",
          lanes = "Lanes",
          road_length_km = "Segment length (km)",
          school_within_500m = "School within 500 m",
          prior_collisions_3yr = "Collisions, prior 3 years")

d2 <- tc0 %>%
  mutate(lighting_poor = as.integer(lighting_quality == "poor")) %>%
  select(segment_id, cameras_installed, y0 = injury_collisions,
         all_of(cov2), lighting_poor) %>%
  left_join(tc1 %>% select(segment_id, y1 = injury_collisions), by = "segment_id") %>%
  mutate(dy = y1 - y0)
d2$score <- fitted(glm(cameras_installed ~ baseline_speed_85th + lanes +
                         road_length_km + school_within_500m +
                         prior_collisions_3yr + lighting_poor,
                       data = d2, family = binomial))
m2 <- find_twins(d2, "score", treat = "cameras_installed")

n_cam      <- sum(d2$cameras_installed == 1)
n_nocam    <- sum(d2$cameras_installed == 0)
c2_used    <- length(unique(attr(m2, "twin_rows")))
c2_ps_c    <- range(d2$score[d2$cameras_installed == 0])
c2_outside <- sum(d2$score[d2$cameras_installed == 1] < c2_ps_c[1] |
                  d2$score[d2$cameras_installed == 1] > c2_ps_c[2])
c2_pct_out <- 100 * c2_outside / n_cam

m2_t <- m2 %>% filter(cameras_installed == 1)
m2_c <- m2 %>% filter(cameras_installed == 0)
c2_base_t <- mean(m2_t$y0); c2_base_c <- mean(m2_c$y0)
c2_base_sd <- (c2_base_t - c2_base_c) / sd(d2$y0)

c2_levels  <- -(mean(m2_t$y1) - mean(m2_c$y1))        # avoided, follow-up levels
pair_chg   <- -(m2_t$dy - m2_c$dy)                     # avoided, change per pair
c2_changes <- mean(pair_chg)
c2_ci      <- c2_changes + c(-1.96, 1.96) * sd(pair_chg) / sqrt(length(pair_chg))
c2_did     <- -coef(lm_robust(injury_collisions ~ cameras_installed * round,
                              data = tc, clusters = segment_id))[["cameras_installed:round"]]

c2_bal <- map_dfr(c(cov2, "y0"), function(v) {
  tibble(Characteristic = if (v == "y0") "Collisions at baseline (the outcome)" else lab2[[v]],
         Before = (mean(d2[[v]][d2$cameras_installed == 1]) -
                   mean(d2[[v]][d2$cameras_installed == 0])) / sd(d2[[v]]),
         After  = (mean(m2_t[[v]]) - mean(m2_c[[v]])) / sd(d2[[v]]))
})
c2_worst_cov <- max(abs(c2_bal$After[c2_bal$Characteristic != "Collisions at baseline (the outcome)"]))

# ---- The two-case comparison table ----------------------------------------
compare_tbl <- tibble(
  Question = c("Every treated unit had a real lookalike?",
               "Comparison units reused as twins",
               "Characteristics balanced after matching?",
               "Outcome's own baseline balanced?",
               "Estimate, follow-up levels",
               "Estimate, changes",
               "Least generous end vs the rule"),
  GreenWaste = c(sprintf("yes, all but %d", n_beyond),
                 sprintf("%s twins for %s businesses", gfmt(twins_used), gfmt(n_took)),
                 sprintf("yes (worst %.2f)", max(abs(bal1$After))),
                 sprintf("yes (gap %s AED)", gfmt(mean(m1$y0[m1$enrolled == 1]) - mean(m1$y0[m1$enrolled == 0]))),
                 sprintf("%s AED", gfmt(s_full$est)),
                 sprintf("%s AED", gfmt(s_change$est)),
                 sprintf("%s against %s: short", gfmt(avoid1[1]), gfmt(RULEBAR))),
  Cameras = c(sprintf("no, %d segments (%d%%) outside", c2_outside, round(c2_pct_out)),
              sprintf("%d twins for %d segments", c2_used, n_cam),
              sprintf("yes (worst %.2f)", c2_worst_cov),
              sprintf("no (std. diff %.2f)", c2_base_sd),
              sprintf("%.2f avoided", c2_levels),
              sprintf("%.2f avoided", c2_changes),
              sprintf("%.2f against %.1f: short", min(c2_ci), CAM_RULE))
)
```


# 1 · The third way to build a comparison

<!-- Hidden setup (webR priming, deck CSS). Kept under the first section heading, not before it: content before the first heading makes Quarto create an empty slide after the title. -->
:::: {style="display:none;"}
```{webr}
#| autorun: true
#| edit: false
#| echo: false
#| output: false
#| warning: false
#| message: false
# ---- Live (webR) setup ----------------------------------------------------
# Runs once in the browser when the deck opens, priming the shared webR session
# so every teaching cell has its data and helpers ready even if the trainer
# skips one. Separate from the knitr `setup` chunk above, which prepares the
# baked charts and tables. The matcher is the SAME function as in knitr.
library(dplyr)
library(ggplot2)
library(estimatr)
library(broom)

gw <- read.csv("evaluation_data_GreenWaste.csv")

# One row per business: baseline characteristics, costs before (y0) and after (y1)
biz <- gw %>% filter(round == 0) %>%
  select(business_identifier, neighborhood_identifier, enrolled,
         efficiency_index, age_manager, educ_manager, female_manager,
         foreign_owned, staff_size, advanced_filtration,
         water_treatment_system, business_area, recycling_center_distance,
         y0 = waste_management_costs) %>%
  left_join(gw %>% filter(round == 1) %>%
              select(business_identifier, y1 = waste_management_costs),
            by = "business_identifier")

f_all <- enrolled ~ efficiency_index + age_manager + educ_manager +
  female_manager + foreign_owned + staff_size + advanced_filtration +
  water_treatment_system + business_area + recycling_center_distance
f_no_index <- update(f_all, . ~ . - efficiency_index)

find_twins <- function(data, score) {
  took <- data[data$enrolled == 1, ]
  rest <- data[data$enrolled == 0, ]
  twin <- sapply(took[[score]], function(s) which.min(abs(s - rest[[score]])))
  bind_rows(took, rest[twin, ])
}

biz$score <- fitted(glm(f_all, data = biz, family = binomial))
matched   <- find_twins(biz, "score")

ACCENT  <- "#0E6E80"; ACCENT2 <- "#A9561F"; ALERT <- "#8E2413"
INK <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
grp_cols <- c("Enrolled" = ACCENT, "Not enrolled" = ACCENT2)

theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(text = element_text(colour = INK),
          axis.title = element_text(colour = MUTED, size = rel(0.85)),
          axis.text = element_text(colour = MUTED),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
          legend.position = "bottom", legend.title = element_blank(),
          plot.margin = margin(4, 10, 4, 4))
}
```
::::

```{=html}
<style>
/* This deck: smaller live-code editors and output, so code plus result fits
   on a 1280x720 slide. The extension default is 1.3rem. */
.reveal .card.exercise-editor { --exercise-font-size: 1.05rem; margin: 0.3rem 0; }
.reveal .card.exercise-editor .card-header { padding: 0.25rem 0.75rem; }
.reveal .cell-output-container pre code,
.reveal .exercise-cell-output pre code { font-size: 0.85rem; line-height: 1.3; }
.reveal .cell-output-container pre { margin: 0.2rem 0; }
.reveal div.sourceCode pre code { font-size: 1.0rem; line-height: 1.35; }
.reveal .cell-output-display img,
.reveal .cell-output-display canvas { max-height: 280px; width: auto !important; }
</style>
```

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1 S3</div> <div class="v">randomisation &middot; the comparison chosen by a lottery</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences &middot; and an assumption that could not be tested</div>
  <div class="k">Day 2 S2</div> <div class="v">regression discontinuity &middot; the comparison the rule creates</div>
  <div class="k">Today</div>     <div class="v">matching &middot; the comparison you build yourself</div>
</div>
```

::: {.panel}
Three methods, three ways of answering **compared to what?**

- Randomisation **creates** a comparison group.
- A cut-off **creates** one at the line.
- Matching **constructs** one, by finding lookalikes.
:::

::: {.ask}
Today is the most intuitive of the three, and the one that is easiest to get
wrong without noticing. Keep that tension in mind.
:::

::: {.notes}
Section 1 is the recap and the idea: aim for about 15 minutes. The room already
knows GreenWaste, the 1,000 AED rule and the selection problem from Day 1
Session 3, so do not re-teach why a raw comparison misleads. The new material
is how matching builds a comparison (Section 2) and what it cannot see
(Section 3).
:::

## `[MENTI PLACEHOLDER]` Would that be fair?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
Two businesses, one enrolled and one did not. They are the same size, the same
age, the same area, in the same kind of neighbourhood.

**Is comparing them fair?**
:::

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Only if nothing else matters

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel .fragment}
**"Only if nothing else matters" is the best answer. Most of this session
explains why.** Matching handles what you measured. Everything you did not measure is
left in the comparison, and you will not be able to see it in the table.
:::

## Matching, in one sentence

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">For each business that got the programme,<br>find one that did not but looks just like it.<br>Then compare the two.</div>
</div>')
```

::: {.panel}
There is no lottery and no cut-off. You build the comparison group yourself,
one matched pair at a time, using the characteristics you measured **before**
the programme started.
:::

::: {.ask}
Say it back before we go on. If you can say this sentence, you can interrogate
any matching study you are handed.
:::

## Your turn: find the twins

:::::: {.columns}

::::: {.column width="50%"}
```{r}
tibble(
  ` ` = c("A", "B", "C"),
  `Manager age` = c(45, 38, 52),
  Size = c("small", "large", "small"),
  `Cost after` = c("900", "1,400", "700")
) %>%
  kable(align = "lccr", caption = "Enrolled") %>%
  kable_styling(full_width = FALSE, font_size = 19)
```
:::::

::::: {.column width="50%"}
```{r}
tibble(
  ` ` = c("1", "2", "3", "4"),
  `Manager age` = c(39, 46, 51, 29),
  Size = c("large", "small", "small", "large"),
  `Cost after` = c("2,300", "1,900", "1,800", "2,600")
) %>%
  kable(align = "lccr", caption = "Not enrolled") %>%
  kable_styling(full_width = FALSE, font_size = 19)
```
:::::

::::::

::: {.ask}
Pair each enrolled business with its closest twin, matching on **who they
are**, not on their cost. Then take the average cost difference across the
three pairs. Ninety seconds.
:::

::: {.panel .fragment}
**A with 2, B with 1, C with 3.** Differences of -1,000, -900 and -1,100:
an average of **-1,000 AED**. Business 4 is nobody's twin, so it is set aside.
:::

::: {.notes}
Hand out the twins card (see the exercise plan). Genuine silence for ninety
seconds. The point they should feel: matching is intuitive, they already know
how to do it, and it quietly sets aside units that resemble nobody. That is
the common-support idea, which comes back in Section 2 and is the heart of
Case 2.

Circulate and listen for "but what if none of them really match?". That
instinct is exactly Case 2. Also listen for anyone matching on cost: that is
matching on the outcome, which would build the answer into the comparison.
:::

## What matching does, on real businesses

```{r fig.height=3.6, fig.width=11.5}
# Three frames. Each trace is present in every frame, with the same number of
# points, so the slider moves cleanly in both directions.
steps <- c("1 · Everyone", "2 · Find each twin", "3 · Keep the pairs")
off <- -100   # a parked point, off the fixed axis range
fr_t <- map_dfr(steps, ~ pic_t %>% mutate(step = .x))
fr_c <- map_dfr(steps, function(s) pic_c %>% mutate(step = s,
  col = case_when(s == steps[3] & !used ~ "#D3D7DA", TRUE ~ ACCENT2)))
links <- map_dfr(seq_len(nrow(pic_t)), function(k) tibble(
  x = c(pic_t$efficiency_index[k], pic_c$efficiency_index[pic_j[k]], NA),
  y = c(pic_t$age_manager[k], pic_c$age_manager[pic_j[k]], NA)))
fr_l <- bind_rows(
  tibble(x = off, y = off, step = steps[1]),
  links %>% mutate(step = steps[2]),
  links %>% mutate(step = steps[3]))

plot_ly() %>%
  add_trace(data = fr_l, x = ~x, y = ~y, frame = ~step, type = "scatter",
            mode = "lines", line = list(color = MUTED, width = 1.5),
            hoverinfo = "skip", showlegend = FALSE) %>%
  add_trace(data = fr_c, x = ~efficiency_index, y = ~age_manager, frame = ~step,
            type = "scatter", mode = "markers",
            marker = list(color = ~col, size = 11, symbol = "circle",
                          line = list(color = "white", width = 1)),
            hovertemplate = "Not enrolled<br>index %{x:.1f}, manager %{y}<extra></extra>",
            showlegend = FALSE) %>%
  add_trace(data = fr_t, x = ~efficiency_index, y = ~age_manager, frame = ~step,
            type = "scatter", mode = "markers",
            marker = list(color = ACCENT, size = 14, symbol = "triangle-up",
                          line = list(color = "white", width = 1)),
            hovertemplate = "Enrolled<br>index %{x:.1f}, manager %{y}<extra></extra>",
            showlegend = FALSE) %>%
  animation_opts(frame = 1200, transition = 0, redraw = FALSE) %>%
  animation_slider(currentvalue = list(visible = FALSE),
                   font = list(size = 12, color = INK), x = 0.08, len = 0.84) %>%
  layout(xaxis = ax("Efficiency index at baseline", range = c(20, 100), zeroline = FALSE),
         yaxis = ax("Manager age", range = c(15, 90), zeroline = FALSE)) %>%
  pl_m2(height = 355, t = 34, b = 92,
        title = "&#9650; enrolled &nbsp;&nbsp; &#9679; not enrolled &nbsp;&nbsp; (71 real GreenWaste businesses, matched on two characteristics)")
```

::: {.ask}
Drag the slider. Where do the triangles sit, and which circles are never
anyone's twin?
:::

::: {.notes}
Real GreenWaste businesses, a random 16 enrolled and 55 not enrolled, matched
here on just two characteristics so the pairs can be drawn on a flat chart. The
real analysis uses ten, and does the same thing in ten dimensions.

Frame 1: the triangles all sit on the left, below an index of 58 (that was the
eligibility rule), and the circles spread across the whole range. The two
groups are different crowds. Frame 2: each triangle is joined to its nearest
circle. Some circles are the twin of more than one triangle: that is allowed.
Frame 3: circles that nobody picked fade out. They are set aside, and the
comparison is now only between the pairs.
:::

# 2 · Implementing matching in R

## Remember: what matching needs from the data

::: {.panel}
1. **An outcome**, measured the same way for everyone
2. **A group that got the programme and a pool that did not**, and a variable
   saying which is which
3. **Characteristics measured before the programme**, never after it
4. **The characteristics that drove who took part**, and that also affect the
   outcome
5. **Enough untreated units that look like the treated ones**, so every
   treated unit has a real twin
:::

::: {.ask}
Same habit as the last two sessions: ask AI *what data do I need for matching,
and does my dataset have it?* Ask it to question you about how people ended up
in the programme before it answers.
:::

::: {.notes}
Items 3 and 4 are the ones that matter most. Item 3: a characteristic measured
after the programme may itself have been changed by the programme, so matching
on it builds part of the effect into the comparison. Item 4 is the one Section 3
turns on: matching can only use what is on the list. Item 5 is the common
support check two slides on, and it is what goes wrong in Case 2.
:::

## Does GreenWaste have what we need?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
biz %>%
  group_by(enrolled) %>%
  summarise(businesses  = n(),
            cost_before = round(mean(y0)),
            cost_after  = round(mean(y1)),
            efficiency  = round(mean(efficiency_index), 1),
            staff       = round(mean(staff_size), 1))
```

:::::

::::: {.column width="44%"}
::: {.panel}
✔ **Outcome, group:** `y1` costs after; `enrolled` 1 or 0

✔ **Before the programme:** ten round-0 characteristics, incl. the efficiency
index

✔ **A big pool:** `r gfmt(n_rest)` possible twins for `r gfmt(n_took)`
enrolled
:::

::: {.warn .fragment}
The enrolled were already cheaper **before** the programme. The raw gap after,
**`r gfmt(raw_gap)` AED**, mixes the programme with who they already were.
:::
:::::

::::::

::: {.notes}
Press Run. `biz` has one row per business: its characteristics at baseline and
its costs before (`y0`) and after (`y1`).

Read the two rows. Enrolled businesses: `r gfmt(raw_t0)` AED before and
`r gfmt(raw_t1)` after. Everyone else: `r gfmt(raw_c0)` before and
`r gfmt(raw_c1)` after. Do not dwell on the raw gap: the room met this problem
on Day 1 Session 3. The point is only that the two groups started in different
places, so there is work for matching to do.

The pool of "not enrolled" includes businesses above the cut-off and eligible
businesses in neighbourhoods that were never offered the programme. Matching
will pick the lookalikes out of it.
:::

## Implementing matching: four steps

:::::: {.columns}

::::: {.column width="56%"}
```{r}
#| echo: true
#| eval: false
# Step 1. A score: how likely was each business
#         to enrol, given what we knew at baseline?
fit <- glm(enrolled ~ efficiency_index + age_manager +
             educ_manager + female_manager + foreign_owned +
             staff_size + advanced_filtration +
             water_treatment_system + business_area +
             recycling_center_distance,
           data = biz, family = binomial)

# Step 2. Each enrolled business takes the
#         not-enrolled business with the closest score.
# Step 3. Check the pairs are alike (balance).
# Step 4. Compare their costs after the programme.
lm_robust(y1 ~ enrolled, data = matched,
          clusters = neighborhood_identifier)
```
:::::

::::: {.column width="44%"}
::: {.panel}
- **Step 1** squeezes ten characteristics into one number between 0 and 1,
  the **likelihood of enrolling**
- **Step 2** is your twins exercise, done `r gfmt(n_took)` times
- **Step 3** is the only evidence the pairs are alike
- **Step 4** is the comparison you wanted all along
:::

::: {.warn}
The **list in Step 1** is the analyst's choice. Section 3 tests it.
:::
:::::

::::::

::: {.notes}
Do not dwell on syntax. The one idea that matters: matching is preparation. It
builds the comparison group first, and only then do we compare. The regression
in Step 4 is deliberately plain: enrolled against not, on the matched pairs.

If asked why a score and not all ten characteristics at once: finding a twin
that is close on ten things at once is hard; the score says how alike two
businesses are in the one respect that matters here, how likely each was to
enrol.
:::

## Steps 1 and 2, live

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
# Step 1. The likelihood-of-enrolling score
fit <- glm(f_all, data = biz, family = binomial)
biz$score <- fitted(fit)

# Step 2. Each enrolled business takes the
#         closest score among the rest
took <- filter(biz, enrolled == 1)
rest <- filter(biz, enrolled == 0)
twin <- sapply(took$score, function(s)
          which.min(abs(s - rest$score)))
matched <- bind_rows(took, rest[twin, ])

c(enrolled   = nrow(took),
  twins_used = length(unique(twin)))
```
:::::

::::: {.column width="42%"}
::: {.ask}
Press **Run**. Every enrolled business gets a twin. How many different
businesses did it take?
:::

::: {.panel .fragment}
**`r gfmt(twins_used)` twins for `r gfmt(n_took)` enrolled businesses.** Some
not-enrolled businesses are the closest match for several enrolled ones, and
are used more than once.
:::

::: {.warn .fragment}
The other `r gfmt(n_rest - twins_used)` not-enrolled businesses are set aside.
Business 4, `r gfmt(n_rest - twins_used)` times over.
:::
:::::

::::::

::: {.notes}
`f_all` is the list of ten characteristics from the previous slide. The
`sapply` line is the whole of matching: for each enrolled business's score,
find the position of the closest score in the pool.

Reusing a twin is allowed (analysts call it matching "with replacement"). It
keeps every pair as close as possible, at the cost of leaning harder on the
businesses that get reused. Worth knowing the phrase if an evaluator uses it.
:::

## Did every enrolled business find a real lookalike?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 2.8
biz %>%
  mutate(group = ifelse(enrolled == 1,
                        "Enrolled", "Not enrolled")) %>%
  ggplot(aes(score, fill = group)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = grp_cols) +
  labs(x = "Likelihood of enrolling (the score)",
       y = "Businesses") +
  theme_m2()
```
:::::

::::: {.column width="44%"}
::: {.panel}
A twin can only come from where the curves **overlap**. Out past every
not-enrolled business, there is no real lookalike, only the least bad one.
:::

::: {.panel .fragment}
Here the ranges overlap almost completely. Only
**`r n_beyond` of `r gfmt(n_took)`** enrolled businesses scores beyond the pool.
:::

::: {.warn .fragment}
**Drop many, and** the result is no longer about the programme, only its
easy-to-match part. Case 2 shows this.
:::
:::::

::::::

::: {.notes}
Press Run. This is the picture that decides whether matching is even possible.
The enrolled curve sits to the right (they were likely to enrol, and did), the
not-enrolled curve to the left, but there are not-enrolled businesses right
across the enrolled range, thinly at the top end.

Ask: if an enrolled business had no counterpart and we dropped it, who would
the result still be about? Only the ones that had a match. Hold that for Case 2.
:::

## Did matching work? Balance

```{r fig.height=3.5, fig.width=11.5}
bal_long <- bal1 %>%
  mutate(Characteristic = fct_reorder(Characteristic, abs(Before))) %>%
  pivot_longer(c(Before, After), names_to = "stage", values_to = "smd") %>%
  mutate(stage = factor(stage, levels = c("Before", "After")))

plot_ly() %>%
  add_trace(data = filter(bal_long, stage == "Before"),
            x = ~abs(smd), y = ~Characteristic, type = "scatter", mode = "markers",
            name = "Before matching",
            marker = list(color = ACCENT2, size = 12),
            hovertemplate = "%{y}: %{x:.2f} before<extra></extra>") %>%
  add_trace(data = filter(bal_long, stage == "After"),
            x = ~abs(smd), y = ~Characteristic, type = "scatter", mode = "markers",
            name = "After matching",
            marker = list(color = ACCENT, size = 12, symbol = "diamond"),
            hovertemplate = "%{y}: %{x:.3f} after<extra></extra>") %>%
  layout(xaxis = ax("How far apart the groups are (standardised difference)",
                    range = c(0, 1.1), zeroline = FALSE),
         yaxis = ax("", tickfont = list(size = 12, color = INK)),
         shapes = list(list(type = "line", x0 = 0.1, x1 = 0.1, y0 = 0, y1 = 1,
                            yref = "paper", line = list(color = ALERT, dash = "dash", width = 1.5))),
         annotations = list(list(x = 0.1, y = 1.02, yref = "paper", xanchor = "left",
                                 text = " 0.1: the usual line", showarrow = FALSE,
                                 font = list(size = 11, color = ALERT)))) %>%
  pl_m2(height = 350, t = 36, b = 80, legend_y = -0.42)
```

::: {.ask}
Each row is one characteristic. Hover to read the values. What happened to the
efficiency index?
:::

::: {.notes}
Orange dots: how far apart enrolled and not-enrolled businesses were before
matching. Blue diamonds: how far apart the matched pairs are. Beyond about 0.1
is usually treated as a problem.

The efficiency index was by far the worst (`r sprintf("%.2f", abs(bal1$Before[bal1$var == "efficiency_index"]))`),
because it decided eligibility. After matching, every characteristic is under
`r sprintf("%.2f", max(abs(bal1$After)))`.

Say it plainly: balance means alike **on the list**. It says nothing about
anything that is not on the list. That is Section 3.
:::

## Step 4: running the comparison

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
out <- lm_robust(y1 ~ enrolled, data = matched,
                 clusters = neighborhood_identifier)

res <- tidy(out)[, c("term", "estimate", "std.error",
                     "conf.low", "conf.high")]
res[, -1] <- round(res[, -1])
res
```
:::::

::::: {.column width="42%"}
::: {.ask}
Press **Run**. Two rows. Which one is the programme's effect?
:::

::: {.panel .fragment}
**Row 2, `enrolled`: `r gfmt(est1)` AED.** The cost difference between each
enrolled business and its twin, averaged.
:::

::: {.warn .fragment}
**Trap:** the intercept, `r gfmt(coef(fit1)[["(Intercept)"]])` AED, is the
twins' average cost, not an effect.
:::
:::::

::::::

::: {.notes}
Run the chunk live, then take a vote before clicking. Most will get it; the
point is the habit of naming the row.

Reminder of the vocabulary (Day 1 Session 1): the **coefficient** is the effect
size in AED; the **confidence interval** (conf.low to conf.high) is the range of
plausible true values; clustering by neighbourhood means businesses in the same
neighbourhood are not treated as independent evidence.

Close the loop: this is the twins card, done for every enrolled business at
once, with a margin of error attached.
:::

## Does it clear the rule?

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the matched estimate (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">Before matching</span></div>
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
</div>', gfmt(est1), gfmt(ci1[1]), gfmt(ci1[2]),
   gfmt(raw_gap), gfmt(RULEBAR), gfmt(avoid1[1])))
```

::: {.panel}
Matching took `r gfmt(abs(raw_gap) - abs(est1))` AED off the raw gap. That part
was **who the businesses already were**, not the programme.
:::

::: {.ask .fragment}
The estimate sits **right on the bar**, and the interval runs from
`r gfmt(avoid1[1])` to `r gfmt(avoid1[2])` AED. We cannot show it clears
`r gfmt(RULEBAR)`. A result this close to the line is decided by the analyst's
choices, which is exactly what we test next.
:::

::: {.notes}
Same reading as the last two sessions: not "is it significant?" but "does my
decision change across this range?". It does: the optimistic end clears the
bar, the pessimistic end does not.

Worth noting if asked: the randomised comparison on Day 1 gave -1,014. Matching
has landed close to it, which is reassuring about the method here. But the
commissioner deciding from a matching study alone would not know that.
:::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: what matching cannot see, and how to
check how much that matters.
:::

# 3 · The one question, and how to check it

## What matching can and cannot see

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.0em;">Matching can only account for what you measured.</div>
</div>')
```

::: {.panel}
Every characteristic in that balance chart was **collected**. Matching cannot
balance on management quality, on how motivated the owner was, on whether the
business was about to invest anyway, unless somebody wrote it down.
:::

::: {.warn}
So the accurate claim for this method is: **we have made the two groups alike in
every way we can observe.** The words doing the work are the last four.
:::

::: {.notes}
DiD rested on parallel trends; RDD on a clean cut-off. Matching rests on
something plainer and harder: everything that drove enrolment and also affects
costs was measured and matched on. This cannot be tested from the data, by
definition. But we can see how much it would matter, which is the next slide.
:::

## Check 1: what if we had never measured the index?

:::::: {.columns}

::::: {.column width="58%"}
```{webr}
# Re-run Steps 1 and 2, pretending the
# efficiency index was never collected
biz$score2 <- fitted(glm(f_no_index, data = biz,
                         family = binomial))
m2 <- find_twins(biz, "score2")

m2 %>%
  group_by(enrolled) %>%
  summarise(cost_after = round(mean(y1)),
            index      = round(mean(efficiency_index), 1),
            staff      = round(mean(staff_size), 1),
            manager    = round(mean(age_manager), 1))
```
:::::

::::: {.column width="42%"}
::: {.ask}
Press **Run**. The nine remaining characteristics still balance. Look at the
index column.
:::

::: {.panel .fragment}
The pairs are now **`r sprintf("%.1f", abs(s_noindex$index_gap))` index
points apart**, and the estimate moves to **`r gfmt(s_noindex$est)` AED**,
across the bar.
:::

::: {.warn .fragment}
The balance check on the nine would still look perfect (worst
`r sprintf("%.2f", s_noindex$worst_other)`). **Nothing in the output would
warn you.** That is what an unmeasured difference looks like.
:::
:::::

::::::

::: {.notes}
This is the session's key demonstration. The efficiency index is the thing
that decided who was eligible, so it is exactly the kind of characteristic that
drives both enrolment and costs. Here we have it, so we can see what leaving it
out does. In a real evaluation, the missing characteristic is one nobody
collected, so you never get to see this comparison.

The estimate moves from `r gfmt(est1)` to `r gfmt(s_noindex$est)`. Not a huge
move, but enough to push the point estimate over the bar. The direction flatters
the programme, which is the usual direction for an omitted "motivated and
already improving" characteristic.
:::

## Check 2: change the recipe, see what moves

```{r fig.height=3.1, fig.width=11.5}
sens_plot <- sens_tbl %>% mutate(spec = factor(spec, levels = rev(spec)))
plot_ly(sens_plot, x = ~est, y = ~spec, type = "scatter", mode = "markers+text",
        error_x = list(type = "data", symmetric = FALSE,
                       array = ~hi - est, arrayminus = ~est - lo,
                       color = ACCENT, thickness = 2.5, width = 0),
        marker = list(color = ACCENT, size = 13),
        text = ~gfmt(est), textposition = "bottom center",
        textfont = list(size = 13, color = INK),
        hovertemplate = "%{y}: %{x:,.0f} AED<extra></extra>", showlegend = FALSE) %>%
  layout(xaxis = ax("Estimated effect on waste costs (AED)", range = c(-1220, -860),
                    zeroline = FALSE),
         yaxis = ax("", tickfont = list(size = 13, color = INK)),
         shapes = list(list(type = "line", x0 = -RULEBAR, x1 = -RULEBAR, y0 = 0, y1 = 1,
                            yref = "paper", line = list(color = ALERT, dash = "dash", width = 2))),
         annotations = list(list(x = -RULEBAR, y = 1.0, yref = "paper",
                                 yanchor = "bottom", yshift = 14,
                                 text = "the bar: -1,000", showarrow = FALSE,
                                 font = list(size = 12, color = ALERT)))) %>%
  pl_m2(height = 320, t = 46, b = 60)
```

::: {.warn}
Same businesses, same method, four defensible choices. The estimates land on
**both sides of the bar**, and every interval crosses it.
:::

::: {.ask .fragment}
Would you sign off a national scale-up on any single one of these rows?
:::

::: {.notes}
Row 1 is the main estimate. Row 2 compares each pair's change in costs rather
than their costs afterwards (the difference-in-differences idea from this
morning, applied to the matched pairs): a useful second opinion when the pairs
might have started at slightly different levels. It agrees. Rows 3 and 4 match
on less: without the index, and on manager age and education alone.

The takeaway for a commissioner: with matching, ask what was matched on, and
whether the answer holds when the list changes. If one choice of list flips the
recommendation, the result is not settled. Here, it flips.
:::

## The question to put to any evaluator

:::::: {.columns}

::::: {.column width="52%"}
::: {.ask style="font-size: 118%;"}
For each business that enrolled, you found one that did not but looked the same.

**What else might be different about those two businesses that you could not
measure?**
:::

It cannot be answered from the table. That is the point of asking it.
:::::

::::: {.column width="48%"}
::: {.panel}
For GreenWaste, the plausible answers are things like:

**How badly did this business want to cut costs?** Businesses that enrol are
often the ones already looking for a solution.

**Was the owner about to invest anyway?** If so, we are crediting the programme
with a decision already made.
:::

::: {.warn}
Neither of these is in the data. Both would make the programme look better than
it was, just as leaving out the index did.
:::
:::::

::::::

::: {.notes}
Get the room to generate the hidden differences before revealing the panel.
Owner motivation always comes up and it is the perfect example: unmeasured,
plausible, and biased in the direction that flatters the programme.
:::

# 4 · Case 2: when the lookalikes run out

## The cameras went where the roads were worst

:::::: {.columns}

::::: {.column width="52%"}
```{r}
d2 %>%
  group_by(`Has a camera` = ifelse(cameras_installed == 1, "Yes", "No")) %>%
  summarise(Segments = n(),
            `85th pct speed` = round(mean(baseline_speed_85th), 1),
            `Collisions, baseline` = round(mean(y0), 2),
            `Collisions, prior 3 yrs` = round(mean(prior_collisions_3yr), 1),
            .groups = "drop") %>%
  kable(align = "lrrrr", digits = 2) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```

::: {.warn .note-ctr}
**Illustrative case, with invented figures.** The roads-authority data from
Day 1; not a real programme.
:::
:::::

::::: {.column width="48%"}
::: {.panel}
You met these cameras on Day 1. The camera segments were **faster** and had
**more collisions** before anything was installed.

In GreenWaste, businesses opted in. Here the authority chose the worst roads on
purpose.
:::

::: {.ask}
Same four steps, same checks. Rule: at least **`r sprintf("%.1f", CAM_RULE)`
fewer collisions** per segment per round. Watch which check fails.
:::
:::::

::::::

::: {.notes}
Section 4 is the second case study the outline asks for: a matching study that
is carried out correctly and still only half works. It stands in for the
"imperfect match" DCD case until those findings arrive. Move at pace: the room
now knows the checks, so this section is about spotting which one fails.
:::

## Check: is there a lookalike for every camera segment?

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.9, fig.width=5.6}
plot_ly() %>%
  add_trace(data = d2 %>% filter(cameras_installed == 0),
            x = ~score, type = "histogram", name = "No camera",
            marker = list(color = ACCENT2, line = list(width = 0)),
            opacity = 0.65, xbins = list(start = 0, end = 1, size = 0.025),
            hovertemplate = "score %{x:.2f}: %{y} segments<extra></extra>") %>%
  add_trace(data = d2 %>% filter(cameras_installed == 1),
            x = ~score, type = "histogram", name = "Has a camera",
            marker = list(color = ACCENT, line = list(width = 0)),
            opacity = 0.65, xbins = list(start = 0, end = 1, size = 0.025),
            hovertemplate = "score %{x:.2f}: %{y} segments<extra></extra>") %>%
  layout(barmode = "overlay",
         xaxis = ax("Likelihood of getting a camera (the score)", zeroline = FALSE),
         yaxis = ax("Segments")) %>%
  pl_m2(width = PL_WC, height = 275)
```
:::::

::::: {.column width="48%"}
::: {.panel}
Camera segments pile up at the **high** end of the score, just where
no-camera segments thin out.
:::

::: {.warn .fragment}
**`r c2_outside` camera segments (`r round(c2_pct_out)`%) score higher than any
segment without a camera.** They have no real lookalike, yet the software still
paired them, with the least bad one.
:::

::: {.warn .fragment}
And `r n_cam` camera segments share just **`r c2_used` twins**. A handful of
roads are doing most of the comparing.
:::
:::::

::::::

::: {.notes}
Contrast with GreenWaste, where the curves overlapped almost completely and
`r n_beyond` enrolled businesses sat outside. Here `r c2_outside` do. Nothing in
a standard results table flags them: the match rate still reads 100%. The
two reasonable options are to drop them (and say the result no longer covers the worst
roads) or to keep them and say the comparison for them is weak. Either way, the
evaluator should tell you which.
:::

## Check: balanced, but not level

:::::: {.columns}

::::: {.column width="52%"}
```{r}
c2_bal %>%
  mutate(across(c(Before, After), ~ sprintf("%.2f", .x))) %>%
  kable(align = "lrr", col.names = c("Characteristic", "Before", "After")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  row_spec(nrow(c2_bal), bold = TRUE, background = "#F7E9E6") %>%
  footnote(general = "Standardised differences. Beyond about 0.1 is usually a problem.",
           general_title = "", footnote_as_chunk = TRUE)
```
:::::

::::: {.column width="48%"}
::: {.panel}
Every characteristic in the model balances. The **outcome itself**, measured
before the cameras went in, does not:
**`r sprintf("%.2f", c2_base_t)`** collisions on camera segments against
**`r sprintf("%.2f", c2_base_c)`** on their twins.
:::

::: {.warn .fragment}
Matching balanced what the model **contained**. Collisions in the baseline year
were not in it, so they were free to differ, and they sit on the wrong side of
0.1.
:::
:::::

::::::

::: {.notes}
The model included collisions over the prior three years, and that balances.
But the baseline-year count, the outcome's own starting point, still differs by
`r sprintf("%.2f", c2_base_t - c2_base_c)` collisions. Ask to see the outcome's
baseline in any balance table: it is often left out.
:::

## Three numbers from the same pairs

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.2f</div>
  <div class="label">collisions avoided per segment, comparing changes (95%% CI: %.2f to %.2f)</div>
</div>
<div class="stat-row">
  <div><span class="n">%.2f</span><span class="k">Comparing follow-up levels</span></div>
  <div><span class="n">%.2f</span><span class="k">Plain DiD, no matching</span></div>
  <div><span class="n">%.1f</span><span class="k">The rule</span></div>
</div>', c2_changes, min(c2_ci), max(c2_ci), c2_levels, c2_did, CAM_RULE))
```

::: {.ask}
**Levels** says `r sprintf("%.2f", c2_levels)` avoided and fails the rule;
**changes** says `r sprintf("%.2f", c2_changes)` and clears it. Which do you
believe?
:::

::: {.panel .fragment}
**Changes.** The pairs started `r sprintf("%.2f", c2_base_t - c2_base_c)`
collisions apart, and changes remove that. It lands near the plain DiD. **But**
the least generous end, `r sprintf("%.2f", min(c2_ci))`, is short of
`r sprintf("%.1f", CAM_RULE)`.
:::

::: {.notes}
Two methods that should agree, agreeing, is the reassurance: the changes-based
matched estimate and the plain DiD from this morning's logic. The levels
comparison was the odd one out because the pairs did not start level. Note the
same check was run on GreenWaste (Check 2, row 2) and there the two agreed,
because the pairs did start level.
:::

## The two cases, side by side

```{r}
compare_tbl %>%
  kable(align = "l", col.names = c("", "Case 1 · GreenWaste", "Case 2 · Cameras")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(c(1, 4), background = "#F7E9E6")
```

::: {.ask}
Read down the table. Case 1 passes every check. Case 2 passes the checks people
usually report, and fails the two they usually do not.
:::

::: {.notes}
Rows 1 and 4 are the two that separate the cases: real lookalikes for everyone,
and the outcome's own baseline. Neither appears in a typical matching results
table, which is why they are on the question card at the end.
:::

# 5 · Group exercise: which would you trust?

## Which case would you trust?

::: {.ask}
In your groups, ten minutes:

1. Which case gives you **more confidence**, and why?
2. What would you ask **Case 2's** evaluator that Case 1 does not raise?
3. Case 2 is the only one whose central estimate clears its rule. Does that
   change your answer?
:::

::: {.panel .fragment}
The uncomfortable answer: **Case 1** is the more trustworthy study, and it
cannot show it clears the rule. **Case 2** gives the more attractive answer,
and it carries the unquantified risk.


:::

::: {.notes}
Verdict round at the end: hands, not discussion. Push them past "Case 1 is
cleaner" to why: overlap, the outcome's baseline, and who the result applies
to. Ask whoever says "send it back" for the single question behind it.
:::

## What to ask the evaluator of Case 2

```{=html}
<div class="facts">
  <div class="k">Ask</div> <div class="v">"What happened to the segments with no real lookalike, and does the answer change without them?"</div>
  <div class="k">Ask</div> <div class="v">"Did the outcome's own baseline balance, and did you compare levels or changes?"</div>
  <div class="k">Ask</div> <div class="v">"What did you match on, and what was left out?"</div>
  <div class="k">Ask</div> <div class="v">"What else might differ between matched pairs that you did not measure?"</div>
</div>
```

::: {.ask}
None of these needs you to run anything. Each of them would change how much
weight you give the headline.

**Question two is the sharpest**, because it catches a real problem that a
reported balance table would not reveal.
:::

# 6 · Put AI to work

## AI Snapshot: comparing two studies

::: {.panel}
**Prompt, with the two-case table and both balance tables pasted in:**
"Summarise the key differences between these two studies and identify what a
commissioner should flag."
:::

::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

**Case 1:** Matching was highly effective, with every business matched and
excellent covariate balance. The programme reduced costs by around 1,000 AED,
with an interval that straddles the 1,000 AED threshold.

**Case 2:** [Matching was similarly strong, with every segment matched and good
balance]{.flag}. [The matched estimate shows cameras avoided 2.10 collisions per
segment, comfortably exceeding the 2.0 threshold]{.flag},
**[so this programme should be scaled]{.flag}**.
:::

::: {.ask}
Which sentences would you sign your name to? Four minutes, in groups.
:::

## Debrief: one clean read, three false assurances

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What it got right.** Case 1 is reported accurately, including the awkward
part: the interval straddles the rule.
:::

::: {.warn}
**1 · "Similarly strong" in Case 2.** It was not. `r c2_outside` segments had
no real lookalike, and "every segment matched" hides that.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**2 · "Comfortably exceeding."** Only the changes-based estimate clears 2.0;
the levels estimate does not, and the interval dips to
`r sprintf("%.2f", min(c2_ci))`.
:::

::: {.warn}
**3 · "So this programme should be scaled."** The recommendation rests entirely
on sentences 1 and 2, and it is stated most confidently of all.
:::
:::::

::::::

::: {.panel}
**The shape of this failure is new.** The model did not invent a number. It
repeated the study's **own summary of itself**, and the summary was the
misleading part. You have to compare the study's claims against the study's
details, not against the world.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Summarise the key differences between these two studies and identify what a
> commissioner should flag.

The model may summarise the tables it was given and repeat their framing.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Two matching studies. For each: how many treated units had **no real
> lookalike**; did the **outcome's baseline** balance; do levels and changes
> agree? Our decision rules are below. **Ask me questions before you
> answer.**
:::
:::::

::::::

::: {.ask}
Run both, then compare: what did each get wrong, and what did each explain
well? Then paste the second answer into a **fresh chat** and
ask it to check for overstatement.
:::

# 7 · The Abu Dhabi case

## `[PLACEHOLDER]` The DCD matching evaluation

::: {.warn}
**Awaiting the real findings.** The second case study in the outline is an Abu
Dhabi Department of Community Development matching study, on real-world data
where some participants had no good match. It has not been shared with this
workspace, so nothing here is presented as its result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>      <div class="v">[PLACEHOLDER]</div>
  <div class="k">Design</div>         <div class="v">Matching on observables</div>
  <div class="k">Match rate</div>     <div class="v">[PLACEHOLDER]</div>
  <div class="k">Outcome</div>        <div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed, and it is the same structure as Case 2 above. When the
DCD findings arrive they drop into these fields without redesigning the session.
:::

## The five questions for the DCD result

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">What was matched on, and what was left out of the model?</div>
  <div class="k">2</div> <div class="v">How many treated units had no real lookalike, and what happened to them?</div>
  <div class="k">3</div> <div class="v">Do the covariates balance after matching, and does the outcome's own baseline balance?</div>
  <div class="k">4</div> <div class="v">Do the levels-based and change-based estimates agree?</div>
  <div class="k">5</div> <div class="v">What else might differ between matched pairs that was never measured?</div>
</div>
```

::: {.ask}
Same five questions we just worked through on the camera case. The programme
changes; the interrogation does not.
:::

# 8 · Close

## Your turn: ask AI first

::: {.ask style="font-size: 110%;"}
Before we show our list of questions to ask someone presenting matching results, ask an AI tool:

**"I am a commissioner about to hear an evaluator present a matching result. What are the key questions I should ask?"**
:::

::: {.panel}
Three minutes. Then compare with our list on the next slide: what did it miss, what did it add, and is anything incorrect?
:::

::: {.notes}
OPTIONAL. If time is short, skip this slide and go straight to the questions
on the next one. If you run it: give three minutes, then reveal the list and
ask what the AI missed, what it added that is worth keeping, and whether
anything it suggested was incorrect.
:::

## Five questions to ask any evaluator presenting matching

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">What did you match on, and what was left out?</div>
  <div class="k">2</div> <div class="v">How many treated units had no real lookalike, and what happened to them?</div>
  <div class="k">3</div> <div class="v">Do the covariates balance, and does the outcome's baseline balance?</div>
  <div class="k">4</div> <div class="v">Does the answer hold if you change what you matched on, or compare changes instead of levels?</div>
  <div class="k">5</div> <div class="v">What else might differ that you could not measure?</div>
</div>
```

::: {.panel}
Question 5 is the one that matters most, and it is the only one the evaluator
cannot answer from their own output.

That should tell you something about how much weight a matching study can carry
on its own.
:::

## The week so far

```{=html}
<div class="facts">
  <div class="k">Day 1 S2</div> <div class="v">descriptive reading &middot; -1.13 against a 2.0 rule</div>
  <div class="k">Day 1 S3</div> <div class="v">randomisation &middot; the comparison created by a lottery</div>
  <div class="k">Day 2 S1</div> <div class="v">DiD &middot; -816 against a 1,000 rule, and an assumption that could not be tested</div>
  <div class="k">Day 2 S2</div> <div class="v">RDD &middot; -249 against a 1,000 rule, and local to the line</div>
  <div class="k">Today</div>  <div class="v">matching &middot; `r gfmt(est1)` against a 1,000 rule, and only as good as the list</div>
</div>
```

::: {.ask}
Each method bought credibility, and each paid for it:

**DiD** assumes parallel trends. **RDD** narrows the answer to a group near a
line. **Matching** assumes everything that matters was measured.

Those three sentences are the whole of Day 2. Tomorrow turns from reading
results to **questioning them**: cost analysis, data visualisation, and then a
clinic where you interrogate an analyst.
:::

::: {.notes}
If there is time at the very end, ask the room: is there a different
question you would ask AI to help you understand a concept from today? For
example: "I am learning about impact evaluation to help me make policy
decisions. How should I think about a confidence interval, and how is it
relevant when I compare an estimate to a decision threshold?" Adding context
about who you are and why you are asking usually gets a more useful answer.
:::
