---
title: "Reading a Regression Discontinuity Result"
subtitle: "Module 2 · Day 2, Session 2: What Happens at the Cut-off"
author: "3ie"
always_allow_html: yes
# revealjs only in this folder: the charts are interactive (plotly/htmlwidgets),
# and pandoc cannot carry an htmlwidget into pptx. Export to PDF for a portable copy.
format:
  revealjs:
    theme: [theme_editorial.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate. knitr sizes every htmlwidget container as
    # fig.width * dpi, so an inherited dpi: 200 silently inflated each chart
    # tenfold (11.5in became 2300px) and pushed slide content off the bottom.
    # At 96, fig.width in inches maps 1:1 onto CSS pixels. If a raster
    # ggplot chunk is ever added, set its dpi per chunk.
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(qrcode)

# ---- Palette (editorial theme; see theme_editorial.scss) -------------------
ACCENT  <- "#0E6E80"   # series 1: programme / eligible side
ACCENT2 <- "#A9561F"   # series 2: comparison / other side
ALERT   <- "#8E2413"   # decision rules and flagged errors, never a series
INK     <- "#16181A"; MUTED <- "#61686D"; RULE <- "#E3E6E8"
CHART_FONT <- "Segoe UI, Helvetica, Arial, sans-serif"

PL_W  <- 1100
PL_WC <- 520
PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000     # must cut costs by at least this many AED

# ---- The GreenWaste eligibility rule ---------------------------------------
# The programme was offered only to businesses whose efficiency index was at or
# BELOW 58. Read that again: the LESS efficient businesses qualified. This is
# the opposite direction to a "higher score wins" rule, and it is the first
# thing to get right when reading the chart.
#
# MEASURED, not assumed: the maximum index among eligible businesses is 57.9598
# and the minimum among ineligible is 58.0371, so the rule is genuinely sharp
# and the cut-off sits in between. Verified agreement: 1.000000.
gw <- read.csv("./evaluation_data_GreenWaste.csv")
CUT <- 58

# ---- Is the rule sharp, and is take-up full? -------------------------------
# The OFFER is a clean function of the index. TAKE-UP is not: only about half of
# the eligible businesses enrolled. That makes this a FUZZY design, and it is
# the distinction that decides whether the headline number is the effect on
# those offered or the effect on those who took it.
gw1 <- gw %>% filter(round == 1)

rule_sharp <- mean(as.integer(gw$efficiency_index <= CUT) == gw$eligible,
                   na.rm = TRUE)

# ---- The local comparison, at a chosen bandwidth ---------------------------
# Everything below is done at +/-2 index points. That is a teaching choice: wide
# enough to have power, narrow enough that the two sides are genuinely alike.
# The deck shows what happens when the bandwidth changes, because that choice
# is the one a commissioner can most easily be misled by.
BAND <- 2

win <- gw1 %>%
  filter(abs(efficiency_index - CUT) <= BAND) %>%
  mutate(below   = as.integer(efficiency_index <= CUT),   # the eligible side
         x       = efficiency_index - CUT)

rdd_fit <- lm_robust(waste_management_costs ~ below * x, data = win)
itt     <- coef(rdd_fit)[["below"]]
itt_ci  <- confint(rdd_fit)["below", ]
itt_p   <- rdd_fit$p.value[["below"]]

# Take-up difference at the cut-off: this is what turns the ITT into the effect
# on those who actually took the programme.
takeup <- mean(win$enrolled[win$below == 1]) - mean(win$enrolled[win$below == 0])
scaled <- itt / takeup

# The decision question. The rule asks the programme to AVOID at least RULEBAR.
# Sign trap: the coefficient is negative, so conf.low is the LARGER effect and
# abs(conf.low) is the OPTIMISTIC end. Sorting the absolute values makes the
# first element the end closest to zero, which is what the decision turns on.
itt_avoided       <- sort(abs(itt_ci))
itt_avoided_least <- itt_avoided[1]
itt_avoided_most  <- itt_avoided[2]
itt_clears        <- itt_avoided_least >= RULEBAR

# ---- What the two sides look like, either side of the line -----------------
side_means <- win %>%
  group_by(below) %>%
  summarise(n = n(), cost = mean(waste_management_costs),
            enrolled_share = mean(enrolled), .groups = "drop")

just_below <- side_means$cost[side_means$below == 1]   # eligible
just_above <- side_means$cost[side_means$below == 0]   # not eligible
raw_gap    <- just_below - just_above

# ---- Sensitivity: the estimate at other bandwidths -------------------------
# Built from the data so the slide cannot drift from the numbers.
band_tbl <- map_dfr(c(1, 2, 3, 5, 8), function(w) {
  d <- gw1 %>% filter(abs(efficiency_index - CUT) <= w) %>%
    mutate(below = as.integer(efficiency_index <= CUT), x = efficiency_index - CUT)
  f <- lm_robust(waste_management_costs ~ below * x, data = d)
  tk <- mean(d$enrolled[d$below == 1]) - mean(d$enrolled[d$below == 0])
  tibble(band = w, n = nrow(d),
         itt = coef(f)[["below"]],
         lo  = confint(f)["below", 1],
         hi  = confint(f)["below", 2],
         takeup = tk, scaled = coef(f)[["below"]] / tk)
})

# ---- Placebo cut-offs: the same test somewhere there is no rule ------------
# If the method reports a jump where no programme exists, its jump at the real
# cut-off is not evidence of anything.
placebo_tbl <- map_dfr(c(50, 54, 62, 66), function(cut) {
  d <- gw1 %>% filter(abs(efficiency_index - cut) <= BAND) %>%
    mutate(below = as.integer(efficiency_index <= cut), x = efficiency_index - cut)
  f <- lm_robust(waste_management_costs ~ below * x, data = d)
  tibble(cut = cut, n = nrow(d), jump = coef(f)[["below"]],
         lo = confint(f)["below", 1], hi = confint(f)["below", 2])
})

# ---- Density at the cut-off: did anyone bend the rule? ---------------------
dens_tbl <- tibble(
  bucket = c("56 to 57", "57 to 58", "58 to 59", "59 to 60"),
  n = c(sum(gw1$efficiency_index >= 56 & gw1$efficiency_index < 57, na.rm = TRUE),
        sum(gw1$efficiency_index >= 57 & gw1$efficiency_index < 58, na.rm = TRUE),
        sum(gw1$efficiency_index >= 58 & gw1$efficiency_index < 59, na.rm = TRUE),
        sum(gw1$efficiency_index >= 59 & gw1$efficiency_index < 60, na.rm = TRUE))
)

# ---- Covariates at the cut-off, at several bandwidths ----------------------
# A covariate that jumps at the cut-off means the two sides are not alike and
# the estimate is not credible. One of these appears to jump - but only at some
# bandwidths, which is the lesson this slide exists to teach.
cov_bands <- c(1, 2, 3, 5, 8)
cov_vars  <- c("recycling_center_distance", "staff_size", "age_manager",
               "business_area")
cov_labs  <- c(recycling_center_distance = "Distance to recycling centre (m)",
               staff_size = "Staff size",
               age_manager = "Manager age (years)",
               business_area = "Business area (units)")

cov_tbl <- map_dfr(cov_vars, function(v) {
  row <- tibble(Characteristic = cov_labs[[v]])
  for (w in cov_bands) {
    d <- gw1 %>% filter(abs(efficiency_index - CUT) <= w) %>%
      mutate(below = as.integer(efficiency_index <= CUT), x = efficiency_index - CUT)
    f <- lm_robust(as.formula(paste(v, "~ below * x")), data = d)
    sig <- confint(f)["below", 1] > 0 | confint(f)["below", 2] < 0
    row[[paste0("b", w)]] <- sprintf("%+.2f%s", coef(f)[["below"]],
                                     if (sig) "*" else "")
  }
  row
})

# ---- Who is near the line, versus everyone else ----------------------------
near_line <- gw1 %>% filter(abs(efficiency_index - CUT) <= BAND)
far_line  <- gw1 %>% filter(abs(efficiency_index - CUT) > BAND)
reach_tbl <- tibble(
  Group = c("Just either side of the cut-off", "Everybody else"),
  n = c(nrow(near_line), nrow(far_line)),
  staff = c(mean(near_line$staff_size), mean(far_line$staff_size)),
  area  = c(mean(near_line$business_area), mean(far_line$business_area)),
  cost  = c(mean(near_line$waste_management_costs),
            mean(far_line$waste_management_costs))
)

# ---- The table an evaluator would report ----------------------------------
rdd_tbl <- rdd_fit %>% tidy() %>%
  mutate(` ` = c("(Intercept)", "Just below the cut-off",
                 "Distance from the cut-off",
                 "Below x Distance"),
         CI = sprintf("[%s, %s]", gfmt(conf.low), gfmt(conf.high))) %>%
  select(` `, estimate, std.error, p.value, CI)
```

# 1 · A different way to find a comparison

## Where we are

```{=html}
<div class="facts">
  <div class="k">Day 1 S1</div> <div class="v">compared to what? &middot; the three questions</div>
  <div class="k">Day 1 S2</div> <div class="v">reading a descriptive output</div>
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons, then randomisation</div>
  <div class="k">Yesterday</div> <div class="v">difference-in-differences &middot; and an assumption that cannot be tested</div>
  <div class="k">Today</div>     <div class="v">regression discontinuity &middot; a comparison the programme created</div>
</div>
```

::: {.ask}
Yesterday's method needed a comparison group that was heading the same way, and
we could not check whether it was.

Today's method does not need you to believe anything about a comparison group.
It builds one out of a rule.
:::

## The rule that decided who got in

:::::: {.columns}

::::: {.column width="54%"}
GreenWaste was not offered to everyone. It went to businesses whose
**efficiency index** was **at or below 58** — the less efficient ones.

Two businesses either side of that line are almost identical. One got the
programme, one did not. That is the whole idea.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Running variable</div> <div class="v">Efficiency index</div>
  <div class="k">Cut-off</div>         <div class="v">58</div>
  <div class="k">Eligible</div>        <div class="v">at or below 58</div>
  <div class="k">The rule is sharp?</div> <div class="v">yes, verified</div>
</div>
```
:::::

::::::

::: {.warn}
**Read the direction carefully.** The rule offers the programme to the **less**
efficient businesses. A "higher is better" rule would put the eligible side on
the right. Getting this backwards inverts every number that follows.
:::

## `[MENTI PLACEHOLDER]` Which side is which?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 120%;"}
Businesses with an efficiency index **at or below 58** were offered the
programme.

On the chart we are about to draw, is the eligible side on the **left** or the
**right**?
:::

Left &nbsp;·&nbsp; Right &nbsp;·&nbsp; Depends how the axis is drawn

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel}
**Left.** A low index is on the left of a normal axis, and low is eligible.

Everyone who answered "right" has just demonstrated exactly the error this
session is about: a reader who assumes the programme went to the *good*
businesses will misread the entire chart, and the estimate's sign with it.
:::

## The picture the method is built on

```{r fig.height=3.2, fig.width=11.5}
# Binned means either side of the cut-off, plus the fitted lines.
binned <- gw1 %>%
  filter(between(efficiency_index, CUT - 10, CUT + 10)) %>%
  mutate(side = if_else(efficiency_index <= CUT, "Eligible", "Not eligible")) %>%
  group_by(side, bucket = round(efficiency_index)) %>%
  summarise(cost = mean(waste_management_costs),
            n = n(), .groups = "drop")

plot_ly() %>%
  add_trace(data = binned, x = ~bucket, y = ~cost, color = ~side,
            type = "scatter", mode = "markers",
            colors = c("Eligible" = ACCENT, "Not eligible" = ACCENT2),
            marker = list(size = ~pmax(8, pmin(18, n / 12))),
            text = ~sprintf("%d businesses", n),
            hovertemplate = "%{x} index: %{y:,.0f} AED<br>%{text}<extra></extra>") %>%
  add_segments(x = CUT, xend = CUT, y = 0, yend = ~max(cost),
               line = list(color = ALERT, width = 3, dash = "dot"),
               showlegend = FALSE, hoverinfo = "skip") %>%
  layout(xaxis = ax("Efficiency index  (eligible is at or below 58)",
                    zeroline = FALSE),
         yaxis = ax("Waste costs (AED)", range = c(0, 2600))) %>%
  pl_m2(height = 310)
```

::: {.ask}
Hover the dots. Is there a step at the dotted line — the cut-off at 58?

Markers are sized by how many businesses sit in each bucket, so you can see
where the evidence is thicker.
:::

# 2 · The estimate, and how to read it

## The comparison the rule creates

:::::: {.columns}

::::: {.column width="48%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, eligible side vs just above the line</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">At or below 58</span></div>
  <div><span class="n">%s</span><span class="k">Just above 58</span></div>
  <div><span class="n">%d</span><span class="k">Within 2 points</span></div>
</div>', gfmt(raw_gap), gfmt(just_below), gfmt(just_above), nrow(win)))
```
:::::

::::: {.column width="52%"}
::: {.panel}
The raw difference, within two index points of the line, is
`r gfmt(abs(raw_gap))` AED. Both sides are businesses that look much the same:
same staff size, same area, same neighbourhoods.
:::

::: {.warn}
That raw difference is the **starting point**, not the answer. It still contains
any slope in the outcome as the index rises, which is what the regression
removes.
:::
:::::

::::::

## The table an evaluator would report

:::::: {.columns}

::::: {.column width="52%"}
```{r}
rdd_tbl %>%
  mutate(across(c(estimate, std.error), ~ round(.x, 0)),
         p.value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("", "Coef.", "Std. err.", "p", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 15) %>%
  row_spec(2, bold = TRUE, background = "#E6F1F5")
```
:::::

::::: {.column width="48%"}
::: {.panel}
**Row 2 is the jump.** Everything else is housekeeping: the intercept is where
the line sits, and the two slope rows describe how costs drift as the index
moves.

The jump is `r gfmt(itt)` AED, interval
[`r gfmt(itt_ci[1])`, `r gfmt(itt_ci[2])`].
:::

::: {.warn}
**The estimate is local.** It is what the programme did for businesses within
two index points of 58. It is not the average effect across every business in
the programme.
:::
:::::

::::::

## Does it clear the rule?

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED, the estimated cut in costs near the cut-off</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">No</span><span class="k">Clears the bar</span></div>
</div>', gfmt(abs(itt)), gfmt(RULEBAR), gfmt(itt_avoided_least)))
```

::: {.warn .fragment}
The estimate is clearly not zero. `r gfmt(abs(itt))` AED is a real effect.

The rule asks for `r gfmt(RULEBAR)`. The interval's most generous end reaches
`r gfmt(itt_avoided_most)`, and its least generous end `r gfmt(itt_avoided_least)`.
**It does not clear the bar.**
:::

::: {.ask}
Fourth session, same verdict. Day 1 S2, Day 1 S3, yesterday's DiD, and now this:
every result has been **real**, and none has been **enough**.

That is not a coincidence in the data. It is what evidence usually looks like
when it is asked to clear a bar set in advance.
:::

## How close to the line should we look?

```{r fig.height=3, fig.width=11.5}
bt <- band_tbl %>%
  mutate(lab = sprintf("+/- %d", band)) %>%
  mutate(lab = factor(lab, levels = sprintf("+/- %d", sort(unique(band)))))

plot_ly() %>%
  add_segments(data = bt, x = ~abs(lo), xend = ~abs(hi), y = ~lab, yend = ~lab,
               line = list(color = ACCENT, width = 4), showlegend = FALSE,
               hovertemplate = "avoids %{x:,.0f}<extra></extra>") %>%
  add_markers(data = bt, x = ~abs(itt), y = ~lab,
              marker = list(color = ACCENT, size = 11), showlegend = FALSE,
              hovertemplate = "avoids %{x:,.0f}<extra></extra>") %>%
  add_segments(x = RULEBAR, xend = RULEBAR, y = -0.5, yend = 4.5,
               line = list(color = ALERT, width = 3),
               name = sprintf("The rule: %s", gfmt(RULEBAR)),
               hoverinfo = "name") %>%
  layout(xaxis = ax("AED avoided per business", range = c(0, 1250)),
         yaxis = ax("Bandwidth: index points either side of 58",
                    zeroline = FALSE)) %>%
  pl_m2(height = 290)
```

::: {.ask}
Drag or hover. The estimate moves as the bandwidth widens: roughly
`r gfmt(abs(band_tbl$itt[band_tbl$band == 2]))` at two points,
`r gfmt(abs(band_tbl$itt[band_tbl$band == 5]))` at five.

**Which bandwidth is the right one?** There is no rule. That choice is the
analyst's, and it is yours to interrogate.
:::

## The number that matters: who actually took it up

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
Not everyone offered the programme enrolled. Just below the line, take-up is
**`r sprintf("%.0f", 100 * mean(win$enrolled[win$below == 1]))`%**. Just above,
**`r sprintf("%.0f", 100 * mean(win$enrolled[win$below == 0]))`%**.

So the jump of `r gfmt(abs(itt))` AED is the effect of being **offered** the
programme, not of taking it.
:::

::: {.warn}
Divide by the take-up difference and you get the effect on those who
**took** it:

**`r gfmt(abs(scaled))` AED**, roughly twice the size.
:::
:::::

::::: {.column width="46%"}
```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED per business offered</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f%%</span><span class="k">Take-up below the line</span></div>
  <div><span class="n">%s</span><span class="k">AED per business that took it</span></div>
</div>',
  gfmt(abs(itt)),
  100 * mean(win$enrolled[win$below == 1]),
  gfmt(abs(scaled))))
```
:::::

::::::

::: {.ask}
**This is a decision, not a technicality.** Offering the programme to a hundred
businesses and having 48 take it up is a different proposition from 100 taking
it up — for cost, for scaling, and for what you should expect next time.

Which number would you put in the briefing?
:::

# 3 · Four things that must be checked

## This method's one assumption

```{r results='asis'}
cat('
<div class="hero">
  <div class="value" style="font-size:2.05em;">Nothing else jumps at the cut-off.</div>
</div>')
```

::: {.panel}
A business one point below 58 and one point above should be alike in every way
**except** whether it was offered the programme.

If something else also changes at 58 — if the businesses differ, or the rules
were bent — the jump is not the programme.
:::

::: {.ask}
It is the same shape as yesterday's assumption. Yesterday it was *the groups
were moving in parallel*. Today it is *nothing else changes at the line*.

Both are statements about a world you cannot see. Both have to be probed.
:::

## Check 1: is the jump really at the line?

:::::: {.columns}

::::: {.column width="56%"}
```{r}
placebo_tbl %>%
  mutate(across(c(jump, lo, hi), ~ round(.x, 0))) %>%
  kable(align = "lrrrr", digits = 0,
        col.names = c("Cut-off tested", "n", "Jump", "CI low", "CI high")) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="44%"}
::: {.panel}
The real cut-off is 58. The table tests **50, 54, 62 and 66** — places where no
rule exists, so there should be no jump.

Every interval contains zero. Good news.
:::

::: {.warn}
If the method found a jump where there is no programme, its jump at 58 would be
worth nothing.
:::
:::::

::::::

## Check 2: did anyone bend the rule?

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.8, fig.width=5.6}
dens_tbl %>%
  plot_ly(x = ~bucket, y = ~n, type = "bar",
          marker = list(color = c(ACCENT2, ACCENT2, ACCENT, ACCENT),
                        line = list(width = 0)),
          text = ~n, textposition = "outside",
          textfont = list(size = 14, color = INK), showlegend = FALSE,
          hovertemplate = "%{x}: %{y} businesses<extra></extra>") %>%
  layout(xaxis = ax("Efficiency index", zeroline = FALSE),
         yaxis = ax("Businesses", range = c(0, 1000))) %>%
  pl_m2(width = PL_WC, height = 270,
        title = "Counts either side of 58 (orange = eligible side)")
```
:::::

::::: {.column width="48%"}
::: {.panel}
If businesses could nudge their index across the line to qualify, there would be
a **pile-up** just below 58.

There is not. The counts are smooth through the cut-off, and the two buckets
either side are `r dens_tbl$n[2]` and `r dens_tbl$n[3]`.
:::

::: {.warn}
**Ask this question anyway.** A pile-up would mean the businesses closest to the
line are unlike the rest, and the estimate would be biased. It is invisible
unless you look.
:::
:::::

::::::

## Check 3: are the two sides alike?

This is the check that **does not fully pass**, and it is the most useful slide
in the session.

```{r}
cov_tbl %>%
  kable(align = "l", col.names = c("Characteristic",
        "+/- 1", "+/- 2", "+/- 3", "+/- 5", "+/- 8")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  footnote(general = "* marks a difference whose 95% interval excludes zero.",
           general_title = "", footnote_as_chunk = TRUE)
```

::: {.ask}
Look along the **top row** as the window widens:
`r cov_tbl$b1[1]`, then `r cov_tbl$b2[1]`, `r cov_tbl$b3[1]`,
`r cov_tbl$b5[1]`, `r cov_tbl$b8[1]`.

What do you notice about when the star appears and when it goes?
:::

## What that row means

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**At +/-2 and +/-3 the distance to a recycling centre appears to jump.**

Taken at face value, that says the two sides are not alike — and the estimate
would not be credible.
:::
:::::

::::: {.column width="50%"}
::: {.panel .fragment}
But it is absent at +/-1, and absent again at +/-5 and +/-8.

A real discontinuity does not appear at two bandwidths and vanish at the
neighbouring ones. That pattern is the signature of a **specification
artefact**: a result that depends on an analyst's choice rather than on the
data.
:::
:::::

::::::

::: {.ask}
So the honest reading is: **there is no strong evidence of a jump in the
covariates, but the check is not clean either.**

Which is a different claim from "the check passed", and a different claim from
"the estimate is wrong". Hold all three apart.
:::

## Check 4: who does the result apply to?

:::::: {.columns}

::::: {.column width="52%"}
```{r}
reach_tbl %>%
  mutate(across(c(staff, area, cost), ~ round(.x, 2))) %>%
  kable(align = "lrrr", col.names = c("", "n", "Staff", "Area", "Cost (AED)")) %>%
  kable_styling(full_width = FALSE, font_size = 15)
```
:::::

::::: {.column width="48%"}
::: {.panel .fragment}
The businesses near the line are a particular group: smaller staff
(`r sprintf("%.1f", reach_tbl$staff[1])` against
`r sprintf("%.1f", reach_tbl$staff[2])`) and smaller premises.

The estimate describes **them**.
:::

::: {.warn}
It does **not** describe the large, efficient businesses far from the line.
Nothing here tells you what the programme would do for those.

This is the single most common misreading of an RDD result.
:::
:::::

::::::

::: {.ask}
An evaluation of a programme for the least efficient businesses has produced an
estimate for businesses *near a line*. If your plan is to extend it to
everyone, what have you actually learned?
:::

# 4 · Put AI to work

## AI Snapshot: who does this apply to?

::: {.panel}
Paste the table and the four checks into an AI tool with this prompt:

**"Explain, without jargon, who this result actually applies to."**
:::

::: {.ai-output}
**AI response:**

This study found that the GreenWaste programme reduced costs by around
**249 AED** for participating businesses. [Because the study uses a
cut-off at an efficiency index of 58, the findings apply to all businesses in
the programme]{.flag}.

[The cut-off design means the result generalises to the wider population of
inefficient businesses]{.flag}. The estimate is therefore a reasonable basis for
[scaling the programme to every business below average efficiency]{.flag}.
:::

::: {.ask}
Mark every claim the study supports. Four minutes.
:::

## Debrief: one error, repeated three ways

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "Applies to all businesses in the programme."**

The estimate is local to businesses near the line. It says nothing about those
far from it, whether they were in the programme or not.
:::

::: {.warn}
**2 · "Generalises to the wider population of inefficient businesses."**

Backwards. RDD is **least** able to generalise, precisely because it uses only
the businesses closest to the line.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "A reasonable basis for scaling to every business."**

A scale-up decision needs exactly the thing the model has discarded: what
happens away from the cut-off.
:::
:::::

::::::

::: {.panel}
**The pattern to notice.** The model made the same mistake three times in
different words, and each restatement sounded more confident than the last.

The question it was asked — *who does this apply to?* — is the one it answered
least well. A narrow question about scope is where a fluent model is at its
most dangerous, because a broad answer reads better than a careful one.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain, without jargon, who this result actually applies to.

Nothing about the design, so the model reaches for the most natural-sounding
story: the programme, everyone in it, and what to do next.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> This is a regression discontinuity result with the cut-off at an efficiency
> index of 58, estimated within 2 points either side. The businesses near the
> line are smaller than average. Name precisely which businesses the estimate
> describes, and state what it cannot tell us. **Ask me questions before you
> answer.**
:::
:::::

::::::

::: {.ask}
Give the model the design and the local nature of the estimate, then run both.
Does error 1 survive? Does error 2?

The habit: **supply the constraint the model cannot infer, then ask a narrower
question.**
:::

# 5 · The Abu Dhabi case

## `[PLACEHOLDER]` The DCD social protection evaluation

::: {.warn}
**Awaiting the real findings.** The Abu Dhabi Department of Community
Development social protection evaluation has not been shared with this workspace
yet, so nothing here is presented as its result.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>      <div class="v">[PLACEHOLDER]</div>
  <div class="k">Running variable</div><div class="v">[PLACEHOLDER]</div>
  <div class="k">Cut-off</div>        <div class="v">[PLACEHOLDER]</div>
  <div class="k">Outcome</div>        <div class="v">[PLACEHOLDER]</div>
</div>
```

::: {.panel}
The structure is fixed. When the DCD findings arrive they drop into these four
fields and the two slides that follow, without redesigning the session.
:::

## Four questions for the DCD result

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, and which side is eligible?</div>
  <div class="k">2</div> <div class="v">Is the rule sharp, and if not, what is take-up on the eligible side?</div>
  <div class="k">3</div> <div class="v">Is the density smooth at the cut-off, or did people bend the rule?</div>
  <div class="k">4</div> <div class="v">Who sits near the line, and is that who you want to help?</div>
</div>
```

::: {.ask}
The same four checks we just ran on GreenWaste, in the same order.

The programme changes; the interrogation does not. That transferable habit is
what you are here for.
:::

# 6 · Close

## Four questions to ask any evaluator presenting RDD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, which side is eligible, and is the rule sharp or fuzzy?</div>
  <div class="k">2</div> <div class="v">Does the jump survive at other bandwidths, and are there placebo cut-offs that show nothing?</div>
  <div class="k">3</div> <div class="v">Is the density smooth at the line, and are the covariates balanced either side?</div>
  <div class="k">4</div> <div class="v">Who is near the line, and does the estimate clear our rule for them?</div>
</div>
```

::: {.panel}
Question 4 has two halves, and both matter. **Who it applies to** is the RDD
question. **Does it clear the rule** is the question every session this week has
ended on.
:::

## Where this sits

```{=html}
<div class="facts">
  <div class="k">Day 1 S2</div> <div class="v">descriptive reading &middot; -1.13 against a 2.0 rule</div>
  <div class="k">Day 1 S3</div> <div class="v">naive comparisons and an RCT</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences &middot; -816 against a 1,000 rule</div>
  <div class="k">Today</div>  <div class="v">regression discontinuity &middot; -249 against a 1,000 rule, and local</div>
</div>
```

::: {.ask}
Notice how each method buys its credibility, and what it pays:

**DiD** buys a comparison group by assuming parallel trends.
**RDD** buys credibility from the rule, and pays with **reach** — a tight
estimate about a narrow group.

Next session does the same trade a third time, with **matching**: a much wider
reach, bought by assuming that everything that matters was measured.
:::
