---
title: "Reading a Regression Discontinuity Result"
subtitle: "Module 2 · Day 2, Session 2: What Happens at the Cut-off"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources must sit
# next to the published .html.
#
# Built from Fiona's Oct13_session2.qmd (Sep 24). The simple GreenWaste case
# (SIMPLE_GREENWASTE_PLAN.md): in the rest of the city every business scoring
# 58 or below took part and none above did, so the rule is sharp. Manager age
# also jumps at 58, which check 3 finds.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
    - gw_lab.R
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(knitr)
library(estimatr)
library(qrcode)

gfmt  <- function(x) format(round(x), big.mark = ",")
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")

# ---- Live Menti poll (placeholder until the in-slide poll is built) --------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))
menti_qr <- function(path = "menti_qr_d2s2.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0)); plot(qr_code(url, ecl = "H")); path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000
CUT     <- 58

# ---- The numbers the static slides quote (live cells recompute them) -------
# greenwaste_case.R computes the headline numbers once. It is sourced into its
# own environment because it defines RULE, the 1,000 AED bar.
case_env <- new.env()
case_env$gw <- read.csv("./evaluation_data_GreenWaste_simple.csv")
sys.source("greenwaste_case.R", envir = case_env)
case <- case_env$case
city <- case_env$city
city$distance <- city$score - CUT
city$below    <- as.integer(city$score <= CUT)

at <- function(d, width, cut = CUT, adjust = FALSE) {
  d$distance <- d$score - cut; d$below <- as.integer(d$score <= cut)
  d <- subset(d, abs(distance) <= width)
  f <- lm_robust(if (adjust) cost_after ~ below * distance + manager_age
                 else cost_after ~ below * distance, data = d)
  list(est = coef(f)[["below"]], ci = confint(f)["below", ], n = nrow(d),
       slope = coef(f)[["distance"]])
}
w2 <- at(city, 2)                                  # the +/-2 window
w2_least <- min(abs(w2$ci))                        # interval end closest to zero
w2_age <- at(city, 2, adjust = TRUE)               # adjusted for manager age

# Check 1: every window the slider offers, and the placebo cut-offs it offers.
windows  <- lapply(1:10, function(w) at(city, w))
win_est  <- abs(sapply(windows, function(x) x$est))
win_best <- max(sapply(windows, function(x) max(abs(x$ci))))   # most generous end

# Check 3: manager age at the line, at every window the slider offers.
age_jumps <- sapply(1:8, function(w) {
  d <- subset(city, abs(distance) <= w)
  coef(lm(manager_age ~ below * distance, data = d))[["below"]]
})
```

# 0 · The GreenWaste case {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The GreenWaste case

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the live cells
# below use is defined here, so each cell on a slide stays a few lines long.
gw <- read.csv("evaluation_data_GreenWaste_simple.csv")
library(estimatr)
source("gw_lab.R")   # the GreenWaste lab: an invented world where the true effect is known
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)
# Warm up estimatr now: its first call in the browser is slow, later ones fast.
invisible(lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1))))

CUT <- 58
# The rest of the city, where the score decided: 58 or below took part.
city <- subset(gw, setting == "city")
city$distance <- city$score - CUT                  # the score measured from the line
city$below    <- as.integer(city$score <= CUT)     # 1 = eligible side

clean_par <- function() par(mar = c(4.2, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.2)

# The jump at the line, inside a window of +/- width points, with its interval.
jump_at <- function(d, width, cut = CUT) {
  d$distance <- d$score - cut
  d$below    <- as.integer(d$score <= cut)
  d <- subset(d, abs(distance) <= width)
  f <- lm_robust(cost_after ~ below * distance, data = d)
  ci <- confint(f)["below", ]
  data.frame(cut_off = cut, window = width, businesses = nrow(d),
             jump = round(coef(f)[["below"]]), lower = round(ci[1]), upper = round(ci[2]),
             row.names = NULL)
}

# Average cost in half-point bins, a fitted line on each side, and the jump.
jump_plot <- function(width = 4, view = 10) {
  d <- subset(city, abs(distance) <= view)
  d$bin <- floor(d$distance * 2) / 2 + 0.25            # half-point bins
  bins <- aggregate(cost_after ~ bin, data = d, FUN = mean)
  names(bins) <- c("bin", "cost")
  clean_par()
  plot(bins$bin + CUT, bins$cost, pch = 19, cex = 1.3, bty = "n",
       col = ifelse(bins$bin < 0, ACCENT, ACCENT2),
       xlab = "Efficiency score (eligible: 58 or below)", ylab = "Waste costs after, AED")
  abline(v = CUT, col = MUTED, lty = 2)
  w <- subset(city, abs(distance) <= width)
  f <- lm(cost_after ~ below * distance, data = w)
  for (s in 0:1) {
    xs <- if (s == 1) seq(-width, 0, length.out = 20) else seq(0, width, length.out = 20)
    lines(xs + CUT, predict(f, data.frame(below = s, distance = xs)),
          col = if (s == 1) ACCENT else ACCENT2, lwd = 4)
  }
  top <- coef(f)[["(Intercept)"]]
  segments(CUT, top, CUT, top + coef(f)[["below"]], col = ALERT, lwd = 6)
}

# Does a characteristic jump at the line? Same regression, a covariate as outcome.
balance <- function(width) {
  vars <- c(manager_age = "manager_age", staff = "staff",
            area = "area", filtration = "filtration")
  d <- subset(city, abs(distance) <= width)
  do.call(rbind, lapply(names(vars), function(v) {
    f  <- lm_robust(reformulate(c("below", "distance", "below:distance"), vars[[v]]), data = d)
    ci <- confint(f)["below", ]
    data.frame(characteristic = v, jump = round(coef(f)[["below"]], 2),
               lower = round(ci[1], 2), upper = round(ci[2], 2),
               differs = ci[1] > 0 | ci[2] < 0, row.names = NULL)
  }))
}

# Businesses per one-point slice of the score, 54 to 62.
density_plot <- function() {
  slice <- subset(city, score >= 54 & score < 62)
  clean_par()
  hist(slice$score, breaks = 54:62, col = ifelse(54:61 < 58, ACCENT, ACCENT2),
       border = "white", main = "", xlab = "Efficiency score", ylab = "Businesses")
  abline(v = 58, col = ALERT, lwd = 3, lty = 2)
}
```

{{< include _case_map.qmd >}}

::: {.notes}
About 1 minute: a reminder of the case. Point at the rest of the city, where the
score decided: this session compares businesses just either side of 58 there.

Open the deck a couple of minutes before the session: R and its statistics
package take about a minute to load in the browser, after which every live cell
runs in a few seconds.
:::

## Today we open the box

::: {.panel}
This morning's DiD leaned on parallel trends, and the pilot's waiting businesses
showed that assumption failing. Today the comparison comes from the rule itself:
businesses just either side of 58. We build the estimate from the data and run
the four checks an RDD result has to pass.
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">The rule and the data: who took part</div>
  <div class="k">2</div> <div class="v">From the picture to a number</div>
  <div class="k">3</div> <div class="v">Can we trust the jump? Checks on the window, bending the rule, balance, and who the result applies to</div>
</div>
```

::: {.warn}
You will not write code. You call the choices; I type and press **Run**.
:::

# 1 · The rule and the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The rule that decided who got in

:::::: {.columns}

::::: {.column width="54%"}
In the rest of the city, GreenWaste went to businesses whose **efficiency
score** was **at or below 58**: the less efficient ones.

Two businesses either side of that line are almost identical. One got the
programme, one did not.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Running variable</div> <div class="v">Efficiency score</div>
  <div class="k">Cut-off</div>          <div class="v">58</div>
  <div class="k">Eligible</div>         <div class="v">at or below 58</div>
  <div class="k">Rule followed?</div>   <div class="v">we check it next</div>
</div>
```
:::::

::::::

::: {.warn}
**Read the direction carefully.** The rule offers the programme to the **less**
efficient businesses, so the eligible side is on the **left** of the chart.
Getting this backwards flips the sign of everything that follows.
:::

## `[MENTI PLACEHOLDER]` Predict the picture

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask}
Businesses at or below 58 got the programme, and it cut costs. On the chart of
costs against the score, **where is the eligible side, and at the line does it
sit above or below the other side?**

**A** Left, lower · **B** Left, higher · **C** Right, lower · **D** Right, higher
:::

Go to **menti.com**, code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='260px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.panel .fragment}
**A.** A low score sits on the left, and low is eligible; a programme that cut
costs puts the eligible side **below** at the line. Anyone who assumed the
programme went to the *better* businesses would read the chart backwards.
:::

::: {.notes}
Two predictions in one vote: the direction of the rule and the sign of the
effect. Keep the result on screen and point back to it on the next chart.
:::

## Did the rule hold?

```{webr}
data.frame(side = c("above 58", "58 or below"),
           businesses = as.vector(table(city$below)),
           took_part  = as.vector(tapply(city$took_part, city$below, mean)))
```

::: {.ask}
Before I press Run: what share of eligible businesses do you expect took part?
:::

::: {.panel .fragment}
Every business at or below 58 took part; none above 58 did. The rule is
**sharp**, and the jump at the line is the effect of taking part.
:::

::: {.notes}
Read the two rows aloud. `took_part` is a share: 1 means every business on that
side, 0 means none.
:::

## The picture the method is built on

```{webr}
#| autorun: true
#| edit: false
#| fig-width: 11
#| fig-height: 3.6
jump_plot(width = 4)
```

::: {.ask}
Dots are average costs in half-point bins. **Is there a step at the dashed line?**
:::

# 2 · From the picture to a number {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## The idea behind the method

::: {.panel}
Compare businesses **just below** the cut-off, who got the programme, with
businesses **just above** it, who did not.
:::

::: {.panel}
Right at the line, scoring 57.9 or 58.1 is close to a matter of chance, so the
two sides are alike except for the programme. The **step in costs at the
cut-off** is the estimated effect.
:::

::: {.notes}
Point back at the previous chart: the step is the gap between the two clouds of
dots at the dashed line.
:::

## A first, rough number

```{webr}
near <- subset(city, abs(distance) <= 2)             # within 2 points of the line
eligible_side <- mean(near$cost_after[near$below == 1])
other_side    <- mean(near$cost_after[near$below == 0])
round(data.frame(eligible_side, other_side, gap = eligible_side - other_side))
```

::: {.ask}
Each group: put your jump on the number line on the wall. **Why are the jumps
so different?** Then Run: the same subtraction on every business within 2 points.
:::

::: {.panel .fragment}
Six businesses a side is too few to trust. And even with many, an average across
the window mixes in the slope of costs along the score. The regression deals
with both.
:::

::: {.notes}
Paper (Oct13_session2_materials.docx): Before pressing Run, hand out the jump cards (pack pages 1-8, eight different cards A-H, one per group). Groups average six businesses on each side, subtract, and put a sticky note with their jump and card letter on the wall number line (print Oct13_session2_board_A1.pdf at A1 and tape it up beforehand: -2,000 to +500 AED, the rule marked in red at -1,000). The jumps run from about -1,630 to +250. Ask why they differ, then press Run: the regression uses the `r format(w2$n, big.mark = ",")` businesses within two points.
:::

## Running the regression

```{webr}
near <- subset(city, abs(distance) <= 2)
fit <- lm_robust(cost_after ~ below * distance, data = near)
round(as.data.frame(summary(fit)$coefficients[, c(1, 2, 4)]), 1)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
`distance` is the score measured from the line (0 = 58); `below` is 1 on the
eligible side. `below * distance` fits a line on each side.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Only one row is the programme's effect. Which one?
:::

::: {.panel .fragment}
**`below`: the jump, `r gfmt(w2$est)` AED.** `distance`, about
`r sprintf("%+.0f", w2$slope)` AED per point, is the slope **along** the line,
not the step **at** it.
:::
:::::

::::::

## Seeing the jump: the red line is `below`

```{webr}
#| autorun: true
#| edit: false
#| fig-width: 11
#| fig-height: 3.4
jump_plot(width = 2, view = 4)
```

::: {.panel}
A line on each side, fitted to businesses within 2 points of 58. The red segment
at the line is the `below` row, `r gfmt(w2$est)` AED. The lines slope up, which
is why the raw average overstated the gap.
:::

## Does it clear the rule?

```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">%s</div>
  <div class="label">AED at the cut-off, within 2 points</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
  <div><span class="n">No</span><span class="k">Clears the bar</span></div>
</div>', gfmt(w2$est), gfmt(RULEBAR), gfmt(w2_least)))
```

::: {.ask}
A real effect, clearly not zero. The interval runs from `r gfmt(w2_least)` to
`r gfmt(max(abs(w2$ci)))`. **Would you scale the programme on this?**
:::

::: {.notes}
The interval is wide because only `r format(w2$n, big.mark = ",")` businesses sit within 2 points. Check 1
shows how the answer moves with that choice.

Wall: draw the regression below the number line now, a star at the estimate
and a line along its interval. How many group notes fall inside the
interval? Which side of the red rule line is most of the interval on?
:::

# 3 · Can we trust the jump at 58? {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## This method's one assumption

```{=html}
<div class="hero">
  <div class="value" style="font-size:1.9em;">Nothing else jumps at the cut-off.</div>
</div>
```

::: {.panel}
A business one point below 58 and one point above should be alike in every way
except whether it got the programme. If something else also changes at 58, the
jump is not only the programme.
:::

::: {.ask}
This morning's difference-in-differences assumed the groups moved in parallel;
this method assumes nothing else changes at the line. **If something else did
change at 58, where in the data would you see it?**
:::

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the checks scorecard ("Can we trust the jump at 58?") (pack page 9, one per group). Groups mark each check as you run it.
:::

## The GreenWaste lab: add RDD

:::::: {.columns}

::::: {.column width="46%"}
```{ojs}
//| echo: false
viewof rdd_hidden = Inputs.range([0, 800], {value: 0, step: 100, label: "Joiners differ in something nobody measured (AED)"})
viewof rdd_diverge = Inputs.range([0, 600], {value: 0, step: 50, label: "Joiners' costs were already falling faster (AED)"})
viewof rdd_jump = Inputs.range([0, 600], {value: 0, step: 50, label: "Something else changes at 58 (AED)"})
```
:::::

::::: {.column width="54%"}
::: {.ask}
RDD joins the lab, with one new problem: the last slider. **Which slider breaks
RDD, and which of DiD's problems does it survive?**
:::

::: {.panel .fragment}
Only a jump at 58 breaks it. Differences that change smoothly along the score
leave the step at the line alone.
:::
:::::

::::::

```{webr}
#| input:
#|   - rdd_hidden
#|   - rdd_diverge
#|   - rdd_jump
#| autorun: true
#| edit: false
#| echo: false
#| fig-width: 10
#| fig-height: 3.2
lab(c("before_after", "with_without", "randomised", "did", "rdd"),
    hidden = rdd_hidden, diverge = rdd_diverge, jump = rdd_jump)
```

::: {.notes}
About 5 minutes, before the checks. The first two sliders break
with-and-without, and the middle one breaks DiD; RDD stays on -900 for both,
because neither changes suddenly at 58. The last slider is this method's one
assumption: if something else changes at the cut-off, the jump is not only the
programme. Check 3 finds exactly that in the real data: manager age jumps at 58.
:::

## Check 1: does the jump survive a different window?

```{ojs}
//| echo: false
viewof width = Inputs.range([1, 10], {value: 2, step: 1, label: "Window either side of 58 (score points)"})
```

```{webr}
#| input:
#|   - width
#| autorun: true
#| edit: false
jump_at(city, width)
```

::: {.ask}
The window is the analyst's choice. Slide from 1 to 10. **Does the verdict against
1,000 AED change?**
:::

::: {.warn .fragment}
The estimate moves between about `r gfmt(min(win_est))` and
`r gfmt(max(win_est))` AED, always short of the bar. Even the most generous end
of the interval stays below 1,000 (at most about `r gfmt(win_best)`). The verdict
holds; still ask which window was used, and why.
:::

## Check 1, continued: a jump where there is no rule?

```{ojs}
//| echo: false
viewof placebo = Inputs.range([50, 66], {value: 62, step: 4, label: "Pretend the cut-off is at"})
```

```{webr}
#| input:
#|   - placebo
#| autorun: true
#| edit: false
jump_at(city, width = 2, cut = placebo)
```

::: {.ask}
Move the cut-off to 50, 54, 62 and 66, where no rule exists. **What should the jump
be?**
:::

::: {.panel .fragment}
Close to zero, and every interval there contains zero. At 58 the jump is far
outside that range. If the method found jumps where there is no programme, its
jump at 58 would mean nothing.
:::

## Check 2: did anyone bend the rule?

```{webr}
#| fig-width: 10
#| fig-height: 2.6
density_plot()
```

::: {.ask}
If businesses could nudge their score to 58 or below to qualify, what would the
bar just left of the line look like?
:::

::: {.panel .fragment}
Taller than its neighbours, with a dip just right. The bars at the line are not
out of pattern: no sign of a pile-up. Ask whether anyone drew this chart; a
pile-up only shows if someone looks.
:::

## Check 3: are the two sides alike?

```{ojs}
//| echo: false
viewof bwidth = Inputs.range([1, 8], {value: 2, step: 1, label: "Window (score points)"})
```

```{webr}
#| input:
#|   - bwidth
#| autorun: true
#| edit: false
balance(bwidth)
```

::: {.ask}
Each row runs the same regression with a business characteristic as the outcome.
If the sides are alike, no row should jump. Change the window. **Which row does
not behave?**
:::

::: {.warn .fragment}
**Manager age**: businesses just below the line have managers
`r floor(min(abs(age_jumps)))` to `r ceiling(max(abs(age_jumps)))` years younger,
at every window. Something other than the programme changes at 58.
This check fails.
:::

## What that row means

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
If younger managers also run their waste costs differently, part of the jump is
the managers, not the programme. Adjusting for manager age moves the jump from
`r gfmt(w2$est)` to `r gfmt(w2_age$est)` AED.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Adjusting for manager age only fixes what was measured. If manager age jumps at
58, something unmeasured may jump too.
:::

::: {.ask .fragment}
What would you ask the evaluator? *Why does manager age change at the cut-off,
and what else might?*
:::
:::::

::::::

## Check 4: who does the result apply to?

```{webr}
tp <- subset(city, took_part == 1)
near <- abs(tp$distance) <= 2
data.frame(took_part = c("within 2 points of 58", "further below 58"),
           businesses    = c(sum(near), sum(!near)),
           average_score = round(c(mean(tp$score[near]), mean(tp$score[!near]))))
```

::: {.panel}
The jump describes businesses scoring close to 58. Most of the businesses that
took part score well below it. In the pilot district the lottery covered scores
from 20 to 58, and found `r gfmt(case$rct)` AED.
:::

::: {.ask .fragment}
The jump here is `r gfmt(w2$est)`. If the plan is to extend the programme to
everyone, what have you learned about the businesses the extension would reach?
:::

::: {.notes}
The two numbers do not contradict each other: the programme cuts costs more for
less efficient businesses, and the jump only sees businesses at 58. This is
also why the adjusted jump is not the answer to the scale-up question.
:::

# 4 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: who does this apply to?

:::::: {.columns}

::::: {.column width="42%"}
::: {.panel}
Paste the regression table and the check results into an AI tool with this
prompt:

**"Explain, without jargon, who this result actually applies to."**
:::

::: {.ask}
Mark every claim the study supports. Four minutes.
:::
:::::

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

This study found that the GreenWaste programme reduced costs by around
**`r gfmt(abs(w2$est))` AED** for participating businesses. [Because the study uses a cut-off at
an efficiency score of 58, the findings apply to all businesses in the
programme]{.flag}. [The cut-off design means the result generalises to the wider
population of inefficient businesses]{.flag}. The estimate is therefore a
reasonable basis for [scaling the programme to every business below average
efficiency]{.flag}.
:::
:::::

::::::

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the AI handout (pack page 6, one per group).
:::

## Debrief: one error, three times

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · "Applies to all businesses in the programme."** The estimate is local to
businesses near the line.
:::

::: {.warn}
**2 · "Generalises to the wider population."** Backwards: RDD uses only the
businesses closest to the line, so it generalises least.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · "A basis for scaling to every business."** A scale-up needs what happens
away from the cut-off, which this design discards.
:::

::: {.panel}
And it never mentions the manager-age jump. The question it was asked, *who does
this apply to?*, is the one it answered least well.
:::
:::::

::::::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Explain, without jargon, who this result actually applies to.

Nothing about the design, so the model may reach for the most natural-sounding
story.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> This is a regression discontinuity result with the cut-off at an efficiency
> score of 58, estimated within 2 points either side. Most participants score
> well below 58, and manager age differs across the line. Name which businesses
> the estimate describes and what it cannot tell us. **Ask me questions before
> you answer.**
:::
:::::

::::::

::: {.ask}
Run both. Does error 1 survive? What did each answer explain well? The habit:
give the model the constraint it cannot infer, then ask a narrower question.
:::

# 5 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Your turn: ask AI first

::: {.ask}
Before we show our list, ask an AI tool: **"I am a commissioner about to hear an
evaluator present a regression discontinuity result. What are the key questions
I should ask?"** Three minutes, then compare with the next slide.
:::

::: {.notes}
Optional. If time is short, skip to the next slide. If you run it, ask what the
AI missed, what it added that is worth keeping, and whether anything was wrong.
:::

## Five questions to ask any evaluator presenting RDD

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Where is the cut-off, which side is eligible, and exactly which units are compared?</div>
  <div class="k">2</div> <div class="v">Does the jump stay when the window is wider or narrower? Is there no jump at placebo (fake) cut-offs?</div>
  <div class="k">3</div> <div class="v">Are there about the same number of units just above and just below the cut-off, or did anyone bend the rule?</div>
  <div class="k">4</div> <div class="v">Do other characteristics (such as size or manager age) change suddenly at the cut-off?</div>
  <div class="k">5</div> <div class="v">Who is near the cut-off, and does the estimate meet our decision rule for them?</div>
</div>
```

::: {.notes}
Paper (Oct13_session2_materials.docx): Hand out the take-away cards (pack page 11). Give a minute to fill in "Back at work" on the card, then take two or three examples from the room.
:::

## Where this sits

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Day 1 S3</div> <div class="v">the lottery in the pilot district: %s AED, interval %s to %s</div>
  <div class="k">Day 2 S1</div> <div class="v">difference-in-differences: %s AED, parallel trends doubtful</div>
  <div class="k">Today</div>    <div class="v">regression discontinuity: %s AED near the line (%s adjusting for manager age), against 1,000</div>
</div>', gfmt(case$rct), gfmt(case$rct_ci[1]), gfmt(case$rct_ci[2]), gfmt(case$did),
   gfmt(w2$est), gfmt(w2_age$est)))
```

::: {.panel}
DiD gets its comparison by assuming parallel trends. RDD gets it from the rule,
and the estimate describes only the businesses near the line. Next session,
matching covers more businesses by assuming everything that matters was
measured.
:::
