---
title: "What would have happened anyway?"
subtitle: "Day 2 | Session 1: Difference-in-differences, the extra change and its assumption"
author: "3ie · Review version"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, day2_session1_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session1_review_interactions.html, _day2_session1_review_interactions.html]
webr:
  render-df: kable
  resources: [evaluation_data_GreenWaste_simple.csv]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day2_session1_review.R', encoding='UTF-8')
```

## Yesterday's saving. Today's question. {#d2s1-p01 .reader-slide .s1-review .d2s1-review}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
city <- subset(gw, setting == 'city')
change <- city$cost_after - city$cost_before
joined <- city$took_part == 1
```

Back to the **city**: 4,794 participants and 5,206 other businesses. Participation followed the score rule, **not the pilot lottery**.

<div class="comparison-pair"><div><h3>Before/after: −669 AED</h3><p>What else changed over the year?</p></div><div><h3>After-only gap: −1,638 AED</h3><p>How different were the groups already?</p></div></div>

::: {.ask}
Could comparing **changes in both groups** help?
:::

::: {.notes}
5 minutes. Two minutes independent recall, two minutes compare reasons, one minute city orientation. Programme is equipment subsidies, installation help and training. Same annual waste cost outcome, measured before and 12 months after. Do not treat pilot as city's true effect. Open webR early for one optional arithmetic demo. Transition: inspect what actually changed in both city groups.
:::

## Both groups changed. What was extra? {#d2s1-p02 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(d4_lines())
```

::: {.ask}
Which change should we subtract from which?
:::

**H4-A on worksheet page 1** identifies the city comparison. Solid lines join observed means; there are only two measurement waves.

::: {.notes}
5 minutes. Read population, axes, period and both directions before arithmetic. Participant 1,432 to 763; others 2,258 to 2,401. Programme-start marker indicates between the two measurements, not an additional observed wave or precise start date. Different starting levels do not alone invalidate DiD. Ask predictions, not answers from a formula yet. Transition: work the two changes, then the difference.
:::

## Four numbers. What is the extra change? {#d2s1-p03 .reader-slide .s1-review .d2s1-review}

Use **Q1 on worksheet page 3**. Work alone, then compare.

1. Participants: **763 − 1,432 ≈ −669 AED**.
2. Other businesses: **2,401 − 2,258 ≈ ______ AED**.
3. Extra change: **participant change − other-group change**.

<button data-open="d4-arithmetic" data-closed="Check the extra change" data-opened="Hide the calculation" aria-controls="d4-arithmetic" aria-expanded="false">Check the extra change</button>
<p id="d4-arithmetic" class="calculation-output" hidden>−669 − (+143) ≈ <strong>−812 AED</strong>.</p>

::: {.ask}
Is this the same as the after-only gap?
:::

::: {.notes}
8 minutes. Preserve Fiona's four-number task. Two minutes work, three minutes pair/group check, three minutes collect signs/units and reveal. Use unrounded means for final -812.446; one-AED display discrepancies acceptable. Q1 includes a blank plot to annotate, not another method exercise. Negative extra change means extra reduction in costs. Not automatically an established effect. Transition: see the same arithmetic produced from individual records.
:::

## The same calculation, using all the records {#d2s1-p04 .reader-slide .s1-review .d2s1-review}

The trainer compares mean **cost changes**, using the unrounded records.

```{webr}
#| echo: true
# Participants' mean change minus others' mean change.
# Costs are annual AED per business.
# Round the result after calculating both means.
round(mean(change[joined]) - mean(change[!joined]))
```

::: {.ask}
What comparison does **−812** describe? What does the calculation leave unanswered?
:::

Paper equivalent: **−669 − (+143) ≈ −812 AED**. You do not need to write or run R.

::: {.notes}
6 minutes. Prediction before trainer Run, then Q2 plain-language interpretation. Code comments explain purpose/units/rounding; no syntax lesson or participant code. Output same interaction estimate as saturated two-group/two-wave regression. This arithmetic does not assess the causal assumption. If runtime unavailable use printed arithmetic within same time. Transition: locate that extra change in a regression table.
:::

## Where is the extra change in this table? {#d2s1-p05 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(d4_regression())
```

<div class="review-annotate"><button data-table-column="1" aria-pressed="false">Coefficient</button><button data-table-column="2" aria-pressed="false">Interval</button><button data-table-column="3" aria-pressed="false">P-value</button></div>

::: {.ask}
**Q2:** Find the **interaction**: after × participant. Which question do the other rows answer?
:::

::: {.notes}
5 minutes. H4-A table, page 1. Reference is other group before, 2,258. After row is other group's +143 change. Participant row is initial gap -826. Interaction -812 is the extra participant change, not their own -669 change or after mean. Highlight columns, then annotate row on paper. Intercept meanings depend on reference coding. Model cost ~ after*took_part, business-clustered robust interval -833 to -792. Avoid coefficients as causal magic. Transition: draw the missing no-programme participant outcome.
:::

## The dashed line is an assumption {#d2s1-p06 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(paste0('<div id="d4-path"><div data-variant>',d4_lines(),'</div><div data-variant hidden>',d4_lines(TRUE),'</div></div>'))
```

<button data-swap="d4-path" data-closed="Show the assumed untreated path" data-opened="Show observed means only" aria-pressed="false">Show the assumed untreated path</button>

Under **parallel trends**, participants' costs without GreenWaste would also rise by **143 AED**, to about **1,575 AED**.

::: {.ask}
Is that dashed participant outcome observed, or estimated under an assumption?
:::

::: {.notes}
6 minutes. Predict before reveal. Dashed path hypothetical counterfactual, not extra measured data. 1,432 +143 approximates1,575; assumed path to observed763 yields812 gap. Parallel trends concerns untreated changes, not equal cost levels. Source §3/page1 and visual/page2. Have learners add dashed path to Q1 blank plot and state assumption in Q3. Transition: what could make untreated changes differ?
:::

## Same city. Different changes without GreenWaste? {#d2s1-p07 .reader-slide .s1-review .d2s1-review}

<div class="comparison-pair"><div><h3>A shared change</h3><p>Suppose a service charge rises by the same AED amount for both groups.</p><p>The other group's change may help account for it.</p></div><div><h3>A different change</h3><p>Suppose a charge changes only for the kinds of businesses that tend to participate.</p><p>Their untreated change could differ.</p></div></div>

These are **hypothetical examples**, not facts established in the GreenWaste file.

::: {.ask}
**Q3 on page 4:** What evidence would help check the second possibility?
:::

::: {.notes}
7 minutes. Two minutes think/request, three minutes pair explanation, two minutes feedback. Need comparable absolute untreated changes for this additive cost DiD, not automatically equal percentage rises. Accept charge schedules linked to business types, concurrent policies, changing business mix or measurement. Note composition/measurement and anticipation/spillovers alongside parallel trends when relevant; no new checklist lecture. A city's shared time period alone is insufficient. Transition: earlier histories could help assess plausibility.
:::

## Which earlier history would make you hesitate? {#d2s1-p08 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(paste0('<div class="review-grid"><div><h3>A</h3>',d4_trend(1),'</div><div><h3>B</h3>',d4_trend(2),'</div></div>'))
```

**Hypothetical histories**, identical scales. These are not extra GreenWaste observations.

<button data-open="d4-trend-answer" data-closed="Check your reason" data-opened="Hide feedback" aria-controls="d4-trend-answer" aria-expanded="false">Check your reason</button>
<p id="d4-trend-answer" class="feedback" hidden>B was already changing differently. A is reassuring, but cannot prove future untreated trends.</p>

::: {.notes}
6 minutes. Q3: choose individually and explain feature to partner before feedback. A shows equal changes at different levels; B participant costs already falling while other group rising. Same cost and time axes. Do not infer real group histories from these examples. No significance test proves parallel trends; several earlier measures help credibility, not certainty. Transition: do we have those earlier measures in this actual file?
:::

## Can one before measure show a trend? {#d2s1-p09 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(d4_timeline())
```

::: {.ask}
Could this file show whether earlier cost trends were similar?
:::

Ask for **several earlier measures in both groups**, and an explanation of other changes. Reassuring histories still cannot prove the unobserved follow-up trend.

::: {.notes}
6 minutes. Q3 evidence request, then distinguish supplied/absent information in H4-A §4. One baseline is a level, not a trend. Do not invent a test or substitute pilot waiting-group changes as proof about this city's different population. Actual file has same records/two waves; stable composition doesn't establish parallel trends. Transition: separate causal credibility from the size needed for the decision.
:::

## A precise result. Enough to meet the rule? {#d2s1-p10 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(d4_interval())
```

::: {.ask}
**Q4:** Where do the estimate and interval sit relative to **1,000 AED**?
:::

The interval describes statistical uncertainty **under the model**. It does not resolve parallel trends or include every source of bias.

::: {.notes}
6 minutes. Estimate812, reductioninterval792-833, both below rule. Reverse negative interval endpoints when converting scale. p<.001/intervalexcludingzero are not proof of causal validity or practical sufficiency. Say conditional on model, not certainty of the true effect if assumptions fail. The fictional rule is annual saving per participant. Transition: read the source as a report reader, not just a calculator.
:::

## What does the source support? {#d2s1-p11 .reader-slide .s1-review .d2s1-review}

Read the **City DiD study extract, H4-A, worksheet page 1**. Work alone first.

Circle the **interaction coefficient**, its units and interval.

Underline the **assumption**. Mark the **data limit** that prevents a claimed earlier-trend check.

::: {.ask}
On **page 4**, write a supported interpretation and one consequential evidence request.
:::

::: {.notes}
9 minutes. Five independent reading/annotation, two pair source-check, two feedback. Source identifies physical paragraphs/table rather than a separate unknown report. Do not reteach all questions; use completed Q1-Q4. Listen for extra recorded reduction conditional on credible untreated trends, not an established effect from arithmetic alone. Require request and decision purpose. Transition: AI can report correct numbers but fabricate assurance.
:::

## The AI says cause. Does the comparison? {#d2s1-p12 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(paste0('<div class="ai-reading">',d4_paragraphs(d4_ai),'</div>'))
```

::: {.ask}
On **page 5**, underline support and cross out overclaims. Check **H4-A on page 1**.
:::

::: {.notes}
8 minutes. Five markup, three source-based debrief. Restore old realistic table-explanation device with corrected812, interval792-833 and richer qualification. Authored exercise, no answer highlighting or naive-prompt run. Supported group changes/estimate/precision and benefit of comparing changes. Unsupported: same follow-up period proves parallel trends, interval proves robustness/attribution, clears1000rule/national guarantee. Different levels accounted for is legitimate mechanics, not proof of untreated trend. Monitoring caveat does not repair unsupported claims. Transition: constrain next prompt to source and missing evidence.
:::

## Ask for an explanation you can verify {#d2s1-p13 .reader-slide .s1-review .d2s1-review}

```{r, results='asis'}
html_out(paste0('<div class="better-prompt">',d4_paragraphs(d4_prompt),'</div>'))
```

Copy **page 5's prompt and H4-A** into **Microsoft Copilot**. Keep your AI good-practice card beside you.

::: {.ask}
Which claim improved? What still needs checking against the source?
:::

::: {.notes}
7 minutes. Four run, three compare; Snapshot total15. No account setup. Page6 prepared response is same-task paper alternative if delayed/unavailable. Ask questions first; source/correctness, not AI agreement, is check. No mandatory second chat or extra best-practice slide. Compare interaction interpretation, causal assumptions, uncheckable pre-trend claim and threshold. Transition: give your own reading and evidence request.
:::

## What would you ask for next? {#d2s1-p14 .reader-slide .s1-review .d2s1-review}

Keep **H4-A open**. Close AI responses and peer comments.

::: {.ask}
Write one plain-language interpretation and one specific request. How could the answer change your advice?
:::

Use the **final box on page 4**. DiD compares changes; its causal reading depends on what would have changed without GreenWaste.

Next: a different comparison, around the score cutoff.

::: {.notes}
6 minutes. Three independent response, one pair purpose check, two feedback. Require comparison/units/conditional effect/decision limit plus useful request, not only more data. Request earlier periods, competing changes or measurement/composition evidence with reason. No claim DiD resolves selection automatically. Total90minutes; new review source and worksheet leave production materials unchanged.
:::
