---
title: "Spot the Problem: Naive Comparisons and Reading an RCT"
subtitle: "Module 2 · Day 1, Session 3: Opening the Box on Module 1's Numbers"
author: "3ie"
always_allow_html: yes
title-slide-attributes:
  data-state: title-slide-state
# live-revealjs: the quarto-live extension (_extensions/r-wasm/live) runs R in
# the browser through webR. The live cells need an internet connection on the
# day (webR loads from a CDN) and the CSV listed under webr: resources
# must sit next to the published .html. Static knitr chunks work as before.
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
webr:
  packages: [estimatr]
  # Data frames print as HTML tables (styled in theme_live.scss) instead of
  # console text. knitr is already loaded by quarto-live, so this costs nothing.
  render-df: kable
  resources:
    - evaluation_data_GreenWaste.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    # dpi 96 is deliberate: knitr sizes htmlwidgets as fig.width * dpi.
    dpi: 96
editor_options:
  chunk_output_type: console
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
library(tidyverse)
library(plotly)
library(knitr)
library(kableExtra)
library(estimatr)
library(qrcode)

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"   # Tech / Light Blue
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"
CHART_FONT <- "Noto Kufi Arabic, Segoe UI, Arial, sans-serif"

PL_W <- 1100; PL_WC <- 520; PL_WT <- 340

pl_m2 <- function(p, width = PL_W, height = 290, t = 30, b = 54,
                  title = NULL, legend_y = 1.02) {
  p <- p %>%
    layout(
      width = width, height = height,
      font = list(family = CHART_FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor  = "rgba(0,0,0,0)",
      margin = list(l = 70, r = 24, t = t, b = b),
      legend = list(orientation = "h", x = 0.5, xanchor = "center",
                    y = legend_y, yanchor = "bottom", title = list(text = "")),
      hoverlabel = list(bgcolor = "white", bordercolor = RULE,
                        font = list(color = INK, size = 12))
    )
  if (!is.null(title)) {
    p <- p %>% layout(title = list(text = title, x = 0, xanchor = "left",
                                   font = list(size = 12, color = MUTED)))
  }
  p %>% config(displayModeBar = FALSE)
}

ax <- function(title = "", ...) {
  modifyList(list(
    title = list(text = title, font = list(size = 12, color = MUTED)),
    gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
    tickfont = list(size = 12, color = MUTED)
  ), list(...))
}

gfmt <- function(x) format(round(x), big.mark = ",")

# Emit an HTML snippet as a raw block. Plain cat() leaves it to pandoc, which
# reads indented lines inside a multi-line <div> as a code block and prints the
# tags on the slide.
FENCE <- strrep("`", 3)   # built, not typed: a literal fence would end the chunk
html_out <- function(x) cat("
", FENCE, "{=html}
", x, "
", FENCE, "
", sep = "")

# ---- Live Menti poll -------------------------------------------------------
MENTI_CODE <- "1234 5678"
MENTI_JOIN <- paste0("https://www.menti.com/", gsub(" ", "", MENTI_CODE))

menti_qr <- function(path = "menti_qr.png", url = MENTI_JOIN) {
  png(path, width = 900, height = 900, res = 300, bg = "white")
  on.exit(dev.off(), add = TRUE)
  par(mar = c(0, 0, 0, 0))
  plot(qr_code(url, ecl = "H"))
  path
}
MENTI_QR <- menti_qr()

RULEBAR <- 1000    # must cut costs by at least this many AED

# ---- The numbers the static slides quote -----------------------------------
# The live cells recompute every one of these in the browser. They are computed
# here too so the static slides (recap, rule charts) cannot drift from them.
gw <- read.csv("./evaluation_data_GreenWaste.csv")
mc <- function(keep) mean(gw$waste_management_costs[keep])

# A. Before-and-after among businesses that enrolled (Module 1: -665).
ba_before <- mc(gw$enrolled == 1 & gw$round == 0)
ba_after  <- mc(gw$enrolled == 1 & gw$round == 1)
ba_change <- ba_after - ba_before

# B. Enrolled vs not enrolled at follow-up, in the offered neighbourhoods
# (Module 1: -1,446). Within those neighbourhoods enrolled == eligible, so the
# "not enrolled" group is the businesses that scored above the cut-off.
# NOTE: the earlier Oct12_session3.qmd filtered to eligible == 1 here, which
# compares enrolled businesses with eligible businesses in the CONTROL
# neighbourhoods. That is the randomised comparison (-1,014), not a naive one.
ww_prog <- mc(gw$treatment_neighborhood == 1 & gw$enrolled == 1 & gw$round == 1)
ww_comp <- mc(gw$treatment_neighborhood == 1 & gw$enrolled == 0 & gw$round == 1)
ww_gap  <- ww_prog - ww_comp
ww_gap0 <- mc(gw$treatment_neighborhood == 1 & gw$enrolled == 1 & gw$round == 0) -
           mc(gw$treatment_neighborhood == 1 & gw$enrolled == 0 & gw$round == 0)

# C. The randomised estimate (Module 1: -1,014), clustered by neighbourhood.
fu  <- gw %>% filter(round == 1, eligible == 1)
rct <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                 data = fu, clusters = neighborhood_identifier)
rct_est     <- coef(rct)[["treatment_neighborhood"]]
rct_ci      <- confint(rct)["treatment_neighborhood", ]
rct_control <- coef(rct)[["(Intercept)"]]
# SIGN TRAP: sort the absolute values so [1] is always the end closest to zero.
rct_avoided       <- sort(abs(rct_ci))
rct_avoided_least <- rct_avoided[1]
rct_avoided_most  <- rct_avoided[2]

# Estimate, interval and rule on one axis.
rule_chart <- function(est, least, most, rule, xlab, xmax, digits = 0) {
  fmt <- paste0("avoids %{x:,.", digits, "f}<extra></extra>")
  tibble(
    what = factor(c("The estimate", "95% interval", "Decision rule"),
                  levels = c("Decision rule", "95% interval", "The estimate")),
    lo = c(abs(est), least, rule), hi = c(abs(est), most, rule),
    col = c(ACCENT, ACCENT, ALERT)
  ) %>%
    plot_ly() %>%
    add_segments(x = ~lo, xend = ~hi, y = ~what, yend = ~what,
                 line = list(color = ~col, width = 5), showlegend = FALSE,
                 hovertemplate = fmt) %>%
    add_markers(x = ~lo, y = ~what, marker = list(color = ~col, size = 11),
                showlegend = FALSE, hovertemplate = fmt) %>%
    layout(xaxis = ax(xlab, range = c(0, xmax)),
           yaxis = ax("", zeroline = FALSE)) %>%
    pl_m2(width = PL_WC, height = 250)
}
```

# 0 · Where Module 1 left us {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Where we left GreenWaste

```{webr}
#| autorun: true
#| include: false
# Runs once, in the browser, when the deck opens. Everything the live cells
# below use is defined here, so each cell on a slide stays two or three lines.
gw <- read.csv("evaluation_data_GreenWaste.csv")
library(estimatr)
RULE <- 1000
ACCENT <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"; MUTED <- "#545860"
options(width = 100, scipen = 6)

# Warm up estimatr now, while the opening slides are talked through: its first
# call in the browser takes about 10 seconds, every later one is instant.
invisible(estimatr::lm_robust(y ~ x, data = data.frame(y = c(1, 3, 2, 5), x = c(0, 1, 0, 1)),
                              clusters = c(1, 1, 2, 2)))

# Defined here as well as on the slides that teach them, so a later slide
# still runs if an earlier one was skipped or the page was reloaded mid-deck.
after <- subset(gw, round == 1 & eligible == 1)
fit   <- lm(waste_management_costs ~ treatment_neighborhood, data = after)

clean_par <- function() par(mar = c(3, 5.5, 1, 1), family = "sans", las = 1,
                            col.axis = MUTED, col.lab = MUTED, fg = MUTED, cex = 1.25)
aed <- function(x) format(round(x), big.mark = ",")

# Bar chart of a few averages, value printed on each bar:
#   bars(Before = before, After = after)
bars <- function(..., col = ACCENT) {
  v <- c(...)
  clean_par()
  par(mar = c(2.5, 1, 2, 1))               # no y axis: the value sits on each bar
  b <- barplot(v, col = col, border = NA, ylim = c(0, max(v) * 1.15), axes = FALSE)
  text(b, v, paste(aed(v), "AED"), pos = 3, cex = 1.2, xpd = NA)
}

# Before-and-after lines for the two groups in the offered neighbourhoods.
plot_two_lines <- function() {
  offered <- subset(gw, treatment_neighborhood == 1)
  took <- tapply(offered$waste_management_costs[offered$enrolled == 1],
                 offered$round[offered$enrolled == 1], mean)
  oth  <- tapply(offered$waste_management_costs[offered$enrolled == 0],
                 offered$round[offered$enrolled == 0], mean)
  clean_par()
  plot(0:1, took, type = "b", pch = 19, lwd = 4, col = ACCENT, bty = "n",
       ylim = c(0, 2500), xlim = c(-0.1, 1.6), xaxt = "n",
       xlab = "", ylab = "Waste costs, AED")
  lines(0:1, oth, type = "b", pch = 19, lwd = 4, col = ACCENT2)
  axis(1, at = 0:1, labels = c("Before", "After"))
  text(1.06, took[2], paste("Took part", aed(took[2])), col = ACCENT, adj = 0)
  text(1.06, oth[2],  paste("Did not", aed(oth[2])),    col = ACCENT2, adj = 0)
}

# Neighbourhood averages at baseline, among eligible businesses: the units the
# real lottery was drawn over.
nb <- aggregate(waste_management_costs ~ neighborhood_identifier,
                data = subset(gw, round == 0 & eligible == 1), FUN = mean)
names(nb)[2] <- "baseline_cost"

draw_lottery <- function(how, seed) {
  set.seed(seed)
  half <- nrow(nb) %/% 2
  picked <- if (how == "Random draw") sample(nrow(nb), half) else
            order(nb$baseline_cost)[1:half]   # the cheapest-to-run go first
  m <- c(Offered = mean(nb$baseline_cost[picked]),
         `Not offered` = mean(nb$baseline_cost[-picked]))
  clean_par()
  b <- barplot(m, col = c(ACCENT, ACCENT2), border = NA, ylim = c(0, 2100),
               ylab = "Baseline costs, AED")
  text(b, m + 110, aed(m), cex = 1.3)
  cat(sprintf("Gap BEFORE the programme started: %+.0f AED\n", m[1] - m[2]))
}

# Re-estimate on a random subset of k neighbourhoods: a smaller trial.
small_trial <- function(k) {
  set.seed(11)
  after <- subset(gw, round == 1 & eligible == 1)
  keep  <- c(sample(unique(after$neighborhood_identifier[after$treatment_neighborhood == 1]), k / 2),
             sample(unique(after$neighborhood_identifier[after$treatment_neighborhood == 0]), k / 2))
  d   <- after[after$neighborhood_identifier %in% keep, ]
  fit <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                   data = d, clusters = neighborhood_identifier)
  est <- -coef(fit)[["treatment_neighborhood"]]
  ci  <- sort(-confint(fit)["treatment_neighborhood", ])
  clean_par()
  plot(NA, xlim = c(0, 2000), ylim = c(0.5, 1.5), yaxt = "n", bty = "n",
       xlab = "AED avoided per business", ylab = "")
  abline(v = RULE, col = ALERT, lwd = 3, lty = 2)
  segments(ci[1], 1, ci[2], 1, col = ACCENT, lwd = 8)
  points(est, 1, pch = 19, cex = 2.5, col = ACCENT)
  text(RULE, 1.4, "the rule", col = ALERT, pos = 4)
  cat(sprintf("%d neighbourhoods: avoids %.0f AED, interval %.0f to %.0f
",
              k, est, ci[1], ci[2]))
}

crowd_gaps <- function(k, draws = 1000) {
  set.seed(7)
  gaps <- replicate(draws, {
    s <- sample(nrow(nb), k)
    t <- s[seq_len(k / 2)]; c <- s[-seq_len(k / 2)]
    mean(nb$baseline_cost[t]) - mean(nb$baseline_cost[c])
  })
  clean_par()
  hist(gaps, breaks = seq(-1500, 1500, by = 50), col = ACCENT, border = "white",
       main = "", xlab = "Baseline gap between the two groups, AED", ylab = "Draws",
       xlim = c(-1200, 1200))
  abline(v = 0, col = ALERT, lwd = 2, lty = 2)
  cat(sprintf("%d neighbourhoods: in 95%% of %d lotteries the groups start within +/- %.0f AED\n",
              k, draws, quantile(abs(gaps), 0.95)))
}
```

:::::: {.columns}

::::: {.column width="52%"}
In Module 1 you met the **GreenWaste Program** and saw three estimates of what
it did to businesses' waste costs. The bar for a national roll-out, fixed for
the whole week: **cut at least `r gfmt(RULEBAR)` AED**.

::: {.ask}
From memory: which one did Module 1 tell you to trust, and what was wrong with
the other two?
:::
:::::

::::: {.column width="48%"}
```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Before and after</div>     <div class="v">%s AED</div>
  <div class="k">Enrolled vs not</div>      <div class="v">%s AED</div>
  <div class="k">Randomised</div>           <div class="v">%s AED</div>
</div>', gfmt(ba_change), gfmt(ww_gap), gfmt(rct_est)))
```

::: {.panel .fragment}
Module 1 called the first two **counterfeit counterfactuals**: one ignores what
else changed over time, the other compares groups that were never alike.
:::
:::::

::::::

::: {.notes}
~3 min. Pure recall; do not re-teach. Take two answers, then reveal the panel.
Open the deck a couple of minutes before the session starts: R and its statistics package take about a minute to load in the browser, after which every live cell runs in a few seconds.

Paper (Oct12_session3_materials.docx): Hand out the session tracker (pack page 1, one per participant); they fill in each row as the number is rebuilt live.
:::

## Today we open the box

::: {.panel}
Module 1 showed you the three numbers. Today we compute them from the raw data,
on screen, and look for each problem in the data itself.
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Meet the data: what one row is, and who was offered what</div>
  <div class="k">2</div> <div class="v">Rebuild the two counterfeit numbers, then catch each one out</div>
  <div class="k">3</div> <div class="v">Run the lottery ourselves, and see why it works</div>
  <div class="k">4</div> <div class="v">Read the randomised result against the 1,000 AED rule</div>
</div>
```

::: {.warn}
You will not write any code. You will call the numbers; I will type them in and
press **Run**. The grey boxes are real R, running in this page.
:::

# 1 · Meet the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## One row, one business, one round

```{webr}
rows <- gw[, c("business_identifier", "round", "treatment_neighborhood",
               "enrolled", "waste_management_costs")]
round(head(rows, 4))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**round** 0 before, 1 after. **treatment_neighborhood** 1: the neighbourhood
was offered the programme. **enrolled** 1: the business took part.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
The file has 19,826 rows. How many **businesses**? 
:::

::: {.panel .fragment}
Half: 9,913 businesses, each measured twice. One row is a business in a round.
:::
:::::

::::::

## Who was offered, who took part?

```{webr}
base <- subset(gw, round == 0)
base$businesses <- 1
aggregate(cbind(businesses, enrolled) ~ treatment_neighborhood + eligible,
          data = base, FUN = sum)
```

::: {.ask}
**Who enrolled in the offered neighbourhoods?** And in the others?
:::

::: {.warn .fragment}
In offered neighbourhoods **every eligible business enrolled, and no other
did.** So "did not take part" means "scored above the cut-off": a different kind
of business.
:::

# 2 · Rebuild the two counterfeit numbers {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Comparison A: before and after

::: {.ask}
Module 1 said **`r gfmt(ba_change)`**. Before I press Run: will the raw rows
give the same number?
:::

```{webr}
#| fig-width: 6
#| fig-height: 2.5
before <- mean(gw$waste_management_costs[gw$enrolled == 1 & gw$round == 0])
after  <- mean(gw$waste_management_costs[gw$enrolled == 1 & gw$round == 1])
round(data.frame(before, after, change = after - before))
bars(Before = before, After = after)
```

::: {.notes}
Press Run live. The room should see -665 appear. Point at each line as you read
it aloud in plain words.
:::

## Comparison B: with and without

```{webr}
#| fig-width: 6
#| fig-height: 2.5
offered <- subset(gw, treatment_neighborhood == 1 & round == 1)
took    <- mean(offered$waste_management_costs[offered$enrolled == 1])
didnt   <- mean(offered$waste_management_costs[offered$enrolled == 0])
round(data.frame(took_part = took, did_not = didnt, gap = took - didnt))
bars("Took part" = took, "Did not" = didnt, col = c(ACCENT, ACCENT2))
```

::: {.warn}
Same neighbourhoods, after the programme. Module 1 said **`r gfmt(ww_gap)`**.
Two readings of one programme, **`r gfmt(ba_change)`** and
**`r gfmt(ww_gap)`**: one misses the 1,000 AED bar, one clears it.
:::

## `[MENTI PLACEHOLDER]` Which way is each one wrong?

:::::: {.columns}

::::: {.column width="62%"}
::: {.ask style="font-size: 115%;"}
You know from Module 1 both numbers are biased. The harder question:

**Is `r gfmt(abs(ba_change))` too big or too small?**

**Is `r gfmt(abs(ww_gap))` too big or too small?**
:::

Go to **menti.com** and enter code **`r MENTI_CODE`**, or scan the code.
:::::

::::: {.column width="38%"}
```{r out.width='330px', out.extra='style="display:block; margin:0 auto;"'}
knitr::include_graphics(MENTI_QR)
```
:::::

::::::

::: {.notes}
~4 min. Do not settle it. The next three slides answer it from the data.
:::

# 3 · Catch each one out, in the data {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## A: what happened to everyone else?

:::::: {.columns}

::::: {.column width="56%"}
```{webr}
#| fig-width: 6.2
#| fig-height: 4.2
plot_two_lines()
```
:::::

::::: {.column width="44%"}
::: {.panel}
Before-and-after compares the programme to **the past**. If costs were moving
anyway, that movement is baked into the `r gfmt(abs(ba_change))`.
:::

::: {.ask}
Look at the businesses that did **not** take part. Which way did their costs go
with no programme at all?
:::

::: {.warn .fragment}
Up, by about 150 AED. Costs were rising anyway, so the
`r gfmt(abs(ba_change))` fall **understates** what the programme did.
:::
:::::

::::::

## B: the same gap, before the programme existed

:::::: {.columns}

::::: {.column width="42%"}
```{ojs}
//| echo: false
viewof when = Inputs.radio(["After the programme", "Before the programme"],
  {value: "After the programme"})
```
:::::

::::: {.column width="58%"}
::: {.ask}
Before the programme, nobody had received anything. **What gap should we see?**
:::
:::::

::::::

```{webr}
#| input:
#|   - when
#| autorun: true
#| fig-width: 6
#| fig-height: 1.8
r <- if (when == "Before the programme") 0 else 1
offered <- subset(gw, treatment_neighborhood == 1 & round == r)
took    <- mean(offered$waste_management_costs[offered$enrolled == 1])
didnt   <- mean(offered$waste_management_costs[offered$enrolled == 0])
round(data.frame(took_part = took, did_not = didnt, gap = took - didnt))
bars("Took part" = took, "Did not" = didnt, col = c(ACCENT, ACCENT2))
```


::: {.warn .fragment}
It should be zero. It is **`r gfmt(ww_gap0)`** AED: a head start, so the
`r gfmt(abs(ww_gap))` gap **overstates** the effect.
:::

## Why they were already different

```{webr}
before <- subset(gw, round == 0 & treatment_neighborhood == 1)
round(aggregate(cbind(staff_size, business_area, age_manager,
                      advanced_filtration) ~ enrolled, data = before, FUN = mean), 2)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Same businesses, **before** the programme. Enrolled (1) vs not (0) on four
things we measured.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Younger managers, smaller premises, more filtration already installed. The two
groups were different kinds of business, and **some of the ways they differ we
did not measure at all.**
:::
:::::

::::::

## Two biases, opposite directions

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">Before and after</div>
  <div class="v">%s AED &middot; folds in the rising costs everyone faced &middot; UNDERSTATES</div>
  <div class="k">With and without</div>
  <div class="v">%s AED &middot; folds in a %s AED head start &middot; OVERSTATES</div>
  <div class="k">So the truth</div>
  <div class="v">sits somewhere between the two, and neither comparison can say where</div>
</div>', gfmt(ba_change), gfmt(ww_gap), gfmt(abs(ww_gap0))))
```

::: {.ask}
Check your Menti answer. What would it take to land on the truth
**deliberately**, instead of guessing between two wrong numbers?
:::

::: {.warn}
Both problems have one root: **the two groups were not comparable.** After the
break: a design that makes them comparable by construction.
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker check: every row should now say too big or too small, with a reason.
:::

# Break {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Five-minute break

::: {.panel}
Back in five minutes. When we return, we run the lottery ourselves.
:::

# 4 · Run the lottery yourself {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## What the lottery buys you

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Module 1 recap.** GreenWaste was offered to **neighbourhoods drawn by lot**.
Inside both kinds of neighbourhood, only eligible businesses count.

So the randomised comparison is: eligible businesses in drawn neighbourhoods
vs eligible businesses in the others.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
The claim is that a draw makes the two groups **alike on average**, on what we
measured *and* on what we did not.
:::

::: {.ask}
We can check that claim on the real neighbourhoods.
:::
:::::

::::::

## Draw the lottery again

```{ojs}
//| echo: false
viewof how = Inputs.radio(["Random draw", "Officials pick"],
  {value: "Random draw", label: "How neighbourhoods are chosen"})
viewof draw = Inputs.button("Draw again")
```

:::::: {.columns}

::::: {.column width="55%"}
```{webr}
#| input:
#|   - how
#|   - draw
#| autorun: true
#| edit: false
#| fig-width: 5.6
#| fig-height: 3.4
draw_lottery(how, seed = draw)
```
:::::

::::: {.column width="45%"}
::: {.panel}
Bars show **baseline** costs, before anyone was offered anything.
:::

::: {.ask}
Press **Draw again** five times. How big does the gap get? Then let officials
pick.
:::

::: {.warn .fragment}
When officials pick, the offered neighbourhoods start hundreds of AED ahead, so
the programme looks successful before it has done anything.
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker: record the largest gap after five draws, then the gap when officials pick.
:::

## Two neighbourhoods or two hundred?

```{ojs}
//| echo: false
viewof k = Inputs.range([4, 196], {value: 4, step: 2, label: "Neighbourhoods in the lottery"})
```

```{webr}
#| input:
#|   - k
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 3.3
crowd_gaps(k)
```

::: {.ask}
One thousand lotteries at each size. Start at 4 neighbourhoods, then slide up.
**At what size would you trust a single draw?**
:::

::: {.notes}
At 4 the gaps run to hundreds of AED either way; at ~196 almost every draw
starts within a few dozen. This is "two people versus two crowds" from Module 1,
seen in the data. The same holds for characteristics nobody measured.
:::

## `[MENTI PLACEHOLDER]` Would you trust this draw?

::: {.panel}
Suppose GreenWaste had drawn only **6 neighbourhoods**, 3 offered and 3 not.
Before the programme the offered three already had costs **300 AED lower**.
Afterwards they are **900 AED lower**.
:::

::: {.ask}
What do you conclude? **A** The effect is 900: it was randomised. **B** About 600:
take off the head start. **C** Too few neighbourhoods to trust either number.
**D** I need to see something else first. Vote, then say what you would need.
:::

::: {.panel .fragment}
C, with D's instinct. A draw makes groups alike **on average over many draws**,
not in every draw: with a handful of units a 300 AED head start is ordinary (the
slider showed about ±400 at 4 neighbourhoods). B looks careful but assumes the
head start would have stayed the same. Ask how many units were randomised, and
for the balance table.
:::

::: {.notes}
Replaces a true/false quiz: this asks the room to apply what the two lottery
slides showed to a result they might actually be handed. Expect a split between
A and B; let one person from each argue before revealing.
:::

# 5 · Reading the randomised result {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## First check: did the real draw work?

```{webr}
before <- subset(gw, round == 0 & eligible == 1)
round(aggregate(cbind(waste_management_costs, staff_size, business_area,
                      age_manager) ~ treatment_neighborhood, data = before, FUN = mean), 1)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Same check as three slides ago, but now the two groups are **drawn**
neighbourhoods (1) vs the others (0), eligible businesses only.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Compare with the enrolled vs not table: there, managers differed by ten years.
Here, the groups look alike **before** the programme. That is what a balance
table is for.
:::
:::::

::::::

## The effect is a difference of two averages

::: {.ask}
Module 1 said **`r gfmt(rct_est)`**. Three lines of averages. Will they match?
:::

```{webr}
#| fig-width: 6
#| fig-height: 2
after <- subset(gw, round == 1 & eligible == 1)
offered     <- mean(after$waste_management_costs[after$treatment_neighborhood == 1])
not_offered <- mean(after$waste_management_costs[after$treatment_neighborhood == 0])
round(data.frame(offered, not_offered, effect = offered - not_offered))
bars(Offered = offered, "Not offered" = not_offered, col = c(ACCENT, ACCENT2))
```

::: {.panel .fragment}
No model, no adjustment: the draw made the groups alike, so **the gap is the effect.**
:::

## The regression gives the same number

```{webr}
fit <- lm(waste_management_costs ~ treatment_neighborhood, data = after)
round(data.frame(coefficient = coef(fit)))
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
The table an evaluator hands you comes from a line like this one.

**(Intercept)** is the not-offered average. **treatment_neighborhood** is the
gap between the groups.
:::
:::::

::::: {.column width="50%"}
::: {.ask}
Which of these two numbers is the impact? And what is the other one?
:::

::: {.warn .fragment}
The coefficient is the difference you just computed by hand. The intercept is
where the comparison group sits: **not an effect**. Reading it as one is a
common mistake, by people and by AI.
:::
:::::

::::::

## How sure? Count neighbourhoods, not businesses

```{webr}
library(estimatr)
fit_nb <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                    data = after, clusters = neighborhood_identifier)
round(as.data.frame(rbind(
  "as if 5,628 separate draws" = confint(fit)["treatment_neighborhood", ],
  "counted by neighbourhood"   = confint(fit_nb)["treatment_neighborhood", ])))
```

::: {.panel}
The lottery drew about 200 **neighbourhoods**, not 5,628 businesses, and
businesses in one neighbourhood are not independent pieces of evidence.
:::

::: {.warn .fragment}
Counting neighbourhoods **widens** the interval: its least generous end moves from
974 to **`r gfmt(rct_avoided_least)`** AED. Ask: *"were your errors clustered
where you randomised?"*
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Tracker: record the interval counted by neighbourhood and whether it clears 1,000 AED.
:::

## Which number is the impact?

```{r results='asis'}
html_out(sprintf('
<div class="facts">
  <div class="k">The impact</div>
  <div class="v">the treatment_neighborhood coefficient: %s AED &middot; the difference the programme made</div>
  <div class="k">Not the impact</div>
  <div class="v">the intercept, %s AED &middot; where the comparison group sat</div>
  <div class="k">Not sufficient either</div>
  <div class="v">the p-value &middot; it says the effect is unlikely to be zero, not that it is big enough</div>
</div>', gfmt(rct_est), gfmt(rct_control)))
```

::: {.ask}
Three ways to misread one table. Session 1 met the third: **statistically
significant does not mean big enough.** Here it is on a clean randomised result.
:::

## Does it clear the rule?

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=2.5, fig.width=6}
rule_chart(rct_est, rct_avoided_least, rct_avoided_most, RULEBAR,
           "AED avoided per business", 1350)
```
:::::

::::: {.column width="48%"}
```{r results='asis'}
html_out(sprintf('
<div class="hero">
  <div class="value">-%s</div>
  <div class="label">AED, the randomised estimate</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule</span></div>
  <div><span class="n">%s</span><span class="k">Least it could be</span></div>
</div>', gfmt(abs(rct_est)), gfmt(RULEBAR), gfmt(rct_avoided_least)))
```

::: {.warn}
The estimate clears the bar. The interval **spans** it: at its least generous
the programme avoids `r gfmt(rct_avoided_least)` AED, short of
`r gfmt(RULEBAR)`.
:::
:::::

::::::

::: {.ask .fragment}
Would you sign a national roll-out on this? Commit, then we ask what a smaller
trial would have told us.
:::

## What if the trial had been smaller?

```{ojs}
//| echo: false
viewof kt = Inputs.range([10, 196], {value: 196, step: 2, label: "Neighbourhoods in the trial"})
```

```{webr}
#| input:
#|   - kt
#| autorun: true
#| edit: false
#| fig-width: 10
#| fig-height: 1.8
small_trial(kt)
```

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Slide down to 20. **What happens to the estimate, and to the interval?**
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The estimate barely moves, but the interval widens to about 590 to 1,290 AED.
At that size the trial cannot tell you whether the bar is cleared.
:::
:::::

::::::

::: {.notes}
Same seed each time, so the subset at a given size is stable while you talk.
Measured: 196 neighbourhoods 1,014 [935, 1,093]; 60 gives 1,022 [864, 1,179];
20 gives 940 [591, 1,289]. Link back to Session 1: a study that is too small
may fail to find a real effect.
:::

# 6 · Put AI to work {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## AI Snapshot: a paragraph for the minister

:::::: {.columns}

::::: {.column width="42%"}
::: {.panel}
Groups, paste this prompt, written the way a busy official might write it:

**"Our pilot cut waste costs by 665 AED per business (costs before vs after,
p < 0.001). Write a short paragraph for the minister on what this shows."**
:::

::: {.ask}
Mark anything the evidence does not support. Three minutes, in your groups. Then
run the same prompt yourselves: did your AI write something similar?
:::
:::::

::::: {.column width="58%"}
::: {.ai-output}
*A response like this is possible, especially with a short or leading prompt:*

The GreenWaste pilot produced a clear result: waste-management costs fell by 665
AED per business, and [with p < 0.001 we can be confident the programme caused
this reduction]{.flag}. [Because the same businesses were measured before and
after, differences between businesses are already accounted for]{.flag}, so
[the 665 AED saving can be attributed to the programme]{.flag}. [This makes a
strong case for national scale-up]{.flag}. As with any pilot, results should be
monitored as the programme expands.
:::
:::::

::::::

::: {.notes}
Paper (Oct12_session3_materials.docx): Hand out the AI handout (pack page 2, one per pair). The better prompt is printed below the flawed paragraph.
:::

## Debrief: four errors

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · Significant is not causal.** p < 0.001 says the fall is unlikely to be
chance. It says nothing about *what* caused it.
:::

::: {.warn}
**2 · Half right, which makes it convincing.** Following the same businesses
removes fixed differences between them. It does not remove what changed for
everyone: costs for businesses that did not take part rose by about 150 AED.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · It reported 665 as the effect.** That is the naive figure. The randomised
estimate is `r gfmt(abs(rct_est))` AED, and the prompt gave the model no way to
know that.
:::

::: {.warn}
**4 · It recommended scale-up without asking** what bar the programme has to
clear, or what 665 was compared to. The closing caveat is generic.
:::
:::::

::::::

::: {.panel}
**The prompt did some of the pushing.** It supplied a p-value, the word "cut" and
a request for the minister, so a model may write a confident summary instead of
questioning the design. People asked for a quick paragraph for a minister often
do the same.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What people actually type**

> Our pilot cut waste costs by 665 AED per business (before vs after,
> p < 0.001). Write a short paragraph for the minister.

The prompt frames the fall as the programme's doing, and the model may follow
that framing.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**What to type instead**

> Costs for participating businesses were 665 AED lower after the pilot than
> before (p < 0.001). We also have businesses that did not take part and a
> randomised comparison. The scale-up rule is at least 1,000 AED. Before writing
> anything, tell me what this before-and-after number can and cannot show.
> **Ask me questions before you answer.**
:::
:::::

::::::

::: {.ask}
Run both. Which one questions the 665 before writing the paragraph? What did
each explain well?

Two habits: make the AI **ask you questions first**, and have a **second chat**
review the first one's answer.
:::

::: {.notes}
The outline's version of this snapshot asked AI to explain why before-and-after
can mislead. Current models usually answer that general question well, so the
exercise uses a leading, realistic prompt instead; the better prompt still asks
the model to explain what a before-and-after number can and cannot show.
:::

# 7 · Close {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

## Your turn: ask AI first

::: {.ask style="font-size: 110%;"}
Before we show our list of questions to ask someone presenting an RCT, ask an AI tool:

**"I am a commissioner about to hear an evaluator present a randomised controlled trial (RCT) result. What are the key questions I should ask?"**
:::

::: {.panel}
Three minutes. Then compare with our list on the next slide: what did it miss, what did it add, and is anything incorrect?
:::

::: {.notes}
OPTIONAL. If time is short, skip this slide and go straight to the questions
on the next one. If you run it: give three minutes, then reveal the list and
ask what the AI missed, what it added that is worth keeping, and whether
anything it suggested was incorrect.
:::

## Three questions to ask the evaluator

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Before-and-after, with-and-without, or a fair comparison?</div>
  <div class="k">2</div> <div class="v">Show me the balance table. How do you know the groups started out alike?</div>
  <div class="k">3</div> <div class="v">Were errors clustered where you randomised, and does the interval clear our rule?</div>
</div>
```

::: {.panel}
Question 2 tells you whether the comparison is fair. Question 3 tells you
whether the result is enough to act on.
:::

::: {.notes}
Paper (Oct12_session3_materials.docx): Hand out the take-away cards (pack page 3). Give a minute to fill in "On my desk" on the card, then take two or three examples from the room.
:::

## What today gives you

```{=html}
<div class="facts">
  <div class="k">Before-and-after</div> <div class="v">folds in everything else that changed &middot; here it understated</div>
  <div class="k">With-and-without</div>  <div class="v">compares groups that already differed &middot; here it overstated</div>
  <div class="k">A lottery</div>         <div class="v">makes groups alike on average, and more so the more units it draws</div>
  <div class="k">Reading the result</div><div class="v">a difference of two averages; judge its interval against the rule</div>
</div>
```

::: {.panel}
Tomorrow: GreenWaste without the lottery. If the groups were not drawn, **was
the gap between them at least stable before the programme started?**
:::
