---
title: "Four statements about GreenWaste: decide what is missing"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    include-after-body: _session2_board.html
    header-includes: |
      <style>
      *{box-sizing:border-box}body{margin:0;font:18px/1.45 Arial,sans-serif;color:#111418;background:#eef2f6}main{max-width:900px;margin:auto}h1,h2,h3{color:#063360}h1{font-size:30px}h2{font-size:28px}.claim-card,.board-page{background:white;padding:38px;margin:24px 0;min-height:820px}.card-number{font-size:72px;font-weight:bold;color:#215a9e;margin:0}.claim-text{font-size:28px;line-height:1.45;border-top:3px solid #215a9e;padding-top:20px}.card-source{font-size:17px;color:#545860;border-top:1px solid #D5DEE8;padding-top:18px;margin-top:30px}.caption{font-size:14px;color:#545860}.choices{display:flex;gap:20px;flex-wrap:wrap;margin-top:25px}.choices span{border:1px solid #7da1c4;padding:8px 14px}.writing{border-bottom:1px solid #c8d6e5;height:65px}.triage{width:100%;table-layout:fixed;border-collapse:collapse;font-size:18px}.triage th,.triage td{border:1px solid #7da1c4;padding:12px;text-align:left;vertical-align:top}.triage td{height:100px}.triage thead{background:#063360;color:white}.placement{display:grid;grid-template-columns:80px 1fr 1fr;gap:10px;align-items:center;margin:12px 0}select,button{font:inherit;padding:7px;border:1px solid #7da1c4;color:#063360;background:white}.badge{display:inline-block;background:#e3ebf5;border:1px solid #215a9e;border-radius:4px;padding:4px 9px;margin:4px}button:focus-visible,select:focus-visible{outline:3px solid #215a9e;outline-offset:3px}@page{size:A4;margin:16mm}@media print{body{background:white;font-size:12pt}main{max-width:none}#title-block-header,.screen-controls{display:none}.claim-card,.board-page{padding:0;margin:0;min-height:0;break-after:page;box-shadow:none}.board-page{break-after:auto}h2{font-size:20pt}.card-number{font-size:50pt}.claim-text{font-size:19pt}.card-source{font-size:11pt}.caption{font-size:9pt}.triage{font-size:10pt}.triage th,.triage td{padding:7px}.triage td{height:27mm}.writing{height:16mm}.choices{font-size:12pt}}</style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('session2_review.R', encoding='UTF-8')
```

<div class="screen-controls"><p>Read each statement about GreenWaste and decide whether to act on it.</p><button onclick="window.print()">Print cards and board</button><p><a href="Oct12_session2_review_worksheet.html">Open the worksheet and H2-A sources</a></p></div>

```{r, results='asis'}
html_out(paste(vapply(1:4,function(i) paste0('<section class="claim-card"><p class="card-number">',i,
  '</p><h2>What would you do with this statement?</h2><p class="claim-text">',s2_cards[i],
  '</p><p class="card-source"><strong>Source:</strong> ',s2_card_sources[i],
  '</p><div class="choices"><span>Act</span><span>Ask first</span><span>Do not act</span></div>',
  '<p><strong>Compared to what? How big? How sure?</strong></p><p>What is missing, and why would the answer matter?</p><div class="writing"></div><div class="writing"></div>',
  if(i==1)'<p class="caption">GreenWaste is a fictional training case. These statements are exercise material for you to check. Do not accept them as conclusions.</p>' else '',
  '</section>'),''),collapse=''))
```

<section class="board-page">

## Place each card, then explain your reason.

Place each card under Act, Ask first or Do not act. Explain your reason, using words and card numbers. More than one missing question can be a good answer. Say which action you mean: acting on a descriptive finding does not have to mean expanding the programme.

<div class="screen-controls" id="placements">

```{r, results='asis'}
html_out(paste(vapply(1:4,function(i)paste0('<div class="placement"><strong>Card ',i,
  '</strong><select aria-label="Decision for card ',i,'" data-decision="',i,
  '"><option value="">Choose decision</option><option>Act</option><option>Ask first</option><option>Do not act</option></select>',
  '<select aria-label="Missing question for card ',i,'" data-question="',i,
  '"><option value="">Choose main question</option><option>Compared to what?</option><option>How big?</option><option>How sure?</option></select></div>'),''),collapse=''))
```

</div>

<table class="triage"><thead><tr><th>Missing question</th><th>Act</th><th>Ask first</th><th>Do not act</th></tr></thead><tbody>
<tr><th>Compared to what?</th><td data-cell="0,0"></td><td data-cell="0,1"></td><td data-cell="0,2"></td></tr>
<tr><th>How big?</th><td data-cell="1,0"></td><td data-cell="1,1"></td><td data-cell="1,2"></td></tr>
<tr><th>How sure?</th><td data-cell="2,0"></td><td data-cell="2,1"></td><td data-cell="2,2"></td></tr>
</tbody></table>

<p id="board-status" role="status" aria-live="polite"></p>

Which evidence supports a disputed placement? What answer would change your advice?

<div class="writing"></div>
</section>
