---
title: "GreenWaste: read the economic model"
format:
  html:
    embed-resources: true
    toc: false
    theme: none
    include-in-header:
      text: |
        <style>
        body{font:16px/1.4 Arial,sans-serif;color:#111418;margin:0;background:#eef2f6}main{max-width:850px;margin:auto}h1,h2,h3{color:#063360}h1{font-size:26px}h2{font-size:22px;margin:0 0 12px}h3{font-size:18px;margin:14px 0 6px}p{margin:8px 0}.sheet-page{background:white;padding:28px 40px;margin:20px 0;box-sizing:border-box;min-height:950px}.reading-table{width:100%;border-collapse:collapse;font-size:13px}.reading-table th,.reading-table td{padding:6px;border-bottom:1px solid #D5DEE8;text-align:left}.reading-table thead{background:#063360;color:white}.writing{height:45px;border-bottom:1px solid #c8d6e5;margin:6px 0}.writing.tall{height:70px}.review-svg{display:block;width:100%;height:130px}.caption{font-size:12px;color:#545860}.ai-source{border-left:3px solid #7da1c4;padding-left:16px}.prompt{border-top:2px solid #215a9e;padding-top:8px;font-size:14px}header button{font:inherit;padding:8px 15px;color:#063360;background:white;border:1px solid #215a9e;cursor:pointer}
        @page{size:A4;margin:14mm}@media print{body{background:white;font-size:11pt;line-height:1.3}main{max-width:none}.sheet-page{padding:0;margin:0;min-height:0;break-after:page}.sheet-page:last-child{break-after:auto}header,#title-block-header{display:none}h2{font-size:17pt}h3{font-size:12pt}.writing{height:10mm}.writing.tall{height:16mm}.reading-table{font-size:9pt}.reading-table th,.reading-table td{padding:5px}.review-svg{height:115px}.caption{font-size:9pt}.prompt{font-size:10pt}}
        </style>
        <style>.writing.part{height:30px;margin:2px 0}.table-title{margin:10px 0 4px}@media print{.writing.part{height:8mm}}</style>
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r, include=FALSE}
source('day3_session1_review.R', encoding='UTF-8')
```

<header><button onclick="window.print()">Print worksheet</button></header>

<div class="sheet-page">

## Page 1 | H7-A: The economic model summary

GreenWaste is a fictional training case. Use this model summary as your source for today's questions.

### §1 | Where does the benefit input come from?

The city difference-in-differences study compares the change in annual waste management cost in 4,794 participant businesses with the change in 5,206 comparison businesses that did not receive GreenWaste. The difference between the two changes is a reduction of **`r fmt(model$saving)` AED per participant business**, measured 12 months after GreenWaste. To call this a causal saving, we need the parallel trends assumption and the other conditions for the comparison. The estimate is provisional and is not a guarantee. The other methods compare different groups of businesses in different ways.

```{r, results='asis'}
html_out(d7_inputs('Table I: Inputs to the model and where each comes from'))
```

### §2 | How are future amounts combined?

The model assumes that equipment delivery costs **1,800 AED per participant business now (year 0)**, that there is no saving in year 1, and that the same annual saving is made in each of **years 2 to 6**. Future savings are discounted at an assumed rate of 5% a year, so that every amount is shown at its value in year 0 (its value today). The city data do not show five years of savings. They show only the first 12 months. For year *t*, the value today of the saving is the annual saving divided by 1.05 raised to the power *t*. Table F and Table O are on page 2. The calculations use the exact inputs, and the figures shown are rounded.

### §3 | Whose perspective, and which costs and benefits?

This model mixes two points of view. It compares a cost to the authority (equipment delivery) with savings to businesses. It is not a full appraisal of the authority's budget, and it is not a full social cost-benefit analysis.

**Table P: Costs and benefits in the model: who bears each one and whether the model includes it**

| Who and what | Included in the model? | Evidence we would need |
|---|---|---|
| Authority: equipment delivery | Yes, as an assumed cost at year 0 | Amounts and coverage from the programme's records |
| Authority: staff and administration | Not known | What is already included, the amount and the timing |
| Business time and maintenance; wider effects | Not known | The point of view, the resource costs and any transfers between actors |

Check that a cost is not already counted before you add it. Financial savings are not automatically net benefits to society. Charges, transfers and wider effects need separate treatment.

### §4 | Sensitivity and the limits of the headline result

The scenarios change the model's inputs. They do not change any observed fact about the programme. The settings for each scenario are on page 2. The scenario results table is given out after your group records its prediction. The results for Cards 1 to 4 depend on the ranges we tested, because each of those cards changes one input. Card 5 changes two inputs together. We have no evidence about how realistic these settings are.

An extra check shows that a discount rate of 40% would give a ratio of **`r fmt(ratio(rate=.4),2)`**, with all other inputs unchanged. Delaying the savings by three more years (years 5 to 9) at 5% gives a ratio of **`r fmt(ratio()/1.05^3,2)`**, which is still above 1. These checks do not show which discount rate the government would use, or how long a delay is realistic. A ratio above 1, or a positive net present value, depends on the model and does not prove value for money. Our rule of at least a 1,000 AED reduction in annual waste management cost is a different test and is not a benefit-cost ratio threshold.

<p class="caption">Source: GreenWaste city data and the economic model. The model inputs, the cost boundary and the scenarios are fictional. Paragraphs §1 to §4 and Tables I, P, F and O, and the scenario results table, are the material to inspect.</p>

</div>

<div class="sheet-page">

## Page 2 | H7-A: Costs and savings by year, base case results and scenario settings

```{r, results='asis'}
html_out(d7_timeline())
```

```{r, results='asis'}
html_out(d7_cash_table('Table F: Delivery cost and annual saving per participant business in each year, with each saving converted to its value today (AED)'))
```

One-year example: 812 divided by 1.05 is about 774 AED. The first saving in the model is in year 2: 812 divided by 1.05² is about **`r fmt(model$saving/1.05^2)` AED**. Add the values today for years 2 to 6. The rounded figures in the table may not add up exactly to the exact total.

```{r, results='asis'}
html_out(d7_output('Table O: Base case results per participant business: total benefits, cost, benefit-cost ratio and net present value (AED)'))
```

The benefit-cost ratio is benefits divided by costs. It has no units. The net present value is benefits minus costs, in AED per participant business. Neither number shows that GreenWaste caused the savings, that the savings will last, or that all costs are included.

### Scenario settings: write your prediction before you see the results

- **Card 1: Savings fade.** After the first year of saving, each year's saving is 30% lower than the year before.
- **Card 2: Hidden costs.** We add 600 AED of costs at year 0 that are not already counted.
- **Card 3: A higher discount rate.** The rate is 8%.
- **Card 4: A shorter life.** Savings last three years (years 2 to 4).
- **Card 5: Savings fade and hidden costs.** This card changes the inputs of Cards 1 and 2 together.

On every card, all other inputs stay at the base values. Keep this page closed while your group makes its prediction. The scenario results table is given out after your group has recorded its prediction and its reason in Q4.

</div>

<div class="sheet-page">

## Page 3 | Q1 to Q3: inputs, timing and whose point of view

### Q1 | What goes into the model? (§1 and §2, Table I)

(a) Name one input that comes from the city study and one that is an assumption, and write where each appears.

<div class="writing part"></div>

(b) In which year does the cost occur? ______

(c) In which year do savings start, and in which year do they end? Start: ______ End: ______

(d) How many years of savings are there? ______

(e) Is there a saving in year 1? ______

(f) What evidence would you need before you trust the future savings?

<div class="writing part"></div>

### Q2 | Value today, ratio and net present value (Tables F and O)

(a) Calculate the year 2 saving at its value today: 812 divided by __________ is about __________ AED.

(b) Why can we not simply multiply the annual saving by five?

<div class="writing part"></div>

(c) Write the benefits at value today: __________ AED. Write the cost at value today: __________ AED.

(d) Show how the ratio is calculated: ____________________

(e) Show how the net present value is calculated: ____________________

(f) Explain what each of these two numbers means, under the model's assumptions.

<div class="writing part"></div>

### Q3 | Whose point of view? (§3, Table P)

(a) Whose decision is this?

<div class="writing part"></div>

(b) Which costs and benefits belong to that point of view?

<div class="writing part"></div>

(c) Name one cost that may be missing.

<div class="writing part"></div>

(d) What might already be counted in the model?

<div class="writing part"></div>

(e) What amount and timing would you request, without counting the cost twice?

<div class="writing part"></div>

</div>

<div class="sheet-page">

## Page 4 | Q4 to Q6: scenarios and your recommendation

### Q4 | Predict before you see the results

(a) Write your card number and the input it changes. Card: ______ Input: ____________________

(b) Which inputs stay the same?

<div class="writing part"></div>

(c) Predict the benefit-cost ratio. Will it be above or below 1? Give your reason. After your group has recorded its prediction, you will receive the scenario results table.

Prediction: ________ Above or below 1? ________

<div class="writing part"></div>

(d) Write the ratio in the table: ________. What explains the difference from your prediction?

<div class="writing part"></div>

(e) What evidence would show whether this change is realistic?

<div class="writing part"></div>

(f) Among Cards 1 to 4, which card lowers the ratio most? Why does this not give a ranking of the assumptions in general?

<div class="writing part"></div>

### Q5 | Read the model yourself (H7-A, pages 1 and 2)

(a) Write one input that comes from the city study and one assumption, and where each appears.

<div class="writing part"></div>

(b) What does the ratio support, under these assumptions?

<div class="writing part"></div>

(c) What does it not show?

<div class="writing part"></div>

### Q6 | Close the source and give a conditional recommendation

(a) Explain what a benefit-cost ratio of 1.86 means.

<div class="writing part"></div>

(b) State your provisional action and one condition.

<div class="writing part"></div>

(c) Request one specific missing fact, and explain how its answer could change your decision.

<div class="writing part"></div>

</div>

<div class="sheet-page">

## Page 5 | AI exercise: check the scenario

### Stage 1: Mark the response

Naive prompt (a prompt with no context, as a busy person might type it): "Generate one plausible scenario where this programme's benefit-cost ratio falls below 1. Tell me whether it is realistic." Source: H7-A. This response was written for the exercise. It is not a real AI answer.

<div class="ai-source">
```{r, results='asis'}
html_out(paste0('<p>', gsub('\n\n', '</p><p>', d7_ai, fixed=TRUE), '</p>'))
```
</div>

Keep the claims that H7-A supports. Cross out the arithmetic, timing or realism claims that it does not support. Write one source reference that changes the advice.

<div class="writing part"></div>

### Stage 2: Run the improved prompt in Microsoft Copilot

<div class="prompt">
```{r, results='asis'}
html_out(paste0('<strong>Improved prompt</strong><p>', d7_prompt, '</p>'))
```
</div>

Use only the information in H7-A, and keep your AI good-practice card beside you. If you are not using Copilot, read the prepared response on page 6 after you finish Stage 1.

(a) Write one number you verified.

<div class="writing part"></div>

(b) Write one useful question that the response asked about missing evidence.

<div class="writing part"></div>

</div>

<div class="sheet-page">

## Page 6 | Prepared response

Use this response if you are not using Copilot. Read it after you have marked the response on page 5. It was written for the exercise. It is not a real AI answer. An AI tool may answer differently.

### Separate inputs from conclusions

H7-A §1 and Table I use an **estimate from the city study** of about 812 AED, provisionally treated as a causal benefit. The cost, the one-year lag, the five-year stream of savings and the 5% rate are **assumptions**, not measured economic facts. Future benefits are modelled from them.

Tables F and O put costs at year 0 and savings in years 2 to 6. The benefits at value today are about **3,350 AED**, the cost is **1,800 AED**, the ratio is **1.86** and the net present value is **1,550 AED per participant business**. These numbers do not verify that GreenWaste caused the savings, how long they last, or that all costs are covered. Table P has a limited boundary that mixes two points of view. It is not a complete appraisal of the authority's budget or of society.

### Quantify the stated changes

The scenario results table gives ratios of **1.07, 1.40, 1.67 and 1.17** for Cards 1 to 4. Compared with the base ratio of 1.86, the largest fall among those settings is the fading savings scenario (about **0.79** ratio units), then the shorter life (**0.69**), the added costs (**0.47**) and the rate of 8% (**0.19**). The ranking depends on the stated changes. Card 5 combines fading savings and added costs and gives **0.80**. It is not a test of one input.

At 40% the arithmetic can produce a ratio below 1 (§4). The model summary does **not** show that this rate is used in government practice. A delay of three more years gives a ratio of about **`r fmt(ratio()/1.05^3,2)`**, which is not below 1. A stress test with a high rate does not show that the programme should be rejected. Suitable guidance and other evidence are needed.

### Questions before recommending an input to investigate

- Whose decision is this, and which point of view should the appraisal take?
- What evidence on lasting results and maintenance supports five years of savings or a 30% fall each year?
- What costs, timing and existing coverage support the extra 600 AED?
- Which discount-rate guidance applies, and how credible is the provisional difference-in-differences benefit?

### Verify this response yourself

(a) Check one number against a named table.

<div class="writing part"></div>

(b) Write one request for evidence that could change the spending decision, and say why.

<div class="writing part"></div>

</div>
