---
title: "Plan Your Own Evaluation"
subtitle: "Module 2 · Day 4, Session 3: Building on Module 1"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [theme_dge.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
editor_options:
  chunk_output_type: console
---

```{r setup}
library(knitr)

# ---- Tables: plain knitr::kable, styled by theme_dge.scss -------------------
# No kableExtra: its JavaScript needs jQuery, which revealjs does not load.
# All columns left-aligned (text tables); font_size in px; table sits left.
tbl <- function(df, ..., font_size = 16) {
  h <- as.character(knitr::kable(df, format = "html", align = "l", ...))
  h <- sub("<table", sprintf('<table style="margin-left:0; font-size:%dpx;"', font_size),
           h, fixed = TRUE)
  cat(h, "\n")
}

# ---- Palette (DGE theme; see theme_dge.scss) --------------------------------
ACCENT  <- "#215a9e"; ACCENT2 <- "#7da1c4"; ALERT <- "#B8272C"
INK     <- "#111418"; MUTED <- "#545860"; RULE <- "#D5DEE8"

# ---- The template ----------------------------------------------------------
# Shared with the printable Evaluation Design Update template.
source("qa_design_update.R")
```

# 0 · The last session {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 5 minutes.
:::

## Four days. One design.

::: {.panel}
On Monday you learned what a comparison is. Tuesday, three more ways of building one.
Wednesday, what a result costs and what a figure hides. Thursday, how to tell
whether evidence is credible and how to translate it.

Today you apply it to your own evaluation. By the end: one completed template
per team and a five-minute pitch. Part 1, 20 minutes; Part 2, 30 minutes;
pitches from about 85 minutes in.

Pick one member's Module 1 design, the most developed one.
:::

::: {.ask .fragment}
**How far along is your design?**
Not started / rough outline / drafted / ready to pitch.
:::

::: {.notes}
Timing for 120 minutes with three teams: opening 5, Part 1 30 (10 of slides, 20 of work), Part 2 45 (10 of slides, 30 of work, 5 of break), pitches 30 (5 of set-up, 3 teams at 8 minutes), close 10. More than three teams: cut each pitch to 3 minutes plus 2 of feedback, or Part 2 to 25 minutes.
:::

# 1 · Where your design stands {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 30 minutes: 10 of slides, then 20 of working time.
:::

## The eight conditions, from Module 1

```{r}
#| results: asis
tab <- data.frame(
  `#` = seq_len(nrow(design_conditions)),
  Condition = design_conditions$condition,
  check.names = FALSE
)
tbl(tab, col.names = c("#", "Condition"),
    caption = "The eight conditions for a credible impact evaluation")
```

::: {.panel}
These are Module 1's eight conditions. Part 1 of the template asks you to write your design against these, so the new
work has something to attach to.
:::

## The one condition that decides the rest

::: {.ask}
Of the eight, which one matters most for a credible estimate?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
**The counterfactual.**

Every method this week has been a different way of operationalising it:
randomisation, a cut-off, a matched comparison, a parallel trend.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Each one has its own way of failing. If you cannot state yours in one sentence,
the design is not finished.
:::
:::::

::::::

::: {.ask .fragment}
What is your counterfactual, in one sentence? Write it now; you will read it out
in the pitch.
:::

## Working time

::: {.panel}
**Twenty minutes.** Part 1 of the template: write your design as it stands,
against the eight conditions.

Roles: scribe, timekeeper, presenter.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Which member's Module 1 design is most developed? Work from it. With no document
to hand, the recap in each template row is enough: write what you would do.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
Do not polish. This is the foundation for Part 2, and Part 2 is what the pitch focuses on.
:::
:::::

::::::

::: {.notes}
Paper (Oct15_session3_materials.docx): Hand out the Evaluation Design Update (pack pages 1 to 3), one per team. Teams start on Part 1 (pack page 1).
:::

# 2 · The four additions {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 45 minutes: 10 of slides, 30 of working time, then the 5-minute break.
:::

## What you are adding

```{r}
#| results: asis
tab <- data.frame(
  Addition = design_additions$addition,
  `From` = design_additions$from,
  check.names = FALSE
)
tbl(tab, col.names = c("Addition", "From"),
    caption = "The four additions, and where each came from")
```

::: {.panel}
Each addition applies one session from this week to your design.
:::

::: {.ask .fragment}
Which of the four is hardest to do **before** you have any data? Why?
:::

## Addition 1: results presentation plan

::: {.ask}
What will you show, to whom, and in what order? Name the one number you will lead
with, and say what the reader must **not** be able to conclude from it.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**This morning's lesson, applied**

The fictitious report led with its largest estimate and called a different one its
preferred specification. A reader who stopped after the summary would have
believed the threshold was met, when the preferred estimate did not meet it.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
So ask: "What will a reader who reads only the headline come away believing,
and is that true?"
:::
:::::

::::::

## Addition 2: cost-benefit framework

::: {.ask}
What are your **decision rule**, **benefit per unit per year**, **horizon** and
**discount rate**? Which of these choices changes the answer most?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

In the GreenWaste example the horizon moved the ratio more than the discount
rate did, and reports often leave the horizon unstated.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
Write your horizon down. In GreenWaste, cutting the horizon from five years to
two moved the ratio from 1.87 to 0.80.
:::
:::::

::::::

## Addition 3: data visualisation sketch

::: {.ask}
What figure carries your headline result? Sketch it, then sketch the **same data
presented a second way**. What does the second version hide?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Wednesday's lesson**

Each pair of figures showed identical numbers and different impressions: the
truncated axis, the bar of means, the denominator, levels versus changes,
discounting.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second sketch matters more. If you cannot say what your preferred figure
hides, expect it to come up in questions.
:::
:::::

::::::

## Addition 4: evidence translation plan

::: {.ask}
What is the **headline sentence** of your brief, with the number in it? What is
the one thing your evaluation did **not** establish, stated plainly?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**This morning's lesson**

A brief that reports the estimate and omits the threshold leaves the reader to
assume the target was met.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
The second half is harder. Naming your own limit makes the pitch more
credible.
:::
:::::

::::::

## Working time

::: {.panel}
**Thirty minutes.** Part 2 of the template: all four additions.

Work as a team. Draft all four; develop two in depth.
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.ask}
Which two additions will your team develop in depth?
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Checkpoint at 15 minutes:** each team reads out its counterfactual sentence.
:::
:::::

::::::

::: {.notes}
Paper (Oct15_session3_materials.docx): Teams turn to Part 2 of the Design Update (pack page 2).

At 15 minutes, stop the room for the checkpoint: each team reads out its counterfactual sentence. If a sentence does not name who the comparison group is, say so and move on.
:::

## Stress-test your design with AI

::: {.panel}
Type your Part 1 answers and your counterfactual sentence into the AI tool. Prompt:

> "You are a sceptical commissioner. Ask me three questions about this design
> before you comment. Then name the one assumption that, if it broke, would make
> the estimate wrong."
:::

::: {.warn}
Paste the AI's answer into a new chat and ask the AI to check that answer.
:::

::: {.notes}
Paper (Oct15_session3_materials.docx): Hand out the AI prompt cards (pack page 4), one per team. This runs inside the Part 2 working time; one team member runs it while the others keep drafting.
:::

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the pitches.
:::

# 3 · The pitch {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 30 minutes: 5 of set-up, then three teams at 5 minutes of pitch and 3 of feedback.
:::

## Five minutes, five parts

```{r}
#| results: asis
tab <- data.frame(
  Part = pitch_structure$part,
  Time = pitch_structure$minutes,
  `What it says` = pitch_structure$says,
  check.names = FALSE
)
tbl(tab, col.names = c("Part", "Time", "What it says"),
    caption = "The pitch structure", font_size = 15)
```

::: {.warn}
Note the order. **The limit comes before the ask.**
:::

::: {.notes}
Paper (Oct15_session3_materials.docx): Teams turn to Part 3 of the Design Update (pack page 3) to plan the pitch.
:::

## Why the limit comes before the ask

::: {.ask}
If you tell the room what your design cannot do, why would that make them more
likely to give you what you want?
:::

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel .fragment}
Because it tells them **which parts you have already thought about**. A pitch with no
stated limits leaves the room to guess them.
:::
:::::

::::: {.column width="50%"}
::: {.warn .fragment}
It also changes what you are asking for. If you know your design cannot establish
X, you can ask for what would, which is a more specific request than "fund the
evaluation".
:::
:::::

::::::

## Feedback from the room

::: {.panel}
Three minutes per team. The room gives feedback in this order:

1. **What is the design?** Say it back in one sentence. If you cannot, that is the
   feedback.
2. **What would make the design fail?** Name the one assumption that would have to break.
3. **What is missing?** One thing you would have wanted to hear.
4. **Which mistake from this week could affect this design?** For example:
   headline vs preferred estimate, time horizon, truncated axis, missing threshold.
:::

::: {.warn}
**Write your feedback down**, so the team can use it. Comment on the design, not on the people.
:::

::: {.notes}
The first question is a check: if the room cannot say the design back in one
sentence, the pitch did not land. That is useful feedback.

Paper (Oct15_session3_materials.docx): Hand out the feedback slips (pack page 5), one slip per participant for each other team. Call time at 5:00; feedback in the fixed order.
:::

# 4 · The week in one slide {background-gradient="linear-gradient(240deg, #7da1c4 0%, #7da1c4 22%, #215a9e 58%, #063360 100%)"}

::: {.notes}
About 10 minutes.
:::

## What you can now do

```{r}
#| results: asis
cat('<div class="panel"><b>1.</b> Say what a number is compared to, and why that comparison matters.</div>\n')
cat('<div class="panel"><b>2.</b> Read DiD, RDD and matching results, and name the assumption each rests on.</div>\n')
cat('<div class="panel"><b>3.</b> Judge whether a benefit-cost ratio clears a threshold, and what decides it.</div>\n')
cat('<div class="panel"><b>4.</b> See the same data drawn two ways, and say what each version hides.</div>\n')
cat('<div class="panel"><b>5.</b> Interrogate an evaluation with a checklist, and rate a report section by section.</div>\n')
cat('<div class="panel"><b>6.</b> Write a brief a decision-maker can act on without being misled.</div>\n')
```

::: {.ask .fragment}
Which of the six would you have said you could do on Monday morning? Which would
you now add?
:::

## The one question

::: {.panel}
The question behind every session this week:

> **What is being compared to what, and why should those two things be
> comparable?**
:::

::: {.ask .fragment}
Which programme in your unit could you ask this about next week?
:::

::: {.notes}
Paper (Oct15_session3_materials.docx): Hand out the take-away cards (pack page 6), one per participant. Give a minute to fill in the card's last line (the first evaluation decision back at work), then take two or three examples on the thank-you slide.
:::

## Thank you

::: {.panel}
The slides and handouts stay available on the course website.
:::

::: {.ask}
What is the first evaluation decision you will make when you get back to your
desk?
:::
