---
title: "Did the programme make the difference?"
subtitle: "Day 1 | Session 3: A pilot lottery, its uncertainty and the limits of expansion"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    css: [session1_review.css, session3_review.css]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: [_reader_interactions.html, _session1_review_interactions.html, _session3_review_interactions.html]
webr:
  render-df: kable
  resources: [evaluation_data_GreenWaste_simple.csv]
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('session3_review.R', encoding='UTF-8')
```

## Who and what are we comparing? The pilot district. {#s3-map .reader-slide .s1-review .s3-review .case-map-slide}

```{r, include=FALSE}
map_highlight <- "pilot"
```

::: {.case-map}
{{< include _case_map.qmd >}}
:::

<div class="map-facts"><div><strong>Unit</strong>one business</div><div><strong>Group</strong>400 businesses in the pilot district: 200 treatment group, 200 control group, chosen by lottery</div><div><strong>Outcome</strong>annual waste management cost, in AED per business</div><div><strong>Period</strong>before GreenWaste and 12 months after</div></div>

::: {.notes}
1 minute. Point to the pilot district only. The lottery is what makes this study different from the city comparisons. The same map appears in every methods session so that participants know the population before the details start.
:::

## We have two city comparisons. Why would a lottery be better? {#s3-p01 .reader-slide .s1-review .s3-review}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
pilot <- subset(gw, setting == 'pilot')
ages <- pilot$manager_age[1:20]
draw <- 0
```

<div class="comparison-pair"><div><h3>Before and after: −669 AED</h3><p>The same participant businesses, measured over time.</p><p>What would have changed without GreenWaste?</p></div><div><h3>With and without: −1,638 AED</h3><p>Participant businesses and comparison businesses, after a year.</p><p>Were the groups already different?</p></div></div>

::: {.ask}
What if chance, rather than the score rule, decided which businesses took part?
:::

Use the summary of the two comparisons from Session 2. Now we move to the **pilot district**.

::: {.notes}
4 minutes. One minute recall, two minutes pairs explain the two missing comparisons, one minute bridge. No new city arithmetic or repeat baseline reveal. Pilot is a different population, not the city's known true effect. Start webR setup early for one optional trainer draw. Printed activity is independent of internet. Transition: there were 400 eligible businesses but only 200 places.
:::

## The pilot had 400 eligible businesses but only 200 places in the programme. {#s3-p02 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_flow())
```

**H3-A, the pilot lottery study summary, §1 and §2 on worksheet page 1:** the outcome is annual waste management cost, measured before GreenWaste and 12 months after. A place is a place in the programme. 200 of the 400 businesses receive GreenWaste (the treatment group) and 200 do not (the control group). A lottery decides which.

::: {.ask}
The lottery assigns **individual businesses**. It does not draw a sample of the country.
:::

::: {.notes}
5 minutes. Read H3-A1/2 source paragraphs. Exactly 200/200, not roughly half; each eligible business has a place in join or wait. All scores at/below 58; city rule has not allocated these pilot places. Distinguish choosing the pilot district from randomly assigning inside it. Ask who the comparison will describe. Transition: why does the method of allocation matter?
:::

## What does a lottery give us? {#s3-p03 .reader-slide .s1-review .s3-review}

<div class="comparison-pair"><div><h3>If managers choose</h3><p>More motivated managers might join first. That could also affect their costs.</p></div><div><h3>If the lottery chooses</h3><p>Managers' motivation does not decide which eligible business gets a place.</p></div></div>

The first example is **hypothetical**. Over many possible lotteries, random assignment does not favour one kind of business.

::: {.ask}
**Q1 on page 3:** Why does the control group give a more reliable comparison than businesses that chose whether to join?
:::

::: {.notes}
7 minutes. Explain random assignment/RCT in plain language. Two minutes write Q1, then discuss. Chance balances observed and unobserved characteristics in expectation; one draw need not be identical. Design still needs implementation, comparable measurement/follow-up and no important effects on the control group. Source §3 explicitly states assignment matches participation and all 400 costs observed in this fictional pilot. Avoid a long validity checklist. Transition: experience the difference between chance and perfect balance.
:::

## Draw the lottery, then draw it again. {#s3-p04 .reader-slide .s1-review .s3-review}

This practice lottery has **20 businesses and 10 places**. The slips show **manager ages before the programme**.

Draw ten age slips for the treatment group. The other ten slips are the control group. Compare the mean manager age of the two groups. Put all the slips back, and draw again.

**R code**

```{webr}
#| echo: true
# Each run makes a new practice lottery.
draw <- draw + 1
set.seed(draw)
# Ten businesses go to the treatment group and ten to the control group.
group <- sample(rep(c('Treatment', 'Control'), each = 10))
# Compare the mean manager ages. These are not programme effects.
round(tapply(ages, group, mean), 1)
```

::: {.ask}
**Q1 (a):** Did both lotteries give exactly the same mean ages?
:::

::: {.notes}
12 minutes. Change of activity, not an extra break: distribute one cut slip set per group, two minutes assign roles/read, three minutes draw/calculate, three minutes replace/redraw, two minutes reflection, two minutes optional R/debrief. Physical slips use the first 20 pilot manager-age records; this is not the allocation used in the pilot. A reader checks IDs and a recorder logs both means on page 3. The trainer run repeats the same pool with a different fixed seed; it need not reproduce a physical draw. Code comments explain operations; no participant code. A calculator or total/10 is enough. If runtime unavailable complete paper route within the same time. Transition: what might change with more businesses?
:::

## Small lotteries can look different {#s3-p05 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<div id="s3-sizes"><div data-variant>',s3_sizes(),'</div><div data-variant hidden>',s3_sizes(TRUE),'</div></div>'))
```

<button data-swap="s3-sizes" data-closed="Show 400-business practice lotteries" data-opened="Show small lotteries only" aria-pressed="false">Show 400-business practice lotteries</button>

In this illustration, we ran **200 practice lotteries of each size** using manager ages from the pilot records. The differences shown are not programme effects.

::: {.ask}
Predict first: would the larger groups usually be closer in mean age?
:::

::: {.notes}
6 minutes. Each dot is the treatment mean age minus the control mean age in one practice allocation; same horizontal scale. For N=20, take a random 20-record subset from the full age pool each time; for N=400 use full pool. This isolates sample-size variation; the physical slips used a fixed selected 20. Explain the display, not probability formulas. Reveal large draws and have Q1 add a sample-size reflection. Larger groups usually closer, no guarantee for a specific draw; no p-value balance proof. Transition: inspect the pilot baseline rather than a redraw.
:::

## Were the treatment and control groups similar before GreenWaste? {#s3-p06 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_baseline())
```

The figure shows the two pilot groups as assigned: 200 businesses in the treatment group and 200 in the control group. The distributions are similar, but the groups are not identical.

::: {.ask}
What reassurance does this give? What does it not prove?
:::

::: {.notes}
6 minutes. Source H3-A §4 and figure, page 1. Means 1,390/1,411; side-by-side bars in each 250 AED band on common axes. Listen for starting distributions similar and small mean difference plausible under chance. No balance p-value, no significance testing to prove randomisation. Similar costs alone cannot establish unobserved balance or correct allocation; check assignment procedure in a real report. Add Q1 baseline reflection. Transition: the outcome comparison is after the programme.
:::

## The result is the difference between two averages. {#s3-p07 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_after())
```

::: {.ask}
**Q2 on page 3:** The treatment group's mean cost minus the control group's mean cost is 769 − 1,783 ≈ **−1,014 AED**.
:::

The treatment group's mean annual waste management cost was lower, **12 months after GreenWaste**, among the eligible pilot businesses.

::: {.notes}
8 minutes. H3-A findings on page 2. First axes/population/period, then two minutes calculate/interpret, then three minutes pair check and feedback. Unrounded 769.125-1783.48=-1014.355; display -1,014. Negative cost difference means positive saving. It is an after-period randomised group comparison, not participants before/after. Assignment equals participation here. Transition: the simple regression expresses this same comparison.
:::

## We create the regression table in R with one line of code. {#s3-p07b .reader-slide .s1-review .s3-review}

We create the table in R. The box shows the **R code** and the numbers it produces.

```{r, echo=TRUE}
# pilot has one row per business (400 rows).
# took_part is 1 for the treatment group and 0 for the control group.
# cost_after is annual waste management cost 12 months after GreenWaste (AED).
fit <- lm(cost_after ~ took_part, data = pilot)
# Row 1 (Intercept): the control group's mean cost.
# Row 2 (took_part): treatment group minus control group.
round(cbind(estimate = coef(fit), confint(fit)))
```

::: {.ask}
Which line of the code gives the control group row, and which gives the comparison row?
:::

You do not need to write or run R. Q2 on your worksheet asks you to read the table.

::: {.notes}
4 minutes. Walk through the comments, not the syntax; participants do not code. Match each printed row to a row of the table on the next slide. Estimates: 1,783 and -1,014; interval -1,167 to -862. This is the same two-group comparison as the previous slide.
:::

## The regression table reports the same comparison. {#s3-p08 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(sub('<p class="caption">.*$', '', s3_regression()))
```

<div class="review-annotate"><button data-table-column="1" aria-pressed="false">Coefficient</button><button data-table-column="2" aria-pressed="false">Interval</button><button data-table-column="3" aria-pressed="false">P-value</button></div>

::: {.ask}
**Q2:** Which row shows the comparison? What does the control group row show?
:::

1,783 + (−1,014) ≈ **769 AED**, which is the treatment group's mean cost. The lottery design supports reading the difference as an effect. The p-value alone does not.

::: {.notes}
6 minutes. Annotate H3-A table on page 2. Control group intercept is the mean cost after 12 months for the reference group; treatment coefficient is the treatment minus control difference. CI -1,167 to -862; ordinary least squares model matches canonical pilot source. Point to code/model note for mechanics, no need fit models. Show each column with native highlights. A causal reading depends on assignment/implementation/measurement and interference assumptions, not regression magic. Transition: a convincing saving does not settle the exact policy minimum.
:::

## Is the whole interval at or above our rule of a 1,000 AED reduction? {#s3-p09 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_interval_svg('trial',TRUE))
html_out('<button data-open="interval-trial" data-closed="Show the 95% interval for the reduction" data-opened="Hide the interval" aria-controls="interval-trial" aria-expanded="false">Show the 95% interval for the reduction</button>')
```

::: {.ask}
**Q3 on page 4:** Does evidence of some reduction in cost show that the reduction is at least **1,000 AED**?
:::

Our rule is a reduction in annual waste management cost of at least 1,000 AED per business per year. GreenWaste is a fictional case, so the rule is fictional too.

::: {.notes}
8 minutes. Predict from point estimate before reveal: 1,014 just above 1,000; reveal positive saving interval 862-1,167, which spans rule. Ask pairs for two statements: evidence of positive saving, uncertain whether minimum met. CI describes mean-effect uncertainty under model, not a probability statement about this fixed interval or individual businesses' savings. Convert negative effect interval by reversing endpoints. No value-for-money conclusion because full costs absent here. Transition: uncertainty is one limit; coverage is another.
:::

## Who does this pilot describe? {#s3-p10 .reader-slide .s1-review .s3-review}

<div class="comparison-pair"><div><h3>What the pilot covers</h3><p>One pilot district.<br>Businesses with scores of 58 or below.<br>Support delivered under pilot conditions.</p></div><div><h3>What wider use could change</h3><p>Other businesses and districts.<br>Delivery capacity and take-up.<br>Programme costs.</p></div></div>

::: {.ask}
**Q4:** What evidence would you request for the expansion being proposed?
:::

A lottery allocates places within the pilot. It does not make the pilot representative of the country.

::: {.notes}
6 minutes. Source §1/3/5. Two minutes specific reach request, then compare. Accept profile/cost/outcome differences, delivery capacity or representative next-stage evidence with purpose. Do not invent observed national outcomes or understate strength of the pilot. No claim result applies only to sampled individuals under all conditions; identify the eligible pilot context and transport assumptions. Transition: place the three estimates beside their comparisons.
:::

## Each of the three numbers compares different groups. {#s3-p11 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(s3_comparisons())
```

::: {.ask}
Which comparison is best designed to estimate an effect? Which businesses does it describe?
:::

The pilot result does **not automatically apply to the city**, because the city's businesses and delivery conditions are different.

::: {.notes}
6 minutes. Use compact reference on worksheet page 2. No new arithmetic. Ask for comparison and population, not just ranking numbers. Stronger pilot design for eligible pilot businesses; do not use its estimate to prove city's before/after bias or city's true effect. All three costs measured after 12 months; city before/after uses its own baseline. Transition: decide what action the pilot evidence supports.
:::

## Would you extend GreenWaste now? {#s3-p12 .reader-slide .s1-review .s3-review}

<div data-choices><button data-choice aria-pressed="false">Extend</button><button data-choice aria-pressed="false">Do not extend</button><button data-choice aria-pressed="false">Ask first</button><p class="feedback" role="status"></p></div>

::: {.ask}
**Q5 on page 4:** Write one strength of the pilot design, one limit that matters for the decision, and one condition that fits the size of the decision.
:::

Choose alone first. Then compare your reasons with your group. The right action depends on the decision, the costs and how much risk the authority will accept.

::: {.notes}
8 minutes. Two minutes individual Q5, four minutes discuss, two minutes hear reasons. Native buttons record choice locally, not shared room voting. Paper labelled cards are equivalent. Do not award ask first by default or force consensus. Good reasons use randomised comparison, threshold uncertainty and reach; condition must address a consequential concern. Transition: turn this into advice the director can read.
:::

## Write your recommendation in three sentences. {#s3-p13 .reader-slide .s1-review .s3-review}

Use the writing space on **worksheet page 4**.

1. **Finding:** the result, the comparison, and the table that shows it.
2. **Action:** your recommendation, or the condition you set.
3. **Limit and request:** the most important limit, and the evidence you would request.

::: {.ask}
Swap notes with a partner. Can your partner find support in H3-A for each claim?
:::

Keep H3-A open. Different recommendations can be defensible.

::: {.notes}
13 minutes. Seven minutes write alone, four minutes peer source-check, two minutes feedback. Partner checks units/period, lottery comparison, interval against rule, reach and specific request; circle a claim to strengthen, not grammar-only correction. Restore explanation time, not more content. Full costs absent, so do not assert value for money. No predetermined scale-up answer. Transition: compare your own advice with a plausible AI paragraph.
:::

## An AI writes a paragraph for the minister. Which claims does the pilot support? {#s3-p14 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<p class="caption">Naive prompt (a prompt with no context, as a busy person might type it): “Our GreenWaste pilot saved 1,014 AED per business (p < 0.001). Write a short paragraph for the minister recommending expansion.” This response was written for the exercise. It is not a real AI answer.</p><div class="ai-reading">',s3_paragraphs(s3_ai),'</div>'))
```

::: {.ask}
On **page 5**, underline the claims that H3-A supports. Cross out the claims it does not support.
:::

::: {.notes}
8 minutes. Five minutes mark individually/pairs, three minutes debrief; do not run naive prompt. Written for the exercise, adapted from the older ministerial-paragraph device, the current pilot result and a clearer source. Supported: lottery N, after means, estimate/interval/p, stronger design, one-district qualification. Unsupported: significance proves minimum met, random allocation establishes city reach, national value for money without cost evidence. Monitoring sensible but cannot supply missing evidence retrospectively. Ask for supported passage and error. Transition: run a prompt constrained by the study summary.
:::

## Ask the AI for a recommendation the pilot supports. {#s3-p15 .reader-slide .s1-review .s3-review}

```{r, results='asis'}
html_out(paste0('<p class="caption"><strong>Improved prompt</strong></p><div class="better-prompt">',s3_paragraphs(s3_prompt),'</div>'))
```

Copy **the prompt on page 5 and H3-A** into **Microsoft Copilot**. Keep your AI good-practice card beside you.

::: {.ask}
What improved? Which advice still needs checking against H3-A?
:::

::: {.notes}
7 minutes. Four minutes run, three compare: AI Snapshot totals 15. Ask questions first; no account setup in session. If unavailable/delayed use page 6 prepared response. Compare with own note, marked draft and source, especially interval/rule and reach/value claims. Agreement between answers is not proof. Shared card supports further review, no second mandatory exercise. Transition: your final request must have a clear decision purpose.
:::

## Which evidence could change your decision? {#s3-p16 .reader-slide .s1-review .s3-review}

Keep **H3-A open**. Close the AI responses and peer comments.

::: {.ask}
Write your final recommendation and one specific evidence request. How could the answer to your request change your advice?
:::

Use the **final box on worksheet page 4**. Tomorrow: what can the city comparison of changes tell us?

::: {.notes}
10 minutes. Four minutes independent revision/request, three minutes partner checks purpose, three minutes feedback. Require evidence/document/result and consequence for advice; accept design verification, interval precision, expansion profile/delivery or costs when relevant. Keep source open but learner writes without AI/peer copy. Daily wrap-up/buffer is separate from this 120-minute session. No internal extra break. The unchanged production deck and handouts remain available; this is a separate review prototype.
:::
