---
title: "Learning R"
subtitle: "Session 2: Data Manipulation in R"
format: 
  revealjs:
    theme: [clean.scss]
    transition: slide
    highlight-style: github
    code-copy: true
    auto-stretch: false
    fig-dpi: 300
---

# Part A: Introduction to Data Manipulation {.section}

## Welcome to Data Manipulation in R!

Data manipulation is a crucial part of data analysis and preparation.

-   Cleaning and transforming data

-   Filtering and selecting relevant information

-   Summarizing and grouping data

-   Handling missing values

In this session we will learn how to examine and clean our data in R.

## Introduction to Tidyverse

The Tidyverse is a collection of R packages designed for data science that can help you import, clean, and understand data.

First, we need some packages to be loaded in. You can install the core packages from the `tidyverse` by simply typing `install.packages("tidyverse")` in the console.

```{r}
#| echo: true
#| eval: false


# Install the packages if needed
# You only need to install a package once per device
install.packages("tidyverse", "dplyr")

# Load the packages
library(tidyverse)
library(dplyr)

```

For more information on Tidyverse, explore the documentation here: <https://www.tidyverse.org/>

Tidyverse tools use pipes and tibbles to make R code more human readable. The pipe, %\>%, allows the output of one command to be used as an input in the next command, which is cleaner than using nested functions. This is an easy and common way to write and decipher R code.

\*Note: a shortcut to write the pipe command is ctrl+SHIFT+M (Windows) or command+SHIFT+M (Mac).

```{r}
## A single command 
sqrt(65)  

## Base R method of running more than one command 
round(sqrt(65), digit = 2)  

## Running more than one command with piping 
sqrt(65) %>% round(digit = 2)
```

## Filtering and Selecting Data with dplyr

One of the most useful tools in the Tidyverse is dplyr. Dplyr has many handy functions that can help with data wrangling, including:

-   `select()` extracts columns of interest.

-   `arrange()` changes the ordering of the rows.

-   `filter()` filters cases based on their values.

-   `mutate()` creates/adds new variables based on conditions of existing variables.

-   `rename()` easily changes the name of a column(s)

-   `summarise()` reduces multiple values down to a single summary.

-   `pull()` extracts a single column as a vector.

-   `_join()` group of functions that merge two data frames together, includes (`inner_join()`, `left_join()`, `right_join()`, and `full_join()`).

### Filtering Rows

Use `filter()` to subset data based on conditions.

```{r}
#| echo: true  

# Load built-in dataset 
data(mtcars)  

# Filter for rows where mpg is greater than 20 
mtcars %>% filter(mpg > 20)
```

### Selecting Columns

Use `select()` to choose specific columns.

```{r}
#| echo: true  

# Select only mpg and hp columns 
mtcars %>% select(mpg, hp)
```

## Sorting and Arranging Data

Use `arrange()` to sort data.

```{r}
#| echo: true  

# Arrange data in ascending order of mpg 
mtcars %>% arrange(mpg)  

# Arrange data in descending order of mpg 
mtcars %>% arrange(desc(mpg))
```

## Creating New Variables with mutate()

Use `mutate()` to add new variables.

```{r}
#| echo: true  

# Create a new variable: power-to-weight ratio 
mtcars %>% mutate(power_to_weight = hp / wt)
```

## 🔨 Practice Time!

Try these exercises in your console:

1.  View the mtcars data set to see the columns available.
2.  Filter the data set for cars with less than 6 cylinders.
3.  

## 🔨 Results

```{r}
#| eval: false
#| echo: true

# view mtcars dataset
head(mtcars)
# or view just the column names
colnames(mtcars)

# Filter rows where cyl is less than 6 
mtcars %>% filter(cyl < 6)
```

# Part B: Grouping and Summarizing Data {.section}

## Grouping Data with group_by()

Use `group_by()` to categorize data before summarization.

```{r}
#| echo: true  

# Group by number of cylinders 
mtcars %>% group_by(cyl)
```

## Summarizing Data with summarize()

Use `summarize()` to aggregate data.

```{r}
#| echo: true  

# Compute mean mpg by cylinder 
mtcars %>% group_by(cyl) %>% summarize(mean_mpg = mean(mpg))
```

## Joining Datasets

Use `left_join()` to combine datasets.

```{r}
#| echo: true  

# Create sample data frames 
df1 <- data.frame(ID = c(1, 2, 3), Score = c(90, 85, 88)) 
df2 <- data.frame(ID = c(1, 2, 3), Name = c("Alice", "Bob", "Charlie"))  

# Join data frames 
merged_df <- left_join(df1, df2, by = "ID")
```

## Handling Missing Values

Use `na.omit()` to remove missing values.

```{r}
#| echo: true  

# Remove rows with missing values 
clean_data <- na.omit(mtcars)
```

## Replacing Missing Values

Use `replace_na()` to fill in missing data.

```{r}
#| echo: true  

# Replace missing values in a column with a default value 
mtcars %>% mutate(mpg = replace_na(mpg, 0))
```

## 🔨 Practice Time!

Try these exercises in your console:

1.  

## 🔨 Results

# Thank You! {.section}

## Additional Resources

-   [Tidyverse Documentation](https://www.tidyverse.org/)

-   [dplyr Cheatsheet](https://rstudio.com/resources/cheatsheets/)

-   [Data Wrangling in R](https://r4ds.had.co.nz/)

## Exercise Solutions

Solutions will be posted after the session!

::: callout-tip Keep practicing for better data manipulation skills! :::
