---
title: "The GreenWaste case"
subtitle: "Preview of the case introduction"
author: "3ie"
format:
  revealjs:
    theme: [theme_dge.scss, theme_live.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    center: false
    width: 1280
    height: 720
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
---

```{r setup}
FENCE <- strrep("`", 3)
html_out <- function(x) cat("\n", FENCE, "{=html}\n", x, "\n", FENCE, "\n", sep = "")
BLUE <- "#215a9e"; GREY <- "#c9d6e6"

# One illustrative city: shop position, efficiency score, and whether it joined.
# Inside the pilot district every score is below 58 and a lottery decided;
# everywhere else, a score of 58 or below meant the shop joined.
pilot_shops <- data.frame(x = c(70, 165, 260, 70, 165, 260), y = c(98, 98, 98, 186, 186, 186),
                          score = c(41, 52, 37, 48, 33, 55), joined = c(1, 0, 1, 0, 0, 1))
city_shops  <- data.frame(
  x = c(415, 525, 635, 745, 415, 525, 635, 745, 70, 180, 290, 415, 525, 635, 745),
  y = c(80, 80, 80, 80, 195, 195, 195, 195, 330, 330, 330, 330, 330, 330, 330),
  score = c(72, 45, 88, 39, 56, 63, 29, 91, 47, 77, 52, 68, 35, 84, 58))
city_shops$joined <- as.integer(city_shops$score <= 58)

shop_svg <- function(d) paste(sprintf(
  '<use href="#shop" x="%d" y="%d" width="62" height="58" color="%s"/><text x="%d" y="%d" font-size="19" text-anchor="middle" fill="%s" font-weight="%s">%d</text>',
  d$x, d$y, ifelse(d$joined == 1, BLUE, GREY), d$x + 31, d$y + 80,
  ifelse(d$joined == 1, "#063360", "#545860"), ifelse(d$joined == 1, "700", "400"), d$score), collapse = "\n")
```

## The GreenWaste programme

<p style="font-size:0.85em; margin: 0 0 4px; max-width: none;">GreenWaste helps businesses cut their waste costs. Here is how it reached them.</p>

```{r city-map, results='asis'}
html_out(paste0('
<svg viewBox="0 0 1200 470" width="100%" role="img"
     aria-label="A city map. In the pilot district a lottery chose which shops joined. In the rest of the city, shops with an efficiency score of 58 or below joined."
     style="font-family: \'Noto Kufi Arabic\', \'Segoe UI\', Arial, sans-serif;">
  <defs>
    <symbol id="shop" viewBox="0 0 60 56">
      <path d="M4 14 L10 2 H50 L56 14 Z" fill="currentColor"/>
      <path d="M4 14 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 10 0 q5 7 12 0" fill="currentColor"/>
      <rect x="8" y="20" width="44" height="34" rx="2" fill="currentColor"/>
      <rect x="14" y="30" width="12" height="24" fill="#fff" opacity="0.85"/>
      <rect x="32" y="28" width="14" height="12" fill="#fff" opacity="0.85"/>
    </symbol>
  </defs>

  <!-- the city -->
  <rect x="20" y="20" width="820" height="440" rx="18" fill="#f4f7fb" stroke="#D5DEE8" stroke-width="2"/>
  <path d="M370 20 V460 M20 300 H840" stroke="#e3e9f1" stroke-width="10"/>
  <text x="820" y="450" font-size="22" font-weight="700" fill="#7f97b5" text-anchor="end">THE CITY</text>

  <!-- the pilot district, inside the city -->
  <rect x="45" y="45" width="305" height="245" rx="14" fill="#e3ebf5" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
  <text x="62" y="80" font-size="22" font-weight="700" fill="#063360">Pilot district</text>
  <g transform="translate(300,55)">
    <rect width="38" height="38" rx="7" fill="#fff" stroke="#063360" stroke-width="2.5"/>
    <circle cx="11" cy="11" r="3.5" fill="#063360"/><circle cx="27" cy="11" r="3.5" fill="#063360"/>
    <circle cx="19" cy="19" r="3.5" fill="#063360"/>
    <circle cx="11" cy="27" r="3.5" fill="#063360"/><circle cx="27" cy="27" r="3.5" fill="#063360"/>
  </g>
', shop_svg(pilot_shops), '
', shop_svg(city_shops), '

  <!-- how each part decided -->
  <g font-size="21" fill="#111418">
    <rect x="875" y="40" width="300" height="150" rx="12" fill="#fff" stroke="#215a9e" stroke-width="3" stroke-dasharray="12 8"/>
    <text x="895" y="82" font-weight="700" fill="#063360">In the pilot district</text>
    <text x="895" y="118">a <tspan font-weight="700">lottery</tspan> chose which</text>
    <text x="895" y="148">shops joined</text>
    <rect x="875" y="215" width="300" height="178" rx="12" fill="#fff" stroke="#D5DEE8" stroke-width="2"/>
    <text x="895" y="257" font-weight="700" fill="#063360">Everywhere else</text>
    <text x="895" y="293">shops scoring <tspan font-weight="700">58 or below</tspan></text>
    <text x="895" y="323">joined</text>
    <text x="895" y="356" font-size="16" fill="#545860">the number under each shop</text>
    <text x="895" y="377" font-size="16" fill="#545860">is its efficiency score</text>
  </g>
  <g font-size="19" fill="#111418">
    <use href="#shop" x="878" y="404" width="30" height="28" color="', BLUE, '"/><text x="918" y="425">joined GreenWaste</text>
    <use href="#shop" x="878" y="436" width="30" height="28" color="', GREY, '"/><text x="918" y="457">did not join</text>
  </g>

</svg>'))
```

::: {.notes}
About 2 minutes. One city; the pilot district is part of it. Point at the pilot
district: every shop there scores below 58, yet only some joined, because a
lottery decided. Then the rest of the city: the number under each shop decided.
Ask: "Where would it be fairer to compare shops that joined with shops that did
not: inside the pilot district or in the rest of the city? Why?" Take answers;
do not resolve it yet.
:::
