---
title: "Did the programme make the difference?"
subtitle: "Day 1 | Session 3: A pilot lottery and the limits of its results"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day1_case.R')
```


## Two savings. Two weak comparisons. {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
pilot <- subset(gw, setting == 'pilot')
ages <- pilot$manager_age[1:20]
draw <- 0
group_ages <- function(groups) {
  out <- aggregate(ages, list(Group = groups), function(x) round(mean(x), 1))
  names(out)[2] <- 'Manager age'
  out
}
```

```{r two-estimates, results='asis'}
html_out(sprintf('<div class="evidence-pair"><div class="evidence"><h3>Same businesses over time</h3><p class="big-number">%s</p><p>AED lower after 12 months</p></div><div class="evidence"><h3>Two groups afterwards</h3><p class="big-number">%s</p><p>AED lower among participants</p></div></div>%s',fmt(abs(case$before_after)),fmt(abs(case$with_without)),fiction))
```

::: {.ask}
Which comparison could support an extension decision?
:::

::: {.notes}
4 minutes. Recall the 1,000 AED rule. The first is before/after city participants; the second compares city participants with non-participants after. Both are observational summaries. Do not endorse the larger number because it passes the threshold. Hidden webR setup belongs under this first heading. Open early; there is one optional trainer demo.
:::

## Which comparison is fair? {.reader-slide}

::: {.lead}
“Participants had lower costs than businesses that did not take part.”
:::

```{=html}
<div data-choices>
<button data-choice aria-pressed="false">A fair comparison</button>
<button data-choice aria-pressed="false">I need their starting costs</button>
<p role="status" class="feedback"></p>
</div>
```

::: {.ask}
Why might the two groups already have different costs?
:::

::: {.notes}
4 minutes. Commit before opening baseline. Invite a reason related to size, management or the score rule. Use the same local choice controls or room poll with a custom question and two choices. No national inference yet.
:::

## How different were they at the start? {.reader-slide}

```{r starting-gap, results='asis'}
html_out(sprintf('<button data-open="starting-costs" data-closed="Open starting costs" data-opened="Hide starting costs" aria-controls="starting-costs" aria-expanded="false">Open starting costs</button><div id="starting-costs" hidden><div class="evidence-pair"><div class="evidence"><h3>Took part</h3><p class="big-number">%s</p></div><div class="evidence"><h3>Did not take part</h3><p class="big-number">%s</p></div></div><p class="caption">Annual waste cost before GreenWaste, AED per business</p></div>%s',fmt(mean(city$cost_before[took])),fmt(mean(city$cost_before[!took])),fiction))
```

::: {.ask}
Can the entire later gap be credited to GreenWaste?
:::

::: {.notes}
5 minutes. Predict who started cheaper, then open. Participants already had lower costs. The later gap includes a starting difference. This is a warning about selection; no need to add a regression slide. Invite someone to state both the evidence and its implication.
:::

## Could something else explain the fall? {.reader-slide}

::: {.lead}
Costs can change while a programme is running.
:::

::: {.ask}
What might have changed costs without GreenWaste?
:::

Compare your idea with a partner’s. Say what evidence would help check it.

::: {.notes}
4 minutes. Possible ideas: waste charges, supply prices, business activity, another waste policy. These are hypotheses, not facts about the invented data. Before/after confounds programme effects with other changes. Day 2 will examine how a comparison of changes helps, with assumptions. Do not teach DiD now.
:::

## What does the lottery protect us from? {.reader-slide}

```{=html}
<div class="evidence-pair">
<div class="evidence"><h3>Chosen by lottery</h3><p>Joined GreenWaste</p></div>
<div class="evidence"><h3>Waiting group</h3><p>Did not join during the study</p></div>
</div>
```

::: {.ask}
Why does chance help make these groups comparable?
:::

Fictional pilot district · `r fmt(nrow(pilot))` businesses · roughly half in each group

::: {.notes}
5 minutes. The pilot picture is already on the case card. Random assignment, an RCT, prevents systematic selection into the two groups on average across possible lotteries. It need not make any one draw perfectly balanced. In this invented pilot, assignment equals participation; real reports require checking follow-up, take-up and effects on the waiting group. Keep the estimand explanation in notes.
:::

## A lottery. Perfectly balanced groups? {.reader-slide}

20 businesses, 10 places. Compare **pre-programme manager ages**.

```{webr}
#| echo: true
# Make a new lottery draw each time the trainer runs this.
draw <- draw + 1
set.seed(draw)
# Randomly give 10 businesses a place and 10 a wait-list place.
group <- sample(rep(c('Join', 'Wait'), each = 10))
# Compare their manager ages: a balance check, not an effect.
group_ages(group)
```

::: {.ask}
Will a small lottery make the groups exactly alike?
:::

::: {.notes}
7 minutes. Predict first; trainer Run, then Run once more. The input is the first 20 pilot businesses’ actual pre-programme manager ages. This redraw is a demonstration of chance balance, not a reanalysis of the trial or an effect estimate. Rename output x orally as average manager age. Each run uses a new fixed seed. Paper fallback: draw ten of twenty numbered slips from the age list in the trainer pack and compare means. Avoid adding a sample-size slider or a sampling theory lecture.
:::

## What did the pilot actually find? {.reader-slide}

```{r trial-table, results='asis'}
html_out(reading_table('trial', TRUE))
```

::: {.ask}
Which comparison does this coefficient describe?
:::

::: {.notes}
6 minutes. Read Sheet 1. The result is the after-period difference between randomly assigned pilot groups. Assignment equals participation in this fictional pilot. Show that the comparison is different from both city estimates. Ask learners to state a saving in AED per business per year, observed 12 months later. The interval is from the canonical ordinary least-squares pilot model in greenwaste_case.R, matching the printed table.
:::

## Break {.reader-slide}

Pause here. Leave the result sheet on the table.

::: {.notes}
5 minutes. Protect the break. Do not use it to add further slides or a lab.
:::

## Does the interval clear our saving rule? {.reader-slide}

```{r trial-interval, results='asis'}
html_out(interval_svg('trial', TRUE))
html_out('<button data-open="interval-trial" data-closed="Open the 95% interval" data-opened="Hide the interval" aria-controls="interval-trial" aria-expanded="false">Open the 95% interval</button>')
```

::: {.ask}
Does the trial settle whether saving reaches the rule?
:::

::: {.notes}
6 minutes. Point estimate first, interval second. The interval spans the 1,000 AED threshold: uncertainty remains about meeting it. On the saving scale, both ends are positive; uncertainty about a threshold is different from uncertainty about any saving. The interval relies on statistical assumptions and does not capture all design failures. The pilot result may not generalise to the whole city.
:::

## Would you extend it now? {.reader-slide}

```{=html}
<div data-choices>
<button data-choice aria-pressed="false">Extend</button>
<button data-choice aria-pressed="false">Do not extend</button>
<button data-choice aria-pressed="false">Ask first</button>
<p role="status" class="feedback"></p>
</div>
```

::: {.ask}
What evidence supports your decision, and what still limits it?
:::

Use the trial’s comparison, interval and decision rule.

::: {.notes}
6 minutes. A provisional judgement may differ with costs, risks and the next evidence opportunity. Require a reason; do not award “ask first” automatically. The interval spans the rule; the pilot covers one district. If using room polling, capture After and compare with Before; invite which evidence changed a mind, without identifying voters.
:::

## Your recommendation in three sentences {.reader-slide}

Use **Sheet 2** with a partner.

State the trial result and comparison. Explain the limit on your recommendation.

::: {.ask}
What would a director need to know before approving scale-up?
:::

Keep your note to three or four sentences. Point to the evidence for each claim.

::: {.notes}
10 minutes. Six to write, two to exchange notes, two for feedback. The note should contain the randomised pilot comparison, annual saving, uncertainty about the threshold and limits on reach/implementation. Do not require every possible caveat or formula. Programme costs are not supplied, so value for money is not settled.
:::

## The AI wants to expand. What is missing? {.reader-slide data-correct="The pilot estimate is supported. The threshold is still uncertain, and nationwide value for money is not established."}

Select every unsupported conclusion in this illustrative AI draft.

```{r ai-trial, results='asis'}
html_out(sprintf('<div class="claim-list"><button data-claim="supported" aria-pressed="false">The pilot trial estimated an annual saving of %s AED per business.</button><button data-claim="unsupported" aria-pressed="false">The trial proves the minimum saving is met.</button><button data-claim="unsupported" aria-pressed="false">Nationwide rollout is good value for money.</button></div><button data-check>Check against the result sheet</button><p data-feedback class="feedback" role="status"></p>%s',fmt(abs(case$rct)),fiction))
```

::: {.notes}
8 minutes. Find two unsupported conclusions, then rewrite them using the source. Ask AI, if used, to quote the evidence behind each conclusion and list unanswered questions. Participants check the quotations, units and comparisons themselves. This authored draft is not a real saved chat. Avoid presenting agreement between two AIs as proof.
:::

## Which question could change your decision? {.reader-slide}

Close the result sheet.

::: {.ask}
What is your most useful question before this pilot becomes a citywide programme?
:::

Write your question and say how its answer could change your recommendation.

::: {.notes}
5 minutes. Individual retrieval, no AI. Accept specific questions about whether the trial was implemented fairly, who was followed up, how the pilot differs from the city, cost, uncertainty about the rule, or the next feasible evidence. Require the reason, not the term. Total 75 minutes. No full lab or additional simulation.
:::
