---
title: "How similar is similar enough?"
subtitle: "Day 2 | Session 3: Matching: similar records can hide differences"
author: "3ie"
always_allow_html: true
format:
  live-revealjs:
    theme: [theme_dge.scss, theme_live.scss, theme_reader.scss]
    logo: assets/dge_logo_horizontal.svg
    slide-number: true
    width: 1280
    height: 720
    slide-level: 2
    embed-resources: true
    include-after-body: _reader_interactions.html
webr:
  render-df: kable
  resources:
    - evaluation_data_GreenWaste_simple.csv
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
source('day2_case.R')
```

## Who would you choose as the comparison? {.reader-slide}

```{webr}
#| context: setup
#| autorun: true
#| echo: false
#| include: false
gw <- read.csv('evaluation_data_GreenWaste_simple.csv')
city <- subset(gw, setting == 'city')
vars <- c('manager_age','staff','area','filtration')
X <- scale(city[,vars])
ids <- which(city$took_part == 1)
controls <- which(city$took_part == 0)
Xc <- t(X[controls,,drop=FALSE])
matched_difference <- function() {
 twins <- controls[vapply(ids,function(i) which.min(colSums((Xc-X[i,])^2)),1L)]
 mean(city$cost_after[ids]-city$cost_after[twins])
}
```
::: {.ask}
Which business would you compare with a participant?
:::
Choose on characteristics recorded **before GreenWaste**.

::: {.notes}
4 minutes. Retrieve comparison question. Matching creates a comparison from observed characteristics; it does not create random assignment. No coding by participants; hidden helper prepares short trainer demonstration.
:::

## What can you tell before seeing costs? {.reader-slide}

```{=html}
<div class="evidence-pair"><div class="evidence"><h3>Business A</h3><p>Manager age: 45<br>Size: small<br>Took part</p></div><div class="evidence"><h3>Business 2</h3><p>Manager age: 46<br>Size: small<br>Did not take part</p></div></div>
```
::: {.ask}
Would you pair these businesses? What is still unknown?
:::
These profiles are invented for the exercise.

::: {.notes}
5 minutes. Show profile fronts only, no costs. Age and size make a plausible pair on two observed features. Other important influences can differ. Do not call them identical.
:::

## Find the twins {.reader-slide}

Lay out your **seven profile cards**.

Pair A, B and C with their closest comparison profiles.

Use manager age and size. Leave costs closed.

Record your pairs on **Sheet 1**.

::: {.ask}
Which comparison business is left over?
:::


::: {.notes}
12 minutes. Preserve Fiona's A45small B38large C52small; 1=39large 2=46small 3=51small 4=29large. Cost slips separately held until next screen. Optional walking version: seven volunteers, eighth group checks. Intended A2 B1 C3; 4 unused. This toy activity differs from full-data matching which allows reuse and matches every participant.
:::

## Now open the costs. Still convinced? {.reader-slide}

Keep your pairs. The trainer now hands out the **cost slips**.

For each pair: participant cost minus comparison cost.

Find the mean of the three differences.

::: {.ask}
Does this small exercise establish what GreenWaste caused?
:::


::: {.notes}
5 minutes. Toy differences -1000,-900,-1100, average -1000. Invented exercise, not the full-city estimate. Unmeasured differences could explain some gap. Do not let groups rematch based on outcomes. Business4 unused in toy; this is not an excluded participant.
:::

## Near the rule. Enough certainty? {.reader-slide}

```{r, results='asis'}
html_out(paste0(as.character(knitr::kable(report_row('match'),format='html',row.names=FALSE,table.attr='class="reading-table"')),fiction))
```
```{r, results='asis'}
html_out(sprintf('<p>%s participants matched to %s different comparison businesses. Some comparison businesses were used more than once.</p>%s',fmt(sum(took)),fmt(match_A$used),fiction))
```
::: {.ask}
Does finding a match for everyone make every match convincing?
:::


::: {.notes}
6 minutes. Full estimator: one nearest neighbour per participant, standardised characteristics, with replacement, no caliper or exclusions. Interval clusters by original business ID to allow reused controls, treating selected matches as given; it does not account fully for the match-selection procedure. Teach reported uncertainty and bias separately. All participants matched is true; closeness/support remains a question.
:::

## Four characteristics. Still different? {.reader-slide}

**Manager age, staff, premises area and filtration**, measured before the programme.

```{webr}
#| echo: true
#| autorun: false
# Pair businesses using the four recorded characteristics.
# Show participant cost minus matched comparison cost, in AED.
round(matched_difference())
```
::: {.ask}
What should the report show about the resulting pairs?
:::


::: {.notes}
5 minutes. Predict sign before one trainer Run. Hidden helper does the same nearest-neighbour calculation as shared source; visible result reproduces the matched coefficient. Ask for similarity, poor matches and control reuse. Standardised distance gives measured variables a common scale. No coding mechanics lesson.
:::

## Take a break {.reader-slide}

Leave your sheets and cards on the table.

We return to one question: **what might matching have missed?**

::: {.notes}
5 minutes. Protect five-minute break. Keep outputs hidden until participants return.
:::

## Drop one characteristic. A new answer? {.reader-slide}

```{r, results='asis'}
html_out(sprintf('<div class="evidence-pair"><div class="evidence"><h3>Four characteristics</h3><p class="big-number">%s</p><p>AED matched difference</p></div><div class="evidence"><h3>Age left out</h3><button data-open="omit-age" data-closed="Open the result" data-opened="Hide the result" aria-controls="omit-age" aria-expanded="false">Open the result</button><div id="omit-age" hidden><p class="big-number">%s</p><p>AED matched difference</p></div></div></div>%s',fmt(match_A$est),fmt(match_B$est),fiction))
```
::: {.ask}
Why could leaving one characteristic out change the answer?
:::


::: {.notes}
6 minutes. Predict direction before reveal. Same city, same algorithm, age excluded. The result is sensitive to information available. The three-feature match can look balanced on its own variables while ages and pre-programme costs differ. No repeated live runs.
:::

## What if motivation is missing? {.reader-slide}

::: {.ask}
Could motivation affect both joining and waste costs?
:::
A matched comparison can miss influences absent from its records.

```{=html}
<button data-open="unmeasured" data-closed="Open the implication" data-opened="Hide feedback" aria-controls="unmeasured" aria-expanded="false">Open the implication</button><div id="unmeasured" class="feedback" hidden>Similar recorded characteristics do not rule out an unmeasured reason for the cost difference.</div>
```


::: {.notes}
5 minutes. Motivation is a hypothetical omitted influence, not an observed column or established cause here. Balance concerns measured features; no balance table establishes absence of all bias.
:::

## What does 'everyone matched' prove? {.reader-slide}

Read **Sheet 2** alone, then with a partner.

Circle the result and interval.

Underline what was used to match.

Flag one missing check and write your next question.

::: {.ask}
Which part of the conclusion needs more evidence?
:::


::: {.notes}
8 minutes. Extract accurately discloses all participants matched, reuse, no maximum acceptable distance, and missing unmeasured-influence evidence. Ask for distances and poor support, baseline outcomes and sensitivity. Avoid saying all observed variables are perfectly balanced; manager age remains somewhat imbalanced.
:::

## Act now, add conditions or ask first? {.reader-slide}

::: {.ask}
Would you act, act with conditions or request more evidence?
:::
Write two sentences: your reason and the evidence that could change it.

Compare with a partner. Different decisions need clear support.

::: {.notes}
6 minutes. Protect individual judgement before discussion. Matching point estimate near rule, interval straddles rule. Causal credibility and programme costs still uncertain. Compare to pilot only if needed, no second case.
:::

## The AI promises no bias. Where is the proof? {.reader-slide}

Illustrative AI draft:

"Every participant has a match, so selection bias is removed. The estimated saving guarantees the minimum saving rule is met."

::: {.ask}
What would you keep, cross out and rewrite?
:::
```{=html}
<button data-open="match-ai" data-closed="Open feedback" data-opened="Hide feedback" aria-controls="match-ai" aria-expanded="false">Open feedback</button><div id="match-ai" class="feedback" hidden>Every participant was matched. That fact does not remove all selection bias. The interval spans the decision rule, so reaching it is uncertain.</div>
```


::: {.notes}
5 minutes. All participants matched is true in this algorithm. Unsupported inference is bias removed and threshold guaranteed. Optional AI prompt asks for claims and source evidence plus missing checks; ask questions before concluding. Verify with extract, no external AI required.
:::

## Close the sheet. What would you ask? {.reader-slide}

::: {.ask}
What would you ask the evaluator about these matches?
:::
Close your sheets. Write one question and explain its purpose.

**Matching:** inspect what made the pairs similar, what was left out and who was reused.

::: {.notes}
3 minutes. Individual exit, no AI. Look for concrete requests about closeness, reuse, baseline outcomes or unmeasured influences. Total 75 minutes.
:::
