---
title: "Reading DiD Results"
subtitle: "Module 2 · Day 2, Session 1 — Demystifying the Impact Analysis"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [clean.scss, module2.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 200
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(estimatr)
library(knitr)
library(kableExtra)

df <- read.csv("./evaluation_data_GreenWaste.csv")
tn <- df %>% filter(treatment_neighborhood == 1)

# ---- The four means -------------------------------------------------------
m <- tn %>% group_by(enrolled, round) %>%
  summarise(mc = mean(waste_management_costs, na.rm = TRUE), .groups = "drop")
gv <- function(e, r) m$mc[m$enrolled == e & m$round == r]
off_b <- gv(1, 0); off_a <- gv(1, 1)
ctl_b <- gv(0, 0); ctl_a <- gv(0, 1)
off_chg <- off_a - off_b
ctl_chg <- ctl_a - ctl_b
did_hand <- off_chg - ctl_chg

# ---- The regression an evaluator would report -----------------------------
out_did <- lm_robust(waste_management_costs ~ round * enrolled,
                     data = tn, clusters = neighborhood_identifier)
b_did  <- unname(out_did$coefficients["round:enrolled"])
se_did <- unname(out_did$std.error["round:enrolled"])
ci_lo  <- b_did - 1.96 * se_did
ci_hi  <- b_did + 1.96 * se_did
b_gap  <- unname(out_did$coefficients["enrolled"])
b_time <- unname(out_did$coefficients["round"])
b_int  <- unname(out_did$coefficients["(Intercept)"])

THRESHOLD <- 1000
CUTOFF    <- 58

# ---- Validated palette (CVD-safe; see CLAUDE.md > Design System) -----------
ACCENT  <- "#00789E"   # programme / offered
ACCENT2 <- "#B45309"   # comparison / not offered
ALERT   <- "#9A2515"   # annotations and decision rules — never a data series
INK     <- "#131516"; MUTED <- "#5C6467"; RULE <- "#DCE1E3"

# ---- One chart theme for the whole module ---------------------------------
theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(
      text             = element_text(colour = INK),
      axis.title       = element_text(colour = MUTED, size = rel(0.85)),
      axis.text        = element_text(colour = MUTED),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
      legend.position  = "bottom",
      legend.title     = element_blank(),
      legend.text      = element_text(size = rel(0.95)),
      strip.text       = element_text(colour = INK, face = "plain", size = rel(0.95)),
      plot.margin      = margin(4, 10, 4, 4)
    )
}

grp_cols <- c("Offered" = ACCENT, "Not offered" = ACCENT2)
tlev <- function(x) factor(x, levels = c("Before", "After"))
```

## What this session gives you

By the end of 90 minutes you will be able to:

::: {.panel}
- Explain, without equations, what a difference-in-differences result means
- Find the one number in an evaluation table that answers **your** question — and recognise the two that are routinely mistaken for it
- Ask the single question that decides whether the result can be trusted
- Spot four specific mistakes AI makes when interpreting an evaluation table
:::

There is no code in this session. You will never be asked to run anything.

::: {.notes}
0:00–0:03 · Set expectations. They are commissioners, not analysts. Their job
is to interrogate evidence, not produce it. Say plainly: "you will not see a
line of code today, and you will not need one."
:::

# 1 · The case [10 min]{.mins}

## The GreenWaste Program

:::::: {.columns}

::::: {.column width="54%"}
Small businesses across the emirate were spending heavily on waste
management, and rising volumes were making it worse. Waste costs had
become one of the larger controllable expenses for a small firm.

The programme offered three things: subsidised recycling collection,
an on-site efficiency audit, and staff training.

::: {.panel}
**The decision on the table:** should this be scaled up nationally?
:::
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Ran</div>        <div class="v">Two years, 100 neighbourhoods</div>
  <div class="k">Offered to</div> <div class="v">Businesses scoring ≤ 58 on the national efficiency index</div>
  <div class="k">Outcome</div>    <div class="v">Annual waste-management costs, in AED</div>
  <div class="k">Data</div>       <div class="v">4,959 businesses, measured before and after</div>
  <div class="k">The bar</div>    <div class="v">Must save at least 1,000 AED per business</div>
</div>
```
:::::

::::::

::: {.notes}
0:03 · Introduce the case properly before any number appears. Participants need
to know what the programme did, who it reached, and what decision hangs on the
answer — otherwise the statistics are abstract.

Point at "The bar" and say it out loud. Everything in the next 90 minutes is
measured against that number, and the session ends by testing the result
against it. Ask: "who sets a number like this in your organisation?"
:::

## The decision rule matters more than the statistics

```{=html}
<div class="hero">
  <div class="value">1,000 AED</div>
  <div class="label">Minimum saving per business to justify national scale-up</div>
</div>
```

::: {.panel}
This number was set **before** anyone saw the results. It reflects what the
programme costs to run and what else the money could buy.

A result can be real, statistically solid, and still not clear this bar.
:::

::: {.notes}
0:06 · Labour this. The single most common failure in evidence use — by people
and by AI — is treating "statistically significant" as "worth doing". Setting
the bar in advance is what protects you from that.
:::

## Who is actually in this evaluation?

```{r fig.height=3.0, fig.width=10}
tibble(
  g = factor(c("Offered the programme", "Not offered"),
             levels = c("Not offered", "Offered the programme")),
  n = c(sum(tn$enrolled == 1 & tn$round == 0), sum(tn$enrolled == 0 & tn$round == 0))
) %>%
  ggplot(aes(n, g, fill = g)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = -0.15, size = 5.4, colour = INK) +
  scale_fill_manual(values = c("Offered the programme" = ACCENT, "Not offered" = ACCENT2)) +
  scale_x_continuous(limits = c(0, 3600), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Businesses", y = NULL) +
  theme_m2() +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

::: {.warn}
Keep an eye on the comparison group. We will come back to exactly who these
businesses are — and it turns out to matter a great deal.
:::

::: {.notes}
0:08 · Plant the seed here; do not explain it yet. The reveal lands in Section 5
and is much stronger if they have been mildly curious about it for an hour.
:::

# 2 · Compared to what? [12 min]{.mins}

## What happened to the businesses that got it

```{r fig.height=3.6, fig.width=8}
tibble(Time = tlev(c("Before", "After")), Cost = c(off_b, off_a)) %>%
  ggplot(aes(Time, Cost, group = 1)) +
  geom_line(linewidth = 1.6, colour = ACCENT) +
  geom_point(size = 4.5, colour = ACCENT) +
  geom_text(aes(label = sprintf("%s AED", format(round(Cost), big.mark = ","))),
            vjust = -1.3, size = 5, colour = INK) +
  scale_y_continuous(limits = c(0, 2750), labels = scales::comma) +
  labs(x = NULL, y = "Waste costs (AED)") +
  theme_m2()
```

Costs fell by **`r sprintf("%.0f", abs(off_chg))` AED** — a drop of
**`r sprintf("%.0f", 100*abs(off_chg)/off_b)`%**.

::: {.notes}
0:10 · Do not editorialise. Present it as an evaluator might present a headline
finding, and let it look good.
:::

## Show of hands

::: {.ask}
On this evidence — did the programme work?
:::

::: {.notes}
0:12 · Count the hands and write the number on the flipchart. Ask two people
why. Do NOT move on until the room has committed; the next slide only works
if they have.

If someone says "we can't tell without a comparison" — excellent, name them
and say you will prove them right in thirty seconds.
:::

## Now add the businesses that did not

```{r fig.height=3.7, fig.width=8.5}
viz <- tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Offered", "Not offered"), each = 2),
  Cost  = c(off_b, off_a, ctl_b, ctl_a))

ggplot(viz, aes(Time, Cost, colour = Group, group = Group)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 4.5) +
  geom_text(aes(label = format(round(Cost), big.mark = ",")), vjust = -1.3,
            size = 4.8, show.legend = FALSE) +
  scale_colour_manual(values = grp_cols) +
  scale_y_continuous(limits = c(0, 2750), labels = scales::comma) +
  labs(x = NULL, y = "Waste costs (AED)") +
  theme_m2()
```

::: {.notes}
0:14 · Silence for a beat. Let them see the comparison line rising. Then ask
the same question again and watch the hands change.
:::

## Costs were rising for everyone

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**What we thought we saw**

Costs fell by `r sprintf("%.0f", abs(off_chg))` AED.

Effect = **`r sprintf("%.0f", off_chg)`**
:::
:::::

::::: {.column width="50%"}
::: {.ask}
**What actually happened**

Costs fell `r sprintf("%.0f", abs(off_chg))` AED *while comparable costs rose
`r sprintf("%.0f", ctl_chg)` AED.*

Effect = **`r sprintf("%.0f", did_hand)`**
:::
:::::

::::::

::: {.warn}
Note the direction: the honest estimate is **larger**, not smaller. The
programme had to overcome a rising tide, so correcting for the comparison
group *increases* the credited effect.
:::

::: {.notes}
0:17 · This surprises people who assume that "adjusting" always deflates a
result. Worth saying explicitly: a comparison group is not a haircut, it is a
correction — and it can go either way.
:::

# 3 · The idea, in four numbers [15 min]{.mins}

## Difference-in-differences in one sentence

::: {.ask}
Compare **the change over time** in the group that got the programme with
**the change over time** in a group that did not.
:::

The difference between those two changes is the estimated effect.

That is the entire method. Everything else is detail.

::: {.notes}
0:22 · Say the sentence, then say it again slowly. This is the sentence they
should be able to repeat to a minister.
:::

## Your turn — fill in the blanks

```{r}
tibble(
  Group  = c("Offered", "Not offered", "Difference of the changes"),
  Before = c(format(round(off_b), big.mark = ","), format(round(ctl_b), big.mark = ","), ""),
  After  = c(format(round(off_a), big.mark = ","), format(round(ctl_a), big.mark = ","), ""),
  Change = c("?", "?", "?")
) %>%
  kable(align = "lrrr", col.names = c("", "Before", "After", "Change")) %>%
  kable_styling(full_width = FALSE, font_size = 26)
```

::: {.ask}
Two subtractions. Ninety seconds.
:::

::: {.notes}
0:24 · Hand out the four-numbers card. Ninety seconds of silence — genuinely
silent, do not fill it. Walk the room. Most people get it, and that is the
point: they own the method rather than receiving it.
:::

## The answer

```{r}
tibble(
  Group  = c("Offered", "Not offered", "Difference of the changes"),
  Before = c(format(round(off_b), big.mark = ","), format(round(ctl_b), big.mark = ","), ""),
  After  = c(format(round(off_a), big.mark = ","), format(round(ctl_a), big.mark = ","), ""),
  Change = c(sprintf("%+.0f", off_chg), sprintf("%+.0f", ctl_chg), sprintf("%.0f", did_hand))
) %>%
  kable(align = "lrrr", col.names = c("", "Before", "After", "Change")) %>%
  kable_styling(full_width = FALSE, font_size = 26) %>%
  row_spec(3, bold = TRUE, background = "#E6F1F5")
```

::: {.ask}
`r sprintf("%+.0f", off_chg)` − (`r sprintf("%+.0f", ctl_chg)`) =
**`r sprintf("%.0f", did_hand)` AED**
:::

::: {.notes}
0:27 · "You have just estimated a causal effect with two subtractions." When
an evaluator shows you a difference-in-differences table, this is what is
underneath it — no more than this.
:::

## The estimate

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.0f AED</div>
  <div class="label">Difference-in-differences estimate</div>
</div>
<div class="stat-row">
  <div><span class="n">%+.0f</span><span class="k">Change, offered</span></div>
  <div><span class="n">%+.0f</span><span class="k">Change, not offered</span></div>
  <div><span class="n">%.0f</span><span class="k">The difference</span></div>
</div>', did_hand, off_chg, ctl_chg, did_hand))
```

::: {.notes}
0:29 · Hold here a moment. Ask: "does this clear the bar?" Most will spot that
it does not. Do not resolve it yet — Section 4 does that properly.
:::

## What the picture shows

```{r fig.height=4.0, fig.width=9}
cf <- off_b + ctl_chg

ggplot(viz, aes(Time, Cost, colour = Group, group = Group)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 4) +
  geom_line(data = tibble(Time = tlev(c("Before", "After")),
                          Cost = c(off_b, cf), Group = "Offered"),
            linetype = "22", linewidth = 1.1, colour = ACCENT) +
  annotate("segment", x = 2, xend = 2, y = off_a, yend = cf,
           arrow = arrow(ends = "both", length = unit(0.1, "inches")),
           colour = ALERT, linewidth = 1) +
  annotate("label", x = 1.76, y = (off_a + cf) / 2,
           label = sprintf("%.0f AED", did_hand), colour = ALERT,
           fontface = "bold", size = 5, label.size = 0, fill = "white") +
  annotate("text", x = 1.52, y = cf + 100, size = 3.9, colour = ACCENT,
           label = "what would have happened\nwithout the programme") +
  scale_colour_manual(values = grp_cols) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Waste costs (AED)") +
  theme_m2()
```

::: {.notes}
0:31 · The dotted line is the counterfactual — the invented number at the heart
of every impact evaluation. It is never observed. Everything rests on how
believable it is, which is exactly what Section 5 interrogates.
:::

# 4 · Reading the table [20 min]{.mins}

## The output, as an evaluator reports it

```{r}
tibble(
  Term = c("(Intercept)", "round", "enrolled", "round:enrolled", "Observations"),
  Estimate = c(sprintf("%.1f", b_int), sprintf("%.1f", b_time),
               sprintf("%.1f", b_gap), sprintf("%.1f ***", b_did),
               format(nobs(out_did), big.mark = ","))
) %>%
  kable(align = "lr", col.names = c("", "Estimate")) %>%
  kable_styling(full_width = FALSE, font_size = 24)
```

::: {.ask}
Three numbers. Only one is the programme's effect. Which one?
:::

::: {.notes}
0:37 · Take a vote across the three. Many will pick `enrolled` because it is
large and has the word "enrolled" in it. That is the trap, and it is worth
letting them fall into it.
:::

## Trap 1 — the baseline gap

```{=html}
<div class="hero">
  <div class="value alert">`r sprintf("%.0f", b_gap)`</div>
  <div class="label"><code>enrolled</code> — how far apart the groups were BEFORE the programme</div>
</div>
```

::: {.warn}
Businesses offered the programme already spent `r sprintf("%.0f", abs(b_gap))`
AED less than those not offered. Nothing the programme did caused this — it
predates the programme entirely.

Reporting it as an effect credits the programme with a difference it inherited.
:::

::: {.notes}
0:40 · This is not hypothetical. Ask whether anyone has seen a summary quote a
baseline difference as an impact. Usually at least one person has.
:::

## Trap 2 — the time trend

```{=html}
<div class="hero">
  <div class="value alert">`r sprintf("%+.0f", b_time)`</div>
  <div class="label"><code>round</code> — what happened to the comparison group anyway</div>
</div>
```

::: {.panel}
Costs rose `r sprintf("%.0f", b_time)` AED for businesses that received
nothing. This is the rising tide — volumes, prices, regulation.

It is the thing difference-in-differences exists to strip out. It is not the
effect.
:::

## The number that answers your question

```{=html}
<div class="hero">
  <div class="value">`r sprintf("%.0f", b_did)` AED</div>
  <div class="label"><code>round:enrolled</code> — the extra change in the programme group</div>
</div>
```

::: {.ask}
Ask for it by name: *"which row is the difference-in-differences estimate?"*
:::

It is the same number you calculated by hand.

::: {.notes}
0:44 · Close the loop back to the four numbers. The regression is not a black
box producing a mystery figure — it is their two subtractions, with a
precision estimate attached.
:::

## How sure are we?

```{r fig.height=2.4, fig.width=9}
# NB: ASCII "-" inside ggplot labels; the Unicode minus does not render here
ggplot(tibble(x = 1), aes(x, 0)) +
  annotate("segment", x = -THRESHOLD, xend = -THRESHOLD, y = -1, yend = 1,
           linetype = "dashed", colour = ALERT, linewidth = 1.1) +
  annotate("segment", x = ci_lo, xend = ci_hi, y = 0, yend = 0,
           linewidth = 3, colour = ACCENT) +
  annotate("point", x = b_did, y = 0, size = 6, colour = ACCENT) +
  annotate("text", x = -THRESHOLD - 25, y = 0.74, colour = ALERT, size = 4.4,
           lineheight = 0.9, hjust = 1,
           label = sprintf("the bar\n-%s AED", format(THRESHOLD, big.mark = ","))) +
  annotate("text", x = b_did, y = -0.62, size = 4.4, colour = INK,
           label = sprintf("%.0f   (%.0f to %.0f)", b_did, ci_lo, ci_hi)) +
  scale_x_continuous(limits = c(-1300, -400)) +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(x = "Change in waste costs (AED)", y = NULL) +
  theme_m2() +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
```

::: {.panel}
The data are consistent with an effect anywhere in that range. For a decision,
the useful question is not *"is it narrow?"* but:

**Does my decision change depending on where in this range the truth sits?**
:::

::: {.notes}
0:47 · The reframing that matters for commissioners. Statisticians ask about
precision; decision-makers should ask about decision-relevance. Here the whole
interval sits below the bar, so the imprecision does not change the call.
:::

## Significant is not the same as sufficient

::: {.warn}
The estimate is **statistically significant** — we are confident it is not zero.

It is also **below the bar you set** — it is not big enough to act on.

These are two different questions, and the table only answers the first.
:::

::: {.notes}
0:50 · The most transferable idea in the session. Flag that it returns in the
AI exercise, because AI conflates the two almost every time.
:::

# 5 · The one question [15 min]{.mins}

## Were the two groups heading the same way?

Everything so far rests on one assumption:

::: {.ask}
Without the programme, the two groups' costs would have moved in parallel.
:::

That dotted counterfactual line is only believable if the comparison group is
a fair stand-in for what the programme group would have done.

::: {.panel}
It cannot be proved. It can only be made more or less plausible.
:::

## Which of these would you trust?

```{r fig.height=3.4, fig.width=10}
mk <- function(lab, tr, ct) {
  tibble(panel = lab, t = rep(1:4, 2),
         grp = rep(c("Offered", "Not offered"), each = 4), y = c(tr, ct))
}
bind_rows(
  mk("A. Moving in parallel", c(40, 42, 44, 46), c(50, 52, 54, 56)),
  mk("B. Already diverging",  c(40, 41, 42, 43), c(50, 53, 56, 59)),
  mk("C. Already converging", c(40, 44, 48, 52), c(50, 52, 54, 56))
) %>%
  ggplot(aes(t, y, colour = grp, group = grp)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.4) +
  facet_wrap(~panel) +
  scale_colour_manual(values = grp_cols) +
  labs(x = "Periods before the programme started", y = "Outcome") +
  theme_m2(14)
```

::: {.warn}
Only **A** supports the assumption. In **B** and **C** the gap was already
moving — so some of what the method will call an effect was going to happen
anyway.
:::

::: {.notes}
0:55 · Vote on each panel before explaining. Make them articulate why B and C
are a problem. Someone usually spots that C would *understate* the effect
while B would overstate it — draw that out if it comes.
:::

## So — were they?

::: {.warn}
We measured costs at **two points in time**. Once before, once after.

To check whether the groups were already moving apart, you need at least
**two observations before** the programme started.

We have one.
:::

::: {.ask}
The parallel trends assumption cannot be tested with this data. Not "was not
tested" — **cannot be**.
:::

This is not unusual. It is the normal situation for most evaluations you
will commission.

::: {.notes}
0:58 · Do not soften this. The honest position is far more useful to them than
the textbook one, and it sets up the AI failure in the next section directly.
:::

## And look at who the comparison group is

```{r fig.height=3.2, fig.width=9.5}
tn %>%
  filter(round == 0) %>%
  mutate(g = ifelse(enrolled == 1, "Offered", "Not offered")) %>%
  ggplot(aes(efficiency_index, fill = g)) +
  geom_histogram(bins = 46, alpha = 0.92, colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = CUTOFF, linetype = "dashed", colour = ALERT, linewidth = 1) +
  annotate("text", x = CUTOFF, y = Inf, vjust = 1.6, hjust = -0.08,
           label = "eligibility cut-off: 58", colour = ALERT, size = 4.2) +
  scale_fill_manual(values = grp_cols) +
  labs(x = "National efficiency index score", y = "Businesses") +
  theme_m2()
```

::: {.warn}
The two groups do not overlap **at all**. The comparison group is not "similar
firms that happened to miss out" — it is, by construction, the better-scoring
half of the market.
:::

::: {.notes}
1:01 · This is the payoff for the seed planted in Section 1. Let the picture do
the work — two blocks of colour that never touch.

Then ask the killer question: "would you expect these two groups to move in
parallel on their own?" The honest answer is that we have no particular reason
to think so, and no way to check.
:::

## What to ask when it cannot be tested

You will rarely get a clean pre-trend check. Ask for the next best things:

::: {.panel}
1. **How was the comparison group chosen, and how are they different?**
2. **What else changed for one group and not the other?** — other programmes, regulations, shocks
3. **Did anyone change behaviour in anticipation of the programme?**
4. **Could the programme have affected the comparison group too?** — spillovers make effects look smaller
5. **What is the result with a different comparison group?**
:::

An evaluator with good answers to these has earned some trust. One who has not
considered them has not.

::: {.notes}
1:04 · These five are on the take-away card. Say that question 1 is the one
that would have caught the problem on the previous slide — and it requires no
statistics at all.
:::

# 6 · AI Snapshot [20 min]{.mins}

## We gave AI the table and asked it to explain

**Prompt:** *"Explain these results."*

::: {.ai-output}
The analysis uses a difference-in-differences design to evaluate the GreenWaste
Program. The results show a statistically significant treatment effect of
−816.3 AED ([p < 0.001]{.flag}), indicating that the programme
[increased savings]{.flag} by 816 AED per business.

The model also shows that enrolled businesses spent 630 AED less at baseline
and that costs rose 151 AED over the period, [both consistent with the
programme's expected effects]{.flag}.

[The parallel trends assumption was tested and holds]{.flag}, confirming that
the groups followed similar trajectories before the programme began. The
estimate is therefore robust.

**Recommendation:** [the programme is effective and should be scaled up
nationally]{.flag}.
:::

::: {.ask}
Four things here are wrong. Three minutes — find them.
:::

::: {.notes}
1:06 · Hand out the printed copy. Three minutes in pairs, marking up the text.
Take answers from the room before revealing. Someone almost always finds the
scale-up recommendation first; the fabricated test is the one they miss.
:::

## What it got wrong

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · It invented a test that is impossible**

With two rounds of data there is nothing to test. The model produced a
confident sentence about a check that was never run and could not be run.
:::

::: {.warn}
**2 · It ignored your decision rule**

−816 is below the 1,000 AED bar. It saw asterisks, read "significant", and
recommended national scale-up.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · It called a pre-existing gap an effect**

The 630 AED baseline difference existed before the programme. Calling it
"consistent with expected effects" credits the programme with it.
:::

::: {.warn}
**4 · It invented precision**

`p < 0.001` appears nowhere in the table it was given. It was inferred from
the asterisks and stated as fact.
:::
:::::

::::::

::: {.notes}
1:09 · Take them in order. Error 1 is the one groups almost never find on their
own — flag that, because it is the most dangerous.
:::

## Notice what they have in common

::: {.ask}
The answer was fluent, confident and correctly formatted.
:::

::: {.warn}
Nothing in its tone distinguishes the true sentences from the invented ones.
There is no surface cue — no hedge, no hesitation, no change of register
around the four things it made up.
:::

::: {.panel}
This is what makes the failures dangerous in a briefing note: they read
exactly like the parts that are right.
:::

::: {.notes}
1:11 · Land the meta-point hard. If they take one thing from the session about
AI, it is this: fluency is not accuracy, and confidence is not a signal.
:::

## The same table, a better prompt

:::::: {.columns}

::::: {.column width="42%"}
### What people type

::: {.panel}
*"Explain these results."*
:::

- No decision rule
- No context about the data
- No invitation to flag uncertainty
- Every gap filled with a confident guess
:::::

::::: {.column width="58%"}
### What to type instead

::: {.ask}
*I am a commissioner, not an analyst. Below is a difference-in-differences
table from a waste-management programme evaluation. Our decision rule: the
programme must cut costs by at least 1,000 AED per business to justify
scale-up. The data has two rounds only.*

*Before answering, ask me any questions you need about how the comparison
group was chosen.*

*Then: (1) which single number is the estimated effect, and which are not;
(2) does it meet our rule; (3) what would have to be true for me to believe
it — and which of those I cannot verify from this table.*
:::
:::::

::::::

::: {.notes}
1:13 · Walk through what each element does. The highest-leverage single
sentence is "ask me any questions you need before answering" — it converts the
model from a guesser into an interviewer.
:::

## Your turn

::: {.panel}
In your groups, using the table on your handout:

1. Run the **short prompt**. Read what you get.
2. Run the **full prompt card**. Answer the questions it asks you.
3. Compare: what did the second answer add, remove, or hedge?
4. Paste the second answer into a **fresh chat** and ask:
   *"Review this interpretation for errors and overstatement."*
:::

::: {.ask}
Step 4 matters. A second, clean conversation catches what the first one is
already committed to.
:::

::: {.notes}
1:15–1:25 · Ten minutes of group work. Circulate and collect one good example
and one bad one for the debrief. If a group gets a genuinely new failure mode,
screenshot it — it goes into the failures library for later sessions.
:::

## Debrief

- What did AI **overstate**?
- What did it **soften** or leave out?
- Did it **invent** anything — a test, a number, a caveat?
- Did the second chat catch what the first missed?

::: {.ask}
Two habits: tell the AI your decision rule, and make it ask you questions
before it answers.
:::

# 7 · Abu Dhabi DOH [15 min]{.mins}

<!-- PLACEHOLDER — Abu Dhabi DOH health evaluation findings not yet received.
     Replace the three slides below with the real case. Keep the structure:
     (1) the case and the decision at stake
     (2) the table, read using the four steps from Section 4
     (3) the credibility question, using the five questions from Section 5   -->

## `[PLACEHOLDER]` The DOH evaluation

::: {.warn}
**Awaiting real findings.** Structure is final; only the numbers and labels change.
:::

```{=html}
<div class="facts">
  <div class="k">Programme</div>      <div class="v">[to be inserted]</div>
  <div class="k">Outcome</div>        <div class="v">[to be inserted]</div>
  <div class="k">Comparison</div>     <div class="v">[to be inserted]</div>
  <div class="k">Decision</div>       <div class="v">[to be inserted]</div>
  <div class="k">Periods of data</div><div class="v">[note whether pre-trends can be checked]</div>
</div>
```

::: {.notes}
Ask Fiona specifically whether the DOH evaluation has more than one pre-period.
If it does, this is the chance to show a real pre-trend check and complete the
contrast with GreenWaste — which would make this section the strongest in the
session rather than a repeat.
:::

## `[PLACEHOLDER]` Reading the DOH result

Apply the four steps, in order:

::: {.panel}
1. Which row is the difference-in-differences estimate?
2. What are the other coefficients telling you — which is the baseline gap?
3. What does the interval mean for **this** decision?
4. Does it clear the threshold that matters here?
:::

::: {.notes}
Group exercise on the handout, not a lecture. Eight minutes.
:::

## `[PLACEHOLDER]` Would you act on it?

::: {.panel}
- How was the comparison group chosen, and how are they different?
- What else changed for one group and not the other?
- Any anticipation effects?
- Any spillovers onto the comparison group?
- What happens with a different comparison group?
:::

::: {.ask}
Verdict round: act on it, act with caveats, or send it back?
:::

::: {.notes}
1:38 · Hands, not discussion — keeps time. Note who says "send it back" and ask
them for the one question that decided it.
:::

# 8 · Closing [5 min]{.mins}

## Three questions to ask any evaluator presenting DiD

::: {.panel}
**1 · "Which number is the difference-in-differences estimate?"**

Not the baseline gap, not the time trend. Make them point at the row.
:::

::: {.panel}
**2 · "Were the two groups already moving apart — and how do you know?"**

If they cannot test it, ask what they offer instead.
:::

::: {.panel}
**3 · "Does this clear the bar we set — and what is the range?"**

Significant is not the same as sufficient.
:::

::: {.notes}
1:40 · The take-away card. Deliberately short enough to remember in a meeting
without notes.
:::

## Key takeaways

1. DiD compares **changes**, not levels — the change in one group against the change in another
2. It is two subtractions; the regression adds precision, not mystery
3. An evaluation table hands you three plausible numbers and only one is the effect
4. It rests on **parallel trends**, which usually cannot be tested — so interrogate the comparison group instead
5. GreenWaste: **`r sprintf("%.0f", b_did)` AED**, significant, and **below** the bar → on this evidence, do not scale up
6. AI will confirm a test that never happened, and recommend scaling against a rule it was never told

## Where this sits — Day 2 scoreboard

```{r}
tibble(
  Method = c("Difference-in-differences", "Regression discontinuity", "Matching", "Randomised trial"),
  Estimate = c(sprintf("%.0f AED", b_did), "—", "—", "—"),
  Verdict = c("Below the bar → do not scale", "(this afternoon)", "(this afternoon)", "(seen on Day 1)")
) %>%
  kable(align = "lcc") %>%
  kable_styling(full_width = FALSE, font_size = 22) %>%
  row_spec(1, bold = TRUE, background = "#E6F1F5")
```

::: {.ask}
Same programme. Same data. The methods will not all agree.
:::

::: {.notes}
1:43 · Tee up the tension deliberately. The disagreement between methods is
Day 2's real lesson, and it is what makes the afternoon worth attending.
:::

## Next session

**Reading RDD Results** — what happens when eligibility is decided by a
cut-off, and why that gives you a different kind of evidence about a
different group of businesses.

::: {.panel}
You have already seen the cut-off that matters: the efficiency index at 58.
This morning it was a problem. This afternoon it becomes the tool.
:::
