---
title: "Compared to What?"
subtitle: "Module 2 · Day 1, Session 1: Know the Language First"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [clean.scss, module2.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 200
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)

# ---- Validated palette (CVD-safe; see CLAUDE.md > Design System) -----------
ACCENT  <- "#00789E"   # primary series
ACCENT2 <- "#B45309"   # comparison / second series
ALERT   <- "#9A2515"   # annotations and decision rules: never a data series
INK     <- "#131516"; MUTED <- "#5C6467"; RULE <- "#DCE1E3"

# ---- One chart theme for the whole module ---------------------------------
theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(
      text             = element_text(colour = INK),
      axis.title       = element_text(colour = MUTED, size = rel(0.85)),
      axis.text        = element_text(colour = MUTED),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
      legend.position  = "bottom",
      legend.title     = element_blank(),
      legend.text      = element_text(size = rel(0.95)),
      strip.text       = element_text(colour = INK, face = "plain", size = rel(0.95)),
      plot.margin      = margin(4, 10, 4, 4)
    )
}

tlev <- function(x) factor(x, levels = c("Before", "After"))

# ---- Illustrative numbers (fictional; not an Abu Dhabi case study) --------
# Al Falah Road Safety Campaign
af_fatal_before <- 420;  af_fatal_after <- 294        # -30%
af_vkm_before   <- 8.0;  af_vkm_after   <- 6.2         # billion km, -22.5%
af_rate_before  <- af_fatal_before / af_vkm_before     # per billion km
af_rate_after   <- af_fatal_after  / af_vkm_after
af_rate_fall    <- 100 * (af_rate_before - af_rate_after) / af_rate_before
af_headline_fall <- 100 * (af_fatal_before - af_fatal_after) / af_fatal_before

# BrightStart After-School Tutoring Program
bs_before <- 54; bs_after <- 71                        # participating schools, pass rate %
bs_gain   <- bs_after - bs_before
nat_before <- 50; nat_after <- 65                      # all schools nationally, pass rate %
nat_gain   <- nat_after - nat_before
bs_extra   <- bs_gain - nat_gain                       # the part BrightStart can actually claim

# BrightStart regression table (illustrative; not fit from real data)
reg_tbl <- tribble(
  ~Variable,                          ~Coef,  ~SE,   ~p,      ~CI_lo, ~CI_hi,
  "(Intercept)",                       40.5,  3.2,   "<0.001", 34.2,   46.8,
  "Program (BrightStart = 1)",          2.1,  0.9,   "0.021",   0.3,    3.9,
  "Baseline pass rate (prior year)",   0.58,  0.04,  "<0.001",  0.50,   0.66,
  "Class size (students)",            -0.32,  0.14,  "0.024",  -0.59,  -0.05,
  "Teacher experience (years)",        0.15,  0.08,  "0.06",   -0.01,   0.31
)
reg_n <- 240
```

## What this session gives you

By the end of this session you will be able to:

::: {.panel}
- Ask the one question every impact evaluation result has to answer: **compared to what?**
- Explain, in plain language, the terms you will hear all week: mean, comparison group, treatment effect, coefficient, p-value, confidence interval
- Recognise when "statistically significant" is being used to mean something it does not
- Use AI to explain an unfamiliar term, and know what to check when it does
:::

The goal of these sessions is to provide you with insight on how we get to impact evaluation results, to read and interpret those results with confidence, and to know what critical questions to ask.

# 1 · Compared to what? [15 min]{.mins}

## The Al Falah Road Safety Campaign

:::::: {.columns}

::::: {.column width="54%"}
A national government runs a one-year awareness campaign: billboards, TV
spots, and school talks promoting seatbelt use and safe speeds.

At the end of the year, someone has to decide: did it work, and should it
run again?

::: {.panel}
*Illustrative example, not a real Abu Dhabi case study.*
:::
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Ran</div>        <div class="v">One year, nationwide</div>
  <div class="k">Delivered</div>  <div class="v">Billboards, TV spots, school talks</div>
  <div class="k">Outcome</div>    <div class="v">Traffic fatalities</div>
  <div class="k">The question</div> <div class="v">Did the campaign save lives?</div>
</div>
```
:::::

::::::

## The headline number

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Traffic fatalities, year on year</div>
</div>
<div class="stat-row">
  <div><span class="n">%d</span><span class="k">Deaths, year before</span></div>
  <div><span class="n">%d</span><span class="k">Deaths, year after</span></div>
  <div><span class="n">-%d</span><span class="k">Fewer deaths</span></div>
</div>', af_headline_fall, af_fatal_before, af_fatal_after,
   af_fatal_before - af_fatal_after))
```

## Show of hands

::: {.ask}
Based on this evidence, did the campaign save lives?
:::

## Before we celebrate, what else happened that year?

:::::: {.columns}

::::: {.column width="50%"}
```{r fig.height=3.6, fig.width=5}
tibble(Time = tlev(c("Before", "After")),
       Deaths = c(af_fatal_before, af_fatal_after)) %>%
  ggplot(aes(Time, Deaths, group = 1)) +
  geom_line(linewidth = 1.6, colour = ACCENT) +
  geom_point(size = 4.5, colour = ACCENT) +
  geom_text(aes(label = Deaths), vjust = -1.3, size = 5, colour = INK) +
  scale_y_continuous(limits = c(0, 480)) +
  labs(x = NULL, y = "Traffic fatalities", title = "Fatalities") +
  theme_m2(13) + theme(plot.title = element_text(size = rel(0.85), colour = MUTED))
```
:::::

::::: {.column width="50%"}
```{r fig.height=3.6, fig.width=5}
tibble(Time = tlev(c("Before", "After")),
       VKM = c(af_vkm_before, af_vkm_after)) %>%
  ggplot(aes(Time, VKM, group = 1)) +
  geom_line(linewidth = 1.6, colour = ACCENT2) +
  geom_point(size = 4.5, colour = ACCENT2) +
  geom_text(aes(label = paste0(VKM, "bn km")), vjust = -1.3, size = 5, colour = INK) +
  scale_y_continuous(limits = c(0, 9)) +
  labs(x = NULL, y = "Vehicle-km driven", title = "Distance driven, nationwide") +
  theme_m2(13) + theme(plot.title = element_text(size = rel(0.85), colour = MUTED))
```
:::::

::::::

::: {.panel}
Fuel prices rose sharply that year. People simply drove less: a 22% drop in
distance driven, nationwide, for reasons that have nothing to do with the
campaign.
:::

## What actually happened to the death rate

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">-%.0f%%</div>
  <div class="label">Fatalities per billion km driven</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f</span><span class="k">Rate before</span></div>
  <div><span class="n">%.0f</span><span class="k">Rate after</span></div>
  <div><span class="n">-%.0f%%</span><span class="k">vs. the -30%% headline</span></div>
</div>', af_rate_fall, af_rate_before, af_rate_after, af_headline_fall))
```

::: {.warn}
Once you account for how much less people drove, the campaign's apparent
effect mostly disappears. The 30% headline was mostly the fuel price, not
the campaign.
:::

## Compared to what?

::: {.hero}
::: {.value style="font-size:1.6em;"}
"Compared to what?"
:::
:::

::: {.panel}
Every evaluation result is a comparison. A before-and-after number compares
the programme to *the past*, not to what would have happened anyway. If you
do not know what a number is being compared to, you do not yet know what it
means.
:::

# 2 · Test yourselves [10 min]{.mins}

## The BrightStart After-School Tutoring Program

:::::: {.columns}

::::: {.column width="46%"}
A tutoring programme runs in a set of schools for two years. Afterwards,
Grade 6 maths pass rates in those schools are compared to where they
started.

::: {.panel}
*Illustrative example, not a real Abu Dhabi case study.*
:::
:::::

::::: {.column width="54%"}
```{r fig.height=3.6, fig.width=6}
tibble(Time = tlev(c("Before", "After")),
       Rate = c(bs_before, bs_after)) %>%
  ggplot(aes(Time, Rate, group = 1)) +
  geom_col(width = 0.45, fill = ACCENT) +
  geom_text(aes(label = paste0(Rate, "%")), vjust = -0.6, size = 5.5, colour = INK) +
  scale_y_continuous(limits = c(0, 85)) +
  labs(x = NULL, y = "Grade 6 maths pass rate") +
  theme_m2()
```
:::::

::::::

## `[MENTI PLACEHOLDER]` Did it work?

::: {.ask}
Live poll: **Based on what you see, did BrightStart cause the improvement?**

Yes &nbsp;·&nbsp; No &nbsp;·&nbsp; Can't tell without more information
:::

::: {.panel}
Results display live as a bar chart. Do not reveal the national comparison
yet, let the room commit first.
:::

## `[MENTI PLACEHOLDER]` What's missing?

::: {.ask}
Open text, rendered as a live word cloud: **In one phrase, what's missing
from this evidence?**
:::

## Meanwhile, across the country

```{r fig.height=3.8, fig.width=8.5}
tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("BrightStart schools", "All schools, nationally"), each = 2),
  Rate  = c(bs_before, bs_after, nat_before, nat_after)) %>%
  ggplot(aes(Time, Rate, colour = Group, group = Group)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 4.5) +
  geom_text(aes(label = paste0(Rate, "%")), vjust = -1.3, size = 4.8, show.legend = FALSE) +
  scale_colour_manual(values = c("BrightStart schools" = ACCENT,
                                  "All schools, nationally" = ACCENT2)) +
  scale_y_continuous(limits = c(0, 85)) +
  labs(x = NULL, y = "Grade 6 maths pass rate") +
  theme_m2()
```

::: {.panel}
An unrelated national curriculum reform landed the same two years.
Pass rates rose everywhere, not just in BrightStart schools.
:::

## Revisit your answer

::: {.ask}
Now that you have seen the comparison: would you change your Menti answer?
What would you want to know before trusting *any* before-and-after claim?
:::

::: {.panel}
BrightStart schools gained `r bs_gain` points. Schools nationally gained
`r nat_gain` points with no programme at all. The gain BrightStart can
actually claim is closer to `r bs_extra` points, not `r bs_gain`.
:::

# 3 · Learn the language [15 min]{.mins}

## Mean (average)

::: {.panel}
**Plain language:** add everything up, divide by how many there are.
:::

A "mean cost of 1,449 AED" is not any single business's cost. It is where
the whole group sits on average, and it can hide a lot of spread
underneath it.

::: {.ask}
**Example:** the `r bs_after`% pass rate on BrightStart's chart is a mean:
the average across every school in the programme, not any one school's
result.
:::

## Comparison group

::: {.panel}
**Plain language:** the group we hold up next to the group that got the
programme, ideally as similar as possible in every way except the
programme itself.
:::

This is the formal name for the thing missing from both stories you just
saw. Without it, "compared to what" has no answer.

::: {.ask}
**Example:** "all schools, nationally" was BrightStart's comparison group:
schools with no tutoring, used to judge what would have happened anyway.
:::

## Treatment effect

::: {.panel}
**Plain language:** the honest difference the programme made, once it has
been compared to the right group.
:::

Not the change over time. Not the gap between groups at the end. The part
of the change that the comparison group did not also experience.

::: {.ask}
**Example:** BrightStart's raw gain was `r bs_gain` points. Its treatment
effect, the part it can actually claim, is closer to `r bs_extra` points,
once the comparison group's own `r nat_gain`-point gain is subtracted out.
:::

## Coefficient

::: {.panel}
**Plain language:** a number a statistical model reports to describe the
size of a relationship. In an evaluation table, the coefficient on the
programme variable is often the treatment effect itself.
:::

You will see coefficients in a table most days this week. The skill is
knowing which one answers your question.

::: {.ask}
**Example:** that same `r bs_extra`-point effect is what you would find as
the coefficient on "Program" in a regression table, next slide.
:::

## A regression table, annotated

An evaluator runs a regression instead of doing the subtraction by hand,
because it can hold other things constant at the same time.

```{r}
reg_tbl %>%
  mutate(CI = sprintf("[%.2f, %.2f]", CI_lo, CI_hi)) %>%
  select(Variable, Coef, SE, p, CI) %>%
  kable(align = "lrrrr", digits = 2,
        col.names = c("Variable", "Coefficient", "Std. error", "p-value", "95% CI")) %>%
  kable_styling(full_width = FALSE, font_size = 22) %>%
  row_spec(2, bold = TRUE, background = "#E6F1F5") %>%
  footnote(general = paste("N =", reg_n, "schools. Outcome: Grade 6 maths pass rate."),
           general_title = "", footnote_as_chunk = TRUE)
```

## The same table, in plain language

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**Coefficient (highlighted row):** the treatment effect. `r sprintf("%.1f", reg_tbl$Coef[2])` points, close to the `r bs_extra` we found by hand; regression refines it slightly by also holding the covariates below constant.

**Covariates:** the other rows (baseline pass rate, class size, teacher experience). Other measured differences between schools, included so they are not confused with the programme's own effect.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
**Std. error and 95% CI:** how much this estimate might move around if you ran the study again. A range that crosses zero (teacher experience) means "we cannot rule out no effect at all."

**p-value:** how surprising each coefficient would be if its true effect were zero.

**N:** the sample size, 240 schools. Keep this number in mind, it matters a great deal in Section 5.
:::
:::::

::::::

## Four terms so far

```{=html}
<div class="facts">
  <div class="k">Mean</div>              <div class="v">The average of a group</div>
  <div class="k">Comparison group</div>  <div class="v">Who we hold the programme group up against</div>
  <div class="k">Treatment effect</div>  <div class="v">The honest difference, once compared properly</div>
  <div class="k">Coefficient</div>       <div class="v">The number in the table that reports it</div>
</div>
```

# 4 · Put AI to work [10 min]{.mins}

## AI Prompt Cards

::: {.panel}
Each group takes one term card. Paste its prompt into your AI tool of
choice, read the answer together, and mark anything you think it got wrong
or overstated.
:::

::: {.ai-output}
**Prompt:** "Explain what a treatment effect is in an evaluation."

**AI:** "The treatment effect is the number that shows how much the
programme changed the outcome. [It tells you the programme caused all of
the change you observed.]{.flag} For example, if costs fell by 500 AED
after the programme, the treatment effect is 500 AED."
:::

## Debrief

::: {.ask}
What did AI get right? What did it skate over?
:::

::: {.warn}
The flagged sentence quietly drops the comparison group. AI answered as if
this were a before-and-after number, the exact trap from Section 1.
:::

# 5 · The most misunderstood term [15 min]{.mins}

## p-value

::: {.panel}
**Plain language:** roughly, how surprising this result would be if the
programme had truly done nothing at all. A small p-value means "this would
be a strange coincidence if there were no effect."
:::

A p-value tells you about confidence that an effect is not exactly zero.
It tells you nothing about whether the effect is big enough to matter.

## Confidence interval

::: {.panel}
**Plain language:** a range of plausible values for the true effect, not
one precise number. A wider range means less certainty, not a bigger
effect.
:::

When someone reports "-816 AED, 95% CI [-1,120, -512]", the honest reading
is: probably somewhere in that range, not exactly -816.

## `.warn` Two traps evaluators fall into

::: {.warn}
**"Statistically significant" does not mean "big" or "important."** It
means we are confident the effect is not exactly zero. A huge sample can
make a tiny, meaningless effect "significant." A small sample can make a
large, real effect look "not significant."
:::

::: {.warn}
**A study that finds nothing may simply have been too small to see
anything.** Absence of evidence is not evidence of absence. Always ask how
many people or businesses the study had before believing a "no effect"
result.
:::

# 6 · Quick check [5 min]{.mins}

## `[MENTI PLACEHOLDER]` True or false?

::: {.panel}
1. A statistically significant result is always a big result.
2. A small study that finds no effect proves the programme did not work.
3. A comparison group should look as similar as possible to the group that
   got the programme.
4. A wide confidence interval means we are very sure of the exact effect.
:::

::: {.ask}
Live Menti quiz, answers revealed one at a time. (1) False (2) False
(3) True (4) False
:::

# 7 · Bring it together [15 min]{.mins}

## Terms Bingo

::: {.panel}
Each of you has a bingo card with today's terms scattered across the grid.
As excerpts from real evaluation findings are read aloud, mark off any term
you hear used. First full row calls "bingo."
:::

::: {.facts}
```{=html}
<div class="k">Card</div>       <div class="v">Mean · Comparison group · Treatment effect · Coefficient · p-value · Confidence interval · Statistically significant · Sample size · Baseline</div>
<div class="k">Play</div>       <div class="v">A short evaluation excerpt is read aloud every 30-45 seconds</div>
```
:::

## Key takeaways

::: {.panel}
- Every result is a comparison. "Compared to what?" comes before any number.
- A comparison group is what turns a before-and-after story into an honest one.
- Significant is not the same as big, and no result is not the same as no effect.
- Ask AI to explain a term, then check what it quietly left out.
:::

## Next session

Session 2 puts this vocabulary to work on the first real case of the week:
descriptive results from the Abu Dhabi Police evaluation, projected live in
R.
