---
title: "Spot the Problem"
subtitle: "Module 2 · Day 1, Session 3: Naive Comparisons and Reading an RCT"
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [clean.scss, module2.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
  pptx:
    slide-level: 2
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 200
editor_options:
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)
library(estimatr)
library(broom)
library(modelsummary)

# ---- Validated palette (CVD-safe; see CLAUDE.md > Design System) -----------
ACCENT  <- "#00789E"   # the programme / treated group
ACCENT2 <- "#B45309"   # comparison / not-treated group
ALERT   <- "#9A2515"   # decision rules, annotations: never a data series
INK     <- "#131516"; MUTED <- "#5C6467"; RULE <- "#DCE1E3"

# ---- One chart theme for the whole module ---------------------------------
theme_m2 <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(
      text             = element_text(colour = INK),
      axis.title       = element_text(colour = MUTED, size = rel(0.85)),
      axis.text        = element_text(colour = MUTED),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = RULE, linewidth = 0.4),
      legend.position  = "bottom",
      legend.title     = element_blank(),
      legend.text      = element_text(size = rel(0.95)),
      strip.text       = element_text(colour = INK, face = "plain", size = rel(0.95)),
      plot.margin      = margin(4, 10, 4, 4)
    )
}

tlev <- function(x) factor(x, levels = c("Before", "After"))

# ---- Data and the numbers this session turns on ---------------------------
df <- read.csv("./evaluation_data_GreenWaste.csv")

# Before-and-after: businesses that took part (enrolled), over time
ba_before <- round(mean(df$waste_management_costs[df$enrolled == 1 & df$round == 0]))
ba_after  <- round(mean(df$waste_management_costs[df$enrolled == 1 & df$round == 1]))
ba_change <- ba_after - ba_before

# The comparison group's own drift over the same period (not-offered, treated nbhds)
comp_before <- round(mean(df$waste_management_costs[df$treatment_neighborhood == 1 &
                          df$eligible == 0 & df$round == 0]))
comp_after  <- round(mean(df$waste_management_costs[df$treatment_neighborhood == 1 &
                          df$eligible == 0 & df$round == 1]))

# With-and-without at follow-up: took part vs did not, same neighbourhoods
ww_prog <- ba_after
ww_comp <- comp_after
ww_gap  <- ww_prog - ww_comp

# The RCT: eligible businesses, treatment vs control neighbourhoods, follow-up
rct <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                 data = df %>% filter(round == 1, eligible == 1),
                 clusters = neighborhood_identifier)
rct_est    <- round(coef(rct)["treatment_neighborhood"])
rct_ci_lo  <- round(rct$conf.low["treatment_neighborhood"])
rct_ci_hi  <- round(rct$conf.high["treatment_neighborhood"])
rct_control <- round(coef(rct)["(Intercept)"])
rct_treat   <- rct_control + rct_est

RULEBAR <- 1000L   # the decision rule used all week: cut costs by >= 1,000 AED
```

## What this session gives you

By the end of this session you will be able to:

::: {.panel}
- Put Session 1's question, **compared to what?**, to work on two comparisons that both look convincing
- Name the hidden problem in a **before-and-after** claim and in a **with-and-without** claim
- Explain, in plain terms, how **randomisation** removes the difference between the groups
- Read a randomised result: which number is the impact, and does it clear the decision rule
:::

You will not run any analysis. Your job is to spot the problem before you act on the number.

::: {.notes}
0:00 · One minute. This is the last session of Day 1 and the pay-off for Session 1.
Today they learn to distrust two comparisons they will see constantly, then meet
the one design that fixes them.
:::

# 1 · Two comparisons, both convincing [18 min]{.mins}

## The programme on the table

:::::: {.columns}

::::: {.column width="54%"}
The **GreenWaste Program** helped businesses cut waste-management costs.
A pilot ran, data was collected before and after, and now someone has to
decide whether to take it national.

The bar it has to clear is fixed for the whole week.
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>    <div class="v">GreenWaste, waste-cost reduction</div>
  <div class="k">Outcome</div>      <div class="v">Waste-management costs (AED)</div>
  <div class="k">The decision</div> <div class="v">Scale up nationally, or not</div>
  <div class="k">The rule</div>     <div class="v">Must cut costs by at least 1,000 AED</div>
</div>
```
:::::

::::::

::: {.notes}
0:01 · Land the decision rule now and repeat it all afternoon. Every number today
gets judged against 1,000 AED, not against zero.
:::

## Comparison A: before and after

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=3.8, fig.width=6}
tibble(Time = tlev(c("Before", "After")),
       Cost = c(ba_before, ba_after)) %>%
  ggplot(aes(Time, Cost, group = 1)) +
  geom_line(linewidth = 1.6, colour = ACCENT) +
  geom_point(size = 4.5, colour = ACCENT) +
  geom_text(aes(label = format(Cost, big.mark = ",")),
            vjust = -1.2, size = 5, colour = INK) +
  scale_y_continuous(limits = c(0, 1700)) +
  labs(x = NULL, y = "Waste costs (AED)",
       title = "Businesses that took part") +
  theme_m2(13) + theme(plot.title = element_text(size = rel(0.9), colour = MUTED))
```
:::::

::::: {.column width="48%"}
::: {.panel}
Businesses that took part paid **`r format(ba_before, big.mark=",")` AED**
before and **`r format(ba_after, big.mark=",")` AED** after.
:::

::: {.hero}
::: {.value style="font-size:1.4em;"}
`r format(ba_change, big.mark=",")` AED
:::
::: {.label}
Fall in costs, before to after
:::
:::

The programme looks like it cut costs by `r format(-ba_change, big.mark=",")` AED.
:::::

::::::

::: {.notes}
0:03 · Say the story the way an evaluator might: "costs fell by 665 after we ran
the programme." Do not challenge it yet.
:::

## Comparison B: with and without

:::::: {.columns}

::::: {.column width="52%"}
```{r fig.height=3.8, fig.width=6}
tibble(Group = factor(c("Took part", "Did not"),
                       levels = c("Took part", "Did not")),
       Cost  = c(ww_prog, ww_comp)) %>%
  ggplot(aes(Group, Cost, fill = Group)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = format(Cost, big.mark = ",")),
            vjust = -0.6, size = 5, colour = INK) +
  scale_fill_manual(values = c("Took part" = ACCENT, "Did not" = ACCENT2)) +
  scale_y_continuous(limits = c(0, 2600)) +
  labs(x = NULL, y = "Waste costs (AED)", title = "At follow-up, same neighbourhoods") +
  theme_m2(13) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(0.9), colour = MUTED))
```
:::::

::::: {.column width="48%"}
::: {.panel}
At the end, businesses that took part paid **`r format(ww_prog, big.mark=",")` AED**;
those that did not paid **`r format(ww_comp, big.mark=",")` AED**.
:::

::: {.hero}
::: {.value style="font-size:1.4em;"}
`r format(ww_gap, big.mark=",")` AED
:::
::: {.label}
Gap between the two groups
:::
:::

This comparison makes the programme look more than twice as good.
:::::

::::::

::: {.notes}
0:05 · Same programme, a much bigger number. Note the two numbers out loud: minus
665 from A, minus 1,446 from B. Hold the tension.
:::

## `[MENTI PLACEHOLDER]` Which would you trust?

::: {.ask}
Both comparisons say GreenWaste worked. **Which would you trust more, and why?**

Before-and-after &nbsp;·&nbsp; With-and-without &nbsp;·&nbsp; Neither on its own
:::

::: {.panel}
Live poll. Let the room commit before anything is revealed. Take two or three
"why" answers out loud, do not correct them yet.
:::

::: {.notes}
0:07 · Resist explaining. The whole point is that they pick one, then discover in
Section 2 that both are biased. A split room is the ideal starting point.
:::

## `[MENTI PLACEHOLDER]` What might be wrong?

::: {.ask}
Open text, shown as a live word cloud: **in one phrase, what might be wrong with
each of these comparisons?**
:::

::: {.notes}
0:09 · Collect the words. If "other things changed" and "different groups" surface,
you have the two answers already; use their words in the next section.
:::

# 2 · The hidden problem in each [15 min]{.mins}

## Before-and-after compares to the past

:::::: {.columns}

::::: {.column width="54%"}
::: {.panel}
A before-and-after number compares the programme to **what came before**, not
to what would have happened anyway. Anything else that changed over the same
period is folded into the number.
:::

::: {.warn}
Over these same two rounds, similar businesses that did **not** take part saw
costs move from `r format(comp_before, big.mark=",")` to
`r format(comp_after, big.mark=",")` AED on their own. Time was not standing
still, so a plain before-and-after does not tell you what the programme alone did.
:::
:::::

::::: {.column width="46%"}
```{r fig.height=3.9, fig.width=5.4}
tibble(
  Time  = tlev(rep(c("Before", "After"), 2)),
  Group = rep(c("Took part", "Did not"), each = 2),
  Cost  = c(ba_before, ba_after, comp_before, comp_after)) %>%
  ggplot(aes(Time, Cost, colour = Group, group = Group)) +
  geom_line(linewidth = 1.5) + geom_point(size = 4) +
  scale_colour_manual(values = c("Took part" = ACCENT, "Did not" = ACCENT2)) +
  scale_y_continuous(limits = c(0, 2600)) +
  labs(x = NULL, y = "Waste costs (AED)") +
  theme_m2(12)
```
:::::

::::::

::: {.notes}
0:10 · The comparison group's costs drifted up. So the eligible group's own costs
would probably not have stayed flat either. A raw before-and-after therefore
understates the true reduction. Keep it intuitive, no DiD arithmetic yet.
:::

## With-and-without compares different groups

::: {.panel}
A with-and-without number compares people who took part to people who did not.
But the two groups were **already different before the programme started**. The
businesses that did not take part were not eligible: they were larger, less
efficient, more costly to begin with.
:::

::: {.warn}
So the `r format(ww_gap, big.mark=",")` AED gap mixes two things together: any
real effect of the programme, **and** the head start the eligible group already
had. It flatters the programme.
:::

::: {.notes}
0:13 · This is selection. The "without" group is not a stand-in for what the
"with" group would have looked like, because they were never comparable. Overstates.
:::

## Two biases, opposite directions

::: {.panel}
The truth is not halfway by coincidence. Each naive comparison is wrong for a
different reason, and they miss in **opposite directions**.
:::

```{=html}
<div class="facts">
  <div class="k">Before-and-after</div> <div class="v">Folds in other changes over time. Here it understates.</div>
  <div class="k">With-and-without</div> <div class="v">Compares groups that already differed. Here it overstates.</div>
  <div class="k">So</div>               <div class="v">The honest answer sits between them, and neither comparison can find it.</div>
</div>
```

::: {.notes}
0:15 · Foreshadow: minus 665 and minus 1,446 bracket something. In a few minutes
a fair comparison lands between them, near minus 1,000.
:::

## `.ask` On your desk

::: {.ask}
Think of an evaluation you have received. Was its headline a before-and-after
number, or a with-and-without one? Which of these two problems could it have had?
:::

::: {.panel}
Ninety seconds with your neighbour. Two or three examples to the room.
:::

::: {.notes}
0:16 · Short and personal. Makes the trap concrete before the break.
:::

# Comfort break [5 min]{.mins}

## Five-minute break

::: {.panel}
Back in five minutes. When we return: the one design that removes the difference
between the groups.
:::

::: {.notes}
0:18 · Hard stop at 5 minutes. Resume promptly.
:::

# 3 · How randomisation fixes it [15 min]{.mins}

## Draw lots

::: {.panel}
Both problems come from the same root: the group that took part and the group
that did not were **not comparable**. Randomisation fixes it at the source.
Decide who is offered the programme by **drawing lots**, not by who is eligible,
keen, or easy to reach.
:::

::: {.hero}
::: {.value style="font-size:1.3em;"}
Same on average, even on what we cannot measure
:::
:::

::: {.notes}
0:19 · The one sentence to remember. Random assignment balances the things we can
see and the things we cannot: motivation, management quality, luck.
:::

## Why it works: two people versus two crowds

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
Pick **two people** at random from this room. They could differ in almost every
way. Neither is a fair stand-in for the other.
:::
:::::

::::: {.column width="50%"}
::: {.panel}
Pick **two large groups** at random. On average they look alike: same mix of
ages, sizes, experience, and the same mix of things you never measured. Any
leftover difference is small and down to chance.
:::
:::::

::::::

::: {.ask}
So when we compare two large randomly-formed groups, any real gap in the outcome
is the programme, not a head start.
:::

::: {.notes}
0:21 · This is the whole intuition for an RCT. Numbers are not needed here; the
crowd-versus-pair contrast does the work.
:::

## `[MENTI PLACEHOLDER]` True or false?

::: {.panel}
1. Randomisation makes the two groups similar only on things we can measure.
2. With enough units, random assignment balances unmeasured things too.
3. A randomised comparison still needs the groups to have started out alike.
:::

::: {.ask}
Live quiz, revealed one at a time. (1) False (2) True (3) It creates that, which
is the point.
:::

::: {.notes}
0:24 · Item 1 is the common misconception; spend the most time there.
:::

# 4 · Reading the randomised result [18 min]{.mins}

## First check: did the random draw work?

::: {.panel}
Check the two groups really did start out alike. A **balance table** compares
them on their starting characteristics: read for whether the differences are
small, with p-values that are **not** tiny.
:::

```{r echo=TRUE, eval=FALSE}
# Compare treated vs control neighbourhoods on each starting characteristic
df %>%
  filter(round == 0, eligible == 1) %>%
  select(treatment_neighborhood, age_manager, educ_manager,
         staff_size, foreign_owned, recycling_center_distance) %>%
  pivot_longer(-treatment_neighborhood) %>%
  group_by(name) %>%
  reframe(tidy(lm_robust(value ~ treatment_neighborhood))) %>%
  filter(term == "treatment_neighborhood")
```

```{r}
vars <- c("age_manager", "educ_manager", "staff_size",
          "foreign_owned", "recycling_center_distance")
var_labs <- c(age_manager = "Manager age", educ_manager = "Manager education",
              staff_size = "Staff size", foreign_owned = "Foreign owned",
              recycling_center_distance = "Distance to recycling centre")

balance <- df %>%
  filter(round == 0, eligible == 1) %>%
  select(treatment_neighborhood, all_of(vars)) %>%
  pivot_longer(-treatment_neighborhood, names_to = "name") %>%
  group_by(name) %>%
  reframe({
    m <- lm_robust(value ~ treatment_neighborhood)
    tibble(
      `Treated` = round(mean(value[treatment_neighborhood == 1]), 1),
      `Control` = round(mean(value[treatment_neighborhood == 0]), 1),
      `Difference` = round(coef(m)["treatment_neighborhood"], 2),
      `p-value` = round(tidy(m)$p.value[tidy(m)$term == "treatment_neighborhood"], 2))
  }) %>%
  mutate(Characteristic = var_labs[name]) %>%
  select(Characteristic, Treated, Control, Difference, `p-value`)

balance %>%
  kable(align = "lrrrr") %>%
  kable_styling(full_width = FALSE, font_size = 18)
```

::: {.notes}
0:25 · Walk one row aloud, e.g. manager age. Treated and control almost identical,
p-value not small. Say what a small p-value here would have meant: a failed draw.
The code above is the same short, checkable calculation that produced the table.
:::

## The estimate

:::::: {.columns}

::::: {.column width="46%"}
```{r echo=TRUE, eval=FALSE}
# Compare costs in treated vs control
# neighbourhoods at follow-up, among
# eligible businesses
lm_robust(
  waste_management_costs ~
    treatment_neighborhood,
  data = filter(df, round == 1,
                eligible == 1),
  clusters = neighborhood_identifier)
```
:::::

::::: {.column width="54%"}
```{r}
modelsummary(
  list("Follow-up" = rct),
  stars = TRUE,
  coef_map = c("treatment_neighborhood" = "Treated vs. control",
               "(Intercept)" = "Control group average"),
  gof_map = c("nobs", "r.squared"),
  output = "kableExtra") %>%
  kable_styling(font_size = 19)
```
:::::

::::::

::: {.notes}
0:31 · One estimate, no covariate model, no robustness zoo. Point to the "Treated
vs. control" row: that number is the impact. Next slide names the mis-reads.
:::

## Which number is the impact?

:::::: {.columns}

::::: {.column width="50%"}
::: {.panel}
**The impact** is the "Treated vs. control" row: about
**`r format(rct_est, big.mark=",")` AED**. That is the whole result. Everything
else is supporting detail.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**Not** the "control group average" (`r format(rct_control, big.mark=",")`): that
is just where the comparison group sat, not an effect.

**Not** the stars on their own: they say the effect is unlikely to be zero, not
that it is big enough to matter.
:::
:::::

::::::

::: {.notes}
0:33 · The two classic mis-reads: quoting the intercept as if it were the effect,
and treating significance as sufficiency. Both come back on Day 2.
:::

## Does it clear the rule?

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%s AED</div>
  <div class="label">Estimated cut in costs (95%% CI: %s to %s)</div>
</div>
<div class="stat-row">
  <div><span class="n">%s</span><span class="k">The rule: cut by at least</span></div>
  <div><span class="n">%s</span><span class="k">Estimated cut</span></div>
  <div><span class="n">Yes</span><span class="k">Clears the bar</span></div>
</div>',
  format(rct_est, big.mark=","),
  format(rct_ci_lo, big.mark=","), format(rct_ci_hi, big.mark=","),
  format(RULEBAR, big.mark=","), format(-rct_est, big.mark=",")))
```

::: {.panel}
The randomised estimate is about `r format(-rct_est, big.mark=",")` AED, and the
confidence interval stays beyond the 1,000 AED bar. On this evidence, GreenWaste
clears the rule.
:::

::: {.notes}
0:35 · Say it plainly: this design says scale up. Do not mention that other
methods will disagree; that is Day 2's reveal. Just note the interval sits past
the bar, so it is not a borderline call.
:::

## The picture

```{r fig.height=4.0, fig.width=8.5}
tibble(Group = factor(c("Control", "Treated"), levels = c("Control", "Treated")),
       Cost  = c(rct_control, rct_treat)) %>%
  ggplot(aes(Group, Cost, colour = Group)) +
  geom_point(size = 5) +
  geom_text(aes(label = format(Cost, big.mark = ",")),
            vjust = -1.1, size = 5, colour = INK, show.legend = FALSE) +
  annotate("segment", x = 2, xend = 2, y = rct_treat, yend = rct_control,
           colour = ALERT, linewidth = 0.7,
           arrow = arrow(ends = "both", length = unit(0.09, "inches"))) +
  annotate("text", x = 2.12, y = mean(c(rct_control, rct_treat)),
           label = paste0(format(rct_est, big.mark = ","), " AED"),
           colour = ALERT, fontface = "bold", hjust = 0) +
  scale_colour_manual(values = c("Control" = ACCENT2, "Treated" = ACCENT)) +
  scale_y_continuous(limits = c(0, 2100),
                     expand = expansion(mult = c(0.02, 0.10))) +
  labs(x = NULL, y = "Waste costs at follow-up (AED)") +
  theme_m2() + theme(legend.position = "none")
```

::: {.notes}
0:37 · The gap between two comparable groups is the effect. Contrast in one line
with the two naive charts from Section 1: same programme, but now the comparison
is fair.
:::

# 5 · AI Snapshot [18 min]{.mins}

## Asking AI to explain the pitfall

Suppose we ask AI: *"Explain in plain language why a before-and-after comparison can be misleading."* Here is the kind of answer it might give. Some of what it says may be wrong.

::: {.ai-output}
A before-and-after comparison looks at the outcome before a programme and after
it, and treats the difference as the programme's effect. It can mislead because
[if the numbers changed, the programme is what changed them]{.flag}.

The fix is simple: [as long as you measure the same group before and after, a
before-and-after comparison gives you the true effect]{.flag}. To be safe, you
can also [run a statistical test, and if the change is significant then the
programme worked]{.flag}.

In this case costs fell, so [we can conclude the programme caused the full fall
of 665 AED]{.flag}.
:::

::: {.ask}
In your groups, mark anything that looks wrong. Three minutes.
:::

::: {.notes}
0:38 · Hand out the printed copy. This is a placeholder example, invented for now.
Standing task: when Fiona runs this prompt herself and the model slips, screenshot
it into AI_failure_exs/ and swap the real capture in. Real captures beat invented ones.
:::

## What might be wrong here

:::::: {.columns}

::::: {.column width="50%"}
::: {.warn}
**1 · It assumed away the whole problem.** "If the numbers changed, the programme
changed them" ignores everything else that moved over the same period.
:::

::: {.warn}
**2 · It called before-and-after a true-effect method.** It is exactly the
comparison we just showed is biased. There is no comparison group in it.
:::
:::::

::::: {.column width="50%"}
::: {.warn}
**3 · It confused significant with correct.** A significant before-and-after
change is still a biased number. Significance does not repair the design.
:::

::: {.warn}
**4 · It quoted the 665 as the causal effect.** That is the naive figure. The
fair estimate is about `r format(-rct_est, big.mark=",")` AED, and the model had
no way to know that.
:::
:::::

::::::

::: {.notes}
0:43 · Take them in order. Error 2 is the dangerous one: the model recommended the
exact design the session exists to warn against, in fluent, confident prose.
:::

## Fluent, confident, and wrong

::: {.ask}
Notice: nothing in the tone flags the invented parts. No hedge, no hesitation.
:::

::: {.warn}
The wrong sentences read exactly like the right ones. In a briefing note you would
not be able to tell them apart. Fluency is not accuracy, and confidence is not a
signal.
:::

::: {.notes}
0:45 · The meta-point. If they remember one thing about AI this week, this is it.
:::

## The same question, a better prompt

:::::: {.columns}

::::: {.column width="42%"}
### What people type

::: {.panel}
*"Explain why a before-and-after comparison can be misleading."*
:::

- No context about the case
- No decision rule
- No invitation to say what it is unsure of
:::::

::::: {.column width="58%"}
### What to type instead

::: {.ask}
*I am a commissioner, not an analyst. A programme's costs fell from 1,449 to 784
AED after it ran, before-and-after, with no comparison group. Our decision rule:
it must cut costs by at least 1,000 AED to justify scale-up.*

*Before you answer, ask me any questions you need about what else changed over the
same period and who was in each group.*

*Then explain, in plain language: what this before-and-after number can and cannot
tell me, and what comparison I would need to trust it.*
:::
:::::

::::::

::: {.notes}
0:47 · The highest-leverage line is "ask me any questions before you answer." It
turns the model from a guesser into an interviewer. Walk through each element.
:::

## Your turn

::: {.panel}
In your groups:

1. Run the **short prompt**. Read what you get.
2. Run the **full prompt card**. Answer the questions it asks you.
3. Compare: what did the second answer add, remove, or hedge?
4. Paste the second answer into a **fresh chat** and ask: *"Review this for errors
   and overstatement."*
:::

::: {.ask}
Step 4 matters: a clean second conversation catches what the first is already
committed to.
:::

::: {.notes}
0:49 to 0:55 · Six minutes of group work. Circulate, collect one good and one bad
example for the debrief. Screenshot any genuinely new failure for the library.
:::

## Debrief

- What did AI **overstate**?
- What did it **soften** or leave out?
- Did it **invent** anything: a test, a number, a caveat?

::: {.ask}
Two habits, every time: tell the AI your decision rule, and make it ask you
questions before it answers.
:::

::: {.notes}
0:56 · Keep it to two minutes. Land the two habits; they recur every session.
:::

# 6 · The real case: Abu Dhabi Police [15 min]{.mins}

## `[PLACEHOLDER]` The Abu Dhabi Police evaluation

<!-- PLACEHOLDER: awaiting Abu Dhabi Police RCT data and findings. Structure is
fixed; real figures drop into the .facts panel and the reading exercise below. -->

:::::: {.columns}

::::: {.column width="54%"}
The same reading skills, now on a real Abu Dhabi case. The Police ran a
randomised evaluation of `[programme name]`. You saw its descriptive results in
Session 2; here you read its randomised estimate.

::: {.warn}
`[PLACEHOLDER]` Figures below are dummy values, marked, until the real findings
are loaded. Nothing here is a real result yet.
:::
:::::

::::: {.column width="46%"}
```{=html}
<div class="facts">
  <div class="k">Programme</div>    <div class="v">[PLACEHOLDER]</div>
  <div class="k">Design</div>       <div class="v">Randomised (drew lots)</div>
  <div class="k">Outcome</div>      <div class="v">[PLACEHOLDER]</div>
  <div class="k">Decision rule</div><div class="v">[PLACEHOLDER]</div>
</div>
```
:::::

::::::

::: {.notes}
0:58 · Bridge from Session 2: same programme, now the causal question. Flag that
the numbers are placeholders so nothing fake is ever read as real.
:::

## `[PLACEHOLDER]` Read it in three steps

::: {.panel}
Using the printed output, in your groups:

1. **Balance:** did the two groups start out alike? Where do you look?
2. **The estimate:** which single number is the impact? Which numbers are not?
3. **The rule:** does the estimate, and its confidence interval, clear the bar?
:::

::: {.ask}
`[PLACEHOLDER]` Then: what is the one thing you would ask the evaluator before
acting on this number?
:::

::: {.notes}
1:00 · Same three-step reading they just used on GreenWaste. When the real data
lands, replace this with the actual table and a short worked read.
:::

## `.ask` Group interpretation

::: {.ask}
Each group reports back: does this evaluation clear the bar, and how confident are
you? What would change your mind?
:::

::: {.panel}
Two minutes per group. Look for groups that quote the intercept, or that read the
stars as the whole answer: those are today's traps in the wild.
:::

::: {.notes}
1:03 · Keep report-backs tight. This is where you see whether the reading skills
landed.
:::

# 7 · Close [6 min]{.mins}

## Three questions to ask the evaluator

::: {.panel}
Take these back to your desk. For any headline number:
:::

```{=html}
<div class="facts">
  <div class="k">1</div> <div class="v">Compared to what? Is this before-and-after, with-and-without, or a fair comparison?</div>
  <div class="k">2</div> <div class="v">If the groups were not randomised, how do you know they started out alike?</div>
  <div class="k">3</div> <div class="v">Does the effect, and its confidence interval, clear our decision rule, not just zero?</div>
</div>
```

::: {.notes}
1:04 · The standing closing slide, tuned to today. These three questions carry
straight into Day 2.
:::

## Key takeaways

::: {.panel}
- Before-and-after folds in everything else that changed. It can understate.
- With-and-without compares groups that already differed. It can overstate.
- Randomisation makes the groups alike on average, even on what we cannot measure.
- Read for the one number that is the impact, and judge it against the rule, not zero.
:::

::: {.notes}
1:06 · Close Day 1 here. Tomorrow: what to do when you cannot draw lots. Same
programme, methods that try to rebuild a fair comparison.
:::
