---
title: "Step-by-Step Linear Regression Visualization with Dichotomous Predictor"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(dplyr)
library(patchwork)
library(ggrepel)
set.seed(123) # For reproducibility
```

## Creating Sample Data

Let's create a dataset with a dichotomous independent variable (0 or 1) and a continuous dependent variable.

```{r data-creation}
# Create dataset
n <- 100  # Sample size
x <- rep(c(0, 1), each = n/2)  # Dichotomous variable: 50 zeros and 50 ones
# Create different means for the two groups
mean_0 <- 10
mean_1 <- 15
sd <- 3  # Standard deviation for both groups

# Generate y values
y <- c(rnorm(n/2, mean = mean_0, sd = sd),  # Group 0
       rnorm(n/2, mean = mean_1, sd = sd))  # Group 1

# Create a data frame
data <- data.frame(x = x, y = y)

# Calculate group means
group_means <- data %>%
  group_by(x) %>%
  summarize(mean_y = mean(y))

# Fit linear regression
model <- lm(y ~ x, data = data)
```

## Step 1: Basic Scatterplot (Points Aligned Vertically)

First, let's plot the raw data points, which will be aligned vertically at x=0 and x=1.

```{r step1}
p1 <- ggplot(data, aes(x = factor(x), y = y)) +
  geom_point(alpha = 0.7, color = "blue", size = 3) +
  labs(title = "Step 1: Basic Scatterplot",
       subtitle = "Points aligned vertically by group (0 or 1)",
       x = "Independent Variable (X)",
       y = "Dependent Variable (Y)") +
  theme_minimal() +
  scale_x_discrete(labels = c("Group 0", "Group 1"))

p1
```

## Step 2: Add Jitter to Visualize Point Density

Now, let's add jitter to see the distribution of points better.

```{r step2}
p2 <- ggplot(data, aes(x = factor(x), y = y)) +
  geom_jitter(width = 0.2, alpha = 0.7, color = "blue", size = 3) +
  labs(title = "Step 2: Scatterplot with Jitter",
       subtitle = "Jittering added to visualize point density",
       x = "Independent Variable (X)",
       y = "Dependent Variable (Y)") +
  theme_minimal() +
  scale_x_discrete(labels = c("Group 0", "Group 1"))

p2
```

## Step 3: Add Group Means

Let's add the mean for each group.

```{r step3}
p3 <- ggplot(data, aes(x = factor(x), y = y)) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "blue", size = 3) +
  geom_point(data = group_means, aes(x = factor(x), y = mean_y), 
             color = "red", size = 5) +
  geom_text_repel(data = group_means, 
                 aes(x = factor(x), y = mean_y, 
                     label = sprintf("Mean: %.2f", mean_y)),
                 nudge_x = 0.3,
                 size = 5) +
  labs(title = "Step 3: Add Group Means",
       subtitle = "Red points represent the mean value for each group",
       x = "Independent Variable (X)",
       y = "Dependent Variable (Y)") +
  theme_minimal() +
  scale_x_discrete(labels = c("Group 0", "Group 1"))

p3
```

## Step 4: Add Regression Line

Finally, let's add the regression line that fits the data.

```{r step4}
# Create a data frame with the regression line
regression_line <- data.frame(
  x = c(0, 1),
  y = predict(model, newdata = data.frame(x = c(0, 1)))
)

# Extract the model coefficients for display
intercept <- round(coef(model)[1], 2)
slope <- round(coef(model)[2], 2)
equation <- paste("y =", intercept, "+", slope, "* x")

p4 <- ggplot() +
  geom_jitter(data = data, aes(x = factor(x), y = y), 
              width = 0.2, alpha = 0.5, color = "blue", size = 3) +
  geom_point(data = group_means, aes(x = factor(x), y = mean_y), 
             color = "red", size = 5) +
  geom_line(data = regression_line, aes(x = c(1, 2), y = y), 
            color = "darkgreen", size = 1.5) +
  geom_text(aes(x = 1.5, y = mean(regression_line$y) + 2), 
            label = equation, size = 5, color = "darkgreen") +
  labs(title = "Step 4: Add Regression Line",
       subtitle = "Green line represents linear regression",
       x = "Independent Variable (X)",
       y = "Dependent Variable (Y)") +
  theme_minimal() +
  scale_x_discrete(limits = c("0", "1"),
                   labels = c("Group 0", "Group 1"))

p4
```

## Combined Visualization

Let's combine all four steps into a single visualization to see the progression.

```{r combined}
combined_plot <- (p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Step-by-Step Linear Regression with Dichotomous Predictor",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

combined_plot
```

## Interactive Visualization (Optional)

If you prefer an interactive visualization where you can toggle between steps, you can use the following code with Shiny:

```{r eval=FALSE}
# This code is not executed here, but can be used in a Shiny app
library(shiny)

ui <- fluidPage(
  titlePanel("Step-by-Step Linear Regression"),
  
  sidebarLayout(
    sidebarPanel(
      radioButtons("step", "Select Step:",
                  choices = list(
                    "Step 1: Basic Scatterplot" = 1,
                    "Step 2: Add Jitter" = 2,
                    "Step 3: Add Group Means" = 3,
                    "Step 4: Add Regression Line" = 4
                  ),
                  selected = 1)
    ),
    
    mainPanel(
      plotOutput("regressionPlot")
    )
  )
)

server <- function(input, output) {
  output$regressionPlot <- renderPlot({
    switch(as.numeric(input$step),
           "1" = p1,
           "2" = p2,
           "3" = p3,
           "4" = p4)
  })
}

shinyApp(ui = ui, server = server)
```

## Explanation

This visualization demonstrates how linear regression works with a dichotomous independent variable:

1.  **Step 1**: We start with a basic scatterplot showing the raw data points aligned vertically at x=0 and x=1.
2.  **Step 2**: We add jitter (horizontal randomness) to the points to better visualize their distribution.
3.  **Step 3**: We calculate and display the mean value for each group (0 and 1).
4.  **Step 4**: We fit a regression line through the data, connecting the mean values of the two groups.

When the independent variable is dichotomous (0/1), the linear regression essentially: - Uses the mean of each group - The slope of the line equals the difference between the means - The intercept is the mean of group 0 - The equation `y = intercept + slope * x` predicts the value of y for each group
