---
title: "Matching Methods"
subtitle: "Econometrics with R"
author: "Dr. Lucas Sempé"
always_allow_html: yes
format:
  pptx:
    incremental: true  
  revealjs:
    theme: [clean.scss]
    slide-number: true
    code-fold: false
    highlight-style: github
editor_options: 
  chunk_output_type: console
---

## Matching Methods: Overview

:::: {.columns}

::: {.column width="60%"}
- **Core idea**: Create comparable treatment and control groups based on observable characteristics
- **Setting**: When treatment assignment is related to observable factors
- **Advantage**: Reduces selection bias from observable differences
- **Assumption**: No unobserved differences affecting both treatment and outcomes
- **Types**: Exact matching, coarsened exact matching, nearest neighbor, etc.
:::

::: {.column width="40%"}
```{r, echo=FALSE, fig.height=4}

# Load required packages
library(tidyverse)
library(MatchIt)
library(estimatr)
library(kableExtra)
library(modelsummary)

# Set rendering options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 10, 
  fig.height = 6
)

set.seed(123)

df <- read.csv("./evaluation_data_GreenWaste.csv")

# Create simulated data for matching visualization
set.seed(123)
n <- 200

# Create two variables that predict both treatment and outcome
x1 <- runif(n, -2, 2)
x2 <- runif(n, -2, 2)

# Generate treatment with higher probability for larger x1, x2
p_treat <- plogis(0.5 + 0.7*x1 + 0.7*x2)
treatment <- rbinom(n, 1, p_treat)

# Create data frame
matching_data <- data.frame(
  x1 = x1,
  x2 = x2,
  treatment = factor(treatment, labels = c("Control", "Treatment"))
)

# Plot
ggplot(matching_data, aes(x = x1, y = x2, color = treatment, shape = treatment)) +
  geom_point(size = 2) +
  labs(title = "Before Matching: Imbalanced Groups",
       x = "Variable X1",
       y = "Variable X2",
       color = "Group",
       shape = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
:::

::::

## Why Matching?


**Problem**: Selection bias

- Treatment group systematically differs from control group
- Direct comparison leads to biased impact estimates

**Solution**: Matching

- Select control units similar to treated units
- Create balance on observable characteristics
- Approximates an experimental setting

## Preparing the Data

Converting from long to wide format:

```{r}
# Create a wide-format dataset
df_w <- df %>%
  # Then create wide format
  pivot_wider(
    id_cols = c(neighborhood_identifier, business_identifier, treatment_neighborhood, 
                eligible, enrolled),
    names_from = round, # variable that determines new columns
    # variables that should be made "wide"
    values_from = c(waste_management_costs, 
                    efficiency_index, age_manager, age_deputy,
                    educ_manager, educ_deputy, female_manager,
                    foreign_owned, staff_size, advanced_filtration,
                    water_treatment_system, business_area,
                    recycling_center_distance, recycling_compliance)) %>%
  # remove the business that has missing values
  # as missing values are not allowed when using matchit
  filter(!is.na(waste_management_costs_0)) 

# Also check the first few rows to confirm format
head(select(df_w, business_identifier, enrolled, 
           waste_management_costs_0, waste_management_costs_1))
```

## Coarsened Exact Matching (CEM)

**Coarsened Exact Matching (CEM)** temporarily coarsens variables into bins, then performs exact matching on those bins:

1. **Coarsen**: Divide each variable into discrete categories (bins)
2. **Match**: Find exact matches within each combination of bins (strata)
3. **Prune**: Drop unmatched units
4. **Analyze**: Use original (uncoarsened) data for analysis

**Advantage over PSM**: CEM guarantees balance improvement—no need to iterate

## CEM: Two Scenarios

We examine two scenarios for matching enrolled vs non-enrolled businesses:

1. **Limited set of variables**:
   - Only using eligibility, based on efficiency_index

2. **Extended set of variables**:
   - Efficiency of business, distance from nearest recycling center, and characteristics of business (education of manager, staff size)

## CEM Intuition: Before Matching

```{r, echo=FALSE, fig.height=5}
# Create simulated data to illustrate CEM concept
set.seed(42)
n_sim <- 400

# Efficiency index
eff_idx <- runif(n_sim, 0, 100)

# Distance from recycling center
distance <- runif(n_sim, 0, 50)

# Treatment probability: only eligible if efficiency <= 58
# Among eligible, higher efficiency = more likely to enroll
p_enroll <- ifelse(eff_idx > 58, 0, plogis(-1.5 + 0.03 * eff_idx))
enrolled_sim <- rbinom(n_sim, 1, p_enroll)

cem_demo <- data.frame(
  efficiency = eff_idx,
  distance = distance,
  enrolled = factor(enrolled_sim, labels = c("Control", "Enrolled"))
)

# Define bin boundaries for 10x10 grid
eff_breaks <- seq(0, 100, by = 10)
dist_breaks <- seq(0, 50, by = 5)

# Plot before matching with grid
ggplot(cem_demo, aes(x = efficiency, y = distance, color = enrolled, shape = enrolled)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = eff_breaks, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = dist_breaks, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 58, linetype = "solid", color = "red", alpha = 0.5) +
  scale_color_manual(values = c("Control" = "#440154", "Enrolled" = "#21918c")) +
  labs(title = "Before CEM: All Units with Coarsened Bins",
       subtitle = "Red line = eligibility threshold (efficiency <= 58)",
       x = "Efficiency Index",
       y = "Distance from Recycling Center",
       color = "Group", shape = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## CEM Intuition: After Matching

```{r, echo=FALSE, fig.height=5}
# Assign bins
cem_demo$eff_bin <- cut(cem_demo$efficiency, breaks = eff_breaks, include.lowest = TRUE)
cem_demo$dist_bin <- cut(cem_demo$distance, breaks = dist_breaks, include.lowest = TRUE)
cem_demo$stratum <- paste(cem_demo$eff_bin, cem_demo$dist_bin)

# Keep only strata with both treated and control
strata_counts <- cem_demo %>%
  group_by(stratum) %>%
  summarise(
    n_enrolled = sum(enrolled == "Enrolled"),
    n_control = sum(enrolled == "Control")
  ) %>%
  filter(n_enrolled > 0 & n_control > 0)

cem_matched <- cem_demo %>%
  filter(stratum %in% strata_counts$stratum)

# Plot after matching
ggplot(cem_matched, aes(x = efficiency, y = distance, color = enrolled, shape = enrolled)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = eff_breaks, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = dist_breaks, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 58, linetype = "solid", color = "red", alpha = 0.5) +
  scale_color_manual(values = c("Control" = "#440154", "Enrolled" = "#21918c")) +
  labs(title = "After CEM: Only Matched Strata Retained (Bins with Both Groups)",
       subtitle = "Ineligible units (efficiency > 58) pruned - no enrolled to match",
       x = "Efficiency Index",
       y = "Distance from Recycling Center",
       color = "Group", shape = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Performing CEM

```{r}
## ----- LIMITED SET -----
# CEM with efficiency index only
cem_r <- matchit(
  enrolled ~ efficiency_index_0,
  data = df_w %>% dplyr::select(-recycling_compliance_0, -recycling_compliance_1),
  method = "cem",
  estimand = "ATT"
)

## ----- FULL SET -----
# CEM with extended set of variables
cem_ur <- matchit(
  enrolled ~ efficiency_index_0 + recycling_center_distance_0 +
    educ_manager_0 + staff_size_0,
  data = df_w %>% dplyr::select(-recycling_compliance_0, -recycling_compliance_1),
  method = "cem",
  estimand = "ATT"
)
```


## CEM Matching Summary

```{r}
# View summary of matching results
summary(cem_r)
```

## CEM Matching Summary (Full Set)

```{r}
# View summary of matching results for full set
summary(cem_ur)
```

## Understanding CEM Output

- **Matched** shows number of units retained after matching
- **Unmatched** shows units pruned (no match in opposite group)
- CEM creates **strata** based on coarsened variable combinations
- Each stratum contains both treated and control units

## Visualizing CEM: Before Matching (Real Data)

```{r, echo=FALSE, fig.height=5}
# Prepare data for visualization - use same data as CEM
cem_viz <- df_w %>%
  dplyr::select(-recycling_compliance_0, -recycling_compliance_1) %>%
  mutate(enrolled_lab = factor(enrolled, labels = c("Control", "Enrolled")))

# Get the cutpoints that CEM actually used from the matchit object
cem_cuts <- cem_ur$info$cuts
eff_breaks <- c(0, cem_cuts$efficiency_index_0, 100)
dist_breaks <- c(0, cem_cuts$recycling_center_distance_0, max(cem_viz$recycling_center_distance_0) + 1)

# Plot before matching with grid
ggplot(cem_viz, aes(x = efficiency_index_0, y = recycling_center_distance_0,
                    color = enrolled_lab, shape = enrolled_lab)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = eff_breaks, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = dist_breaks, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 58, linetype = "solid", color = "red", alpha = 0.5) +
  scale_color_manual(values = c("Control" = "#440154", "Enrolled" = "#21918c")) +
  labs(title = "Before CEM: All Units with Coarsened Bins",
       subtitle = "Red line = eligibility threshold (efficiency <= 58)",
       x = "Efficiency Index (Baseline)",
       y = "Distance from Recycling Center (Baseline)",
       color = "Group", shape = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Visualizing CEM: After Matching (Real Data)

```{r, echo=FALSE, fig.height=5}
# Get matched data - only units with valid subclass (matched strata)
cem_matched_viz <- match.data(cem_ur, drop.unmatched = TRUE) %>%
  mutate(enrolled_lab = factor(enrolled, labels = c("Control", "Enrolled")))

# Plot after matching - only matched units remain
ggplot(cem_matched_viz, aes(x = efficiency_index_0, y = recycling_center_distance_0,
                            color = enrolled_lab, shape = enrolled_lab)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = eff_breaks, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = dist_breaks, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 58, linetype = "solid", color = "red", alpha = 0.5) +
  scale_color_manual(values = c("Control" = "#440154", "Enrolled" = "#21918c")) +
  labs(title = "After CEM: Only Matched Strata Retained",
       subtitle = "Ineligible units (efficiency > 58) pruned - no enrolled to match",
       x = "Efficiency Index (Baseline)",
       y = "Distance from Recycling Center (Baseline)",
       color = "Group", shape = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Checking Balance Before Matching

```{r}
kableExtra::kable(summary(cem_ur)$sum.all,
      caption = "Balance Before Matching") %>%
  kable_styling(font_size = 10)
```

## Checking Balance After Matching

```{r}
kableExtra::kable(summary(cem_ur)$sum.matched,
      caption = "Balance After Matching") %>%
  kable_styling(font_size = 10)
```

## Understanding Balance

- **Mean Diff** should be closer to zero after matching
- Standardized mean differences (not shown) should be < 0.25
- Good balance ensures we're comparing similar facilities

## Estimating Program Impact

First, extract the matched dataset:

```{r}
# Extract matched datasets
match_df_r <- match.data(cem_r, drop.unmatched = TRUE)   # Limited set
match_df_ur <- match.data(cem_ur, drop.unmatched = TRUE) # Full set
```

## Regression with Matched Data

```{r}
# Regression with matched data
out_lm_r <- lm_robust(waste_management_costs_1 ~ enrolled,
                      data = match_df_r,  
                      clusters = neighborhood_identifier)

out_lm_ur <- lm_robust(waste_management_costs_1 ~ enrolled,
                      data = match_df_ur, 
                      clusters = neighborhood_identifier)

# Show results
modelsummary(list("Limited Set" = out_lm_r, 
                  "Full Set" = out_lm_ur),
             title = "Impact on Waste Management Costs: Matching Approach")

```

```{r}

# # # Simplified version
#
# library(tinytable)
#
# modelsummary(
#   list("Limited Covs CEM" = out_lm_r, "All Covs CEM"= out_lm_ur),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "Matching Regression Results",
#   coef_map = c(
#     "enrolled" = "Matching Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   coef_omit = NULL,
#   notes = "Standard errors are in parenthesis",
#   output = "tinytable"
# ) |>
#   style_tt(bootstrap_class = "table caption-top")
#
# ## Output for case study
# library(flextable)
# library(officer)
#
# # Build the table as a flextable (editable in Word/PowerPoint)
# ft <- modelsummary(
#   list("Limited Covs CEM" = out_lm_r, "All Covs CEM"= out_lm_ur),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "Matching Regression Results",
#   coef_map = c(
#     "enrolled" = "Matching Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   notes = "Standard errors are in parentheses",
#   fmt = 2,
#   output = "flextable"
# )
#
# # let flextable size to content, then stretch to a target width
# ft <- autofit(ft)
# ft <- fit_to_width(ft, max_width = 9)  # inches; ~content area on a PPT slide
# ft <- width(ft, j = 1, width = 2.5)    # widen the left/stub column (adjust as needed)
#
# ## 1) Save to Word (.docx)
# doc <- read_docx()
# doc <- body_add_flextable(doc, ft)
# print(doc, target = "C:/Users/FionaKastel/OneDrive - 3ie/Documents/GitHub/R-course/sessions_in_Abu_Dhabi/case_study_outputs/session10_CEM.docx")

``` 

## Matched Difference-in-Differences

Combine matching with difference-in-differences:

```{r}
# Merge matching weights back into long format
df_long_match_r <- df %>%
  left_join(match_df_r %>% dplyr::select(business_identifier, weights)) %>%
  filter(!is.na(weights))

df_long_match_ur <- df %>%
  left_join(match_df_ur %>% dplyr::select(business_identifier, weights)) %>%
  filter(!is.na(weights))
```

## Difference-in-Differences Results

```{r}
# Run DiD regression
did_reg_r <- lm_robust(waste_management_costs ~ enrolled * round,
                        data = df_long_match_r, weights = weights,
                        clusters = neighborhood_identifier)

did_reg_ur <- lm_robust(waste_management_costs ~ enrolled * round,
                        data = df_long_match_ur, weights = weights,
                        clusters = neighborhood_identifier)

# Show results
modelsummary(list("Limited Set" = did_reg_r, 
                  "Full Set" = did_reg_ur),
             coef_map = c('enrolled' = "Offered",
                          'round' = "Round",
                          'enrolled:round' = "Offered × Round"),
             title = "Impact on Waste Management Costs: Matched DiD Approach")
```

## Interpretation of Results

- **Offered** coefficient: Baseline difference between groups offered program vs those not offered program
- **Round** coefficient: Time trend for all facilities
- **Offered × Round**: The program's causal effect

::: {.callout-note}
The interaction coefficient (Offered × Round) represents our estimate of the causal impact of the program on waste management costs.
:::

## Advantages of Matching

- Reduces selection bias
- Makes treatment and control groups comparable
- Can combine with difference-in-differences for robust estimation
- Works even without baseline randomization

## Limitations of Matching

- Only controls for **observed** variables
- Cannot account for unobserved confounders
- Requires good data on pre-treatment characteristics
- Common support may be limited in some contexts

## Summary

1. **Coarsened Exact Matching (CEM)** helps create comparable groups
2. CEM **coarsens variables into bins** and performs exact matching on strata
3. Check for **balance** before and after matching
4. Combine with **regression** or **difference-in-differences** to estimate impacts
5. Interpret results with appropriate caution about unobserved confounders

## Resources for Further Learning

- **Iacus, King, and Porro (2012)**: "Causal Inference without Balance Checking: Coarsened Exact Matching"
- **Ho, Imai, King, and Stuart (2007)**: "Matching as Nonparametric Preprocessing for Reducing Model Dependence in Parametric Causal Inference"
- **Stuart (2010)**: "Matching Methods for Causal Inference: A Review and a Look Forward"
- The **MatchIt** package documentation: [https://kosukeimai.github.io/MatchIt/](https://kosukeimai.github.io/MatchIt/)

## Thank You!


















<!-- ## Comparative Results Across Methods -->

<!-- ```{r, echo=FALSE, fig.height=4} -->
<!-- library(ggplot2) -->

<!-- # Create data frame with results from all methods -->
<!-- all_results <- data.frame( -->
<!--   Method = c("Randomized Assignment", "Instrumental Variables",  -->
<!--              "Regression Discontinuity", "Difference-in-Differences",  -->
<!--              "Matching (Full Set)"), -->
<!--   Estimate = c(-1014, -982, -905, -816, -1000), -->
<!--   CI_Lower = c(-1092, -1168, -992, -879, -1072), -->
<!--   CI_Upper = c(-935, -797, -818, -753, -927) -->
<!-- ) -->

<!-- # Sort by effect size -->
<!-- all_results <- all_results %>% -->
<!--   arrange(Estimate) -->

<!-- # Plot forest plot of results -->
<!-- ggplot(all_results, aes(y = reorder(Method, Estimate), x = Estimate)) + -->
<!--   geom_vline(xintercept = -1000, linetype = "dashed", color = "red") + -->
<!--   geom_point(size = 3) + -->
<!--   geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.3) + -->
<!--   labs(title = "Impact of GreenWaste on Waste Expenditures", -->
<!--        subtitle = "Comparison Across Evaluation Methods", -->
<!--        x = "Effect on Waste Expenditures (AED)", -->
<!--        y = "") + -->
<!--   theme_minimal() + -->
<!--   annotate("text", x = -1000, y = 5.5, label = "Threshold", color = "red", hjust = 1.2) -->
<!-- ``` -->

<!-- ## Final Policy Recommendation -->

<!-- - Methods produce consistent results in the range of -816 to -1033 AED -->
<!-- - Randomized assignment (gold standard): -1,014 AED -->
<!-- - Most quasi-experimental methods: Effect slightly below $10 -->
<!-- - Given uncertainty in estimates and consistency in direction, GreenWaste shows promising results -->

<!-- **RECOMMENDATION**: The preponderance of evidence suggests GreenWaste reduces waste expenditures by approximately $10, which meets the threshold criterion. Given the consistency across methods and the gold-standard randomized result, GreenWaste should be scaled up nationally. -->

<!-- ## Conclusion: Choosing the Right Method -->

<!-- | Method | When to Use | Key Assumption | Strengths | Limitations | -->
<!-- |--------|-------------|----------------|-----------|-------------| -->
<!-- | **Randomized** | Feasible to randomize | Random assignment | Gold standard; controls for all confounders | Implementation challenges; external validity | -->
<!-- | **IV** | Imperfect compliance; valid instrument exists | Exclusion restriction | Addresses selection; natural experiments | Local effect; requires strong instrument | -->
<!-- | **RDD** | Clear eligibility threshold | No manipulation of running variable | Addresses selection bias through psuedo-randomization | Local to threshold; requires threshold | -->
<!-- | **DiD** | Panel data; non-random assignment | Parallel trends | Controls for fixed differences and time trends | Cannot address time-varying confounders | -->
<!-- | **Matching** | Rich observational data | Selection on observables | Uses existing data; intuitive | Cannot address unobserved confounders | -->

<!-- ## Next Steps in Impact Evaluation -->

<!-- 1. **Heterogeneity analysis** -->
<!--    - Does impact vary across subgroups? -->
<!--    - Who benefits most from GreenWaste? -->

<!-- 2. **Cost-effectiveness analysis** -->
<!--    - Beyond impact, is GreenWaste cost-effective? -->
<!--    - How does it compare to alternative programs? -->

<!-- 3. **Implementation research** -->
<!--    - What factors affect successful implementation? -->
<!--    - How to optimize national scale-up? -->

<!-- 4. **Long-term follow-up** -->
<!--    - Do impacts persist over time? -->
<!--    - Are there additional benefits or unintended consequences? -->

<!-- ## Key Course Takeaways -->

<!-- 1. Impact evaluation requires identifying a valid counterfactual -->
<!-- 2. Randomized experiments provide the most credible counterfactual -->
<!-- 3. Quasi-experimental methods can provide rigorous evidence when randomization isn't possible -->
<!-- 4. Multiple methods with consistent results strengthen the evidence base -->
<!-- 5. Statistical significance must be paired with policy significance -->
<!-- 6. The choice of method depends on context, data, and research question -->
<!-- 7. R provides powerful tools for implementing all these methods -->

<!-- ## Thank You! -->

<!-- **Additional Resources**: -->

<!-- - Gertler, P. J., et al. (2016). *Impact Evaluation in Practice*, Second Edition. World Bank. -->
<!-- - Angrist, J. D., & Pischke, J. S. (2008). *Mostly Harmless Econometrics*. Princeton University Press. -->
<!-- - Cunningham, S. (2021). *Causal Inference: The Mixtape*. Yale University Press. -->
<!-- - R packages: `estimatr`, `MatchIt`, `rdrobust`, `did`, `fixest`, `AER` -->