---
title: "Instrumental Variables"
subtitle: "Econometrics with R"
author: "Dr. Lucas Sempé"
always_allow_html: yes
keep-md: true
format: 
  pptx:
    incremental: true  
  revealjs:
    theme: [clean.scss]
    slide-number: true
    code-fold: false
    highlight-style: github
editor_options: 
  chunk_output_type: console
---
## The Challenge of Imperfect Compliance

- Sometimes we can't randomly assign the treatment itself (e.g., can't force uptake of the program after we offer it)
- But we can randomly assign our offering of the treatment (we can randomly offer the program to some businesses)

**Central question:** How do we measure impact when actual participation of those offered the program is voluntary?

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(modelsummary)
library(estimatr)
library(kableExtra)
library(flextable)
```


## Instrumental Variables Framework

:::: {.columns}

::: {.column width="50%"}
**Instrumental Variable (IV)** must:

1. **Affect participation**: The instrument increases likelihood of enrollment
2. **Exclusion restriction**: The instrument affects outcomes only through enrollment
3. **Independence**: The instrument is as good as randomly assigned
4. **Monotonicity**: The instrument doesn't discourage anyone from enrolling
:::

::: {.column width="50%"}

<!-- ```{mermaid} -->
<!-- %%| label: fig-mermaid -->
<!-- %%| fig-width: 6 -->
<!-- %%| fig-cap: | -->
<!-- %%|   A valid instrument affects the Treatment, and **only** affects the outcomes through the treatment. -->
<!-- flowchart LR -->
<!--   A[Instrument] --> B(Treatment) -->
<!--   A -.-> |No| C(Outcome) -->
<!--   B --> C -->
<!-- ``` -->

:::

::::

## Example Case: Randomized Design with Impartial Takeup

- Businesses offered the program, select whether to enroll in the program
- In this case, the effect from our RCT, was the ITT, or intent-to-treat of the program 
- We have a new variable that tells us whether businesses that were randomly assigned the program actually enrolled (enrolled_rp)
- Using this variable, we can estimate the ToT, or actual treatment-on-treated of the program.

```{r, echo=F}
df <- read.csv("./evaluation_data_GreenWaste.csv")
```

If the randomization is effective, then we should have a higher likelihood of participation in the intervention for those who were provided access to it. If we are just interested in the question of whether randomizing access to enrollment in the project improved our outcomes compared to providing the program to everyone, then we can run the regression for those that randomly received access to the program versus those that did not. This is what we learned about randomization so far. 

However, we ideally want an estimate of the program not the random assignment or access to the program, and there also may be some businesses that received access but still didn’t take up the program. This creates selection bias since those that actually take up the program may have higher outcomes to begin with. We want to run our analyses to see if receiving access to the program encouraged participants to take part in the project AND if that participation impacted their waste management costs.

How do we do this? In two stages.

## 1. First Stage: Effect of Randomization on Enrollment
:::: {.columns}

::: {.column width="50%"}

In the first stage, we measure the effect of offering the program on participation in the program (e.g., how much more likely someone is to participate when provided the program).


```{r, eval=T, echo=TRUE}
# Estimate effect of providing access on enrollment (First Stage)
m_first_stage <- lm_robust(enrolled_rp ~ treatment_neighborhood,
                      clusters = business_identifier,
                      data = df %>% filter(round == 1 & eligible ==1))

```


**First Stage Results**:

- Providing access to the program increases enrollment by `r round(m_first_stage$coefficients[2]*100, 1)` percentage points
- As expected, without access, only `r round(m_first_stage$coefficients[1]*100, 1)`% enroll
- With access, `r round((m_first_stage$coefficients[1] + m_first_stage$coefficients[2])*100, 1)`% enroll
- Strong first stage (F-statistic > 10): `r round(m_first_stage[["fstatistic"]][["value"]],1)`

> **What is the F-statistic?**  
> The F-statistic tests whether the instrument (access to program) is a **strong predictor** of the endogenous variable (enrollment).  
> - A **high F-statistic** (greater than 10) suggests the instrument is strong and relevant.  
> - A **low F-statistic** (below 10) would indicate a **weak instrument**, which could bias IV estimates and make them unreliable.
:::

::: {.column width="50%"}
```{r, echo=FALSE}

modelsummary(
  list("Enrollment"= m_first_stage),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  title = "First Stage Regression Results",
  coef_map = c(
    "treatment_neighborhood" = "Random Access Effect",
    "(Intercept)" = "(Intercept)"
  ),
  coef_omit = NULL,
  notes = "Standard errors are in parenthesis",
  output = "tinytable"
) |>
  style_tt(bootstrap_class = "table caption-top")

```
:::

::::

```{r}

# Simplified version

# library(tinytable)
# 
# modelsummary(
#   list("Enrollment"= m_first_stage),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "First Stage Regression Results",
#   coef_map = c(
#     "treatment_neighborhood" = "Random Access Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   coef_omit = NULL,
#   notes = "Standard errors are in parenthesis",
#   output = "tinytable"
# ) |>
#   style_tt(bootstrap_class = "table caption-top")

# ## Output for case study
# library(officer)
# 
# # Build the table as a flextable (editable in Word/PowerPoint)
# ft <- modelsummary(
#   list("Enrollment" = m_first_stage),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "First Stage Regression Results",
#   coef_map = c(
#     "treatment_neighborhood" = "Random Access Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   notes = "Standard errors are in parentheses",
#   fmt = 2,
#   output = "flextable"
# )
# 
# # let flextable size to content, then stretch to a target width
# ft <- autofit(ft)
# ft <- fit_to_width(ft, max_width = 9)  # inches; ~content area on a PPT slide
# ft <- width(ft, j = 1, width = 2.5)    # widen the left/stub column (adjust as needed)
# # optional: a bit of breathing room
# # ft <- padding(ft, padding = 2)
# 
# ## 1) Save to Word (.docx)
# doc <- read_docx()
# doc <- body_add_flextable(doc, ft)
# print(doc, target = "C:/Users/FionaKastel/OneDrive - 3ie/Documents/GitHub/R-course/sessions_in_Abu_Dhabi/case_study_outputs/session7_IV_first_stage.docx")


```  

## 2. Intention-to-Treat (ITT) Effect

The second stage is the link between participation on our outcomes of interest. This second stage is the effect of the *access to the program* on the outcome (referred to as the “reduced form”). This is also known as the Intention-to-Treat (ITT) Effect or the effect on outcomes based on the original randomization, regardless of the actual treatment or compliance with the program.


:::: {.columns}

::: {.column width="40%"}
```{r, eval=T, echo=TRUE}
#| class: small-code

# Estimate ITT effect (reduced form)
m_itt <- lm_robust(waste_management_costs ~ treatment_neighborhood,
                   clusters = neighborhood_identifier,
                   data = df %>% filter(round == 1 & eligible ==1))

# With covariate adjustment
m_itt_wcov <- lm_robust(waste_management_costs ~ treatment_neighborhood + 
                        age_manager + age_deputy +
                        female_manager + foreign_owned + 
                        staff_size +
                        advanced_filtration + 
                        business_area +
                        recycling_center_distance,
                        clusters = neighborhood_identifier,
                        data = df %>% filter(round == 1 & eligible ==1))
```


**ITT Results**:

- Promotion directly reduces waste management costs by `r round(m_itt$coefficients[2], 0)` units
- Promotion directly reduces waste management costs
- This is the policy-relevant effect of offering promotion
- But this underestimates the effect on those who actually enroll
:::

::: {.column width="60%"}
```{r, echo=FALSE}

modelsummary(
  list("Model 1: No covariate adj."= m_itt, "Model 2: With covariate adj." = m_itt_wcov),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  title = "First Stage Regression Results",
  coef_rename = c("treatment_neighborhood" = "ITT Effect"),
  coef_omit = NULL,
  notes = "Standard errors are in parenthesis",
  output = "tinytable"
) |>
  style_tt(bootstrap_class = "table caption-top")

# modelsummary(list("No covariate adj."= m_itt, "With covariate adj." = m_itt_wcov), stars = TRUE,
#              gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
#   kable_styling(font_size = 13) %>% scroll_box(width = "700px", height = "400px")
```
:::

```{r}

# Simplified version

# library(tinytable)
# 
# modelsummary(
#   list("ITT"= m_itt),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "ITT Regression Results",
#   coef_map = c(
#     "treatment_neighborhood" = "RCT ITT Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   coef_omit = NULL, 
#   notes = "Standard errors are in parenthesis",
#   output = "tinytable"
# ) |>
#   style_tt(bootstrap_class = "table caption-top")

# ## Output for case study
# library(flextable)
# library(officer)
# library(openxlsx)
# 
# # Build the table as a flextable (editable in Word/PowerPoint)
# ft <- modelsummary(
#   list("ITT" = m_itt),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "ITT Regression Results",
#   coef_map = c(
#     "treatment_neighborhood" = "RCT ITT Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   notes = "Standard errors are in parentheses",
#   fmt = 2,
#   output = "flextable"
# )
# 
# # let flextable size to content, then stretch to a target width
# ft <- autofit(ft)
# ft <- fit_to_width(ft, max_width = 9)  # inches; ~content area on a PPT slide
# ft <- width(ft, j = 1, width = 2.5)    # widen the left/stub column (adjust as needed)
# # optional: a bit of breathing room
# # ft <- padding(ft, padding = 2)
# 
# ## 1) Save to Word (.docx)
# doc <- read_docx()
# doc <- body_add_flextable(doc, ft)
# print(doc, target = "C:/Users/FionaKastel/OneDrive - 3ie/Documents/GitHub/R-course/sessions_in_Abu_Dhabi/case_study_outputs/session7_IV_ITT.docx")


``` 

::::

## 3. Local Average Treatment Effect (LATE)

We still don’t know the effect of the participation on outcomes, which is what we want. To get the estimate of interest (effect of participation on outcomes), we need to divide the “reduced form” (the effect of offering the program on outcomes) by the first stage (effect of offering the program on participation).

:::: {.columns}

::: {.column width="50%"}
```{r, eval=T, echo=TRUE}
# Estimate LATE using IV regression (2SLS)
m_late <- iv_robust(waste_management_costs ~ enrolled_rp |
                     treatment_neighborhood,
                   clusters = neighborhood_identifier,
                   data = df %>% filter(round == 1 & eligible ==1))

# With covariate adjustment
m_late_wcov <- iv_robust(waste_management_costs ~ enrolled_rp + 
                         age_manager + age_deputy +
                         female_manager + foreign_owned + 
                         staff_size +
                         advanced_filtration + 
                         business_area +
                         recycling_center_distance | 
                         treatment_neighborhood + 
                         age_manager + age_deputy +
                         female_manager + foreign_owned + 
                         staff_size +
                         advanced_filtration + business_area +
                         recycling_center_distance,
                         clusters = neighborhood_identifier,
                         data = df %>% filter(round == 1 & eligible ==1))
```
:::

::: {.column width="50%"}
```{r, echo=FALSE}

modelsummary(list("No covariate adj."= m_late, "With covariate adj." = m_late_wcov), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
  kable_styling(font_size = 15) %>% scroll_box(width = "600px", height = "300px")

```

```{r}

# Simplified version

# ## Output for case study
# library(flextable)
# library(officer)
# 
# # Build the table as a flextable (editable in Word/PowerPoint)
# ft <- modelsummary(
#   list("LATE" = m_late),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "LATE Regression Results",
#   coef_map = c(
#     "enrolled_rp" = "RCT LATE Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   notes = "Standard errors are in parentheses",
#   fmt = 2,
#   output = "flextable"
# )
# 
# # let flextable size to content, then stretch to a target width
# ft <- autofit(ft)
# ft <- fit_to_width(ft, max_width = 9)  # inches; ~content area on a PPT slide
# ft <- width(ft, j = 1, width = 2.5)    # widen the left/stub column (adjust as needed)
# 
# ## 1) Save to Word (.docx)
# doc <- read_docx()
# doc <- body_add_flextable(doc, ft)
# print(doc, target = "C:/Users/FionaKastel/OneDrive - 3ie/Documents/GitHub/R-course/sessions_in_Abu_Dhabi/case_study_outputs/session7_IV_LATE.docx")

``` 

**LATE Explanation**:

- Another way of calculating is: LATE = ITT / First Stage = `r round(m_itt$coefficients[2], 0)` / `r round(m_first_stage$coefficients[2], 3)` = `r round(m_itt$coefficients[2]/m_first_stage$coefficients[2], 0)`
- Ultimately, we are able to isolate the effect of the program, by using the instrument. What we have now is called the LATE (local average treatment effect) – we have an estimate of an impact that is attributable to the program for those who complied with the offering.
- This estimates the effect for "compliers": businesses that enrolled because they were offered the program

:::

::::
## Understanding LATE: The Complier Effect
The IV estimate (LATE) applies to compliers only:

:::: {.columns}
::: {.column width="50%"}
**Types of businesses:**

- Always-takers: Willing to enroll regardless of being offered the program
- Compliers: Enroll only if offered the program
- Never-takers: Don't enroll regardless of whether they are offered the program 
- Defiers: Enroll only without being offered the program (0% by assumption)
:::

::: {.column width="50%"}
**Implications:**

- If treatment effects vary, LATE may not generalize
- External validity depends on how representative compliers are
- Policy relevant if we care about those influenced by being offered the program
:::
::::

## Key Assumptions of IV

- Relevance: Offering the program effectively increases enrollment (✓ Strong first stage)

- Exclusion restriction: Offer affects outcomes only through enrollment
  - Offer must not directly affect waste management costs
  - Careful design focused only on awareness, not waste management behaviors

- Independence: Offer randomly assigned (✓ Baseline balance)

- Monotonicity: Offer doesn't discourage anyone from enrolling
  - No "defiers" who would enroll only without being offered the program

## From Analysis to Policy Decision
**Policy question:** Should the program be scaled up nationally?

**Decision criterion:** Program must reduce waste management costs by at least 1,000 AED.

**Results from IV estimation:**

- LATE: -1,033 AED (from our calculation above)
- Point estimate is above the 1,000 AED threshold
- The IV estimate represents the effect on compliers only

**Recommendation:** Based strictly on the estimate, the program should be scaled up nationally. 

## IV vs. Randomized Assignment
:::: {.columns}
::: {.column width="50%"}
**IV Estimate:**

- LATE = -1,033 AED
- Effect on compliers only
- Larger standard errors

**Policy recommendation:**
Scale up since above threshold
:::
::: {.column width="50%"}
**Randomized Assignment:**

- ATE = -1,014 AED
- Effect on all eligible facilities
- More precise estimate

**Policy recommendation:**
Scale up nationally
:::
::::
Why the difference?

- Different populations: LATE vs. ATE
- Different precision: IV is less efficient
- Different contexts: National scale-up vs. pilot program

## When to Use Instrumental Variables
Use when:

- Random assignment of the actual treatment is not possible
- You have access to a valid instrument
- Partial compliance is expected
- You focus on the effect for compliers

Common applications:

- Randomized encouragement designs
- Natural experiments (lotteries, policy changes)
- Geographic or administrative discontinuities

## Common Pitfalls in IV Analysis

Weak instruments (First-stage F-statistic < 10):

- Leads to biased estimates and poor inference
- Solution: Find stronger instruments or use weak-IV robust methods

Violation of exclusion restriction:

- Instrument affects outcome through other channels
- Often untestable; requires strong theoretical justification

Heterogeneous treatment effects:

- LATE may not generalize to other populations
- Be careful about policy recommendations

Improper standard errors:

- Remember to cluster standard errors when appropriate
- Use robust methods for inference

## Key Takeaways

- IV addresses selection bias when randomization of treatment isn't possible
- LATE estimates the effect for compliers only
- IV requires strong assumptions, especially exclusion restriction
- The program reduced waste management costs by 1,033 AED for compliers, above the threshold
- IV estimates are less precise than randomized assignment estimates

## Next Session

Regression Discontinuity Design: Exploiting eligibility thresholds to estimate program effects