---
title: "Regression Discontinuity Design"
subtitle: "Econometrics with R"
author: "Dr. Lucas Sempé"
always_allow_html: yes
format: 
  pptx:
    incremental: true  
  revealjs:
    theme: [clean.scss]
    slide-number: true
    code-fold: false
    highlight-style: github
editor_options: 
  chunk_output_type: console
---

## Regression Discontinuity Design: Overview

:::: {.columns}

::: {.column width="60%"}
- **Core idea**: Compare units just above and below an eligibility threshold
- **Setting**: Assignment based on a continuous "running variable"
- **Advantage**: Credible causal inference from non-experimental data
- **Assumption**: Units close to threshold are similar across all characteristics (observable and un-observable) except for treatment status and this threshold cannot be manipulated by others and is not being used by other projects provided to the same group
- **Types**: Sharp RDD (deterministic cutoff) vs. Fuzzy RDD (probabilistic cutoff)
:::

::: {.column width="40%"}
```{r, echo=FALSE, fig.height=4.5}
library(tidyverse)
library(estimatr)
library(kableExtra)
library(modelsummary)

set.seed(123)

df <- read.csv("./evaluation_data_GreenWaste.csv")

# Create data subset with only treatment localities
df_treat <- df %>%
  filter(treatment_neighborhood == 1)

# Create simulated RDD data
n <- 200
cutoff <- 0
data <- data.frame(
  running_var = runif(n, -5, 5),
  noise = rnorm(n, 0, 0.5)
)

# Generate outcome with discontinuity at cutoff
data$outcome <- 0.2 * data$running_var + 
                ifelse(data$running_var >= cutoff, -1.5, 0) + 
                data$noise

# Create RDD plot
ggplot(data, aes(x = running_var, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  geom_smooth(data = subset(data, running_var < cutoff), 
              method = "lm", color = "blue") +
  geom_smooth(data = subset(data, running_var >= cutoff), 
              method = "lm", color = "red") +
  annotate("text", x = -2.5, y = 0, label = "Control Group") +
  annotate("text", x = 2.5, y = -1.5, label = "Treatment Group") +
  labs(title = "Regression Discontinuity Design",
       x = "Running Variable",
       y = "Outcome") +
  theme_minimal()
```
:::

::::

## Case: Efficiency Index Threshold

- Program targets businesses using a national efficiency index
- Index assigns scores from 20 to 100 based on business characteristics
- Official efficiency line set at 58
- Businesses with score ≤ 58 are eligible for GreenWaste Program
- Only eligible businesses in treatment neighborhoods were offered the program and could enroll

## Checking for Manipulation

Before applying RDD, we must check for manipulation of the eligibility index:

```{r, eval=T, echo=TRUE}
# Plot density of efficiency index
ggplot(df_treat, aes(x = efficiency_index )) +
  geom_vline(xintercept = 58) +
  geom_density() +
  labs(x = "Efficiency Index") + theme_minimal()

```

:::: {.columns}

::: {.column width="50%"}
- No evidence of "bunching" just below cutoff
- Formal McCrary density test confirms
- Businesses do not appear to manipulate their scores
:::

::: {.column width="50%"}
```{r, echo=FALSE, fig.height=4}

# Formal test for manipulation
library(rddensity)

test_density <- rdplotdensity(rdd = rddensity(df_treat$efficiency_index, c = 58), 
                              X = df_treat$efficiency_index, 
                              type = "both")

```
:::

::::

## Verifying Compliance (Sharp RDD)


:::: {.columns}

::: {.column width="50%"}
- Perfect compliance with eligibility rules
- All businesses with score ≤ 58 enrolled
- No businesses with score > 58 enrolled
- This is a "sharp" RDD with deterministic assignment
:::

::: {.column width="50%"}
```{r, eval=T, echo=TRUE}
# Check if eligibility rules were followed
ggplot(df_treat, aes(y = intent_to_treat, x = efficiency_index)) +
  geom_vline(xintercept = 58) +
  geom_point() +
  labs(title = "Enrollment by Efficiency Index (Sharp RDD)",
       x = "Efficiency Index",
       y = "Enrolled (1 = Yes, 0 = No)") +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_minimal()
```
:::

::::

## Visualizing the RDD Relationship

:::: {.columns}

::: {.column width="50%"}
- Clear discontinuity at cutoff (58)
- Waste expenditures jump higher above the threshold
- Linear relationship within each group
- Visual evidence of program impact
:::

::: {.column width="50%"}

```{r, eval=T, echo=T}
# Plot relationship between efficiency index and waste expenditures
df_treat %>%
  filter(round == 1) %>%
  mutate(enrolled_lab = ifelse(intent_to_treat == 1, "Offered", "Not Offered")) %>%
  ggplot(aes(x = efficiency_index, y = waste_management_costs,
             group = enrolled_lab, colour = enrolled_lab)) +
  geom_point(alpha = 0.03) +
  geom_smooth(method = "lm") +
  geom_vline(xintercept = 58) +
  labs(x = "Efficiency Index", y = "Waste Expenditures") + theme_minimal()
```

:::

::::

## Estimating the RDD Impact

:::: {.columns}

::: {.column width="50%"}

```{r, eval=T, echo=TRUE}
#| class: small-code

# Prepare data by centering the running variable
df_treat <- df_treat %>%
  mutate(efficiency_index_c0 = efficiency_index - 58)

# Basic RDD regression with covariates
out_rdd <- lm_robust(waste_management_costs ~ 
                  efficiency_index_c0 * intent_to_treat +
                  age_manager + age_deputy +
                  female_manager + foreign_owned + 
                  staff_size +
                  advanced_filtration + business_area +
                  recycling_center_distance,
                  data = df_treat %>% filter(round == 1))
```

:::

::: {.column width="50%"}

```{r}

modelsummary(list("Sharp RDD"= out_rdd), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"),output = 'kableExtra') %>%
  kable_styling (font_size = 15) 

```

- Treatment reduces waste expenditures by approximately $9051

```{r}

# Simplified version

# ## Output for case study
# library(flextable)
# library(officer)
# 
# # Build the table as a flextable (editable in Word/PowerPoint)
# ft <- modelsummary(
#   list("Sharp_RDD" = out_rdd),
#   stars = TRUE,
#   gof_map = c("nobs"),
#   title = "RDD Regression Results",
#   coef_map = c(
#     "intent_to_treat" = "RDD Effect",
#     "(Intercept)" = "(Intercept)"
#   ),
#   notes = "Standard errors are in parentheses",
#   fmt = 2,
#   output = "flextable"
# )
# 
# # let flextable size to content, then stretch to a target width
# ft <- autofit(ft)
# ft <- fit_to_width(ft, max_width = 9)  # inches; ~content area on a PPT slide
# ft <- width(ft, j = 1, width = 2.5)    # widen the left/stub column (adjust as needed)
# 
# ## 1) Save to Word (.docx)
# doc <- read_docx()
# doc <- body_add_flextable(doc, ft)
# print(doc, target = "C:/Users/FionaKastel/OneDrive - 3ie/Documents/GitHub/R-course/sessions_in_Abu_Dhabi/case_study_outputs/session8_RDD.docx")

``` 

:::
::::

## Quadratic and polinomial models

In Regression Discontinuity Designs (RDD), the relationship between the running variable and the outcome might not be perfectly linear. To make sure the results are robust, we estimate:

1. A **quadratic model** (2nd degree polynomial) — allowing for a curved relationship.
2. A **cubic model** (3rd degree polynomial) — allowing for even more flexibility in curvature.

This helps test whether the RDD results are sensitive to how we model the relationship near the cutoff.

:::: {.columns}

::: {.column width="50%"}

```{r, eval=T, echo=TRUE}
#| class: small-code

# Quadratic RDD model:
# Include efficiency_index_c0 and its square (efficiency_index_c0^2)
# Interact them with treatment (intent_to_treat) to allow for different slopes
# and curves on each side of the cutoff
out_rdd_quadratic  <- lm_robust(waste_management_costs ~ 
                  efficiency_index_c0 * intent_to_treat +
                    intent_to_treat * I(efficiency_index_c0^2) +
                  age_manager + age_deputy +
                  female_manager + foreign_owned + 
                  staff_size +
                  advanced_filtration + business_area +
                  recycling_center_distance,
                  data = df_treat %>% filter(round == 1))

# Cubic RDD model:
# Same as above, but also add a cubic (efficiency_index_c0^3) term
# for even more flexible modeling of the relationship
out_rdd_cubic  <- lm_robust(waste_management_costs ~ 
                  efficiency_index_c0 * intent_to_treat +
                    intent_to_treat * I(efficiency_index_c0^2) + 
                        intent_to_treat * I(efficiency_index_c0^3) +
                  age_manager + age_deputy +
                  female_manager + foreign_owned + 
                  staff_size +
                  advanced_filtration + business_area +
                  recycling_center_distance,
                  data = df_treat %>% filter(round == 1))
```

:::

::: {.column width="50%"}

```{r}

# Print results in a nice table
# Omit some coefficients for readability and show goodness-of-fit metrics
modelsummary(coef_omit = c(-3),
             list("Quadratic RDD"= out_rdd_quadratic,"Cubic RDD" = out_rdd_cubic), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
  kable_styling (font_size = 15)


```

- RDD estimates are robust across different specifications
- Effect is statistically significant but below the $10,000 threshold
- Results are similar to the IV estimates

:::
::::

## Robustness Checks for RDD

:::: {.columns}

::: {.column width="50%"}

**1. Bandwidth selection**
```{r, eval=T, echo=TRUE}
# Restrict to observations near cutoff
out_rdd5 <- lm_robust(
  waste_management_costs ~ 
    intent_to_treat * efficiency_index_c0 +
          age_manager + age_deputy +
           female_manager + foreign_owned + 
           staff_size +
                  advanced_filtration + business_area +
                  recycling_center_distance,
  data = df_treat %>% 
    filter(round == 1 & 
           abs(efficiency_index_c0) <= 53))

```

```{r}

modelsummary(coef_omit = c(-2), list("Restricted RDD"= out_rdd5),
             stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
  kable_styling (font_size = 15)

```

:::

::: {.column width="50%"}


**2. Placebo tests**

- Test for "effects" at fake cutoff points
- Examine outcomes that shouldn't be affected

:::

::::

## Local Treatment Effect in RDD

:::: {.columns}

::: {.column width="50%"}
- RDD estimates are **local** to the threshold
- They represent the effect for businesses with scores near 58
- Businesses at the margin of eligibility
- Less representative of very low efficiency (low scores)
- External validity depends on whether effects vary with efficiency level
:::

::: {.column width="50%"}
```{r, echo=FALSE, fig.height=4}
library(ggplot2)

# Create data for local effect illustration
x_range <- seq(40, 80, by = 0.1)
cutoff <- 58

# Generate treatment effects that vary with efficiency index
effects <- data.frame(
  efficiency_index = x_range,
  effect_size = -15 + 0.1 * (x_range - 40)
)

# Plot varying treatment effects
ggplot(effects, aes(x = efficiency_index, y = effect_size)) +
  geom_line(size = 1, color = "blue") +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -9, linetype = "dotted") +
  annotate("text", x = 70, y = -9, label = "RDD Estimate", hjust = 0) +
  annotate("segment", x = cutoff, xend = cutoff, y = -15, yend = -9, 
           arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  annotate("text", x = cutoff + 0.5, y = -12, label = "Local Effect", hjust = 0) +
  labs(title = "Treatment Effects May Vary with efficiency Index",
       x = "efficiency Index",
       y = "Treatment Effect on Waste Expenditures") +
  ylim(-15, -5) +
  theme_minimal()
```
:::

::::

## RDD vs. Other Methods

:::: {.columns}

::: {.column width="33%"}
**RDD**:

- Effect: -905 AED
- Local to cutoff
- Uses eligibility threshold
- Sharp discontinuity

**Recommendation**: Do not scale up
:::

::: {.column width="33%"}
**IV**:

- Effect: -1033 AED
- Effect on compliers
- Uses randomized offering of program to get ToT effect

**Recommendation**: Scale up nationally
:::

::: {.column width="33%"}
**Randomized**:

- Effect: -1,014 AED
- Average effect
- Uses random assignment
- Full experimental control

**Recommendation**:  Scale up nationally
:::

::::

**Why the differences?**

- Different populations (local vs. compliers vs. average)
- Different estimation methods
- Different precision levels

## When to Use RDD

**Use when**:

- Clear eligibility threshold exists
- Running variable is continuous and measured precisely
- Units cannot precisely manipulate their position around threshold
- We have sufficient participants/units within the bandwidth that we set
- Interest in effects for marginal participants

**Common applications**:

- Test score cutoffs for educational programs
- Income thresholds for social programs
- Age eligibility for services
- Geographic boundaries for policies

## Implementing RDD in R

```{r, eval=FALSE, echo=TRUE}
# Basic steps for RDD analysis

# Step 1: Check for manipulation
library(rddensity)
rdd_test <- rddensity(data$running_var, c = cutoff)
rdplotdensity(rdd_test, data$running_var)

# Step 2: Center the running variable
data$centered_var <- data$running_var - cutoff

# Step 3: Estimate RDD effect (parametric)
model <- lm_robust(
  outcome ~ centered_var * treatment + controls,
  data = data
)

# Step 4: Non-parametric approach
library(rdrobust)
rd_robust <- rdrobust(
  y = data$outcome,
  x = data$running_var,
  c = cutoff
)
```

## Common Pitfalls in RDD Analysis

1. **Manipulation of the running variable**
   - Units strategically position themselves around threshold
   - Check density for unusual bunching

2. **Inappropriate bandwidth**
   - Too wide: Bias from including dissimilar units
   - Too narrow: Imprecision from small sample
   - Use data-driven bandwidth selection methods

3. **Misspecification of functional form**
   - Linear may not capture true relationship
   - Check robustness to different specifications

4. **Multiple cutoffs or treatments**
   - Other policies may use the same threshold
   - Separate effects of different treatments

## From Analysis to Policy Decision

**Policy question**: Should the Program be scaled up nationally?

**Decision criterion**: Program must reduce waste expenditures by at least 1,000 Dirham.

**Results from RDD estimation**:

- Effect: $`{r} round(out_rdd$coefficients[3],0)` (95% CI: [ $`{r} round(out_rdd$coefficients[3] + 1.96*out_rdd$std.error[3],0)`, $`{r} round(out_rdd$coefficients[3] - 1.96*out_rdd$std.error[3],0)`])
- Point estimate below the 1,000 AED threshold
- Effect applies to businesses near the efficiency threshold

**Recommendation**: Based on the RDD estimate, the program should not be scaled up nationally.

## Key Takeaways

1. RDD exploits discontinuities at eligibility thresholds to estimate causal effects
2. Units just above and below the threshold are comparable except for treatment
3. RDD estimates are local to the cutoff and may not generalize to units far from threshold
4. Multiple specifications yield consistent estimates of approximately 1,000 AED reduction
5. RDD estimate is slightly below the 1,000 AED threshold for scaling up GreenWaste

## Next Session

**Difference-in-Differences**: Comparing changes over time between treatment and comparison groups