---
title: "Instrumental Variables"
subtitle: "Econometrics with R"
author: "Dr. Lucas Sempé"
always_allow_html: yes
keep-md: true
format: 
  pptx:
    incremental: true  
  revealjs:
    theme: [clean.scss]
    slide-number: true
    code-fold: false
    highlight-style: github
editor_options: 
  chunk_output_type: console
---
## The Challenge of Imperfect Compliance

- Sometimes we can't randomly assign the treatment itself
- But we can randomly assign encouragement to take up the treatment
- Example: National scale-up of the Program means we can't deny eligibility
- But we can randomly promote the program in some industries

**Central question:** How do we measure impact when actual participation is voluntary?

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(modelsummary)
library(estimatr)
library(kableExtra)
```


## Instrumental Variables Framework

:::: {.columns}

::: {.column width="50%"}
**Instrumental Variable (IV)** must:

1. **Affect participation**: The instrument increases likelihood of enrollment
2. **Exclusion restriction**: The instrument affects outcomes only through enrollment
3. **Independence**: The instrument is as good as randomly assigned
4. **Monotonicity**: The instrument doesn't discourage anyone from enrolling
:::

::: {.column width="50%"}

<!-- ```{mermaid} -->
<!-- %%| label: fig-mermaid -->
<!-- %%| fig-width: 6 -->
<!-- %%| fig-cap: | -->
<!-- %%|   A valid instrument affects the Treatment, and **only** affects the outcomes through the treatment. -->
<!-- flowchart LR -->
<!--   A[Instrument] --> B(Treatment) -->
<!--   A -.-> |No| C(Outcome) -->
<!--   B --> C -->
<!-- ``` -->

:::

::::

## Example Case: Randomized Promotion Design

- The ministry wants to make the Program available to all industries
- We design an evaluation with random promotion in some zones of the Emirate
- Promotion includes communication and social marketing
- This creates exogenous variation in enrollment

```{r, echo=F}
df <- read.csv("./evaluation_data.csv")
df$waste_management_costs<-df$waste_management_costs*1000
```

## 1. First Stage: Effect of Promotion on Enrollment
:::: {.columns}

::: {.column width="50%"}

```{r, eval=T, echo=TRUE}
# Estimate effect of promotion on enrollment (First Stage)
m_first_stage <- lm_robust(enrolled_rp ~ promotion_zone,
                      clusters = facility_identifier,
                      data = df %>% filter(round == 1))
```


**First Stage Results**:

- Promotion increases enrollment by `r round(m_first_stage$coefficients[2]*100, 1)` percentage points
- Without promotion, only `r round(m_first_stage$coefficients[1]*100, 1)`% enroll
- With promotion, `r round((m_first_stage$coefficients[1] + m_first_stage$coefficients[2])*100, 1)`% enroll
- Strong first stage (F-statistic > 10): `r round(m_first_stage[["fstatistic"]][["value"]],1)`
:::

::: {.column width="50%"}
```{r, echo=FALSE}
modelsummary(list("Enrollment"= m_first_stage), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') 
```
:::

::::

## 2. Intention-to-Treat (ITT) Effect
:::: {.columns}

::: {.column width="40%"}
```{r, eval=T, echo=TRUE}
#| class: small-code

# Estimate ITT effect (reduced form)
m_itt <- lm_robust(waste_management_costs ~ promotion_zone,
                   clusters = zone_identifier,
                   data = df %>% filter(round == 1))

# With covariate adjustment
m_itt_wcov <- lm_robust(waste_management_costs ~ promotion_zone + 
                        age_manager + age_deputy +
                        female_manager + foreign_owned + 
                        staff_size +
                        advanced_filtration + 
                        facility_area +
                        recycling_center_distance,
                        clusters = zone_identifier,
                        data = df %>% filter(round == 1))
```

**ITT Results**:

- Promotion directly reduces waste management costs by `r round(m_itt$coefficients[2], 0)` units
- Promotion directly reduces waste management costs
- This is the policy-relevant effect of offering promotion
- But this underestimates the effect on those who actually enroll
:::

::: {.column width="60%"}
```{r, echo=FALSE}
modelsummary(list("No covariate adj."= m_itt, "With covariate adj." = m_itt_wcov), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
  kable_styling(font_size = 13) %>% scroll_box(width = "700px", height = "400px")
```
:::

::::

## 3. Local Average Treatment Effect (LATE)
:::: {.columns}

::: {.column width="50%"}
```{r, eval=T, echo=TRUE}
# Estimate LATE using IV regression (2SLS)
m_late <- iv_robust(waste_management_costs ~ enrolled_rp |
                     promotion_zone,
                   clusters = zone_identifier,
                   data = df %>% filter(round == 1))

# With covariate adjustment
m_late_wcov <- iv_robust(waste_management_costs ~ enrolled_rp + 
                         age_manager + age_deputy +
                         female_manager + foreign_owned + 
                         staff_size +
                         advanced_filtration + 
                         facility_area +
                         recycling_center_distance | 
                         promotion_zone + 
                         age_manager + age_deputy +
                         female_manager + foreign_owned + 
                         staff_size +
                         advanced_filtration + facility_area +
                         recycling_center_distance,
                         clusters = zone_identifier,
                         data = df %>% filter(round == 1))
```
:::

::: {.column width="50%"}
```{r, echo=FALSE}
modelsummary(list("No covariate adj."= m_late, "With covariate adj." = m_late_wcov), stars = TRUE,
             gof_map = c("nobs", "r.squared","adj.r.squared"), output = 'kableExtra') %>%
  kable_styling(font_size = 15) %>% scroll_box(width = "600px", height = "300px")
```

**LATE Explanation**:

- Another way of calculating is: LATE = ITT / First Stage = `r round(m_itt$coefficients[2], 0)` / `r round(m_first_stage$coefficients[2], 3)` = `r round(m_itt$coefficients[2]/m_first_stage$coefficients[2], 0)`
- This estimates the effect for "compliers": industries that enrolled because of the promotion but would not have enrolled otherwise

:::

::::
## Understanding LATE: The Complier Effect
The IV estimate (LATE) applies to compliers only:

:::: {.columns}
::: {.column width="50%"}
**Types of industries:**

- Always-takers: Enroll regardless of promotion (8%)
- Compliers: Enroll only with promotion (41%)
- Never-takers: Don't enroll regardless of promotion (51%)
- Defiers: Enroll only without promotion (0% by assumption)
:::

::: {.column width="50%"}
**Implications:**

- If treatment effects vary, LATE may not generalize
- External validity depends on how representative compliers are
- Policy relevant if we care about those influenced by promotion
:::
::::

## Key Assumptions of IV

- Relevance: Promotion effectively increases enrollment (✓ Strong first stage)

- Exclusion restriction: Promotion affects outcomes only through enrollment
  - Promotion must not directly affect waste management costs
  - Careful design focused only on awareness, not waste management behaviors

- Independence: Promotion randomly assigned (✓ Baseline balance)

- Monotonicity: Promotion doesn't discourage anyone from enrolling
  - No "defiers" who would enroll only without promotion

## From Analysis to Policy Decision
**Policy question:** Should the program be scaled up nationally?

**Decision criterion:** Program must reduce waste management costs by at least $10,000.

**Results from IV estimation:**

- LATE: -$9,496 (from our calculation above)
- Point estimate is below the $10,000 threshold
- The IV estimate represents the effect on compliers only

**Recommendation:** Based strictly on the estimate, the program should not be scaled up nationally. However, the estimate is very close to the threshold.

## IV vs. Randomized Assignment
:::: {.columns}
::: {.column width="50%"}
**IV Estimate:**

- LATE = -$9,496
- Effect on compliers only
- Larger standard errors

**Policy recommendation:**
Do not scale up (but close to threshold)
:::
::: {.column width="50%"}
**Randomized Assignment:**

- ATE = -$10,140
- Effect on all eligible facilities
- More precise estimate

**Policy recommendation:**
Scale up nationally
:::
::::
Why the difference?

- Different populations: LATE vs. ATE
- Different precision: IV is less efficient
- Different contexts: National scale-up vs. pilot program

## When to Use Instrumental Variables
Use when:

- Random assignment of the actual treatment is not possible
- You have access to a valid instrument
- Partial compliance is expected
- You focus on the effect for compliers

Common applications:

- Randomized encouragement designs
- Natural experiments (lotteries, policy changes)
- Geographic or administrative discontinuities

## Common Pitfalls in IV Analysis

Weak instruments (First-stage F-statistic < 10):

- Leads to biased estimates and poor inference
- Solution: Find stronger instruments or use weak-IV robust methods

Violation of exclusion restriction:

- Instrument affects outcome through other channels
- Often untestable; requires strong theoretical justification

Heterogeneous treatment effects:

- LATE may not generalize to other populations
- Be careful about policy recommendations

Improper standard errors:

- Remember to cluster standard errors when appropriate
- Use robust methods for inference

## Key Takeaways

- IV addresses selection bias when randomization of treatment isn't possible
- LATE estimates the effect for compliers only
- IV requires strong assumptions, especially exclusion restriction
- The program reduced waste management costs by $9,496 for compliers, slightly below the threshold
- IV estimates are less precise than randomized assignment estimates

## Next Session

Regression Discontinuity Design: Exploiting eligibility thresholds to estimate program effects