---
title: "Introduction to Impact Evaluation"
subtitle: "Econometrics with R"
author: "Dr. Lucas Sempé"
always_allow_html: yes
format: 
  pptx:
    incremental: true  
  revealjs:
    theme: [clean.scss]
    slide-number: true
    code-fold: false
    highlight-style: github
---

## What is Impact Evaluation?

:::: {.columns}

::: {.column width="60%"}
- **Objective**: Measure causal effects of programs or policies
- **Challenge**: Identifying what would have happened without the intervention
- **Counterfactual**: The state of the world that would have occurred in absence of the intervention
- **Impact**: The difference between the actual outcome and the counterfactual
:::

::: {.column width="40%"}
```{r, echo=FALSE, fig.cap="Impact = Actual - Counterfactual", fig.height= 12}
library(broom)
library(estimatr)
library(fishmethods)
library(haven)
library(kableExtra)
library(modelsummary)
library(tidyverse)

# Create example data

data <- data.frame(
  Time = c(0, 1, 0, 1),
  Outcome = c(14.5, 7.8, 14.5, 18),
  Group = c("Treatment", "Treatment", "Counterfactual", "Counterfactual")
)

# Create plot
ggplot(data, aes(x = Time, y = Outcome, color = Group, group = Group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  annotate("segment", x = 1, xend = 1, y = 7.8, yend = 18, 
           arrow = arrow(length = unit(0.3, "cm")), linetype = "dashed") +
  annotate("text", x = 1.05, y = 13, label = "Impact", hjust = 0) +
  scale_x_continuous(breaks = c(0, 1), labels = c("Before", "After")) +
  labs(title = "Visualizing Impact", y = "Outcome", x = "") +
  theme_minimal()
```
:::

::::

## The Fundamental Problem of Causal Inference

> We cannot observe the same unit in both treatment and control states simultaneously.

- For each unit, we observe either:
  - The outcome with the intervention
  - The outcome without the intervention
- The challenge: How do we estimate the counterfactual?

## Example: Waste management system

- Government program subsidizing waste management new technologies 
- Goal: Reduce waste management costs
- Policy question: Should the program be scaled up nationally?
- Decision criterion: Program must reduce waste management at least $10,000

## Key Variables in Our Example Dataset

:::: {.columns}

::: {.column width="50%"}
**Outcome Variable**:

- `waste_management_costs`: Waste management costs incurred monthly per industry (USD)

**Treatment Variables**:

- `treatment_zone`: Area selected for program (0/1)
- `enrolled`: Industry enrolled in program (0/1)
- `eligible`: Industry eligible for program (0/1)
:::

::: {.column width="50%"}
**Other Variables**:

- `round`: Survey round (0=baseline, 1=follow-up)
- `efficiency_index`: Score based on industry characteristics
- Leadership Demographics: Age, education, gender, etc.
- `recycling_center_distance`: Distance to nearest recycling centre
:::

::::

## Overview of Impact Evaluation Methods

**Experimental Methods**:

- Randomized assignment (gold standard)

**Quasi-Experimental Methods**:

- Instrumental Variables
- Regression Discontinuity
- Difference-in-Differences
- Matching Methods

## Naive Approaches to Estimating Impact

1. **Before-After Comparison**
   - Compare outcomes of beneficiaries before and after the program
   - Problem: Cannot distinguish program effects from other changes over time

2. **With-Without Comparison**
   - Compare beneficiaries to non-beneficiaries
   - Problem: Selection bias if those who receive the program are systematically different

## Before-After Comparison (Example)

```{r, echo=TRUE}
# Let's start by uploading our data.
# It is important to know in which folder your dataset is so you can use the right path.

df <- read.csv("./evaluation_data.csv")

# We multiply the outcome variable by 1,000

df$waste_management_costs<-df$waste_management_costs*1000

```

```{r, eval=T, echo=TRUE}
# Compare waste management costs before and after for enrolled industries
m_ba1 <- lm_robust(waste_management_costs   ~ round, 
                  clusters = zone_identifier,
                  data = df %>% dplyr::filter(treatment_zone ==1 & enrolled ==1))

```

**Problems**: 

- External factors also change over time (e.g., economic conditions, other government programs)
- Cannot attribute all changes to the program

## With-Without Comparison (HISP Example)

```{r, eval=T, echo=TRUE}
# Compare enrolled vs. non-enrolled households after program implementation
m_ba2 <- lm_robust(waste_management_costs ~ enrolled, 
                  clusters = zone_identifier,
                  data = df %>% filter(treatment_zone==1 & round ==1))

```

**Problems**:

- Selection bias: Enrolled industries differ from non-enrolled
- Cannot attribute all differences to the program

## Compare results

```{r}

modelsummary(list(m_ba1, m_ba2), stars = TRUE,
             gof_map = c("nobs", "r.squared","r2.adjusted"))


```      

## Course Outline

1. **Introduction to Impact Evaluation** ← *You are here*
2. **Randomized Assignment**
3. **Instrumental Variables**
4. **Regression Discontinuity Designs**
5. **Difference-in-Differences** 
6. **Matching Methods**

## Key Takeaways

- Impact evaluation measures the causal effect of programs or policies
- The central challenge is estimating the counterfactual
- Naive before-after or with-without comparisons are susceptible to bias
- Various experimental and quasi-experimental methods address these challenges
- The example will illustrate these methods throughout the course

## Next Session

**Randomized Assignment**: The gold standard for causal inference

- How randomization creates comparable groups
- Implementing and analyzing randomized evaluations
- Interpreting results with confidence