---
author: "3ie"
always_allow_html: yes
format:
  revealjs:
    theme: [style_c.scss]
    slide-number: true
    width: 1280
    height: 720
    fig-align: center
    slide-level: 2
    embed-resources: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    dev: "png"
    dpi: 96
---

```{r setup}
source("probe_setup.R")

INK <- "#131516"; MUTED <- "#5C6467"; RULE <- "#DCE1E3"
ACCENT <- "#00789E"; ALERT <- "#9A2515"
FONT <- "Roboto, 'Segoe UI', Helvetica, Arial, sans-serif"

af_rate_before  <- 420 / 8.0
af_rate_world_a <- 294 / 5.6

probe_chart <- function(plot) {
  plot %>%
    layout(
      width = 1100, height = 300,
      font = list(family = FONT, size = 13, color = INK),
      paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)",
      margin = list(l = 66, r = 20, t = 36, b = 56),
      xaxis = list(title = list(text = "Difference between the two groups (points)",
                                font = list(size = 12, color = MUTED)),
                   gridcolor = RULE, zerolinecolor = RULE, linecolor = RULE,
                   tickfont = list(size = 12, color = MUTED)),
      yaxis = list(title = list(text = "Number of shuffles",
                                font = list(size = 12, color = MUTED)),
                   gridcolor = RULE, zerolinecolor = RULE,
                   linecolor = RULE, tickfont = list(size = 12, color = MUTED)),
      shapes = list(
        list(type = "line", x0 = 2.1, x1 = 2.1, y0 = 0, y1 = 1, yref = "paper",
             line = list(color = ALERT, width = 3))),
      annotations = list(list(x = 2.1, y = 1, yref = "paper", yanchor = "bottom",
                              showarrow = FALSE, xanchor = "right",
                              text = "what we measured (2.1 points)",
                              font = list(size = 12, color = ALERT)))) %>%
    config(displayModeBar = FALSE)
}
```

# 3 · The five words [39 min]{.mins}

## Word 4 · p-value

::: {.panel}
**Plain language:** roughly, how surprising this result would be if the
programme had truly done nothing at all. A small p-value means "this would be a
strange coincidence if there were no effect."
:::

```{r fig.height=3.125, fig.width=11.5}
plot_ly(p_tbl, x = ~mid, y = ~n, type = "bar",
        marker = list(color = ACCENT, line = list(width = 0)),
        showlegend = FALSE,
        hovertemplate = "A shuffled difference of %{x:.1f} points<br>happened in %{y} of 1,000 shuffles<extra></extra>") %>%
  layout(bargap = 0.15) %>%
  probe_chart()
```

::: {.ask}
We took the 240 schools, broke the link between the programme and the outcome,
and did it 1,000 times. The red tails are the shuffles that beat what we
measured: `r sprintf("%.3f", p_val)`, or about `r round(p_val * 1000)` of 1,000.
:::

## World A · the death rate did not move

```{r results='asis'}
cat(sprintf('
<div class="hero">
  <div class="value">%.0f%%</div>
  <div class="label">Deaths per billion km driven</div>
</div>
<div class="stat-row">
  <div><span class="n">%.0f</span><span class="k">Rate before</span></div>
  <div><span class="n">%.0f</span><span class="k">Rate after</span></div>
  <div><span class="n">0</span><span class="k">Lives saved by the campaign</span></div>
</div>', 100 * (af_rate_before - af_rate_world_a) / af_rate_before,
   af_rate_before, af_rate_world_a))
```

::: {.warn}
If the roads emptied by 30% and deaths fell by 30%, the risk of dying on them
did not change. The campaign saved nobody. The headline was the fuel price.
:::
